### 2021-08-31 main.tex
first

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}

\title{}
\author{}
\institution{}

\begin{document}

\end{document}
% EOF

### 2021-09-13 introduction.tex
update 1

\section{Introduction}
In recent years much work has been done in the exploration of the definition and properties of an
  appropriate generalization of the \emph{univariate} generalized Pareto distribution for exceedences
  over a threshold, to a \emph{multivariate} setting.  For a short accounting of the topic, the work of
  \cite{rootzen2006} defines the generalized Pareto distribution, with further analysis on these classes
  of distributions presented in \cite{falk2008} and \cite{michel2008}.  A recent review of the state
  of the art in multivariate peaks over threshold modelling using generalized Pareto is provided in
  \cite{rootzen2018}.  \cite{ferreira2014} presents a constructive definition of the multivariate Pareto
  that decomposes the random vector into independent radial and angular components; the latter
  containing all and only information pertaining to the dependence structure of the random vector.
  Based on this definition, we present a novel approach for modelling that angular component, and thus
  the modelling the dependence structure of the random vector.

To motivate the topic and our proposed solution, we acknowledge the relevance of extreme analysis to
  climatological events.  Outcomes such as heat waves, extreme precipitation, storm surges, and
  such present a similar challenge to investigators, in that the events of interest can be very destructive,
  and occur in probability at the tail end of \emph{normal} behavior.  In this fashion, a model built
  to capture \emph{normal} behavior may struggle to adequately represent extreme behavior.  More
  specialized tools become necessary; especially so in a time of intensifying extreme weather
  events~\citep{jentsch2007,vousdoukas2018,li2019}.  To this end, we use the integrated water vapor
  transport (\emph{IVT}) data for observing atmospheric rivers as an example application for our proposed method.

\subsection{Extreme Value Theory}
\subsubsection{Univariate Maxima}
Regarding the asymptotic behavior of extreme events, there are a couple major strategies for conducting
  inference---developing probabilistic estimates of extreme behavior.  First developed is the theory
  of a limiting distribution on maxima---the largest observation from a sample.  For a sample $\bm{x}$
  where $\bm{x} = (x_1,\ldots,x_n)$ represents a sequence of $n$ independent random variables from a
  distribution function $F$, the distribution of the maximum $M_n$ of this sequence can be derived
  as:
  \begin{equation*}
    \begin{aligned}
      \text{Pr}\left[M_n\leq z\right] &= \text{Pr}\left[X_1 \leq z, \ldots, X_n \leq z\right]\\
        &= \text{Pr}\left[X_1\leq z\right]\times\ldots\times\text{Pr}\left[X_n\leq z\right]\\
        &= F(z)^n.
    \end{aligned}
  \end{equation*}
  In situations where $F$ is unknown, we can seek to approximate the behavior of $F^n$ as
  $n\rightarrow\infty$.  To ensure this does not degenerate to a point mass, we need to consider a
  standardized sequence of maxima. If this sequence stabilizes as $n$ increases, the a limiting
  distribution exists.  More specifically, if there exists some sequence of constants $a_n > 0$, $b_n$
  such that:
  \begin{equation*}
    \text{Pr}\left[\frac{M_n - b_n}{a_n} \leq z\right] \stackrel{d}{\rightarrow} G(z)
  \end{equation*}
  as $n\rightarrow\infty$, then $G$ is a max-stable distribution, and $F$ is in the domain of
  attraction of that max stable distribution.  \cite{frechet1927} originates the
  field, identifying two limiting forms of max-stable distributions, which would become known as the
  Fr{\'e}chet and Weibull distributions. \cite{weibull1951} expands the analysis of the Weibull
  distribution.  \cite{fisher1928}, in what is now called the \emph{Fisher-Tippett Theorem}, identifies
  the Fr{\'e}chet and Weibull distributions, along with an as then unnamed third form, as the three
  limiting forms of the distribution of the maxima of a sample.  \cite{gumbel1935,gumbel1942} offer
  analysis of that third form, now known as the Gumbel distribution.  Later works, including
  \cite{jenkinson1955} reparameterize all three forms as special cases of a single unifying form,
  the generalized extreme value distribution, \emph{GEV}:
  \begin{equation*}
    \label{eqn:gev}
    F(m \mid \mu, \sigma, \xi) = \exp\left\lbrace-\left[1 + \xi\left(\frac{x - \mu}{\sigma}\right)\right]_{}^{-1/{\xi}}\right\rbrace.
  \end{equation*}
  Thus distributions in the domain of attraction of an EVD (Gumbel, Fr{\'e}chet, or Weibull) will be
  in the domain of attraction of the GEV.

As this distribution specifies asymptotic behavior for the maximum of a set of observations,
  inference assuming this distribution requires we that specify some block of data, and for the block
  report only the maximum.  A series of these blocks yielding a series of maxima allows us to conduct
  inference about the parameters of the distribution.  Taking only the
  maximum in a block of observations necessarily leads to the reduction of our sample size by a factor
  of $1/\text{block size}$. In problems where the data occur in natural blocks, such as an hourly time
  series where a natural block might be a day, this might be appropriate.  There is an implicit
  violation of the assumption of independence within a block, in most cases, but that violation is
  generally ignored. In data without a natural block, it may be difficult to justify an induced
  artitifial block. Moreover, by reducing available information, retaining only one datum for each
  block increases the variability of the parameter estimates.

\subsubsection{Univariate exceedences over a high threshold}
The \emph{Pickands-Balkema-de Haan Theorem}~\citep{balkema1974,pickands1975} offers us another means
  of conducting inference that will be less wasteful of data.  It states that if the original distribution
  $F$ is in the domain of attraction of the GEV, then excesses over a high threshold $u$ can be well
  modelled using the Generalized Pareto distribution.  Again, let $X$ follow some distribution function
  $F$. It follows that:
  \begin{equation*}
    \text{Pr}\left[X > u + y\mid X > u\right] = \frac{1 - F(u + y)}{1 - F(u)}
  \end{equation*}
  for $y > 0$.  If $F$ is in the domain of attraction of an EVD, then
  \begin{equation*}
    \lim\limits_{u\to u^{\prime}}\text{Pr}\left[X > u + y\mid X > u\right] = H(y)
  \end{equation*}
  where $u^{\prime}$ represents some upper boundary of $X$ has a functional form---the survival
  function of a Pareto distribution. We approximate this asymptotic relationship by setting some
  large threshold $u$, and then threshold exceedences $Y = X - u$ are modelled under the distrribution
  function
  \begin{equation*}
    \label{eqn:gp}
    H(y) = 1 - \left(1 + \xi\frac{y}{\sigma}\right)^{-\frac{1}{\xi}}.
  \end{equation*}
  This defines the generalized Pareto family of distributions.  Thus, if block maxima have a
  limiting distribution $G$ within the EVD family, then threshold exceedances for a sufficiently
  high threshold have a limiting distribution $H$ within the Generalized Pareto (GP) family.  We can
  identify $\sigma$ as a scale parameter, but $\xi$ deserves special mention as the extremal index.
  We can interpret the Pareto tail probability as
  \begin{equation*}
    \lim\limits_{t\to\infty}\frac{1 - F(ty)}{1 - F(t)} = y^{-\frac{1}{\xi}}
  \end{equation*}
  for $y > 1$.  Importantly, $\xi$ from the limiting distribution of excesses over a threshold is the
  same parameter, and has the same value as the $\xi$ from the limiting distribution of the maximum
  observation of a sample \citep{dehaan2006}.

One other point of note is that for time series, the data are temporally correlated.  Thus the implicit
  assumption of independence for those observations in excess of a threshold is violated.  One means
  of ameliorating the effect of this correlation is to decluster the data---and one method of declustering
  is, for a series of observations in excess of the threshold, to keep only the maximum observed value
  in the series.

\subsubsection{Multivariate threshold EVT}
For the remainder of this paper, we adopt the operators $\wedge$ to denote minima, and the $\vee$
  to denote maxima.  Thus $\wedge_i x_i = \min_i x_i$, $\vee_i x_i = \max_i x_i$.  These operators can
  also be applied component-wise between vectors, such as $\bm{a}\wedge\bm{b} = (a_1\wedge b_1, a_2\wedge b_2,\ldots)$.

The generalization of EVT to multiple dimensions presents a unique challenge in the case of maxima---how
  do we define the maximum of a sample?  How can we order observations in a multivariate scenario?
  \cite{rootzen2006} offers a brief review of the subject.  One possible treatment sidesteps this
  conundrum by using component maxima---let $\bm{M}_n = \left(\vee_i X_{i1},\ldots,\vee_i X_{id}\right)$.
  Then the multivariate GEV establishes the limiting behavior of $\bm{M}_n$ as $n\to\infty$.  Note
  that $\bm{M}_n$ need not be in the observed $\left\lbrace\bm{X}_i; i = 1,\ldots,n\right\rbrace$.
  Then, if there is a set of constants $\bm{a}_n > \bm{0}$, $\bm{b}_n$ such that
  \begin{equation*}
    \lim\limits_{n\to\infty}\text{P}\left(\bigcap_{\ell = 1}^d
      \frac{M_{n\ell} - b_{n\ell}}{a_{n\ell}} \leq x_{\ell}\right) = G(\bm{x})
  \end{equation*}
  with non-degenerate marginals, then $G(\bm{x})$ is a \emph{multivariate extreme value distribution}.
  The marginal distribution of $M_{nl}$ is by construction a GEV with the parameters $a_{n\ell}, b_{n\ell}, \xi$.
  With the existence of the marginal distributions established as a consequence of the existence of the
  GEV, multivariate EVT then splits into two components---estimation of the marginal parameters, and
  estimation of the dependence structure.  This separation is born out extensively in the literature.

As in the multivariate extrema case, analysis in peaks over thresholds modelling generally first develops
  the marginal distributions, then establishes the dependence structure.  In extreme analysis, a frequently
  used method for describing the dependence
  structure is the copula---a family of multivariate cumulative distribution functions with uniform
  marginal distributions~\cite{renard2007,deng2011,falk2019},
  but this presents its own difficulties in modelling and interpretation.  Instead, we propose a novel
  method based on the constructive definition of the multivariate Pareto presented in
  \cite{ferreira2014}, where we assume that the $d$-dimensional vector $\bm{Z}$, after standardization
  is such that for some distribution $\mu$ defined on $\mathcal{R}_+^d$,
  \begin{equation}
    \lim\limits_{n\to\infty}n\text{Pr}\left[\frac{1}{n}\bm{Z}\geq \bm{z}\right] = \mu\left([\bm{0},\bm{z}]^c\right).
  \end{equation}
  In this setting, $\mu$ is the asymptotic distribution of $\bm{Z}$ in extreme regions---the
  \emph{exponent measure}.  $\mu$ features the homogeneity property $\mu(tA) = \frac{1}{t}\mu(A)$.
  By this property, \cite{ferreira2014} factorize $\bm{Z}\to R\bm{V}$, where
  $R = \pnorm{\bm{Z}}{\infty} > 1$ describes the radial component, and
  $\bm{V} = \bm{Z}/\pnorm{\bm{Z}}{\infty} \in \mathcal{S}_{\infty}^{d-1}$ the angular component of
  $\bm{Z}$ in $\mathcal{R}_+^d$.  $\mu(\cdot)$ is similarly factorized as
  \begin{equation}
    \mu\left( \left\lbrace\bm{Z} : R(\bm{Z}) > r, V(\bm{Z}) \in A \right\rbrace \right) = r^{-1}\Phi(A),
  \end{equation}
  where $\Phi(\cdot)$ is the \emph{spectral measure}.  The spectral measure is thus independent and
  separable from the radial component.  Furthermore,
  \begin{equation}
    \text{P}\left[\bm{V} \in A \mid R > r\right]
      = \frac{r\text{P}\left[\bm{V} \in A, R > r\right]}{r\text{P}[R > r]}\hspace{0.2cm}
      \xrightarrow[r\to\infty]{~} \hspace{0.2cm} \frac{\Phi(A)}{\Phi(\mathcal{S}_{\infty}^{d-1})}
  \end{equation}
  Describing the dependence structure of $\bm{Z}$, equivalently describes the spectral measure
  $\Phi$. To accomplish this, we seek to establish a distribution on $\bm{V} \in \mathcal{S}_{\infty}^{d-1}$.

\subsection{Integrated Vapor Transport}
Atmospheric rivers are temporary events, where large elongated regions of high concentrations of
  water vapor are developed in the atmosphere and carry huge amounts of water potentially thousands
  of miles.  The amount of water in transit during these events can dwarf that of terrestrial rivers.
  For the targeted region, the atmospheric river can represent a significant portion of the
  precipitation the region will experience.  Thus, such events are of great interest to meteorologists,
  hydrologists, actuaries, land-use planners, and even farmers.  The ability to estimate the likely magnitude
  and distribution of such events is thus of great importance~\citep{ralph2013,ralph2018}.

One metric by which we might identify and declare atmospheric rivers is the integrated water vapor
  transport, or \emph{IVT}.  This value represents the total amount of water
  vapor being transported in an atmospheric column---that is, a column of the troposphere above a given
  area of the surface of the Earth. Raw data are measured by dropsondes---measurement devices dropped
  out of aircraft, taking measurements as they fall to the earth.~\citep{ralph2017}  A data vector
  includes an estimate of IVT at grid cell at a period in time; where a grid cell specifies the surface
  of the earth under the atmospheric column.  We have two such datasets, in differing spatial
  resolutions---the lower resolution data splits the coast of california into 8 grid cells, while the
  higher resolution does so into 47 grid cells~\citep{guan2015}.

As IVT tracks water vapor in the atmosphere, extreme values in IVT can lead to
  extreme precipitation events.  Atmospheric rivers are necessary events, in that they carry the
  bulk of precipitation that California recieves, but extreme precipitation events can be extremely
  destructive as well.  Thus, we are interested in the distribution of extreme precipitation events,
  including their depenedence structure.  This tells us about the probability of jointly extreme
  events---where IVT is extreme in two or more grid cells, as well as identifying particular
  configurations of IVT that may be anomalous.  In California, groundwater is a managed resource.
  Extreme precipitation events significantly affect groundwater, so anomalous precipitation events
  can produce anomalous configurations of groundwater, leading to destructive flooding.

We are specifically interested in extremal dependence---the relationship between the upper tails of
  dimensions of the distribution.  In this case, that means the relationship between extreme values
  in different grid cells.  We are going to be looking at point-in-time behavior rather than
  considering the time series nature of the data---that relationship may come later.

As we are interested in the extremal dependence, it makes sense that we would choose to represent
  this data using tools from extreme value theory.  Extreme value theory, or \emph{EVT}, seeks to
  model and assess probability of observing extreme events.  Such a topic is applicable generally,
  but it finds particularly strong use among such fields as finance \citep{allen2013},
  climatology \citep{trepanier2018}, and insurance \citep{beirlant1994}.  In these fields, extreme
  events may represent significant loss to the body commissioning the study.  For instance, an
  insurance company might commission a study on extreme weather events, as an extreme weather event
  localized to a particular region could cause a spike in claims from that region.  Extreme value
  theory offers us a tool set for making inference about the tails of a distribution, without having
  observed said tails.  For
  instance, with an extreme weather event like flooding, we can make predictions about return
  levels---the average time until an observation of a particular magnitude occurs---without having seen
  an observation of that magnitude, or having observed that long.  In the context of our motivating
  example, extreme values in the IVT may represent the formation of an atmospheric river, which has
  dramatic effects on precipitation and ground water.  The formation of an IVT may lead to flooding
  and other negative effects, but is also contributes necessary water for irrigation and agriculture.
  Development of statistical tools regarding these extreme events becomes necessary for making informed
  decisions.

  % EOF

### 2021-09-13 methodology_pnorm.tex
update 1

\subsection{Projection onto an arbitrary unit hypersphere}

\begin{figure}[ht]
    \centering
    \includegraphics{images/p_sphere}
    \caption{Unit hyperspheres established on $\mathcal{L}_p$, for $p = 1,\ldots,10$, for two dimensions.\label{fig:psphere}}
\end{figure}

A hypersphere is a geometric object such that the distance from any point to the center takes a fixed,
  constant value.  The unit hypersphere is a hypersphere where that distance is 1. We can define the
  hypersphere under an arbitrary distance measurement, but for our purpose we will use the class of
  hyperspheres defined under the $\mathcal{L}_p$ norm. Let the $\mathcal{L}_p$-norm be defined as
  \begin{equation*}
    \lVert \bm{s} \rVert_p = \left({\textstyle\sum}_{\ell = 1}^d \lvert s_{\ell}\rvert^p\right)^{\frac{1}{p}}.
  \end{equation*}
  From this, we establish the $\mathcal{L}_1$ norm as $p = 1$, or the absolute sum; the
  $\mathcal{L}_2$ norm, as $p = 2$, the Euclidean distance.  From this we also establish the
  $\mathcal{L}_{\infty}$ norm, as
  \begin{equation*}
    \lVert \bm{s} \rVert_{\infty}
      = \lim\limits_{p\to\infty} \lVert \bm{s} \rVert_p
      = \max_{\ell\in\lbrace1,\ldots,d\rbrace}s_{\ell}.
  \end{equation*}
  We are interested in the direction, or angular component, of vectors described in the positive
  orthant, $\mathcal{R}_{+}^d$.  One means of describing the direction of a vector in $\mathcal{R}_+^d$
  is to project that vector onto $\mathcal{S}_{p}^{d-1}$, the positive orthant of the unit hypersphere
  defined on an $\mathcal{L}_p$-norm, denoted as $\mathcal{S}_{p}^{d-1}$.  That is,
  \begin{equation*}
    \mathcal{S}_{p}^{d-1} = \left\lbrace \bm{s} : \bm{s} \in \mathcal{R}_{+}^{d}, \lVert \bm{s}\rVert_{p} = 1\right\rbrace.
  \end{equation*}
  Figure~\ref{fig:psphere} shows $\mathcal{S}_{p}^{1}$ for $p$ at integers between 1 and 10.
  At $p = 1$, we have the simplex.  As $p$ grows larger, the surface moves away from the origin and
  approaches closer to the unit hypercube--the limiting form of the surface. We project an observation
  onto $\mathcal{S}_p^{d-1}$ by dividing said observation by its $p$-norm--let
  $\bm{x}\in \mathcal{R}_{+}^{d}$, then $\bm{y} = \bm{x} / \lVert \bm{x}\rVert_p \in \mathcal{S}_{p}^{d-1}$.
  We denote the $d-1$ to indicate the loss of one degree of freedom relative to the original vector.
  Observations on one hypersphere can be projected without loss of information onto another by
  dividing by the defining norm of the target hypersphere.

Assuming $\bm{y} \in \mathcal{S}_{p}^{d-1}$, then for finite $p$, $y_d$ can always be represented
  as a function of the other dimensions.  That is,
  \begin{equation*}
    y_d = \left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}.
  \end{equation*}
  So the transformation
  \begin{equation}
    \label{eqn:pnormt}
    T(x_1,\ldots,x_d) = \left(\pnorm{\bm{x}}{p}, \frac{x_1}{\pnorm{\bm{x}}{p}},
                          \ldots , \frac{x_{d-1}}{\pnorm{\bm{x}}{p}}\right) = (r,y_1,\ldots,y_{d-1})
  \end{equation}
  does not lose any information.  The reverse of this transformation,
  \begin{equation}
    \label{eqn:pnormtinv}
    T^{-1}\left(r,y_1,\ldots,y_{d-1}\right) =
      \left(ry_1,\ldots,ry_{d-1}, r\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}\right)
  \end{equation}
  equivalently recovers the original data.  To recover the density of the transformed random variables,
  the original density is multiplied by the determinant of the Jacobian--the matrix of derivatives
  of the inverse transformation.  This takes the form:
  \begin{equation}
    \label{eqn:pnormjac}
    r^{d-1}\left[\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}} +
        {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(1 - {\textstyle\sum}_{l=1}^{d-1} y_{\ell}^p\right)^{\frac{1}{p} - 1}\right].
  \end{equation}
  Notice a factor of $r^{d-1}$ independent of $p$. We refer to $\bm{y}$ and $r$ as, respectively,
  the angular and radial components of $\bm{x}$.  If we assume a distribution for $\bm{x}$, then
  by transforming to $r, \bm{y}$ and integrating out $r$, we are left with a distribution on solely
  $\bm{y}$, the projection of $\bm{x}$ onto $\mathcal{S}_{p}^{d-1}$.  If we take the projection to
  its limit as $p\to\infty$, we arrive at a projection onto $\mathcal{S}_{\infty}^{d-1}$.

We can observe a problem at this point in at least two ways.  First, on $\mathcal{S}_{\infty}^{d-1}$, the
  inverse transformation from $\mathcal{S}_{\infty}^{d-1}$ to $R_+^{d}$ in Equation~\ref{eqn:pnormtinv}
  may not recover the original data--that formulation of the projection would only work when $Y_d = 1$.
  Alternatively, we may look at the formula for the determinant of the Jacobian in
  Equation~\ref{eqn:pnormjac}.  In that formula we observe the value
  $(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p)^{\frac{1}{p} - 1}$.  As $p\to\infty$,
  this quantity will go to $0^{-1}$ if $y_d \neq 1$.
  So, the mathematical basis of this transformation is thus dependent upon finite $p$.  However, we can
  project onto $\mathcal{S}_p^{d-1}$ for large $p$, and that projection will be similar to an idealized
  distribution on $\mathcal{S}_{\infty}^{d-1}$.

% EOF

### 2021-09-13 methodology_projgamma.tex
update 1

\subsection{Projected Gamma Family}
As stated previously, one approach we might try in order to generate a distribution on
  $\mathcal{S}_{p}^{d-1}$ would be to take a distribution in $\mathcal{R}_{+}^d$, reparameterize
  it in terms of radial and angular components, and integrate out the radial component in order to
  effectively project the distribution onto $\mathcal{S}_{p}^{d-1}$.  A natural candidate distribution
  to accomplish this is the Gamma.  Let $\bm{ X} \sim \prod_{l = 1}^d\text{Ga}\left(X_l\mid\alpha_l,\beta_l\right)$.
  Then transform $\bm{ X}$ via the transformation in Eqns.~\ref{eqn:pnormt} into $R,\bm{ Y}$, where
  $R = \pnorm{\bm{ X}}{p}$ and $Y\in \mathcal{S}_{p}^{d-1}$.  Completing the transformation, we
  incorporate the determinant of the Jacobian from Eqn.~\ref{eqn:pnormjac}.  Thus we have
  \begin{equation}
    \label{eqn:protopg}
    f(r,\bm{ y}) = \prod_{l = 1}^{d}
      \left[\frac{\beta_l^{\alpha_l}}{\Gamma(\alpha_l)}(ry_l)^{\alpha_l - 1}\exp\lbrace-\beta_lry_l\rbrace\right]
      \times r^{d-1}\left[y_d - {\textstyle \sum}_{l = 1}^{d-1}y_l^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right],
  \end{equation}
  where $y_d = \left(1 - \sum_{l=1}^{d-1}y_l^p\right)^{\frac{1}{p}}$. Integrating out $r$ yields the
  the resulting \emph{Projected Gamma} density
  \begin{equation}
    \label{eqn:projgamma}
    \text{PG}(\bm{ y}\mid\bm{ \alpha},\bm{ \beta}) = \prod_{l = 1}^d\left[\frac{\beta_l^{\alpha_l}}{\Gamma(\alpha_l)}y_l^{\alpha_l - 1}\right]
      \times \left[y_d - {\textstyle \sum}_{l = 1}^{d-1}y_l^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right]
      \times \frac{\Gamma({\textstyle\sum}_{l = 1}^d\alpha_l)}{\left({\textstyle\sum}_{l = 1}^d \beta_ly_l\right)^{{\scriptstyle\sum_{l = 1}^d \alpha_l}}}
  \end{equation}
  established for an arbitrary $\mathcal{L}_p$ norm. This forms a valid density for all finite $p$.
  We can regard the Dirichlet distribution as a special case of this density, where
  $\bm{ y}\in \mathcal{S}_1^{d-1}$, and $\beta_l = \beta$ for all $l$.  The distribution described
  in \cite{nunez2019} is related to this distribution in that it is established on
  $\bm{ y}\in \mathcal{S}_2^{d-1}$, transformed to spherical coordinates via Eqn.~\ref{eqn:spherical}.
  The full conditional for $r$ takes the form
  \begin{equation}
    \label{eqn:rfullcond}
    r\mid\bm{ \alpha},\bm{ \beta}, y \sim \text{Ga}\left(r\mid{\textstyle\sum}_{l = 1}^d \alpha_l, {\textstyle\sum}_{l = 1}^d \beta_ly_l\right).
  \end{equation}

In the estimation of the parameters of this model, we can quickly arrive at a problem, in that
  without some linear restriction on the rate parameters $\beta_l$, the model is not identifiable.
  The restriction most often used in the case of the Generalized Dirichlet, as well as by
  \cite{nunez2019} is to set $\beta_1 := 1$.

\subsubsection{Inference on Projected Gamma Model}
We consider a basic model that assumes all observations are generated by a single projected Gamma
  distribution.  The shape and rate parameters for the latent \text{Gamma} random variables,
  $\alpha_l,\beta_l$, must be greater than 0, so a logical prior to use would be a Gamma prior for
  each parameter, excepting for $\beta_1 := 1$.
  \begin{equation}
    \label{eqn:vanillap}
    \begin{aligned}
      \bm{ y}\mid\alpha,\beta &\sim \text{PG}(\bm{ y}\mid\alpha,\beta)\\
      \bm{ \alpha},\bm{\beta} &\sim {\textstyle \prod}_{l = 1}^d \text{Ga}(\alpha_l \mid \xi_l,\tau_l)
              \times {\textstyle \prod}_{l = 2}^d \text{Ga}(\beta_l\mid \zeta_l,\sigma_l).
    \end{aligned}
  \end{equation}
  A useful feature of the Projected Gamma family is the latent independent Gammas assumption.  Recall
  in Eqn.~\ref{eqn:protopg}, we have a density defined as a product of independent Gamma densities,
  multiplied by a factor of $r^{d-1}$ and some value, constant with respect to the parameters
  $\alpha,\beta$.  If we conduct inference via a Gibbs sampler, then by simulating this latent $r$
  we are able to perform inference of $(\alpha_l,\beta_l)$ independent of that of $(\alpha_k,\beta_k)$.
  The full conditional for $r$ is described in Eqn.~\ref{eqn:rfullcond}.  The full conditional for
  $\beta_l$, $l = 2,\ldots,d$ then becomes
  \begin{equation}
    \label{eqn:betafc}
    \beta_l\mid \bm{ y}, r, \alpha_l \sim \text{Ga}\left(n + \alpha_l + \zeta_l,
                                      {\textstyle \sum}_{i = 1}^nr_iy_{il} + \sigma_l\right).
  \end{equation}
  That is, given $\alpha_l$, $r$, we can sample $\beta_l$ directly as a Gamma random variable.  In
  the sampling for $\alpha_l$, we can integrate $\beta_l$ out leaving us a quantity proportional
  to a density, which we can sample via the Metropolis Hastings algorithm\citep{hastings1970}.  For $\alpha_l$,
  $l = 2,\ldots, d$, This takes the form
  \begin{equation}
    \label{eqn:alphafc}
    f(\alpha_l \mid \bm{ y}, r) \propto
    \frac{\left({\textstyle \prod}_{i = 1}^nr_iy_{il}\right)^{\alpha_l - 1}}{\Gamma^n(\alpha_l)} \times
    \alpha_l^{\xi_l - 1}\exp\{-\tau_l\alpha_l\} \times
    \frac{\Gamma(n\alpha_l + \zeta_l)}{
        \left({\textstyle\sum}_{i = 1}^n r_iy_{il} + \sigma_l\right)^{(n * \alpha_l + \zeta_l)}}.
  \end{equation}
  For $\alpha_1$, as we do not integrate out $\beta_1$ as it is defined to be $1$, this takes the
  slightly simpler form
  \begin{equation}
    \label{eqn:alphafc1}
    f(\alpha_1 \mid \bm{ y}, r) \propto
      \frac{{\textstyle\prod}_{i = 1}^n ry_{i1}^{\alpha_1 - 1}}{\Gamma^n(\alpha_1)} \times
      \alpha_1^{\xi_1 - 1}\exp\{-\tau_1\alpha_1\}.
  \end{equation}
  The \emph{projected restricted Gamma} model sets $\beta_l := 1 \forall l \in \lbrace 1,\ldots,d\rbrace$,
  so in that model inference on all $\alpha_l$ will use the simpler form.

\subsubsection{Finite Mixture Model}
The projected Gamma distribution offers a flexible model for describing direction.  However, it is
  conceivable that a single Projected Gamma distribution will not sufficiently capture the
  variability of real data.  In that case, a logical extension to make to the above model is to
  assume the data, $\bm{ y}$, are generated from a mixture of projected Gammas.  We consider a finite
  mixture,
  \begin{equation}
    \label{eqn:fmixp}
    \begin{aligned}
      \bm{ y}_i &\sim \sum_{j = 1}^J\pi_j\text{PG}\left(\bm{ y}\mid \bm{ \alpha}_j, \bm{ \beta}_j\right)\\
      \bm{ \alpha}_j &\sim {\textstyle \prod}_{l = 1}^d \text{Ga}\left(\alpha_l\mid\xi_l,\tau_l\right)\\
      \bm{ \beta}_j &\sim {\textstyle \prod}_{l = 2}^d \text{Ga}\left(\beta_l\mid\zeta_l,\sigma_l\right)
    \end{aligned}
    \hspace{2cm}
    \begin{aligned}
      \bm{ \xi},\bm{\tau} &\sim {\textstyle \prod}_{l = 1}^d \text{Ga}(\xi_l\mid a,b) \times \text{Ga}(\tau_l\mid c,d)\\
      \bm{ \zeta},\bm{\sigma} &\sim {\textstyle\prod}_{l = 2}^d\text{Ga}(\zeta_l \mid a,b)\times \text{Ga}(\sigma_l\mid c,d)\\
      \bm{ \pi} &\sim \text{Dir}(\pi_0)
    \end{aligned}
  \end{equation}
  In addition to allowing multiple mixture components, we assume those mixture components themselves
  descend from a product of independent Gamma densities. The standard method for sampling from a
  mixture distribution includes a data augmentation step, generating a
  $\delta_i \in \lbrace 1,\ldots, J\rbrace$ indicating to which mixture component a
  particular observation comes from, with
  \begin{equation*}
    \text{P}\left[\delta_i = j\mid\alpha_j,\beta_j\right] \propto \pi_j\text{PG}\left(\bm{ y}_i\mid \bm{ \alpha}_j,\bm{ \beta}_j\right).
  \end{equation*}
  By sampling and conditioning on $\delta_i$, we end up with very similar semi-conjugate forms for
  inference of $\bm{ \alpha}_j$, $\bm{ \beta}_j$ that were used in the basic model.  The changes made
  to Equations~\ref{eqn:betafc},\ref{eqn:alphafc},\ref{eqn:alphafc1} are such that we're conducting inference
  on a mixture component given the mixture component identifiers $\bm{ \delta}$, so instead of indexing
  $i$ over the whole data set, we're indexing $i$ over the set $\{i :\delta_i = j\}$, with $n = n_j$,
  the cardinality of that set. Inference on $\bm{ \xi,\tau,\zeta,\sigma}$ follows a similar form as
  that of $\bm{ \alpha,\beta}$---they are Gamma-Gamma relationships, so rate parameters have a conjugate
  Gamma posterior, and we can integrate them out to perform inference on the shape parameters via
  Metropolis Hastings.

Additionally, we investigate a finite mixture extension of the \emph{Projected Restricted Gamma} model,
  where we define all $\beta_l := 1$, for $l = 1,\ldots,d$, for $j = 1,\ldots,J$.  This affects the above
  model by removing inference steps for $\bm{ \beta},\bm{ \zeta},\bm{ \sigma}$.  The inference for
  $\alpha_l$ for all $l = 1,\ldots,d$ follows the form in Equation~\ref{eqn:alphafc1}; inference on
  $\bm{ \xi,\tau}$ follows the same form as the unrestricted model.  The impetus for the restricted
  gamma model is that in the context of a mixture model, the additional flexibility offered by the unrestricted Gamma model may be better represented with additional clusters.

\subsubsection{Dirichlet Process Mixture Model}
A logical extension to the finite mixture model described in Equation~\ref{eqn:fmixp} would be to assume the
  $\bm{ y}$'s descending from an infinite mixture model.  This to some extent solves the issue of
  selecting the number of mixture components, as that will effectively be done so in the model.
  \begin{equation}
    \label{eqn:infmixp}
    \begin{aligned}
      \bm{ y}_i &\sim \sum_{j = 1}^J\pi_j\text{PG}\left(\bm{ y}\mid \bm{ \alpha}_i, \bm{ \beta}_i\right)\\
      (\alpha_i,\beta_i) &\sim \text{DP}\left(\eta, G_0\right)\\
      &~\hspace{-2cm}G_0 = {\textstyle\prod}_{l = 1}^d\text{Ga}\left(\alpha_{jl}\mid\xi_{l},\tau_{l}\right)
                    \times{\textstyle\prod}_{l=2}^d\text{Ga}\left(\beta_{jl}\mid\zeta_{l}\tau_{l}\right)
    \end{aligned}
    \hspace{1cm}
    \begin{aligned}
      \bm{ \xi},\bm{ \tau} &\sim {\textstyle \prod}_{l = 1}^d \text{Ga}(\xi_l\mid a,b)\times \text{Ga}(\tau_l\mid c,d)\\
      \bm{ \zeta},\bm{\sigma} &\sim {\textstyle\prod}_{l = 2}^d\text{Ga}(\zeta_l \mid a,b)\times \text{Ga}(\sigma_l\mid c,d)\\
      \bm{ \eta} &\sim \text{Ga}(\eta \mid 2, 0.1).
    \end{aligned}
  \end{equation}
  To fit the Dirichlet process, we use a variant of the collapsed Gibbs sampler detailed as Algorithm~8 in \citet{neal2000}.  In this algorithm, we perform a data augmentation to create a cluster identifier, $\delta_i$, for each observation. Let $J$ indicate the number of extant clusters.  Under the collapsed sampler scheme, for existing clusters, the probability of cluster membership is calculated as
  \begin{equation}
    \label{eqn:dpdelta}
    \text{P}\left(\delta_i = j\mid r, \bm{ y_i} \bm{ \alpha}, \bm{ \beta}\right) \propto
        \frac{n_j}{\sum_{k = 1}^{J} n_k + \eta}
        {\textstyle\prod}_{l = 1}^d\text{Ga}\left(r_iy_{il}\mid\alpha_{jl},\beta_{jl}\right).
  \end{equation}
  where $n_j = {\textstyle\sum}_{k\neq i}1_{k == j}$.  Effectively, if the observation under review is a
  singleton, its existing cluster is destroyed.  For new clusters, ostensibly we would integrate out
  cluster parameters to achieve a posterior predictive density; though doing so in a
  non-conjugate model may be difficult.  Instead, the impetus for using this specific DP fitting algorithm
  is it simulates the posterior predictive density by generating $m$ new candidate clusters from their respective priors given
  $\bm{ \xi,\tau,\zeta,\sigma}$.  Then the probability of $y_i$ belonging to a particular candidate
  cluster is calculated as
  \begin{equation}
    \label{eqn:dpdeltanew}
    \text{P}\left[\delta_i = k\mid r, \bm{ \alpha}_k^{\prime}, \bm{ \beta}_k^{\prime}\right] \propto
        \frac{\eta / m}{\sum_{j = 1}^{J} n_j + \eta}
        {\textstyle\prod}_{l = 1}^d\text{Ga}\left(r_iy_{il}\mid\alpha_{kl}^{\prime},\beta_{kl}^{\prime}\right),
  \end{equation}
  where $(\cdot)^{\prime}$ indicates a value as belonging to a new candidate cluster, and $k$ indexes along candidate clusters.  If a new cluster is selected, then $(\bm{\alpha}_{J+1},\bm{\beta}_{J+1}) = (\bm{\alpha}_k^{\prime},\bm{\beta}_k^{\prime})$, and $J = J + 1$.

Inference upon $\bm{ \alpha}_j,\bm{ \beta}_j$ operates in the same manner as with the finite mixture
  model.  For $\bm{ \xi,\tau,\zeta,\sigma}$, we have $J$ cluster components, indicating $J$
  observations that again are under a gamma-gamma model.  The rate parameters have a previously
  described conjugate posterior, that we can integrate out to more directly sample the shape
  parameters.  As before, we will be looking at this model, and the restricted gamma form of this model, where
  $\beta_l := 1$ for $l = 1,\ldots,d$.  This removes inference steps for $\bm{ \zeta},\bm{ \sigma}$
  as well.

\subsubsection{A Log-Normal Prior for Shape Parameters}
A final extension we can make to this model is to allow for some structure, or dependence, in the prior
  for shape parameters within a cluster.  The impetus for this approach is from recognizing that the
  algorithm we are using for sampling the DP may not adequately represent the posterior density for
  a new cluster, as it is assuming sampling from an independent prior.  If we account for dependence
  between dimensions within the prior, then we may be able to more efficiently sample new clusters.
  As shape parameters must be greater than 0, it is logical to assume a log normal prior, as that
  allows us the familiar forms of a mean vector and covariance matrix for representing the source of
  cluster parameters. So,
  \begin{equation}
    \label{eqn:dpln}
    \begin{aligned}
      \bm{ y}_i &\sim \sum_{j = 1}^J\pi_j\text{PG}\left(\bm{ y}\mid \bm{ \alpha}_i, \bm{\beta}_i\right)\\
      (\bm{\alpha}_i, \bm{\beta}_i) &\sim \text{DP}\left(\eta, G_0\right)\\
        &~\hspace{-1cm}G_0 = \mathcal{N}\left(\log\bm{ \alpha}_{j}\mid\mu,\Sigma\right)\times
            {\textstyle\prod}_{l = 2}^d\text{Ga}\left(\beta_{jl}\mid\zeta_l,\sigma_l\right)
    \end{aligned}
    \hspace{1cm}
    \begin{aligned}
      \mu,\Sigma &\sim \mathcal{N}\left(\mu\mid\mu_0,\Sigma_0\right) \times \text{IG}\left(\nu,\Psi\right)\\
      \bm{ \zeta},\bm{\sigma} &\sim {\textstyle\prod}_{l = 2}^d\text{Ga}(\zeta_l \mid a,b) \times \text{Ga}(\sigma_l \mid c,d) \\
      \bm{ \eta} &\sim \text{Ga}(\eta \mid 2, 0.1).
    \end{aligned}
  \end{equation}
  We hope that this will allow us to better incorporate the relationship between dimensions in
  sampling new clusters, allowing for better mixing of the model.  The downside is one of computational
  complexity---this introduces a $d$-dimensional normal distribution into the model, requiring a
  $d$-dimensional matrix inversion step.  As matrix inversion is computationally costly, this limits
  the potential number of dimensions possible.  The posterior inference strategies for $\beta,\zeta,\sigma$
  are the same as for prior models and have been described.  For inference on $\alpha_j$ given
  $r, \mu,\Sigma$, we rely on a $d$-dimensional Metropolis Hastings step, where the log-posterior
  density is proportional to
  \begin{equation}
    \label{eqn:dplnlp}
    \begin{aligned}
    f(\bm{\alpha}_j\mid\bm{\delta},\bm{y},\mu,\Sigma,\zeta,\sigma) &\propto
    \frac{\left(\prod_{i:\delta_i = j}y_{il}\right)^{\alpha_{j1} - 1}}{\Gamma(\alpha_{j1})^{n_j}}
        \prod_{l = 2}^d\left[\frac{\left(\prod_{i:\delta_i = j}y_{il}\right)^{\alpha_{jl} - 1}
        \Gamma(n_j\alpha_{jl} + \zeta)}{\Gamma(\alpha_{jl})^{n_j}\left(\sum_{i:\delta_i = j}y_{il}
        + \sigma\right)^{n_j\alpha_{jl} + \zeta}}\right]\\
        &\hspace{1cm}\times
        \lvert \Sigma\rvert^{-\frac{1}{2}}\prod_{l=1}^d\alpha_{jl}^{-1}\exp\left\lbrace
          -\frac{1}{2}(\log\alpha_j - \mu)^T\Sigma^{-1}(\log\alpha_j - \mu)\right\rbrace.
    \end{aligned}
  \end{equation}
  Updating $\beta_j$ is accomplished with a conjugate Gibbs step, using the full conditional detailed
  in Equation~\ref{eqn:betafc}, again subsetting $i$ such that $\delta_i = j$.  Note that we offer this
  model for completeness---in practice, attempting to learn $\beta$ with $\alpha$ having a log-normal
  prior resulted in extremely poor model performance.  We thus used the restricted gamma likelihood,
  assuming $\beta_{jl} = 1$ for all $j,l$.  This simplifies Equation~\ref{eqn:dplnlp} into
  \begin{equation}
    \label{eqn:dplnlpr}
    \begin{aligned}
    f(\bm{\alpha}_j\mid\bm{\delta},\bm{y},\mu,\Sigma) &\propto
    \prod_{l = 1}^d\left[\frac{\left(\prod_{i:\gamma_i = j}y_{il}\right)^{\alpha_{jl} - 1}}{\alpha_{jl}\Gamma(\alpha_{jl})^{n_j}} \right]\times\lvert\Sigma\rvert^{-\frac{1}{2}}
      \exp\left\lbrace-\frac{1}{2}(\log\alpha_j - \mu)^T\Sigma^{-1}(\log\alpha_j - \mu)\right\rbrace.
    \end{aligned}
  \end{equation}
  This model is conjugate for $\mu,\Sigma$, allowing simple Gibbs step updating.  The full
    conditional distributions for $\mu$, $\Sigma$ take the form
  \begin{equation}
    \begin{aligned}
    \mu\mid\Sigma,\alpha &\sim \mathcal{N}\left((n^*\Sigma^{-1} + S^{-1})^{-1}(n^*\overline{\log\alpha}^T\Sigma^{-1}
      + \mu_0^TS^{-1}),(n^{*}\Sigma^{-1} + S^{-1})^{-1}\right)\\
    \Sigma\mid\mu,\alpha &\sim \mathcal{IW}\left(
        n^* + \nu, {\textstyle\sum}_{j = 1}^{n^*}(\log\alpha_j - \mu)(\log\alpha_j - \mu)^T + \Psi\right),
    \end{aligned}
  \end{equation}
  where $n^*$ specifies the number of extant clusters.  Removing $\beta$ additionally removes the
  inference steps for $\zeta,\sigma$.
  %
  % \bruno{\bf I think that 3.2.2--3.2.4 are pretty much the same model with some elaborations.
  %               You should be able to present the whole story in a much more condensed fashion.}

\subsubsection{Choice of Norm}
In the extreme value case, we have defined our task on $\mathcal{S}_{\infty}^{d-1}$, the positive orthant
  of the hypersphere defined under the $\mathcal{L}_{\infty}$ norm.  We previously described the
  difficulty in establishing a density directly on this geometry.  However, we can build the projected
  gamma distribution on $\mathcal{S}_p^{d-1}$ for a large $p$.  We can then project samples from the
  posterior predictive distribution for a model built on $\mathcal{S}_p^{d-1}$ onto
  $\mathcal{S}_{\infty}^{d-1}$. As $p$ increases, the closer the manifold will be to $\mathcal{S}_{\infty}^{d-1}$, and thus the closer the projected distribution will be to an idealized
  projected Gamma built on $\mathcal{S}_{\infty}^{d-1}$.  In testing, we tried developing models on
  $S_p^{d-1}$ for varying levels of $p$, and found that higher $p$ produced higher performing models.
  Additionally, the difference in model performance became more apparent as dimensionality increased.
  We present results here for models developed on $\mathcal{S}_{10}^{d-1}$.

% EOF

### 2021-09-13 methodology_projgamma.tex
update 2

@@
   effectively project the distribution onto $\mathcal{S}_{p}^{d-1}$.  A natural candidate distribution
-  to accomplish this is the Gamma.  Let $\bm{ X} \sim \prod_{l = 1}^d\text{Ga}\left(X_l\mid\alpha_l,\beta_l\right)$.
+  to accomplish this is the Gamma.  Let
+    $\bm{ X} \sim \prod_{\ell = 1}^d\text{Ga}\left(X_{\ell}\mid\alpha_{\ell},\beta_{\ell}\right)$.
   Then transform $\bm{ X}$ via the transformation in Eqns.~\ref{eqn:pnormt} into $R,\bm{ Y}$, where
@@
     \label{eqn:protopg}
-    f(r,\bm{ y}) = \prod_{l = 1}^{d}
-      \left[\frac{\beta_l^{\alpha_l}}{\Gamma(\alpha_l)}(ry_l)^{\alpha_l - 1}\exp\lbrace-\beta_lry_l\rbrace\right]
-      \times r^{d-1}\left[y_d - {\textstyle \sum}_{l = 1}^{d-1}y_l^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right],
+    f(r,\bm{ y}) = \prod_{\ell = 1}^{d}
+      \left[\frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell})}(ry_{\ell})^{\alpha_{\ell} - 1}
+          \exp\lbrace-\beta_{\ell}ry_{\ell}\rbrace\right]
+      \times r^{d-1}\left[y_d -
+            {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right],
   \end{equation}
-  where $y_d = \left(1 - \sum_{l=1}^{d-1}y_l^p\right)^{\frac{1}{p}}$. Integrating out $r$ yields the
-  the resulting \emph{Projected Gamma} density
+  where $y_d = \left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}$. Integrating out $r$
+  yields the resulting \emph{Projected Gamma} density
   \begin{equation}
     \label{eqn:projgamma}
-    \text{PG}(\bm{ y}\mid\bm{ \alpha},\bm{ \beta}) = \prod_{l = 1}^d\left[\frac{\beta_l^{\alpha_l}}{\Gamma(\alpha_l)}y_l^{\alpha_l - 1}\right]
-      \times \left[y_d - {\textstyle \sum}_{l = 1}^{d-1}y_l^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right]
-      \times \frac{\Gamma({\textstyle\sum}_{l = 1}^d\alpha_l)}{\left({\textstyle\sum}_{l = 1}^d \beta_ly_l\right)^{{\scriptstyle\sum_{l = 1}^d \alpha_l}}}
+    \text{PG}(\bm{ y}\mid\bm{ \alpha},\bm{ \beta}) =
+          \prod_{\ell = 1}^d\left[\frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell})}
+                y_{\ell}^{\alpha_{\ell} - 1}\right]
+      \times \left[y_d -
+          {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right]
+      \times \frac{\Gamma({\textstyle\sum}_{\ell = 1}^d\alpha_{\ell})}{\left({\textstyle\sum}_{\ell = 1}^d
+                    \beta_{\ell}y_{\ell}\right)^{{\scriptstyle\sum_{\ell = 1}^d \alpha_{\ell}}}}
   \end{equation}
@@
   We can regard the Dirichlet distribution as a special case of this density, where
-  $\bm{ y}\in \mathcal{S}_1^{d-1}$, and $\beta_l = \beta$ for all $l$.  The distribution described
-  in \cite{nunez2019} is related to this distribution in that it is established on
-  $\bm{ y}\in \mathcal{S}_2^{d-1}$, transformed to spherical coordinates via Eqn.~\ref{eqn:spherical}.
+  $\bm{ y}\in \mathcal{S}_1^{d-1}$, and $\beta_{\ell} = \beta$ for all $\ell$.  The distribution
+  described in \cite{nunez2019} is related to this distribution in that it is established on
+  $\bm{ y}\in \mathcal{S}_2^{d-1}$, then transformed to spherical coordinates via Eqn.~\ref{eqn:spherical}.
   The full conditional for $r$ takes the form
@@
     \label{eqn:rfullcond}
-    r\mid\bm{ \alpha},\bm{ \beta}, y \sim \text{Ga}\left(r\mid{\textstyle\sum}_{l = 1}^d \alpha_l, {\textstyle\sum}_{l = 1}^d \beta_ly_l\right).
+    r\mid\bm{ \alpha},\bm{ \beta}, y \sim \text{Ga}\left(r\mid{\textstyle\sum}_{\ell = 1}^d
+                    \alpha_{\ell}, {\textstyle\sum}_{\ell = 1}^d \beta_{\ell}y_{\ell}\right).
   \end{equation}
@@
 In the estimation of the parameters of this model, we can quickly arrive at a problem, in that
-  without some linear restriction on the rate parameters $\beta_l$, the model is not identifiable.
+  without some linear restriction on the rate parameters $\beta_{\ell}$, the model is not identifiable.
   The restriction most often used in the case of the Generalized Dirichlet, as well as by
@@
   distribution.  The shape and rate parameters for the latent \text{Gamma} random variables,
-  $\alpha_l,\beta_l$, must be greater than 0, so a logical prior to use would be a Gamma prior for
+  $\alpha_{\ell},\beta_{\ell}$, must be greater than 0, so a logical prior to use would be a Gamma prior for
   each parameter, excepting for $\beta_1 := 1$.
@@
       \bm{ y}\mid\alpha,\beta &\sim \text{PG}(\bm{ y}\mid\alpha,\beta)\\
-      \bm{ \alpha},\bm{\beta} &\sim {\textstyle \prod}_{l = 1}^d \text{Ga}(\alpha_l \mid \xi_l,\tau_l)
-              \times {\textstyle \prod}_{l = 2}^d \text{Ga}(\beta_l\mid \zeta_l,\sigma_l).
+      \bm{ \alpha},\bm{\beta} &\sim {\textstyle \prod}_{\ell = 1}^d \text{Ga}(\alpha_{\ell} \mid \xi_{\ell},\tau_{\ell})
+              \times {\textstyle \prod}_{\ell = 2}^d \text{Ga}(\beta_{\ell}\mid \zeta_{\ell},\sigma_{\ell}).
     \end{aligned}
   \end{equation}
-  A useful feature of the Projected Gamma family is the latent independent Gammas assumption.  Recall
+  A useful property of the Projected Gamma family is the latent independent Gammas assumption.  Recall
   in Eqn.~\ref{eqn:protopg}, we have a density defined as a product of independent Gamma densities,
-  multiplied by a factor of $r^{d-1}$ and some value, constant with respect to the parameters
-  $\alpha,\beta$.  If we conduct inference via a Gibbs sampler, then by simulating this latent $r$
-  we are able to perform inference of $(\alpha_l,\beta_l)$ independent of that of $(\alpha_k,\beta_k)$.
-  The full conditional for $r$ is described in Eqn.~\ref{eqn:rfullcond}.  The full conditional for
-  $\beta_l$, $l = 2,\ldots,d$ then becomes
+  multiplied by a factor of $r^{d-1}C$, where $C$ is constant with respect to the parameters
+  $\alpha,\beta$, as well as $r$.  If we conduct inference via a Gibbs sampler, then by simulating
+  this latent $r$ we are able to perform inference of $(\alpha_{\ell},\beta_{\ell})$ independent of
+  that of $(\alpha_k,\beta_k)$.  The full conditional for $r$ is described in Eqn.~\ref{eqn:rfullcond}.
+  The full conditional for $\beta_{\ell}$, $l = 2,\ldots,d$ then becomes
   \begin{equation}
     \label{eqn:betafc}
-    \beta_l\mid \bm{ y}, r, \alpha_l \sim \text{Ga}\left(n + \alpha_l + \zeta_l,
-                                      {\textstyle \sum}_{i = 1}^nr_iy_{il} + \sigma_l\right).
+    \beta_{\ell}\mid \bm{ y}, r, \alpha_{\ell} \sim \text{Ga}\left(n + \alpha_{\ell} + \zeta_{\ell},
+                                      {\textstyle \sum}_{i = 1}^nr_iy_{il} + \sigma_{\ell}\right).
   \end{equation}
-  That is, given $\alpha_l$, $r$, we can sample $\beta_l$ directly as a Gamma random variable.  In
-  the sampling for $\alpha_l$, we can integrate $\beta_l$ out leaving us a quantity proportional
-  to a density, which we can sample via the Metropolis Hastings algorithm\citep{hastings1970}.  For $\alpha_l$,
-  $l = 2,\ldots, d$, This takes the form
+  That is, given $\alpha_{\ell}$, $r$, we can sample $\beta_{\ell}$ directly as a Gamma random
+  variable.  In the sampling for $\alpha_{\ell}$, we can integrate $\beta_{\ell}$ out leaving us a
+  quantity proportional to a density, which we can sample via the Metropolis Hastings
+  algorithm\citep{hastings1970}.  For $\alpha_{\ell}$, $\ell = 2,\ldots, d$, This takes the form
   \begin{equation}
     \label{eqn:alphafc}
-    f(\alpha_l \mid \bm{ y}, r) \propto
-    \frac{\left({\textstyle \prod}_{i = 1}^nr_iy_{il}\right)^{\alpha_l - 1}}{\Gamma^n(\alpha_l)} \times
-    \alpha_l^{\xi_l - 1}\exp\{-\tau_l\alpha_l\} \times
-    \frac{\Gamma(n\alpha_l + \zeta_l)}{
-        \left({\textstyle\sum}_{i = 1}^n r_iy_{il} + \sigma_l\right)^{(n * \alpha_l + \zeta_l)}}.
+    f(\alpha_{\ell} \mid \bm{ y}, r) \propto
+    \frac{\left({\textstyle \prod}_{i = 1}^nr_iy_{il}\right)^{\alpha_{\ell} - 1}}{\Gamma^n(\alpha_{\ell})} \times
+    \alpha_{\ell}^{\xi_{\ell} - 1}\exp\{-\tau_{\ell}\alpha_{\ell}\} \times
+    \frac{\Gamma(n\alpha_{\ell} + \zeta_{\ell})}{
+        \left({\textstyle\sum}_{i = 1}^n r_iy_{il} + \sigma_{\ell}\right)^{(n * \alpha_{\ell} + \zeta_{\ell})}}.
   \end{equation}
@@
   \end{equation}
-  The \emph{projected restricted Gamma} model sets $\beta_l := 1 \forall l \in \lbrace 1,\ldots,d\rbrace$,
-  so in that model inference on all $\alpha_l$ will use the simpler form.
+  The \emph{projected restricted Gamma} model sets $\beta_{\ell} := 1 \forall \ell \in \lbrace 1,\ldots,d\rbrace$,
+  so in that model inference on all $\alpha_{\ell}$ will use the simpler form.
 
@@
       \bm{ y}_i &\sim \sum_{j = 1}^J\pi_j\text{PG}\left(\bm{ y}\mid \bm{ \alpha}_j, \bm{ \beta}_j\right)\\
-      \bm{ \alpha}_j &\sim {\textstyle \prod}_{l = 1}^d \text{Ga}\left(\alpha_l\mid\xi_l,\tau_l\right)\\
-      \bm{ \beta}_j &\sim {\textstyle \prod}_{l = 2}^d \text{Ga}\left(\beta_l\mid\zeta_l,\sigma_l\right)
+      \bm{ \alpha}_j &\sim {\textstyle \prod}_{\ell = 1}^d \text{Ga}\left(\alpha_{\ell}\mid\xi_{\ell},\tau_{\ell}\right)\\
+      \bm{ \beta}_j &\sim {\textstyle \prod}_{\ell = 2}^d \text{Ga}\left(\beta_{\ell}\mid\zeta_{\ell},\sigma_{\ell}\right)
     \end{aligned}
@@
     \begin{aligned}
-      \bm{ \xi},\bm{\tau} &\sim {\textstyle \prod}_{l = 1}^d \text{Ga}(\xi_l\mid a,b) \times \text{Ga}(\tau_l\mid c,d)\\
-      \bm{ \zeta},\bm{\sigma} &\sim {\textstyle\prod}_{l = 2}^d\text{Ga}(\zeta_l \mid a,b)\times \text{Ga}(\sigma_l\mid c,d)\\
+      \bm{ \xi},\bm{\tau} &\sim {\textstyle \prod}_{\ell = 1}^d \text{Ga}(\xi_{\ell}\mid a,b) \times \text{Ga}(\tau_{\ell}\mid c,d)\\
+      \bm{ \zeta},\bm{\sigma} &\sim {\textstyle\prod}_{\ell = 2}^d\text{Ga}(\zeta_{\ell} \mid a,b)\times \text{Ga}(\sigma_{\ell}\mid c,d)\\
       \bm{ \pi} &\sim \text{Dir}(\pi_0)
@@
 Additionally, we investigate a finite mixture extension of the \emph{Projected Restricted Gamma} model,
-  where we define all $\beta_l := 1$, for $l = 1,\ldots,d$, for $j = 1,\ldots,J$.  This affects the above
+  where we define all $\beta_{\ell} := 1$, for $l = 1,\ldots,d$, for $j = 1,\ldots,J$.  This affects the above
   model by removing inference steps for $\bm{ \beta},\bm{ \zeta},\bm{ \sigma}$.  The inference for
-  $\alpha_l$ for all $l = 1,\ldots,d$ follows the form in Equation~\ref{eqn:alphafc1}; inference on
+  $\alpha_{\ell}$ for all $\ell = 1,\ldots,d$ follows the form in Equation~\ref{eqn:alphafc1}; inference on
   $\bm{ \xi,\tau}$ follows the same form as the unrestricted model.  The impetus for the restricted
-  gamma model is that in the context of a mixture model, the additional flexibility offered by the unrestricted Gamma model may be better represented with additional clusters.
+  gamma model is that in the context of a mixture model, the additional flexibility offered by the
+  unrestricted Gamma model may be better represented with additional clusters.
 
@@
       (\alpha_i,\beta_i) &\sim \text{DP}\left(\eta, G_0\right)\\
-      &~\hspace{-2cm}G_0 = {\textstyle\prod}_{l = 1}^d\text{Ga}\left(\alpha_{jl}\mid\xi_{l},\tau_{l}\right)
-                    \times{\textstyle\prod}_{l=2}^d\text{Ga}\left(\beta_{jl}\mid\zeta_{l}\tau_{l}\right)
+      &~\hspace{-2cm}G_0 = {\textstyle\prod}_{\ell = 1}^d\text{Ga}\left(\alpha_{jl}\mid\xi_{l},\tau_{l}\right)
+                    \times{\textstyle\prod}_{\ell = 2}^d\text{Ga}\left(\beta_{jl}\mid\zeta_{l}\tau_{l}\right)
     \end{aligned}
@@
     \begin{aligned}
-      \bm{ \xi},\bm{ \tau} &\sim {\textstyle \prod}_{l = 1}^d \text{Ga}(\xi_l\mid a,b)\times \text{Ga}(\tau_l\mid c,d)\\
-      \bm{ \zeta},\bm{\sigma} &\sim {\textstyle\prod}_{l = 2}^d\text{Ga}(\zeta_l \mid a,b)\times \text{Ga}(\sigma_l\mid c,d)\\
+      \bm{ \xi},\bm{ \tau} &\sim {\textstyle \prod}_{\ell = 1}^d \text{Ga}(\xi_{\ell}\mid a,b)\times \text{Ga}(\tau_{\ell}\mid c,d)\\
+      \bm{ \zeta},\bm{\sigma} &\sim {\textstyle\prod}_{\ell = 2}^d\text{Ga}(\zeta_{\ell} \mid a,b)\times \text{Ga}(\sigma_{\ell}\mid c,d)\\
       \bm{ \eta} &\sim \text{Ga}(\eta \mid 2, 0.1).
@@
   \end{equation}
-  To fit the Dirichlet process, we use a variant of the collapsed Gibbs sampler detailed as Algorithm~8 in \citet{neal2000}.
-In this algorithm, we perform a data augmentation to create a cluster identifier, $\delta_i$, for each observation.
-Let $J$ indicate the number of extant clusters.
-Under the collapsed sampler scheme, for existing clusters, the probability of cluster membership is calculated as
+  To fit the Dirichlet process, we use a variant of the collapsed Gibbs sampler detailed as
+  Algorithm~8 in \citet{neal2000}.  In this algorithm, we perform a data augmentation to create a
+  cluster identifier, $\delta_i$, for each observation. Let $J$ indicate the number of extant clusters.
+  Under the collapsed sampler scheme, for existing clusters, the probability of cluster membership is
+  calculated as
   \begin{equation}
@@
     \text{P}\left(\delta_i = j\mid r, \bm{ y_i} \bm{ \alpha}, \bm{ \beta}\right) \propto
-        \frac{n_j}{\sum_{k = 1}^{J} n_k + \eta}
-        {\textstyle\prod}_{l = 1}^d\text{Ga}\left(r_iy_{il}\mid\alpha_{jl},\beta_{jl}\right).
+        \frac{n_j^{\not i}}{\sum_{k = 1}^{J} n_k + \eta}
+        {\textstyle\prod}_{\ell = 1}^d\text{Ga}\left(r_iy_{il}\mid\alpha_{jl},\beta_{jl}\right).
   \end{equation}
-  where $n_j = {\textstyle\sum}_{k\neq i}1_{k == j}$.  Effectively, if the observation under review is a
+  where $n_j^{\not i} = {\textstyle\sum}_{k\neq i}1_{k == j}$.  Effectively, if the observation under review is a
   singleton, its existing cluster is destroyed.  For new clusters, ostensibly we would integrate out
@@
   non-conjugate model may be difficult.  Instead, the impetus for using this specific DP fitting algorithm
-  is it simulates the posterior predictive density by generating $m$ new candidate clusters from their respective priors given
-  $\bm{ \xi,\tau,\zeta,\sigma}$.  Then the probability of $y_i$ belonging to a particular candidate
-  cluster is calculated as
+  is it simulates the posterior predictive density by generating $m$ new candidate clusters from
+  their respective priors given $\bm{ \xi,\tau,\zeta,\sigma}$.  Then the probability of $y_i$
+  belonging to a particular candidate cluster is calculated as
   \begin{equation}
@@
         \frac{\eta / m}{\sum_{j = 1}^{J} n_j + \eta}
-        {\textstyle\prod}_{l = 1}^d\text{Ga}\left(r_iy_{il}\mid\alpha_{kl}^{\prime},\beta_{kl}^{\prime}\right),
+        {\textstyle\prod}_{\ell = 1}^d\text{Ga}\left(r_iy_{il}\mid\alpha_{kl}^{\prime},\beta_{kl}^{\prime}\right),
   \end{equation}
-  where $(\cdot)^{\prime}$ indicates a value as belonging to a new candidate cluster, and $k$ indexes along candidate clusters.
-If a new cluster is selected, then $(\bm{\alpha}_{J+1},\bm{\beta}_{J+1}) = (\bm{\alpha}_k^{\prime},\bm{\beta}_k^{\prime})$, and $J = J + 1$.
+  where $(\cdot)^{\prime}$ indicates a value as belonging to a new candidate cluster, and $k$ indexes
+  along candidate clusters.  If a new cluster is selected, then
+  $(\bm{\alpha}_{J+1},\bm{\beta}_{J+1}) = (\bm{\alpha}_k^{\prime},\bm{\beta}_k^{\prime})$, and $J = J + 1$.
 
@@
   parameters.  As before, we will be looking at this model, and the restricted gamma form of this model, where
-  $\beta_l := 1$ for $l = 1,\ldots,d$.  This removes inference steps for $\bm{ \zeta},\bm{ \sigma}$
+  $\beta_{\ell} := 1$ for $\ell = 1,\ldots,d$.  This removes inference steps for $\bm{ \zeta},\bm{ \sigma}$
   as well.
@@
         &~\hspace{-1cm}G_0 = \mathcal{N}\left(\log\bm{ \alpha}_{j}\mid\mu,\Sigma\right)\times
-            {\textstyle\prod}_{l = 2}^d\text{Ga}\left(\beta_{jl}\mid\zeta_l,\sigma_l\right)
+            {\textstyle\prod}_{\ell = 2}^d\text{Ga}\left(\beta_{j,\ell}\mid\zeta_{\ell},\sigma_{\ell}\right)
     \end{aligned}
@@
       \mu,\Sigma &\sim \mathcal{N}\left(\mu\mid\mu_0,\Sigma_0\right) \times \text{IG}\left(\nu,\Psi\right)\\
-      \bm{ \zeta},\bm{\sigma} &\sim {\textstyle\prod}_{l = 2}^d\text{Ga}(\zeta_l \mid a,b) \times \text{Ga}(\sigma_l \mid c,d) \\
+      \bm{ \zeta},\bm{\sigma} &\sim {\textstyle\prod}_{\ell = 2}^d\text{Ga}(\zeta_{\ell} \mid a,b) \times \text{Ga}(\sigma_{\ell} \mid c,d) \\
       \bm{ \eta} &\sim \text{Ga}(\eta \mid 2, 0.1).
@@
     f(\bm{\alpha}_j\mid\bm{\delta},\bm{y},\mu,\Sigma,\zeta,\sigma) &\propto
-    \frac{\left(\prod_{i:\delta_i = j}y_{il}\right)^{\alpha_{j1} - 1}}{\Gamma(\alpha_{j1})^{n_j}}
-        \prod_{l = 2}^d\left[\frac{\left(\prod_{i:\delta_i = j}y_{il}\right)^{\alpha_{jl} - 1}
-        \Gamma(n_j\alpha_{jl} + \zeta)}{\Gamma(\alpha_{jl})^{n_j}\left(\sum_{i:\delta_i = j}y_{il}
-        + \sigma\right)^{n_j\alpha_{jl} + \zeta}}\right]\\
+    \frac{\left(\prod_{i:\delta_i = j}y_{i,\ell}\right)^{\alpha_{j,\ell} - 1}}{\Gamma(\alpha_{j,\ell})^{n_j}}
+        \prod_{\ell = 2}^d\left[\frac{\left(\prod_{i:\delta_i = j}y_{i,\ell}\right)^{\alpha_{j,\ell} - 1}
+        \Gamma(n_j\alpha_{j,\ell} + \zeta)}{\Gamma(\alpha_{j,\ell})^{n_j}\left(\sum_{i:\delta_i = j}y_{i,\ell}
+        + \sigma\right)^{n_j\alpha_{j,\ell} + \zeta}}\right]\\
         &\hspace{1cm}\times
-        \lvert \Sigma\rvert^{-\frac{1}{2}}\prod_{l=1}^d\alpha_{jl}^{-1}\exp\left\lbrace
+        \lvert \Sigma\rvert^{-\frac{1}{2}}\prod_{\ell = 1}^d\alpha_{j,\ell}^{-1}\exp\left\lbrace
           -\frac{1}{2}(\log\alpha_j - \mu)^T\Sigma^{-1}(\log\alpha_j - \mu)\right\rbrace.
@@
   prior resulted in extremely poor model performance.  We thus used the restricted gamma likelihood,
-  assuming $\beta_{jl} = 1$ for all $j,l$.  This simplifies Equation~\ref{eqn:dplnlp} into
+  assuming $\beta_{j,\ell} = 1$ for all $j,\ell$.  This simplifies Equation~\ref{eqn:dplnlp} into
   \begin{equation}
@@
     f(\bm{\alpha}_j\mid\bm{\delta},\bm{y},\mu,\Sigma) &\propto
-    \prod_{l = 1}^d\left[\frac{\left(\prod_{i:\gamma_i = j}y_{il}\right)^{\alpha_{jl} - 1}}{\alpha_{jl}\Gamma(\alpha_{jl})^{n_j}} \right]\times\lvert\Sigma\rvert^{-\frac{1}{2}}
+    \prod_{\ell = 1}^d\left[\frac{\left(\prod_{i:\gamma_i = j}
+        y_{i,\ell}\right)^{\alpha_{j,\ell} - 1}}{\alpha_{j,\ell}\Gamma(\alpha_{j,\ell})^{n_j}} \right]
+        \times\lvert\Sigma\rvert^{-\frac{1}{2}}
       \exp\left\lbrace-\frac{1}{2}(\log\alpha_j - \mu)^T\Sigma^{-1}(\log\alpha_j - \mu)\right\rbrace.
@@
   inference steps for $\zeta,\sigma$.
-  %
-  % \bruno{\bf I think that 3.2.2--3.2.4 are pretty much the same model with some elaborations.
-  %               You should be able to present the whole story in a much more condensed fashion.}
 
@@
   posterior predictive distribution for a model built on $\mathcal{S}_p^{d-1}$ onto
-  $\mathcal{S}_{\infty}^{d-1}$. As $p$ increases, the closer the manifold will be to $\mathcal{S}_{\infty}^{d-1}$, and thus the closer the projected distribution will be to an idealized
+  $\mathcal{S}_{\infty}^{d-1}$. As $p$ increases, the closer the manifold will be to
+  $\mathcal{S}_{\infty}^{d-1}$, and thus the closer the projected distribution will be to an idealized
   projected Gamma built on $\mathcal{S}_{\infty}^{d-1}$.  In testing, we tried developing models on

### 2021-09-13 applications.tex

\section{Applications to EVT Analysis}
\label{ref:applications}
Representing the angular component of the Pareto process with a (semi) parametric distribution grants
  us opportunities in posterior analysis.

\subsection{Pairwise Extremal Dependence Coefficients}
A summary measure of extremal dependence between dimensions can be offered by the extremal dependence
  coefficient.  This can be thought of as something like a correlation coefficient for extreme
  observations.  For a random variable $\bm{Z}$, which has been standardized such that the marginal
  distributions are common between dimensions, it is defined as
  \begin{equation}
    \chi_{k\ell} = \lim\limits_{u\to\infty}\text{P}\left(Z_k > u\mid Z_{\ell} > u\right).
  \end{equation}
  The value $\chi_{k\ell} \in [0,1]$.  When $\chi_{k\ell}$ approaches 1, that indicates complete dependence;
  the variables are linked, in much the same manner as if a correlation coefficient approaches 1.
  When $\chi_{k\ell}$ is greater than 0, $X_k$ and $X_{\ell}$ are considered asymptotically dependent.  For
  a $\bm{Z}$ distributed by a Pareto process, \cite{warner2018} transform $Z$ to the unit square to
  re-express this quantity,
  \begin{equation}
    \label{eqn:extdepcoef}
    \chi_{k\ell} = \text{E}\left[\frac{V_k}{\text{E}(V_k)}{\bigwedge}\frac{V_{\ell}}{\text{E}(V_{\ell})}\right]
  \end{equation}
  where $\bm{V} = \bm{Z} / \pnorm{\bm{Z}}{\infty}$.  This calculation is wholly dependent on the
  angular component, as radial measure is independent of angular measure.  However, it is Important
  to note that complete asymptotic independence, where $\chi_{k\ell} := 0$, can not be described
  under a Pareto process based model.

  \begin{figure}[ht!]
    \centering
    \caption{Pairwise extremal dependence coefficients for IVT data, computed from fitted DP
      mixture of projected restricted gamma model.  The left corresponds to the 8-cell data, the
      right the 47-cell data.\label{fig:chi_ij}}
    \begin{minipage}{.49\textwidth}
      \centering
      \includegraphics[width=0.99\linewidth]{./images/chi_ij_8}
    \end{minipage}
    \begin{minipage}{.49\textwidth}
      \centering
      \includegraphics[width=0.99\linewidth]{./images/chi_ij_46}
    \end{minipage}
  \end{figure}

Figure~\ref{fig:chi_ij} plots the pairwise extremal dependence coefficients for the IVT data, in 8
  cells (left) and 47 cells (right), computed fom a fitted DP mixture of projected restricted gammas.
  We see a higher extremal dependence between neighboring columns which is not unexpected, as
  neighboring columns correspond to neighboring cells.  We recognize that pairwise asymptotic
  dependence coefficients tell a limited story, as a particular dependence may structure may include
  more than two columns.   We can, however, glean some information from the patterns that emerge in
  two dimensions.  On the 8-cell data, we see a stronger association between cells 5-8, indicating a
  greater dependence among these cells.  On the 47 cell data, we see at least 3 groupings of columns,
  indicating greater asymptotic dependence among these groups.

\subsection{Conditional Survival Curves}
Another interesting inference we can make from a fitted model is a conditional survival curve.
  Following rules of conditional probability, we can develop survival curves based on
  threshold exceedences. For ${\bf Z} = (Z_1,\ldots,Z_d)$ following a $d$-variate Pareto
  distribution, we make the proposition:

\begin{prop}
    In one dimension, the conditional survival curve can be calculated as
 \begin{equation}
    \label{eqn:condsurv1df}
    \text{P}\left[Z_{\ell} > z_{\ell}\mid {\bf Z}_{\neg(\ell)} > {\bf z}_{\neg(\ell)}\right] =
      \frac{\text{E}\left[\wedge_{k = 1}^d \frac{V_k}{z_k}\right]}{\text{E}\left[\bigwedge_{k \neq l}\frac{V_k}{z_k}\right]}
  \end{equation}
    where $\bm{V} = \bm{Z} / \pnorm{\bm{Z}}{\infty}$.
\end{prop}

By simple application of conditional probability, we expand the conditional survival function into
  a ratio of probabilities:
  \begin{equation}
    \label{eqn:condsurv1d}
    \text{P}\left[Z_{\ell} > z_{\ell}\mid {\bf Z}_{\neg(\ell)} > {\bf z}_{\neg(\ell)}\right] =
      \frac{\text{P}\left[\cap_{k = 1}^d Z_k > z_k\right]}{\text{P}\left[\cap_{k \neq l} Z_k > z_k\right]}.
  \end{equation}
  Let $R = \pnorm{{\bf Z}}{\infty}$, ${\bf V} = \frac{{\bf Z}}{R}$, such that ${\bf V}\in \mathcal{S}_{\infty}^{d-1}$.
  For evaluating these probabilities, recall that ${\bf Z} = R{\bf V}$.  Then,
  \begin{equation}
    \text{P}\left(\cap_{k = 1}^d Z_k > z_k\right) = \text{P}\left(\cap_{k = 1}^d RV_k > z_k\right)
  \end{equation}
  Also, for standard Pareto, recall that $\text{P}(R > r) = 1\wedge\frac{1}{r}$.  Then,
  \begin{equation}
    \text{P}\left[\bigcap_{k = 1}^d R > \frac{z_k}{V_k}\right] =
      \text{P}\left[R  > \bigvee_{k=1}^d\frac{z_k}{V_k}\right] =
      \text{E}\left[1 \bigwedge \left(\bigvee_{k = 1}^d\frac{z_k}{V_k}\right)^{-1}\right].
  \end{equation}
  As $V_k \in [0,1]$, and given our interest in the tail of the distribution, we can assume
  $z_k > 1$, therefore $\frac{z_k}{V_k} > 1$.  We thus arrive at
  \begin{equation}
    \text{P}\left[\cap_{k = 1}^d Z_k > z_k\right] = \text{E}\left[\wedge_{k = 1}^d\frac{V_i}{z_i}\right].
  \end{equation}
  We employ a similar calculation in the denominator, yielding
  \begin{equation}
    \text{P}\left[Z_{\ell} > z_{\ell}\mid {\bf Z}_{\neg(l)} > {\bf z}_{\neg(l)}\right] =
      \frac{\text{E}\left[\wedge_{k = 1}^d \frac{V_k}{z_k}\right]}{\text{E}\left[\bigwedge_{k \neq \ell}\frac{V_k}{z_k}\right]}
  \end{equation}
  Figure~\ref{fig:condsurv1d} displays this conditional survival function for the 8 cell data. For each
  cell, the survival function is computed given all other cells are at or above their $90$th marginal
  percentile.

\begin{figure}[b]
  \caption{
    Single-variable conditional survival function in extreme
    regions---$\text{Pr}[Z_{\ell} > z_{\ell}\mid Z_{\neg l} > z_{\neg l}]$,
    evaluated where $z_{\neg l} = F_{\neg l}^{-1}(0.9)$---for the 8-cell data.  The model used to
    generate this estimate is the DP mixture of projected restricted Gammas.\label{fig:condsurv1d}
    }
  \centering
  \includegraphics[width = 5in]{./images/condsurv_1d}
\end{figure}

Following similar calculations, for a dimension set $\alpha \subset \{1,\ldots, d\}$, a multivariate
  conditional survival function can be similarly estimated as
  \begin{equation}
    \label{eqn:condsurv2df}
    \text{P}\left[\bigcap_{\ell \in \alpha} Z_{\ell} > z_{\ell} \mid \bigcap_{\ell\not\in\alpha} Z_{\ell} > z_{\ell}\right] =
      \frac{\text{E}\left[\bigwedge_{k = 1}^d \frac{V_k}{z_k}\right]}{\text{E}\left[\bigwedge_{k \not\in\alpha}\frac{V_k}{z_k}\right]}.
  \end{equation}
  Figure~\ref{fig:condsurv2d} displays selected bi-variate conditional survival surfaces, for the 8-cell
  data.  For each pair of cells, the survival function is again computed conditional on  all other cells
  at or above their $90$th percentile.  Note, for cell pairs (1,2) and (7,8), the strong asymptotic dependence
  observed in Figure~\ref{fig:chi_ij} exhibits itself in a greater tendency towards jointly extreme
  behavior in the survival function; whereas cell pairs (1,6) and (2,8) show their low asymptotic
  dependence with a low propensity towards jointly extreme behavior.

\begin{figure}[t]
  \caption{
    Selected bi-variate conditional survival surfaces in extreme regions for the 8 cell data.
    The DP mixture of projected restricted Gammas model was used to generate these surfaces.
    \label{fig:condsurv2d}
    }
  \includegraphics{./images/condsurv_2d}
\end{figure}

% EOF

### 2021-09-13 conclusion.tex

In this paper, we have built upon the definition of the multivariate Pareto described in \cite{ferreira2014}
  to establish a useful parametric representation of its dependence structure, through describing
  the distribution of its angular component; it being supported on the positive orthant of the unit
  hypersphere under the $\mathcal{L}_{\infty}$ norm, $\mathcal{S}_{\infty}^{d-1}$.  We then demonstrated an
  inherent difficulty in establishing a distribution supported in this space.  To build that representation,
  we presented a distribution based on a product of independent Gammas, projected onto a supporting manifold
  $\mathcal{S}_{p}^{d-1}$, which will converge to $\mathcal{S}_{\infty}^{d-1}$ as $p\to\infty$.  We then
  established a means of model comparison on $\mathcal{S}_{\infty}^{d-1}$ through the use of a negative
  definite kernel defined in that space.

We used integrated vapor transport data---a measure of water in the atmosphere---as a motivating example for
  our model.  This data subsets California into a series of grid cells, and provides daily IVT values for
  each cell.  We used our model to develop measures of pairwise extremal dependence between grid cells, which
  could be of interest to researchers, actuaries, hydrologists and even farmers.  We additionally developed
  conditional survival functions---$\text{P}\left[Z_{\ell} > z_{\ell}\mid Z_{\not\ell} > z_{\not\ell}\right]$---which
  provide a more complete understanding of the dependence structure.

### 2021-09-13 results.tex

\section{Results}

\subsection{Simulation}
We have introduced this projected gamma distribution existing on $\mathcal{S}_{\infty}^{p-1}$.  As a
  means of evaluating how well we can recover a distribution under the proposed model, we have
  generated a series of datasets on $\mathcal{S}_{\infty}^{d-1}$, for $d$ ranging between 3 and 20.
  We generated each dataset from a mixture of projected Gammas, with the number of mixture components
  ranging between 3 and 12.  The generation procedure is detailed in Algorithm~\ref{algo:simulated}.
  \begin{algorithm}[htb]
    \For{$n_{\text{mix}}$ in $[3, 6, 9, 12]$}{
      Generate $n_{\text{mix}} \times 20$ shape Parameters $\alpha$\\
      Generate $n_{\text{mix}} \times 20$ Rate Parameters $\beta$\\
      Generate $500$ Mixture Component Identifiers $\delta$\\
      \For{$i$ in $1,\ldots,500$}{
        Generate ${\bf X}_i \sim \prod_{\ell = 1}^d\text{Ga}\left(X_{il}\mid\alpha_{\delta_i,l},\beta_{\delta_i, l}\right)$
        }
      \For{$n_{\text{col}}$ in $[3,6,12,20]$}{
        Project columns 1 to $n_{\text{col}}$ of ${\bf X}$ onto $\mathcal{S}_{\infty}^{n_{\text{col}} - 1}$ and save.
        }
    }
    \caption{Simulated Angular Dataset Generation Routine\label{algo:simulated}}
  \end{algorithm}
  The choice of generation routine for $\alpha = \alpha_0 + \alpha_1$ where
  $\alpha_0 \sim \text{Unif}(\alpha_0\mid 0,4)$, $\alpha_1\sim \text{Gamma}(\alpha_1\mid 1,1)$.
  The choice of generation routine for $\beta$ is $\beta\sim\text{Unif}(\beta\mid 0.25, 2.5)$.
  This choice was made to ensure that $\beta$ does not approach 0 in the generation routine.

\begin{figure}[ht]
  \caption{Simulation Energy Score versus Column Count, by Mixture Component Count.\label{fig:simes}}
  \centering
  \includegraphics[width=0.8\textwidth]{./images/simulation_es}
\end{figure}

In Figure~\ref{fig:simes}, we fit DP Mixtures of the various models presented and plot the resulting
  energy score against the number of columns, grouped by the number of mixture components used in
  dataset generation.  The \emph{Generative} model refers to a dataset generated
  in the same manner, from the same parameters as was used to create the original dataset.  Against
  this generative dataset, we compare DP Mixtures of projected gamma, projected restricted gamma, and
  projected restricted gamma with a multivariate log-normal prior.  We observe that the projected
  restricted Gamma model dominates in most situations, but the difference between projected restricted
  Gamma and the other gamma-based models tends to shrink as the number of columns increases.
  Alternatively, that difference appears to grow as the number of mixture components increases.

\begin{figure}[ht]
  \caption{Simulation Posterior Predictive Loss versus Column Count, by Mixture Component Count.\label{fig:simppl}}
  \centering
  \includegraphics[width=0.8\textwidth]{./images/simulation_ppl}
\end{figure}

The posterior predictive loss criterion in Figure~\ref{fig:simppl} tells a similar story: the base
  model is inadequate for describing a mixture; and the DP mixture of projected restricted gammas
  performs the best of the presented models in most cases.  Interestingly, under both metrics we see
  the probit-normal model has the best performance when the number of columns is low.  That is
  unexpected behavior, as its performance very quickly degrades as the number of columns increases.

We frequently see all of the gamma-based mixture models accomplish a lower energy score than the generative
  distribution.  As \cite{nunez2019} noted, the projected Gamma model is very flexible and can generate
  multimodal distributions from a single mixture component.  When we allow a mixture model, we allow
  the possibility to break those modes into two different mixture components.  This will tend to
  result in a better energy score, as posterior predictive replicates for a given observation in a local
  model will tend to be concentrate in that local mode rather than being spread between multiple
  modes.  As for the dominance of projected restricted Gamma, By specifying $\beta_{\ell} := 1$ for all
  $l$, if there were multi-modal mixture components, we are in effect breaking the mixture components
  into separate modes. As for whether this represents \emph{overfitting}, If we had reason to
  believe that the generative distribution for real data followed this form, then such a case could
  be made.  However, we are using projected Gamma as a parametric stand-in for an unknown distribution.
  In this sense, the argument for overfitting is less clear.

\begin{figure}[ht]
  \caption{Estimated $k$NN-KL Divergence metric on simulated data, averaged between 1 and 10 nearest
    neighbors, against column count, for selected models.  Graphs are facetted according to the
    number of mixture components in the generating distribution.  The \emph{generative} model
    represents another replicate of the underlying distribution, so that should be regarded as
    \emph{truth}.\label{fig:simkld}}
  \includegraphics[width = 5.5in, height = 3.75in]{./images/simulation_knn_kld}
\end{figure}

in Figure~\ref{fig:simkld} we see the application of the $k$NN-based K.L. divergence metric detailed
  in Equation~\ref{eqn:knnkld} to the simulated data sets.  For display, this metric has been averaged
  between $1$ and $10$ neighbors.  In general, the K.L. divergence metric is a log-ratio
  of densities, integrated over the entire space.  In this fashion, we should interpret that model
  is \emph{best} that stays closest to 0, as that indicates the two densities are equivalent.  The
  equation also includes a scaling factor for the size of the data sets, as all else being equal, a
  larger data set will decrease the distance between any point and its neighbors.  We can regard the
  \emph{generative} model as another replicate data set from the generating distribution---another
  view of truth.  As its distribution should be almost identical, we would assume that the KL divergence
  metric from this model will come closest to 0.  It does not.  Knowing this to be the case, we should
  instead regard the generative model as truth, and treat as best the model which comes closest to the
  \emph{generative} model.  In this regard, we observe that the projected gamma model comes closest,
  with the projected restricted gamma model following behind.  The probit-normal model quickly diverges
  as the number of dimensions increases.  This is not unexpected, as we saw similar behavior in the
  energy score and posterior predictive loss criteria.

\subsection{Integrated Vapor Transport}
We use data from the integrated water vapor transport (\emph{IVT}) model \citep{guan2015}
  of atmospheric rivers as a means of testing these models.  An atmospheric river is a meteorological
  event of local concentration of water vapor in the atmosphere that moves with wind patterns.
  Understanding extremal dependence in atmospheric rivers, for instance, grants hydrological engineers
  a greater understanding of how and where to build ground water capacity to handle precipitation.
  California suffers persistent, extended droughts---if ground water transportation and holding capacity
  (canals, reservoirs) are not built to manage extreme precipitation without adequate understanding of
  the extremal dependence, then we may not install adequate capacity and thus experience flooding.
  Alternatively, if we install too much capacity everywhere, then we are wasting money that could be
  better spent elsewhere.

Fitting our models to this data requires some pre-processing. First, we subset the data to the rainy
  season, which appears from November to March.  The marginal distributions of the
  IVT data appear naturally log-normal, which falls into the domain of attraction of a Gumbel
  distribution.  Given that, we can apply a high threshold, and exceedances over that threshold can
  be modelled as Generalized Pareto.  Our modelling begins with that assessment.  As estimating
  the Pareto parameters is not yet our focus in this analysis, we choose to apply the threshold
  using the empirical CDF.  That is, for a given $t$, let $b_{t,\ell} = \hat{F}_{\ell}^{-1}(1 - t^{-1})$.  For
  this analysis, we set $t = 20$, indicating the marginal $95$ percentile.  The other parameters of the
  generalized Pareto---the scale parameter $\alpha_{\ell}$ and the extremal index $\chi_{\ell}$---are set via
  maximum likelihood.  A fully Bayesian model formulation will allow their varying within a
  distribution, but such a model formulation will not be conducive to fitting by Markov-chain Monte
  Carlo methods.

After the thresholding and maximum likelihood estimation of the parameters of the Pareto, we scale
  the data to the standard multivariate Pareto.  Dividing each standardized observation by its
  $\mathcal{L}_{\infty}$ norm, we project the standardized data onto $\mathcal{S}_{\infty}^{d-1}$.
  Data in sequence represents observations in time, and these are heavily correlated.  As such, we
  choose to \emph{decluster} the observations, such that, by observing a sequence of observations
  $\bm{ z}$ for which $\inorm{z_i} > 1$ for each observation in sequence, we keep only the observation
  with the greatest observed $\mathcal{L}_{\infty}$ norm.  The complete procedure is outlined in
  Algorithm~\ref{algo:processing}.

\begin{algorithm}
  \KwResult{$\bm{ r} : r_i \sim \text{Pareto}(1)$, $\bm{ v} \in \mathcal{S}_{\infty}^{d-1}$}
  \For{$\ell = 1,\ldots,d$}{
    Set $b_{t,\ell} = \hat{F}_{\ell}^{-1}\left(1 - \frac{1}{t}\right)$.\\
    With $\bm{ x}_{\ell} > b_{t,\ell}$, fit $a_{\ell}$, $\chi_{\ell}$ via MLE according to generalized Pareto likelihood.\\
    }
  \For{$i = 1,\ldots,n$}{
    Define $z_{i,\ell} = \left(1 + \xi_{\ell}\frac{x_{i,\ell} - b_{t,\ell}}{a_{\ell}}\right)_{+}^{1/\xi_{\ell}}$\\
    Define $r_i = \pnorm{\bm{ z}_i}{\infty}$, $\bm{ V}_i = \frac{\bm{ z}_i}{\pnorm{\bm{ z}_i}{\infty}}$\\
    }
  Subset $\bm{ r},\bm{ v}$ such that $r_i \geq 1$ for all $r_i\in \bm{r}$.\\
  \If{declustering}{
    \For{$i = 1,\ldots,n$}{
      If $r_i \geq 1$ and $r_{i-1} \geq 1$, drop the lesser (and associated $v_i$) from data set.\\
    }
  }
 \caption{Data preprocessing to isolate and transform data exhibiting extreme behavior.  $r_i$
   represents the radial component, and $\bm{v}_i$ the angular component.  The declustering
   portion is relevant if data is correlated in time.\label{algo:processing}}
\end{algorithm}

We have, at our disposal, two data sets from the IVT model, totalling approximately 30 years of daily
  values omitting leap days.  One records data from 8 grid cells, covering
  generally the coast of California.  The other, with a higher resolution, records data from 47 grid
  cells covering the same area.  For the 8-cell data, after subsetting to the rainy season, we have
  $5587$ observations.  After the data processing steps in Algorithm~\ref{algo:processing}, where we
  omit observations not extreme in any dimension, and decluster temporally correlated data, we are left
  with $511$ observations.  For the 47-cell data, after subsetting to the rainy season, we have $6342$
  observations.  After the same data processing, we are left with
  $532$ observations.  We fit our models to both data sets.

\begin{table}[b]
  \centering
  \caption{Model comparison metrics: Posterior Predictive Loss and Energy Score criteria from fitted
    models against the IVT data.  All presented models are DP mixtures; the \emph{Model} field
    identifies the kernel distribution.  For both criteria, lower is better.
  \label{tab:dev}}
  \include{table_dev}
\end{table}

In Table~\ref{tab:dev} we see the same model preference towards the projected restricted gamma models
  that we saw in Figures~\ref{fig:simes},\ref{fig:simppl}.  The unrestricted gamma model is penalized much
  more strongly on real data than we saw with our simulation, performing comparably to the probit-normal
  model.  We also see the restricted gamma model with the log-normal prior performing significantly
  better than the restricted gamma model with the gamma prior on the higher dimensional data, a reversal
  of what we saw on the low dimensional data and indeed from what we saw in the simulation study.

That said, there is much we can glean from this table.  The probit-normal model, developed on a
  mapping from $\mathcal{S}_{2}^{d-1}$ to $(-\infty,\infty)^{d-1}$ via probit transformation of
  spherical coordinates was not competitive with the gamma models on real data, similar to what we
  saw in the simulation study. We also established finite mixture models for all presented
  distributions, but after tuning the number of mixture components, their model performance was,
  and could only be \emph{as good} as the DP mixture.  In the interest of parsimony, we only present
  the DP models as representative of mixture distributions.

\begin{figure}[ht]
  \centering
  \caption{KL Divergence Curves calculated through the KNN-KL Metric, evaluated between the
  empirical dataset and posterior predictive datasets of various models.  The top row corresponds
  to the 8-dimensional data, while the bottom row the 47.  All presented models use a DP prior,
  the nominal model being the kernel distribution.\label{fig:knnkl}}
  \includegraphics[width = 5.5in, height = 2.75in]{./images/knn_kld}
\end{figure}

Interpreting the KL divergence curves in Figure~\ref{fig:knnkl} is more difficult.  As we saw in
  the simulated case in Figure~\ref{fig:simkld}, the general logic of \emph{best} being that which
  is closest to zero is not necessarily in use here.  We interpret a positive value in the KL
  divergence metric as indicating that, for a given point in the given distribution, we must
  travel \emph{farther} on the target distribution to reach the $k$th neighbor as compared to
  the given distribution.  In this regard, we should regard \emph{large} positive values as
  indicating a particularly poor fit.  By this metric, the projected gamma and probit-normal
  models do not appear to scale well as they performed very poorly on the high dimensional data.
  For the projected restricted gamma model, the log-normal prior appears to scale better than the
  gamma prior.  This is in keeping with what we observed in Table~\ref{tab:dev}, but is a
  different result than what we observed in the simulation study, where we observed that the
  log-normal prior only tended towards the gamma prior, never performing better.

% EOF

### 2021-09-15 introduction.tex
Bruno's edits

@@
 \section{Introduction}
-In recent years much work has been done in the exploration of the definition and properties of an
-  appropriate generalization of the \emph{univariate} generalized Pareto distribution for exceedences
-  over a threshold, to a \emph{multivariate} setting.  For a short accounting of the topic, the work of
+The statistical analysis of extreme values focuses on inference for
+rare events that correspond to the tails of probability distributions.
+As such, it is a key ingredient in the assessment of the risk of
+phenomena that can have strong societal impacts like floods, heat waves,
+high concentration of pollutants, crashes in the financial markets,
+among others. The fundamental challenge of extreme value theory (EVT) is
+to use information, collected over limited periods of time, to
+extrapolate to long time horizons. This sets EVT apart from most of
+statistical inference, where the focus is on the bulk of the
+distribution. Extrapolation to the tails of the distributions is
+possible thanks to theoretical results that give asymptotic descriptions
+of the probability distributions of extreme values.
+
+Inferential methods for the extreme values of univariate observations
+are well established and software is widely available \cite[see, for
+example,][]{coles2001}.
+For variables in one dimension the application of EVT methods consider the asymptotic distribution of either the maxima calculated for regular blocks of data, or the values that exceed a certain threshold.
+The former leads to a Generalized Extreme Value (GEV) distribution, that depends on three parameters.
+The latter, which is the focus of this paper, leads to a Generalized Pareto (GP) distribution, that depends on a shape and a scale parameter.
+Likelihood-based approaches to inference can be readily implemented in both cases.
+In the multivariate case the GEV theory is well established \citep[see, for example][]{dehaan2006}, but the inferential problem is more complicated, as there is no parametric representation of the GEV.
+This problem is inherited by the PoT approach and compunded by the fact that there is no unique definition of an exceedance of a multivariate threshold, as there is an obvious dependence on the norm that used to measure the size of a vector.
+
+During the last decade much work has been done in the exploration of the definition and properties of an
+  appropriate generalization of the univariate GP distribution 
+  to a multivariate setting.  To mention some of the papers on the topic, the work of
   \cite{rootzen2006} defines the generalized Pareto distribution, with further analysis on these classes
@@
   of the art in multivariate peaks over threshold modelling using generalized Pareto is provided in
-  \cite{rootzen2018}.  \cite{ferreira2014} presents a constructive definition of the multivariate Pareto
-  that decomposes the random vector into independent radial and angular components; the latter
-  containing all and only information pertaining to the dependence structure of the random vector.
-  Based on this definition, we present a novel approach for modelling that angular component, and thus
-  the modelling the dependence structure of the random vector.
+  \cite{rootzen2018} while \cite{RoSeWa2018a} provides insight on the theoretical properties of possible parametrizations. 
+  A frequently used method for describing the dependence
+  in multivariate distributions is to use a copula.
+\cite{renard2007,deng2011,falk2019}, provide successful examples of this approach in an EVT framework.
+{\bf is \cite{deng2011} really a relevant reference?
+It looks weird.}
+  \cite{ferreira2014} presents a constructive definition of the Pareto process, that generalizes the GP to an infinite dimensional setting.
+It consists of decomposing the process into independent radial and angular components.
+Such approach can be used in the finite dimensional case, where the angular component
+  contains the information pertaining to the dependence structure of the random vector.
+  Based on this definition, we present a novel approach for modelling the distribution of the angular component
+  with families of distributions that provide flexibility and can be applied in a moderately large dimensional setting.
+  Models for the angular measure are important for multivariate EVT applications.
+\cite{SaNa2014} and \cite{HaCaCh2017} consider Bayesian non-parametric approaches to model the angular measure of the multivariate GEV.
+Particular attention is paid to the problem of satisfying the so called moment constraint.
+
+  The reminder of this paper is outlined as follows.
 
@@
 
-\subsection{Extreme Value Theory}
-\subsubsection{Univariate Maxima}
-Regarding the asymptotic behavior of extreme events, there are a couple major strategies for conducting
-  inference---developing probabilistic estimates of extreme behavior.  First developed is the theory
-  of a limiting distribution on maxima---the largest observation from a sample.  For a sample $\bm{x}$
-  where $\bm{x} = (x_1,\ldots,x_n)$ represents a sequence of $n$ independent random variables from a
-  distribution function $F$, the distribution of the maximum $M_n$ of this sequence can be derived
-  as:
-  \begin{equation*}
-    \begin{aligned}
-      \text{Pr}\left[M_n\leq z\right] &= \text{Pr}\left[X_1 \leq z, \ldots, X_n \leq z\right]\\
-        &= \text{Pr}\left[X_1\leq z\right]\times\ldots\times\text{Pr}\left[X_n\leq z\right]\\
-        &= F(z)^n.
-    \end{aligned}
-  \end{equation*}
-  In situations where $F$ is unknown, we can seek to approximate the behavior of $F^n$ as
-  $n\rightarrow\infty$.  To ensure this does not degenerate to a point mass, we need to consider a
-  standardized sequence of maxima. If this sequence stabilizes as $n$ increases, the a limiting
-  distribution exists.  More specifically, if there exists some sequence of constants $a_n > 0$, $b_n$
-  such that:
-  \begin{equation*}
-    \text{Pr}\left[\frac{M_n - b_n}{a_n} \leq z\right] \stackrel{d}{\rightarrow} G(z)
-  \end{equation*}
-  as $n\rightarrow\infty$, then $G$ is a max-stable distribution, and $F$ is in the domain of
-  attraction of that max stable distribution.  \cite{frechet1927} originates the
-  field, identifying two limiting forms of max-stable distributions, which would become known as the
-  Fr{\'e}chet and Weibull distributions. \cite{weibull1951} expands the analysis of the Weibull
-  distribution.  \cite{fisher1928}, in what is now called the \emph{Fisher-Tippett Theorem}, identifies
-  the Fr{\'e}chet and Weibull distributions, along with an as then unnamed third form, as the three
-  limiting forms of the distribution of the maxima of a sample.  \cite{gumbel1935,gumbel1942} offer
-  analysis of that third form, now known as the Gumbel distribution.  Later works, including
-  \cite{jenkinson1955} reparameterize all three forms as special cases of a single unifying form,
-  the generalized extreme value distribution, \emph{GEV}:
-  \begin{equation*}
-    \label{eqn:gev}
-    F(m \mid \mu, \sigma, \xi) = \exp\left\lbrace-\left[1 + \xi\left(\frac{x - \mu}{\sigma}\right)\right]_{}^{-1/{\xi}}\right\rbrace.
-  \end{equation*}
-  Thus distributions in the domain of attraction of an EVD (Gumbel, Fr{\'e}chet, or Weibull) will be
-  in the domain of attraction of the GEV.
-
-As this distribution specifies asymptotic behavior for the maximum of a set of observations,
-  inference assuming this distribution requires we that specify some block of data, and for the block
-  report only the maximum.  A series of these blocks yielding a series of maxima allows us to conduct
-  inference about the parameters of the distribution.  Taking only the
-  maximum in a block of observations necessarily leads to the reduction of our sample size by a factor
-  of $1/\text{block size}$. In problems where the data occur in natural blocks, such as an hourly time
-  series where a natural block might be a day, this might be appropriate.  There is an implicit
-  violation of the assumption of independence within a block, in most cases, but that violation is
-  generally ignored. In data without a natural block, it may be difficult to justify an induced
-  artitifial block. Moreover, by reducing available information, retaining only one datum for each
-  block increases the variability of the parameter estimates.
-
-\subsubsection{Univariate exceedences over a high threshold}
-The \emph{Pickands-Balkema-de Haan Theorem}~\citep{balkema1974,pickands1975} offers us another means
-  of conducting inference that will be less wasteful of data.  It states that if the original distribution
-  $F$ is in the domain of attraction of the GEV, then excesses over a high threshold $u$ can be well
-  modelled using the Generalized Pareto distribution.  Again, let $X$ follow some distribution function
-  $F$. It follows that:
-  \begin{equation*}
-    \text{Pr}\left[X > u + y\mid X > u\right] = \frac{1 - F(u + y)}{1 - F(u)}
-  \end{equation*}
-  for $y > 0$.  If $F$ is in the domain of attraction of an EVD, then
-  \begin{equation*}
-    \lim\limits_{u\to u^{\prime}}\text{Pr}\left[X > u + y\mid X > u\right] = H(y)
-  \end{equation*}
-  where $u^{\prime}$ represents some upper boundary of $X$ has a functional form---the survival
-  function of a Pareto distribution. We approximate this asymptotic relationship by setting some
-  large threshold $u$, and then threshold exceedences $Y = X - u$ are modelled under the distrribution
-  function
-  \begin{equation*}
-    \label{eqn:gp}
-    H(y) = 1 - \left(1 + \xi\frac{y}{\sigma}\right)^{-\frac{1}{\xi}}.
-  \end{equation*}
-  This defines the generalized Pareto family of distributions.  Thus, if block maxima have a
-  limiting distribution $G$ within the EVD family, then threshold exceedances for a sufficiently
-  high threshold have a limiting distribution $H$ within the Generalized Pareto (GP) family.  We can
-  identify $\sigma$ as a scale parameter, but $\xi$ deserves special mention as the extremal index.
-  We can interpret the Pareto tail probability as
-  \begin{equation*}
-    \lim\limits_{t\to\infty}\frac{1 - F(ty)}{1 - F(t)} = y^{-\frac{1}{\xi}}
-  \end{equation*}
-  for $y > 1$.  Importantly, $\xi$ from the limiting distribution of excesses over a threshold is the
-  same parameter, and has the same value as the $\xi$ from the limiting distribution of the maximum
-  observation of a sample \citep{dehaan2006}.
-
-One other point of note is that for time series, the data are temporally correlated.  Thus the implicit
-  assumption of independence for those observations in excess of a threshold is violated.  One means
-  of ameliorating the effect of this correlation is to decluster the data---and one method of declustering
-  is, for a series of observations in excess of the threshold, to keep only the maximum observed value
-  in the series.
-
-\subsubsection{Multivariate threshold EVT}
 For the remainder of this paper, we adopt the operators $\wedge$ to denote minima, and the $\vee$
@@
 
-The generalization of EVT to multiple dimensions presents a unique challenge in the case of maxima---how
-  do we define the maximum of a sample?  How can we order observations in a multivariate scenario?
-  \cite{rootzen2006} offers a brief review of the subject.  One possible treatment sidesteps this
-  conundrum by using component maxima---let $\bm{M}_n = \left(\vee_i X_{i1},\ldots,\vee_i X_{id}\right)$.
-  Then the multivariate GEV establishes the limiting behavior of $\bm{M}_n$ as $n\to\infty$.  Note
-  that $\bm{M}_n$ need not be in the observed $\left\lbrace\bm{X}_i; i = 1,\ldots,n\right\rbrace$.
-  Then, if there is a set of constants $\bm{a}_n > \bm{0}$, $\bm{b}_n$ such that
-  \begin{equation*}
-    \lim\limits_{n\to\infty}\text{P}\left(\bigcap_{\ell = 1}^d
-      \frac{M_{n\ell} - b_{n\ell}}{a_{n\ell}} \leq x_{\ell}\right) = G(\bm{x})
-  \end{equation*}
-  with non-degenerate marginals, then $G(\bm{x})$ is a \emph{multivariate extreme value distribution}.
-  The marginal distribution of $M_{nl}$ is by construction a GEV with the parameters $a_{n\ell}, b_{n\ell}, \xi$.
-  With the existence of the marginal distributions established as a consequence of the existence of the
-  GEV, multivariate EVT then splits into two components---estimation of the marginal parameters, and
-  estimation of the dependence structure.  This separation is born out extensively in the literature.
+\section{A model for the multivariate GP}
+%\subsubsection{Univariate Maxima}
+%Regarding the asymptotic behavior of extreme events, there are a couple major strategies for conducting
+%  inference---developing probabilistic estimates of extreme behavior.  First developed is the theory
+%  of a limiting distribution on maxima---the largest observation from a sample.  For a sample $\bm{x}$
+%  where $\bm{x} = (x_1,\ldots,x_n)$ represents a sequence of $n$ independent random variables from a
+%  distribution function $F$, the distribution of the maximum $M_n$ of this sequence can be derived
+%  as:
+%  \begin{equation*}
+%    \begin{aligned}
+%      \text{Pr}\left[M_n\leq z\right] &= \text{Pr}\left[X_1 \leq z, \ldots, X_n \leq z\right]\\
+%        &= \text{Pr}\left[X_1\leq z\right]\times\ldots\times\text{Pr}\left[X_n\leq z\right]\\
+%        &= F(z)^n.
+%    \end{aligned}
+%  \end{equation*}
+%  In situations where $F$ is unknown, we can seek to approximate the behavior of $F^n$ as
+%  $n\rightarrow\infty$.  To ensure this does not degenerate to a point mass, we need to consider a
+%  standardized sequence of maxima. If this sequence stabilizes as $n$ increases, the a limiting
+%  distribution exists.  More specifically, if there exists some sequence of constants $a_n > 0$, $b_n$
+%  such that:
+%  \begin{equation*}
+%    \text{Pr}\left[\frac{M_n - b_n}{a_n} \leq z\right] \stackrel{d}{\rightarrow} G(z)
+%  \end{equation*}
+%  as $n\rightarrow\infty$, then $G$ is a max-stable distribution, and $F$ is in the domain of
+%  attraction of that max stable distribution.  \cite{frechet1927} originates the
+%  field, identifying two limiting forms of max-stable distributions, which would become known as the
+%  Fr{\'e}chet and Weibull distributions. \cite{weibull1951} expands the analysis of the Weibull
+%  distribution.  \cite{fisher1928}, in what is now called the \emph{Fisher-Tippett Theorem}, identifies
+%  the Fr{\'e}chet and Weibull distributions, along with an as then unnamed third form, as the three
+%  limiting forms of the distribution of the maxima of a sample.  \cite{gumbel1935,gumbel1942} offer
+%  analysis of that third form, now known as the Gumbel distribution.  Later works, including
+%  \cite{jenkinson1955} reparameterize all three forms as special cases of a single unifying form,
+%  the generalized extreme value distribution, \emph{GEV}:
+%  \begin{equation*}
+%    \label{eqn:gev}
+%    F(m \mid \mu, \sigma, \xi) = \exp\left\lbrace-\left[1 + \xi\left(\frac{x - \mu}{\sigma}\right)\right]_{}^{-1/{\xi}}\right\rbrace.
+%  \end{equation*}
+%  Thus distributions in the domain of attraction of an EVD (Gumbel, Fr{\'e}chet, or Weibull) will be
+%  in the domain of attraction of the GEV.
 
-As in the multivariate extrema case, analysis in peaks over thresholds modelling generally first develops
-  the marginal distributions, then establishes the dependence structure.  In extreme analysis, a frequently
-  used method for describing the dependence
-  structure is the copula---a family of multivariate cumulative distribution functions with uniform
-  marginal distributions~\cite{renard2007,deng2011,falk2019},
-  but this presents its own difficulties in modelling and interpretation.  Instead, we propose a novel
-  method based on the constructive definition of the multivariate Pareto presented in
-  \cite{ferreira2014}, where we assume that the $d$-dimensional vector $\bm{Z}$, after standardization
-  is such that for some distribution $\mu$ defined on $\mathcal{R}_+^d$,
-  \begin{equation}
-    \lim\limits_{n\to\infty}n\text{Pr}\left[\frac{1}{n}\bm{Z}\geq \bm{z}\right] = \mu\left([\bm{0},\bm{z}]^c\right).
+%As this distribution specifies asymptotic behavior for the maximum of a set of observations,
+%  inference assuming this distribution requires we that specify some block of data, and for the block
+%  report only the maximum.  A series of these blocks yielding a series of maxima allows us to conduct
+%  inference about the parameters of the distribution.  Taking only the
+%  maximum in a block of observations necessarily leads to the reduction of our sample size by a factor
+%  of $1/\text{block size}$. In problems where the data occur in natural blocks, such as an hourly time
+%  series where a natural block might be a day, this might be appropriate.  There is an implicit
+%  violation of the assumption of independence within a block, in most cases, but that violation is
+%  generally ignored. In data without a natural block, it may be difficult to justify an induced
+%  artitifial block. Moreover, by reducing available information, retaining only one datum for each
+%  block increases the variability of the parameter estimates.
+
+%\subsubsection{Univariate exceedences over a high threshold}
+%The \emph{Pickands-Balkema-de Haan Theorem}~\citep{balkema1974,pickands1975} offers us another means
+%  of conducting inference that will be less wasteful of data.  It states that if the original distribution
+%  $F$ is in the domain of attraction of the GEV, then excesses over a high threshold $u$ can be well
+%  modelled using the Generalized Pareto distribution.  Again, let $X$ follow some distribution function
+%  $F$. It follows that:
+%  \begin{equation*}
+%    \text{Pr}\left[X > u + y\mid X > u\right] = \frac{1 - F(u + y)}{1 - F(u)}
+%  \end{equation*}
+%  for $y > 0$.  If $F$ is in the domain of attraction of an EVD, then
+%  \begin{equation*}
+%    \lim\limits_{u\to u^{\prime}}\text{Pr}\left[X > u + y\mid X > u\right] = H(y)
+%  \end{equation*}
+%  where $u^{\prime}$ represents some upper boundary of $X$ has a functional form---the survival
+%  function of a Pareto distribution. We approximate this asymptotic relationship by setting some
+%  large threshold $u$, and then threshold exceedences $Y = X - u$ are modelled under the distrribution
+%  function
+%  \begin{equation*}
+%    \label{eqn:gp}
+%    H(y) = 1 - \left(1 + \xi\frac{y}{\sigma}\right)^{-\frac{1}{\xi}}.
+%  \end{equation*}
+%  This defines the generalized Pareto family of distributions.  Thus, if block maxima have a
+%  limiting distribution $G$ within the EVD family, then threshold exceedances for a sufficiently
+%  high threshold have a limiting distribution $H$ within the Generalized Pareto (GP) family.  We can
+%  identify $\sigma$ as a scale parameter, but $\xi$ deserves special mention as the extremal index.
+%  We can interpret the Pareto tail probability as
+%  \begin{equation*}
+%    \lim\limits_{t\to\infty}\frac{1 - F(ty)}{1 - F(t)} = y^{-\frac{1}{\xi}}
+%  \end{equation*}
+%  for $y > 1$.  Importantly, $\xi$ from the limiting distribution of excesses over a threshold is the
+%  same parameter, and has the same value as the $\xi$ from the limiting distribution of the maximum
+%  observation of a sample \citep{dehaan2006}.
+
+%One other point of note is that for time series, the data are temporally correlated.  Thus the implicit
+%  assumption of independence for those observations in excess of a threshold is violated.  One means
+%  of ameliorating the effect of this correlation is to decluster the data---and one method of declustering
+%  is, for a series of observations in excess of the threshold, to keep only the maximum observed value
+%  in the series.
+
+%\subsubsection{Multivariate threshold EVT}
+
+%The generalization of EVT to multiple dimensions presents a unique challenge in the case of maxima---how
+%  do we define the maximum of a sample?  How can we order observations in a multivariate scenario?
+%  \cite{rootzen2006} offers a brief review of the subject.  One possible treatment sidesteps this
+%  conundrum by using component maxima---let $\bm{M}_n = \left(\vee_i X_{i1},\ldots,\vee_i X_{id}\right)$.
+%  Then the multivariate GEV establishes the limiting behavior of $\bm{M}_n$ as $n\to\infty$.  Note
+%  that $\bm{M}_n$ need not be in the observed $\left\lbrace\bm{X}_i; i = 1,\ldots,n\right\rbrace$.
+%  Then, if there is a set of constants $\bm{a}_n > \bm{0}$, $\bm{b}_n$ such that
+%  \begin{equation*}
+%    \lim\limits_{n\to\infty}\text{P}\left(\bigcap_{\ell = 1}^d
+%      \frac{M_{n\ell} - b_{n\ell}}{a_{n\ell}} \leq x_{\ell}\right) = G(\bm{x})
+%  \end{equation*}
+%  with non-degenerate marginals, then $G(\bm{x})$ is a \emph{multivariate extreme value distribution}.
+%  The marginal distribution of $M_{nl}$ is by construction a GEV with the parameters $a_{n\ell}, b_{n\ell}, \xi$.
+%  With the existence of the marginal distributions established as a consequence of the existence of the
+%  GEV, multivariate EVT then splits into two components---estimation of the marginal parameters, and
+%  estimation of the dependence structure.  This separation is born out extensively in the literature.
+
+To develop a multivariate PoT model for extreme values consider a $d$-dimensional random vector $\bm{Z}$, whose marginal distributions have been standardized.
+Thus $\bm{Z} \in {\mathbb R}^d_+ = [0,+\infty]^d$, the positive $d$-dimensional cone.
+Following the discussion in \cite{goix2017} we assume a form of regular variation, consisting on assuming the existing of a limiting measure $\mu$, such that.
+\[
+    \lim\limits_{n\to\infty}n\text{Pr}\left[\frac{1}{n}\bm{Z}\geq \bm{z}\right] = \mu\left([\bm{0},\bm{z}]^c\right),
+\]
+where the inequality and the interval are taken componentwise. 
+In this setting, $\mu$ is the asymptotic distribution of $\bm{Z}$ in extreme regions, and it is referred to as the
+  \emph{exponent measure}.  $\mu$ features the homogeneity property $\mu(tA) = \frac{1}{t}\mu(A)$. This
+corresponds to the second condition in Theorem 2.1 of
+\cite{ferreira2014}, applied to the finite-dimensional case.
+Thus, we factorize $\bm{Z} =  R\bm{V}$, where
+  $R = \pnorm{\bm{Z}}{\infty} $ describes the radial component, and
+  $\bm{V} = \bm{Z}/\pnorm{\bm{Z}}{\infty} \in \mathcal{S}_{\infty}^{d-1}$ the angular component of
+  $\bm{Z}$,  and ${\mathbb S}_\infty^{d-1}$ denotes the positive
+orthant of the unit sphere in infinite
+norm.  $\mu(\cdot)$ is then factorized as
+\[
+    \mu\left( \left\lbrace\bm{Z} : V(\bm{Z}) \in A , R(\bm{Z}) > r\right\rbrace \right) = r^{-1}\Phi(A),
+\]
+  where $\Phi(\cdot)$ is the \emph{angular measure}.  The angular measure is thus independent and
+  separable from the radial component.  Furthermore,
+\begin{equation} \label{MGP}
+    \text{Pr}\left[\bm{V} \in A \mid R > r\right]
+      = \frac{r\text{Pr}\left[\bm{V} \in A, R > r\right]}{r\text{Pr}[R > r]}\hspace{0.2cm}
+      \xrightarrow[r\to\infty]{~} \hspace{0.2cm} \frac{\Phi(A)}{\Phi(\mathcal{S}_{\infty}^{d-1})}.
 \end{equation}
-  In this setting, $\mu$ is the asymptotic distribution of $\bm{Z}$ in extreme regions---the
-  \emph{exponent measure}.  $\mu$ features the homogeneity property $\mu(tA) = \frac{1}{t}\mu(A)$.
-  By this property, \cite{ferreira2014} factorize $\bm{Z}\to R\bm{V}$, where
-  $R = \pnorm{\bm{Z}}{\infty} > 1$ describes the radial component, and
-  $\bm{V} = \bm{Z}/\pnorm{\bm{Z}}{\infty} \in \mathcal{S}_{\infty}^{d-1}$ the angular component of
-  $\bm{Z}$ in $\mathcal{R}_+^d$.  $\mu(\cdot)$ is similarly factorized as
-  \begin{equation}
-    \mu\left( \left\lbrace\bm{Z} : R(\bm{Z}) > r, V(\bm{Z}) \in A \right\rbrace \right) = r^{-1}\Phi(A),
-  \end{equation}
-  where $\Phi(\cdot)$ is the \emph{spectral measure}.  The spectral measure is thus independent and
-  separable from the radial component.  Furthermore,
-  \begin{equation}
-    \text{P}\left[\bm{V} \in A \mid R > r\right]
-      = \frac{r\text{P}\left[\bm{V} \in A, R > r\right]}{r\text{P}[R > r]}\hspace{0.2cm}
-      \xrightarrow[r\to\infty]{~} \hspace{0.2cm} \frac{\Phi(A)}{\Phi(\mathcal{S}_{\infty}^{d-1})}
-  \end{equation}
-  Describing the dependence structure of $\bm{Z}$, equivalently describes the spectral measure
-  $\Phi$. To accomplish this, we seek to establish a distribution on $\bm{V} \in \mathcal{S}_{\infty}^{d-1}$.
+  Thus, the angular measure corresponds to the distribution of $\bm{V}$, conditional on $\bm{R}$ being large.
+  As such, it controls the dependence structure of $\bm{Z}$ in the tails. To obtain an EVT model for $\bm{Z}$, we seek to establish a flexible distribution on $\bm{V} \in \mathcal{S}_{\infty}^{d-1}$.
+
+An alternative approach to defining the multivariate GP in Equation \eqref{MGP} is provided by Theorem 6 in \cite{RoSeWa2018a}.
+Using the notation in that theorem, define $\bm{S} = \log(\bm{V})$.
+Then $\bm{S}$ is a spectral random vector, according to Definition 5.
+Taking $E \sim \text{Exp}(1)$, independent of $\bm{V}$, we have that $\bm{Y} = \bm{S} + E$ follows a multivariate GP, according to Theorem 6.
+Then, using Proposition 4, $\bm{Z} = e^{Y}$ is also a multivariate GP.
+Notice that $\bm{R} = e^E$ is distributed as a standard Pareto random variable.
 

### 2021-09-15 evaluation.tex
evaluation

\section{Model Comparison on the Hypersphere}
\label{sec:evaluation}
It is not immediately obvious what criteria to use to judge these models and decide which best
  represents the data's generating distribution.  As the task resides on $\mathcal{S}_{\infty}^{d-1}$,
  our options are somewhat limited in terms of what model criteria we can use.  Because of the aforementioned
  difficulty in evaluating a density directly in this space, we choose to use model evaluation
  criteria that can operate on samples from the posterior predictive distribution.  For this reason,
  we have opted to use the \emph{posterior predictive loss} criterion of \cite{gelfand1998} and the
  \emph{energy score} criterion of \cite{gneiting2007}.  Both of these metrics require calculating
  some distance in the target space, and this section will be devoted to that end.  As both these
  metrics operate on the posterior predictive distribution for each observation, we will also be
  including a metric operating on the overall posterior predictive distribution, in the form of a
  modified Kullbeck-Liebler divergence.

\subsection{Posterior Predictive Loss Criterion}
The posterior predictive loss criterion, \emph{PPL} is introduced in \cite{gelfand1998}.  When we
  assume a squared error loss function, then for the $i$th observation, the posterior predictive
  loss criterion is computed as
  \begin{equation}
    \label{eq:ppl}
    D_{k}^{(i)} = \text{Var}\left(X_i\right) + \frac{k}{k + 1}\left(\text{E}[X_i] - x_i\right)^2,
  \end{equation}
  where $X_i$ is a random variable from the posterior predictive distribution for $x_i$.  The
  scalar $k$ is a weighting factor by which we arbitrarily scale the importance of goodness of fit
  relative to precision.  In our analysis, we take the limit as $k\to\infty$, weighting both
  parts equally.  Interpreting this criterion, a smaller $\text{Var}(\bm{X}_l)$ indicates a higher
  precision, and a smaller $(\text{E}[\bm{X}_l] - \bm{x}_l)^2$  indicates a better fit.  Thus,
  smaller is better.  Note that this is defined for a univariate distribution.  As we are dealing
  with a multivariate distribution, we will be taking the posterior predictive loss
  summed over all dimensions.  That is,
  \begin{equation}
    \label{eq:ppl2}
    D_k^{(i)} = \frac{1}{d}\sum_{l = 1}^{d}\left[\text{Var}(X_{il}) + \frac{k}{k+1}\left(\text{E}[X_{il}] - x_{il}\right)^2\right]
  \end{equation}
  Then we report the average $D_k$, taken over all $i$.

\subsection{Energy Score}
The energy score of Gneiting, et al\cite{gneiting2007}, is a generalization of the continuous ranked
  probability score, or \emph{crps}, defined for a multi-dimensional random variable.
  \begin{equation}
    \label{eq:es}
    \text{ES}\left(P,x_i\right) =  \text{E}_p g\left(\bm{X}_i, \bm{x}_i\right) - \frac{1}{2}\text{E}_p g\left(\bm{X}_i,\bm{X}_i^{\prime}\right)
  \end{equation}
  where $g(\cdot)$ identifies a negative definite kernel, and $\bm{X}_i^{\prime}$ represents another
  replicate from the posterior predictive distribution of $\bm{x}_i$. A function $g$ is a negative definite kernel if it is symmetric in its arguments, $g(x_1,x_2) = g(x_2,x_1)$, and for which
  $\sum_{i =1}^n\sum_{j=1}^n\alpha_i\alpha_jg(x_1,x_2) \leq 0$ for all positive integers n, with the
  restriction that $\sum_{i=1}^n\alpha_i = 0$.  Euclidean distance is one example of a negative definite kernel,
  and the one most often used.  However,   on $\mathcal{S}_p^{d-1}$ for $p > 1$, Euclidean distance
  can under-report the actual distance required for travel between two points. That distortion is
  maximized on $\mathcal{S}_{\infty}^{d-1}$.

\subsection{Distance on the positive orthant of the Hypercube}
  \label{subsec:distance}
The positive orthant of the unit hypercube, defined in Euclidean geometry, is that structure for
  which, in a given point on the hypercube, all dimensions of that point are between 0 and 1, and
  at least one dimension must be 1.  Developing terminology, we can consider observations for which
  the $j$th dimension is equal to 1, to be on the $j$th \emph{face}.  The intersection of the $i$th
  and $j$th face is a $d-2$ dimensional cube, and observations in this space have dimensions $i,j$
  equal to 1.

A \emph{distance} in this space is a \emph{geodesic} on this space. From geometry, we know that
  the geodesic, or shortest path between 2 points along the surface of a $d$ dimensional figure
  corresponds to at least one \emph{unfolding}, or \emph{rotation} of the $d$-dimensional
  figure into a $d-1$ dimensional space.  The appropriate term for the structure generated by this
  unfolding is a \emph{net}.  For the appropriate net, a line segment connecting the two points and
  staying within the boundaries of the net corresponds to the shortest path between those points\citep{pappas1989}, and is thus a geodesic.  The length of that line segment is properly
  defines the distance required for travel between those points.

Consider a 3-dimensional cube.  Consider 2 points on this 3-dimensional cube,
  $\bm{a}_1 = (x_1,y_1,z_1)$, and $\bm{a}_2 = (x_2,y_2,z_2)$. Let's say that the two points are
  on the same face.  Then the distance between those two points, the distance one has to travel
  along the space to move from one point to the other, is calculated by Euclidean norm.  Now,
  consider two points on separate faces.  All faces are pairwise adjacent, as we have stated, so
  in order to move to the other point, we must \emph{at least} move to the intersection between the
  faces, then to the other point.  Let $\bm{a}_1$ lie on the $x$ face, and $\bm{a}_2$ lie on the
  $y$ face.  That is, $\bm{a}_1 = (1, y_1, z_1)$, $\bm{a}_2 = (x_2, 1, z_2)$.  Then traveling
  between these points we must at least pass through the intersection of faces $x,y$.  One possible
  net representation of this is unfolding the $y$ face alongside the $x$ face.  We accomplish this
  by applying a rotation and translation to $a_2$, corresponding to the following:
  \begin{equation}
    \label{eq:1drotation}
    a_2^{\prime} = \begin{bmatrix}
    1 \\
    2 \\
    0
    \end{bmatrix}
    +
    \begin{bmatrix}
    0  & 0 & 0 \\
    -1 & 0 & 0 \\
    0  & 0 & 1
    \end{bmatrix}
    \begin{bmatrix}
    x_2 \\
    1 \\
    z_2 \\
    \end{bmatrix} = \begin{bmatrix}
    1 \\
    2 - x_2 \\
    z_2
    \end{bmatrix}
  \end{equation}
  Then, if this is the appropriate net, the distance from $\bm{a}$ to $\bm{b}$ becomes
  $\lVert \bm{a}_1 - \bm{a}_2^{\prime}\rVert_2$. However, there
  is another possible net we must consider, travelling first through the $z$ face then to the
  $y$ face.  The rotation for that becomes:
  \begin{equation}
    \label{eq:2drotation}
    a_2^{\prime} = \begin{bmatrix}
    1 \\
    2 \\
    2
    \end{bmatrix}
    +
    \begin{bmatrix}
    0 & 0 & 0  \\
    0 & 0 & -1 \\
    -1 & 0 & 0
    \end{bmatrix}
    \begin{bmatrix}
    x_2 \\
    1 \\
    z_2
    \end{bmatrix} = \begin{bmatrix}
    1 \\
    2 - z_2 \\
    2 - x_2
    \end{bmatrix}
  \end{equation}
Every successive rotation is relative to the last face.  So, as the number of dimensions grows, the
  number of possible rotations grows as well.  As we have $d$ faces, if 2 observations are on
  different faces, then there are $\sum_{j = 1}^{d-2}\binom{d-2}{j} + 1$ possible rotations to
  consider\footnote{There are truly $d!$ possible nets, but when we consider starting and ending
  faces fixed, and that portions of the net that diverge after the ending face are irrelevant,
  we arrive at that number of rotations that we need consider}.  While this is not
  insurmountable, it is computationally costly. However, for a valid energy score, a true distance
  is not necessary.  What is needed is a negative definite kernel.

\begin{prop}
For points $a,b \in \mathcal{S}_{\infty}^{d-1}$ a valid negative definite kernel can be formed as
  \begin{equation}
    g(\bm{a},\bm{b}) = \begin{cases}
        \pnorm{\bm{b}-\bm{a}}{2} &\text{ if }\argmax_l\bm{a} = \argmax_l\bm{b}\\
        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} &\text{ otherwise}
    \end{cases}
  \end{equation}
  where $\bm{c}$ resides on the intersection between the faces of $\bm{a}$ and $\bm{b}$, and minimizes $g(\bm{a},\bm{b})$.
\end{prop}

In the first case, where $\bm{a}$ and $\bm{b}$ reside on the same face, then that is trivially a
  negative definite kernel.  In the second case, where $\bm{a}$ and $\bm{b}$ reside on separate
  faces, we can be assured of symmetry in the arguments as $\bm{c}$ is unique, and $\pnorm{\bm{c} - \bm{a}}{2} = \pnorm{\bm{a} - \bm{c}}{2}$.  Then for the other property of negative definiteness,
  \begin{equation*}
    \begin{aligned}
      \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j g(\bm{x}_1,\bm{x}_2) &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j \bigg[\pnorm{\bm{c} - \bm{x}_1}{2} + \pnorm{\bm{x}_2 - \bm{c}}{2}\bigg]\\
      &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{c} - \bm{x}_1}{2} + \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{x}_2 - \bm{c}}{2}
    \end{aligned}
  \end{equation*}
  which is less than $0$ in both terms by definition.

To compute this energy score as described would require optimization of the interim point $\bm{c}$.
  If we again consider the one-dimensional rotation in Equation~\ref{eq:1drotation}--we
  can view that as a numerically easier analogue accomplishing the same goal: the distance from
  the starting point, to some optimal point along the intersection between the starting and
  ending faces, to the ending point.  The strait line will pass through the optimal point, so
  the optimization of the interim point is accomplished \emph{en passant}.

% EOF

### 2021-09-15 evaluation.tex
evaluation edits

@@
   relative to precision.  In our analysis, we take the limit as $k\to\infty$, weighting both
-  parts equally.  Interpreting this criterion, a smaller $\text{Var}(\bm{X}_l)$ indicates a higher
-  precision, and a smaller $(\text{E}[\bm{X}_l] - \bm{x}_l)^2$  indicates a better fit.  Thus,
+  parts equally.  Interpreting this criterion, a smaller $\text{Var}(\bm{X}_{\ell})$ indicates a higher
+  precision, and a smaller $(\text{E}[\bm{X}_{\ell}] - \bm{x}_{\ell})^2$  indicates a better fit.  Thus,
   smaller is better.  Note that this is defined for a univariate distribution.  As we are dealing
@@
     \label{eq:ppl2}
-    D_k^{(i)} = \frac{1}{d}\sum_{l = 1}^{d}\left[\text{Var}(X_{il}) + \frac{k}{k+1}\left(\text{E}[X_{il}] - x_{il}\right)^2\right]
+    D_k^{(i)} = \frac{1}{d}\sum_{\ell = 1}^{d}\left[\text{Var}(X_{i,\ell}) +
+                  \frac{k}{k+1}\left(\text{E} [X_{i,\ell}] - x_{i,\ell}\right)^2\right]
   \end{equation}
@@
     \label{eq:es}
-    \text{ES}\left(P,x_i\right) =  \text{E}_p g\left(\bm{X}_i, \bm{x}_i\right) - \frac{1}{2}\text{E}_p g\left(\bm{X}_i,\bm{X}_i^{\prime}\right)
+    \text{ES}\left(P, \bm{x}_i\right) =  \text{E}_p g\left(\bm{X}_i, \bm{x}_i\right) -
+                                    \frac{1}{2}\text{E}_p g\left(\bm{X}_i,\bm{X}_i^{\prime}\right)
   \end{equation}
@@
   replicate from the posterior predictive distribution of $\bm{x}_i$. A function $g$ is a negative definite kernel if it is symmetric in its arguments, $g(x_1,x_2) = g(x_2,x_1)$, and for which
-  $\sum_{i =1}^n\sum_{j=1}^n\alpha_i\alpha_jg(x_1,x_2) \leq 0$ for all positive integers n, with the
+  $\sum_{i =1}^n\sum_{j=1}^n\alpha_i\alpha_jg(\bm{x}_1,\bm{x}_2) \leq 0$ for all positive integers n, with the
   restriction that $\sum_{i=1}^n\alpha_i = 0$.  Euclidean distance is one example of a negative definite kernel,
@@
     g(\bm{a},\bm{b}) = \begin{cases}
-        \pnorm{\bm{b}-\bm{a}}{2} &\text{ if }\argmax_l\bm{a} = \argmax_l\bm{b}\\
+        \pnorm{\bm{b}-\bm{a}}{2} &\text{ if }\argmax_{\ell}\bm{a} = \argmax_{\ell}\bm{b}\\
         \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} &\text{ otherwise}
@@
 To compute this energy score as described would require optimization of the interim point $\bm{c}$.
-  If we again consider the one-dimensional rotation in Equation~\ref{eq:1drotation}--we
+  If we again consider the one-dimensional rotation in Equation~\ref{eq:1drotation}---we
   can view that as a numerically easier analogue accomplishing the same goal: the distance from

### 2021-09-16 evaluation.tex
bruno's edits 2

@@
-\section{Model Comparison on the Hypersphere}
+\section{Scores for distributions in $\mathcal{S}_\infty^{d-1}$}
 \label{sec:evaluation}
+
+{\bf After a sentence or two motivating the need for scoring distributions in $\mathcal{S}_\infty^{d-1}$ that can be calculated with a sample-based approach, mention the paper of \cite{gneiting2007} on proper scoring rules, and introduce the energy score.
+Cite the result in that paper that gives conditions for the energy score to be a proper scoring rule.
+Then discuss the problem of finding a metric in $\mathcal{S}_\infty^{d-1}$ that corresponds to a negative definite kernel.
+That is, the material in Section 4.3.
+Introduce and discuss the idea, the write the proposition, with its proof.
+I don't know if the example in the 3-dimensional cube is needed or not.
+Finally, say how the score will be calculated in our case.
+}
 It is not immediately obvious what criteria to use to judge these models and decide which best

### 2021-09-16 introduction.tex
bruno's edits 2

@@
 The former leads to a Generalized Extreme Value (GEV) distribution, that depends on three parameters.
-The latter, which is the focus of this paper, leads to a Generalized Pareto (GP) distribution, that depends on a shape and a scale parameter.
+The latter leads to a Generalized Pareto (GP) distribution, that depends on a shape and a scale parameters.
 Likelihood-based approaches to inference can be readily implemented in both cases.
 In the multivariate case the GEV theory is well established \citep[see, for example][]{dehaan2006}, but the inferential problem is more complicated, as there is no parametric representation of the GEV.
-This problem is inherited by the PoT approach and compunded by the fact that there is no unique definition of an exceedance of a multivariate threshold, as there is an obvious dependence on the norm that used to measure the size of a vector.
+This problem is inherited by the PoT approach and compounded by the fact that there is no unique definition of an exceedance of a multivariate threshold, as there is an obvious dependence on the norm that is used to measure the size of a vector.
 
@@
 {\bf is \cite{deng2011} really a relevant reference?
-It looks weird.}
+It looks weird.
+Can we say something more about the copula approach?
+I confess my ignorance here.}
   \cite{ferreira2014} presents a constructive definition of the Pareto process, that generalizes the GP to an infinite dimensional setting.
@@
   contains the information pertaining to the dependence structure of the random vector.
-  Based on this definition, we present a novel approach for modelling the distribution of the angular component
+  Based on this definition, we present a novel approach for modelling  the angular component
   with families of distributions that provide flexibility and can be applied in a moderately large dimensional setting.
-  Models for the angular measure are important for multivariate EVT applications.
-\cite{SaNa2014} and \cite{HaCaCh2017} consider Bayesian non-parametric approaches to model the angular measure of the multivariate GEV.
-Particular attention is paid to the problem of satisfying the so called moment constraint.
+  Our focus on the  angular measure is similar to that in \cite{SaNa2014} and \cite{HaCaCh2017}, that consider Bayesian non-parametric approaches.
+Yet, our approach avoids the problem of dealing with the so called moment constraint by considering a constructive definition of the multivariate GP based on the infinity norm.
 
   The reminder of this paper is outlined as follows.
+{\bf We need a roadmap.
+Part of the roadmap should mention the analysis of the IVT.
+Basically condensing the paragraph below in a couple of sentences.}
 
@@
 
-For the remainder of this paper, we adopt the operators $\wedge$ to denote minima, and the $\vee$
+Throughout the paper, we adopt the operators $\wedge$ to denote minima, and the $\vee$
   to denote maxima.  Thus $\wedge_i x_i = \min_i x_i$, $\vee_i x_i = \max_i x_i$.  These operators can
@@
 
-\section{A model for the multivariate GP}
+\section{A multivariate PoT model}
 %\subsubsection{Univariate Maxima}
@@
 
-To develop a multivariate PoT model for extreme values consider a $d$-dimensional random vector $\bm{Z}$, whose marginal distributions have been standardized.
-Thus $\bm{Z} \in {\mathbb R}^d_+ = [0,+\infty]^d$, the positive $d$-dimensional cone.
-Following the discussion in \cite{goix2017} we assume a form of regular variation, consisting on assuming the existing of a limiting measure $\mu$, such that.
+To develop a multivariate PoT model for extreme values consider a $d$-dimensional random vector $\bm{Z} \in {\mathbb R}^d_+ = [0,+\infty)^d$, the positive $d$-dimensional cone.
+Following the discussion in \cite{goix2017} we assume a form of regular variation, consisting on assuming the existence of a limiting measure $\mu$, such that.
 \[
@@
 In this setting, $\mu$ is the asymptotic distribution of $\bm{Z}$ in extreme regions, and it is referred to as the
-  \emph{exponent measure}.  $\mu$ features the homogeneity property $\mu(tA) = \frac{1}{t}\mu(A)$. This
-corresponds to the second condition in Theorem 2.1 of
-\cite{ferreira2014}, applied to the finite-dimensional case.
-Thus, we factorize $\bm{Z} =  R\bm{V}$, where
-  $R = \pnorm{\bm{Z}}{\infty} $ describes the radial component, and
-  $\bm{V} = \bm{Z}/\pnorm{\bm{Z}}{\infty} \in \mathcal{S}_{\infty}^{d-1}$ the angular component of
-  $\bm{Z}$,  and ${\mathbb S}_\infty^{d-1}$ denotes the positive
-orthant of the unit sphere in infinite
-norm.  $\mu(\cdot)$ is then factorized as
+  \emph{exponent measure}.  $\mu$ features the homogeneity property $\mu(tA) = \frac{1}{t}\mu(A)$. 
+Assume that $R = \max_{i} Z_i>1$ with probability one, and that $V_i = Z_i/\max_i Z_i$ is such that $E(V_i)>0$, then
+we have that the second condition in Theorem 2.1 of
+\cite{ferreira2014}, applied to the finite-dimensional case, is satisfied.
+Thus, we have the factorization $\bm{Z} =  R\bm{V}$, where $R$ and $\bm{V}$  are independent. Thus,
+for some measure $\Phi$, $\mu(\cdot)$ is factorized as
 \[
-    \mu\left( \left\lbrace\bm{Z} : V(\bm{Z}) \in A , R(\bm{Z}) > r\right\rbrace \right) = r^{-1}\Phi(A),
+    \mu\left( \left\lbrace\bm{Z} : V \in A , R > r\right\rbrace \right) = r^{-1}\Phi(A).
 \]
-  where $\Phi(\cdot)$ is the \emph{angular measure}.  The angular measure is thus independent and
-  separable from the radial component.  Furthermore,
+$R$ describes the radial component, and $\bm{V} \in \mathcal{S}_{\infty}^{d-1}$ the 
+angular component of $\bm{Z}$.  Here ${\mathbb S}_\infty^{d-1}$ denotes the positive
+orthant of the unit sphere in infinite norm.  $\Phi(\cdot)$ is referred to as the \emph{angular measure}.   
+Moreover, $R$ is distributed as a standard Pareto random variable. Conditioning on $R$, we have that
 \begin{equation} \label{MGP}
@@
 \end{equation}
-  Thus, the angular measure corresponds to the distribution of $\bm{V}$, conditional on $\bm{R}$ being large.
-  As such, it controls the dependence structure of $\bm{Z}$ in the tails. To obtain an EVT model for $\bm{Z}$, we seek to establish a flexible distribution on $\bm{V} \in \mathcal{S}_{\infty}^{d-1}$.
+ Thus, the angular measure corresponds to the distribution of $\bm{V}$, conditional on $R$ being large.
+ As such, it controls the dependence structure of $\bm{Z}$ in the tails. In fact, following \cite{ferreira2014} the distribution function of $\bm{Z}$ is given as
+\begin{equation*}
+	\text{Pr}[Z_1\leq z_1, \ldots ,Z_d\leq z_d] = E\left(\bigvee_{i=1}^d
+\frac{V_i}{z_i\wedge 1}\right) - E\left(\bigvee_{i=1}^d
+\frac{V_i}{z_i}\right) 
+\end{equation*}
+where the expectation is taken with respect to $\Phi$.
+In view of this, to obtain a PoT model for $\bm{Z}$, we seek 
+a flexible distribution on $\bm{V} \in \mathcal{S}_{\infty}^{d-1}$.
 
-An alternative approach to defining the multivariate GP in Equation \eqref{MGP} is provided by Theorem 6 in \cite{RoSeWa2018a}.
-Using the notation in that theorem, define $\bm{S} = \log(\bm{V})$.
-Then $\bm{S}$ is a spectral random vector, according to Definition 5.
-Taking $E \sim \text{Exp}(1)$, independent of $\bm{V}$, we have that $\bm{Y} = \bm{S} + E$ follows a multivariate GP, according to Theorem 6.
-Then, using Proposition 4, $\bm{Z} = e^{Y}$ is also a multivariate GP.
-Notice that $\bm{R} = e^E$ is distributed as a standard Pareto random variable.
+\cite{ferreira2014} refer to the model defined in Equation \eqref{MGP} as a simple 
+Pareto process indexed on a finite dimensional set. An alternative approach to 
+define such distribution that provides a link to the multivariate GEV is discussed 
+in Remark 1 of \cite{RoSeWa2018a}. More specifically, 
+using the notation in Theorem 6 of \cite{RoSeWa2018a}, define $\bm{S} = \log(\bm{V})$. 
+Then, according to Definition 5, $\bm{S}$ is a spectral random vector. Taking 
+$E \sim \text{Exp}(1)$, independent of $\bm{V}$, we have that $\bm{Y} = \bm{S} + E$ 
+follows a multivariate GP, according to Theorem 6.  Then, using Proposition 4, 
+$\bm{Z} = e^{Y}$ is also a multivariate GP. Notice that $R = e^E$ is distributed as 
+a standard Pareto random variable.
 
-\subsection{Integrated Vapor Transport}
-Atmospheric rivers are temporary events, where large elongated regions of high concentrations of
-  water vapor are developed in the atmosphere and carry huge amounts of water potentially thousands
-  of miles.  The amount of water in transit during these events can dwarf that of terrestrial rivers.
-  For the targeted region, the atmospheric river can represent a significant portion of the
-  precipitation the region will experience.  Thus, such events are of great interest to meteorologists,
-  hydrologists, actuaries, land-use planners, and even farmers.  The ability to estimate the likely magnitude
-  and distribution of such events is thus of great importance~\citep{ralph2013,ralph2018}.
+A measure that is used to characterize the strength of the dependence,
+in the tail, for two random variables $W_1$ and $W_2$, with marginal
+distributions $F_1$ and $F_2$ is given by
+\[	\chi_{12} = \lim_{u\uparrow 1} \text{Pr}[ F_1(W_1)>u|F_2(W_2)>u]  \]
+\citep{coles2001}.
+$\chi_{12}$ provides information about the distribution of extremes for the variable $W_1$
+given that $W_2$ is very large.  When $\chi_{12}>0$, $W_1$ and $W_2$ 
+are said to
+be asymptotically dependent, otherwise they are asymptotically
+independent. The following result provides the asymptotic dependence coefficient for two components of $\bm{Z}$. 
+\begin{prop} \label{ppchi}
+Suppose that $\bm{Z} = R\bm{V}$ with $R\sim Pa(1)$,
+$\text{Pr}[V_i>0] = 1$ and $E(V_i)$ exists ,
+$i=1, \ldots ,d$, then
+\begin{equation}
+	\chi_{ij} = E\left(\frac{V_i}{E(V_i)} \wedge 
+\frac{V_j}{E(V_j)}\right).
+\end{equation}
+\end{prop}
+{\em Proof:}
+Denote as $F_i$ the marginal distribution of $Z_i$. To obtain $\chi_{ij}$ we need $Pr(Z_1>z_1,Z_2>z_2)$, where $z_i =
+F_i^{-1}(u) = E(V_i)/(1 - u), \;i=1,\dots, d$.
+Using the fact that
+$V_i>0, \forall i$ almost surely, we have that the former is equal to
+\[	\text{Pr}\left[R>\frac{z_i}{V_i}\vee\frac{z_2}{V_j}\right] =
+E\left(1\wedge\left(\frac{z_i}{V_i}\vee\frac{z_j}{V_j}\right)
+^{-1} \right)
+= E\left( \frac{V_i}{z_i}\wedge\frac{V_j}{z_j}
+\right) = (1 - u)E\left(
+\frac{V_i}{EV_i}\wedge\frac{V_j}{EV_j}
+\right) ,
+\]
+where the second identity is justified by the fact that $V_i$ is
+bounded and $z_i\rightarrow\infty$. The proof is completed by noting
+that $Pr(F_i(Z_i)>u ) = 1 - u.\hfill\Box$
 
-One metric by which we might identify and declare atmospheric rivers is the integrated water vapor
-  transport, or \emph{IVT}.  This value represents the total amount of water
-  vapor being transported in an atmospheric column---that is, a column of the troposphere above a given
-  area of the surface of the Earth. Raw data are measured by dropsondes---measurement devices dropped
-  out of aircraft, taking measurements as they fall to the earth.~\citep{ralph2017}  A data vector
-  includes an estimate of IVT at grid cell at a period in time; where a grid cell specifies the surface
-  of the earth under the atmospheric column.  We have two such datasets, in differing spatial
-  resolutions---the lower resolution data splits the coast of california into 8 grid cells, while the
-  higher resolution does so into 47 grid cells~\citep{guan2015}.
+For the analysis of extreme values it is of interest to calculate the multivariate conditional survival function.
+The following result provides the relevant expression, as a function of the angular measure.
+\begin{prop}
+Assume the same conditions of Proposition \ref{ppchi}. 
+Let $\alpha \subset \{1, \ldots ,p\}$ be a collections of indexes. 
+Then     
+\begin{equation} \label{eqn:condsurv2df}
+    \text{Pr}\left[\bigcap_{l \in \alpha} Z_l > z_l \right|\left. \bigcap_{l\not\in\alpha} Z_l > z_l\right] =
+      \frac{E\left(\bigwedge_{k = 1}^d 1\wedge \frac{V_k}{z_k}\right)}{E\left(\bigwedge_{k \not\in\alpha}1\wedge\frac{V_k}{z_k}\right)}.
+  \end{equation}
+\end{prop}  
+The proof uses a similar approach to the proof of Proposition \ref{ppchi}.
 
-As IVT tracks water vapor in the atmosphere, extreme values in IVT can lead to
-  extreme precipitation events.  Atmospheric rivers are necessary events, in that they carry the
-  bulk of precipitation that California recieves, but extreme precipitation events can be extremely
-  destructive as well.  Thus, we are interested in the distribution of extreme precipitation events,
-  including their depenedence structure.  This tells us about the probability of jointly extreme
-  events---where IVT is extreme in two or more grid cells, as well as identifying particular
-  configurations of IVT that may be anomalous.  In California, groundwater is a managed resource.
-  Extreme precipitation events significantly affect groundwater, so anomalous precipitation events
-  can produce anomalous configurations of groundwater, leading to destructive flooding.
+%\subsection{Integrated Vapor Transport}
+%Atmospheric rivers are temporary events, where large elongated regions of high concentrations of
+%  water vapor are developed in the atmosphere and carry huge amounts of water potentially thousands
+%  of miles.  The amount of water in transit during these events can dwarf that of terrestrial rivers.
+%  For the targeted region, the atmospheric river can represent a significant portion of the
+%  precipitation the region will experience.  Thus, such events are of great interest to meteorologists,
+%  hydrologists, actuaries, land-use planners, and even farmers.  The ability to estimate the likely magnitude
+%  and distribution of such events is thus of great importance~\citep{ralph2013,ralph2018}.
 
-We are specifically interested in extremal dependence---the relationship between the upper tails of
-  dimensions of the distribution.  In this case, that means the relationship between extreme values
-  in different grid cells.  We are going to be looking at point-in-time behavior rather than
-  considering the time series nature of the data---that relationship may come later.
+%One metric by which we might identify and declare atmospheric rivers is the integrated water vapor
+%  transport, or \emph{IVT}.  This value represents the total amount of water
+%  vapor being transported in an atmospheric column---that is, a column of the troposphere above a given
+%  area of the surface of the Earth. Raw data are measured by dropsondes---measurement devices dropped
+%  out of aircraft, taking measurements as they fall to the earth.~\citep{ralph2017}  A data vector
+%  includes an estimate of IVT at grid cell at a period in time; where a grid cell specifies the surface
+%  of the earth under the atmospheric column.  We have two such datasets, in differing spatial
+%  resolutions---the lower resolution data splits the coast of california into 8 grid cells, while the
+%  higher resolution does so into 47 grid cells~\citep{guan2015}.
 
-As we are interested in the extremal dependence, it makes sense that we would choose to represent
-  this data using tools from extreme value theory.  Extreme value theory, or \emph{EVT}, seeks to
-  model and assess probability of observing extreme events.  Such a topic is applicable generally,
-  but it finds particularly strong use among such fields as finance \citep{allen2013},
-  climatology \citep{trepanier2018}, and insurance \citep{beirlant1994}.  In these fields, extreme
-  events may represent significant loss to the body commissioning the study.  For instance, an
-  insurance company might commission a study on extreme weather events, as an extreme weather event
-  localized to a particular region could cause a spike in claims from that region.  Extreme value
-  theory offers us a tool set for making inference about the tails of a distribution, without having
-  observed said tails.  For
-  instance, with an extreme weather event like flooding, we can make predictions about return
-  levels---the average time until an observation of a particular magnitude occurs---without having seen
-  an observation of that magnitude, or having observed that long.  In the context of our motivating
-  example, extreme values in the IVT may represent the formation of an atmospheric river, which has
-  dramatic effects on precipitation and ground water.  The formation of an IVT may lead to flooding
-  and other negative effects, but is also contributes necessary water for irrigation and agriculture.
-  Development of statistical tools regarding these extreme events becomes necessary for making informed
-  decisions.
+%As IVT tracks water vapor in the atmosphere, extreme values in IVT can lead to
+%  extreme precipitation events.  Atmospheric rivers are necessary events, in that they carry the
+%  bulk of precipitation that California recieves, but extreme precipitation events can be extremely
+%  destructive as well.  Thus, we are interested in the distribution of extreme precipitation events,
+%  including their depenedence structure.  This tells us about the probability of jointly extreme
+%  events---where IVT is extreme in two or more grid cells, as well as identifying particular
+%  configurations of IVT that may be anomalous.  In California, groundwater is a managed resource.
+%  Extreme precipitation events significantly affect groundwater, so anomalous precipitation events
+%  can produce anomalous configurations of groundwater, leading to destructive flooding.
+
+%We are specifically interested in extremal dependence---the relationship between the upper tails of
+%  dimensions of the distribution.  In this case, that means the relationship between extreme values
+%  in different grid cells.  We are going to be looking at point-in-time behavior rather than
+%  considering the time series nature of the data---that relationship may come later.
+
+%As we are interested in the extremal dependence, it makes sense that we would choose to represent
+%  this data using tools from extreme value theory.  Extreme value theory, or \emph{EVT}, seeks to
+%  model and assess probability of observing extreme events.  Such a topic is applicable generally,
+%  but it finds particularly strong use among such fields as finance \citep{allen2013},
+%  climatology \citep{trepanier2018}, and insurance \citep{beirlant1994}.  In these fields, extreme
+%  events may represent significant loss to the body commissioning the study.  For instance, an
+%  insurance company might commission a study on extreme weather events, as an extreme weather event
+%  localized to a particular region could cause a spike in claims from that region.  Extreme value
+%  theory offers us a tool set for making inference about the tails of a distribution, without having
+%  observed said tails.  For
+%  instance, with an extreme weather event like flooding, we can make predictions about return
+%  levels---the average time until an observation of a particular magnitude occurs---without having seen
+%  an observation of that magnitude, or having observed that long.  In the context of our motivating
+%  example, extreme values in the IVT may represent the formation of an atmospheric river, which has
+%  dramatic effects on precipitation and ground water.  The formation of an IVT may lead to flooding
+%  and other negative effects, but is also contributes necessary water for irrigation and agriculture.
+%  Development of statistical tools regarding these extreme events becomes necessary for making informed
+%  decisions.
 

### 2021-09-16 methodology_pnorm.tex
bruno's edits 2

@@
 
-\begin{figure}[ht]
-    \centering
-    \includegraphics{images/p_sphere}
-    \caption{Unit hyperspheres established on $\mathcal{L}_p$, for $p = 1,\ldots,10$, for two dimensions.\label{fig:psphere}}
-\end{figure}
+%\begin{figure}[ht]
+%    \centering
+%    \includegraphics{images/p_sphere}
+%    \caption{Unit hyperspheres established on $\mathcal{L}_p$, for $p = 1,\ldots,10$, for two dimensions.\label{fig:psphere}}
+%\end{figure}
 
-A hypersphere is a geometric object such that the distance from any point to the center takes a fixed,
-  constant value.  The unit hypersphere is a hypersphere where that distance is 1. We can define the
-  hypersphere under an arbitrary distance measurement, but for our purpose we will use the class of
-  hyperspheres defined under the $\mathcal{L}_p$ norm. Let the $\mathcal{L}_p$-norm be defined as
+Consider a vector $\bm{s} \in {\mathbb R}^d$, and let the $\mathcal{L}_p$-norm be defined as
   \begin{equation*}
-    \lVert \bm{s} \rVert_p = \left({\textstyle\sum}_{\ell = 1}^d \lvert s_{\ell}\rvert^p\right)^{\frac{1}{p}}.
+    \lVert \bm{s} \rVert_p = \left({\textstyle\sum}_{\ell = 1}^d \lvert s_{\ell}\rvert^p\right)^{\frac{1}{p}},
   \end{equation*}
-  From this, we establish the $\mathcal{L}_1$ norm as $p = 1$, or the absolute sum; the
-  $\mathcal{L}_2$ norm, as $p = 2$, the Euclidean distance.  From this we also establish the
-  $\mathcal{L}_{\infty}$ norm, as
+for $p>0$. The absolute and Euclidean norms are obtained for $p=1$ and $p=2$ respectively. 
+The $\mathcal{L}_{\infty}$ norm can be obtained as a limit, 
   \begin{equation*}
@@
   \end{equation*}
-  We are interested in the direction, or angular component, of vectors described in the positive
-  orthant, $\mathcal{R}_{+}^d$.  One means of describing the direction of a vector in $\mathcal{R}_+^d$
-  is to project that vector onto $\mathcal{S}_{p}^{d-1}$, the positive orthant of the unit hypersphere
-  defined on an $\mathcal{L}_p$-norm, denoted as $\mathcal{S}_{p}^{d-1}$.  That is,
+To define an angular measure,  we are interested in the direction of
+vectors in ${\mathbb R}_{+}^d$, thus, we project them onto
   \begin{equation*}
-    \mathcal{S}_{p}^{d-1} = \left\lbrace \bm{s} : \bm{s} \in \mathcal{R}_{+}^{d}, \lVert \bm{s}\rVert_{p} = 1\right\rbrace.
+    \mathcal{S}_{p}^{d-1} = \left\lbrace \bm{s} : \bm{s} \in {\mathbb R}_{+}^{d}, \lVert \bm{s}\rVert_{p} = 1\right\rbrace,
   \end{equation*}
-  Figure~\ref{fig:psphere} shows $\mathcal{S}_{p}^{1}$ for $p$ at integers between 1 and 10.
-  At $p = 1$, we have the simplex.  As $p$ grows larger, the surface moves away from the origin and
-  approaches closer to the unit hypercube--the limiting form of the surface. We project an observation
-  onto $\mathcal{S}_p^{d-1}$ by dividing said observation by its $p$-norm--let
-  $\bm{x}\in \mathcal{R}_{+}^{d}$, then $\bm{y} = \bm{x} / \lVert \bm{x}\rVert_p \in \mathcal{S}_{p}^{d-1}$.
-  We denote the $d-1$ to indicate the loss of one degree of freedom relative to the original vector.
+  the $\mathcal{L}_p$-norm unit hypersphere.
+  Given $\bm{x}\in {\mathbb R}^d_+$ the projection onto $\mathcal{S}_{p}^{d-1}$ is given as
+  $\bm{y} = \bm{x} / \lVert \bm{x}\rVert_p \in \mathcal{S}_{p}^{d-1}$.
   Observations on one hypersphere can be projected without loss of information onto another by
-  dividing by the defining norm of the target hypersphere.
-
-Assuming $\bm{y} \in \mathcal{S}_{p}^{d-1}$, then for finite $p$, $y_d$ can always be represented
-  as a function of the other dimensions.  That is,
+  dividing by the defining norm of the target hypersphere. In fact, letting 
   \begin{equation*}
@@
   \end{equation*}
-  So the transformation
+  the transformation
   \begin{equation}
@@
   \end{equation}
-  does not lose any information.  The reverse of this transformation,
+  is invertible with
   \begin{equation}
@@
     T^{-1}\left(r,y_1,\ldots,y_{d-1}\right) =
-      \left(ry_1,\ldots,ry_{d-1}, r\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}\right)
+      \left(ry_1,\ldots,ry_{d-1}, r\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}\right).
   \end{equation}
-  equivalently recovers the original data.  To recover the density of the transformed random variables,
-  the original density is multiplied by the determinant of the Jacobian--the matrix of derivatives
-  of the inverse transformation.  This takes the form:
+  The Jacobian of the transformation takes the form
   \begin{equation}
@@
   \end{equation}
-  Notice a factor of $r^{d-1}$ independent of $p$. We refer to $\bm{y}$ and $r$ as, respectively,
-  the angular and radial components of $\bm{x}$.  If we assume a distribution for $\bm{x}$, then
-  by transforming to $r, \bm{y}$ and integrating out $r$, we are left with a distribution on solely
-  $\bm{y}$, the projection of $\bm{x}$ onto $\mathcal{S}_{p}^{d-1}$.  If we take the projection to
-  its limit as $p\to\infty$, we arrive at a projection onto $\mathcal{S}_{\infty}^{d-1}$.
-
-We can observe a problem at this point in at least two ways.  First, on $\mathcal{S}_{\infty}^{d-1}$, the
-  inverse transformation from $\mathcal{S}_{\infty}^{d-1}$ to $R_+^{d}$ in Equation~\ref{eqn:pnormtinv}
-  may not recover the original data--that formulation of the projection would only work when $Y_d = 1$.
-  Alternatively, we may look at the formula for the determinant of the Jacobian in
-  Equation~\ref{eqn:pnormjac}.  In that formula we observe the value
-  $(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p)^{\frac{1}{p} - 1}$.  As $p\to\infty$,
-  this quantity will go to $0^{-1}$ if $y_d \neq 1$.
-  So, the mathematical basis of this transformation is thus dependent upon finite $p$.  However, we can
-  project onto $\mathcal{S}_p^{d-1}$ for large $p$, and that projection will be similar to an idealized
-  distribution on $\mathcal{S}_{\infty}^{d-1}$.
+  Equations (\ref{eqn:pnormt}) -- (\ref{eqn:pnormjac}) can be used to define a density on ${\mathbb R}^p_+$, and then obtain a density on $\mathcal{S}_{p}^{d-1}$, for any finite $p>0$.
+Unfortunately in the limit, when $p\rightarrow +\infty$, the transformation is not differentiable.
+Thus, even though the projected distribution on $\mathcal{S}_\infty^{d-1}$ is defined, there is no corresponding density.
 

### 2021-09-16 methodology_projgamma.tex
bruno's edits 2

@@
-\subsection{Projected Gamma Family}
-As stated previously, one approach we might try in order to generate a distribution on
-  $\mathcal{S}_{p}^{d-1}$ would be to take a distribution in $\mathcal{R}_{+}^d$, reparameterize
-  it in terms of radial and angular components, and integrate out the radial component in order to
-  effectively project the distribution onto $\mathcal{S}_{p}^{d-1}$.  A natural candidate distribution
-  to accomplish this is the Gamma.  Let
-    $\bm{ X} \sim \prod_{\ell = 1}^d\text{Ga}\left(X_{\ell}\mid\alpha_{\ell},\beta_{\ell}\right)$.
-  Then transform $\bm{ X}$ via the transformation in Eqns.~\ref{eqn:pnormt} into $R,\bm{ Y}$, where
-  $R = \pnorm{\bm{ X}}{p}$ and $Y\in \mathcal{S}_{p}^{d-1}$.  Completing the transformation, we
-  incorporate the determinant of the Jacobian from Eqn.~\ref{eqn:pnormjac}.  Thus we have
-  \begin{equation}
-    \label{eqn:protopg}
+\subsection{Projected gamma family}
+  A natural  distribution to consider in ${\mathbb R}^p_+$ is given by a product of independent
+  univariate Gamma distributions. Let
+    $\bm{ X} \sim \prod_{\ell = 1}^d\text{Ga}\left(X_{\ell}\mid\alpha_{\ell},\beta_{\ell}\right)$, 
+    $a_\ell$ and $b_\ell$ are the shape and scale parameters, repectively. Using Equations (\ref{eqn:pnormt}) -- (\ref{eqn:pnormjac}) we have the joint density
+  \begin{equation*}
     f(r,\bm{ y}) = \prod_{\ell = 1}^{d}
@@
             {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right],
-  \end{equation}
-  where $y_d = \left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}$. Integrating out $r$
-  yields the resulting \emph{Projected Gamma} density
+  \end{equation*}
+  Integrating out $r$ yields the resulting \emph{Projected Gamma} density
   \begin{equation}
@@
   \end{equation}
-  established for an arbitrary $\mathcal{L}_p$ norm. This forms a valid density for all finite $p$.
-  We can regard the Dirichlet distribution as a special case of this density, where
-  $\bm{ y}\in \mathcal{S}_1^{d-1}$, and $\beta_{\ell} = \beta$ for all $\ell$.  The distribution
-  described in \cite{nunez2019} is related to this distribution in that it is established on
-  $\bm{ y}\in \mathcal{S}_2^{d-1}$, then transformed to spherical coordinates via Eqn.~\ref{eqn:spherical}.
-  The full conditional for $r$ takes the form
-  \begin{equation}
-    \label{eqn:rfullcond}
-    r\mid\bm{ \alpha},\bm{ \beta}, y \sim \text{Ga}\left(r\mid{\textstyle\sum}_{\ell = 1}^d
-                    \alpha_{\ell}, {\textstyle\sum}_{\ell = 1}^d \beta_{\ell}y_{\ell}\right).
+  defined for $\bm{y}\in \mathcal{S}_p^{d-1}$, and for any $p>0$. To avoid idetifiability problems when estimating the shape and scale parameters, we set $\beta_1 = 1$.
+  \cite{nunez2019} obtain the density in Equation (\ref{eqn:projgamma})
+  for $p=2$ as a multivariate distribution for directional data, using spherical coordinates.
+  For $\bm{ y}\in \mathcal{S}_1^{d-1}$, and $\beta_{\ell} = \beta, \forall\ell$, the density in Equation (\ref{eqn:projgamma}) corresponds to that of a Dirichlet distribution.
+
+  The projected gamma family is simple to specify and has very tractable computational properties.
+Thus, we use it as a building block for the angular measure $\Phi$ models.
+To build a flexible family of distributions in $\mathcal{S}_p^{d-1}$ we consider mixtures of projected gamma densities defined as
+  \begin{equation} \label{PGmix}     
+     f(\bm{y}) = \int_\Theta \text{PG}(\bm{y}|\bm{\theta}) dG(\bm{\theta}), 
   \end{equation} 
+  where $\bm{\theta} = (\bm{\alpha}, \bm{\beta})$.
+Following a Bayesian non-parametric approach \citep{Ferguson74,Antoniak1974} we assume that $G$ is drawn from a random measure.
+In particular, assuming a Dirichlet process prior, we have a hierarchical formulation of the mixture model given by
+  \[    \bm{y} \sim \text{PG}(\bm{y}|\bm{\theta}) , \;\;\; \bm{\theta} \sim G, \;\;\; G \sim \text{DP}(\eta, G_0),\]
+  where DP denotes a Dirichlet process, $\eta$ is the precision parameter, and $G_0$ is the centering distribution.
 
-In the estimation of the parameters of this model, we can quickly arrive at a problem, in that
-  without some linear restriction on the rate parameters $\beta_{\ell}$, the model is not identifiable.
-  The restriction most often used in the case of the Generalized Dirichlet, as well as by
-  \cite{nunez2019} is to set $\beta_1 := 1$.
+  Our strategy consists of describing the angular distribution $\Phi$ using a sample based approach with the following steps: (i) Apply the transformation in Equation (\ref{eqn:standardization}) to the original data; (ii) Obtain the subsample of the standardized observations that satisfy $R>1$; (iii) Take a finite $p$ and project the observations onto $\mathcal{S}_p^{d-1}$; (iv) Fit the model in Equation (\ref{PGmix}) to the resulting data and obtain samples from the fitted model; (v) Project the resulting samples to $\mathcal{S}_\infty^{d-1}$.
+  {\bf Instead of the olf Figure 1, it would be nice to add a figure where you illustrate this procedure. Give it a try.}
+%  The full conditional for $r$ takes the form
+%  \begin{equation}
+%    \label{eqn:rfullcond}
+%    r\mid\bm{ \alpha},\bm{ \beta}, y \sim \text{Ga}\left(r\mid{\textstyle\sum}_{\ell = 1}^d
+%                    \alpha_{\ell}, {\textstyle\sum}_{\ell = 1}^d \beta_{\ell}y_{\ell}\right).
+%  \end{equation}
 
-\subsubsection{Inference on Projected Gamma Model}
+\subsubsection{Inference for the projected gamma mixture model}
+{\bf Our focus is on model (\ref{PGmix}).
+Describe how you build a sample based approach to fitting that model.
+How you introduce latent variables, both to break the mixture and to produce the PG.
+You can then discuss variations that essentially consists on different priors.
+So, basically, you need to condense the next three pages in about a page.}
+
 We consider a basic model that assumes all observations are generated by a single projected Gamma

### 2021-09-16 results.tex
bruno's edits 2

@@
-\section{Results}
+\section{Data illustrations}
 
-\subsection{Simulation}
+\subsection{Simulations}
 We have introduced this projected gamma distribution existing on $\mathcal{S}_{\infty}^{p-1}$.  As a
@@
 \subsection{Integrated Vapor Transport}
+One metric by which we might identify and declare atmospheric rivers is the integrated water vapor
+  transport, or \emph{IVT}.  This value represents the total amount of water
+  vapor being transported in an atmospheric column---that is, a column of the troposphere above a given
+  area of the surface of the Earth. Raw data are measured by dropsondes---measurement devices dropped
+  out of aircraft, taking measurements as they fall to the earth.~\citep{ralph2017}  A data vector
+  includes an estimate of IVT at grid cell at a period in time; where a grid cell specifies the surface
+  of the earth under the atmospheric column.  We have two such datasets, in differing spatial
+  resolutions---the lower resolution data splits the coast of california into 8 grid cells, while the
+  higher resolution does so into 47 grid cells~\citep{guan2015}.
+
+As IVT tracks water vapor in the atmosphere, extreme values in IVT can lead to
+  extreme precipitation events.  Atmospheric rivers are necessary events, in that they carry the
+  bulk of precipitation that California recieves, but extreme precipitation events can be extremely
+  destructive as well.  Thus, we are interested in the distribution of extreme precipitation events,
+  including their depenedence structure.  This tells us about the probability of jointly extreme
+  events---where IVT is extreme in two or more grid cells, as well as identifying particular
+  configurations of IVT that may be anomalous.  In California, groundwater is a managed resource.
+  Extreme precipitation events significantly affect groundwater, so anomalous precipitation events
+  can produce anomalous configurations of groundwater, leading to destructive flooding.
+
 We use data from the integrated water vapor transport (\emph{IVT}) model \citep{guan2015}
@@
   better spent elsewhere.
+
+  {\bf We need a one paragraph summary of the stuff in the last three paragraphs. Say what IVT, how it is obtained, why it is important and describe the two datasets you have.}
 

### 2021-09-18 introduction.tex
asasd

@@
 Likelihood-based approaches to inference can be readily implemented in both cases.
-In the multivariate case the GEV theory is well established \citep[see, for example][]{dehaan2006}, but the inferential problem is more complicated, as there is no parametric representation of the GEV.
+In the multivariate case the GEV theory is well developed \citep[see, for example][]{dehaan2006}, but the inferential problem is more complicated, as there is no parametric representation of the GEV.
 This problem is inherited by the PoT approach and compounded by the fact that there is no unique definition of an exceedance of a multivariate threshold, as there is an obvious dependence on the norm that is used to measure the size of a vector.
 
-During the last decade much work has been done in the exploration of the definition and properties of an
+During the last decade or so much work has been done in the exploration of the definition and properties of an
   appropriate generalization of the univariate GP distribution 
@@
 In this setting, $\mu$ is the asymptotic distribution of $\bm{Z}$ in extreme regions, and it is referred to as the
-  \emph{exponent measure}.  $\mu$ features the homogeneity property $\mu(tA) = \frac{1}{t}\mu(A)$. 
-Assume that $R = \max_{i} Z_i>1$ with probability one, and that $V_i = Z_i/\max_i Z_i$ is such that $E(V_i)>0$, then
-we have that the second condition in Theorem 2.1 of
-\cite{ferreira2014}, applied to the finite-dimensional case, is satisfied.
-Thus, we have the factorization $\bm{Z} =  R\bm{V}$, where $R$ and $\bm{V}$  are independent. Thus,
+  \emph{exponent measure}. Under this condition $\mu$ features the homogeneity property $\mu(tA) = \frac{1}{t}\mu(A)$. 
+Define that $R = \max_{i} Z_i$, and assume that $R>1$ with probability one. Also, define $V_i = Z_i/\max_i Z_i$, and assume that $E(V_i)>0$.
+These two assumption together with the homogeneity of $\mu$ correspond to the second condition in Theorem 2.1 of
+\cite{ferreira2014}, applied to the finite-dimensional case.
+Thus, we have the factorization $\bm{Z} =  R\bm{V}$, where $R$ and $\bm{V}$  are independent. Moreover,
 for some measure $\Phi$, $\mu(\cdot)$ is factorized as
-\[
+\begin{equation}\label{mufact}
     \mu\left( \left\lbrace\bm{Z} : V \in A , R > r\right\rbrace \right) = r^{-1}\Phi(A).
-\]
+\end{equation}
 $R$ describes the radial component, and $\bm{V} \in \mathcal{S}_{\infty}^{d-1}$ the 
@@
 orthant of the unit sphere in infinite norm.  $\Phi(\cdot)$ is referred to as the \emph{angular measure}.   
-Moreover, $R$ is distributed as a standard Pareto random variable. Conditioning on $R$, we have that
+From Equation (\ref{mufact}) we see that $R$ is distributed as a standard Pareto random variable. Conditioning on $R$, we have that
 \begin{equation} \label{MGP}
@@
 Pareto process indexed on a finite dimensional set. An alternative approach to 
-define such distribution that provides a link to the multivariate GEV is discussed 
+define such distribution, that provides a link to the multivariate GEV, is discussed 
 in Remark 1 of \cite{RoSeWa2018a}. More specifically, 

### 2021-09-18 methodology_projgamma.tex
asasd

@@
 
-  Our strategy consists of describing the angular distribution $\Phi$ using a sample based approach with the following steps: (i) Apply the transformation in Equation (\ref{eqn:standardization}) to the original data; (ii) Obtain the subsample of the standardized observations that satisfy $R>1$; (iii) Take a finite $p$ and project the observations onto $\mathcal{S}_p^{d-1}$; (iv) Fit the model in Equation (\ref{PGmix}) to the resulting data and obtain samples from the fitted model; (v) Project the resulting samples to $\mathcal{S}_\infty^{d-1}$.
+  Our strategy consists of describing the angular distribution $\Phi$ using a sample based approach with the following steps: (i) Apply the transformation in Equation (\ref{eqn:standardization}) to the original data; (ii) Obtain the subsample of the standardized observations that satisfy $R>1$; (iii) Take a finite $p$ and project the observations onto $\mathcal{S}_p^{d-1}$; (iv) Fit the model in Equation (\ref{PGmix}) to the resulting data and obtain samples from the fitted model; (v) Project the resulting samples onto $\mathcal{S}_\infty^{d-1}$.
+  For step (iv) we use a Bayesian approach that is implemented using a purposely developed Markov chain Monte Carlo that is described in the next section. 
   {\bf Instead of the olf Figure 1, it would be nice to add a figure where you illustrate this procedure. Give it a try.}

### 2021-10-12 conclusion.tex
update

@@
+\section{Conclusion\label{sec:conclusion}}
 In this paper, we have built upon the definition of the multivariate Pareto described in \cite{ferreira2014}
@@
   the distribution of its angular component; it being supported on the positive orthant of the unit
-  hypersphere under the $\mathcal{L}_{\infty}$ norm, $\mathcal{S}_{\infty}^{d-1}$.  We then demonstrated an
+  hypersphere under the $\mathcal{L}_{\infty}$ norm, ${\mathbb S}_{\infty}^{d-1}$.  We then demonstrated an
   inherent difficulty in establishing a distribution supported in this space.  To build that representation,
   we presented a distribution based on a product of independent Gammas, projected onto a supporting manifold
-  $\mathcal{S}_{p}^{d-1}$, which will converge to $\mathcal{S}_{\infty}^{d-1}$ as $p\to\infty$.  We then
-  established a means of model comparison on $\mathcal{S}_{\infty}^{d-1}$ through the use of a negative
+  ${\mathbb S}_{p}^{d-1}$, which will converge to ${\mathbb S}_{\infty}^{d-1}$ as $p\to\infty$.  We then
+  established a means of model comparison on ${\mathbb S}_{\infty}^{d-1}$ through the use of a negative
   definite kernel defined in that space.

### 2021-10-12 evaluation.tex
update

--- 2021-09-15 evaluation.tex
@@
-\section{Model Comparison on the Hypersphere}
-\label{sec:evaluation}
-It is not immediately obvious what criteria to use to judge these models and decide which best
-  represents the data's generating distribution.  As the task resides on $\mathcal{S}_{\infty}^{d-1}$,
-  our options are somewhat limited in terms of what model criteria we can use.  Because of the aforementioned
-  difficulty in evaluating a density directly in this space, we choose to use model evaluation
-  criteria that can operate on samples from the posterior predictive distribution.  For this reason,
-  we have opted to use the \emph{posterior predictive loss} criterion of \cite{gelfand1998} and the
-  \emph{energy score} criterion of \cite{gneiting2007}.  Both of these metrics require calculating
-  some distance in the target space, and this section will be devoted to that end.  As both these
-  metrics operate on the posterior predictive distribution for each observation, we will also be
-  including a metric operating on the overall posterior predictive distribution, in the form of a
-  modified Kullbeck-Liebler divergence.
+\section{Scores for distributions in ${\mathbb S}_\infty^{d-1}$\label{sec:evaluation}}
 
-\subsection{Posterior Predictive Loss Criterion}
-The posterior predictive loss criterion, \emph{PPL} is introduced in \cite{gelfand1998}.  When we
-  assume a squared error loss function, then for the $i$th observation, the posterior predictive
-  loss criterion is computed as
-  \begin{equation}
-    \label{eq:ppl}
-    D_{k}^{(i)} = \text{Var}\left(X_i\right) + \frac{k}{k + 1}\left(\text{E}[X_i] - x_i\right)^2,
-  \end{equation}
-  where $X_i$ is a random variable from the posterior predictive distribution for $x_i$.  The
-  scalar $k$ is a weighting factor by which we arbitrarily scale the importance of goodness of fit
-  relative to precision.  In our analysis, we take the limit as $k\to\infty$, weighting both
-  parts equally.  Interpreting this criterion, a smaller $\text{Var}(\bm{X}_{\ell})$ indicates a higher
-  precision, and a smaller $(\text{E}[\bm{X}_{\ell}] - \bm{x}_{\ell})^2$  indicates a better fit.  Thus,
-  smaller is better.  Note that this is defined for a univariate distribution.  As we are dealing
-  with a multivariate distribution, we will be taking the posterior predictive loss
-  summed over all dimensions.  That is,
-  \begin{equation}
-    \label{eq:ppl2}
-    D_k^{(i)} = \frac{1}{d}\sum_{\ell = 1}^{d}\left[\text{Var}(X_{i,\ell}) +
-                  \frac{k}{k+1}\left(\text{E} [X_{i,\ell}] - x_{i,\ell}\right)^2\right]
-  \end{equation}
-  Then we report the average $D_k$, taken over all $i$.
-
-\subsection{Energy Score}
-The energy score of Gneiting, et al\cite{gneiting2007}, is a generalization of the continuous ranked
-  probability score, or \emph{crps}, defined for a multi-dimensional random variable.
+As explained in~\ref{subsubsec:norm}, we can not establish a \emph{density} on ${\mathbb S}_{\infty}^{d-1}$, restricting our ability to construct model selection criteria to sample-based approaches.
+In \cite{gneiting2007}, Theorem~4 states that a score,
     \begin{equation}
     \label{eq:es}
-    \text{ES}\left(P, \bm{x}_i\right) =  \text{E}_p g\left(\bm{X}_i, \bm{x}_i\right) -
-                                    \frac{1}{2}\text{E}_p g\left(\bm{X}_i,\bm{X}_i^{\prime}\right)
+    \text{S}\left(P, \bm{x}_i\right) =  \text{E}_p g\left(\bm{X}_i, \bm{x}_i\right) -
+                                    \frac{1}{2}\text{E}_p g\left(\bm{X}_i,\bm{X}_i^{\prime}\right),
     \end{equation}
-  where $g(\cdot)$ identifies a negative definite kernel, and $\bm{X}_i^{\prime}$ represents another
-  replicate from the posterior predictive distribution of $\bm{x}_i$. A function $g$ is a negative definite kernel if it is symmetric in its arguments, $g(x_1,x_2) = g(x_2,x_1)$, and for which
-  $\sum_{i =1}^n\sum_{j=1}^n\alpha_i\alpha_jg(\bm{x}_1,\bm{x}_2) \leq 0$ for all positive integers n, with the
-  restriction that $\sum_{i=1}^n\alpha_i = 0$.  Euclidean distance is one example of a negative definite kernel,
-  and the one most often used.  However, on $\mathcal{S}_p^{d-1}$ for $p > 1$, Euclidean distance
-  can under-report the actual distance required for travel between two points. That distortion is
-  maximized on $\mathcal{S}_{\infty}^{d-1}$.
+where $g(\cdot)$ is a negative definite kernel, is a \emph{proper} scoring rule.
+A real valued function $g$ is a negative definite kernel if it is symmetric in its arguments, and $\sum_{i=1}^n\sum_{j=1}^na_ia_jg(x_i,x_j)\leq 0$ for all positive integers $n$, $a_1,\ldots,a_n\in{\mathbb R}$, and $\sum_{i = 1}^na_i = 0$.
+\cite{gneiting2007} introduces this quantity as the \emph{energy score}.
 
-\subsection{Distance on the positive orthant of the Hypercube}
-  \label{subsec:distance}
-The positive orthant of the unit hypercube, defined in Euclidean geometry, is that structure for
-  which, in a given point on the hypercube, all dimensions of that point are between 0 and 1, and
-  at least one dimension must be 1.  Developing terminology, we can consider observations for which
-  the $j$th dimension is equal to 1, to be on the $j$th \emph{face}.  The intersection of the $i$th
-  and $j$th face is a $d-2$ dimensional cube, and observations in this space have dimensions $i,j$
-  equal to 1.
+In Euclidean space, the conditions for a negative definite kernel are satisfied by Euclidean distance.
+On ${\mathbb S}_{\infty}^{d-1}$, however, Euclidean distance will tend to under-estimate actual distance required for travel between two points.
+Developing terminology, consider ${\mathbb C}_{\ell}^{d-2} = \lbrace x : x_{\ell} = 1\rbrace$ the \emph{$\ell$th face}.
+Then for points on the same face, Euclidean distance will report the length of the shortest path.
+For points on different faces, Euclidean distance will under-report that length.
 
-A \emph{distance} in this space is a \emph{geodesic} on this space. From geometry, we know that
-  the geodesic, or shortest path between 2 points along the surface of a $d$ dimensional figure
-  corresponds to at least one \emph{unfolding}, or \emph{rotation} of the $d$-dimensional
-  figure into a $d-1$ dimensional space.  The appropriate term for the structure generated by this
-  unfolding is a \emph{net}.  For the appropriate net, a line segment connecting the two points and
-  staying within the boundaries of the net corresponds to the shortest path between those points\citep{pappas1989}, and is thus a geodesic.  The length of that line segment is properly
-  defines the distance required for travel between those points.
+The space, ${\mathbb S}_{\infty}^{d-1}$ exists as the limit of a sequence of manifolds, ${\mathbb S}_p^{d-1}$ for $p\to\infty$.
+For a given manifold, the length of the shortest path between two points is a \emph{geodesic}; as the length of the geodesic is symmetric, and by definition satisfies the triangle inequality, we call it a \emph{distance} in that manifold.
+Ideally, we would want to establish such a distance in ${\mathbb S}_{\infty}^{d-1}$, but as the surface is not differentiable, routines for establishing geodesics are not available.
+However, as ${\mathbb S}_{\infty}^{d-1}$ is a portion of of a $d$-cube, we can borrow a result from geometry \citep{pappas1989} in that the length of the shortest path between 2 points on a geometric figure corresponds to the length of a straight line drawn between the points on an appropriate unfolding, rotation, or \emph{net} of the figure from $d$-dimensional to $d-1$-dimensional space.
+The optimal net will have the shortest straight line between the points, so long as that line does not exit the net.
+As ${\mathbb S}_{\infty}^{d-1}$ has $d$ faces---each face pairwise adjacent, there are $d!$ possible nets.
+However, as we are only interested in nets that begin and end on the source and destination faces respectively, and are not concerned with divergence after the destination face.
+That reduces the number of nets for consideration to $\sum_{k = 0}^{d-2}\binom{d-2}{k}$.
+This is still computationally burdensome for a large number of dimensions.
 
-Consider a 3-dimensional cube.  Consider 2 points on this 3-dimensional cube,
-  $\bm{a}_1 = (x_1,y_1,z_1)$, and $\bm{a}_2 = (x_2,y_2,z_2)$. Let's say that the two points are
-  on the same face.  Then the distance between those two points, the distance one has to travel
-  along the space to move from one point to the other, is calculated by Euclidean norm.  Now,
-  consider two points on separate faces.  All faces are pairwise adjacent, as we have stated, so
-  in order to move to the other point, we must \emph{at least} move to the intersection between the
-  faces, then to the other point.  Let $\bm{a}_1$ lie on the $x$ face, and $\bm{a}_2$ lie on the
-  $y$ face.  That is, $\bm{a}_1 = (1, y_1, z_1)$, $\bm{a}_2 = (x_2, 1, z_2)$.  Then traveling
-  between these points we must at least pass through the intersection of faces $x,y$.  One possible
-  net representation of this is unfolding the $y$ face alongside the $x$ face.  We accomplish this
-  by applying a rotation and translation to $a_2$, corresponding to the following:
-  \begin{equation}
-    \label{eq:1drotation}
-    a_2^{\prime} = \begin{bmatrix}
-    1 \\
-    2 \\
-    0
-    \end{bmatrix}
-    +
-    \begin{bmatrix}
-    0  & 0 & 0 \\
-    -1 & 0 & 0 \\
-    0  & 0 & 1
-    \end{bmatrix}
-    \begin{bmatrix}
-    x_2 \\
-    1 \\
-    z_2 \\
-    \end{bmatrix} = \begin{bmatrix}
-    1 \\
-    2 - x_2 \\
-    z_2
-    \end{bmatrix}
-  \end{equation}
-  Then, if this is the appropriate net, the distance from $\bm{a}$ to $\bm{b}$ becomes
-  $\lVert \bm{a}_1 - \bm{a}_2^{\prime}\rVert_2$. However, there
-  is another possible net we must consider, travelling first through the $z$ face then to the
-  $y$ face.  The rotation for that becomes:
-  \begin{equation}
-    \label{eq:2drotation}
-    a_2^{\prime} = \begin{bmatrix}
-    1 \\
-    2 \\
-    2
-    \end{bmatrix}
-    +
-    \begin{bmatrix}
-    0 & 0 & 0  \\
-    0 & 0 & -1 \\
-    -1 & 0 & 0
-    \end{bmatrix}
-    \begin{bmatrix}
-    x_2 \\
-    1 \\
-    z_2
-    \end{bmatrix} = \begin{bmatrix}
-    1 \\
-    2 - z_2 \\
-    2 - x_2
-    \end{bmatrix}
-  \end{equation}
-Every successive rotation is relative to the last face.  So, as the number of dimensions grows, the
-  number of possible rotations grows as well.  As we have $d$ faces, if 2 observations are on
-  different faces, then there are $\sum_{j = 1}^{d-2}\binom{d-2}{j} + 1$ possible rotations to
-  consider\footnote{There are truly $d!$ possible nets, but when we consider starting and ending
-  faces fixed, and that portions of the net that diverge after the ending face are irrelevant,
-  we arrive at that number of rotations that we need consider}.  While this is not
-  insurmountable, it is computationally costly. However, for a valid energy score, a true distance
-  is not necessary.  What is needed is a negative definite kernel.
-
+Instead, we develop as our kernel an upper bound on distance between points on separate faces as the length of the path from the source point, to an optimal point on the intersection between faces, to the destination point.
 \begin{prop}
-For points $a,b \in \mathcal{S}_{\infty}^{d-1}$ a valid negative definite kernel can be formed as
+For points $a,b \in {\mathbb S}_{\infty}^{d-1}$ a valid negative definite kernel can be formed as
   \begin{equation}
@@
 
+{\em Proof:}
 In the first case, where $\bm{a}$ and $\bm{b}$ reside on the same face, then that is trivially a
@@
       \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j g(\bm{x}_1,\bm{x}_2) &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j \bigg[\pnorm{\bm{c} - \bm{x}_1}{2} + \pnorm{\bm{x}_2 - \bm{c}}{2}\bigg]\\
-      &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{c} - \bm{x}_1}{2} + \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{x}_2 - \bm{c}}{2}
+      &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{c} - \bm{x}_1}{2} + \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{x}_2 - \bm{c}}{2},
     \end{aligned}
   \end{equation*}
-  which is less than $0$ in both terms by definition.
+  which is less than $0$ in both terms.$\hfill\Box$
 
-To compute this energy score as described would require optimization of the interim point $\bm{c}$.
-  If we again consider the one-dimensional rotation in Equation~\ref{eq:1drotation}---we
-  can view that as a numerically easier analogue accomplishing the same goal: the distance from
-  the starting point, to some optimal point along the intersection between the starting and
-  ending faces, to the ending point.  The strait line will pass through the optimal point, so
-  the optimization of the interim point is accomplished \emph{en passant}.
+Evaluating $g$ as described requires optimization of the interim point $\bm{c}$; a $d-2$-dimensional optimization problem.
+However, let us consider a computationally easier analogue: the aforementioned \emph{net}, where we move directly from the source face to the target face in one jump.
+Let $\bm{a} \in {\mathbb C}_{\ell}^{d-2}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-2}$.
+The rotation required to move the $\jmath$th face along the $\ell$th axis is
+\begin{equation}
+    \bm{b}^{\prime} = 
+    \begin{cases}
+        b_{i} &\text{for }i\neq \jmath,\ell\\
+        1 &\text{for }i = \ell\\
+        2 - b_{\ell} &\text{for }i = \jmath
+    \end{cases}
+\end{equation}
+where $\cdot^{\prime}$ indicates a rotated point.
+Then $g(\bm{a},\bm{b}) = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}$.
+By fitting the straight line between $\bm{a}$ and $\bm{b}^{\prime}$, the optimization of $\bm{c}$ is accomplished \emph{en passant}.
+
+% {\bf After a sentence or two motivating the need for scoring distributions in ${\mathbb S}_\infty^{d-1}$ that can be calculated with a sample-based approach, mention the paper of \cite{gneiting2007} on proper scoring rules, and introduce the energy score.
+Cite the result in that paper that gives conditions for the energy score to be a proper scoring rule.
+Then discuss the problem of finding a metric in ${\mathbb S}_\infty^{d-1}$ that corresponds to a negative definite kernel.
+That is, the material in Section 4.3.
+Introduce and discuss the idea, the write the proposition, with its proof.
+I don't know if the example in the 3-dimensional cube is needed or not.
+Finally, say how the score will be calculated in our case.
+% }
+% It is not immediately obvious what criteria to use to judge these models and decide which best
+%   represents the data's generating distribution.  As the task resides on ${\mathbb S}_{\infty}^{d-1}$,
+%   our options are somewhat limited in terms of what model criteria we can use.  Because of the aforementioned
+%   difficulty in evaluating a density directly in this space, we choose to use model evaluation
+%   criteria that can operate on samples from the posterior predictive distribution.  For this reason,
+%   we have opted to use the \emph{posterior predictive loss} criterion of \cite{gelfand1998} and the
+%   \emph{energy score} criterion of \cite{gneiting2007}.  Both of these metrics require calculating
+%   some distance in the target space, and this section will be devoted to that end.  As both these
+%   metrics operate on the posterior predictive distribution for each observation, we will also be
+%   including a metric operating on the overall posterior predictive distribution, in the form of a
+%   modified Kullbeck-Liebler divergence.
+
+% \subsection{Posterior Predictive Loss Criterion}
+% The posterior predictive loss criterion, \emph{PPL} is introduced in \cite{gelfand1998}.  When we
+%   assume a squared error loss function, then for the $i$th observation, the posterior predictive
+%   loss criterion is computed as
+%   \begin{equation}
+%     \label{eq:ppl}
+%     D_{k}^{(i)} = \text{Var}\left(X_i\right) + \frac{k}{k + 1}\left(\text{E}[X_i] - x_i\right)^2,
+%   \end{equation}
+%   where $X_i$ is a random variable from the posterior predictive distribution for $x_i$.  The
+%   scalar $k$ is a weighting factor by which we arbitrarily scale the importance of goodness of fit
+%   relative to precision.  In our analysis, we take the limit as $k\to\infty$, weighting both
+%   parts equally.  Interpreting this criterion, a smaller $\text{Var}(\bm{X}_{\ell})$ indicates a higher
+%   precision, and a smaller $(\text{E}[\bm{X}_{\ell}] - \bm{x}_{\ell})^2$  indicates a better fit.  Thus,
+%   smaller is better.  Note that this is defined for a univariate distribution.  As we are dealing
+%   with a multivariate distribution, we will be taking the posterior predictive loss
+%   summed over all dimensions.  That is,
+%   \begin{equation}
+%     \label{eq:ppl2}
+%     D_k^{(i)} = \frac{1}{d}\sum_{\ell = 1}^{d}\left[\text{Var}(X_{i,\ell}) +
+%                   \frac{k}{k+1}\left(\text{E} [X_{i,\ell}] - x_{i,\ell}\right)^2\right]
+%   \end{equation}
+%   Then we report the average $D_k$, taken over all $i$.
+
+% \subsection{Energy Score}
+% The energy score of Gneiting, et al\cite{gneiting2007}, is a generalization of the continuous ranked
+%   probability score, or \emph{crps}, defined for a multi-dimensional random variable.
+%   \begin{equation}
+%     \text{ES}\left(P, \bm{x}_i\right) =  \text{E}_p g\left(\bm{X}_i, \bm{x}_i\right) -
+%                                     \frac{1}{2}\text{E}_p g\left(\bm{X}_i,\bm{X}_i^{\prime}\right)
+%   \end{equation}
+%   where $g(\cdot)$ identifies a negative definite kernel, and $\bm{X}_i^{\prime}$ represents another
+%   replicate from the posterior predictive distribution of $\bm{x}_i$. A function $g$ is a negative definite kernel if it is symmetric in its arguments, $g(x_1,x_2) = g(x_2,x_1)$, and for which
+%   $\sum_{i =1}^n\sum_{j=1}^n\alpha_i\alpha_jg(\bm{x}_1,\bm{x}_2) \leq 0$ for all positive integers n, with the
+%   restriction that $\sum_{i=1}^n\alpha_i = 0$.  Euclidean distance is one example of a negative definite kernel,
+%   and the one most often used.  However, on ${\mathbb S}_p^{d-1}$ for $p > 1$, Euclidean distance
+%   can under-report the actual distance required for travel between two points. That distortion is
+%   maximized on ${\mathbb S}_{\infty}^{d-1}$.
+
+% \subsection{Distance on the positive orthant of the Hypercube}
+%   \label{subsec:distance}
+% The positive orthant of the unit hypercube, defined in Euclidean geometry, is that structure for
+%   which, in a given point on the hypercube, all dimensions of that point are between 0 and 1, and
+%   at least one dimension must be 1.  Developing terminology, we can consider observations for which
+%   the $j$th dimension is equal to 1, to be on the $j$th \emph{face}.  The intersection of the $i$th
+%   and $j$th face is a $d-2$ dimensional cube, and observations in this space have dimensions $i,j$
+%   equal to 1.
+
+% A \emph{distance} in this space is a \emph{geodesic} on this space. From geometry, we know that
+%   the geodesic, or shortest path between 2 points along the surface of a $d$ dimensional figure
+%   corresponds to at least one \emph{unfolding}, or \emph{rotation} of the $d$-dimensional
+%   figure into a $d-1$ dimensional space.  The appropriate term for the structure generated by this
+%   unfolding is a \emph{net}.  For the appropriate net, a line segment connecting the two points and
+%   staying within the boundaries of the net corresponds to the shortest path between those points\citep{pappas1989}, and is thus a geodesic.  The length of that line segment is properly
+%   defines the distance required for travel between those points.
+
+% Consider a 3-dimensional cube.  Consider 2 points on this 3-dimensional cube,
+%   $\bm{a}_1 = (x_1,y_1,z_1)$, and $\bm{a}_2 = (x_2,y_2,z_2)$. Let's say that the two points are
+%   on the same face.  Then the distance between those two points, the distance one has to travel
+%   along the space to move from one point to the other, is calculated by Euclidean norm.  Now,
+%   consider two points on separate faces.  All faces are pairwise adjacent, as we have stated, so
+%   in order to move to the other point, we must \emph{at least} move to the intersection between the
+%   faces, then to the other point.  Let $\bm{a}_1$ lie on the $x$ face, and $\bm{a}_2$ lie on the
+%   $y$ face.  That is, $\bm{a}_1 = (1, y_1, z_1)$, $\bm{a}_2 = (x_2, 1, z_2)$.  Then traveling
+%   between these points we must at least pass through the intersection of faces $x,y$.  One possible
+%   net representation of this is unfolding the $y$ face alongside the $x$ face.  We accomplish this
+%   by applying a rotation and translation to $a_2$, corresponding to the following:
+%   \begin{equation}
+%     \label{eq:1drotation}
+%     a_2^{\prime} = \begin{bmatrix}
+%     1 \\
+%     2 \\
+%     0
+%     \end{bmatrix}
+%     +
+%     \begin{bmatrix}
+%     0  & 0 & 0 \\
+%     -1 & 0 & 0 \\
+%     0  & 0 & 1
+%     \end{bmatrix}
+%     \begin{bmatrix}
+%     x_2 \\
+%     1 \\
+%     z_2 \\
+%     \end{bmatrix} = \begin{bmatrix}
+%     1 \\
+%     2 - x_2 \\
+%     z_2
+%     \end{bmatrix}
+%   \end{equation}
+%   Then, if this is the appropriate net, the distance from $\bm{a}$ to $\bm{b}$ becomes
+%   $\lVert \bm{a}_1 - \bm{a}_2^{\prime}\rVert_2$. However, there
+%   is another possible net we must consider, travelling first through the $z$ face then to the
+%   $y$ face.  The rotation for that becomes:
+%   \begin{equation}
+%     \label{eq:2drotation}
+%     a_2^{\prime} = \begin{bmatrix}
+%     1 \\
+%     2 \\
+%     2
+%     \end{bmatrix}
+%     +
+%     \begin{bmatrix}
+%     0 & 0 & 0  \\
+%     0 & 0 & -1 \\
+%     -1 & 0 & 0
+%     \end{bmatrix}
+%     \begin{bmatrix}
+%     x_2 \\
+%     1 \\
+%     z_2
+%     \end{bmatrix} = \begin{bmatrix}
+%     1 \\
+%     2 - z_2 \\
+%     2 - x_2
+%     \end{bmatrix}
+%   \end{equation}
+% Every successive rotation is relative to the last face.  So, as the number of dimensions grows, the
+%   number of possible rotations grows as well.  As we have $d$ faces, if 2 observations are on
+%   different faces, then there are $\sum_{j = 1}^{d-2}\binom{d-2}{j} + 1$ possible rotations to
+%   consider\footnote{There are truly $d!$ possible nets, but when we consider starting and ending
+%   faces fixed, and that portions of the net that diverge after the ending face are irrelevant,
+%   we arrive at that number of rotations that we need consider}.  While this is not
+%   insurmountable, it is computationally costly. However, for a valid energy score, a true distance
+%   is not necessary.  What is needed is a negative definite kernel.
+
+% \begin{prop}
+% For points $a,b \in {\mathbb S}_{\infty}^{d-1}$ a valid negative definite kernel can be formed as
+%   \begin{equation}
+%     g(\bm{a},\bm{b}) = \begin{cases}
+%         \pnorm{\bm{b}-\bm{a}}{2} &\text{ if }\argmax_{\ell}\bm{a} = \argmax_{\ell}\bm{b}\\
+%         \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} &\text{ otherwise}
+%     \end{cases}
+%   \end{equation}
+%   where $\bm{c}$ resides on the intersection between the faces of $\bm{a}$ and $\bm{b}$, and minimizes $g(\bm{a},\bm{b})$.
+% \end{prop}
+
+% In the first case, where $\bm{a}$ and $\bm{b}$ reside on the same face, then that is trivially a
+%   negative definite kernel.  In the second case, where $\bm{a}$ and $\bm{b}$ reside on separate
+%   faces, we can be assured of symmetry in the arguments as $\bm{c}$ is unique, and $\pnorm{\bm{c} - \bm{a}}{2} = \pnorm{\bm{a} - \bm{c}}{2}$.  Then for the other property of negative definiteness,
+%   \begin{equation*}
+%     \begin{aligned}
+%       \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j g(\bm{x}_1,\bm{x}_2) &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j \bigg[\pnorm{\bm{c} - \bm{x}_1}{2} + \pnorm{\bm{x}_2 - \bm{c}}{2}\bigg]\\
+%       &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{c} - \bm{x}_1}{2} + \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{x}_2 - \bm{c}}{2}
+%     \end{aligned}
+%   \end{equation*}
+%   which is less than $0$ in both terms by definition.
+
+% To compute this energy score as described would require optimization of the interim point $\bm{c}$.
+%   If we again consider the one-dimensional rotation in Equation~\ref{eq:1drotation}---we
+%   can view that as a numerically easier analogue accomplishing the same goal: the distance from
+%   the starting point, to some optimal point along the intersection between the starting and
+%   ending faces, to the ending point.  The strait line will pass through the optimal point, so
+%   the optimization of the interim point is accomplished \emph{en passant}.
 

### 2021-10-12 introduction.tex
updates
update

@@
 \section{Introduction}
+
 The statistical analysis of extreme values focuses on inference for
@@
 Inferential methods for the extreme values of univariate observations
-are well established and software is widely available \cite[see, for
-example,][]{coles2001}.
+are well established and software is widely available \cite[see, for example,][]{coles2001}.
 For variables in one dimension the application of EVT methods consider the asymptotic distribution of either the maxima calculated for regular blocks of data, or the values that exceed a certain threshold.
@@
   The reminder of this paper is outlined as follows.
-{\bf We need a roadmap.
-Part of the roadmap should mention the analysis of the IVT.
-Basically condensing the paragraph below in a couple of sentences.}
+Section~\ref{sec:multivariatepot} comprises a brief review of multivariate PoT, detailing the separation of the radial measure from the angular measure.
+  Section~\ref{sec:methodology} details our routine for estimating the angular measure, where~\ref{subsec:projection} elaborates the process of projecting an arbitrary distribution supported in ${\mathbb R}_+^d$ onto an a series of manifolds ${\mathbb S}_p^{d-1}$, $p = 1,2,\ldots$ approaching the support of the angular measure, on to which~\ref{subsec:projgamma} builds to develop a \emph{projected Gamma} family of distributions.
+Section~\ref{sec:evaluation} is concerned with developing criteria for model selection in the support of the angular measure.
+Section~\ref{sec:results} details the efficacy of the projected gamma model in recovering the distribution of data on the support of the angular measure---in~\ref{subsec:simulated}, this is done on simulated data, and in~\ref{subsec:ivt}, acknowledging the relevance of of extreme value theory to climatological events~\citep{jentsch2007,vousdoukas2018,li2019}, we estimate the extremal dependence structure for the integrated vapor transport---a measure of water vapor in the atmosphere, used for identifying atmospheric rivers.
+Inferences from the dependence structure detailed in~\ref{subsec:ped} and~\ref{subsec:condsurv} are applied.
+Finally, Section~\ref{sec:conclusion} concludes.
 
-To motivate the topic and our proposed solution, we acknowledge the relevance of extreme analysis to
-  climatological events.  Outcomes such as heat waves, extreme precipitation, storm surges, and
-  such present a similar challenge to investigators, in that the events of interest can be very destructive,
-  and occur in probability at the tail end of \emph{normal} behavior.  In this fashion, a model built
-  to capture \emph{normal} behavior may struggle to adequately represent extreme behavior.  More
-  specialized tools become necessary; especially so in a time of intensifying extreme weather
-  events~\citep{jentsch2007,vousdoukas2018,li2019}.  To this end, we use the integrated water vapor
-  transport (\emph{IVT}) data for observing atmospheric rivers as an example application for our proposed method.
+% Extreme value theory bears strong relevance to climate analysis, as climatological events of interest occur at the tail of the distribution, both in terms of magnitude and frequency.
+We motivate our proposed model with an analysis of the integrated vapor transport data for observing atmospheric rivers---high concentrations of water vapor that lead to extreme precipitation events.
+
+% To motivate the topic and our proposed solution, we acknowledge the relevance of extreme analysis to
+%   climatological events.  Outcomes such as heat waves, extreme precipitation, storm surges, and
+%   such present a similar challenge to investigators, in that the events of interest can be very destructive,
+%   and occur in probability at the tail end of \emph{normal} behavior.  In this fashion, a model built
+%   to capture \emph{normal} behavior may struggle to adequately represent extreme behavior.  More
+%   specialized tools become necessary; especially so in a time of intensifying extreme weather
+%   events.  To this end, we use the integrated water vapor
+%   transport (\emph{IVT}) data for observing atmospheric rivers as an example application for our proposed method.
 
@@
 
-\section{A multivariate PoT model}
+\section{A multivariate PoT model\label{sec:multivariatepot}}
 %\subsubsection{Univariate Maxima}
@@
 \[
-    \lim\limits_{n\to\infty}n\text{Pr}\left[\frac{1}{n}\bm{Z}\geq \bm{z}\right] = \mu\left([\bm{0},\bm{z}]^c\right),
+    \lim\limits_{n\to\infty}n\prob{\frac{1}{n}\bm{Z}\geq \bm{z}} = \mu\left([\bm{0},\bm{z}]^c\right),
 \]
@@
   \emph{exponent measure}. Under this condition $\mu$ features the homogeneity property $\mu(tA) = \frac{1}{t}\mu(A)$. 
-Define that $R = \max_{i} Z_i$, and assume that $R>1$ with probability one. Also, define $V_i = Z_i/\max_i Z_i$, and assume that $E(V_i)>0$.
+Define  $R = \max_{i} Z_i$, and assume that $R>1$ with probability one. Also, define $V_i = Z_i/\max_i Z_i$, and assume that $E(V_i)>0$.
 These two assumption together with the homogeneity of $\mu$ correspond to the second condition in Theorem 2.1 of
@@
 \end{equation}
-$R$ describes the radial component, and $\bm{V} \in \mathcal{S}_{\infty}^{d-1}$ the 
+$R$ describes the radial component, and $\bm{V} \in {\mathbb S}_{\infty}^{d-1}$ the 
 angular component of $\bm{Z}$.  Here ${\mathbb S}_\infty^{d-1}$ denotes the positive
@@
 \begin{equation} \label{MGP}
-    \text{Pr}\left[\bm{V} \in A \mid R > r\right]
-      = \frac{r\text{Pr}\left[\bm{V} \in A, R > r\right]}{r\text{Pr}[R > r]}\hspace{0.2cm}
-      \xrightarrow[r\to\infty]{~} \hspace{0.2cm} \frac{\Phi(A)}{\Phi(\mathcal{S}_{\infty}^{d-1})}.
+    \prob{\bm{V} \in A \mid R > r} = \frac{r\prob{\bm{V} \in A, R > r}}{r\prob{R > r}}\hspace{0.2cm}
+      \xrightarrow[r\to\infty]{~} \hspace{0.2cm} \frac{\Phi(A)}{\Phi({\mathbb S}_{\infty}^{d-1})}.
 \end{equation}
@@
 \begin{equation*}
-	\text{Pr}[Z_1\leq z_1, \ldots ,Z_d\leq z_d] = E\left(\bigvee_{i=1}^d
-\frac{V_i}{z_i\wedge 1}\right) - E\left(\bigvee_{i=1}^d
-\frac{V_i}{z_i}\right) 
+    \prob{Z_1\leq z_1, \ldots ,Z_d\leq z_d} = \expect{\bigvee_{i=1}^d
+\frac{V_i}{z_i\wedge 1}} - \expect{\bigvee_{i=1}^d
+\frac{V_i}{z_i}}
 \end{equation*}
@@
 In view of this, to obtain a PoT model for $\bm{Z}$, we seek 
-a flexible distribution on $\bm{V} \in \mathcal{S}_{\infty}^{d-1}$.
+a flexible distribution on $\bm{V} \in {\mathbb S}_{\infty}^{d-1}$.
 
@@
 
+\subsection{Pairwise Extremal Dependence\label{subsec:ped}}
 A measure that is used to characterize the strength of the dependence,
@@
 distributions $F_1$ and $F_2$ is given by
-\[	\chi_{12} = \lim_{u\uparrow 1} \text{Pr}[ F_1(W_1)>u|F_2(W_2)>u]  \]
+\[	\chi_{12} = \lim_{u\uparrow 1} \prob{F_1(W_1)>u|F_2(W_2)>u}  \]
 \citep{coles2001}.
@@
 independent. The following result provides the asymptotic dependence coefficient for two components of $\bm{Z}$. 
-\begin{prop} \label{ppchi}
+\begin{prop}\label{ppchi}
 Suppose that $\bm{Z} = R\bm{V}$ with $R\sim Pa(1)$,
-$\text{Pr}[V_i>0] = 1$ and $E(V_i)$ exists ,
+$\prob{V_i > 0} = 1$ and $\expect{V_i}$ exists ,
 $i=1, \ldots ,d$, then
 \begin{equation}
-	\chi_{ij} = E\left(\frac{V_i}{E(V_i)} \wedge 
-\frac{V_j}{E(V_j)}\right).
+	\chi_{ij} = \expect{\frac{V_i}{\expect{V_i}} \wedge \frac{V_j}{\expect{V_j}}}
 \end{equation}
@@
 Denote as $F_i$ the marginal distribution of $Z_i$. To obtain $\chi_{ij}$ we need $Pr(Z_1>z_1,Z_2>z_2)$, where $z_i =
-F_i^{-1}(u) = E(V_i)/(1 - u), \;i=1,\dots, d$.
+F_i^{-1}(u) = \expect{V_i}/(1 - u), \;i=1,\dots, d$.
 Using the fact that
 $V_i>0, \forall i$ almost surely, we have that the former is equal to
-\[	\text{Pr}\left[R>\frac{z_i}{V_i}\vee\frac{z_2}{V_j}\right] =
-E\left(1\wedge\left(\frac{z_i}{V_i}\vee\frac{z_j}{V_j}\right)
-^{-1} \right)
-= E\left( \frac{V_i}{z_i}\wedge\frac{V_j}{z_j}
-\right) = (1 - u)E\left(
-\frac{V_i}{EV_i}\wedge\frac{V_j}{EV_j}
-\right) ,
-\]
+\begin{equation*}
+\prob{R>\frac{z_i}{V_i}\vee\frac{z_2}{V_j}}
+    = \expect{1\wedge\left(\frac{z_i}{V_i}\vee\frac{z_j}{V_j}\right)^{-1}}
+    = \expect{\frac{V_i}{z_i}\wedge\frac{V_j}{z_j}}
+    = (1 - u)\expect{\frac{V_i}{\expect{V_i}}\wedge\frac{V_j}{\expect{V_j}}},
+\end{equation*}
 where the second identity is justified by the fact that $V_i$ is
 bounded and $z_i\rightarrow\infty$. The proof is completed by noting
-that $Pr(F_i(Z_i)>u ) = 1 - u.\hfill\Box$
+that $\prob{F_i(Z_i)>u} = 1 - u.\hfill\Box$
 
+\subsection{Multivariate Conditional Survival Function\label{subsec:condsurv}}
 For the analysis of extreme values it is of interest to calculate the multivariate conditional survival function.
@@
     \text{Pr}\left[\bigcap_{l \in \alpha} Z_l > z_l \right|\left. \bigcap_{l\not\in\alpha} Z_l > z_l\right] =
-      \frac{E\left(\bigwedge_{k = 1}^d 1\wedge \frac{V_k}{z_k}\right)}{E\left(\bigwedge_{k \not\in\alpha}1\wedge\frac{V_k}{z_k}\right)}.
+      \frac{\expect{\bigwedge_{k = 1}^d 1\wedge \frac{V_k}{z_k}}}{\expect{\bigwedge_{k \not\in\alpha}1\wedge\frac{V_k}{z_k}}}.
   \end{equation}

### 2021-10-12 methodology_pnorm.tex
update

@@
-\subsection{Projection onto an arbitrary unit hypersphere}
+\subsection{Projection onto an arbitrary unit hypersphere\label{subsec:projection}}
 
@@
   \begin{equation*}
-    \mathcal{S}_{p}^{d-1} = \left\lbrace \bm{s} : \bm{s} \in {\mathbb R}_{+}^{d}, \lVert \bm{s}\rVert_{p} = 1\right\rbrace,
+    {\mathbb S}_{p}^{d-1} = \left\lbrace \bm{s} : \bm{s} \in {\mathbb R}_{+}^{d}, \lVert \bm{s}\rVert_{p} = 1\right\rbrace,
   \end{equation*}
   the $\mathcal{L}_p$-norm unit hypersphere.
-  Given $\bm{x}\in {\mathbb R}^d_+$ the projection onto $\mathcal{S}_{p}^{d-1}$ is given as
-  $\bm{y} = \bm{x} / \lVert \bm{x}\rVert_p \in \mathcal{S}_{p}^{d-1}$.
+  Given $\bm{x}\in {\mathbb R}^d_+$ the projection onto ${\mathbb S}_{p}^{d-1}$ is given as
+  $\bm{y} = \bm{x} / \lVert \bm{x}\rVert_p \in {\mathbb S}_{p}^{d-1}$.
   Observations on one hypersphere can be projected without loss of information onto another by
@@
   \end{equation}
-  Equations (\ref{eqn:pnormt}) -- (\ref{eqn:pnormjac}) can be used to define a density on ${\mathbb R}^p_+$, and then obtain a density on $\mathcal{S}_{p}^{d-1}$, for any finite $p>0$.
+  Equations (\ref{eqn:pnormt}) -- (\ref{eqn:pnormjac}) can be used to define a density on ${\mathbb R}^p_+$, and then obtain a density on ${\mathbb S}_{p}^{d-1}$, for any finite $p>0$.
 Unfortunately in the limit, when $p\rightarrow +\infty$, the transformation is not differentiable.
-Thus, even though the projected distribution on $\mathcal{S}_\infty^{d-1}$ is defined, there is no corresponding density.
+Thus, even though the projected distribution on ${\mathbb S}_\infty^{d-1}$ is defined, there is no corresponding density.
 

### 2021-10-12 methodology_projgamma.tex
update

@@
-\subsection{Projected gamma family}
+\subsection{Projected gamma family\label{subsec:projgamma}}
   A natural  distribution to consider in ${\mathbb R}^p_+$ is given by a product of independent
@@
     $\bm{ X} \sim \prod_{\ell = 1}^d\text{Ga}\left(X_{\ell}\mid\alpha_{\ell},\beta_{\ell}\right)$, 
-    $a_\ell$ and $b_\ell$ are the shape and scale parameters, repectively. Using Equations (\ref{eqn:pnormt}) -- (\ref{eqn:pnormjac}) we have the joint density
+    $\alpha_\ell$ and $\beta_\ell$ are the shape and scale parameters, respectively. Using Equations (\ref{eqn:pnormt}) -- (\ref{eqn:pnormjac}) we have the joint density
   \begin{equation*}
@@
           \exp\lbrace-\beta_{\ell}ry_{\ell}\rbrace\right]
-      \times r^{d-1}\left[y_d -
+      \times r^{d-1}\left[y_d +
             {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right],
@@
                 y_{\ell}^{\alpha_{\ell} - 1}\right]
-      \times \left[y_d -
+      \times \left[y_d +
           {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right]
@@
   \end{equation}
-  defined for $\bm{y}\in \mathcal{S}_p^{d-1}$, and for any $p>0$. To avoid idetifiability problems when estimating the shape and scale parameters, we set $\beta_1 = 1$.
+  defined for $\bm{y}\in {\mathbb S}_p^{d-1}$, and for any $p>0$. To avoid idetifiability problems when estimating the shape and scale parameters, we set $\beta_1 = 1$.
   \cite{nunez2019} obtain the density in Equation (\ref{eqn:projgamma})
   for $p=2$ as a multivariate distribution for directional data, using spherical coordinates.
-  For $\bm{ y}\in \mathcal{S}_1^{d-1}$, and $\beta_{\ell} = \beta, \forall\ell$, the density in Equation (\ref{eqn:projgamma}) corresponds to that of a Dirichlet distribution.
+  For $\bm{ y}\in {\mathbb S}_1^{d-1}$, and $\beta_{\ell} = \beta, \forall\ell$, the density in Equation (\ref{eqn:projgamma}) corresponds to that of a Dirichlet distribution.
 
@@
 Thus, we use it as a building block for the angular measure $\Phi$ models.
-To build a flexible family of distributions in $\mathcal{S}_p^{d-1}$ we consider mixtures of projected gamma densities defined as
-  \begin{equation} \label{PGmix}     
+To build a flexible family of distributions in ${\mathbb S}_p^{d-1}$ we consider mixtures of projected gamma densities defined as
+  \begin{equation} \label{eqn:PGmix}     
      f(\bm{y}) = \int_\Theta \text{PG}(\bm{y}|\bm{\theta}) dG(\bm{\theta}), 
@@
 
-  Our strategy consists of describing the angular distribution $\Phi$ using a sample based approach with the following steps: (i) Apply the transformation in Equation (\ref{eqn:standardization}) to the original data; (ii) Obtain the subsample of the standardized observations that satisfy $R>1$; (iii) Take a finite $p$ and project the observations onto $\mathcal{S}_p^{d-1}$; (iv) Fit the model in Equation (\ref{PGmix}) to the resulting data and obtain samples from the fitted model; (v) Project the resulting samples onto $\mathcal{S}_\infty^{d-1}$.
+  Our strategy consists of describing the angular distribution $\Phi$ using a sample based approach with the following steps: (i) Apply the transformation in Equation (\ref{eqn:standardization}) to the original data; (ii) Obtain the subsample of the standardized observations that satisfy $R>1$; (iii) Take a finite $p$ and project the observations onto ${\mathbb S}_p^{d-1}$; (iv) Fit the model in Equation (\ref{PGmix}) to the resulting data and obtain samples from the fitted model; (v) Project the resulting samples onto ${\mathbb S}_\infty^{d-1}$.
   For step (iv) we use a Bayesian approach that is implemented using a purposely developed Markov chain Monte Carlo that is described in the next section. 
@@
 
-\subsubsection{Inference for the projected gamma mixture model}
-{\bf Our focus is on model (\ref{PGmix}).
+\subsubsection{Inference for the projected Gamma mixture model}
+A sample-based method of inference for the Dirichlet process
+
+Introducing $\bm{\pi}$, $\sum_j\pi_j = 1$ as the mixture weighting, Equation~\ref{eqn:PGmix} can be recomposed as
+\begin{equation*}
+    f(\bm{y}\mid\bm{\alpha},\bm{\beta},\bm{\pi}) = \int_{\Omega}\sum_{j = 1}^{\infty}\left[\pi_j\text{PG}(\bm{y}\mid\bm{\theta}_j)\right]dG_0(\bm{\theta}_j)dF(\bm{\pi}),
+\end{equation*}
+which in turn can be further recomposed as
+\begin{equation}
+    \label{eqn:dppgmix}
+    f(\bm{y}\mid\bm{\alpha},\bm{\beta},\bm{\pi}) = {\int\int}_{\Omega} \prod_{j = 1}^{\infty}\left[\pi_j\text{PG}\left(\bm{y}\mid\alpha_j,\beta_j\right)\right]^{(\delta = j)}dG_0(\bm{\theta}_j)dF(\bm{\delta})dF(\bm{\pi}).
+\end{equation}
+Thus, by sampling the latent parameter $\delta$ as a mixture component identifier, we can break the mixture into components and perform inference on components.
+By integrating out $\bm{\pi}$ from Equation~\ref{eqn:dppgmix}, we arrive at the collapsed sampler formulation of the Dirichlet process\needcite.
+Under this interpretation, for extant clusters the probability of cluster membership is proportional to
+\begin{equation*}
+    \prob{\delta_i = j\mid \bm{y}_i, \bm{\theta}_j} \propto \frac{n_j^{\neg i}}{\sum_{k = 1}^{J}n_k^{\neg i} + \eta}\text{PG}\left(\bm{y}_i\mid\bm{\theta}_j\right),
+\end{equation*}
+where $n_j^{\neg i}$ is the size of cluster $j$ not counting observation $i$, and $J$ the number of extant clusters.
+Formation of new clusters is controlled by $\eta$, the concentration parameter for the Dirichlet process.
+Ideally, the probability of a new cluster is proportional to
+\begin{equation*}
+    \prob{\delta_i = J + 1\mid\theta_j,\eta} \propto \int\frac{\eta}{\sum_{k = 1}^{J}n_j^{\neg i} + \eta}\text{PG}\left(y_i\mid\bm{\theta}\right)dG_0(\bm{\theta}).
+\end{equation*}
+Unfortunately, in a model where $G_0(\bm{\theta})$ is not conjugate to the kernel distribution, this integral is not directly tractable (and a method for sampling the new cluster parameters given a single observation not obvious).
+Instead, following Algorithm 8 from~\cite{neal2000}, we compute this integral implicitly (and find samples for the new cluster parameters) by sampling $m$ \emph{candidate clusters} from $dG_0(\bm{\theta})$.
+Then the probability of cluster membership in any of these candidate clusters is proportional to
+\begin{equation*}
+    \prob{\delta_i = j^{*}\mid\theta_{j^*}, \eta} \propto \frac{\eta / m}{\sum_{k = 1}^{J}n_k^{\neg i} + \eta}\text{PG}\left(\bm{y}_i\mid\theta_{j^*}\right),
+\end{equation*}
+where $j^*$ specifies the index for a candidate cluster.  Upon sampling $\delta$ from extant and candidate clusters, if a candidate cluster is selected, then its cluster parameters are saved.
+
+As stated before, the projected Gamma model offers very tractable computational properties.  After breaking the mixture, the likelihood for the $j$th cluster is proportional to
+\begin{equation}
+    \label{eqn:pglikelihood}
+    L(\bm{\theta}_j\mid \bm{Y},\bm{\delta}) = \prod_{i:\delta_i = j}\text{PG}\left(\bm{y}_i\mid \bm{\theta_j}\right) \propto \prod_{i:\delta_i = j}\int_0^{\infty}\prod_{\ell = 1}^d\text{Ga}\left(r_iy_{i\ell}\mid\theta_{j\ell}\right)r_i^{d-1}dr_i.
+\end{equation}
+By reintroducing the latent parameter $r$, the likelihood with respect to the distributional parameters becomes independent between dimensions.
+An appropriately chosen centering distribution that preserves that independence between dimensions will enable us to perform inference on each dimension independently.
+
+The form of centering distribution $G_0$ deserves further mention.
+In the Gibbs sampler, for inference on $\bm{\theta}_j$, the centering distribution acts as a prior.
+Given $r$, the conjugate prior for $\beta_{j\ell}$ is another Gamma distribution---this allows us to sample it directly, and integrate it out for inference on $\alpha_{j\ell}$.
+A conjugate prior for $\alpha_{j\ell}$ exists, though sampling from it is not straightforward.
+Instead, a logical prior for $\alpha_{j\ell}$ is another Gamma distribution, and exploring the posterior of $\alpha_{j\ell}$ is accomplished with a Metropolis-within-Gibbs step.
+This logic develops the \emph{Gamma} form for the centering distribution:
+\begin{equation*}
+    G_0(\bm{\alpha}_j,\bm{\beta}_j) =
+    \prod_{\ell = 1}^d\text{Ga}\left(\alpha_{j\ell}\mid\xi_\ell,\tau_\ell\right) \times \prod_{\ell = 2}^d\text{Ga}\left(\beta_{j\ell}\mid\zeta_\ell,\sigma_\ell\right),
+\end{equation*}
+with Gamma priors for $\xi_\ell$, $\tau_\ell$, $\zeta_\ell$, and $\sigma_\ell$.  Under this model, given $r$, inference is independent between dimensions.
+
+A slightly different choice of centering distribution also presents promise when we consider how the DP forms new clusters, in the algorithm we are using.
+As stated, at each iteration the algorithm generates $m$ new candidate clusters from $G_0$.
+A form of centering distribution that forces independence between dimensions may have detrimental effects on performance in the formation of new clusters.
+Instead, we might choose a centering distribution that shares information between dimensions in a covariance structure.
+This logic develops the \emph{log-normal} form for the centering distribution:
+\begin{equation}
+    G_0(\bm{\alpha}_j,\bm{\beta}_j) = \mathcal{LN}\left(\bm{\alpha}_j\mid\bm{\mu},\Sigma\right)\prod_{\ell = 2}^d\text{Ga}\left(\beta_{j\ell}\mid\zeta_\ell,\sigma_\ell\right)
+\end{equation}
+with a normal prior for $\bm{\mu}$, an inverse-Wishart prior for $\Sigma$, and Gamma priors for $\zeta_{\ell}$ and $\sigma_{\ell}$.
+This form enables us to integrate out $\beta_{j\ell}$ for $\ell = 2,\ldots, d$ to perform inference on $\bm{\alpha}_j$.
+Inference on $\bm{\alpha}_j$ can be accomplished with a $d$-dimensional Metropolis step.
+
+The \emph{restricted} form for the projected Gamma model holds $\beta_{j\ell} := 1$ for all $\ell$.  This simplifies inference for both the Gamma and log-normal models.
+
+% {\bf Our focus is on model (\ref{eqn:PGmix}).
 Describe how you build a sample based approach to fitting that model.
@@
 
-We consider a basic model that assumes all observations are generated by a single projected Gamma
-  distribution.  The shape and rate parameters for the latent \text{Gamma} random variables,
-  $\alpha_{\ell},\beta_{\ell}$, must be greater than 0, so a logical prior to use would be a Gamma prior for
-  each parameter, excepting for $\beta_1 := 1$.
-  \begin{equation}
-    \label{eqn:vanillap}
-    \begin{aligned}
-      \bm{ y}\mid\alpha,\beta &\sim \text{PG}(\bm{ y}\mid\alpha,\beta)\\
-      \bm{ \alpha},\bm{\beta} &\sim {\textstyle \prod}_{\ell = 1}^d \text{Ga}(\alpha_{\ell} \mid \xi_{\ell},\tau_{\ell})
-              \times {\textstyle \prod}_{\ell = 2}^d \text{Ga}(\beta_{\ell}\mid \zeta_{\ell},\sigma_{\ell}).
-    \end{aligned}
-  \end{equation}
-  A useful property of the Projected Gamma family is the latent independent Gammas assumption.  Recall
-  in Eqn.~\ref{eqn:protopg}, we have a density defined as a product of independent Gamma densities,
-  multiplied by a factor of $r^{d-1}C$, where $C$ is constant with respect to the parameters
-  $\alpha,\beta$, as well as $r$.  If we conduct inference via a Gibbs sampler, then by simulating
-  this latent $r$ we are able to perform inference of $(\alpha_{\ell},\beta_{\ell})$ independent of
-  that of $(\alpha_k,\beta_k)$.  The full conditional for $r$ is described in Eqn.~\ref{eqn:rfullcond}.
-  The full conditional for $\beta_{\ell}$, $l = 2,\ldots,d$ then becomes
-  \begin{equation}
-    \label{eqn:betafc}
-    \beta_{\ell}\mid \bm{ y}, r, \alpha_{\ell} \sim \text{Ga}\left(n + \alpha_{\ell} + \zeta_{\ell},
-                                      {\textstyle \sum}_{i = 1}^nr_iy_{il} + \sigma_{\ell}\right).
-  \end{equation}
-  That is, given $\alpha_{\ell}$, $r$, we can sample $\beta_{\ell}$ directly as a Gamma random
-  variable.  In the sampling for $\alpha_{\ell}$, we can integrate $\beta_{\ell}$ out leaving us a
-  quantity proportional to a density, which we can sample via the Metropolis Hastings
-  algorithm\citep{hastings1970}.  For $\alpha_{\ell}$, $\ell = 2,\ldots, d$, This takes the form
-  \begin{equation}
-    \label{eqn:alphafc}
-    f(\alpha_{\ell} \mid \bm{ y}, r) \propto
-    \frac{\left({\textstyle \prod}_{i = 1}^nr_iy_{il}\right)^{\alpha_{\ell} - 1}}{\Gamma^n(\alpha_{\ell})} \times
-    \alpha_{\ell}^{\xi_{\ell} - 1}\exp\{-\tau_{\ell}\alpha_{\ell}\} \times
-    \frac{\Gamma(n\alpha_{\ell} + \zeta_{\ell})}{
-        \left({\textstyle\sum}_{i = 1}^n r_iy_{il} + \sigma_{\ell}\right)^{(n * \alpha_{\ell} + \zeta_{\ell})}}.
-  \end{equation}
-  For $\alpha_1$, as we do not integrate out $\beta_1$ as it is defined to be $1$, this takes the
-  slightly simpler form
-  \begin{equation}
-    \label{eqn:alphafc1}
-    f(\alpha_1 \mid \bm{ y}, r) \propto
-      \frac{{\textstyle\prod}_{i = 1}^n ry_{i1}^{\alpha_1 - 1}}{\Gamma^n(\alpha_1)} \times
-      \alpha_1^{\xi_1 - 1}\exp\{-\tau_1\alpha_1\}.
-  \end{equation}
-  The \emph{projected restricted Gamma} model sets $\beta_{\ell} := 1 \forall \ell \in \lbrace 1,\ldots,d\rbrace$,
-  so in that model inference on all $\alpha_{\ell}$ will use the simpler form.
+% We consider a basic model that assumes all observations are generated by a single projected Gamma
+%   distribution.  The shape and rate parameters for the latent \text{Gamma} random variables,
+%   $\alpha_{\ell},\beta_{\ell}$, must be greater than 0, so a logical prior to use would be a Gamma prior for
+%   each parameter, excepting for $\beta_1 := 1$.
+%   \begin{equation}
+%     \label{eqn:vanillap}
+%     \begin{aligned}
+%       \bm{ y}\mid\alpha,\beta &\sim \text{PG}(\bm{ y}\mid\alpha,\beta)\\
+%       \bm{ \alpha},\bm{\beta} &\sim {\textstyle \prod}_{\ell = 1}^d \text{Ga}(\alpha_{\ell} \mid \xi_{\ell},\tau_{\ell})
+%               \times {\textstyle \prod}_{\ell = 2}^d \text{Ga}(\beta_{\ell}\mid \zeta_{\ell},\sigma_{\ell}).
+%     \end{aligned}
+%   \end{equation}
+%   A useful property of the Projected Gamma family is the latent independent Gammas assumption.  Recall
+%   in Eqn.~\ref{eqn:protopg}, we have a density defined as a product of independent Gamma densities,
+%   multiplied by a factor of $r^{d-1}C$, where $C$ is constant with respect to the parameters
+%   $\alpha,\beta$, as well as $r$.  If we conduct inference via a Gibbs sampler, then by simulating
+%   this latent $r$ we are able to perform inference of $(\alpha_{\ell},\beta_{\ell})$ independent of
+%   that of $(\alpha_k,\beta_k)$.  The full conditional for $r$ is described in Eqn.~\ref{eqn:rfullcond}.
+%   The full conditional for $\beta_{\ell}$, $l = 2,\ldots,d$ then becomes
+%   \begin{equation}
+%     \label{eqn:betafc}
+%     \beta_{\ell}\mid \bm{ y}, r, \alpha_{\ell} \sim \text{Ga}\left(n + \alpha_{\ell} + \zeta_{\ell},
+%                                       {\textstyle \sum}_{i = 1}^nr_iy_{il} + \sigma_{\ell}\right).
+%   \end{equation}
+%   That is, given $\alpha_{\ell}$, $r$, we can sample $\beta_{\ell}$ directly as a Gamma random
+%   variable.  In the sampling for $\alpha_{\ell}$, we can integrate $\beta_{\ell}$ out leaving us a
+%   quantity proportional to a density, which we can sample via the Metropolis Hastings
+%   algorithm\citep{hastings1970}.  For $\alpha_{\ell}$, $\ell = 2,\ldots, d$, This takes the form
+%   \begin{equation}
+%     \label{eqn:alphafc}
+%     f(\alpha_{\ell} \mid \bm{ y}, r) \propto
+%     \frac{\left({\textstyle \prod}_{i = 1}^nr_iy_{il}\right)^{\alpha_{\ell} - 1}}{\Gamma^n(\alpha_{\ell})} \times
+%     \alpha_{\ell}^{\xi_{\ell} - 1}\exp\{-\tau_{\ell}\alpha_{\ell}\} \times
+%     \frac{\Gamma(n\alpha_{\ell} + \zeta_{\ell})}{
+%         \left({\textstyle\sum}_{i = 1}^n r_iy_{il} + \sigma_{\ell}\right)^{(n * \alpha_{\ell} + \zeta_{\ell})}}.
+%   \end{equation}
+%   For $\alpha_1$, as we do not integrate out $\beta_1$ as it is defined to be $1$, this takes the
+%   slightly simpler form
+%   \begin{equation}
+%     \label{eqn:alphafc1}
+%     f(\alpha_1 \mid \bm{ y}, r) \propto
+%       \frac{{\textstyle\prod}_{i = 1}^n ry_{i1}^{\alpha_1 - 1}}{\Gamma^n(\alpha_1)} \times
+%       \alpha_1^{\xi_1 - 1}\exp\{-\tau_1\alpha_1\}.
+%   \end{equation}
+%   The \emph{projected restricted Gamma} model sets $\beta_{\ell} := 1 \forall \ell \in \lbrace 1,\ldots,d\rbrace$,
+%   so in that model inference on all $\alpha_{\ell}$ will use the simpler form.
 
-\subsubsection{Finite Mixture Model}
-The projected Gamma distribution offers a flexible model for describing direction.  However, it is
-  conceivable that a single Projected Gamma distribution will not sufficiently capture the
-  variability of real data.  In that case, a logical extension to make to the above model is to
-  assume the data, $\bm{ y}$, are generated from a mixture of projected Gammas.  We consider a finite
-  mixture,
-  \begin{equation}
-    \label{eqn:fmixp}
-    \begin{aligned}
-      \bm{ y}_i &\sim \sum_{j = 1}^J\pi_j\text{PG}\left(\bm{ y}\mid \bm{ \alpha}_j, \bm{ \beta}_j\right)\\
-      \bm{ \alpha}_j &\sim {\textstyle \prod}_{\ell = 1}^d \text{Ga}\left(\alpha_{\ell}\mid\xi_{\ell},\tau_{\ell}\right)\\
-      \bm{ \beta}_j &\sim {\textstyle \prod}_{\ell = 2}^d \text{Ga}\left(\beta_{\ell}\mid\zeta_{\ell},\sigma_{\ell}\right)
-    \end{aligned}
-    \hspace{2cm}
-    \begin{aligned}
-      \bm{ \xi},\bm{\tau} &\sim {\textstyle \prod}_{\ell = 1}^d \text{Ga}(\xi_{\ell}\mid a,b) \times \text{Ga}(\tau_{\ell}\mid c,d)\\
-      \bm{ \zeta},\bm{\sigma} &\sim {\textstyle\prod}_{\ell = 2}^d\text{Ga}(\zeta_{\ell} \mid a,b)\times \text{Ga}(\sigma_{\ell}\mid c,d)\\
-      \bm{ \pi} &\sim \text{Dir}(\pi_0)
-    \end{aligned}
-  \end{equation}
-  In addition to allowing multiple mixture components, we assume those mixture components themselves
-  descend from a product of independent Gamma densities. The standard method for sampling from a
-  mixture distribution includes a data augmentation step, generating a
-  $\delta_i \in \lbrace 1,\ldots, J\rbrace$ indicating to which mixture component a
-  particular observation comes from, with
-  \begin{equation*}
-    \text{P}\left[\delta_i = j\mid\alpha_j,\beta_j\right] \propto \pi_j\text{PG}\left(\bm{ y}_i\mid \bm{ \alpha}_j,\bm{ \beta}_j\right).
-  \end{equation*}
-  By sampling and conditioning on $\delta_i$, we end up with very similar semi-conjugate forms for
-  inference of $\bm{ \alpha}_j$, $\bm{ \beta}_j$ that were used in the basic model.  The changes made
-  to Equations~\ref{eqn:betafc},\ref{eqn:alphafc},\ref{eqn:alphafc1} are such that we're conducting inference
-  on a mixture component given the mixture component identifiers $\bm{ \delta}$, so instead of indexing
-  $i$ over the whole data set, we're indexing $i$ over the set $\{i :\delta_i = j\}$, with $n = n_j$,
-  the cardinality of that set. Inference on $\bm{ \xi,\tau,\zeta,\sigma}$ follows a similar form as
-  that of $\bm{ \alpha,\beta}$---they are Gamma-Gamma relationships, so rate parameters have a conjugate
-  Gamma posterior, and we can integrate them out to perform inference on the shape parameters via
-  Metropolis Hastings.
+% \subsubsection{Finite Mixture Model}
+% The projected Gamma distribution offers a flexible model for describing direction.  However, it is
+%   conceivable that a single Projected Gamma distribution will not sufficiently capture the
+%   variability of real data.  In that case, a logical extension to make to the above model is to
+%   assume the data, $\bm{ y}$, are generated from a mixture of projected Gammas.  We consider a finite
+%   mixture,
+%   \begin{equation}
+%     \label{eqn:fmixp}
+%     \begin{aligned}
+%       \bm{ y}_i &\sim \sum_{j = 1}^J\pi_j\text{PG}\left(\bm{ y}\mid \bm{ \alpha}_j, \bm{ \beta}_j\right)\\
+%       \bm{ \alpha}_j &\sim {\textstyle \prod}_{\ell = 1}^d \text{Ga}\left(\alpha_{\ell}\mid\xi_{\ell},\tau_{\ell}\right)\\
+%       \bm{ \beta}_j &\sim {\textstyle \prod}_{\ell = 2}^d \text{Ga}\left(\beta_{\ell}\mid\zeta_{\ell},\sigma_{\ell}\right)
+%     \end{aligned}
+%     \hspace{2cm}
+%     \begin{aligned}
+%       \bm{ \xi},\bm{\tau} &\sim {\textstyle \prod}_{\ell = 1}^d \text{Ga}(\xi_{\ell}\mid a,b) \times \text{Ga}(\tau_{\ell}\mid c,d)\\
+%       \bm{ \zeta},\bm{\sigma} &\sim {\textstyle\prod}_{\ell = 2}^d\text{Ga}(\zeta_{\ell} \mid a,b)\times \text{Ga}(\sigma_{\ell}\mid c,d)\\
+%       \bm{ \pi} &\sim \text{Dir}(\pi_0)
+%     \end{aligned}
+%   \end{equation}
+%   In addition to allowing multiple mixture components, we assume those mixture components themselves
+%   descend from a product of independent Gamma densities. The standard method for sampling from a
+%   mixture distribution includes a data augmentation step, generating a
+%   $\delta_i \in \lbrace 1,\ldots, J\rbrace$ indicating to which mixture component a
+%   particular observation comes from, with
+%   \begin{equation*}
+%     \text{P}\left[\delta_i = j\mid\alpha_j,\beta_j\right] \propto \pi_j\text{PG}\left(\bm{ y}_i\mid \bm{ \alpha}_j,\bm{ \beta}_j\right).
+%   \end{equation*}
+%   By sampling and conditioning on $\delta_i$, we end up with very similar semi-conjugate forms for
+%   inference of $\bm{ \alpha}_j$, $\bm{ \beta}_j$ that were used in the basic model.  The changes made
+%   to Equations~\ref{eqn:betafc},\ref{eqn:alphafc},\ref{eqn:alphafc1} are such that we're conducting inference
+%   on a mixture component given the mixture component identifiers $\bm{ \delta}$, so instead of indexing
+%   $i$ over the whole data set, we're indexing $i$ over the set $\{i :\delta_i = j\}$, with $n = n_j$,
+%   the cardinality of that set. Inference on $\bm{ \xi,\tau,\zeta,\sigma}$ follows a similar form as
+%   that of $\bm{ \alpha,\beta}$---they are Gamma-Gamma relationships, so rate parameters have a conjugate
+%   Gamma posterior, and we can integrate them out to perform inference on the shape parameters via
+%   Metropolis Hastings.
 
-Additionally, we investigate a finite mixture extension of the \emph{Projected Restricted Gamma} model,
-  where we define all $\beta_{\ell} := 1$, for $l = 1,\ldots,d$, for $j = 1,\ldots,J$.  This affects the above
-  model by removing inference steps for $\bm{ \beta},\bm{ \zeta},\bm{ \sigma}$.  The inference for
-  $\alpha_{\ell}$ for all $\ell = 1,\ldots,d$ follows the form in Equation~\ref{eqn:alphafc1}; inference on
-  $\bm{ \xi,\tau}$ follows the same form as the unrestricted model.  The impetus for the restricted
-  gamma model is that in the context of a mixture model, the additional flexibility offered by the
-  unrestricted Gamma model may be better represented with additional clusters.
+% Additionally, we investigate a finite mixture extension of the \emph{Projected Restricted Gamma} model,
+%   where we define all $\beta_{\ell} := 1$, for $l = 1,\ldots,d$, for $j = 1,\ldots,J$.  This affects the above
+%   model by removing inference steps for $\bm{ \beta},\bm{ \zeta},\bm{ \sigma}$.  The inference for
+%   $\alpha_{\ell}$ for all $\ell = 1,\ldots,d$ follows the form in Equation~\ref{eqn:alphafc1}; inference on
+%   $\bm{ \xi,\tau}$ follows the same form as the unrestricted model.  The impetus for the restricted
+%   gamma model is that in the context of a mixture model, the additional flexibility offered by the
+%   unrestricted Gamma model may be better represented with additional clusters.
 
-\subsubsection{Dirichlet Process Mixture Model}
-A logical extension to the finite mixture model described in Equation~\ref{eqn:fmixp} would be to assume the
-  $\bm{ y}$'s descending from an infinite mixture model.  This to some extent solves the issue of
-  selecting the number of mixture components, as that will effectively be done so in the model.
-  \begin{equation}
-    \label{eqn:infmixp}
-    \begin{aligned}
-      \bm{ y}_i &\sim \sum_{j = 1}^J\pi_j\text{PG}\left(\bm{ y}\mid \bm{ \alpha}_i, \bm{ \beta}_i\right)\\
-      (\alpha_i,\beta_i) &\sim \text{DP}\left(\eta, G_0\right)\\
-      &~\hspace{-2cm}G_0 = {\textstyle\prod}_{\ell = 1}^d\text{Ga}\left(\alpha_{jl}\mid\xi_{l},\tau_{l}\right)
-                    \times{\textstyle\prod}_{\ell = 2}^d\text{Ga}\left(\beta_{jl}\mid\zeta_{l}\tau_{l}\right)
-    \end{aligned}
-    \hspace{1cm}
-    \begin{aligned}
-      \bm{ \xi},\bm{ \tau} &\sim {\textstyle \prod}_{\ell = 1}^d \text{Ga}(\xi_{\ell}\mid a,b)\times \text{Ga}(\tau_{\ell}\mid c,d)\\
-      \bm{ \zeta},\bm{\sigma} &\sim {\textstyle\prod}_{\ell = 2}^d\text{Ga}(\zeta_{\ell} \mid a,b)\times \text{Ga}(\sigma_{\ell}\mid c,d)\\
-      \bm{ \eta} &\sim \text{Ga}(\eta \mid 2, 0.1).
-    \end{aligned}
-  \end{equation}
-  To fit the Dirichlet process, we use a variant of the collapsed Gibbs sampler detailed as
-  Algorithm~8 in \citet{neal2000}.  In this algorithm, we perform a data augmentation to create a
-  cluster identifier, $\delta_i$, for each observation. Let $J$ indicate the number of extant clusters.
-  Under the collapsed sampler scheme, for existing clusters, the probability of cluster membership is
-  calculated as
-  \begin{equation}
-    \label{eqn:dpdelta}
-    \text{P}\left(\delta_i = j\mid r, \bm{ y_i} \bm{ \alpha}, \bm{ \beta}\right) \propto
-        \frac{n_j^{\not i}}{\sum_{k = 1}^{J} n_k + \eta}
-        {\textstyle\prod}_{\ell = 1}^d\text{Ga}\left(r_iy_{il}\mid\alpha_{jl},\beta_{jl}\right).
-  \end{equation}
-  where $n_j^{\not i} = {\textstyle\sum}_{k\neq i}1_{k == j}$.  Effectively, if the observation under review is a
-  singleton, its existing cluster is destroyed.  For new clusters, ostensibly we would integrate out
-  cluster parameters to achieve a posterior predictive density; though doing so in a
-  non-conjugate model may be difficult.  Instead, the impetus for using this specific DP fitting algorithm
-  is it simulates the posterior predictive density by generating $m$ new candidate clusters from
-  their respective priors given $\bm{ \xi,\tau,\zeta,\sigma}$.  Then the probability of $y_i$
-  belonging to a particular candidate cluster is calculated as
-  \begin{equation}
-    \label{eqn:dpdeltanew}
-    \text{P}\left[\delta_i = k\mid r, \bm{ \alpha}_k^{\prime}, \bm{ \beta}_k^{\prime}\right] \propto
-        \frac{\eta / m}{\sum_{j = 1}^{J} n_j + \eta}
-        {\textstyle\prod}_{\ell = 1}^d\text{Ga}\left(r_iy_{il}\mid\alpha_{kl}^{\prime},\beta_{kl}^{\prime}\right),
-  \end{equation}
-  where $(\cdot)^{\prime}$ indicates a value as belonging to a new candidate cluster, and $k$ indexes
-  along candidate clusters.  If a new cluster is selected, then
-  $(\bm{\alpha}_{J+1},\bm{\beta}_{J+1}) = (\bm{\alpha}_k^{\prime},\bm{\beta}_k^{\prime})$, and $J = J + 1$.
+% \subsubsection{Dirichlet Process Mixture Model}
+% A logical extension to the finite mixture model described in Equation~\ref{eqn:fmixp} would be to assume the
+%   $\bm{ y}$'s descending from an infinite mixture model.  This to some extent solves the issue of
+%   selecting the number of mixture components, as that will effectively be done so in the model.
+%   \begin{equation}
+%     \label{eqn:infmixp}
+%     \begin{aligned}
+%       \bm{ y}_i &\sim \sum_{j = 1}^J\pi_j\text{PG}\left(\bm{ y}\mid \bm{ \alpha}_i, \bm{ \beta}_i\right)\\
+%       (\alpha_i,\beta_i) &\sim \text{DP}\left(\eta, G_0\right)\\
+%       &~\hspace{-2cm}G_0 = {\textstyle\prod}_{\ell = 1}^d\text{Ga}\left(\alpha_{jl}\mid\xi_{l},\tau_{l}\right)
+%                     \times{\textstyle\prod}_{\ell = 2}^d\text{Ga}\left(\beta_{jl}\mid\zeta_{l}\tau_{l}\right)
+%     \end{aligned}
+%     \hspace{1cm}
+%     \begin{aligned}
+%       \bm{ \xi},\bm{ \tau} &\sim {\textstyle \prod}_{\ell = 1}^d \text{Ga}(\xi_{\ell}\mid a,b)\times \text{Ga}(\tau_{\ell}\mid c,d)\\
+%       \bm{ \zeta},\bm{\sigma} &\sim {\textstyle\prod}_{\ell = 2}^d\text{Ga}(\zeta_{\ell} \mid a,b)\times \text{Ga}(\sigma_{\ell}\mid c,d)\\
+%       \bm{ \eta} &\sim \text{Ga}(\eta \mid 2, 0.1).
+%     \end{aligned}
+%   \end{equation}
+%   To fit the Dirichlet process, we use a variant of the collapsed Gibbs sampler detailed as
+%   Algorithm~8 in \citet{neal2000}.  In this algorithm, we perform a data augmentation to create a
+%   cluster identifier, $\delta_i$, for each observation. Let $J$ indicate the number of extant clusters.
+%   Under the collapsed sampler scheme, for existing clusters, the probability of cluster membership is
+%   calculated as
+%   \begin{equation}
+%     \label{eqn:dpdelta}
+%     \text{P}\left(\delta_i = j\mid r, \bm{ y_i} \bm{ \alpha}, \bm{ \beta}\right) \propto
+%         \frac{n_j^{\not i}}{\sum_{k = 1}^{J} n_k + \eta}
+%         {\textstyle\prod}_{\ell = 1}^d\text{Ga}\left(r_iy_{il}\mid\alpha_{jl},\beta_{jl}\right).
+%   \end{equation}
+%   where $n_j^{\not i} = {\textstyle\sum}_{k\neq i}1_{k == j}$.  Effectively, if the observation under review is a
+%   singleton, its existing cluster is destroyed.  For new clusters, ostensibly we would integrate out
+%   cluster parameters to achieve a posterior predictive density; though doing so in a
+%   non-conjugate model may be difficult.  Instead, the impetus for using this specific DP fitting algorithm
+%   is it simulates the posterior predictive density by generating $m$ new candidate clusters from
+%   their respective priors given $\bm{ \xi,\tau,\zeta,\sigma}$.  Then the probability of $y_i$
+%   belonging to a particular candidate cluster is calculated as
+%   \begin{equation}
+%     \label{eqn:dpdeltanew}
+%     \text{P}\left[\delta_i = k\mid r, \bm{ \alpha}_k^{\prime}, \bm{ \beta}_k^{\prime}\right] \propto
+%         \frac{\eta / m}{\sum_{j = 1}^{J} n_j + \eta}
+%         {\textstyle\prod}_{\ell = 1}^d\text{Ga}\left(r_iy_{il}\mid\alpha_{kl}^{\prime},\beta_{kl}^{\prime}\right),
+%   \end{equation}
+%   where $(\cdot)^{\prime}$ indicates a value as belonging to a new candidate cluster, and $k$ indexes
+%   along candidate clusters.  If a new cluster is selected, then
+%   $(\bm{\alpha}_{J+1},\bm{\beta}_{J+1}) = (\bm{\alpha}_k^{\prime},\bm{\beta}_k^{\prime})$, and $J = J + 1$.
 
-Inference upon $\bm{ \alpha}_j,\bm{ \beta}_j$ operates in the same manner as with the finite mixture
-  model.  For $\bm{ \xi,\tau,\zeta,\sigma}$, we have $J$ cluster components, indicating $J$
-  observations that again are under a gamma-gamma model.  The rate parameters have a previously
-  described conjugate posterior, that we can integrate out to more directly sample the shape
-  parameters.  As before, we will be looking at this model, and the restricted gamma form of this model, where
-  $\beta_{\ell} := 1$ for $\ell = 1,\ldots,d$.  This removes inference steps for $\bm{ \zeta},\bm{ \sigma}$
-  as well.
+% Inference upon $\bm{ \alpha}_j,\bm{ \beta}_j$ operates in the same manner as with the finite mixture
+%   model.  For $\bm{ \xi,\tau,\zeta,\sigma}$, we have $J$ cluster components, indicating $J$
+%   observations that again are under a gamma-gamma model.  The rate parameters have a previously
+%   described conjugate posterior, that we can integrate out to more directly sample the shape
+%   parameters.  As before, we will be looking at this model, and the restricted gamma form of this model, where
+%   $\beta_{\ell} := 1$ for $\ell = 1,\ldots,d$.  This removes inference steps for $\bm{ \zeta},\bm{ \sigma}$
+%   as well.
 
-\subsubsection{A Log-Normal Prior for Shape Parameters}
-A final extension we can make to this model is to allow for some structure, or dependence, in the prior
-  for shape parameters within a cluster.  The impetus for this approach is from recognizing that the
-  algorithm we are using for sampling the DP may not adequately represent the posterior density for
-  a new cluster, as it is assuming sampling from an independent prior.  If we account for dependence
-  between dimensions within the prior, then we may be able to more efficiently sample new clusters.
-  As shape parameters must be greater than 0, it is logical to assume a log normal prior, as that
-  allows us the familiar forms of a mean vector and covariance matrix for representing the source of
-  cluster parameters. So,
-  \begin{equation}
-    \label{eqn:dpln}
-    \begin{aligned}
-      \bm{ y}_i &\sim \sum_{j = 1}^J\pi_j\text{PG}\left(\bm{ y}\mid \bm{ \alpha}_i, \bm{\beta}_i\right)\\
-      (\bm{\alpha}_i, \bm{\beta}_i) &\sim \text{DP}\left(\eta, G_0\right)\\
-        &~\hspace{-1cm}G_0 = \mathcal{N}\left(\log\bm{ \alpha}_{j}\mid\mu,\Sigma\right)\times
-            {\textstyle\prod}_{\ell = 2}^d\text{Ga}\left(\beta_{j,\ell}\mid\zeta_{\ell},\sigma_{\ell}\right)
-    \end{aligned}
-    \hspace{1cm}
-    \begin{aligned}
-      \mu,\Sigma &\sim \mathcal{N}\left(\mu\mid\mu_0,\Sigma_0\right) \times \text{IG}\left(\nu,\Psi\right)\\
-      \bm{ \zeta},\bm{\sigma} &\sim {\textstyle\prod}_{\ell = 2}^d\text{Ga}(\zeta_{\ell} \mid a,b) \times \text{Ga}(\sigma_{\ell} \mid c,d) \\
-      \bm{ \eta} &\sim \text{Ga}(\eta \mid 2, 0.1).
-    \end{aligned}
-  \end{equation}
-  We hope that this will allow us to better incorporate the relationship between dimensions in
-  sampling new clusters, allowing for better mixing of the model.  The downside is one of computational
-  complexity---this introduces a $d$-dimensional normal distribution into the model, requiring a
-  $d$-dimensional matrix inversion step.  As matrix inversion is computationally costly, this limits
-  the potential number of dimensions possible.  The posterior inference strategies for $\beta,\zeta,\sigma$
-  are the same as for prior models and have been described.  For inference on $\alpha_j$ given
-  $r, \mu,\Sigma$, we rely on a $d$-dimensional Metropolis Hastings step, where the log-posterior
-  density is proportional to
-  \begin{equation}
-    \label{eqn:dplnlp}
-    \begin{aligned}
-    f(\bm{\alpha}_j\mid\bm{\delta},\bm{y},\mu,\Sigma,\zeta,\sigma) &\propto
-    \frac{\left(\prod_{i:\delta_i = j}y_{i,\ell}\right)^{\alpha_{j,\ell} - 1}}{\Gamma(\alpha_{j,\ell})^{n_j}}
-        \prod_{\ell = 2}^d\left[\frac{\left(\prod_{i:\delta_i = j}y_{i,\ell}\right)^{\alpha_{j,\ell} - 1}
-        \Gamma(n_j\alpha_{j,\ell} + \zeta)}{\Gamma(\alpha_{j,\ell})^{n_j}\left(\sum_{i:\delta_i = j}y_{i,\ell}
-        + \sigma\right)^{n_j\alpha_{j,\ell} + \zeta}}\right]\\
-        &\hspace{1cm}\times
-        \lvert \Sigma\rvert^{-\frac{1}{2}}\prod_{\ell = 1}^d\alpha_{j,\ell}^{-1}\exp\left\lbrace
-          -\frac{1}{2}(\log\alpha_j - \mu)^T\Sigma^{-1}(\log\alpha_j - \mu)\right\rbrace.
-    \end{aligned}
-  \end{equation}
-  Updating $\beta_j$ is accomplished with a conjugate Gibbs step, using the full conditional detailed
-  in Equation~\ref{eqn:betafc}, again subsetting $i$ such that $\delta_i = j$.  Note that we offer this
-  model for completeness---in practice, attempting to learn $\beta$ with $\alpha$ having a log-normal
-  prior resulted in extremely poor model performance.  We thus used the restricted gamma likelihood,
-  assuming $\beta_{j,\ell} = 1$ for all $j,\ell$.  This simplifies Equation~\ref{eqn:dplnlp} into
-  \begin{equation}
-    \label{eqn:dplnlpr}
-    \begin{aligned}
-    f(\bm{\alpha}_j\mid\bm{\delta},\bm{y},\mu,\Sigma) &\propto
-    \prod_{\ell = 1}^d\left[\frac{\left(\prod_{i:\gamma_i = j}
-        y_{i,\ell}\right)^{\alpha_{j,\ell} - 1}}{\alpha_{j,\ell}\Gamma(\alpha_{j,\ell})^{n_j}} \right]
-        \times\lvert\Sigma\rvert^{-\frac{1}{2}}
-      \exp\left\lbrace-\frac{1}{2}(\log\alpha_j - \mu)^T\Sigma^{-1}(\log\alpha_j - \mu)\right\rbrace.
-    \end{aligned}
-  \end{equation}
-  This model is conjugate for $\mu,\Sigma$, allowing simple Gibbs step updating.  The full
-    conditional distributions for $\mu$, $\Sigma$ take the form
-  \begin{equation}
-    \begin{aligned}
-    \mu\mid\Sigma,\alpha &\sim \mathcal{N}\left((n^*\Sigma^{-1} + S^{-1})^{-1}(n^*\overline{\log\alpha}^T\Sigma^{-1}
-      + \mu_0^TS^{-1}),(n^{*}\Sigma^{-1} + S^{-1})^{-1}\right)\\
-    \Sigma\mid\mu,\alpha &\sim \mathcal{IW}\left(
-        n^* + \nu, {\textstyle\sum}_{j = 1}^{n^*}(\log\alpha_j - \mu)(\log\alpha_j - \mu)^T + \Psi\right),
-    \end{aligned}
-  \end{equation}
-  where $n^*$ specifies the number of extant clusters.  Removing $\beta$ additionally removes the
-  inference steps for $\zeta,\sigma$.
+% \subsubsection{A Log-Normal Prior for Shape Parameters}
+% A final extension we can make to this model is to allow for some structure, or dependence, in the prior
+%   for shape parameters within a cluster.  The impetus for this approach is from recognizing that the
+%   algorithm we are using for sampling the DP may not adequately represent the posterior density for
+%   a new cluster, as it is assuming sampling from an independent prior.  If we account for dependence
+%   between dimensions within the prior, then we may be able to more efficiently sample new clusters.
+%   As shape parameters must be greater than 0, it is logical to assume a log normal prior, as that
+%   allows us the familiar forms of a mean vector and covariance matrix for representing the source of
+%   cluster parameters. So,
+%   \begin{equation}
+%     \label{eqn:dpln}
+%     \begin{aligned}
+%       \bm{ y}_i &\sim \sum_{j = 1}^J\pi_j\text{PG}\left(\bm{ y}\mid \bm{ \alpha}_i, \bm{\beta}_i\right)\\
+%       (\bm{\alpha}_i, \bm{\beta}_i) &\sim \text{DP}\left(\eta, G_0\right)\\
+%         &~\hspace{-1cm}G_0 = \mathcal{N}\left(\log\bm{ \alpha}_{j}\mid\mu,\Sigma\right)\times
+%             {\textstyle\prod}_{\ell = 2}^d\text{Ga}\left(\beta_{j,\ell}\mid\zeta_{\ell},\sigma_{\ell}\right)
+%     \end{aligned}
+%     \hspace{1cm}
+%     \begin{aligned}
+%       \mu,\Sigma &\sim \mathcal{N}\left(\mu\mid\mu_0,\Sigma_0\right) \times \text{IG}\left(\nu,\Psi\right)\\
+%       \bm{ \zeta},\bm{\sigma} &\sim {\textstyle\prod}_{\ell = 2}^d\text{Ga}(\zeta_{\ell} \mid a,b) \times \text{Ga}(\sigma_{\ell} \mid c,d) \\
+%       \bm{ \eta} &\sim \text{Ga}(\eta \mid 2, 0.1).
+%     \end{aligned}
+%   \end{equation}
+%   We hope that this will allow us to better incorporate the relationship between dimensions in
+%   sampling new clusters, allowing for better mixing of the model.  The downside is one of computational
+%   complexity---this introduces a $d$-dimensional normal distribution into the model, requiring a
+%   $d$-dimensional matrix inversion step.  As matrix inversion is computationally costly, this limits
+%   the potential number of dimensions possible.  The posterior inference strategies for $\beta,\zeta,\sigma$
+%   are the same as for prior models and have been described.  For inference on $\alpha_j$ given
+%   $r, \mu,\Sigma$, we rely on a $d$-dimensional Metropolis Hastings step, where the log-posterior
+%   density is proportional to
+%   \begin{equation}
+%     \label{eqn:dplnlp}
+%     \begin{aligned}
+%     f(\bm{\alpha}_j\mid\bm{\delta},\bm{y},\mu,\Sigma,\zeta,\sigma) &\propto
+%     \frac{\left(\prod_{i:\delta_i = j}y_{i,\ell}\right)^{\alpha_{j,\ell} - 1}}{\Gamma(\alpha_{j,\ell})^{n_j}}
+%         \prod_{\ell = 2}^d\left[\frac{\left(\prod_{i:\delta_i = j}y_{i,\ell}\right)^{\alpha_{j,\ell} - 1}
+%         \Gamma(n_j\alpha_{j,\ell} + \zeta)}{\Gamma(\alpha_{j,\ell})^{n_j}\left(\sum_{i:\delta_i = j}y_{i,\ell}
+%         + \sigma\right)^{n_j\alpha_{j,\ell} + \zeta}}\right]\\
+%         &\hspace{1cm}\times
+%         \lvert \Sigma\rvert^{-\frac{1}{2}}\prod_{\ell = 1}^d\alpha_{j,\ell}^{-1}\exp\left\lbrace
+%           -\frac{1}{2}(\log\alpha_j - \mu)^T\Sigma^{-1}(\log\alpha_j - \mu)\right\rbrace.
+%     \end{aligned}
+%   \end{equation}
+%   Updating $\beta_j$ is accomplished with a conjugate Gibbs step, using the full conditional detailed
+%   in Equation~\ref{eqn:betafc}, again subsetting $i$ such that $\delta_i = j$.  Note that we offer this
+%   model for completeness---in practice, attempting to learn $\beta$ with $\alpha$ having a log-normal
+%   prior resulted in extremely poor model performance.  We thus used the restricted gamma likelihood,
+%   assuming $\beta_{j,\ell} = 1$ for all $j,\ell$.  This simplifies Equation~\ref{eqn:dplnlp} into
+%   \begin{equation}
+%     \label{eqn:dplnlpr}
+%     \begin{aligned}
+%     f(\bm{\alpha}_j\mid\bm{\delta},\bm{y},\mu,\Sigma) &\propto
+%     \prod_{\ell = 1}^d\left[\frac{\left(\prod_{i:\gamma_i = j}
+%         y_{i,\ell}\right)^{\alpha_{j,\ell} - 1}}{\alpha_{j,\ell}\Gamma(\alpha_{j,\ell})^{n_j}} \right]
+%         \times\lvert\Sigma\rvert^{-\frac{1}{2}}
+%       \exp\left\lbrace-\frac{1}{2}(\log\alpha_j - \mu)^T\Sigma^{-1}(\log\alpha_j - \mu)\right\rbrace.
+%     \end{aligned}
+%   \end{equation}
+%   This model is conjugate for $\mu,\Sigma$, allowing simple Gibbs step updating.  The full
+%     conditional distributions for $\mu$, $\Sigma$ take the form
+%   \begin{equation}
+%     \begin{aligned}
+%     \mu\mid\Sigma,\alpha &\sim \mathcal{N}\left((n^*\Sigma^{-1} + S^{-1})^{-1}(n^*\overline{\log\alpha}^T\Sigma^{-1}
+%       + \mu_0^TS^{-1}),(n^{*}\Sigma^{-1} + S^{-1})^{-1}\right)\\
+%     \Sigma\mid\mu,\alpha &\sim \mathcal{IW}\left(
+%         n^* + \nu, {\textstyle\sum}_{j = 1}^{n^*}(\log\alpha_j - \mu)(\log\alpha_j - \mu)^T + \Psi\right),
+%     \end{aligned}
+%   \end{equation}
+%   where $n^*$ specifies the number of extant clusters.  Removing $\beta$ additionally removes the
+%   inference steps for $\zeta,\sigma$.
 
-\subsubsection{Choice of Norm}
-In the extreme value case, we have defined our task on $\mathcal{S}_{\infty}^{d-1}$, the positive orthant
+\subsubsection{Choice of Norm\label{subsubsec:norm}}
+In the extreme value case, we have defined our task on ${\mathbb S}_{\infty}^{d-1}$, the positive orthant
   of the hypersphere defined under the $\mathcal{L}_{\infty}$ norm.  We previously described the
   difficulty in establishing a density directly on this geometry.  However, we can build the projected
-  gamma distribution on $\mathcal{S}_p^{d-1}$ for a large $p$.  We can then project samples from the
-  posterior predictive distribution for a model built on $\mathcal{S}_p^{d-1}$ onto
-  $\mathcal{S}_{\infty}^{d-1}$. As $p$ increases, the closer the manifold will be to
-  $\mathcal{S}_{\infty}^{d-1}$, and thus the closer the projected distribution will be to an idealized
-  projected Gamma built on $\mathcal{S}_{\infty}^{d-1}$.  In testing, we tried developing models on
+  gamma distribution on ${\mathbb S}_p^{d-1}$ for a large $p$.  We can then project samples from the
+  posterior predictive distribution for a model built on ${\mathbb S}_p^{d-1}$ onto
+  ${\mathbb S}_{\infty}^{d-1}$. As $p$ increases, the closer the manifold will be to
+  ${\mathbb S}_{\infty}^{d-1}$, and thus the closer the projected distribution will be to an idealized
+  projected Gamma built on ${\mathbb S}_{\infty}^{d-1}$.  In testing, we tried developing models on
   $S_p^{d-1}$ for varying levels of $p$, and found that higher $p$ produced higher performing models.
   Additionally, the difference in model performance became more apparent as dimensionality increased.
-  We present results here for models developed on $\mathcal{S}_{10}^{d-1}$.
+  We present results here for models developed on ${\mathbb S}_{10}^{d-1}$.
 

### 2021-10-12 results.tex
update

@@
-\section{Data illustrations}
+\section{Data illustrations\label{sec:results}}
 
-\subsection{Simulations}
+\subsection{Simulations\label{subsec:simulated}}
 We have introduced this projected gamma distribution existing on $\mathcal{S}_{\infty}^{p-1}$.  As a
@@
 
-\subsection{Integrated Vapor Transport}
+\subsection{Integrated Vapor Transport\label{subsec:ivt}}
 One metric by which we might identify and declare atmospheric rivers is the integrated water vapor

### 2021-10-27 abstract.tex
overleaf updates

{\bf We need an abstract}
Extreme analysis

Drawing on the constructive definition for the multivariate Pareto introduced in~\cite{ferreira2014} which separates the radial component of the RV from the angular, we introduce a method for inferring the distribution of the angular component.

### 2021-10-27 conclusion.tex
overleaf updates

@@
   could be of interest to researchers, actuaries, hydrologists and even farmers.  We additionally developed
-  conditional survival functions---$\text{P}\left[Z_{\ell} > z_{\ell}\mid Z_{\not\ell} > z_{\not\ell}\right]$---which
+  conditional survival functions---$\text{P}\left[Z_{\ell} > z_{\ell}\mid Z_{\neg\ell} > z_{\neg\ell}\right]$---which
   provide a more complete understanding of the dependence structure.

### 2021-10-27 dp_projgamma.tex
overleaf updates

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
% \usepackage{setspace}
% \linespread{1.8}

\begin{document}
\paragraph{DP mixture of projected Gammas}
The DP mixture of projected Gammas model uses projected Gamma as a kernel density.
\begin{equation}
    \begin{aligned}
        \bm{y}_i\mid\bm{\alpha}_i,\bm{\beta}_i &\sim \mathcal{PG}(\bm{y}_i\mid\bm{\alpha}_i,\bm{\beta}_i)\\
        \bm{\alpha}_i,\bm{\beta}_i &\sim G\\
        G &\sim \mathcal{DP}\left(\eta,G_0\right)
    \end{aligned}
\end{equation}
Let $n_j^{-i}$ be the number of observations in cluster $j$ not including observation $i$.  Let $J^{-i}$ be the number of extant observations, not including any singletons that observation $i$ are in. 
Under this model, the probability of cluster membership is proportional to
\begin{equation*}
    \text{P}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
        \eta\int\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)dG_0(\bm{\alpha}_j,\bm{\beta}_j)\hspace{0.5cm} &\text{~for~}j = J^{-i} + 1.
        \end{cases}
\end{equation*}
Of course, if $G_0$ is not a conjugate prior for the kernel distribution, then the integral is difficult.  Instead, using algorithm 8 from Neal~(2000), we draw $m$ candidate clusters, $\bm{\alpha}_j,\bm{\beta}_j$ for $j = J^{-i} + 1,\ldots, J^{-i} + m$ from $G_0$. Then, the probability of cluster membership is proportional to
\begin{equation}
    \text{P}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
        \frac{\eta}{m}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace J^{-i} + 1,\ldots, J^{-i} + m\rbrace.
        \end{cases}
\end{equation}
Of course, if a candidate cluster is selected, then $\delta_j = J + 1$, and $J = J^{-i} + 1$.

A key feature of the the projected Gamma distribution is its computational properties.  With respect to the distributional parameters $\bm{\alpha}_i,\bm{\beta}_i$, the projected Gamma density is proportional to
\begin{equation*}
    \mathcal{PG}(\bm{y}\mid\bm{\alpha}_i,\bm{\beta}_i) \propto \int\prod_{\ell = 1}^d\mathcal{G}\left(ry_{i\ell}\mid\alpha_{i\ell},\beta_{o\ell}\right)r^{d-1}dr,
\end{equation*}
which implies that, by reintroducing the latent radial component $r_i$ via sampling from its full conditional,
\begin{equation}
    r_i\mid\bm{\alpha}_i,\bm{\beta}_i \sim \mathcal{G}\left(\sum_{\ell = 1}^d\alpha_{i\ell}, \sum_{\ell = 1}^d\beta_{\ell} y_{i\ell}\right),
\end{equation}
the likelihood for $\bm{\alpha},\bm{\beta}$ becomes separable by dimension.  We stress this point: \emph{given} $r$, the distributional parameters $(\alpha_j,\beta_j)_{\ell}$ are independent in the likelihood.

The full conditional for $\bm{\alpha}_j,\bm{\beta}_j$ is then proportional to
\begin{equation}
    f(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{Y},\bm{r},\bm{\delta},\ldots) \propto \prod_{i:\gamma_i = j}\prod_{\ell = 1}^d\mathcal{G}\left(r_iy_{i\ell}\mid\alpha_{j\ell},\beta_{j\ell}\right) \times dG_0(\alpha_{\ell},\beta_{\ell})
\end{equation}
The first form of the centering distribution we explore is a product of independent Gammas.
\begin{equation}
    G_0(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{\xi},\bm{\tau},\bm{\zeta},\bm{\sigma}) = \prod_{\ell = 1}^d\mathcal{G}(\alpha_{j\ell}\mid \xi_{\ell},\tau_{\ell})\times\prod_{\ell = 2}^d\mathcal{G}(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell})
\end{equation}
This model is completed with Gamma priors on $\xi_{\ell}$, $\tau_{\ell}$, $\zeta_{\ell}$, $\sigma_{\ell}$.  Optionally, placing a Gamma prior on $\eta$ allows updating via the procedure outlined in Escobar \& West (1995).  In further analysis, we refer to this model as the \emph{projected Gamma--Gamma} (PG--G) model.  An advantage to using a Gamma centering distribution for the rate parameter is, as a conjugate prior, the rate parameters can easily be integrated out for inference on $\bm{\alpha}_j$.  After doing so, the full conditional for $\alpha_{j\ell}$ takes the form
\begin{equation}
    \pi(\alpha_{j\ell}\mid \bm{r},\bm{Y},\bm{\gamma},\xi_\ell,\tau_\ell,\zeta_\ell,\sigma_\ell) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i\ell}\right)^{\alpha_{j\ell} - 1}\alpha_{j\ell}^{\xi_\ell - 1}e^{-\tau_\ell \alpha_{j\ell}}\times\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\Gamma^{n_j}(\alpha_{j\ell})\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
\end{equation}
for $\ell = 2,\ldots,d$.  For $\ell = 1$, as $\beta_{1} := 1$, the full conditional takes the form
\begin{equation}
    \label{eqn:alpha1update}
    \pi(\alpha_{j1}\mid\bm{r},\bm{Y},\bm{\gamma},\xi_1,\tau_1) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}\alpha_{j1}^{\zeta_1 - 1}e^{-\tau_1\alpha_{j1}}}{\Gamma^{n_j}(\alpha_{j1})}.
\end{equation}
Updating these parameters can be done via a simple Metropolis-within-Gibbs step.  In our analysis, we first transform $\alpha_{j\ell}$ to log scale, and use a normal proposal density.
The full conditional for $\beta$ is then
\begin{equation}
    \beta_{j\ell}\mid\bm{r},\bm{Y},\alpha,\zeta_{\ell},\sigma_{\ell} \sim \mathcal{G}\left(\beta_{j\ell}\mid n_j\alpha_{j\ell} + \zeta_\ell, \sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right),
\end{equation}
for $\ell = 2,\ldots, d$.  Updating $\beta_{j\ell}$ is done via a Gibbs step.  The hyper-parameters $\xi_{\ell},\tau_{\ell},\zeta_{\ell},\sigma_{\ell}$ follow similar Gamma-Gamma update relationships.  We also explore a restricted form of this model, where $\beta_{\ell} = 1$ for all $\ell$.  Under this model, we use the full conditional in Equation~\ref{eqn:alpha1update} for all $\ell$, and omit inference on $\bm{\zeta},\bm{\sigma}$.  We refer to this model as the \emph{projected restricted Gamma--Gamma} (PRG--G) model.

The second form of centering distribution we explore is a multivariate log-normal distribution on the shape parameters $\bm{\alpha}_j$, with independent Gamma $\beta_{j\ell}$ rate parameters.  
\begin{equation}
    G_0\left(\bm{\alpha}_j,\bm{\beta}_j\mid\bm{\mu},\bm{\Sigma},\zeta,\sigma\right) = \mathcal{LN}\left(\bm{\alpha}_j\mid\bm{\mu},\bm{\Sigma}\right)\times\prod_{\ell = 2}^d\mathcal{G}\left(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}\right)
\end{equation}
This model is completed with a normal prior on $\bm{\mu}$, an inverse Wishart prior on $\bm{\Sigma}$, and Gamma priors on $\zeta_{\ell}$ and $\sigma_{\ell}$, and $\eta$.  This model is denoted as the \emph{projected Gamma--log-normal} (PG--LN) model.  We also explore a restricted Gamma form of this model as above, where $\beta_{\ell} = 1$ for all $\ell$.  This is denoted as the \emph{projected restricted Gamma--log-normal} (PRG--LN) model.  in the interest of brevity, we omit the inferential forms for this model, sufficing to say that updating $\bm{\alpha}_j$ is accomplished with a joint Metropolis-within-Gibbs step; the normal prior for $\mu$ is conjugate for the log-normal $\log\bm{\alpha}_j$, and updated via a Gibbs step; and an inverse Wishart prior for $\Sigma$ is again conjugate to the log-normal $\bm{\alpha}_j$, meaning it too can be updated via a Gibbs step.

\end{document}

% EOF

### 2021-10-27 evaluation.tex
overleaf updates

--- 2021-09-15 evaluation.tex
@@
-\section{Model Comparison on the Hypersphere}
-\label{sec:evaluation}
-It is not immediately obvious what criteria to use to judge these models and decide which best
-  represents the data's generating distribution.  As the task resides on $\mathcal{S}_{\infty}^{d-1}$,
-  our options are somewhat limited in terms of what model criteria we can use.  Because of the aforementioned
-  difficulty in evaluating a density directly in this space, we choose to use model evaluation
-  criteria that can operate on samples from the posterior predictive distribution.  For this reason,
-  we have opted to use the \emph{posterior predictive loss} criterion of \cite{gelfand1998} and the
-  \emph{energy score} criterion of \cite{gneiting2007}.  Both of these metrics require calculating
-  some distance in the target space, and this section will be devoted to that end.  As both these
-  metrics operate on the posterior predictive distribution for each observation, we will also be
-  including a metric operating on the overall posterior predictive distribution, in the form of a
-  modified Kullbeck-Liebler divergence.
+\section{Scoring criteria for distributions in ${\mathbb S}_\infty^{d-1}$\label{sec:evaluation}}
 
-\subsection{Posterior Predictive Loss Criterion}
-The posterior predictive loss criterion, \emph{PPL} is introduced in \cite{gelfand1998}.  When we
+In order to assess and compare the estimation of a distribution on ${\mathbb S}_\infty^{d-1}$ we consider the theory of proper scoring rules developed in \cite{gneiting2007}.
+As mentioned in Section~\ref{subsubsec:norm}, our approach does not provide a density on ${\mathbb S}_{\infty}^{d-1}$, restricting our ability to construct model selection criteria to sample-based approaches.
+In this manner, we consider two methods of scoring:  \emph{posterior predictive loss} (PPL) introduced in \cite{gelfand1998}, and \emph{energy score} (ES) introduced in \cite{gneiting2007}.
+
+\subsection{Posterior predictive loss}
+The posterior predictive loss criterion is introduced in \cite{gelfand1998}.  When we
   assume a squared error loss function, then for the $i$th observation, the posterior predictive
@@
     \label{eq:ppl}
-    D_{k}^{(i)} = \text{Var}\left(X_i\right) + \frac{k}{k + 1}\left(\text{E}[X_i] - x_i\right)^2,
+      S_{\text{PPL}}^k\left(P,x_i\right) = \text{Var}_P\left(X_i\right) + \frac{k}{k+1}\left(\text{E}_P\left[X_{i}\right] - x_i\right)^2
   \end{equation}
   where $X_i$ is a random variable from the posterior predictive distribution for $x_i$.  The
-  scalar $k$ is a weighting factor by which we arbitrarily scale the importance of goodness of fit
+  scalar $k$ is a weighting factor by which we can scale the importance of goodness of fit
   relative to precision.  In our analysis, we take the limit as $k\to\infty$, weighting both
-  parts equally.  Interpreting this criterion, a smaller $\text{Var}(\bm{X}_{\ell})$ indicates a higher
-  precision, and a smaller $(\text{E}[\bm{X}_{\ell}] - \bm{x}_{\ell})^2$  indicates a better fit.  Thus,
+  equally.  Interpreting this criterion, a smaller $\text{Var}(X_{i})$ indicates a higher
+  precision, and a smaller $(\text{E}[X_{i}] - x_{i})^2$  indicates a better fit.  Thus,
   smaller is better.  Note that this is defined for a univariate distribution.  As we are dealing
@@
     \label{eq:ppl2}
-    D_k^{(i)} = \frac{1}{d}\sum_{\ell = 1}^{d}\left[\text{Var}(X_{i,\ell}) +
-                  \frac{k}{k+1}\left(\text{E} [X_{i,\ell}] - x_{i,\ell}\right)^2\right]
+      S_{\text{PPL}}\left(P,\bm{x}_i\right) = \frac{1}{d}\sum_{\ell = 1}^d\left[\text{Var}_P\left(X_{i\ell}\right) + \left(\text{E}_P\left[X_{i\ell}\right] - x_{i\ell}\right)^2\right]
   \end{equation}
-  Then we report the average $D_k$, taken over all $i$.
+  Then we report the average $S$, taken over all $i$.
 
-\subsection{Energy Score}
-The energy score of Gneiting, et al\cite{gneiting2007}, is a generalization of the continuous ranked
-  probability score, or \emph{crps}, defined for a multi-dimensional random variable.
+\subsection{Energy score}
+We focus on \emph{energy scores} defined for a general probability distribution $P$, with finite expectation, as 
     \begin{equation}
     \label{eq:es}
-    \text{ES}\left(P, \bm{x}_i\right) =  \text{E}_p g\left(\bm{X}_i, \bm{x}_i\right) -
-                                    \frac{1}{2}\text{E}_p g\left(\bm{X}_i,\bm{X}_i^{\prime}\right)
+    \text{S}_{\text{ES}}\left(P, \bm{x}_i\right) =  \text{E}_p g\left(\bm{X}_i, \bm{x}_i\right) -
+                                    \frac{1}{2}\text{E}_p g\left(\bm{X}_i,\bm{X}_i^{\prime}\right),
     \end{equation}
-  where $g(\cdot)$ identifies a negative definite kernel, and $\bm{X}_i^{\prime}$ represents another
-  replicate from the posterior predictive distribution of $\bm{x}_i$. A function $g$ is a negative definite kernel if it is symmetric in its arguments, $g(x_1,x_2) = g(x_2,x_1)$, and for which
-  $\sum_{i =1}^n\sum_{j=1}^n\alpha_i\alpha_jg(\bm{x}_1,\bm{x}_2) \leq 0$ for all positive integers n, with the
-  restriction that $\sum_{i=1}^n\alpha_i = 0$.  Euclidean distance is one example of a negative definite kernel,
-  and the one most often used.  However, on $\mathcal{S}_p^{d-1}$ for $p > 1$, Euclidean distance
-  can under-report the actual distance required for travel between two points. That distortion is
-  maximized on $\mathcal{S}_{\infty}^{d-1}$.
+where $g$ is a given kernel. The score defined in Equation \eqref{eq:es} can be evaluated using samples from $P$, with the help of the law of large numbers.
+Moreover, Theorem~4 in \cite{gneiting2007}, states that if $g(\cdot,\cdot)$ is a negative definite kernel, then $\text{S}(P,\bm{x})$ is a \emph{proper} scoring rule.
+Recall that a real valued function $g$ is a negative definite kernel if it is symmetric in its arguments, and $\sum_{i=1}^n\sum_{j=1}^na_ia_jg(x_i,x_j)\leq 0$ for all positive integers $n$, and any collection $a_1,\ldots,a_n\in{\mathbb R}$ such that  $\sum_{i = 1}^na_i = 0$.
 
-\subsection{Distance on the positive orthant of the Hypercube}
-  \label{subsec:distance}
-The positive orthant of the unit hypercube, defined in Euclidean geometry, is that structure for
-  which, in a given point on the hypercube, all dimensions of that point are between 0 and 1, and
-  at least one dimension must be 1.  Developing terminology, we can consider observations for which
-  the $j$th dimension is equal to 1, to be on the $j$th \emph{face}.  The intersection of the $i$th
-  and $j$th face is a $d-2$ dimensional cube, and observations in this space have dimensions $i,j$
-  equal to 1.
+In an Euclidean space, the conditions for a negative definite kernel are satisfied by the Euclidean distance.
+However, the Euclidean distance between two points will likely under-estimate the actual distance required to connect those points fully within ${\mathbb S}_{\infty}^{d-1}$.
+Let
+\begin{equation*}
+    {\mathbb C}_{\ell}^{d-1} = \lbrace \bm{x} : \bm{x} \in {\mathbb S}_{\infty}^{d-1}, x_{\ell} = 1\rbrace
+\end{equation*}
+comprise the $\ell$th \emph{face}.
+For points on the same face, Euclidean distance corresponds to the length of the shortest possible path in ${\mathbb S}_{\infty}^{d-1}$.
+For points on different faces, the Euclidean distance is a lower bound for that length.
 
-A \emph{distance} in this space is a \emph{geodesic} on this space. From geometry, we know that
-  the geodesic, or shortest path between 2 points along the surface of a $d$ dimensional figure
-  corresponds to at least one \emph{unfolding}, or \emph{rotation} of the $d$-dimensional
-  figure into a $d-1$ dimensional space.  The appropriate term for the structure generated by this
-  unfolding is a \emph{net}.  For the appropriate net, a line segment connecting the two points and
-  staying within the boundaries of the net corresponds to the shortest path between those points\citep{pappas1989}, and is thus a geodesic.  The length of that line segment is properly
-  defines the distance required for travel between those points.
+For a finite $p$, the shortest connecting path between two points in 
+${\mathbb S}_p^{d-1}$ is the minimum geodesic; its length satisfying 
+the definition of a distance.
+Thus its length can be used as a negative definite kernel for the purpose of defining an energy score.
+Unfortunately as $p\to\infty$ the resulting surface ${\mathbb S}_{\infty}^{d-1}$ is not differentiable, implying that routines to calculate geodesics are not readily available.
+However, as ${\mathbb S}_{\infty}^{d-1}$ is a portion of of a $d$-cube, we can borrow a result from geometry \citep{pappas1989} stating that the length of the shortest path between two points on a geometric figure corresponds to the length of a straight line drawn between the points on an appropriate unfolding, rotation, or \emph{net} of the figure from a $d$-dimensional to a $d-1$-dimensional space.
+The optimal net will have the shortest straight line between the points, as long as that line is fully contained within such net.
+As ${\mathbb S}_{\infty}^{d-1}$ has $d$ faces---each face pairwise adjacent, there are $d!$ possible nets.
+However, we are only interested in nets that begin and end on the source and destination faces respectively.
+That reduces the number of nets under consideration to $\sum_{k = 0}^{d-2}\binom{d-2}{k}$.
+This is still computationally burdensome for a large number of dimensions.
 
-Consider a 3-dimensional cube.  Consider 2 points on this 3-dimensional cube,
-  $\bm{a}_1 = (x_1,y_1,z_1)$, and $\bm{a}_2 = (x_2,y_2,z_2)$. Let's say that the two points are
-  on the same face.  Then the distance between those two points, the distance one has to travel
-  along the space to move from one point to the other, is calculated by Euclidean norm.  Now,
-  consider two points on separate faces.  All faces are pairwise adjacent, as we have stated, so
-  in order to move to the other point, we must \emph{at least} move to the intersection between the
-  faces, then to the other point.  Let $\bm{a}_1$ lie on the $x$ face, and $\bm{a}_2$ lie on the
-  $y$ face.  That is, $\bm{a}_1 = (1, y_1, z_1)$, $\bm{a}_2 = (x_2, 1, z_2)$.  Then traveling
-  between these points we must at least pass through the intersection of faces $x,y$.  One possible
-  net representation of this is unfolding the $y$ face alongside the $x$ face.  We accomplish this
-  by applying a rotation and translation to $a_2$, corresponding to the following:
-  \begin{equation}
-    \label{eq:1drotation}
-    a_2^{\prime} = \begin{bmatrix}
-    1 \\
-    2 \\
-    0
-    \end{bmatrix}
-    +
-    \begin{bmatrix}
-    0  & 0 & 0 \\
-    -1 & 0 & 0 \\
-    0  & 0 & 1
-    \end{bmatrix}
-    \begin{bmatrix}
-    x_2 \\
-    1 \\
-    z_2 \\
-    \end{bmatrix} = \begin{bmatrix}
-    1 \\
-    2 - x_2 \\
-    z_2
-    \end{bmatrix}
-  \end{equation}
-  Then, if this is the appropriate net, the distance from $\bm{a}$ to $\bm{b}$ becomes
-  $\lVert \bm{a}_1 - \bm{a}_2^{\prime}\rVert_2$. However, there
-  is another possible net we must consider, travelling first through the $z$ face then to the
-  $y$ face.  The rotation for that becomes:
-  \begin{equation}
-    \label{eq:2drotation}
-    a_2^{\prime} = \begin{bmatrix}
-    1 \\
-    2 \\
-    2
-    \end{bmatrix}
-    +
-    \begin{bmatrix}
-    0 & 0 & 0  \\
-    0 & 0 & -1 \\
-    -1 & 0 & 0
-    \end{bmatrix}
-    \begin{bmatrix}
-    x_2 \\
-    1 \\
-    z_2
-    \end{bmatrix} = \begin{bmatrix}
-    1 \\
-    2 - z_2 \\
-    2 - x_2
-    \end{bmatrix}
-  \end{equation}
-Every successive rotation is relative to the last face.  So, as the number of dimensions grows, the
-  number of possible rotations grows as well.  As we have $d$ faces, if 2 observations are on
-  different faces, then there are $\sum_{j = 1}^{d-2}\binom{d-2}{j} + 1$ possible rotations to
-  consider\footnote{There are truly $d!$ possible nets, but when we consider starting and ending
-  faces fixed, and that portions of the net that diverge after the ending face are irrelevant,
-  we arrive at that number of rotations that we need consider}.  While this is not
-  insurmountable, it is computationally costly. However, for a valid energy score, a true distance
-  is not necessary.  What is needed is a negative definite kernel.
-
-\begin{prop}
-For points $a,b \in \mathcal{S}_{\infty}^{d-1}$ a valid negative definite kernel can be formed as
+To reduce the computations needed to calculate the energy score we define a kernel as {\bf Please write a clearer description here.} our kernel an upper bound on distance between points on separate faces as the length of the path from the source point, to an optimal point on the intersection between faces, to the destination point.
+\begin{prop}\label{prop:g}
+For points $a,b \in {\mathbb S}_{\infty}^{d-1}$ the kernel defined as  
   \begin{equation}
@@
         \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} &\text{ otherwise}
-    \end{cases}
+    \end{cases} ,
   \end{equation}
-  where $\bm{c}$ resides on the intersection between the faces of $\bm{a}$ and $\bm{b}$, and minimizes $g(\bm{a},\bm{b})$.
+  where $\bm{c}$ is in the intersection of the faces corresponding to $\bm{a}$ and $\bm{b}$, and minimizes $g(\bm{a},\bm{b})$, is negative definite.
 \end{prop}
 
-In the first case, where $\bm{a}$ and $\bm{b}$ reside on the same face, then that is trivially a
-  negative definite kernel.  In the second case, where $\bm{a}$ and $\bm{b}$ reside on separate
-  faces, we can be assured of symmetry in the arguments as $\bm{c}$ is unique, and $\pnorm{\bm{c} - \bm{a}}{2} = \pnorm{\bm{a} - \bm{c}}{2}$.  Then for the other property of negative definiteness,
+{\em Proof:}
+When $\bm{a}$ and $\bm{b}$ are on the same face, then $g$ is the Euclidean distance, which is a 
+  negative definite kernel.  When $\bm{a}$ and $\bm{b}$ are on separate
+  faces, we have that $g(\bm{a}, \bm{b}) = g(\bm{b}, \bm{a})$ due to the uniqueness of $\bm{c}$ and the symmetry of the Euclidean distance.
+Additionally, given $n$ and  collection $\alpha_1,\ldots,\alpha_n\in{\mathbb R}$ such that  $\sum_{i = 1}^n\alpha_i = 0$, then
   \begin{equation*}
@@
       \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j g(\bm{x}_1,\bm{x}_2) &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j \bigg[\pnorm{\bm{c} - \bm{x}_1}{2} + \pnorm{\bm{x}_2 - \bm{c}}{2}\bigg]\\
-      &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{c} - \bm{x}_1}{2} + \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{x}_2 - \bm{c}}{2}
+      &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{c} - \bm{x}_1}{2} + \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{x}_2 - \bm{c}}{2} \leq0,
     \end{aligned}
   \end{equation*}
-  which is less than $0$ in both terms by definition.
-
-To compute this energy score as described would require optimization of the interim point $\bm{c}$.
-  If we again consider the one-dimensional rotation in Equation~\ref{eq:1drotation}---we
-  can view that as a numerically easier analogue accomplishing the same goal: the distance from
-  the starting point, to some optimal point along the intersection between the starting and
-  ending faces, to the ending point.  The strait line will pass through the optimal point, so
-  the optimization of the interim point is accomplished \emph{en passant}.
+  as the Euclidean norm is a negative definite kernel.$\hfill\Box$
+Evaluating $g$ as defined in Proposition \ref{prop:g} requires the solution of a $d-2$-dimensional 
+  optimization problem in order to obtain $\bm{c}$.  However, a computationally easier \makenote{rewrite sentence} method exists.
+\begin{prop}
+    Let $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-1}$.  The 
+  rotation $P_{\jmath\ell}(\cdot)$ required to move the $\jmath$th face along the $\ell$th axis 
+  produces the vector $\bm{b}^\prime$, where
+  \begin{equation}
+    \label{eqn:rotation}
+    P_{\jmath\ell}(\bm{b}) = \bm{b}^{\prime}_i = 
+    \begin{cases}
+        b_{i} &\text{for }i\neq \jmath,\ell\\
+        1 &\text{for }i = \ell\\
+        2 - b_{\ell} &\text{for }i = \jmath
+    \end{cases}.
+  \end{equation}
+  Then $g(\bm{a},\bm{b}) = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}$.
+\end{prop}
+{\em Proof:}
+Note that for 
+  $\bm{c} \in {\mathbb C}_{\jmath\ell}^{d-2} = {\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}$,
+  $P_{\jmath\ell}(\bm{c}) = \bm{c}$.  Then
+  \begin{equation*}
+     \pnorm{\bm{b} - \bm{c}}{2} = \pnorm{P_{\jmath\ell}(\bm{b} - \bm{c})}{2} = \pnorm{\bm{b}^{\prime} - \bm{c}}{2},
+  \end{equation*}
+  as rotations preserve Euclidean norms.  Then,
+  \begin{equation*}
+    \begin{aligned}
+    g(\bm{a}, \bm{b}) &= \pnorm{\bm{a} - \bm{c}}{2} + \pnorm{\bm{b} - \bm{c}}{2}\\
+    &= \pnorm{\bm{a} - \bm{c}}{2} + \pnorm{\bm{b}^{\prime} - \bm{c}}{2} = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}
+    \end{aligned}
+  \end{equation*}
+  when $\bm{c} \in \overrightarrow{\bm{a}\bm{b}^{\prime}}$, as the lower bound of the triangle inequality.
+We have already declared $\bm{c}\in{\mathbb C}_{\jmath\ell}^{d-2}$, so it must reside at $\overrightarrow{\bm{a}\bm{b}^{\prime}}\cap{\mathbb C}_{\jmath\ell}^{d-2}$.
+Because the straight line between $\bm{a}$ and $\bm{b}^{\prime}$ passes through this point without further analysis, the optimization of $\bm{c}$ is accomplished \emph{en passant}.$\hfill\Box$
 

### 2021-10-27 introduction.tex
overleaf updates

@@
 rare events that correspond to the tails of probability distributions.
-As such, it is a key ingredient in the assessment of the risk of
+As such, it is a key ingredient in the risk assessment of
 phenomena that can have strong societal impacts like floods, heat waves,
@@
 are well established and software is widely available \cite[see, for example,][]{coles2001}.
-For variables in one dimension the application of EVT methods consider the asymptotic distribution of either the maxima calculated for regular blocks of data, or the values that exceed a certain threshold.
+For variables in one dimension the application of EVT methods considers the asymptotic distribution of either the maxima calculated for regular blocks of data, or the values that exceed a certain threshold.
 The former leads to a Generalized Extreme Value (GEV) distribution, that depends on three parameters.
@@
 Section~\ref{sec:multivariatepot} comprises a brief review of multivariate PoT, detailing the separation of the radial measure from the angular measure.
-  Section~\ref{sec:methodology} details our routine for estimating the angular measure, where~\ref{subsec:projection} elaborates the process of projecting an arbitrary distribution supported in ${\mathbb R}_+^d$ onto an a series of manifolds ${\mathbb S}_p^{d-1}$, $p = 1,2,\ldots$ approaching the support of the angular measure, on to which~\ref{subsec:projgamma} builds to develop a \emph{projected Gamma} family of distributions.
-Section~\ref{sec:evaluation} is concerned with developing criteria for model selection in the support of the angular measure.
-Section~\ref{sec:results} details the efficacy of the projected gamma model in recovering the distribution of data on the support of the angular measure---in~\ref{subsec:simulated}, this is done on simulated data, and in~\ref{subsec:ivt}, acknowledging the relevance of of extreme value theory to climatological events~\citep{jentsch2007,vousdoukas2018,li2019}, we estimate the extremal dependence structure for the integrated vapor transport---a measure of water vapor in the atmosphere, used for identifying atmospheric rivers.
-Inferences from the dependence structure detailed in~\ref{subsec:ped} and~\ref{subsec:condsurv} are applied.
-Finally, Section~\ref{sec:conclusion} concludes.
+  Section~\ref{sec:methodology} details our approach for estimating the angular measure, based on projecting an arbitrary distribution supported in ${\mathbb R}_+^d$ onto unit hyper-spheres defined for $p$-norms.
+Section~\ref{sec:evaluation} develops criteria for model selection in the support of the angular measure.
+Section~\ref{sec:results} explores the efficacy of the proposed approach on a set simulated data, and, acknowledging the relevance of of extreme value theory to climatological events~\citep{jentsch2007,vousdoukas2018,li2019}, estimates the extremal dependence structure for a measure of water vapor flow in the atmosphere, used for identifying atmospheric rivers.
+Finally, Section~\ref{sec:conclusion} presents our conclusions and discussion.
 
@@
 Define  $R = \max_{i} Z_i$, and assume that $R>1$ with probability one. Also, define $V_i = Z_i/\max_i Z_i$, and assume that $E(V_i)>0$.
-These two assumption together with the homogeneity of $\mu$ correspond to the second condition in Theorem 2.1 of
+These two assumptions together with the homogeneity of $\mu$ correspond to the second condition in Theorem 2.1 of
 \cite{ferreira2014}, applied to the finite-dimensional case.
@@
 
-\subsection{Pairwise Extremal Dependence\label{subsec:ped}}
+%\subsection{Pairwise Extremal Dependence\label{subsec:ped}}
 A measure that is used to characterize the strength of the dependence,
@@
 
-\subsection{Multivariate Conditional Survival Function\label{subsec:condsurv}}
+%\subsection{Multivariate Conditional Survival Function\label{subsec:condsurv}}
 For the analysis of extreme values it is of interest to calculate the multivariate conditional survival function.

### 2021-10-27 methodology_projgamma.tex
overleaf updates

--- 2021-09-18 methodology_projgamma.tex
@@
   \end{equation}
-  defined for $\bm{y}\in \mathcal{S}_p^{d-1}$, and for any $p>0$. To avoid idetifiability problems when estimating the shape and scale parameters, we set $\beta_1 = 1$.
+  defined for $\bm{y}\in {\mathbb S}_p^{d-1}$, and for any $p>0$. To avoid identifiability problems when estimating the shape and scale parameters, we set $\beta_1 = 1$.
   \cite{nunez2019} obtain the density in Equation (\ref{eqn:projgamma})
   for $p=2$ as a multivariate distribution for directional data, using spherical coordinates.
-  For $\bm{ y}\in \mathcal{S}_1^{d-1}$, and $\beta_{\ell} = \beta, \forall\ell$, the density in Equation (\ref{eqn:projgamma}) corresponds to that of a Dirichlet distribution.
+  For $\bm{ y}\in {\mathbb S}_1^{d-1}$, and $\beta_{\ell} = \beta$ for all $\ell$, the density in Equation (\ref{eqn:projgamma}) corresponds to that of a Dirichlet distribution.
 
@@
 Thus, we use it as a building block for the angular measure $\Phi$ models.
-To build a flexible family of distributions in $\mathcal{S}_p^{d-1}$ we consider mixtures of projected gamma densities defined as
-  \begin{equation} \label{PGmix}     
-     f(\bm{y}) = \int_\Theta \text{PG}(\bm{y}|\bm{\theta}) dG(\bm{\theta}), 
+To build a flexible family of distributions in ${\mathbb S}_p^{d-1}$ we consider mixtures of projected gamma densities defined as
+  \begin{equation} \label{eqn:PGmix}     
+     f(\bm{y}) = \int_\Theta \mathcal{PG}(\bm{y}|\bm{\theta}) dG(\bm{\theta}), 
   \end{equation} 
@@
 Following a Bayesian non-parametric approach \citep{Ferguson74,Antoniak1974} we assume that $G$ is drawn from a random measure.
-In particular, assuming a Dirichlet process prior, we have a hierarchical formulation of the mixture model given by
-  \[    \bm{y} \sim \text{PG}(\bm{y}|\bm{\theta}) , \;\;\; \bm{\theta} \sim G, \;\;\; G \sim \text{DP}(\eta, G_0),\]
-  where DP denotes a Dirichlet process, $\eta$ is the precision parameter, and $G_0$ is the centering distribution.
+In particular, assuming a Dirichlet process prior, we have a hierarchical formulation of the mixture model that, for a vector of observations $\bm{y}_i$, is given by
+  \begin{equation}\label{eqn:dppg}    \bm{y}_i \sim \text{PG}(\bm{y}_i|\bm{\theta}_i) , \;\;\; \bm{\theta}_i \sim G, \;\;\; G \sim \mathcal{DP}(\eta, G_0),
+  \end{equation}
+  where $\mathcal{DP}$ denotes a Dirichlet process, $\eta$ is the precision parameter, and $G_0$ is the centering distribution.
 
-  Our strategy consists of describing the angular distribution $\Phi$ using a sample based approach with the following steps: (i) Apply the transformation in Equation (\ref{eqn:standardization}) to the original data; (ii) Obtain the subsample of the standardized observations that satisfy $R>1$; (iii) Take a finite $p$ and project the observations onto $\mathcal{S}_p^{d-1}$; (iv) Fit the model in Equation (\ref{PGmix}) to the resulting data and obtain samples from the fitted model; (v) Project the resulting samples onto $\mathcal{S}_\infty^{d-1}$.
+  Our strategy consists of describing the angular distribution $\Phi$ using a sample based approach with the following steps: (i) Apply the transformation in Equation (\ref{eqn:standardization}) to the original data; (ii) Obtain the subsample of the standardized observations that satisfy $R>1$; (iii) Take a finite $p$ and project the observations onto ${\mathbb S}_p^{d-1}$; (iv) Fit the model in Equation (\ref{eqn:PGmix}) to the resulting data and obtain samples from the fitted model; (v) Project the resulting samples onto ${\mathbb S}_\infty^{d-1}$.
   For step (iv) we use a Bayesian approach that is implemented using a purposely developed Markov chain Monte Carlo that is described in the next section. 
-  {\bf Instead of the olf Figure 1, it would be nice to add a figure where you illustrate this procedure. Give it a try.}
-%  The full conditional for $r$ takes the form
-%  \begin{equation}
-%    \label{eqn:rfullcond}
-%    r\mid\bm{ \alpha},\bm{ \beta}, y \sim \text{Ga}\left(r\mid{\textstyle\sum}_{\ell = 1}^d
-%                    \alpha_{\ell}, {\textstyle\sum}_{\ell = 1}^d \beta_{\ell}y_{\ell}\right).
-%  \end{equation}
+  {\bf Instead of the old Figure 1, it would be nice to add a figure where you illustrate this procedure. Give it a try.}
 
-\subsubsection{Inference for the projected gamma mixture model}
-{\bf Our focus is on model (\ref{PGmix}).
-Describe how you build a sample based approach to fitting that model.
-How you introduce latent variables, both to break the mixture and to produce the PG.
-You can then discuss variations that essentially consists on different priors.
-So, basically, you need to condense the next three pages in about a page.}
+\subsubsection{Inference for the projected Gamma mixture model}
+The Dirichlet process mixture model groups observations into stochastically assigned \emph{clusters}.
+Building out the methods of inference for Equation~\ref{eqn:dppg}, let $n_j^{-i}$ be the number of observations in cluster $j$ not including observation $i$.
+Let $J^{-i}$ be the number of extant observations, not including any singleton that observation $i$ is in.
+Under this model, the probability of cluster membership is proportional to
+\begin{equation*}
+    \text{P}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
+        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
+        \eta\int\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)dG_0(\bm{\alpha}_j,\bm{\beta}_j)\hspace{0.5cm} &\text{~for~}j = J^{-i} + 1.
+        \end{cases}
+\end{equation*}
+Of course, if $G_0$ is not a conjugate prior for the kernel distribution, then the integral is difficult.
+Instead, using algorithm 8 from \cite{neal2000}, we draw $m$ candidate clusters, $\bm{\alpha}_j,\bm{\beta}_j$ for $j = J^{-i} + 1,\ldots, J^{-i} + m$ from $G_0$.
+Then, the probability of cluster membership is proportional to
+\begin{equation}
+    \text{P}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
+        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
+        \frac{\eta}{m}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace J^{-i} + 1,\ldots, J^{-i} + m\rbrace.
+        \end{cases}
+\end{equation}
+Of course, if a candidate cluster is selected, then $\delta_j = J + 1$, and $J = J^{-i} + 1$.
 
-We consider a basic model that assumes all observations are generated by a single projected Gamma
-  distribution.  The shape and rate parameters for the latent \text{Gamma} random variables,
-  $\alpha_{\ell},\beta_{\ell}$, must be greater than 0, so a logical prior to use would be a Gamma prior for
-  each parameter, excepting for $\beta_1 := 1$.
+A key feature of the the projected Gamma distribution is its computational properties.
+With respect to the distributional parameters $\bm{\alpha}_i,\bm{\beta}_i$, the projected Gamma density is proportional to
+\begin{equation*}
+    \mathcal{PG}(\bm{y}\mid\bm{\alpha}_i,\bm{\beta}_i) \propto \int\prod_{\ell = 1}^d\mathcal{G}\left(ry_{i\ell}\mid\alpha_{i\ell},\beta_{i\ell}\right)r^{d-1}dr,
+\end{equation*}
+which implies that, by reintroducing the latent radial component $r_i$ via sampling from its full conditional,
 \begin{equation}
-    \label{eqn:vanillap}
-    \begin{aligned}
-      \bm{ y}\mid\alpha,\beta &\sim \text{PG}(\bm{ y}\mid\alpha,\beta)\\
-      \bm{ \alpha},\bm{\beta} &\sim {\textstyle \prod}_{\ell = 1}^d \text{Ga}(\alpha_{\ell} \mid \xi_{\ell},\tau_{\ell})
-              \times {\textstyle \prod}_{\ell = 2}^d \text{Ga}(\beta_{\ell}\mid \zeta_{\ell},\sigma_{\ell}).
-    \end{aligned}
+    r_i\mid\bm{\alpha}_i,\bm{\beta}_i \sim \mathcal{G}\left(\sum_{\ell = 1}^d\alpha_{i\ell}, \sum_{\ell = 1}^d\beta_{\ell} y_{i\ell}\right),
 \end{equation}
-  A useful property of the Projected Gamma family is the latent independent Gammas assumption.  Recall
-  in Eqn.~\ref{eqn:protopg}, we have a density defined as a product of independent Gamma densities,
-  multiplied by a factor of $r^{d-1}C$, where $C$ is constant with respect to the parameters
-  $\alpha,\beta$, as well as $r$.  If we conduct inference via a Gibbs sampler, then by simulating
-  this latent $r$ we are able to perform inference of $(\alpha_{\ell},\beta_{\ell})$ independent of
-  that of $(\alpha_k,\beta_k)$.  The full conditional for $r$ is described in Eqn.~\ref{eqn:rfullcond}.
-  The full conditional for $\beta_{\ell}$, $l = 2,\ldots,d$ then becomes
+the likelihood for $\bm{\alpha},\bm{\beta}$ becomes separable by dimension.
+We stress this point: \emph{given} $r$, the distributional parameters $(\alpha_j,\beta_j)_{\ell}$ are independent in the likelihood.
+
+The full conditional for $\bm{\alpha}_j,\bm{\beta}_j$ is then proportional to
 \begin{equation}
-    \label{eqn:betafc}
-    \beta_{\ell}\mid \bm{ y}, r, \alpha_{\ell} \sim \text{Ga}\left(n + \alpha_{\ell} + \zeta_{\ell},
-                                      {\textstyle \sum}_{i = 1}^nr_iy_{il} + \sigma_{\ell}\right).
+    f(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{Y},\bm{r},\bm{\delta},\ldots) \propto \prod_{i:\gamma_i = j}\prod_{\ell = 1}^d\mathcal{G}\left(r_iy_{i\ell}\mid\alpha_{j\ell},\beta_{j\ell}\right) \times dG_0(\alpha_{\ell},\beta_{\ell})
 \end{equation}
-  That is, given $\alpha_{\ell}$, $r$, we can sample $\beta_{\ell}$ directly as a Gamma random
-  variable.  In the sampling for $\alpha_{\ell}$, we can integrate $\beta_{\ell}$ out leaving us a
-  quantity proportional to a density, which we can sample via the Metropolis Hastings
-  algorithm\citep{hastings1970}.  For $\alpha_{\ell}$, $\ell = 2,\ldots, d$, This takes the form
+The first form of the centering distribution we explore is a product of independent Gammas.
 \begin{equation}
-    \label{eqn:alphafc}
-    f(\alpha_{\ell} \mid \bm{ y}, r) \propto
-    \frac{\left({\textstyle \prod}_{i = 1}^nr_iy_{il}\right)^{\alpha_{\ell} - 1}}{\Gamma^n(\alpha_{\ell})} \times
-    \alpha_{\ell}^{\xi_{\ell} - 1}\exp\{-\tau_{\ell}\alpha_{\ell}\} \times
-    \frac{\Gamma(n\alpha_{\ell} + \zeta_{\ell})}{
-        \left({\textstyle\sum}_{i = 1}^n r_iy_{il} + \sigma_{\ell}\right)^{(n * \alpha_{\ell} + \zeta_{\ell})}}.
+    G_0(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{\xi},\bm{\tau},\bm{\zeta},\bm{\sigma}) = \prod_{\ell = 1}^d\mathcal{G}(\alpha_{j\ell}\mid \xi_{\ell},\tau_{\ell})\times\prod_{\ell = 2}^d\mathcal{G}(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell})
 \end{equation}
-  For $\alpha_1$, as we do not integrate out $\beta_1$ as it is defined to be $1$, this takes the
-  slightly simpler form
+This model is completed with Gamma priors on $\xi_{\ell}$, $\tau_{\ell}$, $\zeta_{\ell}$, $\sigma_{\ell}$.
+Optionally, placing a Gamma prior on $\eta$ allows updating via the procedure outlined in \cite{escobar1995}.
+In further analysis, we refer to this model as the \emph{projected Gamma--Gamma} (PG--G) model.
+An advantage to using a Gamma centering distribution for the rate parameter is, as a conjugate prior, the rate parameters can easily be integrated out for inference on $\bm{\alpha}_j$.
+After doing so, the full conditional for $\alpha_{j\ell}$ takes the form
 \begin{equation}
-    \label{eqn:alphafc1}
-    f(\alpha_1 \mid \bm{ y}, r) \propto
-      \frac{{\textstyle\prod}_{i = 1}^n ry_{i1}^{\alpha_1 - 1}}{\Gamma^n(\alpha_1)} \times
-      \alpha_1^{\xi_1 - 1}\exp\{-\tau_1\alpha_1\}.
+    \pi(\alpha_{j\ell}\mid \bm{r},\bm{Y},\bm{\gamma},\xi_\ell,\tau_\ell,\zeta_\ell,\sigma_\ell) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i\ell}\right)^{\alpha_{j\ell} - 1}\alpha_{j\ell}^{\xi_\ell - 1}e^{-\tau_\ell \alpha_{j\ell}}\times\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\Gamma^{n_j}(\alpha_{j\ell})\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
 \end{equation}
-  The \emph{projected restricted Gamma} model sets $\beta_{\ell} := 1 \forall \ell \in \lbrace 1,\ldots,d\rbrace$,
-  so in that model inference on all $\alpha_{\ell}$ will use the simpler form.
+for $\ell = 2,\ldots,d$.  For $\ell = 1$, as $\beta_{1} := 1$, the full conditional takes the form
+\begin{equation}
+    \label{eqn:alpha1update}
+    \pi(\alpha_{j1}\mid\bm{r},\bm{Y},\bm{\gamma},\xi_1,\tau_1) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}\alpha_{j1}^{\zeta_1 - 1}e^{-\tau_1\alpha_{j1}}}{\Gamma^{n_j}(\alpha_{j1})}.
+\end{equation}
+Updating these parameters can be done via a simple Metropolis-within-Gibbs step.  In our analysis, we first transform $\alpha_{j\ell}$ to log scale, and use a normal proposal density.
+The full conditional for $\beta$ is then
+\begin{equation}
+    \beta_{j\ell}\mid\bm{r},\bm{Y},\alpha,\zeta_{\ell},\sigma_{\ell} \sim \mathcal{G}\left(\beta_{j\ell}\mid n_j\alpha_{j\ell} + \zeta_\ell, \sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right),
+\end{equation}
+for $\ell = 2,\ldots, d$.
+Updating $\beta_{j\ell}$ is done via a Gibbs step.
+The hyper-parameters $\xi_{\ell},\tau_{\ell},\zeta_{\ell},\sigma_{\ell}$ follow similar Gamma-Gamma update relationships.
+We also explore a restricted form of this model, where $\beta_{\ell} = 1$ for all $\ell$.
+Under this model, we use the full conditional in Equation~\ref{eqn:alpha1update} for all $\ell$, and omit inference on $\bm{\zeta},\bm{\sigma}$.
+We refer to this model as the \emph{projected restricted Gamma--Gamma} (PRG--G) model.
 
-\subsubsection{Finite Mixture Model}
-The projected Gamma distribution offers a flexible model for describing direction.  However, it is
-  conceivable that a single Projected Gamma distribution will not sufficiently capture the
-  variability of real data.  In that case, a logical extension to make to the above model is to
-  assume the data, $\bm{ y}$, are generated from a mixture of projected Gammas.  We consider a finite
-  mixture,
+The second form of centering distribution we explore is a multivariate log-normal distribution on the shape parameters $\bm{\alpha}_j$, with independent Gamma $\beta_{j\ell}$ rate parameters.  
 \begin{equation}
-    \label{eqn:fmixp}
-    \begin{aligned}
-      \bm{ y}_i &\sim \sum_{j = 1}^J\pi_j\text{PG}\left(\bm{ y}\mid \bm{ \alpha}_j, \bm{ \beta}_j\right)\\
-      \bm{ \alpha}_j &\sim {\textstyle \prod}_{\ell = 1}^d \text{Ga}\left(\alpha_{\ell}\mid\xi_{\ell},\tau_{\ell}\right)\\
-      \bm{ \beta}_j &\sim {\textstyle \prod}_{\ell = 2}^d \text{Ga}\left(\beta_{\ell}\mid\zeta_{\ell},\sigma_{\ell}\right)
-    \end{aligned}
-    \hspace{2cm}
-    \begin{aligned}
-      \bm{ \xi},\bm{\tau} &\sim {\textstyle \prod}_{\ell = 1}^d \text{Ga}(\xi_{\ell}\mid a,b) \times \text{Ga}(\tau_{\ell}\mid c,d)\\
-      \bm{ \zeta},\bm{\sigma} &\sim {\textstyle\prod}_{\ell = 2}^d\text{Ga}(\zeta_{\ell} \mid a,b)\times \text{Ga}(\sigma_{\ell}\mid c,d)\\
-      \bm{ \pi} &\sim \text{Dir}(\pi_0)
-    \end{aligned}
+    G_0\left(\bm{\alpha}_j,\bm{\beta}_j\mid\bm{\mu},\bm{\Sigma},\zeta,\sigma\right) = \mathcal{LN}\left(\bm{\alpha}_j\mid\bm{\mu},\bm{\Sigma}\right)\times\prod_{\ell = 2}^d\mathcal{G}\left(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}\right)
 \end{equation}
-  In addition to allowing multiple mixture components, we assume those mixture components themselves
-  descend from a product of independent Gamma densities. The standard method for sampling from a
-  mixture distribution includes a data augmentation step, generating a
-  $\delta_i \in \lbrace 1,\ldots, J\rbrace$ indicating to which mixture component a
-  particular observation comes from, with
-  \begin{equation*}
-    \text{P}\left[\delta_i = j\mid\alpha_j,\beta_j\right] \propto \pi_j\text{PG}\left(\bm{ y}_i\mid \bm{ \alpha}_j,\bm{ \beta}_j\right).
-  \end{equation*}
-  By sampling and conditioning on $\delta_i$, we end up with very similar semi-conjugate forms for
-  inference of $\bm{ \alpha}_j$, $\bm{ \beta}_j$ that were used in the basic model.  The changes made
-  to Equations~\ref{eqn:betafc},\ref{eqn:alphafc},\ref{eqn:alphafc1} are such that we're conducting inference
-  on a mixture component given the mixture component identifiers $\bm{ \delta}$, so instead of indexing
-  $i$ over the whole data set, we're indexing $i$ over the set $\{i :\delta_i = j\}$, with $n = n_j$,
-  the cardinality of that set. Inference on $\bm{ \xi,\tau,\zeta,\sigma}$ follows a similar form as
-  that of $\bm{ \alpha,\beta}$---they are Gamma-Gamma relationships, so rate parameters have a conjugate
-  Gamma posterior, and we can integrate them out to perform inference on the shape parameters via
-  Metropolis Hastings.
+This model is completed with a normal prior on $\bm{\mu}$, an inverse Wishart prior on $\bm{\Sigma}$, and Gamma priors on $\zeta_{\ell}$ and $\sigma_{\ell}$, and $\eta$.
+This model is denoted as the \emph{projected Gamma--log-normal} (PG--LN) model.
+We also explore a restricted Gamma form of this model as above, where $\beta_{\ell} = 1$ for all $\ell$.
+This is denoted as the \emph{projected restricted Gamma--log-normal} (PRG--LN) model.
+in the interest of brevity, we omit the inferential forms for this model, sufficing to say that updating $\bm{\alpha}_j$ is accomplished with a joint Metropolis-within-Gibbs step; the normal prior for $\mu$ is conjugate for the log-normal $\log\bm{\alpha}_j$, and updated via a Gibbs step; and an inverse Wishart prior for $\Sigma$ is again conjugate to the log-normal $\bm{\alpha}_j$, meaning it too can be updated via a Gibbs step.
 
-Additionally, we investigate a finite mixture extension of the \emph{Projected Restricted Gamma} model,
-  where we define all $\beta_{\ell} := 1$, for $l = 1,\ldots,d$, for $j = 1,\ldots,J$.  This affects the above
-  model by removing inference steps for $\bm{ \beta},\bm{ \zeta},\bm{ \sigma}$.  The inference for
-  $\alpha_{\ell}$ for all $\ell = 1,\ldots,d$ follows the form in Equation~\ref{eqn:alphafc1}; inference on
-  $\bm{ \xi,\tau}$ follows the same form as the unrestricted model.  The impetus for the restricted
-  gamma model is that in the context of a mixture model, the additional flexibility offered by the
-  unrestricted Gamma model may be better represented with additional clusters.
-
-\subsubsection{Dirichlet Process Mixture Model}
-A logical extension to the finite mixture model described in Equation~\ref{eqn:fmixp} would be to assume the
-  $\bm{ y}$'s descending from an infinite mixture model.  This to some extent solves the issue of
-  selecting the number of mixture components, as that will effectively be done so in the model.
-  \begin{equation}
-    \label{eqn:infmixp}
-    \begin{aligned}
-      \bm{ y}_i &\sim \sum_{j = 1}^J\pi_j\text{PG}\left(\bm{ y}\mid \bm{ \alpha}_i, \bm{ \beta}_i\right)\\
-      (\alpha_i,\beta_i) &\sim \text{DP}\left(\eta, G_0\right)\\
-      &~\hspace{-2cm}G_0 = {\textstyle\prod}_{\ell = 1}^d\text{Ga}\left(\alpha_{jl}\mid\xi_{l},\tau_{l}\right)
-                    \times{\textstyle\prod}_{\ell = 2}^d\text{Ga}\left(\beta_{jl}\mid\zeta_{l}\tau_{l}\right)
-    \end{aligned}
-    \hspace{1cm}
-    \begin{aligned}
-      \bm{ \xi},\bm{ \tau} &\sim {\textstyle \prod}_{\ell = 1}^d \text{Ga}(\xi_{\ell}\mid a,b)\times \text{Ga}(\tau_{\ell}\mid c,d)\\
-      \bm{ \zeta},\bm{\sigma} &\sim {\textstyle\prod}_{\ell = 2}^d\text{Ga}(\zeta_{\ell} \mid a,b)\times \text{Ga}(\sigma_{\ell}\mid c,d)\\
-      \bm{ \eta} &\sim \text{Ga}(\eta \mid 2, 0.1).
-    \end{aligned}
-  \end{equation}
-  To fit the Dirichlet process, we use a variant of the collapsed Gibbs sampler detailed as
-  Algorithm~8 in \citet{neal2000}.  In this algorithm, we perform a data augmentation to create a
-  cluster identifier, $\delta_i$, for each observation. Let $J$ indicate the number of extant clusters.
-  Under the collapsed sampler scheme, for existing clusters, the probability of cluster membership is
-  calculated as
-  \begin{equation}
-    \label{eqn:dpdelta}
-    \text{P}\left(\delta_i = j\mid r, \bm{ y_i} \bm{ \alpha}, \bm{ \beta}\right) \propto
-        \frac{n_j^{\not i}}{\sum_{k = 1}^{J} n_k + \eta}
-        {\textstyle\prod}_{\ell = 1}^d\text{Ga}\left(r_iy_{il}\mid\alpha_{jl},\beta_{jl}\right).
-  \end{equation}
-  where $n_j^{\not i} = {\textstyle\sum}_{k\neq i}1_{k == j}$.  Effectively, if the observation under review is a
-  singleton, its existing cluster is destroyed.  For new clusters, ostensibly we would integrate out
-  cluster parameters to achieve a posterior predictive density; though doing so in a
-  non-conjugate model may be difficult.  Instead, the impetus for using this specific DP fitting algorithm
-  is it simulates the posterior predictive density by generating $m$ new candidate clusters from
-  their respective priors given $\bm{ \xi,\tau,\zeta,\sigma}$.  Then the probability of $y_i$
-  belonging to a particular candidate cluster is calculated as
-  \begin{equation}
-    \label{eqn:dpdeltanew}
-    \text{P}\left[\delta_i = k\mid r, \bm{ \alpha}_k^{\prime}, \bm{ \beta}_k^{\prime}\right] \propto
-        \frac{\eta / m}{\sum_{j = 1}^{J} n_j + \eta}
-        {\textstyle\prod}_{\ell = 1}^d\text{Ga}\left(r_iy_{il}\mid\alpha_{kl}^{\prime},\beta_{kl}^{\prime}\right),
-  \end{equation}
-  where $(\cdot)^{\prime}$ indicates a value as belonging to a new candidate cluster, and $k$ indexes
-  along candidate clusters.  If a new cluster is selected, then
-  $(\bm{\alpha}_{J+1},\bm{\beta}_{J+1}) = (\bm{\alpha}_k^{\prime},\bm{\beta}_k^{\prime})$, and $J = J + 1$.
-
-Inference upon $\bm{ \alpha}_j,\bm{ \beta}_j$ operates in the same manner as with the finite mixture
-  model.  For $\bm{ \xi,\tau,\zeta,\sigma}$, we have $J$ cluster components, indicating $J$
-  observations that again are under a gamma-gamma model.  The rate parameters have a previously
-  described conjugate posterior, that we can integrate out to more directly sample the shape
-  parameters.  As before, we will be looking at this model, and the restricted gamma form of this model, where
-  $\beta_{\ell} := 1$ for $\ell = 1,\ldots,d$.  This removes inference steps for $\bm{ \zeta},\bm{ \sigma}$
-  as well.
-
-\subsubsection{A Log-Normal Prior for Shape Parameters}
-A final extension we can make to this model is to allow for some structure, or dependence, in the prior
-  for shape parameters within a cluster.  The impetus for this approach is from recognizing that the
-  algorithm we are using for sampling the DP may not adequately represent the posterior density for
-  a new cluster, as it is assuming sampling from an independent prior.  If we account for dependence
-  between dimensions within the prior, then we may be able to more efficiently sample new clusters.
-  As shape parameters must be greater than 0, it is logical to assume a log normal prior, as that
-  allows us the familiar forms of a mean vector and covariance matrix for representing the source of
-  cluster parameters. So,
-  \begin{equation}
-    \label{eqn:dpln}
-    \begin{aligned}
-      \bm{ y}_i &\sim \sum_{j = 1}^J\pi_j\text{PG}\left(\bm{ y}\mid \bm{ \alpha}_i, \bm{\beta}_i\right)\\
-      (\bm{\alpha}_i, \bm{\beta}_i) &\sim \text{DP}\left(\eta, G_0\right)\\
-        &~\hspace{-1cm}G_0 = \mathcal{N}\left(\log\bm{ \alpha}_{j}\mid\mu,\Sigma\right)\times
-            {\textstyle\prod}_{\ell = 2}^d\text{Ga}\left(\beta_{j,\ell}\mid\zeta_{\ell},\sigma_{\ell}\right)
-    \end{aligned}
-    \hspace{1cm}
-    \begin{aligned}
-      \mu,\Sigma &\sim \mathcal{N}\left(\mu\mid\mu_0,\Sigma_0\right) \times \text{IG}\left(\nu,\Psi\right)\\
-      \bm{ \zeta},\bm{\sigma} &\sim {\textstyle\prod}_{\ell = 2}^d\text{Ga}(\zeta_{\ell} \mid a,b) \times \text{Ga}(\sigma_{\ell} \mid c,d) \\
-      \bm{ \eta} &\sim \text{Ga}(\eta \mid 2, 0.1).
-    \end{aligned}
-  \end{equation}
-  We hope that this will allow us to better incorporate the relationship between dimensions in
-  sampling new clusters, allowing for better mixing of the model.  The downside is one of computational
-  complexity---this introduces a $d$-dimensional normal distribution into the model, requiring a
-  $d$-dimensional matrix inversion step.  As matrix inversion is computationally costly, this limits
-  the potential number of dimensions possible.  The posterior inference strategies for $\beta,\zeta,\sigma$
-  are the same as for prior models and have been described.  For inference on $\alpha_j$ given
-  $r, \mu,\Sigma$, we rely on a $d$-dimensional Metropolis Hastings step, where the log-posterior
-  density is proportional to
-  \begin{equation}
-    \label{eqn:dplnlp}
-    \begin{aligned}
-    f(\bm{\alpha}_j\mid\bm{\delta},\bm{y},\mu,\Sigma,\zeta,\sigma) &\propto
-    \frac{\left(\prod_{i:\delta_i = j}y_{i,\ell}\right)^{\alpha_{j,\ell} - 1}}{\Gamma(\alpha_{j,\ell})^{n_j}}
-        \prod_{\ell = 2}^d\left[\frac{\left(\prod_{i:\delta_i = j}y_{i,\ell}\right)^{\alpha_{j,\ell} - 1}
-        \Gamma(n_j\alpha_{j,\ell} + \zeta)}{\Gamma(\alpha_{j,\ell})^{n_j}\left(\sum_{i:\delta_i = j}y_{i,\ell}
-        + \sigma\right)^{n_j\alpha_{j,\ell} + \zeta}}\right]\\
-        &\hspace{1cm}\times
-        \lvert \Sigma\rvert^{-\frac{1}{2}}\prod_{\ell = 1}^d\alpha_{j,\ell}^{-1}\exp\left\lbrace
-          -\frac{1}{2}(\log\alpha_j - \mu)^T\Sigma^{-1}(\log\alpha_j - \mu)\right\rbrace.
-    \end{aligned}
-  \end{equation}
-  Updating $\beta_j$ is accomplished with a conjugate Gibbs step, using the full conditional detailed
-  in Equation~\ref{eqn:betafc}, again subsetting $i$ such that $\delta_i = j$.  Note that we offer this
-  model for completeness---in practice, attempting to learn $\beta$ with $\alpha$ having a log-normal
-  prior resulted in extremely poor model performance.  We thus used the restricted gamma likelihood,
-  assuming $\beta_{j,\ell} = 1$ for all $j,\ell$.  This simplifies Equation~\ref{eqn:dplnlp} into
-  \begin{equation}
-    \label{eqn:dplnlpr}
-    \begin{aligned}
-    f(\bm{\alpha}_j\mid\bm{\delta},\bm{y},\mu,\Sigma) &\propto
-    \prod_{\ell = 1}^d\left[\frac{\left(\prod_{i:\gamma_i = j}
-        y_{i,\ell}\right)^{\alpha_{j,\ell} - 1}}{\alpha_{j,\ell}\Gamma(\alpha_{j,\ell})^{n_j}} \right]
-        \times\lvert\Sigma\rvert^{-\frac{1}{2}}
-      \exp\left\lbrace-\frac{1}{2}(\log\alpha_j - \mu)^T\Sigma^{-1}(\log\alpha_j - \mu)\right\rbrace.
-    \end{aligned}
-  \end{equation}
-  This model is conjugate for $\mu,\Sigma$, allowing simple Gibbs step updating.  The full
-    conditional distributions for $\mu$, $\Sigma$ take the form
-  \begin{equation}
-    \begin{aligned}
-    \mu\mid\Sigma,\alpha &\sim \mathcal{N}\left((n^*\Sigma^{-1} + S^{-1})^{-1}(n^*\overline{\log\alpha}^T\Sigma^{-1}
-      + \mu_0^TS^{-1}),(n^{*}\Sigma^{-1} + S^{-1})^{-1}\right)\\
-    \Sigma\mid\mu,\alpha &\sim \mathcal{IW}\left(
-        n^* + \nu, {\textstyle\sum}_{j = 1}^{n^*}(\log\alpha_j - \mu)(\log\alpha_j - \mu)^T + \Psi\right),
-    \end{aligned}
-  \end{equation}
-  where $n^*$ specifies the number of extant clusters.  Removing $\beta$ additionally removes the
-  inference steps for $\zeta,\sigma$.
-
-\subsubsection{Choice of Norm}
-In the extreme value case, we have defined our task on $\mathcal{S}_{\infty}^{d-1}$, the positive orthant
+\subsubsection{Choice of Norm\label{subsubsec:norm}}
+In the extreme value case, we have defined our task on ${\mathbb S}_{\infty}^{d-1}$, the positive orthant
   of the hypersphere defined under the $\mathcal{L}_{\infty}$ norm.  We previously described the
   difficulty in establishing a density directly on this geometry.  However, we can build the projected
-  gamma distribution on $\mathcal{S}_p^{d-1}$ for a large $p$.  We can then project samples from the
-  posterior predictive distribution for a model built on $\mathcal{S}_p^{d-1}$ onto
-  $\mathcal{S}_{\infty}^{d-1}$. As $p$ increases, the closer the manifold will be to
-  $\mathcal{S}_{\infty}^{d-1}$, and thus the closer the projected distribution will be to an idealized
-  projected Gamma built on $\mathcal{S}_{\infty}^{d-1}$.  In testing, we tried developing models on
+  gamma distribution on ${\mathbb S}_p^{d-1}$ for a large $p$.  We can then project samples from the
+  posterior predictive distribution for a model built on ${\mathbb S}_p^{d-1}$ onto
+  ${\mathbb S}_{\infty}^{d-1}$. As $p$ increases, the closer the manifold will be to
+  ${\mathbb S}_{\infty}^{d-1}$, and thus the closer the projected distribution will be to an idealized
+  projected Gamma built on ${\mathbb S}_{\infty}^{d-1}$.  In testing, we tried developing models on
   $S_p^{d-1}$ for varying levels of $p$, and found that higher $p$ produced higher performing models.
   Additionally, the difference in model performance became more apparent as dimensionality increased.
-  We present results here for models developed on $\mathcal{S}_{10}^{d-1}$.
+  We present results here for models developed on ${\mathbb S}_{10}^{d-1}$.
 

### 2021-10-27 results.tex
overleaf updates

--- 2021-09-13 results.tex
@@
-\section{Results}
+\section{Data illustrations\label{sec:results}}
 
-\subsection{Simulation}
+\subsection{Simulations\label{subsec:simulated}}
 We have introduced this projected gamma distribution existing on $\mathcal{S}_{\infty}^{p-1}$.  As a
@@
   ranging between 3 and 12.  The generation procedure is detailed in Algorithm~\ref{algo:simulated}.
-  \begin{algorithm}[htb]
+  \begin{algorithm}[ht]
+    \caption{Simulated Angular Dataset Generation Routine\label{algo:simulated}}
     \For{$n_{\text{mix}}$ in $[3, 6, 9, 12]$}{
@@
       \For{$n_{\text{col}}$ in $[3,6,12,20]$}{
-        Project columns 1 to $n_{\text{col}}$ of ${\bf X}$ onto $\mathcal{S}_{\infty}^{n_{\text{col}} - 1}$ and save.
+        Project columns 1 to $n_{\text{col}}$ of ${\bf X}$ onto 
+        $\mathcal{S}_{\infty}^{n_{\text{col}} - 1}$ and save.
         }
     }
-    \caption{Simulated Angular Dataset Generation Routine\label{algo:simulated}}
   \end{algorithm}
@@
 \begin{figure}[ht]
-  \caption{Simulation Energy Score versus Column Count, by Mixture Component Count.\label{fig:simes}}
+  \caption{Simulation Energy Score versus Column Count, by Mixture Component Count. Lower is better.\label{fig:simes}}
   \centering
@@
   energy score against the number of columns, grouped by the number of mixture components used in
-  dataset generation.  The \emph{Generative} model refers to a dataset generated
+  dataset generation.  The \emph{Gen} model refers to a dataset generated
   in the same manner, from the same parameters as was used to create the original dataset.  Against
@@
 \begin{figure}[ht]
-  \caption{Simulation Posterior Predictive Loss versus Column Count, by Mixture Component Count.\label{fig:simppl}}
+  \caption{Simulation posterior predictive loss versus column count, by mixture component count.\label{fig:simppl}}
   \centering
@@
 
-\begin{figure}[ht]
-  \caption{Estimated $k$NN-KL Divergence metric on simulated data, averaged between 1 and 10 nearest
-    neighbors, against column count, for selected models.  Graphs are facetted according to the
-    number of mixture components in the generating distribution.  The \emph{generative} model
-    represents another replicate of the underlying distribution, so that should be regarded as
-    \emph{truth}.\label{fig:simkld}}
-  \includegraphics[width = 5.5in, height = 3.75in]{./images/simulation_knn_kld}
-\end{figure}
-
-in Figure~\ref{fig:simkld} we see the application of the $k$NN-based K.L. divergence metric detailed
-  in Equation~\ref{eqn:knnkld} to the simulated data sets.  For display, this metric has been averaged
-  between $1$ and $10$ neighbors.  In general, the K.L. divergence metric is a log-ratio
-  of densities, integrated over the entire space.  In this fashion, we should interpret that model
-  is \emph{best} that stays closest to 0, as that indicates the two densities are equivalent.  The
-  equation also includes a scaling factor for the size of the data sets, as all else being equal, a
-  larger data set will decrease the distance between any point and its neighbors.  We can regard the
-  \emph{generative} model as another replicate data set from the generating distribution---another
-  view of truth.  As its distribution should be almost identical, we would assume that the KL divergence
-  metric from this model will come closest to 0.  It does not.  Knowing this to be the case, we should
-  instead regard the generative model as truth, and treat as best the model which comes closest to the
-  \emph{generative} model.  In this regard, we observe that the projected gamma model comes closest,
-  with the projected restricted gamma model following behind.  The probit-normal model quickly diverges
-  as the number of dimensions increases.  This is not unexpected, as we saw similar behavior in the
-  energy score and posterior predictive loss criteria.
-
-\subsection{Integrated Vapor Transport}
-We use data from the integrated water vapor transport (\emph{IVT}) model \citep{guan2015}
-  of atmospheric rivers as a means of testing these models.  An atmospheric river is a meteorological
-  event of local concentration of water vapor in the atmosphere that moves with wind patterns.
-  Understanding extremal dependence in atmospheric rivers, for instance, grants hydrological engineers
-  a greater understanding of how and where to build ground water capacity to handle precipitation.
-  California suffers persistent, extended droughts---if ground water transportation and holding capacity
-  (canals, reservoirs) are not built to manage extreme precipitation without adequate understanding of
-  the extremal dependence, then we may not install adequate capacity and thus experience flooding.
-  Alternatively, if we install too much capacity everywhere, then we are wasting money that could be
-  better spent elsewhere.
+\subsection{Integrated Vapor Transport\label{subsec:ivt}}
+The \emph{integrated vapor transport} (IVT) tracks the total water volume in a column of air over a given
+  area \citep{ralph2017}.  IVT is the means by which we track the formation and path of atmospheric 
+  rivers---elongated areas of high local concentration of water vapor in the atmosphere that moves with
+  wind patterns.  Extreme behavior in IVT can lead to extreme precipitation events, causing flooding; 
+  but in California, those same events are responsible for the bulk of precipitation that inland 
+  California recieves---water necessary for agriculture and household consumption.  Thus it becomes 
+  necessary to manage precipitation to prevent inundation and support agriculture.  To this end, 
+  understanding the dependence structure of extreme events in the IVT is useful---for instance, for 
+  hydrological engineers in planning the capacity of groundwater systems to capture precipitation before 
+  wasteful and destructive flooding occurs.  Our data tracks daily average values for the IVT, dividing 
+  the coast of California into grid cells, with two spatial resolutions of 8 and 47 cells, respectively 
+  \citep{guan2015}.
 
@@
 \begin{algorithm}
-  \KwResult{$\bm{ r} : r_i \sim \text{Pareto}(1)$, $\bm{ v} \in \mathcal{S}_{\infty}^{d-1}$}
+  \KwResult{$\bm{ r},\bm{v} : r_i \sim \text{Pareto}(1)$, $\bm{ v}_i \in \mathcal{S}_{\infty}^{d-1}$}
   \For{$\ell = 1,\ldots,d$}{
@@
     Define $z_{i,\ell} = \left(1 + \xi_{\ell}\frac{x_{i,\ell} - b_{t,\ell}}{a_{\ell}}\right)_{+}^{1/\xi_{\ell}}$\\
-    Define $r_i = \pnorm{\bm{ z}_i}{\infty}$, $\bm{ V}_i = \frac{\bm{ z}_i}{\pnorm{\bm{ z}_i}{\infty}}$\\
+    Define $r_i = \pnorm{\bm{ z}_i}{\infty}$, $\bm{ v}_i = \frac{\bm{ z}_i}{\pnorm{\bm{ z}_i}{\infty}}$\\
     }
@@
 
-\begin{table}[b]
+\begin{table}[ht]
   \centering
   \caption{Model comparison metrics: Posterior Predictive Loss and Energy Score criteria from fitted
-    models against the IVT data.  All presented models are DP mixtures; the \emph{Model} field
-    identifies the kernel distribution.  For both criteria, lower is better.
+    models against the IVT data.  Lower is better.
   \label{tab:dev}}
-  \include{table_dev}
+  \include{./tables/table_dev}
 \end{table}
@@
 In Table~\ref{tab:dev} we see the same model preference towards the projected restricted gamma models
-  that we saw in Figures~\ref{fig:simes},\ref{fig:simppl}.  The unrestricted gamma model is penalized much
+  that we saw in Figures~\ref{fig:simes}.  The unrestricted gamma model is penalized much
   more strongly on real data than we saw with our simulation, performing comparably to the probit-normal
@@
 
-That said, there is much we can glean from this table.  The probit-normal model, developed on a
-  mapping from $\mathcal{S}_{2}^{d-1}$ to $(-\infty,\infty)^{d-1}$ via probit transformation of
-  spherical coordinates was not competitive with the gamma models on real data, similar to what we
-  saw in the simulation study. We also established finite mixture models for all presented
-  distributions, but after tuning the number of mixture components, their model performance was,
-  and could only be \emph{as good} as the DP mixture.  In the interest of parsimony, we only present
-  the DP models as representative of mixture distributions.
-
-\begin{figure}[ht]
-  \centering
-  \caption{KL Divergence Curves calculated through the KNN-KL Metric, evaluated between the
-  empirical dataset and posterior predictive datasets of various models.  The top row corresponds
-  to the 8-dimensional data, while the bottom row the 47.  All presented models use a DP prior,
-  the nominal model being the kernel distribution.\label{fig:knnkl}}
-  \includegraphics[width = 5.5in, height = 2.75in]{./images/knn_kld}
-\end{figure}
-
-Interpreting the KL divergence curves in Figure~\ref{fig:knnkl} is more difficult.  As we saw in
-  the simulated case in Figure~\ref{fig:simkld}, the general logic of \emph{best} being that which
-  is closest to zero is not necessarily in use here.  We interpret a positive value in the KL
-  divergence metric as indicating that, for a given point in the given distribution, we must
-  travel \emph{farther} on the target distribution to reach the $k$th neighbor as compared to
-  the given distribution.  In this regard, we should regard \emph{large} positive values as
-  indicating a particularly poor fit.  By this metric, the projected gamma and probit-normal
-  models do not appear to scale well as they performed very poorly on the high dimensional data.
-  For the projected restricted gamma model, the log-normal prior appears to scale better than the
-  gamma prior.  This is in keeping with what we observed in Table~\ref{tab:dev}, but is a
-  different result than what we observed in the simulation study, where we observed that the
-  log-normal prior only tended towards the gamma prior, never performing better.
-
 % EOF

### 2021-10-30 evaluation.tex
updates

@@
   as the Euclidean norm is a negative definite kernel.$\hfill\Box$
+
 Evaluating $g$ as defined in Proposition \ref{prop:g} requires the solution of a $d-2$-dimensional 
@@
 We have already declared $\bm{c}\in{\mathbb C}_{\jmath\ell}^{d-2}$, so it must reside at $\overrightarrow{\bm{a}\bm{b}^{\prime}}\cap{\mathbb C}_{\jmath\ell}^{d-2}$.
-Because the straight line between $\bm{a}$ and $\bm{b}^{\prime}$ passes through this point without further analysis, the optimization of $\bm{c}$ is accomplished \emph{en passant}.$\hfill\Box$
+Because the straight line between $\bm{a}$ and $\bm{b}^{\prime}$ passes through this point without further analysis, the optimization of $\bm{c}$ is accomplished \emph{en~passant}.$\hfill\Box$
+
+Interpretation of the energy score as we have formulated it here is the same as that of posterior predictive loss: smaller is better.
 

### 2021-10-30 introduction.tex
updates

@@
   in multivariate distributions is to use a copula.
-\cite{renard2007,deng2011,falk2019}, provide successful examples of this approach in an EVT framework.
-{\bf is \cite{deng2011} really a relevant reference?
-It looks weird.
-Can we say something more about the copula approach?
-I confess my ignorance here.}
+\cite{renard2007,falk2019}, provide successful examples of this approach in an EVT framework.
 \cite{ferreira2014} presents a constructive definition of the Pareto process, that generalizes the GP to an infinite dimensional setting.
@@
 
-% Extreme value theory bears strong relevance to climate analysis, as climatological events of interest occur at the tail of the distribution, both in terms of magnitude and frequency.
-We motivate our proposed model with an analysis of the integrated vapor transport data for observing atmospheric rivers---high concentrations of water vapor that lead to extreme precipitation events.
-
-% To motivate the topic and our proposed solution, we acknowledge the relevance of extreme analysis to
-%   climatological events.  Outcomes such as heat waves, extreme precipitation, storm surges, and
-%   such present a similar challenge to investigators, in that the events of interest can be very destructive,
-%   and occur in probability at the tail end of \emph{normal} behavior.  In this fashion, a model built
-%   to capture \emph{normal} behavior may struggle to adequately represent extreme behavior.  More
-%   specialized tools become necessary; especially so in a time of intensifying extreme weather
-%   events.  To this end, we use the integrated water vapor
-%   transport (\emph{IVT}) data for observing atmospheric rivers as an example application for our proposed method.
-
 Throughout the paper, we adopt the operators $\wedge$ to denote minima, and the $\vee$
@@
 \section{A multivariate PoT model\label{sec:multivariatepot}}
-%\subsubsection{Univariate Maxima}
-%Regarding the asymptotic behavior of extreme events, there are a couple major strategies for conducting
-%  inference---developing probabilistic estimates of extreme behavior.  First developed is the theory
-%  of a limiting distribution on maxima---the largest observation from a sample.  For a sample $\bm{x}$
-%  where $\bm{x} = (x_1,\ldots,x_n)$ represents a sequence of $n$ independent random variables from a
-%  distribution function $F$, the distribution of the maximum $M_n$ of this sequence can be derived
-%  as:
-%  \begin{equation*}
-%    \begin{aligned}
-%      \text{Pr}\left[M_n\leq z\right] &= \text{Pr}\left[X_1 \leq z, \ldots, X_n \leq z\right]\\
-%        &= \text{Pr}\left[X_1\leq z\right]\times\ldots\times\text{Pr}\left[X_n\leq z\right]\\
-%        &= F(z)^n.
-%    \end{aligned}
-%  \end{equation*}
-%  In situations where $F$ is unknown, we can seek to approximate the behavior of $F^n$ as
-%  $n\rightarrow\infty$.  To ensure this does not degenerate to a point mass, we need to consider a
-%  standardized sequence of maxima. If this sequence stabilizes as $n$ increases, the a limiting
-%  distribution exists.  More specifically, if there exists some sequence of constants $a_n > 0$, $b_n$
-%  such that:
-%  \begin{equation*}
-%    \text{Pr}\left[\frac{M_n - b_n}{a_n} \leq z\right] \stackrel{d}{\rightarrow} G(z)
-%  \end{equation*}
-%  as $n\rightarrow\infty$, then $G$ is a max-stable distribution, and $F$ is in the domain of
-%  attraction of that max stable distribution.  \cite{frechet1927} originates the
-%  field, identifying two limiting forms of max-stable distributions, which would become known as the
-%  Fr{\'e}chet and Weibull distributions. \cite{weibull1951} expands the analysis of the Weibull
-%  distribution.  \cite{fisher1928}, in what is now called the \emph{Fisher-Tippett Theorem}, identifies
-%  the Fr{\'e}chet and Weibull distributions, along with an as then unnamed third form, as the three
-%  limiting forms of the distribution of the maxima of a sample.  \cite{gumbel1935,gumbel1942} offer
-%  analysis of that third form, now known as the Gumbel distribution.  Later works, including
-%  \cite{jenkinson1955} reparameterize all three forms as special cases of a single unifying form,
-%  the generalized extreme value distribution, \emph{GEV}:
-%  \begin{equation*}
-%    \label{eqn:gev}
-%    F(m \mid \mu, \sigma, \xi) = \exp\left\lbrace-\left[1 + \xi\left(\frac{x - \mu}{\sigma}\right)\right]_{}^{-1/{\xi}}\right\rbrace.
-%  \end{equation*}
-%  Thus distributions in the domain of attraction of an EVD (Gumbel, Fr{\'e}chet, or Weibull) will be
-%  in the domain of attraction of the GEV.
-
-%As this distribution specifies asymptotic behavior for the maximum of a set of observations,
-%  inference assuming this distribution requires we that specify some block of data, and for the block
-%  report only the maximum.  A series of these blocks yielding a series of maxima allows us to conduct
-%  inference about the parameters of the distribution.  Taking only the
-%  maximum in a block of observations necessarily leads to the reduction of our sample size by a factor
-%  of $1/\text{block size}$. In problems where the data occur in natural blocks, such as an hourly time
-%  series where a natural block might be a day, this might be appropriate.  There is an implicit
-%  violation of the assumption of independence within a block, in most cases, but that violation is
-%  generally ignored. In data without a natural block, it may be difficult to justify an induced
-%  artitifial block. Moreover, by reducing available information, retaining only one datum for each
-%  block increases the variability of the parameter estimates.
-
-%\subsubsection{Univariate exceedences over a high threshold}
-%The \emph{Pickands-Balkema-de Haan Theorem}~\citep{balkema1974,pickands1975} offers us another means
-%  of conducting inference that will be less wasteful of data.  It states that if the original distribution
-%  $F$ is in the domain of attraction of the GEV, then excesses over a high threshold $u$ can be well
-%  modelled using the Generalized Pareto distribution.  Again, let $X$ follow some distribution function
-%  $F$. It follows that:
-%  \begin{equation*}
-%    \text{Pr}\left[X > u + y\mid X > u\right] = \frac{1 - F(u + y)}{1 - F(u)}
-%  \end{equation*}
-%  for $y > 0$.  If $F$ is in the domain of attraction of an EVD, then
-%  \begin{equation*}
-%    \lim\limits_{u\to u^{\prime}}\text{Pr}\left[X > u + y\mid X > u\right] = H(y)
-%  \end{equation*}
-%  where $u^{\prime}$ represents some upper boundary of $X$ has a functional form---the survival
-%  function of a Pareto distribution. We approximate this asymptotic relationship by setting some
-%  large threshold $u$, and then threshold exceedences $Y = X - u$ are modelled under the distrribution
-%  function
-%  \begin{equation*}
-%    \label{eqn:gp}
-%    H(y) = 1 - \left(1 + \xi\frac{y}{\sigma}\right)^{-\frac{1}{\xi}}.
-%  \end{equation*}
-%  This defines the generalized Pareto family of distributions.  Thus, if block maxima have a
-%  limiting distribution $G$ within the EVD family, then threshold exceedances for a sufficiently
-%  high threshold have a limiting distribution $H$ within the Generalized Pareto (GP) family.  We can
-%  identify $\sigma$ as a scale parameter, but $\xi$ deserves special mention as the extremal index.
-%  We can interpret the Pareto tail probability as
-%  \begin{equation*}
-%    \lim\limits_{t\to\infty}\frac{1 - F(ty)}{1 - F(t)} = y^{-\frac{1}{\xi}}
-%  \end{equation*}
-%  for $y > 1$.  Importantly, $\xi$ from the limiting distribution of excesses over a threshold is the
-%  same parameter, and has the same value as the $\xi$ from the limiting distribution of the maximum
-%  observation of a sample \citep{dehaan2006}.
-
-%One other point of note is that for time series, the data are temporally correlated.  Thus the implicit
-%  assumption of independence for those observations in excess of a threshold is violated.  One means
-%  of ameliorating the effect of this correlation is to decluster the data---and one method of declustering
-%  is, for a series of observations in excess of the threshold, to keep only the maximum observed value
-%  in the series.
-
-%\subsubsection{Multivariate threshold EVT}
-
-%The generalization of EVT to multiple dimensions presents a unique challenge in the case of maxima---how
-%  do we define the maximum of a sample?  How can we order observations in a multivariate scenario?
-%  \cite{rootzen2006} offers a brief review of the subject.  One possible treatment sidesteps this
-%  conundrum by using component maxima---let $\bm{M}_n = \left(\vee_i X_{i1},\ldots,\vee_i X_{id}\right)$.
-%  Then the multivariate GEV establishes the limiting behavior of $\bm{M}_n$ as $n\to\infty$.  Note
-%  that $\bm{M}_n$ need not be in the observed $\left\lbrace\bm{X}_i; i = 1,\ldots,n\right\rbrace$.
-%  Then, if there is a set of constants $\bm{a}_n > \bm{0}$, $\bm{b}_n$ such that
-%  \begin{equation*}
-%    \lim\limits_{n\to\infty}\text{P}\left(\bigcap_{\ell = 1}^d
-%      \frac{M_{n\ell} - b_{n\ell}}{a_{n\ell}} \leq x_{\ell}\right) = G(\bm{x})
-%  \end{equation*}
-%  with non-degenerate marginals, then $G(\bm{x})$ is a \emph{multivariate extreme value distribution}.
-%  The marginal distribution of $M_{nl}$ is by construction a GEV with the parameters $a_{n\ell}, b_{n\ell}, \xi$.
-%  With the existence of the marginal distributions established as a consequence of the existence of the
-%  GEV, multivariate EVT then splits into two components---estimation of the marginal parameters, and
-%  estimation of the dependence structure.  This separation is born out extensively in the literature.
-
 To develop a multivariate PoT model for extreme values consider a $d$-dimensional random vector $\bm{Z} \in {\mathbb R}^d_+ = [0,+\infty)^d$, the positive $d$-dimensional cone.
@@
 Suppose that $\bm{Z} = R\bm{V}$ with $R\sim Pa(1)$,
-$\prob{V_i > 0} = 1$ and $\expect{V_i}$ exists ,
-$i=1, \ldots ,d$, then
+$\prob{V_\ell > 0} = 1$ and $\expect{V_\ell}$ exists ,
+$\ell=1, \ldots ,d$, then
 \begin{equation}
-	\chi_{ij} = \expect{\frac{V_i}{\expect{V_i}} \wedge \frac{V_j}{\expect{V_j}}}
+    \label{eqn:chi_ij}
+	\chi_{\jmath\ell} = \expect{\frac{V_\jmath}{\expect{V_\jmath}} \wedge \frac{V_\ell}{\expect{V_\ell}}}
 \end{equation}
@@
 {\em Proof:}
-Denote as $F_i$ the marginal distribution of $Z_i$. To obtain $\chi_{ij}$ we need $Pr(Z_1>z_1,Z_2>z_2)$, where $z_i =
-F_i^{-1}(u) = \expect{V_i}/(1 - u), \;i=1,\dots, d$.
+Denote as $F_\ell$ the marginal distribution of $Z_\ell$. To obtain $\chi_{\jmath\ell}$ we need $Pr(Z_\jmath>z_\jmath,Z_\ell>z_\ell)$, where $z_\ell =
+F_\ell^{-1}(u) = \expect{V_\ell}/(1 - u), \;\ell=1,\dots, d$.
 Using the fact that
-$V_i>0, \forall i$ almost surely, we have that the former is equal to
+$V_\ell>0, \forall \ell$ almost surely, we have that the former is equal to
 \begin{equation*}
-\prob{R>\frac{z_i}{V_i}\vee\frac{z_2}{V_j}}
-    = \expect{1\wedge\left(\frac{z_i}{V_i}\vee\frac{z_j}{V_j}\right)^{-1}}
-    = \expect{\frac{V_i}{z_i}\wedge\frac{V_j}{z_j}}
-    = (1 - u)\expect{\frac{V_i}{\expect{V_i}}\wedge\frac{V_j}{\expect{V_j}}},
+\prob{R>\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}}
+    = \expect{1\wedge\left(\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}\right)^{-1}}
+    = \expect{\frac{V_\jmath}{z_\jmath}\wedge\frac{V_\ell}{z_\ell}}
+    = (1 - u)\expect{\frac{V_\jmath}{\expect{V_\jmath}}\wedge\frac{V_\ell}{\expect{V_\ell}}},
 \end{equation*}
@@
 \begin{equation} \label{eqn:condsurv2df}
-    \text{Pr}\left[\bigcap_{l \in \alpha} Z_l > z_l \right|\left. \bigcap_{l\not\in\alpha} Z_l > z_l\right] =
+    \text{Pr}\left[\bigcap_{\ell \in \alpha} Z_\ell > z_\ell \right|\left. \bigcap_{\ell\not\in\alpha} Z_\ell > z_\ell\right] =
       \frac{\expect{\bigwedge_{k = 1}^d 1\wedge \frac{V_k}{z_k}}}{\expect{\bigwedge_{k \not\in\alpha}1\wedge\frac{V_k}{z_k}}}.
@@
 
-%\subsection{Integrated Vapor Transport}
-%Atmospheric rivers are temporary events, where large elongated regions of high concentrations of
-%  water vapor are developed in the atmosphere and carry huge amounts of water potentially thousands
-%  of miles.  The amount of water in transit during these events can dwarf that of terrestrial rivers.
-%  For the targeted region, the atmospheric river can represent a significant portion of the
-%  precipitation the region will experience.  Thus, such events are of great interest to meteorologists,
-%  hydrologists, actuaries, land-use planners, and even farmers.  The ability to estimate the likely magnitude
-%  and distribution of such events is thus of great importance~\citep{ralph2013,ralph2018}.
-
-%One metric by which we might identify and declare atmospheric rivers is the integrated water vapor
-%  transport, or \emph{IVT}.  This value represents the total amount of water
-%  vapor being transported in an atmospheric column---that is, a column of the troposphere above a given
-%  area of the surface of the Earth. Raw data are measured by dropsondes---measurement devices dropped
-%  out of aircraft, taking measurements as they fall to the earth.~\citep{ralph2017}  A data vector
-%  includes an estimate of IVT at grid cell at a period in time; where a grid cell specifies the surface
-%  of the earth under the atmospheric column.  We have two such datasets, in differing spatial
-%  resolutions---the lower resolution data splits the coast of california into 8 grid cells, while the
-%  higher resolution does so into 47 grid cells~\citep{guan2015}.
-
-%As IVT tracks water vapor in the atmosphere, extreme values in IVT can lead to
-%  extreme precipitation events.  Atmospheric rivers are necessary events, in that they carry the
-%  bulk of precipitation that California recieves, but extreme precipitation events can be extremely
-%  destructive as well.  Thus, we are interested in the distribution of extreme precipitation events,
-%  including their depenedence structure.  This tells us about the probability of jointly extreme
-%  events---where IVT is extreme in two or more grid cells, as well as identifying particular
-%  configurations of IVT that may be anomalous.  In California, groundwater is a managed resource.
-%  Extreme precipitation events significantly affect groundwater, so anomalous precipitation events
-%  can produce anomalous configurations of groundwater, leading to destructive flooding.
-
-%We are specifically interested in extremal dependence---the relationship between the upper tails of
-%  dimensions of the distribution.  In this case, that means the relationship between extreme values
-%  in different grid cells.  We are going to be looking at point-in-time behavior rather than
-%  considering the time series nature of the data---that relationship may come later.
-
-%As we are interested in the extremal dependence, it makes sense that we would choose to represent
-%  this data using tools from extreme value theory.  Extreme value theory, or \emph{EVT}, seeks to
-%  model and assess probability of observing extreme events.  Such a topic is applicable generally,
-%  but it finds particularly strong use among such fields as finance \citep{allen2013},
-%  climatology \citep{trepanier2018}, and insurance \citep{beirlant1994}.  In these fields, extreme
-%  events may represent significant loss to the body commissioning the study.  For instance, an
-%  insurance company might commission a study on extreme weather events, as an extreme weather event
-%  localized to a particular region could cause a spike in claims from that region.  Extreme value
-%  theory offers us a tool set for making inference about the tails of a distribution, without having
-%  observed said tails.  For
-%  instance, with an extreme weather event like flooding, we can make predictions about return
-%  levels---the average time until an observation of a particular magnitude occurs---without having seen
-%  an observation of that magnitude, or having observed that long.  In the context of our motivating
-%  example, extreme values in the IVT may represent the formation of an atmospheric river, which has
-%  dramatic effects on precipitation and ground water.  The formation of an IVT may lead to flooding
-%  and other negative effects, but is also contributes necessary water for irrigation and agriculture.
-%  Development of statistical tools regarding these extreme events becomes necessary for making informed
-%  decisions.
-
   % EOF

### 2021-10-30 methodology_projgamma.tex
updates

@@
 \end{equation}
-Of course, if a candidate cluster is selected, then $\delta_j = J + 1$, and $J = J^{-i} + 1$.
+Of course, if a candidate cluster is selected, then $\delta_j = J = J^{- i} + 1$, and the associated cluster parameters are saved.
 
@@
 \begin{equation*}
-    \mathcal{PG}(\bm{y}\mid\bm{\alpha}_i,\bm{\beta}_i) \propto \int\prod_{\ell = 1}^d\mathcal{G}\left(ry_{i\ell}\mid\alpha_{i\ell},\beta_{i\ell}\right)r^{d-1}dr,
+    \mathcal{PG}(\bm{y}\mid\bm{\alpha}_i,\bm{\beta}_i) = \int\prod_{\ell = 1}^d\mathcal{G}\left(ry_{i\ell}\mid\alpha_{i\ell},\beta_{i\ell}\right)r^{d-1}\left[y_d +
+            {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right]dr,
 \end{equation*}
@@
 the likelihood for $\bm{\alpha},\bm{\beta}$ becomes separable by dimension.
-We stress this point: \emph{given} $r$, the distributional parameters $(\alpha_j,\beta_j)_{\ell}$ are independent in the likelihood.
+We stress this point: \emph{given} $r$, the dimensions of the distributional parameters $\bm{\alpha}_j,\bm{\beta}_j$ are independent in the likelihood.
 
@@
 \end{equation}
-The first form of the centering distribution we explore is a product of independent Gammas.
+The first form of the centering distribution we explore is a product of independent Gammas:
 \begin{equation}
-    G_0(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{\xi},\bm{\tau},\bm{\zeta},\bm{\sigma}) = \prod_{\ell = 1}^d\mathcal{G}(\alpha_{j\ell}\mid \xi_{\ell},\tau_{\ell})\times\prod_{\ell = 2}^d\mathcal{G}(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell})
+    G_0(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{\xi},\bm{\tau},\bm{\zeta},\bm{\sigma}) = \prod_{\ell = 1}^d\mathcal{G}(\alpha_{j\ell}\mid \xi_{\ell},\tau_{\ell})\times\prod_{\ell = 2}^d\mathcal{G}(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}).
 \end{equation}
 This model is completed with Gamma priors on $\xi_{\ell}$, $\tau_{\ell}$, $\zeta_{\ell}$, $\sigma_{\ell}$.
-Optionally, placing a Gamma prior on $\eta$ allows updating via the procedure outlined in \cite{escobar1995}.
+We also placed a Gamma prior on $\eta$, updating via the procedure outlined in \cite{escobar1995}.
 In further analysis, we refer to this model as the \emph{projected Gamma--Gamma} (PG--G) model.

### 2021-10-30 results.tex
updates

@@
 
+% \begin{figure}[ht]
+%   \caption{Simulation Energy Score versus Column Count, by Mixture Component Count. Lower is better.\label{fig:simes}}
+%   \centering
+%   \includegraphics[width=0.8\textwidth]{./images/simulation_es}
+% \end{figure}
+
 \begin{figure}[ht]
-  \caption{Simulation Energy Score versus Column Count, by Mixture Component Count. Lower is better.\label{fig:simes}}
-  \centering
-  \includegraphics[width=0.8\textwidth]{./images/simulation_es}
+  \begin{subfigure}[b]{0.45\textwidth}
+    \includegraphics[width=\textwidth]{./images/simulation_ppl}
+    \caption{Posterior Predictive Loss\label{fig:simppl}}
+  \end{subfigure}
+  %
+  \begin{subfigure}[b]{0.45\textwidth}
+    \includegraphics[width=\textwidth]{./images/simulation_es}
+    \caption{Energy Score\label{fig:simes}}
+  \end{subfigure}
+  \caption{Simulation PPL (left) and ES (left) versus column count, by mixture component count.}
 \end{figure}
@@
 
-\begin{figure}[ht]
-  \caption{Simulation posterior predictive loss versus column count, by mixture component count.\label{fig:simppl}}
-  \centering
-  \includegraphics[width=0.8\textwidth]{./images/simulation_ppl}
-\end{figure}
+% \begin{figure}[ht]
+%   \caption{Simulation posterior predictive loss versus column count, by mixture component count.\label{fig:simppl}}
+%   \centering
+%   \includegraphics[width=0.8\textwidth]{./images/simulation_ppl}
+% \end{figure}
 
@@
   but in California, those same events are responsible for the bulk of precipitation that inland 
-  California recieves---water necessary for agriculture and household consumption.  Thus it becomes 
+  California receives---water necessary for agriculture and household consumption.  Thus it becomes 
   necessary to manage precipitation to prevent inundation and support agriculture.  To this end, 
@@
   \citep{guan2015}.
+
+\begin{algorithm}[h]
+  \KwResult{$\bm{ r},\bm{v} : r_i \sim \text{Pareto}(1)$, $\bm{ v}_i \in {\mathbb S}_{\infty}^{d-1}$}
+  \For{$\ell = 1,\ldots,d$}{
+    Set $b_{t,\ell} = \hat{F}_{\ell}^{-1}\left(1 - \frac{1}{t}\right)$.\\
+    With $\bm{ x}_{\ell} > b_{t,\ell}$, fit $a_{\ell}$, $\chi_{\ell}$ via MLE according to generalized Pareto likelihood.\\
+    }
+  \For{$i = 1,\ldots,n$}{
+    Define $z_{i,\ell} = \left(1 + \xi_{\ell}\frac{x_{i,\ell} - b_{t,\ell}}{a_{\ell}}\right)_{+}^{1/\xi_{\ell}}$\\
+    Define $r_i = \pnorm{\bm{ z}_i}{\infty}$, $\bm{ v}_i = \frac{\bm{ z}_i}{\pnorm{\bm{ z}_i}{\infty}}$\\
+    }
+  Subset $\bm{ r},\bm{ v}$ such that $r_i \geq 1$\\
+  \If{declustering}{
+    \For{$i = 1,\ldots,n$}{
+      If $r_i \geq 1$ and $r_{i-1} \geq 1$, drop the lesser (and associated $v_i$) from data set.\\
+    }
+  }
+ \caption{Data preprocessing to isolate and transform data exhibiting extreme behavior.  $r_i$
+   represents the radial component, and $\bm{v}_i$ the angular component.  The declustering
+   portion is relevant if data is correlated in time.\label{algo:processing}}
+\end{algorithm}
 
@@
   generalized Pareto---the scale parameter $\alpha_{\ell}$ and the extremal index $\chi_{\ell}$---are set via
-  maximum likelihood.  A fully Bayesian model formulation will allow their varying within a
-  distribution, but such a model formulation will not be conducive to fitting by Markov-chain Monte
-  Carlo methods.
+  maximum likelihood.
 
 After the thresholding and maximum likelihood estimation of the parameters of the Pareto, we scale
-  the data to the standard multivariate Pareto.  Dividing each standardized observation by its
-  $\mathcal{L}_{\infty}$ norm, we project the standardized data onto $\mathcal{S}_{\infty}^{d-1}$.
-  Data in sequence represents observations in time, and these are heavily correlated.  As such, we
+  the data to the standard multivariate Pareto following Equation~\ref{eqn:standardization}.  
+  Dividing each standardized observation by its $\mathcal{L}_{\infty}$ norm, we project the 
+  standardized data onto $\mathcal{S}_{\infty}^{d-1}$. Data in sequence represents observations
+  in time, which for this application are temporally correlated.  As such, we
   choose to \emph{decluster} the observations, such that, by observing a sequence of observations
-  $\bm{ z}$ for which $\inorm{z_i} > 1$ for each observation in sequence, we keep only the observation
+  $\bm{ z}$ for which $\inorm{z_i} > 1$ for each observation in sequence, we keep only that observation
   with the greatest observed $\mathcal{L}_{\infty}$ norm.  The complete procedure is outlined in
-  Algorithm~\ref{algo:processing}.
-
-\begin{algorithm}
-  \KwResult{$\bm{ r},\bm{v} : r_i \sim \text{Pareto}(1)$, $\bm{ v}_i \in \mathcal{S}_{\infty}^{d-1}$}
-  \For{$\ell = 1,\ldots,d$}{
-    Set $b_{t,\ell} = \hat{F}_{\ell}^{-1}\left(1 - \frac{1}{t}\right)$.\\
-    With $\bm{ x}_{\ell} > b_{t,\ell}$, fit $a_{\ell}$, $\chi_{\ell}$ via MLE according to generalized Pareto likelihood.\\
-    }
-  \For{$i = 1,\ldots,n$}{
-    Define $z_{i,\ell} = \left(1 + \xi_{\ell}\frac{x_{i,\ell} - b_{t,\ell}}{a_{\ell}}\right)_{+}^{1/\xi_{\ell}}$\\
-    Define $r_i = \pnorm{\bm{ z}_i}{\infty}$, $\bm{ v}_i = \frac{\bm{ z}_i}{\pnorm{\bm{ z}_i}{\infty}}$\\
-    }
-  Subset $\bm{ r},\bm{ v}$ such that $r_i \geq 1$ for all $r_i\in \bm{r}$.\\
-  \If{declustering}{
-    \For{$i = 1,\ldots,n$}{
-      If $r_i \geq 1$ and $r_{i-1} \geq 1$, drop the lesser (and associated $v_i$) from data set.\\
-    }
-  }
- \caption{Data preprocessing to isolate and transform data exhibiting extreme behavior.  $r_i$
-   represents the radial component, and $\bm{v}_i$ the angular component.  The declustering
-   portion is relevant if data is correlated in time.\label{algo:processing}}
-\end{algorithm}
-
-We have, at our disposal, two data sets from the IVT model, totalling approximately 30 years of daily
-  values omitting leap days.  One records data from 8 grid cells, covering
-  generally the coast of California.  The other, with a higher resolution, records data from 47 grid
-  cells covering the same area.  For the 8-cell data, after subsetting to the rainy season, we have
-  $5587$ observations.  After the data processing steps in Algorithm~\ref{algo:processing}, where we
-  omit observations not extreme in any dimension, and decluster temporally correlated data, we are left
-  with $511$ observations.  For the 47-cell data, after subsetting to the rainy season, we have $6342$
-  observations.  After the same data processing, we are left with
-  $532$ observations.  We fit our models to both data sets.
+  Algorithm~\ref{algo:processing}.  In the 8-cell data, after subsetting to the rainy season we 
+  have $5587$ observations, which reduces to $511$ observations after processing and declustering.  
+  In the 47 cell data, after subsetting we have $6342$ observations, which reduces to $532$ after 
+  processing and declustering.  We fit PG--G, PRG--G, PG--LN, and PRG--LN models to both datasets.
 
@@
 
-In Table~\ref{tab:dev} we see the same model preference towards the projected restricted gamma models
-  that we saw in Figures~\ref{fig:simes}.  The unrestricted gamma model is penalized much
-  more strongly on real data than we saw with our simulation, performing comparably to the probit-normal
-  model.  We also see the restricted gamma model with the log-normal prior performing significantly
-  better than the restricted gamma model with the gamma prior on the higher dimensional data, a reversal
+In Table~\ref{tab:dev} we see the same model preference towards the projected restricted 
+  gamma models that we saw in Figures~\ref{fig:simes}.  The unrestricted gamma model is 
+  penalized much more strongly on real data than we saw with our simulation.  We also see
+  the restricted gamma model with the log-normal prior performing significantly better 
+  than the restricted gamma model with the gamma prior on the higher dimensional data, a reversal
   of what we saw on the low dimensional data and indeed from what we saw in the simulation study.
 
+\begin{figure}[ht!]
+    \centering
+    \caption{Pairwise extremal dependence coefficients for IVT data, from fitted PRG--G.  
+    The left corresponds to the 8-cell data, the right the 47-cell data.\label{fig:chi_ij}}
+    \begin{minipage}{.49\textwidth}
+      \centering
+      \includegraphics[width=0.99\linewidth]{./images/chi_ij_8}
+    \end{minipage}
+    \begin{minipage}{.49\textwidth}
+      \centering
+      \includegraphics[width=0.99\linewidth]{./images/chi_ij_46}
+    \end{minipage}
+\end{figure}
+
+Figure~\ref{fig:chi_ij} shows our implementation of Equation~\ref{eqn:chi_ij}, plotting the 
+  pairwise extremal dependence coefficients for the IVT data computed from PRG--G models 
+  fitted to the 8 (left) and 47 (right) cell data.  
+  That we see a higher extremal dependence between neighboring columns is not unexpected, as
+  neighboring columns correspond to neighboring cells.  We recognize that pairwise asymptotic
+  dependence coefficients tell a limited story, as a particular dependence may structure may include
+  more than two columns.   We can, however, glean some information from the patterns that emerge in
+  two dimensions.  On the 8-cell data, we see a stronger association between cells 5-8, indicating a
+  greater dependence among these cells.  On the 47 cell data, we see at least 3 groupings of columns,
+  indicating greater asymptotic dependence among these groups.
+
 % EOF

### 2021-11-06 abstract.tex
overleaf updates

@@
 {\bf We need an abstract}
-Extreme analysis
 
-Drawing on the constructive definition for the multivariate Pareto introduced in~\cite{ferreira2014} which separates the radial component of the RV from the angular, we introduce a method for inferring the distribution of the angular component.
+The constructive definition of the multivariate Pareto introduced 
+    in~\cite{ferreira2014} separates the Pareto RV $\bm{z}\in {\mathbb R}^d\not[0,1]^d$
+    into independent radial,  $r \in (1,\infty)$ and angular 
+    $\bm{y} \in {\mathbb S}_{\infty}^{d-1}$ components.
+    This paper proposes a method for inferring a density on ${\mathbb S}_{\infty}^{d-1}$
+    as the limit of a series of manifolds, ${\mathbb S}_{p}^{d-1}$.  We project a
+    distribution in ${\mathbb R}_+^d$ to ${\mathbb S}_p^{d-1}$; then as $p$ grows
+    large, we approach a density on ${\mathbb S}_{\infty}^{d-1}$.
+
+We discuss model selection criteria appropriate to ${\mathbb S}_{\infty}^{d-1}$.  We
+    use the energy score criterion introduced in \cite{gneiting2007}, incorporating a
+    novel kernel metric we develop suitable to the space.  We use this energy score, along
+    with posterior predictive loss to inform model choice in how well we recover the distribution of simulated data.  We then fit these models to real data
+
+%

### 2021-11-06 evaluation.tex
overleaf updates

@@
 In order to assess and compare the estimation of a distribution on ${\mathbb S}_\infty^{d-1}$ we consider the theory of proper scoring rules developed in \cite{gneiting2007}.
-As mentioned in Section~\ref{subsubsec:norm}, our approach does not provide a density on ${\mathbb S}_{\infty}^{d-1}$, restricting our ability to construct model selection criteria to sample-based approaches.
-In this manner, we consider two methods of scoring:  \emph{posterior predictive loss} (PPL) introduced in \cite{gelfand1998}, and \emph{energy score} (ES) introduced in \cite{gneiting2007}.
+As mentioned in Section~\ref{subsec:projgamma}, our approach does not provide a density on ${\mathbb S}_{\infty}^{d-1}$, restricting our ability to construct model selection criteria to sample-based approaches.
+We consider two scoring methods:  \emph{posterior predictive loss} (PPL) introduced in \cite{gelfand1998}, and \emph{energy score} (ES) introduced in \cite{gneiting2007}.
 
@@
   where $X_i$ is a random variable from the posterior predictive distribution for $x_i$.  The
-  scalar $k$ is a weighting factor by which we can scale the importance of goodness of fit
-  relative to precision.  In our analysis, we take the limit as $k\to\infty$, weighting both
-  equally.  Interpreting this criterion, a smaller $\text{Var}(X_{i})$ indicates a higher
-  precision, and a smaller $(\text{E}[X_{i}] - x_{i})^2$  indicates a better fit.  Thus,
-  smaller is better.  Note that this is defined for a univariate distribution.  As we are dealing
+  scalar $k$ is a weighting factor that scales the importance of goodness of fit
+  relative to precision.  In fact a small $\text{Var}(X_{i})$ indicates high
+  precision, and a small $(\text{E}[X_{i}] - x_{i})^2$  indicates a good fit.  In our analysis, we take the limit as $k\to\infty$, weighting both
+  equally. Overall, the smaller the score, the better.  Note that the PPLC score is defined for a univariate distribution.  As we are dealing
   with a multivariate distribution, we will be taking the posterior predictive loss
@@
     \label{eq:ppl2}
-      S_{\text{PPL}}\left(P,\bm{x}_i\right) = \frac{1}{d}\sum_{\ell = 1}^d\left[\text{Var}_P\left(X_{i\ell}\right) + \left(\text{E}_P\left[X_{i\ell}\right] - x_{i\ell}\right)^2\right]
+      S_{\text{PPL}}\left(P,\bm{x}_i\right) = \frac{1}{d}\sum_{\ell = 1}^d\left[\text{Var}_P\left(X_{i\ell}\right) + \left(\text{E}_P\left[X_{i\ell}\right] - x_{i\ell}\right)^2\right] .
   \end{equation}
-  Then we report the average $S$, taken over all $i$.
 
@@
     Let $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-1}$.  The 
-  rotation $P_{\jmath\ell}(\cdot)$ required to move the $\jmath$th face along the $\ell$th axis 
+  rotation $P_{\jmath\ell}(\cdot)$ required to rotate the $\jmath$th face along the $\ell$th axis 
   produces the vector $\bm{b}^\prime$, where
@@
     \label{eqn:rotation}
-    P_{\jmath\ell}(\bm{b}) = \bm{b}^{\prime}_i = 
+     \bm{b}^{\prime}_i = P_{\jmath\ell}(\bm{b}) = 
     \begin{cases}
@@
 
-Interpretation of the energy score as we have formulated it here is the same as that of posterior predictive loss: smaller is better.
+We report the average Interpretation of the energy score as we have formulated it here is the same as that of posterior predictive loss: smaller is better.
 

### 2021-11-06 methodology_projgamma.tex

--- 2021-10-27 methodology_projgamma.tex
@@
     $\alpha_\ell$ and $\beta_\ell$ are the shape and scale parameters, respectively. Using Equations (\ref{eqn:pnormt}) -- (\ref{eqn:pnormjac}) we have the joint density
-  \begin{equation*}
+  \begin{equation}
+  \label{joint}
     f(r,\bm{ y}) = \prod_{\ell = 1}^{d}
@@
       \times r^{d-1}\left[y_d +
-            {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right],
-  \end{equation*}
+            {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right].
+  \end{equation}
   Integrating out $r$ yields the resulting \emph{Projected Gamma} density
@@
       \times \frac{\Gamma({\textstyle\sum}_{\ell = 1}^d\alpha_{\ell})}{\left({\textstyle\sum}_{\ell = 1}^d
-                    \beta_{\ell}y_{\ell}\right)^{{\scriptstyle\sum_{\ell = 1}^d \alpha_{\ell}}}}
+                    \beta_{\ell}y_{\ell}\right)^{{\scriptstyle\sum_{\ell = 1}^d \alpha_{\ell}}}} ,
   \end{equation}
@@
 Building out the methods of inference for Equation~\ref{eqn:dppg}, let $n_j^{-i}$ be the number of observations in cluster $j$ not including observation $i$.
-Let $J^{-i}$ be the number of extant observations, not including any singleton that observation $i$ is in.
-Under this model, the probability of cluster membership is proportional to
+Let $J^{-i}$ be the number of extant
+% {\bf (remaining?)} observations
+clusters, not including any singleton that observation $i$ is in. Under this model, the probability of cluster membership for a given observation is proportional to
 \begin{equation*}
-    \text{P}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
+    \text{Pr}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
         n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
@@
 \end{equation*}
-Of course, if $G_0$ is not a conjugate prior for the kernel distribution, then the integral is difficult.
-Instead, using algorithm 8 from \cite{neal2000}, we draw $m$ candidate clusters, $\bm{\alpha}_j,\bm{\beta}_j$ for $j = J^{-i} + 1,\ldots, J^{-i} + m$ from $G_0$.
-Then, the probability of cluster membership is proportional to
+If $G_0$ is not a conjugate prior for the kernel, the integral in the above formula can not be obtained in closed form.
+Using Algorithm 8 from \cite{neal2000}, we draw $m$ candidate clusters, $\bm{\alpha}_j,\bm{\beta}_j$ for $j = J^{-i} + 1,\ldots, J^{-i} + m$ from $G_0$.
+Then, we sample the cluster indicator $\delta_i$ from extant or candidate clusters, where the probability of cluster membership is proportional to
 \begin{equation}
-    \text{P}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
+    \text{Pr}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
         n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
@@
 \end{equation}
-Of course, if a candidate cluster is selected, then $\delta_j = J + 1$, and $J = J^{-i} + 1$.
+If a candidate cluster is selected,
+%{\bf you first have to say that you sample from this distribution. What is the ``candidate cluster''?}
+then $\delta_i = J = J^{- i} + 1$, and the associated cluster parameters are saved.
 
 A key feature of the the projected Gamma distribution is its computational properties.
-With respect to the distributional parameters $\bm{\alpha}_i,\bm{\beta}_i$, the projected Gamma density is proportional to
+We augment  $\mathcal{PG}(\bm{y}_i\mid\bm{\alpha}_i,\bm{\beta}_i) $ by introducing a latent radial component $r_i$, for each observation.
+Using Equation \eqref{joint} we observe that the
+full conditional of $r_i$ is easy to sample from, as it is given as
+\begin{equation}
+    r_i\mid\bm{\alpha}_i,\bm{\beta}_i \sim \mathcal{G}\left(\sum_{\ell = 1}^d\alpha_{i\ell}, \sum_{\ell = 1}^d\beta_{\ell} y_{i\ell}\right).
+\end{equation}
+Moreover,  the full conditional for $\bm{\alpha}_j,\bm{\beta}_j$ is then proportional to
+
+\begin{equation}
+    f(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{Y},\bm{r},\bm{\delta},\ldots) \propto \prod_{i:\gamma_i = j}\prod_{\ell = 1}^d\mathcal{G}\left(r_iy_{i\ell}\mid\alpha_{j\ell},\beta_{j\ell}\right) \times dG_0(\bm{\alpha}_j,\bm{\beta}_j)
+\end{equation}
+
+% \begin{equation}
+%     f(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{Y},\bm{r},\bm{\delta},\ldots) \propto \prod_{i:\gamma_i = j}\prod_{\ell = 1}^d\mathcal{G}\left(r_iy_{i\ell}\mid\alpha_{j\ell},\beta_{j\ell}\right) \times dG_0(\alpha_{\ell},\beta_{\ell})
+% \end{equation}
+% {\bf I am confused here. Shouldn't it be: }
+
+% \begin{equation}
+%     f(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{Y},\bm{r},\bm{\delta},\ldots) \propto \prod_{i:\gamma_i = j}\left(\prod_{\ell = 1}^d\mathcal{G}\left(r_iy_{i\ell}\mid\alpha_{j\ell},\beta_{j\ell}\right)\right) \times dG_0(\bm{\alpha}_{i},\bm{\beta}_{i})
+% \end{equation}
+We first consider a centering distribution given by a product of independent Gammas:
 \begin{equation*}
-    \mathcal{PG}(\bm{y}\mid\bm{\alpha}_i,\bm{\beta}_i) \propto \int\prod_{\ell = 1}^d\mathcal{G}\left(ry_{i\ell}\mid\alpha_{i\ell},\beta_{i\ell}\right)r^{d-1}dr,
+    G_0(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{\xi},\bm{\tau},\bm{\zeta},\bm{\sigma}) = \prod_{\ell = 1}^d\mathcal{G}(\alpha_{j\ell}\mid \xi_{\ell},\tau_{\ell})\times\prod_{\ell = 2}^d\mathcal{G}(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}).
 \end{equation*}
-which implies that, by reintroducing the latent radial component $r_i$ via sampling from its full conditional,
-\begin{equation}
-    r_i\mid\bm{\alpha}_i,\bm{\beta}_i \sim \mathcal{G}\left(\sum_{\ell = 1}^d\alpha_{i\ell}, \sum_{\ell = 1}^d\beta_{\ell} y_{i\ell}\right),
-\end{equation}
-the likelihood for $\bm{\alpha},\bm{\beta}$ becomes separable by dimension.
-We stress this point: \emph{given} $r$, the distributional parameters $(\alpha_j,\beta_j)_{\ell}$ are independent in the likelihood.
-
-The full conditional for $\bm{\alpha}_j,\bm{\beta}_j$ is then proportional to
-\begin{equation}
-    f(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{Y},\bm{r},\bm{\delta},\ldots) \propto \prod_{i:\gamma_i = j}\prod_{\ell = 1}^d\mathcal{G}\left(r_iy_{i\ell}\mid\alpha_{j\ell},\beta_{j\ell}\right) \times dG_0(\alpha_{\ell},\beta_{\ell})
-\end{equation}
-The first form of the centering distribution we explore is a product of independent Gammas.
-\begin{equation}
-    G_0(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{\xi},\bm{\tau},\bm{\zeta},\bm{\sigma}) = \prod_{\ell = 1}^d\mathcal{G}(\alpha_{j\ell}\mid \xi_{\ell},\tau_{\ell})\times\prod_{\ell = 2}^d\mathcal{G}(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell})
-\end{equation}
-This model is completed with Gamma priors on $\xi_{\ell}$, $\tau_{\ell}$, $\zeta_{\ell}$, $\sigma_{\ell}$.
-Optionally, placing a Gamma prior on $\eta$ allows updating via the procedure outlined in \cite{escobar1995}.
-In further analysis, we refer to this model as the \emph{projected Gamma--Gamma} (PG--G) model.
-An advantage to using a Gamma centering distribution for the rate parameter is, as a conjugate prior, the rate parameters can easily be integrated out for inference on $\bm{\alpha}_j$.
-After doing so, the full conditional for $\alpha_{j\ell}$ takes the form
-\begin{equation}
+This model is completed with independent Gamma priors on $\xi_{\ell}$, $\tau_{\ell}$, $\zeta_{\ell}$, $\sigma_{\ell}$.
+We also assume a Gamma prior on $\eta$, that is updated via the procedure outlined in \cite{escobar1995}.
+We refer to this model as the \emph{projected Gamma--Gamma} (PG--G) model.
+An advantage of the PG--G model is that, thanks to conjugacy, the rate parameters can easily be integrated out for inference on $\bm{\alpha}_j$.
+Then, the full conditional for $\alpha_{j\ell}$ takes the form
+\begin{equation*}
     \pi(\alpha_{j\ell}\mid \bm{r},\bm{Y},\bm{\gamma},\xi_\ell,\tau_\ell,\zeta_\ell,\sigma_\ell) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i\ell}\right)^{\alpha_{j\ell} - 1}\alpha_{j\ell}^{\xi_\ell - 1}e^{-\tau_\ell \alpha_{j\ell}}\times\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\Gamma^{n_j}(\alpha_{j\ell})\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
-\end{equation}
+\end{equation*}
 for $\ell = 2,\ldots,d$.  For $\ell = 1$, as $\beta_{1} := 1$, the full conditional takes the form
@@
 \end{equation}
-Updating these parameters can be done via a simple Metropolis-within-Gibbs step.  In our analysis, we first transform $\alpha_{j\ell}$ to log scale, and use a normal proposal density.
-The full conditional for $\beta$ is then
+Sample of $\alpha_{j\ell}$ can thus be obtained using a Metropolis step. In our analysis, we first transform $\alpha_{j\ell}$ to the log scale, and use a normal proposal density.
+The full conditional for $\beta$ is 
 \begin{equation}
@@
 \begin{equation}
-    G_0\left(\bm{\alpha}_j,\bm{\beta}_j\mid\bm{\mu},\bm{\Sigma},\zeta,\sigma\right) = \mathcal{LN}\left(\bm{\alpha}_j\mid\bm{\mu},\bm{\Sigma}\right)\times\prod_{\ell = 2}^d\mathcal{G}\left(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}\right)
+    G_0\left(\bm{\alpha}_j,\bm{\beta}_j\mid\bm{\mu},\bm{\Sigma},\zeta,\sigma\right) = \mathcal{LN}\left(\bm{\alpha}_j\mid\bm{\mu},\bm{\Sigma}\right)\times\prod_{\ell = 2}^d\mathcal{G}\left(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}\right).
 \end{equation}
@@
 This is denoted as the \emph{projected restricted Gamma--log-normal} (PRG--LN) model.
-in the interest of brevity, we omit the inferential forms for this model, sufficing to say that updating $\bm{\alpha}_j$ is accomplished with a joint Metropolis-within-Gibbs step; the normal prior for $\mu$ is conjugate for the log-normal $\log\bm{\alpha}_j$, and updated via a Gibbs step; and an inverse Wishart prior for $\Sigma$ is again conjugate to the log-normal $\bm{\alpha}_j$, meaning it too can be updated via a Gibbs step.
+In the interest of brevity, we omit the inferential forms for this model, sufficing to say that updating $\bm{\alpha}_j$ is accomplished with a joint Metropolis-within-Gibbs step; the normal prior for $\mu$ is conjugate for the log-normal $\log\bm{\alpha}_j$, and updated via a Gibbs step; and an inverse Wishart prior for $\Sigma$ is again conjugate to the log-normal $\bm{\alpha}_j$, implying that it can also be updated via a Gibbs step.
 
-\subsubsection{Choice of Norm\label{subsubsec:norm}}
-In the extreme value case, we have defined our task on ${\mathbb S}_{\infty}^{d-1}$, the positive orthant
-  of the hypersphere defined under the $\mathcal{L}_{\infty}$ norm.  We previously described the
-  difficulty in establishing a density directly on this geometry.  However, we can build the projected
-  gamma distribution on ${\mathbb S}_p^{d-1}$ for a large $p$.  We can then project samples from the
-  posterior predictive distribution for a model built on ${\mathbb S}_p^{d-1}$ onto
-  ${\mathbb S}_{\infty}^{d-1}$. As $p$ increases, the closer the manifold will be to
-  ${\mathbb S}_{\infty}^{d-1}$, and thus the closer the projected distribution will be to an idealized
-  projected Gamma built on ${\mathbb S}_{\infty}^{d-1}$.  In testing, we tried developing models on
-  $S_p^{d-1}$ for varying levels of $p$, and found that higher $p$ produced higher performing models.
-  Additionally, the difference in model performance became more apparent as dimensionality increased.
-  We present results here for models developed on ${\mathbb S}_{10}^{d-1}$.
+%\subsubsection{Choice of Norm\label{subsubsec:norm}}
+%In the extreme value case, we have defined our task on ${\mathbb S}_{\infty}^{d-1}$, the positive orthant
+%  of the hypersphere defined under the $\mathcal{L}_{\infty}$ norm.  We previously described the
+%  difficulty in establishing a density directly on this geometry.  However, we can build the projected
+%  gamma distribution on ${\mathbb S}_p^{d-1}$ for a large $p$.  We can then project samples from the
+%  posterior predictive distribution for a model built on ${\mathbb S}_p^{d-1}$ onto
+%  ${\mathbb S}_{\infty}^{d-1}$. As $p$ increases, the closer the manifold will be to
+%  ${\mathbb S}_{\infty}^{d-1}$, and thus the closer the projected distribution will be to an idealized
+%  projected Gamma built on ${\mathbb S}_{\infty}^{d-1}$.  In testing, we tried developing models on
+%  $S_p^{d-1}$ for varying levels of $p$, and found that higher $p$ produced higher performing models.
+%  Additionally, the difference in model performance became more apparent as dimensionality increased.
+%  We present results here for models developed on ${\mathbb S}_{10}^{d-1}$.
 

### 2021-11-11 evaluation.tex
updates

@@
   relative to precision.  In fact a small $\text{Var}(X_{i})$ indicates high
-  precision, and a small $(\text{E}[X_{i}] - x_{i})^2$  indicates a good fit.  In our analysis, we take the limit as $k\to\infty$, weighting both
-  equally. Overall, the smaller the score, the better.  Note that the PPLC score is defined for a univariate distribution.  As we are dealing
+  precision, and a small $(\text{E}[X_{i}] - x_{i})^2$  indicates a good fit.  In our analysis,
+  we take the limit as $k\to\infty$, weighting both
+  equally. Overall, the smaller the score, the better.  Note that the PPLC score is defined 
+  for a univariate distribution.  As we are dealing
   with a multivariate distribution, we will be taking the posterior predictive loss
@@
   \end{equation}
+  We report the average $S_{\text{PPL}}$ taken from all observed data.
 
@@
     \end{equation}
-where $g$ is a given kernel. The score defined in Equation \eqref{eq:es} can be evaluated using samples from $P$, with the help of the law of large numbers.
-Moreover, Theorem~4 in \cite{gneiting2007}, states that if $g(\cdot,\cdot)$ is a negative definite kernel, then $\text{S}(P,\bm{x})$ is a \emph{proper} scoring rule.
-Recall that a real valued function $g$ is a negative definite kernel if it is symmetric in its arguments, and $\sum_{i=1}^n\sum_{j=1}^na_ia_jg(x_i,x_j)\leq 0$ for all positive integers $n$, and any collection $a_1,\ldots,a_n\in{\mathbb R}$ such that  $\sum_{i = 1}^na_i = 0$.
+where $g$ is a kernel function. The score defined in Equation \eqref{eq:es} can be evaluated
+  using samples from $P$, with the help of the law of large numbers.
+  Moreover, Theorem~4 in \cite{gneiting2007}, states that if $g(\cdot,\cdot)$ is a negative 
+  definite kernel, then $\text{S}(P,\bm{x})$ is a \emph{proper} scoring rule.  Recall that a 
+  real valued function $g$ is a negative definite kernel if it is symmetric in its arguments, 
+  and $\sum_{i=1}^n\sum_{j=1}^na_ia_jg(x_i,x_j)\leq 0$ for all positive integers $n$, and any 
+  collection $a_1,\ldots,a_n\in{\mathbb R}$ such that  $\sum_{i = 1}^na_i = 0$.
 
-In an Euclidean space, the conditions for a negative definite kernel are satisfied by the Euclidean distance.
-However, the Euclidean distance between two points will likely under-estimate the actual distance required to connect those points fully within ${\mathbb S}_{\infty}^{d-1}$.
-Let
+In a Euclidean space, these conditions are satisfied by the Euclidean distance. However, in
+  ${\mathbb S}_{\infty}^{d-1}$, Euclidean distance will tend to under-estimate the actual 
+  distance required to travel between two points.  Let
 \begin{equation*}
@@
 \end{equation*}
-comprise the $\ell$th \emph{face}.
-For points on the same face, Euclidean distance corresponds to the length of the shortest possible path in ${\mathbb S}_{\infty}^{d-1}$.
+comprise the $\ell$th \emph{face}.  For points on the same face, Euclidean distance 
+  corresponds to the length of the shortest possible path in ${\mathbb S}_{\infty}^{d-1}$.  
   For points on different faces, the Euclidean distance is a lower bound for that length.
 
-For a finite $p$, the shortest connecting path between two points in 
-${\mathbb S}_p^{d-1}$ is the minimum geodesic; its length satisfying 
-the definition of a distance.
-Thus its length can be used as a negative definite kernel for the purpose of defining an energy score.
-Unfortunately as $p\to\infty$ the resulting surface ${\mathbb S}_{\infty}^{d-1}$ is not differentiable, implying that routines to calculate geodesics are not readily available.
-However, as ${\mathbb S}_{\infty}^{d-1}$ is a portion of of a $d$-cube, we can borrow a result from geometry \citep{pappas1989} stating that the length of the shortest path between two points on a geometric figure corresponds to the length of a straight line drawn between the points on an appropriate unfolding, rotation, or \emph{net} of the figure from a $d$-dimensional to a $d-1$-dimensional space.
+For a finite $p$, the shortest connecting path between two points in ${\mathbb S}_p^{d-1}$
+  is the minimum geodesic; its length satisfying the definition of a distance.  Thus its
+  length can be used as a negative definite kernel for the purpose of defining an energy
+  score. Unfortunately as $p\to\infty$ the resulting surface ${\mathbb S}_{\infty}^{d-1}$
+  is not differentiable, implying that routines to calculate geodesics are not readily
+  available.  However, as ${\mathbb S}_{\infty}^{d-1}$ is a portion of of a $d$-cube, we
+  can borrow a result from geometry \citep{pappas1989} stating that the length of the
+  shortest path between two points on a geometric figure corresponds to the length of a
+  straight line drawn between the points on an appropriate unfolding, rotation, or \emph{net} of the figure from a $d$-dimensional to a $d-1$-dimensional space.
 The optimal net will have the shortest straight line between the points, as long as that line is fully contained within such net.
@@
 \begin{prop}\label{prop:g}
-For points $a,b \in {\mathbb S}_{\infty}^{d-1}$ the kernel defined as  
-  \begin{equation}
-    g(\bm{a},\bm{b}) = \begin{cases}
-        \pnorm{\bm{b}-\bm{a}}{2} &\text{ if }\argmax_{\ell}\bm{a} = \argmax_{\ell}\bm{b}\\
-        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} &\text{ otherwise}
-    \end{cases} ,
-  \end{equation}
-  where $\bm{c}$ is in the intersection of the faces corresponding to $\bm{a}$ and $\bm{b}$, and minimizes $g(\bm{a},\bm{b})$, is negative definite.
+Let $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-1}$, for $\ell, \jmath \in \{1, \ldots , d\}$. Define $g$ as
+\[  
+    g(\bm{a},\bm{b}) = \min_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
+        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
+\]
+Then $g$  is negative definite kernel.
 \end{prop}
@@
 {\em Proof:}
-When $\bm{a}$ and $\bm{b}$ are on the same face, then $g$ is the Euclidean distance, which is a 
-  negative definite kernel.  When $\bm{a}$ and $\bm{b}$ are on separate
-  faces, we have that $g(\bm{a}, \bm{b}) = g(\bm{b}, \bm{a})$ due to the uniqueness of $\bm{c}$ and the symmetry of the Euclidean distance.
-Additionally, given $n$ and  collection $\alpha_1,\ldots,\alpha_n\in{\mathbb R}$ such that  $\sum_{i = 1}^n\alpha_i = 0$, then
+When $\bm{a}$ and $\bm{b}$ are on the same face, then $\ell = \jmath$, and $g$ is the Euclidean distance between $\bm{a}$ and $\bm{b}$, which is a  negative definite kernel.
+When $\bm{a}$ and $\bm{b}$ are on separate
+  faces, let 
+  \[\bm{d} = 
+    \argmin_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
+        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
+  \]
+  We have that $g(\bm{a}, \bm{b}) = g(\bm{b}, \bm{a})$ due to the uniqueness of $\bm{d}$ and the symmetry of the Euclidean distance.
+Additionally, given $n$ and  a collection $\alpha_1,\ldots,\alpha_n\in{\mathbb R}$ such that  $\sum_{i = 1}^n\alpha_i = 0$, then,
   \begin{equation*}
     \begin{aligned}
-      \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j g(\bm{x}_1,\bm{x}_2) &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j \bigg[\pnorm{\bm{c} - \bm{x}_1}{2} + \pnorm{\bm{x}_2 - \bm{c}}{2}\bigg]\\
-      &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{c} - \bm{x}_1}{2} + \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{x}_2 - \bm{c}}{2} \leq0,
+      \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j g(\bm{x}_1,\bm{x}_2) &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j \bigg[\pnorm{\bm{d} - \bm{x}_1}{2} + \pnorm{\bm{x}_2 - \bm{d}}{2}\bigg]\\
+      &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{d} - \bm{x}_1}{2} + \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{x}_2 - \bm{d}}{2} \leq0,
     \end{aligned}
@@
 
-Evaluating $g$ as defined in Proposition \ref{prop:g} requires the solution of a $d-2$-dimensional 
-  optimization problem in order to obtain $\bm{c}$.  However, a computationally easier \makenote{rewrite sentence} method exists.
+Evaluating $g$ as defined in Proposition \ref{prop:g} requires the solution of a $(d-2)$-dimensional 
+  optimization problem.  The following proposition provides a straightforward solution.
 \begin{prop}
-    Let $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-1}$.  The 
-  rotation $P_{\jmath\ell}(\cdot)$ required to rotate the $\jmath$th face along the $\ell$th axis 
+    Let $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-1}$, for $\ell, \jmath \in \{1, \ldots , d\}$.  For $\ell\neq \jmath$, the 
+  transformation $P_{\jmath\ell}(\cdot)$ required to rotate the $\jmath$th face along the $\ell$th axis 
   produces the vector $\bm{b}^\prime$, where
@@
 {\em Proof:}
-Note that for 
-  $\bm{c} \in {\mathbb C}_{\jmath\ell}^{d-2} = {\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}$,
-  $P_{\jmath\ell}(\bm{c}) = \bm{c}$.  Then
-  \begin{equation*}
-     \pnorm{\bm{b} - \bm{c}}{2} = \pnorm{P_{\jmath\ell}(\bm{b} - \bm{c})}{2} = \pnorm{\bm{b}^{\prime} - \bm{c}}{2},
-  \end{equation*}
-  as rotations preserve Euclidean norms.  Then,
+Notice that for 
+  $\bm{c} \in {\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}$,
+     $\pnorm{\bm{b} - \bm{c}}{2} =  \pnorm{\bm{b}^{\prime} - \bm{c}}{2}$.
+{\bf This is true just because you do the math, the invariance of Euclidean norms is not needed.} Then, recalling the definition of $\bm{d}$ in the proof of Proposition \ref{prop:g}, we have that
   \begin{equation*}
     \begin{aligned}
-    g(\bm{a}, \bm{b}) &= \pnorm{\bm{a} - \bm{c}}{2} + \pnorm{\bm{b} - \bm{c}}{2}\\
-    &= \pnorm{\bm{a} - \bm{c}}{2} + \pnorm{\bm{b}^{\prime} - \bm{c}}{2} = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}
+    g(\bm{a}, \bm{b}) &= \pnorm{\bm{a} - \bm{d}}{2} + \pnorm{\bm{b} - \bm{d}}{2}\\
+    &= \pnorm{\bm{a} - \bm{d}}{2} + \pnorm{\bm{b}^{\prime} - \bm{d}}{2} = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}.
     \end{aligned}
   \end{equation*}
-  when $\bm{c} \in \overrightarrow{\bm{a}\bm{b}^{\prime}}$, as the lower bound of the triangle inequality.
-We have already declared $\bm{c}\in{\mathbb C}_{\jmath\ell}^{d-2}$, so it must reside at $\overrightarrow{\bm{a}\bm{b}^{\prime}}\cap{\mathbb C}_{\jmath\ell}^{d-2}$.
-Because the straight line between $\bm{a}$ and $\bm{b}^{\prime}$ passes through this point without further analysis, the optimization of $\bm{c}$ is accomplished \emph{en~passant}.$\hfill\Box$
+  The last equality is due to the fact that $\bm{a}$ and $\bm{b}'$ belong to the same hyperplane.
+  $\hfill\Box$
 
-We report the average Interpretation of the energy score as we have formulated it here is the same as that of posterior predictive loss: smaller is better.
-
+We report the average $S_{\text{ES}}$ taken across all observed data.  In this form, the interpretation of $S_{\text{ES}}$ is similar to that of $S_{\text{PPL}}$: the smaller the score, the better.
 % EOF

### 2021-11-11 methodology_projgamma.tex
updates

@@
 Moreover,  the full conditional for $\bm{\alpha}_j,\bm{\beta}_j$ is then proportional to
-
 \begin{equation}
-    f(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{Y},\bm{r},\bm{\delta},\ldots) \propto \prod_{i:\gamma_i = j}\prod_{\ell = 1}^d\mathcal{G}\left(r_iy_{i\ell}\mid\alpha_{j\ell},\beta_{j\ell}\right) \times dG_0(\bm{\alpha}_j,\bm{\beta}_j)
+    f(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{Y},\bm{r},\bm{\delta},\ldots) \propto \prod_{i:\gamma_i = j}\prod_{\ell = 1}^d\mathcal{G}\left(r_iy_{i\ell}\mid\alpha_{j\ell},\beta_{j\ell}\right) \times dG_0(\bm{\alpha}_j,\bm{\beta}_j).
 \end{equation}
-
-% \begin{equation}
-%     f(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{Y},\bm{r},\bm{\delta},\ldots) \propto \prod_{i:\gamma_i = j}\prod_{\ell = 1}^d\mathcal{G}\left(r_iy_{i\ell}\mid\alpha_{j\ell},\beta_{j\ell}\right) \times dG_0(\alpha_{\ell},\beta_{\ell})
-% \end{equation}
-% {\bf I am confused here. Shouldn't it be: }
-
-% \begin{equation}
-%     f(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{Y},\bm{r},\bm{\delta},\ldots) \propto \prod_{i:\gamma_i = j}\left(\prod_{\ell = 1}^d\mathcal{G}\left(r_iy_{i\ell}\mid\alpha_{j\ell},\beta_{j\ell}\right)\right) \times dG_0(\bm{\alpha}_{i},\bm{\beta}_{i})
-% \end{equation}
 We first consider a centering distribution given by a product of independent Gammas:
@@
 This is denoted as the \emph{projected restricted Gamma--log-normal} (PRG--LN) model.
-In the interest of brevity, we omit the inferential forms for this model, sufficing to say that updating $\bm{\alpha}_j$ is accomplished with a joint Metropolis-within-Gibbs step; the normal prior for $\mu$ is conjugate for the log-normal $\log\bm{\alpha}_j$, and updated via a Gibbs step; and an inverse Wishart prior for $\Sigma$ is again conjugate to the log-normal $\bm{\alpha}_j$, implying that it can also be updated via a Gibbs step.
+In the interest of brevity, we omit the inferential forms for this model, sufficing to say that the Gamma prior for $\beta_{j\ell}$ means it can again be integrated out, improving mixing on sampling $\bm{\alpha}_j$ with a joint Metropolis step; the normal prior for $\mu$ is conjugate for the log-normal $\log\bm{\alpha}_j$, and updated via a Gibbs step; and an inverse Wishart prior for $\Sigma$ is again conjugate to the log-normal $\bm{\alpha}_j$, implying that it can also be updated via a Gibbs step.
 

### 2021-11-11 results.tex
updates

@@
 
-\begin{figure}[ht!]
+\begin{figure}[ht]
     \centering
@@
 
+\begin{figure}[ht]
+    \centering
+    \caption{Conditional survival curves for selected dimensions, conditioning on all other dimensions at greater than 90th percentile (fitted)\label{fig:condsurv1d}}
+    \includegraphics{./images/condsurv_1d}
+\end{figure}
+
+\begin{figure}[htb]
+    \centering
+    \caption{Pairwise conditional survival curves for selected dimensions, conditioning on all other dimensions at greater than 90th percentile (fitted)\label{fig:condsurv2d}}
+    \includegraphics[height=5in, width=5in]{./images/condsurv_2d}
+\end{figure}
+
+In Figure~\ref{fig:condsurv1d}, using the PRG--G model on the 8-dimensional data, we implement 
+    the conditional survival curve defined in Equation~\ref{eqn:condsurv}, for one dimension, 
+    conditioned on all other dimensions being greater than their (fitted) $90$th percentile.  
+    On the left, the survival curve is in original units, whereas the right is the same plot 
+    given in standardized units.  
+    Figure~\ref{fig:condsurv2d} implements the same conditional survival equation in pairwise form,
+    conditioning on all other dimensions.  Here we see another representation of extremal dependence,
+    affecting the shape of the conditional survival curves.  The top pair represents the joint survival function between grid zones 1 and 2, which we saw from Figure~\ref{fig:chi_ij} to exhibit strong
+    extremal dependence.  We see that again here, where the joint survival surface in standardized form
+    is strongly convex.
+The bottom pair represents the joint survival surface between grid zones 2 and 8, which exhibited low extremal dependence.
+We see the effects of that, in the concave pairwise survival surface.
 % EOF

### 2021-11-17 abstract.tex
updates

--- 2021-10-27 abstract.tex
@@
 {\bf We need an abstract}
-Extreme analysis
 
-Drawing on the constructive definition for the multivariate Pareto introduced in~\cite{ferreira2014} which separates the radial component of the RV from the angular, we introduce a method for inferring the distribution of the angular component.
+The constructive definition of the multivariate Pareto separates the Pareto RV into
+    a radial component, following a univariate Pareto distribution, and independently
+    an angular component, defined on the surface of the positive orthant of the unit 
+    hypercube.  In this paper, we 
+    propose a method for inferring the distribution of that angular component.  We  
+    identify its support as the limit of the unit $p$--norm
+    spheres, and describe a means by which we project a distribution onto those spheres.
+    We then introduce a family of \emph{projected Gamma} distributions, defined by the
+    power of the unit norm sphere on which they have support.  This projected Gamma family
+    serves as a building block to a Dirichlet process mixture of projected Gammas model,
+    which we ultimately present.
+
+We discuss model scoring methods appropriate to this space, and develop a kernel metric
+    to be used with the \emph{energy score} criterion for evaluating models.  We compare
+    these results with those obtained using the posterior predictive loss criterion.  
+    We compute model scores
+    on various iterations of the aforementioned model, fitted to simulated data.  We then
+    apply the model to the \emph{integrated vapor transport} (IVT)---a dataset describing 
+    the quantity of water vapor in the atmosphere, used for identifying atmospheric 
+    rivers---to demonstrate the model on \emph{real} data.
+
+% EOF

### 2021-11-30 abstract.tex
updates

--- 2021-10-27 abstract.tex
@@
-{\bf We need an abstract}
-Extreme analysis
+We consider a constructive definition of the multivariate Pareto that factorizes the 
+random vector into a radial component and an independent angular component. The former
+follows a univariate Pareto distribution. The latter is defined on the surface of the 
+positive orthant of the unit  hypercube.  In this paper, we  propose a method for
+inferring the distribution of the angular component.  We   identify its support 
+as the limit of the positive orthant of the unit $p$--norm spheres, 
+and describe a method to project a distribution onto those spheres. To define a 
+distribution in the $p$--norm spheres, we introduce a family of projected Gamma 
+distributions. This family serves as a building block for a flexible family of 
+distributions obtained as  a Dirichlet process mixture of projected Gammas. For 
+model comparison and assessment, we discuss model scoring methods appropriate 
+to distributions on the unit hypercube.  In particular, we develop a kernel metric 
+that produces a proper scoring rule.   We present a simulation study to compare 
+different modeling choices using the proposed  scoring rules. We then apply our 
+approach to study the extreme values of the  magnitude of the integrated vapor 
+transport (IVT), that describes the rate of flow of moisture in the atmosphere 
+along the coast of California for the years of 1979 through 2020.
 
-Drawing on the constructive definition for the multivariate Pareto introduced in~\cite{ferreira2014} which separates the radial component of the RV from the angular, we introduce a method for inferring the distribution of the angular component.
+% EOF

### 2021-11-30 evaluation.tex
updates

@@
 
-To reduce the computations needed to calculate the energy score we define a kernel as {\bf Please write a clearer description here.} our kernel an upper bound on distance between points on separate faces as the length of the path from the source point, to an optimal point on the intersection between faces, to the destination point.
+To reduce the computations needed to calculate the energy score we define a kernel as 
+in the following proposition.
 \begin{prop}\label{prop:g}
@@
   faces, let 
-  \[\bm{d} = 
+  \[\bm{e} = 
     \argmin_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
@@
   \]
-  We have that $g(\bm{a}, \bm{b}) = g(\bm{b}, \bm{a})$ due to the uniqueness of $\bm{d}$ and the symmetry of the Euclidean distance.
+  We have that $g(\bm{a}, \bm{b}) = g(\bm{b}, \bm{a})$ due to the uniqueness of $\bm{e}$ and the symmetry of the Euclidean distance.
 Additionally, given $n$ and  a collection $\alpha_1,\ldots,\alpha_n\in{\mathbb R}$ such that  $\sum_{i = 1}^n\alpha_i = 0$, then,
@@
     \begin{aligned}
-      \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j g(\bm{x}_1,\bm{x}_2) &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j \bigg[\pnorm{\bm{d} - \bm{x}_1}{2} + \pnorm{\bm{x}_2 - \bm{d}}{2}\bigg]\\
-      &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{d} - \bm{x}_1}{2} + \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{x}_2 - \bm{d}}{2} \leq0,
+      \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j g(\bm{x}_1,\bm{x}_2) &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j \bigg[\pnorm{\bm{e} - \bm{x}_1}{2} + \pnorm{\bm{x}_2 - \bm{e}}{2}\bigg]\\
+      &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{e} - \bm{x}_1}{2} + \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{x}_2 - \bm{e}}{2} \leq0,
     \end{aligned}
@@
   $\bm{c} \in {\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}$,
-     $\pnorm{\bm{b} - \bm{c}}{2} =  \pnorm{\bm{b}^{\prime} - \bm{c}}{2}$.
-{\bf This is true just because you do the math, the invariance of Euclidean norms is not needed.} Then, recalling the definition of $\bm{d}$ in the proof of Proposition \ref{prop:g}, we have that
+     $\pnorm{\bm{b} - \bm{c}}{2} =  \pnorm{\bm{b}^{\prime} - \bm{c}}{2}$. Then, recalling the definition of 
+     $\bm{e}$ in the proof of Proposition \ref{prop:g}, we have that
   \begin{equation*}
     \begin{aligned}
-    g(\bm{a}, \bm{b}) &= \pnorm{\bm{a} - \bm{d}}{2} + \pnorm{\bm{b} - \bm{d}}{2}\\
-    &= \pnorm{\bm{a} - \bm{d}}{2} + \pnorm{\bm{b}^{\prime} - \bm{d}}{2} = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}.
+    g(\bm{a}, \bm{b}) &= \pnorm{\bm{a} - \bm{e}}{2} + \pnorm{\bm{b} - \bm{e}}{2}\\
+    &= \pnorm{\bm{a} - \bm{e}}{2} + \pnorm{\bm{b}^{\prime} - \bm{e}}{2} = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}.
     \end{aligned}

### 2021-11-30 methodology_projgamma.tex
updates

@@
   {\bf Instead of the old Figure 1, it would be nice to add a figure where you illustrate this procedure. Give it a try.}
+
+\begin{figure}[ht]
+    \centering
+  \begin{subfigure}[b]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{./images/p_sphere}
+    \caption{${\mathbb S}_p^1$ for $p = 1,\ldots, 10$\label{fig:psphere}}
+  \end{subfigure}
+  %
+  \begin{subfigure}[b]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{./images/p_project}
+    \caption{Projection of data on ${\mathbb S}_{\infty}^1$ onto ${\mathbb S}_p^1$\label{fig:pproject}}
+  \end{subfigure}
+\end{figure}
 
@@
 This is denoted as the \emph{projected restricted Gamma--log-normal} (PRG--LN) model.
-In the interest of brevity, we omit the inferential forms for this model, sufficing to say that the Gamma prior for $\beta_{j\ell}$ means it can again be integrated out, improving mixing on sampling $\bm{\alpha}_j$ with a joint Metropolis step; the normal prior for $\mu$ is conjugate for the log-normal $\log\bm{\alpha}_j$, and updated via a Gibbs step; and an inverse Wishart prior for $\Sigma$ is again conjugate to the log-normal $\bm{\alpha}_j$, implying that it can also be updated via a Gibbs step.
+In the interest of brevity, we omit the inferential forms for this model, sufficing to say that the Gamma prior for $\beta_{j\ell}$ means it can again be integrated out, improving mixing on sampling $\bm{\alpha}_j$ with a joint Metropolis step; the normal prior for $\mu$ is conjugate for the log-normal $\bm{\alpha}_j$, and sampled via a Gibbs step; and an inverse Wishart prior for $\Sigma$ is again conjugate to the log-normal $\bm{\alpha}_j$, implying that it can also be sampled via a Gibbs step.
 

### 2021-11-30 results.tex
updates

--- 2021-10-30 results.tex
@@
 \subsection{Simulations\label{subsec:simulated}}
-We have introduced this projected gamma distribution existing on $\mathcal{S}_{\infty}^{p-1}$.  As a
-  means of evaluating how well we can recover a distribution under the proposed model, we have
-  generated a series of datasets on $\mathcal{S}_{\infty}^{d-1}$, for $d$ ranging between 3 and 20.
-  We generated each dataset from a mixture of projected Gammas, with the number of mixture components
+To evaluate our proposed approach for angular measure estimation 
+we consider simulated datasets on $\mathcal{S}_{\infty}^{d-1}$, for 
+$d$ ranging between 3 and 20. We generated each dataset from a mixture of projected Gammas, with the number of mixture components
   ranging between 3 and 12.  The generation procedure is detailed in Algorithm~\ref{algo:simulated}.
@@
   \end{algorithm}
-  The choice of generation routine for $\alpha = \alpha_0 + \alpha_1$ where
+Each one of the shape parameters $\alpha$ were generated 
+$\alpha = \alpha_0 + \alpha_1$ where
   $\alpha_0 \sim \text{Unif}(\alpha_0\mid 0,4)$, $\alpha_1\sim \text{Gamma}(\alpha_1\mid 1,1)$.
-  The choice of generation routine for $\beta$ is $\beta\sim\text{Unif}(\beta\mid 0.25, 2.5)$.
-  This choice was made to ensure that $\beta$ does not approach 0 in the generation routine.
+The scale parameters $\beta$ were obtained as $\beta\sim\text{Unif}(\beta\mid 0.25, 2.5)$.
 
@@
 \begin{figure}[ht]
-  \begin{subfigure}[b]{0.45\textwidth}
+\begin{center}
+  \begin{subfigure}[b]{0.48\textwidth}
     \includegraphics[width=\textwidth]{./images/simulation_ppl}
@@
   %
-  \begin{subfigure}[b]{0.45\textwidth}
+  \begin{subfigure}[b]{0.48\textwidth}
     \includegraphics[width=\textwidth]{./images/simulation_es}
@@
   \end{subfigure}
-  \caption{Simulation PPL (left) and ES (left) versus column count, by mixture component count.}
+  \caption{Scores for the simulated datasets calculated using different models:
+  GEN - baseline; PG - mixture of projected gammas; PRG - mixture of projected restricted gamma; PRG--LN - mixture of projected restricted gammas with log-normal prior.}
+\end{center}
 \end{figure}
 
-In Figure~\ref{fig:simes}, we fit DP Mixtures of the various models presented and plot the resulting
-  energy score against the number of columns, grouped by the number of mixture components used in
-  dataset generation.  The \emph{Gen} model refers to a dataset generated
-  in the same manner, from the same parameters as was used to create the original dataset.  Against
-  this generative dataset, we compare DP Mixtures of projected gamma, projected restricted gamma, and
-  projected restricted gamma with a multivariate log-normal prior.  We observe that the projected
+Figure~\ref{fig:simes}, shows two sets of panels that illustrate the comparison
+between four different models used to fit the simulated data. We produced 16 
+different data sets, one for each of four different numbers of mixture components 
+and four different dimensions. We considered three different models: DP mixtures 
+of projected gamma; Projected restricted gamma; and projected restricted gamma 
+with a multivariate log-normal prior. We calculated $S^\infty_{\text{PPL}}$ and 
+$S_{\text{ES}}$ using predictive samples from each of fitted models. To provide 
+a comparative baseline, we calculated the scores using samples from generated 
+directly using the same model and parameters that produced the simulated 
+observations. The results are denoted as \emph{Gen} in the figure. 
+We observe that the projected
   restricted Gamma model dominates in most situations, but the difference between projected restricted
-  Gamma and the other gamma-based models tends to shrink as the number of columns increases.
-  Alternatively, that difference appears to grow as the number of mixture components increases.
+  Gamma and the other gamma-based models tends to shrink as the number of dimensions increases.
+  Alternatively, that difference appears to grow as the number of mixture components increases. The results from both scoring criteria are pretty comparable.
 
@@
 % \end{figure}
-
-The posterior predictive loss criterion in Figure~\ref{fig:simppl} tells a similar story: the base
-  model is inadequate for describing a mixture; and the DP mixture of projected restricted gammas
-  performs the best of the presented models in most cases.  Interestingly, under both metrics we see
-  the probit-normal model has the best performance when the number of columns is low.  That is
-  unexpected behavior, as its performance very quickly degrades as the number of columns increases.
 
@@
   In this sense, the argument for overfitting is less clear.
+  {\bf This paragraph is not very clear. Please make an effort to tell the story more clearly.}
 
 \subsection{Integrated Vapor Transport\label{subsec:ivt}}
-The \emph{integrated vapor transport} (IVT) tracks the total water volume in a column of air over a given
-  area \citep{ralph2017}.  IVT is the means by which we track the formation and path of atmospheric 
-  rivers---elongated areas of high local concentration of water vapor in the atmosphere that moves with
-  wind patterns.  Extreme behavior in IVT can lead to extreme precipitation events, causing flooding; 
-  but in California, those same events are responsible for the bulk of precipitation that inland 
-  California receives---water necessary for agriculture and household consumption.  Thus it becomes 
-  necessary to manage precipitation to prevent inundation and support agriculture.  To this end, 
-  understanding the dependence structure of extreme events in the IVT is useful---for instance, for 
-  hydrological engineers in planning the capacity of groundwater systems to capture precipitation before 
-  wasteful and destructive flooding occurs.  Our data tracks daily average values for the IVT, dividing 
-  the coast of California into grid cells, with two spatial resolutions of 8 and 47 cells, respectively 
-  \citep{guan2015}.
+The \emph{integrated vapor transport} (IVT) is a two component vector 
+that tracks the flow of the total water volume in a column of air 
+over a given area \citep{ralph2017}.  IVT is increasingly used in the study of atmospheric rivers because of its direct 
+relationship with orographically induced precipitation
+\citep{neiman2009water}. Atmospheric rivers (AR) are elongated areas of high 
+local concentration of water vapor in the atmosphere that transport water
+from the tropics around the world. AR can cause extreme 
+precipitation,  something that usually associated with very large values
+of the IVT magnitude over a whole geographical area. In spite of this, AR
+are fundamental for the water supply of areas like California. Thus the
+importance of understanding the extreme behavior of IVT, including 
+extreme tail dependence. We consider datasets that correspond to IVT
+estimated at two different spatial resolutions. The coarse resolution dataset
+is obtained from the European Centre for Medium-Range Weather Forecasts
+(ECMWF) Interim reanalysis (ERA-Interim) \citep{berrisford2011atmospheric,dee2011era}. 
+The high resolution dataset corresponds to the latest ECMWF 
+observational product, ERA5  \citep{hersbach2020era5}.
+
+Our data corresponds to daily average values for the IVT magnitude
+along the coast of California.  The ERA-Interim data used covers the time period 
+1979 through 2014 (37 years) omitting leap days, and provides 8 grid cells.  The ERA5 data
+covers the time period 1979 through 2019 (42 years) with the same restriction, and provides 
+47 grid cells.  This gives us the opportunity to illustrate the performance of our method 
+in multivariate settings of very different dimensions. {\bf Here it will be nice to include a plot showing the location of the grid cells on a map in each one of the cases.}
 
@@
    represents the radial component, and $\bm{v}_i$ the angular component.  The declustering
-   portion is relevant if data is correlated in time.\label{algo:processing}}
+   portion is relevant for data correlated in time.\label{algo:processing}}
 \end{algorithm}
 
-Fitting our models to this data requires some pre-processing. First, we subset the data to the rainy
-  season, which appears from November to March.  The marginal distributions of the
-  IVT data appear naturally log-normal, which falls into the domain of attraction of a Gumbel
-  distribution.  Given that, we can apply a high threshold, and exceedances over that threshold can
-  be modelled as Generalized Pareto.  Our modelling begins with that assessment.  As estimating
-  the Pareto parameters is not yet our focus in this analysis, we choose to apply the threshold
-  using the empirical CDF.  That is, for a given $t$, let $b_{t,\ell} = \hat{F}_{\ell}^{-1}(1 - t^{-1})$.  For
-  this analysis, we set $t = 20$, indicating the marginal $95$ percentile.  The other parameters of the
-  generalized Pareto---the scale parameter $\alpha_{\ell}$ and the extremal index $\chi_{\ell}$---are set via
-  maximum likelihood.
+Fitting our models to the IVT data requires some pre-processing. First, we subset the data to the rainy
+  season, which in California runs roughly from November to March.
+Following the approach described in Section \ref{sec:methodology} we estimate the shape and scale parameters of a univariate GP, in each dimension, using maximum likelihood.
+We set the threshold in each dimensions $\ell$ as  $b_{t,\ell} = \hat{F}_{\ell}^{-1}(1 - t^{-1})$, where $\hat{F}$ is the empirical CDF and $t=20$, that corresponds to the $95$ percentile.
+We then use the transformation in Equation \eqref{eqn:standardization} to standardize the observations.
+  Dividing each standardized observation by its $\mathcal{L}_{\infty}$ 
+  norm, we obtain a projection onto $\mathcal{S}_{\infty}^{d-1}$.
+As the data correspond to a daily time series, the observations are temporally correlated.
+For each group of consecutive standardized vectors $z_i$ such  that  $\inorm{z_i} > 1$, we retain only the vector
+  with the largest $\mathcal{L}_{\infty}$ norm.  The complete procedure is outlined in
+  Algorithm~\ref{algo:processing}.  After subsetting the ERA-Interim data to the rainy season we 
+  have $5587$ observations. This reduces to $511$ observations after processing and declustering.  
+  For the ERA5 data, after subsetting we have $6342$ observations, which reduces to $532$ observations after 
+  processing and declustering.  We fit the PG--G, PRG--G, PG--LN, and PRG--LN models to both datasets.
+  {\bf At this point it would be nice to include a figure with a plot of the data after all the pre-processing. You show a couple of panels for the ERA-Iterim and a couple of the ERA5.}
+  \makenote{Do you mean marginal histograms, or, like, a couple selected ternary scatter-plots?}
 
-After the thresholding and maximum likelihood estimation of the parameters of the Pareto, we scale
-  the data to the standard multivariate Pareto following Equation~\ref{eqn:standardization}.  
-  Dividing each standardized observation by its $\mathcal{L}_{\infty}$ norm, we project the 
-  standardized data onto $\mathcal{S}_{\infty}^{d-1}$. Data in sequence represents observations
-  in time, which for this application are temporally correlated.  As such, we
-  choose to \emph{decluster} the observations, such that, by observing a sequence of observations
-  $\bm{ z}$ for which $\inorm{z_i} > 1$ for each observation in sequence, we keep only that observation
-  with the greatest observed $\mathcal{L}_{\infty}$ norm.  The complete procedure is outlined in
-  Algorithm~\ref{algo:processing}.  In the 8-cell data, after subsetting to the rainy season we 
-  have $5587$ observations, which reduces to $511$ observations after processing and declustering.  
-  In the 47 cell data, after subsetting we have $6342$ observations, which reduces to $532$ after 
-  processing and declustering.  We fit PG--G, PRG--G, PG--LN, and PRG--LN models to both datasets.
-
-\begin{table}[ht]
+\begin{table}[t]
   \centering
-  \caption{Model comparison metrics: Posterior Predictive Loss and Energy Score criteria from fitted
-    models against the IVT data.  Lower is better.
-  \label{tab:dev}}
+  \caption{Model comparison metrics: Posterior~Predictive~Loss~($S_{\text{PPL}}$) and 
+    Energy~Score~($S_{\text{ES}}$) criteria from fitted models against the IVT data.  
+    Lower is better. \label{tab:dev}}
   \include{./tables/table_dev}
@@
 
-In Table~\ref{tab:dev} we see the same model preference towards the projected restricted 
-  gamma models that we saw in Figures~\ref{fig:simes}.  The unrestricted gamma model is 
-  penalized much more strongly on real data than we saw with our simulation.  We also see
-  the restricted gamma model with the log-normal prior performing significantly better 
-  than the restricted gamma model with the gamma prior on the higher dimensional data, a reversal
-  of what we saw on the low dimensional data and indeed from what we saw in the simulation study.
+Table~\ref{tab:dev} shows the values of the estimated PPL and energy scores for the different models considered.
+We observe that, as in the case of the simulation study shown in Figures~\ref{fig:simes}, the preferred model is the projected restricted
+  gamma models.  Penalization for the unrestricted gamma model is 
+  stronger for the real data than for the simulated data. We also observe that
+  the restricted gamma model with the log-normal prior performs significantly better 
+  than the restricted gamma model with the gamma prior on the higher dimensional data. This is the opposite of the result obtained in the simulation example.
 
-\begin{figure}[ht!]
+\begin{figure}[ht]
     \centering
-    \caption{Pairwise extremal dependence coefficients for IVT data, from fitted PRG--G.  
-    The left corresponds to the 8-cell data, the right the 47-cell data.\label{fig:chi_ij}}
+    \caption{Pairwise extremal dependence coefficients for IVT data using the PRG--G model.  
+    The left panel corresponds to the ERA-Interim data, the right panel to the ERA5 data.\label{fig:chi_ij}}
     \begin{minipage}{.49\textwidth}
@@
 
-Figure~\ref{fig:chi_ij} shows our implementation of Equation~\ref{eqn:chi_ij}, plotting the 
-  pairwise extremal dependence coefficients for the IVT data computed from PRG--G models 
-  fitted to the 8 (left) and 47 (right) cell data.  
-  That we see a higher extremal dependence between neighboring columns is not unexpected, as
-  neighboring columns correspond to neighboring cells.  We recognize that pairwise asymptotic
-  dependence coefficients tell a limited story, as a particular dependence may structure may include
-  more than two columns.   We can, however, glean some information from the patterns that emerge in
-  two dimensions.  On the 8-cell data, we see a stronger association between cells 5-8, indicating a
-  greater dependence among these cells.  On the 47 cell data, we see at least 3 groupings of columns,
-  indicating greater asymptotic dependence among these groups.
+We consider an exploration of the pairwise extremal dependence using Monte Carlo estimates of the 
+coefficients in  Equation~\ref{eqn:chi_ij}. For this we use samples obtained fromn the PRG--G model.
+Figure~\ref{fig:chi_ij} provides a graphical analysis of the results. 
+The coefficients achieve values between {\bf say the range of values} for the ERA-Interim data and between {\bf say the range here} for the ERA5 data.
+Distance between locations mostly determines the strength of the pairwise asymptotic dependence.
+The highest coefficients are {\bf XXX for locations YYY} in the ERA-Interim data and {\bf XXX for locations YYY} in the ERA5 data.
+Clearly, pairwise asymptotic
+  dependence coefficients tell a limited story, as a particular dependence may include
+  more than two locations.   We can, however, glean some information from the patterns that emerge in
+  two dimensions.  For the ERA-Interim data, we observe a possible cluster between cells 5-8, indicating a
+  strong dependence among these cells.  Analogously, for the ERA5 data, we observe three possible groups  of locations.
 
+\begin{figure}[ht]
+    \centering
+    \caption{Conditional survival curves for selected locations, conditioning on all other dimensions at greater than 90th percentile (fitted)\label{fig:condsurv1d}.
+The left panel uses original units.
+Right panel uses standardized units.}
+    \includegraphics{./images/condsurv_1d}
+\end{figure}
+
+\begin{figure}[htb]
+    \centering
+    \caption{Pairwise conditional survival curves for selected locations, conditioning on all other dimensions at greater than 90th percentile (fitted)\label{fig:condsurv2d} {\bf The numbers in the plots do not correspond to the number in the text}}
+    \includegraphics[height=5in, width=5in]{./images/condsurv_2d}
+\end{figure}
+
+Figure~\ref{fig:condsurv1d} shows, for the ERA-Interim data under the PRG--G model,  
+    the conditional survival curve defined in Equation~\ref{eqn:condsurv}, for one dimension, 
+    conditioned on all other dimensions being greater than their (fitted) $90$th percentile. 
+    Figure~\ref{fig:condsurv2d} presents the bi-variate conditional survival function,
+    conditioning on all other dimensions.  These results illustrate quantitatively how extremal dependence
+    affects the shape of the conditional survival curves.  The two top panels represent the joint survival function 
+    between grid locations 1 and 2, which are shown in Figure~\ref{fig:chi_ij} to exhibit strong
+    extremal dependence.  We observe that the joint survival surface 
+    is strongly convex.  The bottom panels represent the joint survival surface between grid locations 2 and 8, 
+    which exhibited low extremal dependence.  In this case the shape of the contours tend to be concave, quite different from the shapes observed in the top panels.
 % EOF

### 2021-11-30 results.tex

@@
 47 grid cells.  This gives us the opportunity to illustrate the performance of our method 
-in multivariate settings of very different dimensions. {\bf Here it will be nice to include a plot showing the location of the grid cells on a map in each one of the cases.}
+in multivariate settings of very different dimensions. Figure~\ref{fig:gridlocs} provides a visual representation of the area these grid cells represent.
+
+\begin{figure}[ht]
+    \begin{center}
+    \begin{subfigure}[b]{0.48\textwidth}
+        \centering 
+        \includegraphics[height=3in]{./images/erai_grid}
+    \end{subfigure}
+    %
+    \begin{subfigure}[b]{0.48\textwidth}
+        \centering
+        \includegraphics[height=3in]{./images/era5_grid}
+    \end{subfigure}
+    \caption{Grid cell locations for ERA-Interim (left) and ERA5 (right).\label{fig:gridlocs}}
+    \end{center}
+\end{figure}
 
@@
 
-Table~\ref{tab:dev} shows the values of the estimated PPL and energy scores for the different models considered.
+Table~\ref{tab:dev} shows the values of the estimated posterior predictive loss and energy scores for the different models considered.
 We observe that, as in the case of the simulation study shown in Figures~\ref{fig:simes}, the preferred model is the projected restricted
@@
   Figure~\ref{fig:chi_ij} provides a graphical analysis of the results. 
-The coefficients achieve values between {\bf say the range of values} for the ERA-Interim data and between {\bf say the range here} for the ERA5 data.
-Distance between locations mostly determines the strength of the pairwise asymptotic dependence.
-The highest coefficients are {\bf XXX for locations YYY} in the ERA-Interim data and {\bf XXX for locations YYY} in the ERA5 data.
-Clearly, pairwise asymptotic
+  The coefficients achieve values between $0.286$ and $0.759$ for the ERA-Interim data and 
+  between $0.181$ and $0.840$ for the ERA5 data.  The greater range in dependence scores observed 
+  with the ERA5 data speaks to the greater granularity of the ERA5 data. Distance between locations 
+  mostly determines the strength of the pairwise asymptotic dependence. The highest coefficients 
+  are $0.759$ for locations 4 and 5 in the ERA-Interim data and
+  $0.840$ for locations 1 and 2 in the ERA5 data.  Clearly, pairwise asymptotic
   dependence coefficients tell a limited story, as a particular dependence may include

### 2021-12-01 results.tex
Updates from Overleaf
ccchanges

@@
     \centering
-    \caption{Pairwise extremal dependence coefficients for IVT data using the PRG--G model.  
-    The left panel corresponds to the ERA-Interim data, the right panel to the ERA5 data.\label{fig:chi_ij}}
-    \begin{minipage}{.49\textwidth}
-      \centering
-      \includegraphics[width=0.99\linewidth]{./images/chi_ij_8}
-    \end{minipage}
-    \begin{minipage}{.49\textwidth}
-      \centering
-      \includegraphics[width=0.99\linewidth]{./images/chi_ij_46}
-    \end{minipage}
+    \caption{Pairwise extremal dependence coefficients for IVT data using the PRG--G model.\label{fig:chi_ij}}
+    \includegraphics[width=0.9\textwidth]{./images/chi_ij_c}
 \end{figure}
@@
     affects the shape of the conditional survival curves.  The two top panels represent the joint survival function 
-    between grid locations 1 and 2, which are shown in Figure~\ref{fig:chi_ij} to exhibit strong
+    between grid locations 4 and 5, which are shown in Figure~\ref{fig:chi_ij} to exhibit strong
     extremal dependence.  We observe that the joint survival surface 
-    is strongly convex.  The bottom panels represent the joint survival surface between grid locations 2 and 8, 
+    is strongly convex.  The bottom panels represent the joint survival surface between grid locations 1 and 5, 
     which exhibited low extremal dependence.  In this case the shape of the contours tend to be concave, quite different from the shapes observed in the top panels.

### 2021-12-12 methodology_projgamma.tex
sdf

@@
   For step (iv) we use a Bayesian approach that is implemented using a purposely developed Markov chain Monte Carlo that is described in the next section.
-  {\bf Instead of the old Figure 1, it would be nice to add a figure where you illustrate this procedure. Give it a try.}
 
@@
 \end{equation*}
-If $G_0$ is not a conjugate prior for the kernel, the integral in the above formula can not be obtained in closed form.
-Using Algorithm 8 from \cite{neal2000}, we draw $m$ candidate clusters, $\bm{\alpha}_j,\bm{\beta}_j$ for $j = J^{-i} + 1,\ldots, J^{-i} + m$ from $G_0$.
-Then, we sample the cluster indicator $\delta_i$ from extant or candidate clusters, where the probability of cluster membership is proportional to
+If $G_0$ is not a conjugate prior for the kernel, the integral in the above formula may not be
+available in closed form. Using Algorithm 8 from \cite{neal2000}, by Monte Carlo integration, drawing  $m$ candidate
+clusters, $\bm{\alpha}_j,\bm{\beta}_j$ for $j = J^{-i} + 1,\ldots, J^{-i} + m$ from $G_0$. 
+Then, we sample the cluster indicator $\delta_i$ from extant or candidate clusters, where
+the probability of cluster membership is proportional to
 \begin{equation}

### 2021-12-12 results.tex
sdf

@@
     \centering
-    \caption{Pairwise conditional survival curves for selected locations, conditioning on all other dimensions at greater than 90th percentile (fitted)\label{fig:condsurv2d} {\bf The numbers in the plots do not correspond to the number in the text}}
+    \caption{Pairwise conditional survival curves for selected locations, conditioning on all other dimensions at greater than 90th percentile (fitted)\label{fig:condsurv2d}}
     \includegraphics[height=5in, width=5in]{./images/condsurv_2d}

### 2021-12-30 abstract.tex
sfdf

--- 2021-11-06 abstract.tex
@@
-{\bf We need an abstract}
+We consider a constructive definition of the multivariate Pareto that factorizes the 
+random vector into a radial component and an independent angular component; the former
+following a univariate Pareto distribution, and the latter defined on the surface of the 
+positive orthant of the unit hypercube.  In this paper, we  propose a method for
+inferring the distribution of the angular component.  We identify its support 
+as the limit of the positive orthants of the unit $p$--norm spheres.
+To this effect, we introduce a projected gamma family of
+distributions defined as the projection of a vector of independent gamma random variables
+onto the $p$--norm sphere.  This family serves as a building block for a flexible family
+of distributions obtained as a Dirichlet process mixture of projected gammas.  For 
+model assessment and comparison, we discuss model scoring methods appropriate 
+to distributions on the unit hypercube.  In particular, working with the energy score
+criterion, we develop a kernel metric appropriate to the hypercube that produces a 
+proper scoring rule.  We then present a simulation study to compare 
+different modeling choices using the proposed scoring rules.  Finally we apply our 
+approach to describe the dependence structure of the extreme values of the magnitude 
+of the integrated vapor  transport (IVT), a variable that describes the rate of flow 
+of moisture in the atmosphere  along the coast of California for the years of 1979 
+through 2020.  We find a clear but heterogeneous geographical dependence.
+%We present the results 
+%of this analysis to describe the dependence structure of extreme events in IVT using 
+%pairwise extremal dependence coefficients---a summary measure of jointly extreme behavior.
+%We also present conditional survival curves---the survival function for some dimensions
+%conditional on other dimensions.
 
-The constructive definition of the multivariate Pareto introduced 
-    in~\cite{ferreira2014} separates the Pareto RV $\bm{z}\in {\mathbb R}^d\not[0,1]^d$
-    into independent radial,  $r \in (1,\infty)$ and angular 
-    $\bm{y} \in {\mathbb S}_{\infty}^{d-1}$ components.
-    This paper proposes a method for inferring a density on ${\mathbb S}_{\infty}^{d-1}$
-    as the limit of a series of manifolds, ${\mathbb S}_{p}^{d-1}$.  We project a
-    distribution in ${\mathbb R}_+^d$ to ${\mathbb S}_p^{d-1}$; then as $p$ grows
-    large, we approach a density on ${\mathbb S}_{\infty}^{d-1}$.
-
-We discuss model selection criteria appropriate to ${\mathbb S}_{\infty}^{d-1}$.  We
-    use the energy score criterion introduced in \cite{gneiting2007}, incorporating a
-    novel kernel metric we develop suitable to the space.  We use this energy score, along
-    with posterior predictive loss to inform model choice in how well we recover the distribution of simulated data.  We then fit these models to real data
-
-%
+% EOF

### 2021-12-30 conclusion.tex
sfdf

--- 2021-09-13 conclusion.tex
@@
+\section{Conclusion\label{sec:conclusion}}
 In this paper, we have built upon the definition of the multivariate Pareto described in \cite{ferreira2014}
-  to establish a useful parametric representation of its dependence structure, through describing
-  the distribution of its angular component; it being supported on the positive orthant of the unit
-  hypersphere under the $\mathcal{L}_{\infty}$ norm, $\mathcal{S}_{\infty}^{d-1}$.  We then demonstrated an
-  inherent difficulty in establishing a distribution supported in this space.  To build that representation,
-  we presented a distribution based on a product of independent Gammas, projected onto a supporting manifold
-  $\mathcal{S}_{p}^{d-1}$, which will converge to $\mathcal{S}_{\infty}^{d-1}$ as $p\to\infty$.  We then
-  established a means of model comparison on $\mathcal{S}_{\infty}^{d-1}$ through the use of a negative
-  definite kernel defined in that space.
+  to establish a useful parametric representation of its dependence structure, through
+  the distribution of its angular component, which is supported on the positive orthant of the unit
+  hypersphere under the $\mathcal{L}_{\infty}$ norm.  Due to the an
+  inherent difficulty of obtaining a distribution with support on ${\mathbb S}^{d-1}_\infty$
+  we projected distributions based on products of independent gammas, 
+  onto the manifold
+  ${\mathbb S}_{p}^{d-1}$, and the projected samples of the resulting
+  probability distribution onto ${\mathbb S}_{\infty}^{d-1}$.  
+  As ${\mathbb S}_{p}^{d-1}$ converges to ${\mathbb S}_{\infty}^{d-1}$ as $
+  \to\infty$ we expect the resampling to be efficient for large enough
+  $p$. In fact, our exploration of the simulated and real data indicate
+  that the procedure is robust to the choice of moderately large values
+  of $p$.
 
-We used integrated vapor transport data---a measure of water in the atmosphere---as a motivating example for
-  our model.  This data subsets California into a series of grid cells, and provides daily IVT values for
-  each cell.  We used our model to develop measures of pairwise extremal dependence between grid cells, which
-  could be of interest to researchers, actuaries, hydrologists and even farmers.  We additionally developed
-  conditional survival functions---$\text{P}\left[Z_{\ell} > z_{\ell}\mid Z_{\not\ell} > z_{\not\ell}\right]$---which
-  provide a more complete understanding of the dependence structure.
+Our simulations explored simulated data from
+    mixtures of projected Gammas with varying degrees of complexity---both in terms of dimensionality and 
+    in number of mixture components.  From this, we learned that the additional flexibility 
+    offered by varying the $\beta$ term---the rate parameter of the Gamma distribution---does not result 
+    in increased model fidelity in a mixture setting.  Within the tested range, between 3 and 20 
+    dimensions, we observed that the additional information that could be captured using the log--normal 
+    centering distribution did not translate to additional model fidelity.  However, when we looked at 
+    real data, with 47 dimensions, there was an effect.
+
+The computations performed in this paper were performed on a desktop computer with an AMD Ryzen 5950X processor.
+  The program is largely single-threaded, so computation time is not dependent on available core count.  In each
+  case, we run the MCMC chain for 50\,000 iterations, with a burn-in of 20\,000 samples.  Fitting the PG--G model
+  on the ERA5 dataset took approximately 30 minutes.  Work is in progress to optimize the code, and explore
+  parallelization where possible.  We are also exploring alternative computational approaches that will make it
+  feasible to tackle very high dimensional problems. In fact, to elaborate on the study of IVT, there is a need 
+  to consider several hundreds, if not thousands, of grid cells over the Pacific Ocean in order to obtain a 
+  good decsription of atmospheric events responsible for large storm activity over California.
+
+% {\bf ( describe
+% the computer)} and in each case we run the MCMC for {\bf XXX} iterations
+% with a burn in of {\bf XXX} samples. For the large simulated data and the
+% ERA5 dataset the process of fitting the model took {\bd YYY} minutes. Work is in progress to optimize the
+% code to perform the MCMC {\bf (insert some comment here if you think it is appropriate)}. We are also 
+% exploring alternative computational approaches that will make it feasible to tackle very high dimensional
+% problems. In fact, to elaborate on the study of IVT, there is a need to consider several hundreds, if not
+% thousands, of grid cells over the Pacific Ocean in order to obtain a good decsription of atmospheric events responsible for large storm activity over California.
+
+%We used integrated vapor transport data---a measure of water in the atmosphere---from the ERA-Interim and ERA5
+%    models as a motivating example for our model.  This data subsets California into a series of grid cells,
+%    and provides daily IVT values for each cell.  For the ERA-Interim data, having determined from
+%    Table~\ref{tab:dev} that the optimal model presented was PRG--G, we used this fitted model to develop
+%    $\Chi$, a measure of pairwise extremal dependence between grid cells.  This provides a useful summary
+%    statistic for describing the relationship between two sites.  Granted this relationship is pairwise, it
+%    can not represent the full dependence structure; so recognizing these limits, we also presented 
+%    conditional survival 
+%    functions---$\text{P}\left[Z_{\ell} > z_{\ell}\mid Z_{\neg\ell} > z_{\neg\ell}\right]$---which
+%    provide a more complete understanding of the dependence structure of $\bm{Z}$.

### 2021-12-30 introduction.tex
updates
sfdf

@@
 
-During the last decade or so much work has been done in the exploration of the definition and properties of an
+During the last decade or so, much work has been done in the exploration of the definition and properties of an
   appropriate generalization of the univariate GP distribution 
@@
 Assume the same conditions of Proposition \ref{ppchi}. 
-Let $\alpha \subset \{1, \ldots ,p\}$ be a collections of indexes. 
+Let $\alpha \subset \{1, \ldots ,d\}$ be a collections of indexes. 
 Then     
-\begin{equation} \label{eqn:condsurv2df}
-    \text{Pr}\left[\bigcap_{\ell \in \alpha} Z_\ell > z_\ell \right|\left. \bigcap_{\ell\not\in\alpha} Z_\ell > z_\ell\right] =
+\begin{equation} \label{eqn:condsurv}
+    \mathrm{Pr}\left[\bigcap_{\ell \in \alpha} Z_\ell > z_\ell \right|\left. \bigcap_{\ell\not\in\alpha} Z_\ell > z_\ell\right] =
       \frac{\expect{\bigwedge_{k = 1}^d 1\wedge \frac{V_k}{z_k}}}{\expect{\bigwedge_{k \not\in\alpha}1\wedge\frac{V_k}{z_k}}}.

### 2021-12-30 methodology_projgamma.tex
sfdf

@@
     \centering
+    \caption{The positive orthant of the $p$-norm sphere.}
   \begin{subfigure}[b]{0.47\textwidth}
@@
 \subsubsection{Inference for the projected Gamma mixture model}
-The Dirichlet process mixture model groups observations into stochastically assigned \emph{clusters}.
+The Dirichlet process mixture model groups observations into stochastically assigned clusters.
 Building out the methods of inference for Equation~\ref{eqn:dppg}, let $n_j^{-i}$ be the number of observations in cluster $j$ not including observation $i$.
@@
 \end{equation}
-Sample of $\alpha_{j\ell}$ can thus be obtained using a Metropolis step. In our analysis, we first transform $\alpha_{j\ell}$ to the log scale, and use a normal proposal density.
+Samples of $\alpha_{j\ell}$ can thus be obtained using a Metropolis step. In our analysis, we first transform $\alpha_{j\ell}$ to the log scale, and use a normal proposal density.
 The full conditional for $\beta$ is 

### 2021-12-30 results.tex
sfdf

@@
+\begin{figure}[htb]
+\begin{center}
+    \centering
+  \caption{Posterior predictive loss (Left) and energy score (right) for various models on simulated data.  Gen indicates baseline.\label{fig:simpples}}
+  \begin{subfigure}[b]{0.48\textwidth}
+    \centering
+    \includegraphics[height=2in, width=\textwidth]{./images/simulation_ppl}
+  \end{subfigure}
+  %
+  \begin{subfigure}[b]{0.48\textwidth}
+    \centering
+    \includegraphics[height=2in, width=\textwidth]{./images/simulation_es}
+  \end{subfigure}
+\end{center}
+\end{figure}
+
 \section{Data illustrations\label{sec:results}}
@@
 \subsection{Simulations\label{subsec:simulated}}
+
 To evaluate our proposed approach for angular measure estimation 
@@
   \begin{algorithm}[ht]
+    \small 
     \caption{Simulated Angular Dataset Generation Routine\label{algo:simulated}}
@@
 
-\begin{figure}[ht]
-\begin{center}
-  \begin{subfigure}[b]{0.48\textwidth}
-    \includegraphics[width=\textwidth]{./images/simulation_ppl}
-    \caption{Posterior Predictive Loss\label{fig:simppl}}
-  \end{subfigure}
-  %
-  \begin{subfigure}[b]{0.48\textwidth}
-    \includegraphics[width=\textwidth]{./images/simulation_es}
-    \caption{Energy Score\label{fig:simes}}
-  \end{subfigure}
-  \caption{Scores for the simulated datasets calculated using different models:
-  GEN - baseline; PG - mixture of projected gammas; PRG - mixture of projected restricted gamma; PRG--LN - mixture of projected restricted gammas with log-normal prior.}
-\end{center}
-\end{figure}
-
-Figure~\ref{fig:simes}, shows two sets of panels that illustrate the comparison
+Figure~\ref{fig:simpples}, shows two sets of panels that illustrate the comparison
 between four different models used to fit the simulated data. We produced 16 
@@
 $S_{\text{ES}}$ using predictive samples from each of fitted models. To provide 
-a comparative baseline, we calculated the scores using samples from generated 
+a comparative baseline, we calculated the scores using samples generated 
 directly using the same model and parameters that produced the simulated 
@@
 
-% \begin{figure}[ht]
-%   \caption{Simulation posterior predictive loss versus column count, by mixture component count.\label{fig:simppl}}
-%   \centering
-%   \includegraphics[width=0.8\textwidth]{./images/simulation_ppl}
-% \end{figure}
-
-We frequently see all of the gamma-based mixture models accomplish a lower energy score than the generative
-  distribution.  As \cite{nunez2019} noted, the projected Gamma model is very flexible and can generate
-  multimodal distributions from a single mixture component.  When we allow a mixture model, we allow
-  the possibility to break those modes into two different mixture components.  This will tend to
-  result in a better energy score, as posterior predictive replicates for a given observation in a local
-  model will tend to be concentrate in that local mode rather than being spread between multiple
-  modes.  As for the dominance of projected restricted Gamma, By specifying $\beta_{\ell} := 1$ for all
-  $l$, if there were multi-modal mixture components, we are in effect breaking the mixture components
-  into separate modes. As for whether this represents \emph{overfitting}, If we had reason to
-  believe that the generative distribution for real data followed this form, then such a case could
-  be made.  However, we are using projected Gamma as a parametric stand-in for an unknown distribution.
-  In this sense, the argument for overfitting is less clear.
-  {\bf This paragraph is not very clear. Please make an effort to tell the story more clearly.}
+% We frequently see all of the gamma-based mixture models accomplish a lower energy score
+%     than the generative distribution.  As \cite{nunez2019} noted, the projected Gamma model
+%     is flexible and can generate multimodal distributions from a single mixture component.
+%     When we allow a mixture model, we allow the possibility to break those modes into 
+%     different mixture components.  This will tend to result in a better energy score, as 
+%     posterior predictive replicates for a given observation in a local model will tend to 
+%     be concentrated in that local mode rather than being spread between multiple modes.  
+%     As for the dominance of projected restricted Gamma, By specifying $\beta_{\ell} := 1$ 
+%     for all $l$, if there were multi-modal mixture components, we are in effect breaking 
+%     the mixture components into separate modes. As for whether this represents 
+%     \emph{overfitting}, If we had reason to believe that the generative distribution for
+%     real data followed this form, then such a case could be made.  However, we are using 
+%     projected Gamma as a parametric stand-in for an unknown distribution. In this sense, 
+%     the argument for overfitting is less clear.
+%     {\bf This paragraph is not very clear. Please make an effort to tell the story more clearly.}
 
 \subsection{Integrated Vapor Transport\label{subsec:ivt}}
+
+\begin{figure}[htb]
+    \begin{center}
+    \caption{Grid cell locations for ERA-Interim (left) and ERA5 (right).\label{fig:gridlocs}}
+    \begin{subfigure}[b]{0.48\textwidth}
+        \centering 
+        \includegraphics[height=2.5in]{./images/erai_grid}
+    \end{subfigure}
+    %
+    \begin{subfigure}[b]{0.48\textwidth}
+        \centering
+        \includegraphics[height=2.5in]{./images/era5_grid}
+    \end{subfigure}
+    \end{center}
+\end{figure}
+
 The \emph{integrated vapor transport} (IVT) is a two component vector 
@@
 from the tropics around the world. AR can cause extreme 
-precipitation,  something that usually associated with very large values
+precipitation,  something that is usually associated with very large values
 of the IVT magnitude over a whole geographical area. In spite of this, AR
@@
 along the coast of California.  The ERA-Interim data used covers the time period 
-1979 through 2014 (37 years) omitting leap days, and provides 8 grid cells.  The ERA5 data
-covers the time period 1979 through 2019 (42 years) with the same restriction, and provides 
-47 grid cells.  This gives us the opportunity to illustrate the performance of our method 
+1979 through 2014 (37 years) omitting leap days, and eight grid cells that correspond to the coast of California.  The ERA5 data
+covers the time period 1979 through 2019 (42 years) with the same restriction, and 
+47 grid cells for the coast of California.  This gives us the opportunity to illustrate the performance of our method 
 in multivariate settings of very different dimensions. Figure~\ref{fig:gridlocs} provides a visual representation of the area these grid cells represent.
 
-\begin{figure}[ht]
-    \begin{center}
-    \begin{subfigure}[b]{0.48\textwidth}
-        \centering 
-        \includegraphics[height=3in]{./images/erai_grid}
-    \end{subfigure}
-    %
-    \begin{subfigure}[b]{0.48\textwidth}
-        \centering
-        \includegraphics[height=3in]{./images/era5_grid}
-    \end{subfigure}
-    \caption{Grid cell locations for ERA-Interim (left) and ERA5 (right).\label{fig:gridlocs}}
-    \end{center}
-\end{figure}
-
-\begin{algorithm}[h]
+\begin{algorithm}[htb]
+    \small
   \KwResult{$\bm{ r},\bm{v} : r_i \sim \text{Pareto}(1)$, $\bm{ v}_i \in {\mathbb S}_{\infty}^{d-1}$}
@@
   \For{$i = 1,\ldots,n$}{
-    Define $z_{i,\ell} = \left(1 + \xi_{\ell}\frac{x_{i,\ell} - b_{t,\ell}}{a_{\ell}}\right)_{+}^{1/\xi_{\ell}}$\\
-    Define $r_i = \pnorm{\bm{ z}_i}{\infty}$, $\bm{ v}_i = \frac{\bm{ z}_i}{\pnorm{\bm{ z}_i}{\infty}}$\\
+    Define $z_{i,\ell} = \left(1 + \xi_{\ell}\frac{x_{i,\ell} - b_{t,\ell}}{a_{\ell}}\right)_{+}^{1/\xi_{\ell}}$; $\;\;\;$ then $r_i = \pnorm{\bm{ z}_i}{\infty}$, $\;\;\bm{ v}_i = \frac{\bm{ z}_i}{\pnorm{\bm{ z}_i}{\infty}}$\\
     }
@@
 
+\begin{figure}[ht]%{l}{0.45\textwidth}
+    \centering
+    \caption{Pairwise plots from ERA-Interim data after transformation and projection to ${\mathbb S}_{\infty}^{7}$.
+Down the diagonal are marginal kernel densities, with two-dimensional histograms on the off-diagonal.
+In those plots, red indicates a higher density.
+All data is between 0 and 1.\label{fig:erai_data}}
+    \includegraphics[width=.8\linewidth]{./images/data_transformed}
+\end{figure}
+
 Fitting our models to the IVT data requires some pre-processing. First, we subset the data to the rainy
-  season, which in California runs roughly from November to March.
-Following the approach described in Section \ref{sec:methodology} we estimate the shape and scale parameters of a univariate GP, in each dimension, using maximum likelihood.
-We set the threshold in each dimensions $\ell$ as  $b_{t,\ell} = \hat{F}_{\ell}^{-1}(1 - t^{-1})$, where $\hat{F}$ is the empirical CDF and $t=20$, that corresponds to the $95$ percentile.
-We then use the transformation in Equation \eqref{eqn:standardization} to standardize the observations.
-  Dividing each standardized observation by its $\mathcal{L}_{\infty}$ 
-  norm, we obtain a projection onto $\mathcal{S}_{\infty}^{d-1}$.
-As the data correspond to a daily time series, the observations are temporally correlated.
-For each group of consecutive standardized vectors $z_i$ such  that  $\inorm{z_i} > 1$, we retain only the vector
-  with the largest $\mathcal{L}_{\infty}$ norm.  The complete procedure is outlined in
-  Algorithm~\ref{algo:processing}.  After subsetting the ERA-Interim data to the rainy season we 
-  have $5587$ observations. This reduces to $511$ observations after processing and declustering.  
-  For the ERA5 data, after subsetting we have $6342$ observations, which reduces to $532$ observations after 
-  processing and declustering.  We fit the PG--G, PRG--G, PG--LN, and PRG--LN models to both datasets.
-  {\bf At this point it would be nice to include a figure with a plot of the data after all the pre-processing. You show a couple of panels for the ERA-Iterim and a couple of the ERA5.}
-  \makenote{Do you mean marginal histograms, or, like, a couple selected ternary scatter-plots?}
+  season, which in California runs roughly from November to March.  Following the approach described 
+  in Section \ref{sec:methodology} we estimate the shape and scale parameters of a univariate GP, in each
+  dimension, using maximum likelihood. We set the threshold in each dimensions $\ell$ as  
+  $b_{t,\ell} = \hat{F}_{\ell}^{-1}(1 - t^{-1})$, where $\hat{F}$ is the empirical CDF and $t=20$, 
+  that corresponds to the $95$ percentile. We then use the transformation in Equation
+  \eqref{eqn:standardization} to standardize the observations.  Dividing each standardized observation
+  by its $\mathcal{L}_{\infty}$ norm, we obtain a projection onto $\mathcal{S}_{\infty}^{d-1}$. As 
+  the data correspond to a daily time series, the observations are temporally correlated.  For each
+  group of consecutive standardized vectors $z_i$ such  that  $\inorm{z_i} > 1$, we retain only the
+  vector with the largest $\mathcal{L}_{\infty}$ norm.  The complete procedure is outlined in
+  Algorithm~\ref{algo:processing}.
 
-\begin{table}[t]
+After subsetting the ERA-Interim data to the rainy season we 
+  have $5587$ observations. After the processing and declustering described in
+  Algorithm~\ref{algo:processing}, this number reduces to $511$ observations. A pairwise plot 
+  of the transformed data after processing and declustering is presented in 
+  Figure~\ref{fig:erai_data}.  From this, we note that the marginal densities display strong
+  similarities, with a large spike near 0 and a small spike near 1. A value of 1 in a particular 
+  axis indicates that the standardized threshold exceedance was largest in that dimension.  
+  The off-diagonal plots correspond to pairwise density plots.  We observe that some site pairs, such 
+  as $(1,2)$, $(7,8)$, and especially $(4,5)$ have the bulk of their data concentrated in a 
+  small arc along the $45^{\circ}$, while other site combinations such as $(3,6)$, $(2,7)$, or
+  $(1,8)$ the data is split, favoring one side or the other of the $45^{\circ}$ line. For the 
+  ERA5 data, after subsetting we have $6342$ observations, which reduces to $532$ observations 
+  after processing and declustering. We fit the PG--G, PRG--G, PG--LN, and PRG--LN models to 
+  both datasets.
+%  {\bf At this point it would be nice to include a figure with a plot of the data after all the pre-processing. You show a couple of panels for the ERA-Iterim and a couple of the ERA5.}
+%  \makenote{Do you mean marginal histograms, or, like, a couple selected ternary scatter-plots?}
+
+\begin{table}[htb]
   \centering
@@
 Table~\ref{tab:dev} shows the values of the estimated posterior predictive loss and energy scores for the different models considered.
-We observe that, as in the case of the simulation study shown in Figures~\ref{fig:simes}, the preferred model is the projected restricted
+We observe that, as in the case of the simulation study shown in Figures~\ref{fig:simpples}, the preferred model is the projected restricted
   gamma models.  Penalization for the unrestricted gamma model is 
@@
   between $0.181$ and $0.840$ for the ERA5 data.  The greater range in dependence scores observed 
-  with the ERA5 data speaks to the greater granularity of the ERA5 data. Distance between locations 
-  mostly determines the strength of the pairwise asymptotic dependence. The highest coefficients 
-  are $0.759$ for locations 4 and 5 in the ERA-Interim data and
+  with the ERA5 data versus ERA-Interim speaks to the greater granularity of the ERA5 data--distance between locations is a strong contributor to the strength of the 
+  pairwise asymptotic dependence. The highest coefficients are $0.759$ for 
+  locations 4 and 5 in the ERA-Interim data and
   $0.840$ for locations 1 and 2 in the ERA5 data.  Clearly, pairwise asymptotic
   dependence coefficients tell a limited story, as a particular dependence may include
-  more than two locations.   We can, however, glean some information from the patterns that emerge in
-  two dimensions.  For the ERA-Interim data, we observe a possible cluster between cells 5-8, indicating a
-  strong dependence among these cells.  Analogously, for the ERA5 data, we observe three possible groups  of locations.
+  more than two locations.   We can, however, glean some information from the patterns that
+  emerge in two dimensions.  For the ERA-Interim data, we observe a possible cluster 
+  between cells 5-8, indicating a strong dependence among these cells.  Analogously, for
+  the ERA5 data, we observe three possible groups of locations.
 
-\begin{figure}[ht]
+\begin{figure}[htb]
     \centering
@@
     which exhibited low extremal dependence.  In this case the shape of the contours tend to be concave, quite different from the shapes observed in the top panels.
+
+We also explored what effect the power of the projection would have---that is, which $p$-norm
+    sphere the data was projected on to for model fitting.  Using the simulated data, coming
+    from a mixture of projected Gammas, we were unable to observe an effect from $p$ when 
+    looking at a range from 1 to 15 for $p$, on our simulated data.  However, when we repeated 
+    this analysis on the IVT data, we were able to observe a drop in the energy score associated 
+    with higher $p$, with diminishing returns, as tested up to $p=15$.  The relative performance
+    between $p=10$, where the modelling for our presented analysis was conducted, and $p=15$, 
+    was not significant.
 % EOF

### 2022-01-05 abstract.tex
update

@@
 We consider a constructive definition of the multivariate Pareto that factorizes the 
-random vector into a radial component and an independent angular component; the former
+random vector into a radial component and an independent angular component; The former
 following a univariate Pareto distribution, and the latter defined on the surface of the 

### 2022-01-05 conclusion.tex
update

@@
 In this paper, we have built upon the definition of the multivariate Pareto described in \cite{ferreira2014}
-  to establish a useful parametric representation of its dependence structure, through
+  to establish a useful representation of its dependence structure, through
   the distribution of its angular component, which is supported on the positive orthant of the unit
-  hypersphere under the $\mathcal{L}_{\infty}$ norm.  Due to the an
+  hypersphere under the $\mathcal{L}_{\infty}$ norm.  Due to the 
   inherent difficulty of obtaining a distribution with support on ${\mathbb S}^{d-1}_\infty$
-  we projected distributions based on products of independent gammas, 
-  onto the manifold
-  ${\mathbb S}_{p}^{d-1}$, and the projected samples of the resulting
-  probability distribution onto ${\mathbb S}_{\infty}^{d-1}$.  
+  our method projects distributions, supported on the positive quadrant and
+  based on products of independent gammas,  onto the manifold
+  ${\mathbb S}_{p}^{d-1}$. Samples of the resulting probability distribution 
+  are then projected onto ${\mathbb S}_{\infty}^{d-1}$.  
   As ${\mathbb S}_{p}^{d-1}$ converges to ${\mathbb S}_{\infty}^{d-1}$ as $
-  \to\infty$ we expect the resampling to be efficient for large enough
+  p\to\infty$, we expect the resampling to be efficient for large enough
   $p$. In fact, our exploration of the simulated and real data indicate
@@
 
-Our simulations explored simulated data from
+We explored the behavior of the proposed model using simulated data from
     mixtures of projected Gammas with varying degrees of complexity---both in terms of dimensionality and 
@@
     in increased model fidelity in a mixture setting.  Within the tested range, between 3 and 20 
-    dimensions, we observed that the additional information that could be captured using the log--normal 
-    centering distribution did not translate to additional model fidelity.  However, when we looked at 
-    real data, with 47 dimensions, there was an effect.
+    dimensions, we observed that the additional information provided by a log--normal 
+    centering distribution did not translate to additional model fidelity.  However, for 
+    real data, with 47 dimensions, we did observe an effect.
 
-The computations performed in this paper were performed on a desktop computer with an AMD Ryzen 5950X processor.
+The computations in this paper were performed on a desktop computer with an AMD Ryzen 5950X processor.
   The program is largely single-threaded, so computation time is not dependent on available core count.  In each
@@
   to consider several hundreds, if not thousands, of grid cells over the Pacific Ocean in order to obtain a 
-  good decsription of atmospheric events responsible for large storm activity over California.
+  good description of atmospheric events responsible for large storm activity over California.
 

### 2022-01-05 evaluation.tex
sfdf
update

@@
     \label{eq:ppl}
-      S_{\text{PPL}}^k\left(P,x_i\right) = \text{Var}_P\left(X_i\right) + \frac{k}{k+1}\left(\text{E}_P\left[X_{i}\right] - x_i\right)^2
+      S_{\text{PPL}}^k\left(P,x_i\right) = \text{Var}_P\left[X_i\right] + \frac{k}{k+1}\left(\text{E}_P\left[X_{i}\right] - x_i\right)^2
   \end{equation}
@@
     \label{eq:ppl2}
-      S_{\text{PPL}}\left(P,\bm{x}_i\right) = \frac{1}{d}\sum_{\ell = 1}^d\left[\text{Var}_P\left(X_{i\ell}\right) + \left(\text{E}_P\left[X_{i\ell}\right] - x_{i\ell}\right)^2\right] .
+      S_{\text{PPL}}\left(P,\bm{x}_i\right) = \frac{1}{d}\sum_{\ell = 1}^d\left[\text{Var}_P\left[X_{i\ell}\right] + \left(\text{E}_P\left[X_{i\ell}\right] - x_{i\ell}\right)^2\right] .
   \end{equation}
@@
     \label{eq:es}
-    \text{S}_{\text{ES}}\left(P, \bm{x}_i\right) =  \text{E}_p g\left(\bm{X}_i, \bm{x}_i\right) -
-                                    \frac{1}{2}\text{E}_p g\left(\bm{X}_i,\bm{X}_i^{\prime}\right),
+    \text{S}_{\text{ES}}\left(P, \bm{x}_i\right) =  \text{E}_p \left[g\left(\bm{X}_i, \bm{x}_i\right)\right] -
+                                    \frac{1}{2}\text{E}_p \left[g\left(\bm{X}_i,\bm{X}_i^{\prime}\right)\right],
     \end{equation}
@@
 \]
-Then $g$  is negative definite kernel.
+Then $g$  is a negative definite kernel.
 \end{prop}
@@
   optimization problem.  The following proposition provides a straightforward solution.
-\begin{prop}
+\begin{prop}\label{prop:rot}
     Let $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-1}$, for $\ell, \jmath \in \{1, \ldots , d\}$.  For $\ell\neq \jmath$, the 
@@
 
-We report the average $S_{\text{ES}}$ taken across all observed data.  In this form, the interpretation of $S_{\text{ES}}$ is similar to that of $S_{\text{PPL}}$: the smaller the score, the better.
+Proposition \ref{prop:rot} provides a computational efficient way to evaluate the proper scoring rule $S_{\text{ES}}$ defined on ${\mathbb S}_\infty^{d-1}$, for each observation.
+For the purpose of model assessment and comparison, we report the average $S_{\text{ES}}$ taken across all observed data.
+In this form, the interpretation of $S_{\text{ES}}$ is similar to that of $S_{\text{PPL}}$: The smaller the score, the better.
 % EOF

### 2022-01-05 introduction.tex
update

@@
   to a multivariate setting.  To mention some of the papers on the topic, the work of
-  \cite{rootzen2006} defines the generalized Pareto distribution, with further analysis on these classes
+  \citep{rootzen2006} defines the generalized Pareto distribution, with further analysis on these classes
   of distributions presented in \cite{falk2008} and \cite{michel2008}.  A recent review of the state

### 2022-01-05 methodology_projgamma.tex
update

@@
 We first consider a centering distribution given by a product of independent Gammas:
-\begin{equation*}
+\begin{equation}
     G_0(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{\xi},\bm{\tau},\bm{\zeta},\bm{\sigma}) = \prod_{\ell = 1}^d\mathcal{G}(\alpha_{j\ell}\mid \xi_{\ell},\tau_{\ell})\times\prod_{\ell = 2}^d\mathcal{G}(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}).
-\end{equation*}
+\end{equation}
 This model is completed with independent Gamma priors on $\xi_{\ell}$, $\tau_{\ell}$, $\zeta_{\ell}$, $\sigma_{\ell}$.
@@
 Then, the full conditional for $\alpha_{j\ell}$ takes the form
-\begin{equation*}
+\begin{equation}
     \pi(\alpha_{j\ell}\mid \bm{r},\bm{Y},\bm{\gamma},\xi_\ell,\tau_\ell,\zeta_\ell,\sigma_\ell) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i\ell}\right)^{\alpha_{j\ell} - 1}\alpha_{j\ell}^{\xi_\ell - 1}e^{-\tau_\ell \alpha_{j\ell}}\times\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\Gamma^{n_j}(\alpha_{j\ell})\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
-\end{equation*}
+\end{equation}
 for $\ell = 2,\ldots,d$.  For $\ell = 1$, as $\beta_{1} := 1$, the full conditional takes the form
@@
 \begin{equation}
+    \label{eqn:betafc}
     \beta_{j\ell}\mid\bm{r},\bm{Y},\alpha,\zeta_{\ell},\sigma_{\ell} \sim \mathcal{G}\left(\beta_{j\ell}\mid n_j\alpha_{j\ell} + \zeta_\ell, \sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right),
@@
 This is denoted as the \emph{projected restricted Gamma--log-normal} (PRG--LN) model.
-In the interest of brevity, we omit the inferential forms for this model, sufficing to say that the Gamma prior for $\beta_{j\ell}$ means it can again be integrated out, improving mixing on sampling $\bm{\alpha}_j$ with a joint Metropolis step; the normal prior for $\mu$ is conjugate for the log-normal $\bm{\alpha}_j$, and sampled via a Gibbs step; and an inverse Wishart prior for $\Sigma$ is again conjugate to the log-normal $\bm{\alpha}_j$, implying that it can also be sampled via a Gibbs step.
+Updates for $\bm{\alpha}$ can be accomplished using a joint Metropolis step, where $\beta_{j\ell}$ for $\ell = 2,\ldots,d$ have been integrated out of the log-density.
+That is,
+\begin{equation*}
+    \begin{aligned}
+    \pi(\bm{\alpha}_j\mid\bm{Y},\bm{r},\bm{\delta},\bm{\mu},\bm{\Sigma},\bm{\zeta},\bm{\sigma})
+    &\propto
+    \frac{1}{\prod_{\ell = 1}^d\alpha_{j\ell}}\exp\left\lbrace-\frac{1}{2}(\log\bm{\alpha}_j - \mu)^T\bm{\Sigma}^{-1}(\log\bm{\alpha}_j - \mu)\right\rbrace \\
+    &\hspace{-2cm}\times \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}}{\prod_{\ell = 1}^d\Gamma^{n_j}(\alpha_{j\ell})}\times \prod_{\ell = 2}^d\frac{\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
+    \end{aligned}
+\end{equation*}
+The inferential forms for $\beta_{j\ell}$ and its priors are the same as for PG--G.  The 
+  normal prior for $\mu$ is conjugate for the log-normal $\bm{\alpha}_j$, and can be sampled
+  via a Gibbs step.  Finally, the inverse Wishart prior for $\Sigma$ is again conjugate to
+  the log-normal $\bm{\alpha}_j$, implying that it can also be sampled via a Gibbs step.
 

### 2022-01-05 results.tex
update

@@
-\begin{figure}[htb]
+\begin{figure}[tb]
 \begin{center}
@@
 
-\begin{figure}[htb]
+\begin{figure}[tb]
     \begin{center}
@@
 We consider an exploration of the pairwise extremal dependence using Monte Carlo estimates of the 
-  coefficients in  Equation~\ref{eqn:chi_ij}. For this we use samples obtained fromn the PRG--G model.
+  coefficients in  Equation~\ref{eqn:chi_ij}. For this we use samples obtained from the PRG--G model.
   Figure~\ref{fig:chi_ij} provides a graphical analysis of the results. 
@@
 Right panel uses standardized units.}
-    \includegraphics{./images/condsurv_1d}
+    \includegraphics[width=\linewidth]{./images/condsurv_1d}
 \end{figure}
@@
 
-We also explored what effect the power of the projection would have---that is, which $p$-norm
-    sphere the data was projected on to for model fitting.  Using the simulated data, coming
-    from a mixture of projected Gammas, we were unable to observe an effect from $p$ when 
-    looking at a range from 1 to 15 for $p$, on our simulated data.  However, when we repeated 
-    this analysis on the IVT data, we were able to observe a drop in the energy score associated 
-    with higher $p$, with diminishing returns, as tested up to $p=15$.  The relative performance
-    between $p=10$, where the modelling for our presented analysis was conducted, and $p=15$, 
-    was not significant.
+Using our proposed scoring criteria, we explored the effect of the choice of $p$ on the final results. Using the simulated data, generated
+    from a mixture of projected Gammas, we were unable to observe sizeable differences in the scores for $p$ 
+    ranging between 1 and 15.   However, 
+    for the IVT data, we observed a drop in the energy score associated 
+    with higher $p$, with diminishing effect as $p$ increased.  
+    We observed no significant differences in the performance
+    of the model that uses $p=10$, which corresponds to the analysis 
+    presented, relative to the one that uses $p=15$.
 % EOF

### 2022-01-05 results.tex
asd

@@
+% \begin{figure}[tb]
+% \begin{center}
+%     \centering
+%   \caption{Posterior predictive loss (Left) and energy score (right) for various models on simulated data.  Gen indicates baseline.\label{fig:simpples}}
+%   \begin{subfigure}[b]{0.48\textwidth}
+%     \centering
+%     \includegraphics[height=2in, width=\textwidth]{./images/simulation_ppl}
+%   \end{subfigure}
+%   %
+%   \begin{subfigure}[b]{0.48\textwidth}
+%     \centering
+%     \includegraphics[height=2in, width=\textwidth]{./images/simulation_es}
+%   \end{subfigure}
+% \end{center}
+% \end{figure}
 \begin{figure}[tb]
-\begin{center}
     \centering
-  \caption{Posterior predictive loss (Left) and energy score (right) for various models on simulated data.  Gen indicates baseline.\label{fig:simpples}}
-  \begin{subfigure}[b]{0.48\textwidth}
-    \centering
-    \includegraphics[height=2in, width=\textwidth]{./images/simulation_ppl}
-  \end{subfigure}
-  %
-  \begin{subfigure}[b]{0.48\textwidth}
-    \centering
-    \includegraphics[height=2in, width=\textwidth]{./images/simulation_es}
-  \end{subfigure}
-\end{center}
+    \caption{Energy score (top) and Posterior Predictive Loss (bottom) for various models on simulated data with ascending count of mixture components (indicated by plot heading).
+Gen indicates baseline.\label{fig:simpples}}
+    \includegraphics[height=3in, width = \textwidth]{./images/simulation_score}
 \end{figure}

### 2022-05-27 comments.tex
pgpareto github

\documentclass[10pt]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[margin=2cm]{geometry}
\usepackage{bm}
\usepackage{xcolor}

\newcommand{\bruno}[1]{\textcolor{blue}{#1}} % Bruno edits - blue
\newcommand{\peter}[1]{\textcolor{red}{#1}}  % Peter edits - red
\newcommand{\handled}[1]{\textcolor{green}{#1}} % mark item as handled.

\begin{document}

\subsection*{Review 1}
\subsubsection*{Soundness}
Soundness The presentation of the probabilistic setting for the multivariate POT model (Section 2) and estimation method (Section 3) contains significant inconsistencies and unclear points. I list some of them below.
\begin{itemize}
    \item  the variable $Z$ is sometimes seen as the observed variable, sometimes as the unobserved variable following the limit distribution $\mu$ (after suitable restriction and normalization of the latter).

    \peter{I'm not sure what to say.  $\bm{Z}$ is always a multivariate pareto random vector.  Maybe he's saying I mis-capitalized something when referencing observed data?  I'll check.}

    \bruno{I have not checked carefully, but I think that we are consistent in using capital letters for random variables or vectors, and bold face for vectors. I think it would be useful to make that explicit at the beginning of the paper.}

    \item \handled{Also the writing of Equation (1) is confusing: the measure $\mu$ takes as input measurable sets of points, not a random variable $Z$.}

    \bruno{This comment is really annoying, it seems that he referee did not bother to understand the difference between Z and z. Anyway, making the notation explicitly, as I mentioned before, will help. You could add this in the last paragraph of Section 1.}

    \item  In the proof of Proposition 1 the ambiguity about the law of Z is quite problematic: It looks like the $Z_{\jmath}$, $Z_{\ell}$ variables are the asymptotic ones but then I cannot make sense of the proof in view of the definition for $\chi_{\jmath\ell}$.

    \bruno{Boy, this guy did not really make an effort to read the paper! The proposition is very clear, and it is valid for any random vector that be factorized as RV, no asymptotic result needed.}

    \item  It is not clear what is the link between the variables $w_l$ (beginning of Section 3) and the objects introduced in Section 2. The standardization (5) does not follow from the background material presented in Section 2.

    \peter{I think this is on me.  I introduced $\bm{w}$ as representing observed data in original scale before standardization, but I don't think that took in the reader's mind.  In equation 6, i'm using $\bm{x}$ as a generic input to the transformation function, and immediatelyu before i'm using $\bm{s}$ as a generic measureable point set.  I should probably standardize notation to use $\bm{s}$ as the generic input; replace $\bm{w}$ with $\bm{x}$ as the original data (removing $\bm{w}$ entirely).  I'm not sure where else I use $\bm{x}$, if at all.  Either that, or use $\bm{z}$ in equation 6.  $\bm{x}$ has a very strong bias in the mind towards being original data, so I should probably use that.}

    \bruno{I actually think that there is something deeper here. I think the referee is trying to figure out why the limiting behavior in th marginals implies the asymptotic behavior we considered in the previous section. I say we don't get there, we just mention, in the very beginning that we are considering the components of $\bm{Z}$ to be standardized.}

    \item  The discussion about the non-existence of a density on the infinity sphere is unclear. One may certainly start with a random variable with a density on the infinity sphere. Indeed some transformation is not differentiable so the usual change of variable formula cannot be used but this does not prove the claim. In particular I don’t see why the construction of the Gamma model on the $\mathcal{L}_p$ sphere should not result in a density for $p = \infty$ even though computing the density explicitly may be difficult.

    \peter{both reviewers had trouble in this point.  The transformation explicitly as described is not recoverable at $p = \infty$ when $\underset{\ell}{\arg\max}z_{\ell} \neq d$.  It's possible to proceed under a conditional transformation (that is, given $\underset{\ell}{\arg\max}z_{\ell} = \jmath$, but a closed form expression then requires the probability of the conditioning statement.  The interesting thing is the density of a point on the $\mathcal{L}_p$ sphere does asymptotically approach some value as $p\to\infty$ (or, at least, the rest of the Jacobian does.  I didn't calculate for the density in general, but the rest of it is just gamma stuff).}

    \bruno{OK, I am not going to discuss if there is or there is not a density WRT some meausre in some subspace. Let's make it clear that we are considering projected gamma distributions as the building block of our model, and that the transformation that projects from a produjct of independent gammas to the unit hypercube for norm $p$ breaks down for $p=\infty$, due to lack of differentiability. }
    \item Section 3.2.1: The Dirichlet Process is not properly introduced. It is unclear why and how clusters of observations occur and what are these clusters.

    \peter{I think this is the sentence which generates this comment: "The Dirichlet process mixture model groups observations into stochastically assigned clusters."  That's \emph{true} for a given sampled set of parameters, but I should probably state that better.  The sentence was used to introduce the full conditionals for the cluster membership identifiers.  Will require a bit of word-smithing.}

    \bruno{I would not pay much attention to this.}

\end{itemize}
\subsubsection*{Vagueness}
 Many vague formulations, here is a sample
\begin{itemize}
    \item What does the first sentence of section 3) mean? [‘Consider a collection of observations $w_{\ell}$ ,$\ell = 1,...,n$ that exhibit extreme behavior’].

    \peter{This was where I tried to introduce the $\bm{w}$ notation as indicating data at original scale.  I think the issue here is I'm using the subscript $\ell$ rather than $i$ for indexing along observations.  So in this case,  not vague, just wrong notation.  But also, should switch to using $\bm{x}$ for observed data at original scale.}

    \item  ‘To define an angular measure, we are interested in the direction of vectors in ${\mathbb R}_+^d$ thus, we project them onto [...]’ : The angular variable variable is not really a projection (at least not an orthogonal one)

    \peter{I'm not really sure I understand what he means here.}
\end{itemize}
\subsubsection*{Unclear goals/method assessment}
It is not clear to me what is the purpose of estimating the angular measure for the infinity norm rather than the $\mathcal{L}_p$ norm. In other words, since the proposed Gamma family on the $\mathcal{L}_p$ sphere has a workable expression, what is the purpose of changing the norm and trying to estimate the $\mathcal{L}_{\infty}$ angular measure? (As I understand from the description of the proposed strategy p.7). On the other hand, There does not \peter{[appear]} to be any theoretical result in the paper supporting the proposed strategy.

\peter{We're not evaluating on the $\mathcal{L}_{\infty}$ norm either. We're trying to approximate geodesic distance on the surface of a sphere defined by the $\mathcal{L}_{\infty}$ norm.  A direct rebuttal to this question would be to ask whether we would use Euclidean distance for evaluating the distance between two cities.}

\peter{On re-read, I'm not clear whether he's wondering why switch the evaluation regime back to $S_{\infty}^{d-1}$ rather than $S_{p}^{d-1}$.  That seems obvious---the problem is defined on $S_{\infty}^{d-1}$, and if we could do the entire process in that space, we would.  Also, a proper geodesic distance in $\mathbb{S}_p^{d-1}$ is possible, but also difficult.}

\peter{Maybe he's suggesting using density-based rather than distance-based metrics, evaluated on $\mathbb{S}_p^{d-1}$?}

\bruno{We can use any norm we want. The problem is that if you use a norm that is not the infinite norm, then you have to deal with a moment restriction on the measure, and that is a pain. We do say that \emph{Our focus on the angular measure is similar to that in Sabourin
and Naveau (2014) and Hanson et al. (2017), that consider Bayesian non-parametric ap-
proaches. Yet, our approach avoids the problem of dealing with the so called moment
constraint by considering a constructive definition of the multivariate GP based on the
infinity norm.}, but maybe it should be emphasized.}
\subsubsection*{Organization of the paper}
The construction of a negative kernel on the infinity sphere for the Energy Score is described at length, whereas this looks to me not to be a central contribution of the paper. I suggest moving some details to some appendix section in order not to interrupt the flow of ideas.

\peter{Really?  I thought it was kind of important, as it makes evaluation feasible; not merely a minor algorithmic detail.}

\bruno{No way!}

\subsubsection*{Comparison with other methods aiming at same goal}
There is none. One may e.g. think of comparing with the results obtained with the empirical angular measure or smoothed versions of it in terms of extremal dependence and conditional survival probability.

\peter{This is probably a good point.}

\bruno{Sure, this is a good point, and should be on your todo list. There is a suggestion by the other referee in this respect.}

\subsubsection*{Minor Points}
\begin{itemize}
    \item (Section 3, p. 5) ‘We are, as yet, unaware of any known distributions with a support in this space’ (that is the positive orthant of the infinity sphere): Even though I guess I see the author’s point the formulation is not appropriate because one may think of many simple distributions on this space which is just a union of hyper-cubes.

    \item  \handled{(typos): Some confusions between ${\mathbb R}^d$ and $\mathbb{R}^p$.}

    \peter{Yes, 2 instances of $\mathbb{R}^p$ (line 45 of methodology\_pnorm.tex, line 2 of methodology\_projgamma.tex}

\end{itemize}

\subsection*{Review 2}
\subsubsection*{Comment 1}
In the beginning of Section 2, the probability theory underlying multivariate extremes is incorrectly described. In the first stated equation, the event on the left-hand side should be $n^{-1}\bm{Z} \not < \bm{z}$, which, for $d \geq 2$, is different from $n^{-1}\bm{Z}\leq \bm{z}$;

\peter{Yeah.  Looking back through this and the original theory, this is a pretty egregious typo that escaped my notice.}

also, it is worth specifying that $\bm{z}\in (0, \infty]^d$, otherwise the limit $\mu([\bm{0},\bm{z}]^c)$ is infinity. Further, Eq. (1) is not correct: the limit measure $\mu$ does not determine the distribution of $\bm{Z}$ as stated in that formula. The measure $\mu$ has infinite mass and its radial component is not restricted to the set $\lbrace \bm{z} \in {\mathbb R}_+^d : \max(z_1,\ldots,z_d) > 1\rbrace$, in contrast to what the requirement $R > 1$ suggests.

\peter{Equation 1 which is called out here is a transcription of equation 4 from the Goix et al paper, which in turn comes from Ferreira and de Haan.  Maybe I need to explicitly state that we're considering the measure \emph{within} the support of the standard multivariate Pareto: $[\bm{0},\bm{\infty})\backslash [\bm{0},\bm{1}]$?}

On the bottom of p3, an expectation is taken with respect to $\Phi$: but $\Phi$, but $\Phi$ is a probability measure only if its total mass is equal to one, its total mass being
\[
    \Phi(\mathbb{S}_{\infty}^{d-1}) = \lim\limits_{n\to\infty} n\text{Pr}\left[\max(Z_1,\ldots,Z_n) > n\right]
\]
The requirement that $\Phi$ is a probability measure and thus that $\Phi(\mathbb{S}_{\infty}^{d-1}) = 1$ is a further assumption on $\bm{Z}$ which does not follow from multivariate regular variation as assumed in the beginning of the section. In (2), since it has already been assumed that $R$ and $\bm{V}$ are independent (one line above (1)), we simply have $\text{Pr}[\bm{V} \in A \mid R > r ] = \text{Pr}[ \bm{V} \in A ]$. A correct way to describe the theory, in line with the authors’ intentions, is that provided $\bm{Z}$ is of the form \[\bm{Z} = R\bm{V}\] with $R$ independent of $\bm{V}$ and if $R$ is unit Pareto while $\bm{V}$ is concentrated on $\mathbb{S}_{\infty}^{d-1}$ and has probability distribution $\Phi$ [in particular, $\Phi(\mathbb{S}_{\infty}^{d-1}$ by assumption], then the convergence to $\mu$ stated in the beginning of the section (to be corrected as above) holds with limit measure $\mu$ determined by
\[
    \mu\left(\left\lbrace  \bm{z} \in [0,\infty)^d : \lVert \bm{z} \rVert_{\infty} > r,\;\bm{z}/\lVert\bm{z}\rVert_{\infty} \in A\right\rbrace \right) = r^{-1}\Phi(A)
\]
for $r > 0$ and for Borel sets $A\subseteq \mathbb{S}_{\infty}^{d-1}$

\peter{The suggested edit essentially follows the same line: I need to be more explicit about the support of $\bm{Z}$ (along with $A\subseteq{\mathbb S}_{\infty}^{d-1}$).  I was sloppy in the writing because I was assuming a defined support, where I apparently hadn't sufficiently defined the support.}

\bruno{OK, this is complicated. I have agonized with this for a while, and I thought we had it pinned down. The typo did not help, of course, but the referee makes a typo himself! ;-) I thought that, by essentially copying the approach that Goix et al were using, we had killed the issue. Anyway, the referee is being really helpful here, he/she is suggesting how to write things. I propose we actually follow the advice. The only point is that we need a quote for \emph{the convergence to $\mu$ holds}, I mean, why does it hold? There is a quote in Goix et al paper, that one may be helpful.}

\bruno{I think it is obvious that the convergence holds. Define  $R = \max_{i} Z_i$, and assume that $R>1$ with probability one. Also, define $V_i = Z_i/R$, and assume that $E(V_i)>0$. Assume that $ \bm{Z}$ and $R$ are independent and that $R$ is distributed as a Pareto. Then these two assumptions correspond to the third condition in Theorem 2.1 of
\cite{ferreira2014}, applied to the finite-dimensional case. This is equivalent to the second condition that implies that
\[
    Pr(\bm{Z}\in r A) = \frac{1}{r} Pr(\bm{Z} \in A) 
\]
Then
\[
  \lim_{n\rightarrow\infty} n Pr(\frac{1}{n}\bm{Z} \not < \bm{z}) =  Pr(\bm{Z}\not < \bm{z})
\]
}
\subsubsection*{Comment 2}
In the review, it is worth mentioning that various parametric models for multivariate generalized Pareto distributions are considered in \cite{kiriliouk2018}.

On p5, the authors write: “We are, as yet, unaware of any known distributions with a support in this space.” Passing to logarithmic coordinates via $\bm{S} = \log\bm{V}$ (component-wise), we can construct $\bm{S}$ via $\bm{S} = \bm{T} - \max_i T_i$  for an arbitrary random vector $\bm{T}$. This is the representation in \cite{rootzen2018mgpd}, for which a wealth of parametric models are proposed in Kiriliouk et al. Generators with independent (log-)Gamma components are proposed in that article, and these are associated to the Dirichlet max-stable distribution. As this model resembles the proposal in the paper, a comparison would be appropriate.

\peter{Adding both to the reading stack.}

\bruno{Good. This is an important one.}
\subsubsection*{Comment 3}
On p5, below (5), do the authors mean $r_i = \lVert\bm{z}_i\rVert_{\infty}$? There is no reason why $\lVert \bm{w}_i\rVert_{\infty}$ would exceed 1, since the marginal scales are arbitrary.

\peter{Yeah.  This is a pretty serious typo; I missed it.}

Note that to make multivariate GP modelling appropriate, the effective sample should consist of those observations for which there exists $\ell = 1,\ldots,d$ such that $w\bm{w}_{i\ell} > b_{t\ell}$, rather than for all $\ell = 1,\ldots,d$ simultaneously. By construction, we automatically have $\lVert \bm{z}_i\rVert_{\infty} > 1$ as soon as there exists $\ell = 1,\ldots,d$ such that $w_{i\ell} > b_{t\ell}$.

\peter{We specify as such in algorithm 2.  I think, fixing that typo, that we effectively say as such in the referenced paragraph.}

\bruno{Fix the typo and make clear what we are doing.}

\subsubsection*{Comment 4}
p6, below (8): Do you mean $\mathbb{R}_+^d$?

\peter{same typo referenced by previous reviewer.}

More importantly, it is written that the projected distribution on $\mathbb{S}_{\infty}^{d-1}$ will not have a density. I believe it does, actually. The set $\mathbb{S}_{\infty}^{d-1}$ is the upper boundary of the unit hypercube in $\mathbb{R}^d$, and this boundary is a union of $d$ faces of dimension $d-1$. The projected distribution can be expected to have a density with respect to $(d - 1)$-dimensional Lebesgue measure on each of those faces.

\peter{The other reviewer has a similar comment; I don't think there's a way to create a density in closed form.}

\subsubsection*{Comment 5}
In Section 3.2.1 on p7, clusters suddenly appear, but I found it hard to follow what they are or where they originate from. Do they form a (random) partition of the observations? Can an observation belong to multiple clusters? What is $\delta_i$? What does it mean to “draw a cluster from $G_0$”? On top of the page, $G_0$ is a distribution of the Gamma parameter $\bm{\theta}$.

\peter{I'm thinking I need a greater discussion to motivate the use of the Dirichlet process prior?  As well as, from what it seems, a small discussion on how the Dirichlet process works?  I thought my explanation of algorithm 8 was fairly straightforward...but reading now it seems only straightforward \emph{assuming} the audience is familiar with the Chinese restaurant process representation of the Dirichlet process.}

\bruno{Yeah, not everybody has taken the BNP class! ;-) I don't know that I would go into more explanation of how the DP works, I would quote papers where it is used as an efficient way to obtain flexible classes of densities. }

\subsubsection*{Comment 6}
The inference method in Section 3.2.1 focuses on the dependence structure. However, the marginal GP parameters in (5) are uncertain too, while the transformation to the standard Pareto scale from which the angular components are extracted is done using these parameters. Therefore, it also needs to be said how the uncertainty on the marginal GP parameters is incorporated in the procedure.

In the case study (p14), the marginal parameters are estimated by maximum likelihood and then held fixed. In this way, the true uncertainty is probably underestimated.

\peter{We agreed fairly early on that trying to introduce the uncertainty from the marginal GP parameters would be a \emph{mess}.  I guess we should explicitly state as such?  We side-step the problem using MLE, but as far as I can tell side-stepping the marginalization uncertainty problem is fairly standard practice...}

\subsubsection*{Comment 7}
The simulation study in Section 5.1 is not conducted in the proper multivariate PoT setting. Given a data-set, we need to choose a high threshold for each variable, select those observations such that there is a threshold excess in at least one of the $d$ variables, transform the margins to a standard scale (using estimated univariate GP parameters or via ranks), and then compute the radial and angular components of the standardized points. Instead, the authors immediately generate the angular components by a mixture of projected Gammas. A simulation study mimicking the procedure in Algorithm 2 would be more appropriate.

\peter{The simulation study was meant to evaluate the efficacy of the model in recovering the distribution of angular data; not specifically in recovering the angular distribution of GP data.  Should we approach this in another manner?}

Moreover, the analysis is based on 16 data-sets only. Each point in Figure 2 represents just a single data-set. In case of 9 mixture components and in dimension 6, the PRG-LN method seems to do much worse than the other methods. Is this a real effect or is it just due to this particular data-set?

\peter{Weirdly enough, I thought the same, and simulated another set of trial datasets to run the models on again.  Same result---same weird spike in log-normal model.  It seems to be a real effect, but I don't understand why.}

Also, as soon as the dimension is not too small, it is in practice not always likely that all variables are large simultaneously. That is, for real-life data, many angular vectors may thus have a large fraction of components that are close to zero. If it is not already the case, it would be informative to allow for this in the simulation study as well.

\peter{Of course.   Gamma with random shape/rate is a pretty unstable distribution.}

\subsubsection*{Minor Comments}
\begin{itemize}
    \item Proposition 1 is well-known: see for instance Eq. (47) in Rootzén, Segers and Wadsworth (JMVA 2018) or also Theorem 6 in the main paper. Coles’ $\chi$ is equal to the bivariate tail copula evaluated in the point $(1 , 1)$, i.e., $R(1,1) = 2-\ell(1 , 1)$ (with $\ell$ the stable tail dependence function), which gives the formula in the proposition via Eq. (21) in the cited article and the connection $e^{\bm{S}} = \bm{V}$.

    Proposition 2 is an immediate consequence of the above representation $\bm{Z} = R\bm{V}$  via \[\text{Pr}[\forall\ell\in \beta:Z_{\ell}>z_{\ell}] = \text{Pr}\left[1/R < \bigwedge_{\ell \in B}V_{\ell}/z_{\ell}\right] = \text{E}\left[1\wedge\bigwedge_{\ell \in \beta}V_{\ell}/z_{\ell}\right]\] for non-empty $\beta\subset\lbrace 1,\ldots, d\rbrace$, since $1 /R$ is uniformly distributed between 0 and 1 and is independent of $\bm{V}$.

    \item  In (22), note that $\text{Var}_P[X_{i\ell}] + (\text{E}_p[X_{i\ell} - x_{\beta\ell})^2 = \text{E}_P[(X_{i\ell} - x_{i\ell})^2]$
    \item On p10, could you give a reference confirming that Euclidean distance defines a negative definite kernel?

    \peter{It's stated as such in the reference for proper scoring rules---I guess mention again?}

    \item On p10–11, it is argued that the computation of the geodesic distance between two points on different faces of $\mathbb{S}_{\infty}^{d-1}$ is difficult. I wonder if the minimum defined in Proposition 3 doesn’t actually give that geodesic distance. Let $\bm{x} = (x_1,\ldots,x_d)$ and $\bm{y} = (y_1,\ldots,y_d)$ be two such points, with $x_i = 1 = y_j$ for two distinct indices $i,j \in \lbrace1,\ldots,d\rbrace$, whereas the other coordinates of $\bm{x}$ and $\bm{y}$ belong to $[0 , 1)$. Intuitively, it seems obvious that the geodesic connecting $\bm{x}$ and $\bm{y}$ will be the union of two line segments of the form $[\bm{x},\bm{z}]$, and $\bm{z},\bm{y}$, for a certain point $\bm{z} = (z_1,\ldots,z_d)$ on the intersection of the two faces, i.e., $z_i = z_j = 1$.  All that is required is then to find this point $\bm{z}$, 
    but this is precisely the kernel in Proposition 3 and studied in Proposition 4.

    \peter{Intuitive, but wrong (we got caught in the same idea).  It's easy to consider a 3-dimensional example where this doesn't hold true. In the worst case in said example, the proposed kernel is approaching $\sqrt{2}$ times longer than the "geodesic".  Thinking about it, we may be able to generalize that to $\sqrt{d-1}$ times longer than the "geodesic".  Should we put in this caveat?  I think that upon reading this, the viewer would overestimate the probability of this "worst case scenario".}

    \item On p14 and in Figure 4, it makes perfect sense that for nearby grid cells, extremal dependence is strong (threshold excesses occur simultaneously and points lie along the diagonal) whereas for grid cells far apart (C1 and C8), extremal dependence is weak (threshold excesses occur rarely simultaneously and points are hugging the two axes). This is typical for spatial data.

    \peter{Yes...  that's what we expected to see.  It's nice to see that the model bears it out too?}

    \item In the case study, the pairwise extremal coefficients in Eq. (3) are computed from the fitted parametric models and shown in Figure 5. However, non-parametric estimates of these coefficients may be calculated easily too by counting points with a threshold excess in the two variables simultaneously, given that there is an excess in at least one of the variables (this is the empirical tail copula or the empirical tail dependence coefficient). Comparing parametric and non-parametric estimates would provide a diagnostic for the goodness-of-fit.

    \peter{True.  Not sure if we can fit such into the paper.}

    \item Figure 6: the caption should mention which of the two data-sets the figure is about.

    \peter{Oops.  Yeah, caption etiquette.}

    \item Writing
    \begin{itemize}
        \item p2, top: “a scale parameter”
        \item p2, middle: the references Renard and Lang (2007) and Falk et al. (2019) are separated by a “;”, even though the sentence continues.
        \item p2, bottom: “a set of simulated data”
        \item p3, top: “the relevance of extreme value theory”
        \item p10, bottom: “a portion of a d-cube”
        \item p12, beginning of Section 5.1: $\mathbb{S}_{\infty}^{d-1}$ rather than $\mathcal{S}_{\infty}^{d-1}$
        \item p15, Algorithm 2: $\chi_{\ell}$ should be $\xi_{\ell}$
    \end{itemize}
\end{itemize}

\bibliographystyle{plain}
\bibliography{refs}

\end{document}

### 2022-05-27 cover.tex
pgpareto github

\documentclass[letterpaper,11pt]{letter}

\name{Peter Trubey}
\signature{Peter Trubey}
\address{Peter Trubey \\ Department of Statistics \\ University of California, Santa Cruz \\ USA}

\begin{document}%

\begin{letter}{Editorial Board \\ Computational Statistics \& Data Analysis}

\opening{Dear Editorial Board,}

We submit to you our manuscript for review.  It details a method of inference for the dependence structure of
extremes in a multivariate peaks over thresholds model based mixtures of projections of vectors of gamma variables, 
along with an appropriate model selection criteria and applications of the fitted model.  A simulation study 
is performed to explore properties of the distribution, and the model then applied to atmospheric data.

We confirm the content of this work is original, and is not under consideration at any other journal.

We further confirm that there are no competing interests to declare.

Sincerely,

\closing{~}

\end{letter}
\end{document}

% EOF

### 2022-05-27 evaluation.tex
pgpareto github

@@
   is not differentiable, implying that routines to calculate geodesics are not readily
-  available.  However, as ${\mathbb S}_{\infty}^{d-1}$ is a portion of of a $d$-cube, we
+  available.  However, as ${\mathbb S}_{\infty}^{d-1}$ is a portion of a $d$-cube, we
   can borrow a result from geometry \citep{pappas1989} stating that the length of the

### 2022-05-27 introduction.tex
pgpareto github

@@
 Inferential methods for the extreme values of univariate observations
-are well established and software is widely available \cite[see, for example,][]{coles2001}.
-For variables in one dimension the application of EVT methods considers the asymptotic distribution of either the maxima calculated for regular blocks of data, or the values that exceed a certain threshold.
-The former leads to a Generalized Extreme Value (GEV) distribution, that depends on three parameters.
-The latter leads to a Generalized Pareto (GP) distribution, that depends on a shape and a scale parameters.
-Likelihood-based approaches to inference can be readily implemented in both cases.
-In the multivariate case the GEV theory is well developed \citep[see, for example][]{dehaan2006}, but the inferential problem is more complicated, as there is no parametric representation of the GEV.
-This problem is inherited by the PoT approach and compounded by the fact that there is no unique definition of an exceedance of a multivariate threshold, as there is an obvious dependence on the norm that is used to measure the size of a vector.
+are well established and software is widely available \cite[see, for example,][]{coles2001}. For variables 
+in one dimension the application of EVT methods considers the asymptotic distribution of either the maxima 
+calculated for regular blocks of data, or the values that exceed a certain threshold. The former leads to
+a Generalized Extreme Value (GEV) distribution, that depends on three parameters. The latter leads to a 
+Generalized Pareto (GP) distribution, that depends on a shape and a scale parameter. Likelihood-based 
+approaches to inference can be readily implemented in both cases.
+In the multivariate case the GEV theory is well developed \citep[see, for example][]{dehaan2006}, but the
+inferential problem is more complicated, as there is no parametric representation of the GEV. This problem
+is inherited by the peaks over threshold (PoT) approach and compounded by the fact that there is no unique definition of an 
+exceedance of a multivariate threshold, as there is an obvious dependence on the norm that is used to
+measure the size of a vector.
 
-During the last decade or so, much work has been done in the exploration of the definition and properties of an
-  appropriate generalization of the univariate GP distribution 
+During the last decade or so, much work has been done in the exploration of the definition 
+  and properties of an appropriate generalization of the univariate GP distribution 
   to a multivariate setting.  To mention some of the papers on the topic, the work of
@@
   \cite{rootzen2018} while \cite{RoSeWa2018a} provides insight on the theoretical properties of possible parametrizations.
+These are use in \cite{KiRoSeWa2019} for likelihood-based models for PoT estimation.
   A frequently used method for describing the dependence
   in multivariate distributions is to use a copula.
-\cite{renard2007,falk2019}, provide successful examples of this approach in an EVT framework.
+\cite{renard2007}, and \cite{falk2019} provide successful examples of this approach in an EVT framework.
 \cite{ferreira2014} presents a constructive definition of the Pareto process, that generalizes the GP to an infinite dimensional setting.
@@
 Section~\ref{sec:evaluation} develops criteria for model selection in the support of the angular measure.
-Section~\ref{sec:results} explores the efficacy of the proposed approach on a set simulated data, and, acknowledging the relevance of of extreme value theory to climatological events~\citep{jentsch2007,vousdoukas2018,li2019}, estimates the extremal dependence structure for a measure of water vapor flow in the atmosphere, used for identifying atmospheric rivers.
+Section~\ref{sec:results} explores the efficacy of the proposed approach on a set of simulated data, and, acknowledging the relevance of extreme value theory to climatological events~\citep{jentsch2007,vousdoukas2018,li2019}, estimates the extremal dependence structure for a measure of water vapor flow in the atmosphere, used for identifying atmospheric rivers.
 Finally, Section~\ref{sec:conclusion} presents our conclusions and discussion.
 
-Throughout the paper, we adopt the operators $\wedge$ to denote minima, and the $\vee$
-  to denote maxima.  Thus $\wedge_i x_i = \min_i x_i$, $\vee_i x_i = \max_i x_i$.  These operators can
-  also be applied component-wise between vectors, such as $\bm{a}\wedge\bm{b} = (a_1\wedge b_1, a_2\wedge b_2,\ldots)$.
+Throughout the paper, we adopt the operators $\wedge$ to denote minima, and 
+    the $\vee$ to denote maxima.  Thus $\wedge_i s_i = \min_i s_i$, and $\vee_i s_i = \max_i s_i$.  
+    These operators can also be applied component-wise between vectors, such as
+    $\bm{a}\wedge\bm{b} = (a_1\wedge b_1, a_2\wedge b_2,\ldots)$.  We use 
+    uppercase to indicate random variables, lowercase to indicate points, and
+    bold-face to indicate vectors or matrices thereof.
 
 \section{A multivariate PoT model\label{sec:multivariatepot}}
-To develop a multivariate PoT model for extreme values consider a $d$-dimensional random vector $\bm{Z} \in {\mathbb R}^d_+ = [0,+\infty)^d$, the positive $d$-dimensional cone.
-Following the discussion in \cite{goix2017} we assume a form of regular variation, consisting on assuming the existence of a limiting measure $\mu$, such that.
+%To develop a multivariate PoT model for extreme values consider a 
+%$d$-dimensional random vector $\bm{W} = (W_1, \ldots ,W_d)$  with 
+%marginal distributions $F_1(w_1), \dots , F_d(w_d)$, and define its
+%standardization as $\bm{Z} = (Z_1, \ldots , Z_d)$, where $Z_i = 1/ (1 - F_i(W_i)), \; i=1, \ldots , d$.
+%Following the discussion in \cite{goix2017} we assume a form of 
+%regular variation, consisting on assuming the existence of a 
+%limiting measure $\mu$ on $(0,\infty]^d$, such that.
+%\begin{equation}\label{eq:reg-var}
+%        \lim\limits_{n\to\infty}n\prob{\frac{1}{n}\bm{Z}\not\leq \bm{z}} = \mu\left([\bm{0},\bm{z}]^c\right),
+%\end{equation}
+%where the inequality and the interval are taken componentwise. 
+%In this setting, $\mu$ is the asymptotic distribution of $\bm{Z}$ 
+%in extreme regions, and it is  referred to as the \emph{exponent
+%measure}. In addition, we assume that $\lim_{n\rightarrow\infty} 
+%n\prob{\max\{Z_1, \ldots , Z_d\}>n} = 1$. Equation \ref{eq:reg-var}
+%implies that $\mu$ has the homogeneity property $\mu(t\cdot) = 
+%\frac{1}{t}\mu(\cdot)$i, and it can be shown that  $\mu$ 
+%can be decomposed into a radial and an angular component. In fact, 
+%for $\bm{z} = (z_1, \ldots , z_n) \in (0,\infty]^d$, let $r(\bm{z}) = 
+%||\bm{z}||_\infty$ and $\bm{v}(\bm{z}) = \bm{z}/r(\bm{z})$. Define 
+%$\Phi(B) = \mu(\{\bm{z} : r(\bm{z})>1, \bm{v}(\bm{z}) \in B\})$, for any
+%$B\subset \mathbb{S}_{\infty}^{d-1}$, the positive orthant of the unit 
+%sphere in infinite norm. $\Phi$ is referred to as the {\em angular 
+%measure}. Then 
+%\begin{equation}\label{eq:exp-measure}
+%    \mu\left(\left\lbrace  \bm{z} \in [0,\infty)^d : \lVert \bm{z} \rVert_{\infty} > r,\;\bm{z}/\lVert\bm{z}\rVert_{\infty} \in B\right\rbrace \right) = r^{-1}\Phi(B).
+%\end{equation}
+%Thus, the radial and angular components of the exponent measure are
+%independent. To obtain a multivariate peak over threshold model we 
+%consider the radial and angular component of the random variable $\bm{Z}$
+%denoted, respectively, as $R = \max_{i} Z_i$ and $\bm{V} = \bm{Z}/R$. 
+%Then, using Equations \ref{eq:reg-var} and \ref{eq:exp-measure}, we have
+%that
+
+%To obtain a statistical approach to estimating $\mu$, we follow the approach proposed in .
+%Define  , and assume that it is distributed as a standard Pareto, and thus $R>1$ with probability one.
+Also, define , and assume that $\bm{V}$ has probability distribution $\Phi$ concentrated in $\mathbb{S}_{\infty}^{d-1}$, the positive orthant of the unit sphere in infinite norm.
+Also, assume that $E(V_i)>0$, and that $ \bm{Z}$ and $R$ are independent.
+%We refer to $R$ and $\bm{V}$ as, respectively, the  of $\bm{Z}$. Under those assumptions,  the convergence stated in Equation \ref{eq:reg-var} holds with limit measure $\mu$ determined by
+
+%\citep[e.g.][applied to the finite-dimensional case]{ferreira2014}
+%\[
+%    \mu\left(\left\lbrace  \bm{z} \in [0,\infty)^d : \lVert \bm{z} \rVert_{\infty} > r,\;\bm{z}/\lVert\bm{z}\rVert_{\infty} \in A\right\rbrace \right) = r^{-1}\Phi(A)
+%\]
+%for $r > 0$ and for Borel sets $A\subseteq \mathbb{S}_{\infty}^{d-1}$.
+
+%    $\mu(tA) = \frac{1}{t}\mu(A)$.  Define  $R = \max_{i} Z_i$, and assume that $R>1$ with probability one. 
+%    Also, define $V_i = Z_i/\max_i Z_i$, and assume that $E(V_i)>0$. These two assumptions together with the 
+%    homogeneity of $\mu$ correspond to the second condition in Theorem 2.1 of
+%    \cite{ferreira2014}, applied to the finite-dimensional case.
+%    Thus, we have the factorization $\bm{Z} =  R\bm{V}$, where $R$ and $\bm{V}$  are independent. Moreover,
+%    for some measure $\Phi$, $\mu(\cdot)$ is factorized as
+%    \temp{\begin{equation}\label{mufact}
+%        %\mu\left( \left\lbrace\bm{Z} : V \in A , R > r\right\rbrace \right) = r^{-1}\Phi(A).
+%        \mu\left(\left\lbrace \bm{z}\in \mathbb{R}_+^d : R(\bm{z}) > r, V(\bm{z}) \in A\right\rbrace\right) = r^{-1}\Phi(A)
+%    \end{equation}
+%    for $R(\bm{Z}) > 0$, and Borel sets $A \subseteq \mathbb{S}_{\infty}^{d-1}$.  Hereafter $R$ is referred
+%    to as the radial component, and $\bm{V}$ the angular component of $\bm{Z}$
+%    \st{$R$ describes the radial component, and $\bm{V} \in {\mathbb S}_{\infty}^{d-1}$ the 
+%    angular component of $\bm{Z}$.}}  Here ${\mathbb S}_\infty^{d-1}$ denotes the positive
+%    orthant of the unit sphere in infinite norm.  $\Phi(\cdot)$ is referred to as the \emph{angular measure}.   
+%    From Equation (\ref{mufact}) we see that $R$ is distributed as a standard Pareto random variable. Conditioning on $R$, we have that
+%    \begin{equation} \label{MGP}
+%        \prob{\bm{V} \in B \mid R > r} = \frac{r\prob{\bm{V} \in B, R > r}}{r\prob{R > r}}\hspace{0.2cm}
+%            \xrightarrow[r\to\infty]{~} \hspace{0.2cm} \Phi(B),
+%    \end{equation}
+%as $\lim_{r\rightarrow\infty} r\prob{R>r} = 1$.
+%    Thus, the angular measure corresponds to the distribution of $\bm{V}$, conditional on $R$ being large.
+%    As such, it controls the dependence structure of $\bm{Z}$ in the tails. 
+%    In fact, assuming that $E(V_i)>0$ 
+%    and following \cite{ferreira2014}, in the limit the distribution function of $\bm{Z}$ is given as
+%    \begin{equation*}
+%        \prob{Z_1\leq z_1, \ldots ,Z_d\leq z_d} = \expect{\bigvee_{i=1}^d
+%    \frac{V_i}{z_i\wedge 1}} - \expect{\bigvee_{i=1}^d
+%    \frac{V_i}{z_i}}
+%    \end{equation*}
+%    where the expectation is taken with respect to $\Phi$
+%In view of this, to obtain a PoT model we consider a limiting 
+%distribution for $\bm{Z}$ that corresponds to the distribution
+%of the product of two independent random variables $\tilde{R}$ 
+%and $\tilde{\bm{V}}$, with distributions, respectively, standard 
+%Pareto and $\Phi$. \cite{ferreira2014} refer to such model as a 
+%simple  Pareto process indexed on a finite dimensional set.  We 
+%seek a flexible model for the distribution
+%of $\tilde{\bm{V}} \in {\mathbb S}_{\infty}^{d-1}$.
+
+%An alternative approach to justify the above mentioned limiting
+%distribution, that provides a link to the multivariate GEV and the
+%theory of max-stable distributions, is discussed 
+%in Remark 1 of \cite{RoSeWa2018a}. More specifically, 
+%using the notation in Theorem 6 of \cite{RoSeWa2018a}, define $\bm{S} 
+%=  \log(\tilde{\bm{V}})$.  Then, according to Definition 5, $\bm{S}$ 
+%is a spectral  random vector. Taking  $E \sim \text{Exp}(1)$, 
+%independent of  $\tilde{\bm{V}}$,  we have that $\bm{Y} = \bm{S} + E$  
+%follows a multivariate GP, according  to Theorem 6.  Then, using 
+%Proposition 4,  $\tilde{\bm{V}}\tilde{R} = e^{\bm{Y}}$ will also follow
+%a multivariate GP. Notice that $\tilde{R} = e^E$ is distributed as 
+%a standard Pareto random variable. This is one of several stochastic
+%representations of a multivariate GP presented in \cite{RoSeWa2018a}.
+
+To develop a multivariate PoT model for extreme values consider a 
+$d$-dimensional random vector $\bm{W} = (W_1, \ldots ,W_d)$ with
+cumulative distribution $F$. Following \cite{RoSeWa2018a}, assume 
+that there exists sequences of vectors $\bm{a}_n$ and $\bm{b}_n$,
+and a $d$-variate distribution $G$ such that 
+$\lim_{n\rightarrow\infty} F^n(\bm{a}_n \bm{w} + \bm{b}_n) = 
+G(\bm{w})$. $G$ is a $d$-variate generalized extreme value 
+distribution. It follows that
 \[
-    \lim\limits_{n\to\infty}n\prob{\frac{1}{n}\bm{Z}\geq \bm{z}} = \mu\left([\bm{0},\bm{z}]^c\right),
+    \lim_{n\rightarrow\infty}\prob{\bm{a}^{-1}_n (\bm{W} - \bm{b}_n)
+    \leq \bm{w} | \bm{W} \not\leq \bm{b}_n)} = \frac{\log G(\bm{w} 
+    \wedge \bm{0}) - \log G(\bm{w}) }{\log G(\bm{0})} = H(\bm{w}).
 \]
-where the inequality and the interval are taken componentwise. 
-In this setting, $\mu$ is the asymptotic distribution of $\bm{Z}$ in extreme regions, and it is referred to as the
-  \emph{exponent measure}. Under this condition $\mu$ features the homogeneity property $\mu(tA) = \frac{1}{t}\mu(A)$. 
-Define  $R = \max_{i} Z_i$, and assume that $R>1$ with probability one. Also, define $V_i = Z_i/\max_i Z_i$, and assume that $E(V_i)>0$.
-These two assumptions together with the homogeneity of $\mu$ correspond to the second condition in Theorem 2.1 of
-\cite{ferreira2014}, applied to the finite-dimensional case.
-Thus, we have the factorization $\bm{Z} =  R\bm{V}$, where $R$ and $\bm{V}$  are independent. Moreover,
-for some measure $\Phi$, $\mu(\cdot)$ is factorized as
-\begin{equation}\label{mufact}
-    \mu\left( \left\lbrace\bm{Z} : V \in A , R > r\right\rbrace \right) = r^{-1}\Phi(A).
-\end{equation}
-$R$ describes the radial component, and $\bm{V} \in {\mathbb S}_{\infty}^{d-1}$ the 
-angular component of $\bm{Z}$.  Here ${\mathbb S}_\infty^{d-1}$ denotes the positive
-orthant of the unit sphere in infinite norm.  $\Phi(\cdot)$ is referred to as the \emph{angular measure}.   
-From Equation (\ref{mufact}) we see that $R$ is distributed as a standard Pareto random variable. Conditioning on $R$, we have that
-\begin{equation} \label{MGP}
-    \prob{\bm{V} \in A \mid R > r} = \frac{r\prob{\bm{V} \in A, R > r}}{r\prob{R > r}}\hspace{0.2cm}
-      \xrightarrow[r\to\infty]{~} \hspace{0.2cm} \frac{\Phi(A)}{\Phi({\mathbb S}_{\infty}^{d-1})}.
-\end{equation}
- Thus, the angular measure corresponds to the distribution of $\bm{V}$, conditional on $R$ being large.
- As such, it controls the dependence structure of $\bm{Z}$ in the tails. In fact, following \cite{ferreira2014} the distribution function of $\bm{Z}$ is given as
-\begin{equation*}
-    \prob{Z_1\leq z_1, \ldots ,Z_d\leq z_d} = \expect{\bigvee_{i=1}^d
-\frac{V_i}{z_i\wedge 1}} - \expect{\bigvee_{i=1}^d
-\frac{V_i}{z_i}}
-\end{equation*}
-where the expectation is taken with respect to $\Phi$.
-In view of this, to obtain a PoT model for $\bm{Z}$, we seek 
-a flexible distribution on $\bm{V} \in {\mathbb S}_{\infty}^{d-1}$.
+$H$ is the multivariate Pareto distribution. \cite{RoSeWa2018a}
+provide a number of stochastic representations for $H$. In particular
+Remark 1 justifies the representation given in \cite{ferreira2014},
+consisting of taking $\bm{W}$, in the limit, as $\bm{W} = \bm{V}R$ 
+where $R$ and $\bm{V}$ are independent. $R = ||\bm{W}||_\infty$ is 
+distributed as a standard Pareto  random variable, and $\bm{V} = 
+\bm{W}/||\bm{W}||_\infty$ is a random vector in  
+$\mathbb{S}_{\infty}^{d-1}$, the positive orthant of the 
+unit  sphere in infinite norm, with distribution $\Phi$. 
+%More  specifically, 
+%using the notation in Theorem 6 of \cite{RoSeWa2018a}, define $\bm{S} 
+%=  \log(\bm{V})$.  Then, according to Definition 5, $\bm{S}$ 
+%is a spectral  random vector. Taking  $E \sim \text{Exp}(1)$, 
+%independent of  $\bm{V}$,  we have that $\bm{Y} = \bm{S} + E$  
+%follows a multivariate GP, according  to Theorem 6.  Then, using 
+%Proposition 4,  $\bm{V}R = e^{\bm{Y}}$ will also follow
+%a multivariate GP. Notice that $R = e^E$ is distributed as 
+%a standard Pareto random variable. 
+$R$ and $\bm{V}$ are referred, respectively, as the {\em radial} 
+and {\em angular}  components of $H$. The angular measure controls 
+the dependence  structure of $\bm{W}$  in the tails. In view of 
+this, to obtain a  PoT model we seek a flexible model for the 
+distribution of $\bm{V} \in {\mathbb S}_{\infty}^{d-1}$.
 
-\cite{ferreira2014} refer to the model defined in Equation \eqref{MGP} as a simple 
-Pareto process indexed on a finite dimensional set. An alternative approach to 
-define such distribution, that provides a link to the multivariate GEV, is discussed 
-in Remark 1 of \cite{RoSeWa2018a}. More specifically, 
-using the notation in Theorem 6 of \cite{RoSeWa2018a}, define $\bm{S} = \log(\bm{V})$. 
-Then, according to Definition 5, $\bm{S}$ is a spectral random vector. Taking 
-$E \sim \text{Exp}(1)$, independent of $\bm{V}$, we have that $\bm{Y} = \bm{S} + E$ 
-follows a multivariate GP, according to Theorem 6.  Then, using Proposition 4, 
-$\bm{Z} = e^{Y}$ is also a multivariate GP. Notice that $R = e^E$ is distributed as 
-a standard Pareto random variable.
-
-%\subsection{Pairwise Extremal Dependence\label{subsec:ped}}
+%\subection{Pairwise Extremal Dependence\label{subsec:ped}}
 A measure that is used to characterize the strength of the dependence,
-in the tail, for two random variables $W_1$ and $W_2$, with marginal
+in the tail, for two random variables $Z_1$ and $Z_2$, with marginal
 distributions $F_1$ and $F_2$ is given by
-\[	\chi_{12} = \lim_{u\uparrow 1} \prob{F_1(W_1)>u|F_2(W_2)>u}  \]
+\[	\chi_{12} = \lim_{u\uparrow 1} \prob{F_1(Z_1)>u|F_2(Z_2)>u}  \]
 \citep{coles2001}.
 $\chi_{12}$ provides information about the distribution of extremes for the variable $W_1$
-given that $W_2$ is very large.  When $\chi_{12}>0$, $W_1$ and $W_2$ 
+given that $W_2$ is very large.  When $\chi_{12}>0$, $Z_1$ and $Z_2$ 
 are said to
 be asymptotically dependent, otherwise they are asymptotically
-independent. The following result provides the asymptotic dependence coefficient for two components of $\bm{Z}$. 
+independent. The following result provides the asymptotic dependence coefficient between two components of $\bm{Z}$
+after PoT limit. 
 \begin{prop}\label{ppchi}
 Suppose that $\bm{Z} = R\bm{V}$ with $R\sim Pa(1)$,
-$\prob{V_\ell > 0} = 1$ and $\expect{V_\ell}$ exists ,
-$\ell=1, \ldots ,d$, then
+$\prob{V_\ell > 0} = 1$ and $\expect{V_\ell}$ exists, for $\ell=1, \ldots ,d$, then
 \begin{equation}
@@
 %\subsection{Multivariate Conditional Survival Function\label{subsec:condsurv}}
+Equation \ref{eqn:chi_ij} implies that $\chi_{\jmath\ell}>0$, and so, no asymptotic independence is possible under our proposed model.
 For the analysis of extreme values it is of interest to calculate the multivariate conditional survival function.

### 2022-05-27 main.tex
update 1
Updates from Overleaf
bibliography
Bruno's edits
updates
update
sdf
sfdf
pgpareto github

@@
-\documentclass{article}
-\usepackage[margin=1in]{geometry}
+\documentclass[12pt,authoryear]{elsarticle}
+%\documentclass[12pt]{article}
+% \usepackage[margin=1in]{geometry}
+% \usepackage[margin=0.5in]{caption}
 \usepackage{amsmath}
 \usepackage{amssymb}
+%\usepackage{natbib}
+\usepackage{amsthm}
+\usepackage{hyperref}
+%\usepackage{graphicx}
+\usepackage{xcolor}
+\usepackage{booktabs}
+\usepackage{soul}
+\usepackage[ruled]{algorithm2e}
 \usepackage{bm}
+\usepackage{subcaption}
+\usepackage{wrapfig}
+% \usepackage{setspace}
+% \linespread{1.8}
+\usepackage{anysize}
+\marginsize{1in}{1in}{1in}{1in}
 
-\title{}
-\author{}
-\institution{}
+\newtheorem{prop}{Proposition}
+\DeclareMathOperator{\tr}{Tr}
+\DeclareMathOperator*{\argmin}{arg\,min}
+\DeclareMathOperator*{\argmax}{arg\,max}
+\newcommand{\bigmid}{\,\middle\vert\,}
+% \newcommand{\bruno}[1]{\textcolor{blue}{#1}} % Bruno Sanso edits - blue
+% \newcommand{\makenote}[1]{\textcolor{red}{#1}} % Peter edits - red
+
+\newcommand{\findcite}{\textcolor{red}{[Find Citation]}}
+\newcommand{\needcite}{\findcite}
+\newcommand{\Chi}{\mbox{\Large$\chi$}}
+\newcommand{\norm}[1]{\left\lVert #1 \right\rVert}
+\newcommand{\inorm}[1]{\norm{#1}_{\infty}}
+\newcommand{\pnorm}[2]{\norm{#1}_{#2}}
+\newcommand{\prob}[1]{\text{Pr}\left[#1\right]}
+\newcommand{\expect}[1]{\text{E}\left[#1\right]}
+
+\SetAlCapNameFnt{\footnotesize}
+\SetAlCapFnt{\footnotesize}
+\SetAlCapSty{}
 
@@
 
+\begin{frontmatter}
+    %\title{Inference for multivariate peaks over threshold models}
+    \title{Bayesian non-parametric inference for multivariate peaks-over-threshold models}
+    \author{Peter Trubey\corref{cor1}}
+    \ead{ptrubey@ucsc.edu}
+
+    \author{Bruno Sans\'o}
+    \ead{bruno@soe.ucsc.edu}
+    \address{Department of Statistics, University of California, Santa Cruz,  USA}
+    \cortext[cor1]{Corresponding author}
+
+    \begin{abstract}
+        \input{abstract}
+    \end{abstract}
+
+    \begin{keyword}
+        Multivariate extremes \sep Peak over threshold models \sep
+        Bayesian non-parametric models \sep Dirichlet process mixtures
+    \end{keyword}
+\end{frontmatter}
+
+%--------content
+\input{introduction}
+
+\input{methodology}
+
+\input{evaluation}
+
+\input{results}
+
+\input{conclusion}
+
+\section*{Acknowledgements}
+This material is based upon work supported by the National Science Foundation under Grant No. SES-2050012 and DMS-2153277.
+
+\bibliographystyle{elsarticle-harv}
+\bibliography{./refs}
+
 \end{document}
+
 % EOF

### 2022-05-27 methodology_pnorm.tex
updates
pgpareto github

@@
 \subsection{Projection onto an arbitrary unit hypersphere\label{subsec:projection}}
-
-%\begin{figure}[ht]
-%    \centering
-%    \includegraphics{images/p_sphere}
-%    \caption{Unit hyperspheres established on $\mathcal{L}_p$, for $p = 1,\ldots,10$, for two dimensions.\label{fig:psphere}}
-%\end{figure}
 
@@
       = \lim\limits_{p\to\infty} \lVert \bm{s} \rVert_p
-      = \max_{\ell\in\lbrace1,\ldots,d\rbrace}s_{\ell}.
+      = \underset{\ell = 1}{\overset{d}{\vee}}s_{\ell}.
   \end{equation*}
@@
   Observations on one hypersphere can be projected without loss of information onto another by
-  dividing by the defining norm of the target hypersphere. In fact, letting 
+  dividing by the defining norm of the target hypersphere. In fact, for finite $p$, letting 
   \begin{equation*}
-    y_d = \left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}.
+    y_d = \left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}},
   \end{equation*}
@@
   \end{equation}
-  Equations (\ref{eqn:pnormt}) -- (\ref{eqn:pnormjac}) can be used to define a density on ${\mathbb R}^p_+$, and then obtain a density on ${\mathbb S}_{p}^{d-1}$, for any finite $p>0$.
-Unfortunately in the limit, when $p\rightarrow +\infty$, the transformation is not differentiable.
-Thus, even though the projected distribution on ${\mathbb S}_\infty^{d-1}$ is defined, there is no corresponding density.
+  Equations (\ref{eqn:pnormt}) -- (\ref{eqn:pnormjac}) can be used to define a density on ${\mathbb R}^d_+$, and then obtain a density on ${\mathbb S}_{p}^{d-1}$, for any finite $p>0$. 
+  %Unfortunately in the limit, when $p\rightarrow +\infty$, the transformation is not differentiable.
+Thus, even though the projected distribution on ${\mathbb S}_\infty^{d-1}$ is defined, there is no corresponding density \temp{when the distribution is projected in this manner.
+However, for large but finite $p$, $\mathbb{S}_p^{d-1}$ will tend to approach $\mathbb{S}_{\infty}^{d-1}$}.
 
+%   \temp{
+%   We might also ask what degree of distortion in the distribution is imposed by the difference in projections
+%   from $\mathbb{S}_{\infty}^{d-1}$ to $\mathbb{S}_p^{d-1}$.
+Ideally, we might consider the ratio of surface areas between the two surfaces.
+As there is no general closed form expression for the surface area of a unit $n$-sphere under the $\mathcal{L}_p$ norm, it is difficult to assess in this manner.
+There is, however, a closed-form expression for the \emph{volume} of the positive orthant of the unit $n$-sphere under the $\mathcal{L}_p$ norm.
+This is not directly comparable, however, the closer the volume is to 1, the closer ${\mathbb S}_{p}^{d-1}$ is to ${\mathbb S}_{\infty}^{d-1}$.
+%     \[
+%         \text{Volume}~\left({\mathbb S}_{p}^{d-1}\right) = \frac{\Gamma\left(\frac{1}{p} + 1\right)^{d}}{\Gamma\left(\frac{d}{p} + 1\right)}
+%     \]
+%     }
 % EOF

### 2022-05-27 methodology_projgamma.tex
pgpareto github

@@
 \subsection{Projected gamma family\label{subsec:projgamma}}
-  A natural  distribution to consider in ${\mathbb R}^p_+$ is given by a product of independent
+  A natural  distribution to consider in ${\mathbb R}^d_+$ is given by a product of independent
   univariate Gamma distributions. Let
@@
 Following a Bayesian non-parametric approach \citep{Ferguson74,Antoniak1974} we assume that $G$ is drawn from a random measure.
-In particular, assuming a Dirichlet process prior, we have a hierarchical formulation of the mixture model that, for a vector of observations $\bm{y}_i$, is given by
+In particular, assuming a Dirichlet process prior for $G$, we have a hierarchical formulation of the mixture model that, for a vector of observations $\bm{y}_i$, is given by
   \begin{equation}\label{eqn:dppg}    \bm{y}_i \sim \text{PG}(\bm{y}_i|\bm{\theta}_i) , \;\;\; \bm{\theta}_i \sim G, \;\;\; G \sim \mathcal{DP}(\eta, G_0),
@@
 
+  Unfortunately, in the limit when $p\rightarrow \infty$, the 
+  projection transformation is not differentiable. Thus, a simple
+  formula like Equation \ref{eqn:projgamma} for the
+  projected gamma density on ${\mathbb S}_\infty^{d-1}$ is not 
+  available.
 Our strategy consists of describing the angular distribution $\Phi$ using a sample based approach with the following steps: (i) Apply the transformation in Equation (\ref{eqn:standardization}) to the original data; (ii) Obtain the subsample of the standardized observations that satisfy $R>1$; (iii) Take a finite $p$ and project the observations onto ${\mathbb S}_p^{d-1}$; (iv) Fit the model in Equation (\ref{eqn:PGmix}) to the resulting data and obtain samples from the fitted model; (v) Project the resulting samples onto ${\mathbb S}_\infty^{d-1}$.
@@
 \subsubsection{Inference for the projected Gamma mixture model}
-The Dirichlet process mixture model groups observations into stochastically assigned clusters.
-Building out the methods of inference for Equation~\ref{eqn:dppg}, let $n_j^{-i}$ be the number of observations in cluster $j$ not including observation $i$.
-Let $J^{-i}$ be the number of extant
-% {\bf (remaining?)} observations
-clusters, not including any singleton that observation $i$ is in. Under this model, the probability of cluster membership for a given observation is proportional to
+In a sample-based inference approach, for a given iteration the Dirichlet process mixture model groups observations
+    into stochastically assigned clusters, where members of a cluster share distributional parameters. Building out 
+    the methods of inference for Equation~\ref{eqn:dppg}, 
+    let $n_j^{-i}$ be the number of observations in cluster $j$ not including observation $i$.  Let $J^{-i}$ be 
+    the number of extant clusters, not including any singleton that observation $i$ is in. Under this model, 
+    the probability of cluster membership for a given observation is proportional to
 \begin{equation*}
@@
 \end{equation*}
-If $G_0$ is not a conjugate prior for the kernel, the integral in the above formula may not be
-available in closed form. Using Algorithm 8 from \cite{neal2000}, by Monte Carlo integration, drawing  $m$ candidate
-clusters, $\bm{\alpha}_j,\bm{\beta}_j$ for $j = J^{-i} + 1,\ldots, J^{-i} + m$ from $G_0$. 
+Note the top case is iterating over extant clusters, and the bottom case is for a \emph{new} cluster. 
+If $G_0$ is not a conjugate prior for the kernel density, the integral in the above formula may not be
+available in closed form. We solve this by using Algorithm 8 from \cite{neal2000}: by Monte Carlo integration, 
+we draw $m$ candidate clusters, $\bm{\alpha}_j,\bm{\beta}_j$ for $j = J^{-i} + 1,\ldots, J^{-i} + m$ from $G_0$. 
 Then, we sample the cluster indicator $\delta_i$ from extant or candidate clusters, where
@@
 
-A key feature of the the projected Gamma distribution is its computational properties.
-We augment  $\mathcal{PG}(\bm{y}_i\mid\bm{\alpha}_i,\bm{\beta}_i) $ by introducing a latent radial component $r_i$, for each observation.
-Using Equation \eqref{joint} we observe that the
+A key feature of the the projected Gamma distribution is its computational properties. We augment 
+$\mathcal{PG}(\bm{y}_i\mid\bm{\alpha}_i,\bm{\beta}_i) $ by introducing a latent radial component $r_i$, 
+for each observation. Using Equation \eqref{joint} we observe that the 
 full conditional of $r_i$ is easy to sample from, as it is given as
 \begin{equation}
-    r_i\mid\bm{\alpha}_i,\bm{\beta}_i \sim \mathcal{G}\left(\sum_{\ell = 1}^d\alpha_{i\ell}, \sum_{\ell = 1}^d\beta_{\ell} y_{i\ell}\right).
+    r_i\mid\bm{\alpha}_i,\bm{\beta}_i \sim \mathcal{G}\left(r_i \bigmid \sum_{\ell = 1}^d\alpha_{i\ell}, \sum_{\ell = 1}^d\beta_{\ell} y_{i\ell}\right).
 \end{equation}
@@
 \end{equation}
+Note that the ordering of the products can be reversed in the likelihood, indicating that given $\bm{r}$, 
+the likelihood becomes separable by dimension.
 We first consider a centering distribution given by a product of independent Gammas:
@@
 \end{equation}
-for $\ell = 2,\ldots,d$.  For $\ell = 1$, as $\beta_{1} := 1$, the full conditional takes the form
+for $\ell = 2,\ldots,d$.  For $\ell = 1$, as $\beta_{1} := 1$, the full conditional takes the simpler form
 \begin{equation}
     \label{eqn:alpha1update}
-    \pi(\alpha_{j1}\mid\bm{r},\bm{Y},\bm{\gamma},\xi_1,\tau_1) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}\alpha_{j1}^{\zeta_1 - 1}e^{-\tau_1\alpha_{j1}}}{\Gamma^{n_j}(\alpha_{j1})}.
+    \pi(\alpha_{j1}\mid\bm{r},\bm{Y},\bm{\gamma},\xi_1,\tau_1) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}\alpha_{j1}^{\xi_1 - 1}e^{-\tau_1\alpha_{j1}}}{\Gamma^{n_j}(\alpha_{j1})}.
 \end{equation}
@@
     \label{eqn:betafc}
-    \beta_{j\ell}\mid\bm{r},\bm{Y},\alpha,\zeta_{\ell},\sigma_{\ell} \sim \mathcal{G}\left(\beta_{j\ell}\mid n_j\alpha_{j\ell} + \zeta_\ell, \sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right),
+    \beta_{j\ell}\mid\bm{r},\bm{Y},\alpha,\zeta_{\ell},\sigma_{\ell} \sim \mathcal{G}\left(\beta_{j\ell}\bigmid n_j\alpha_{j\ell} + \zeta_\ell, \sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right),
 \end{equation}

### 2022-05-27 results.tex
pgpareto github

@@
 % \end{figure}
-\begin{figure}[tb]
+
+\begin{figure}[htb]
     \centering
-    \caption{Energy score (top) and Posterior Predictive Loss (bottom) for various models on simulated data with ascending count of mixture components (indicated by plot heading).
+    \caption{Energy score (top row) and Posterior Predictive Loss (bottom row) for various models on simulated data with ascending count of mixture components (indicated by plot heading) and number of dimensions (indicated by horizontal axis).
 Gen indicates baseline.\label{fig:simpples}}
@@
 To evaluate our proposed approach for angular measure estimation 
-we consider simulated datasets on $\mathcal{S}_{\infty}^{d-1}$, for 
+we consider simulated datasets on $\mathbb{S}_{\infty}^{d-1}$, for 
 $d$ ranging between 3 and 20. We generated each dataset from a mixture of projected Gammas, with the number of mixture components
@@
   \begin{algorithm}[ht]
-    \small 
+    \footnotesize
     \caption{Simulated Angular Dataset Generation Routine\label{algo:simulated}}
@@
   Gamma and the other gamma-based models tends to shrink as the number of dimensions increases.
-  Alternatively, that difference appears to grow as the number of mixture components increases. The results from both scoring criteria are pretty comparable.
+  Alternatively, that difference appears to grow as the number of mixture components increases. 
+  The results from both scoring criteria are comparable.
 
@@
 \subsection{Integrated Vapor Transport\label{subsec:ivt}}
-
-\begin{figure}[tb]
-    \begin{center}
-    \caption{Grid cell locations for ERA-Interim (left) and ERA5 (right).\label{fig:gridlocs}}
-    \begin{subfigure}[b]{0.48\textwidth}
-        \centering 
-        \includegraphics[height=2.5in]{./images/erai_grid}
-    \end{subfigure}
-    %
-    \begin{subfigure}[b]{0.48\textwidth}
-        \centering
-        \includegraphics[height=2.5in]{./images/era5_grid}
-    \end{subfigure}
-    \end{center}
-\end{figure}
-
 The \emph{integrated vapor transport} (IVT) is a two component vector 
@@
 
+\begin{figure}[tb]
+    \begin{center}
+    \caption{Grid cell locations for ERA-Interim (left) and ERA5 (right).\label{fig:gridlocs}}
+    \begin{subfigure}[b]{0.48\textwidth}
+        \centering 
+        \includegraphics[height=2.5in]{./images/erai_grid}
+    \end{subfigure}
+    %
+    \begin{subfigure}[b]{0.48\textwidth}
+        \centering
+        \includegraphics[height=2.5in]{./images/era5_grid}
+    \end{subfigure}
+    \end{center}
+\end{figure}
+
 Our data corresponds to daily average values for the IVT magnitude
@@
 \begin{algorithm}[htb]
-    \small
+    \footnotesize
+   \caption{Data preprocessing to isolate and transform data exhibiting extreme behavior.  $r_i$
+   represents the radial component, and $\bm{v}_i$ the angular component.  The declustering
+   portion is relevant for data correlated in time.\label{algo:processing}}
   \KwResult{$\bm{ r},\bm{v} : r_i \sim \text{Pareto}(1)$, $\bm{ v}_i \in {\mathbb S}_{\infty}^{d-1}$}
@@
     Set $b_{t,\ell} = \hat{F}_{\ell}^{-1}\left(1 - \frac{1}{t}\right)$.\\
-    With $\bm{ x}_{\ell} > b_{t,\ell}$, fit $a_{\ell}$, $\chi_{\ell}$ via MLE according to generalized Pareto likelihood.\\
+    With $\bm{ x}_{\ell} > b_{t,\ell}$, fit $a_{\ell}$, $\xi_{\ell}$ via MLE according to generalized Pareto likelihood.\\
     }
@@
   }
- \caption{Data preprocessing to isolate and transform data exhibiting extreme behavior.  $r_i$
-   represents the radial component, and $\bm{v}_i$ the angular component.  The declustering
-   portion is relevant for data correlated in time.\label{algo:processing}}
 \end{algorithm}
@@
     \centering
-    \caption{Conditional survival curves for selected locations, conditioning on all other dimensions at greater than 90th percentile (fitted)\label{fig:condsurv1d}.
+    \caption{Conditional survival curves for selected locations, using ERA-Interim, and PRG-G model,  conditioning on all other dimensions at greater than 90th percentile (fitted)\label{fig:condsurv1d}.
 The left panel uses original units.
@@
     \centering
-    \caption{Pairwise conditional survival curves for selected locations, conditioning on all other dimensions at greater than 90th percentile (fitted)\label{fig:condsurv2d}}
+    \caption{Pairwise conditional survival curves for selected locations, using ERA-Interim, and PRG-G model, conditioning on all other dimensions at greater than 90th percentile (fitted).\label{fig:condsurv2d}}
     \includegraphics[height=5in, width=5in]{./images/condsurv_2d}

### 2022-06-15 abstract.tex
pgpareto github
first pass at springer-nature class

@@
+% We consider a constructive definition of the multivariate Pareto that factorizes the 
+% random vector into a radial component and an independent angular component; The former
+% following a univariate Pareto distribution, and the latter defined on the surface of the 
+% positive orthant of the unit hypercube.  In this paper, we  propose a method for
+% inferring the distribution of the angular component.  We identify its support 
+% as the limit of the positive orthants of the unit $p$--norm spheres,
+% and introduce a projected gamma family of
+% distributions defined as the projection of a vector of independent gamma random variables
+% onto the $p$--norm sphere.  This family serves as a building block for a flexible family
+% of distributions obtained as a Dirichlet process mixture of projected gammas.  For 
+% model assessment and comparison, we discuss model scoring methods appropriate 
+% to distributions on the unit hypercube.  In particular, working with the energy score
+% criterion, we develop a kernel metric appropriate to the hypercube that produces a 
+% proper scoring rule.  We then present a simulation study to compare 
+% different modeling choices using the proposed scoring rules.  Finally we apply our 
+% approach to describe the dependence structure of the extreme values of the magnitude 
+% of the integrated vapor  transport (IVT), a variable that describes the rate of flow 
+% of moisture in the atmosphere  along the coast of California for the years of 1979 
+% through 2020.  We find a clear but heterogeneous geographical dependence.
+
 We consider a constructive definition of the multivariate Pareto that factorizes the 
@@
 inferring the distribution of the angular component.  We identify its support 
-as the limit of the positive orthants of the unit $p$--norm spheres.
-To this effect, we introduce a projected gamma family of
+as the limit of the positive orthant of the unit $p$--norm spheres,
+and introduce a projected gamma family of
 distributions defined as the projection of a vector of independent gamma random variables
@@
 of distributions obtained as a Dirichlet process mixture of projected gammas.  For 
-model assessment and comparison, we discuss model scoring methods appropriate 
+model assessment, we discuss scoring methods appropriate 
 to distributions on the unit hypercube.  In particular, working with the energy score
-criterion, we develop a kernel metric appropriate to the hypercube that produces a 
-proper scoring rule.  We then present a simulation study to compare 
+criterion, we develop a kernel metric that produces a 
+proper scoring rule, and present a simulation study to compare 
 different modeling choices using the proposed scoring rules.  Finally we apply our 
 approach to describe the dependence structure of the extreme values of the magnitude 
-of the integrated vapor  transport (IVT), a variable that describes the rate of flow 
-of moisture in the atmosphere  along the coast of California for the years of 1979 
-through 2020.  We find a clear but heterogeneous geographical dependence.
-%We present the results 
-%of this analysis to describe the dependence structure of extreme events in IVT using 
-%pairwise extremal dependence coefficients---a summary measure of jointly extreme behavior.
-%We also present conditional survival curves---the survival function for some dimensions
-%conditional on other dimensions.
+of the integrated vapor transport (IVT), data describing the flow 
+of atmospheric moisture along the coast of California for the years of 1979 
+through 2020.  We find clear but heterogeneous geographical dependence.
 

### 2022-06-15 introduction.tex
first pass at springer-nature class

@@
 It consists of decomposing the process into independent radial and angular components.
-Such approach can be used in the finite dimensional case, where the angular component
+Such an approach can be used in the finite dimensional case, where the angular component
   contains the information pertaining to the dependence structure of the random vector.
@@
 \citep{coles2001}.
-$\chi_{12}$ provides information about the distribution of extremes for the variable $W_1$
-given that $W_2$ is very large.  When $\chi_{12}>0$, $Z_1$ and $Z_2$ 
+$\chi_{12}$ provides information about the distribution of extremes for the variable $Z_1$
+given that $Z_2$ is very large.  When $\chi_{12}>0$, $Z_1$ and $Z_2$ 
 are said to

### 2022-06-15 main.tex
first pass at springer-nature class

@@
-\documentclass[12pt,authoryear]{elsarticle}
+\documentclass[sn-basic]{sn-jnl}
+%\documentclass[12pt,authoryear]{elsarticle}
 %\documentclass[12pt]{article}
@@
 %\usepackage{graphicx}
-\usepackage{xcolor}
+%\usepackage{xcolor}
 \usepackage{booktabs}
@@
 
-\SetAlCapNameFnt{\footnotesize}
-\SetAlCapFnt{\footnotesize}
-\SetAlCapSty{}
+% \SetAlCapNameFnt{\footnotesize}
+% \SetAlCapFnt{\footnotesize}
+% \SetAlCapSty{}
+
+\jyear{2022}
+\theoremstyle{thmstyleone}
+
+\raggedbottom
 
@@
 
-\begin{frontmatter}
-    %\title{Inference for multivariate peaks over threshold models}
-    \title{Bayesian non-parametric inference for multivariate peaks-over-threshold models}
-    \author{Peter Trubey\corref{cor1}}
-    \ead{ptrubey@ucsc.edu}
+\title[Bayesian non-parametric inference for multivariate PoT models]{Bayesian non-parametric inference for multivariate peaks-over-threshold models}
 
-    \author{Bruno Sans\'o}
-    \ead{bruno@soe.ucsc.edu}
-    \address{Department of Statistics, University of California, Santa Cruz,  USA}
-    \cortext[cor1]{Corresponding author}
+\author*[1]{\fnm{Peter} \sur{Trubey}}\email{ptrubey@ucsc.edu}
+\author[1]{\fnm{Bruno} \sur{Sans\'{o}}}\email{bruno@soe.ucsc.edu}
 
-    \begin{abstract}
-        \input{abstract}
-    \end{abstract}
+\affil*[1]{\orgdiv{Department of Statistics}, \orgname{University of California Santa Cruz}, 
+            \orgaddress{\street{1156 High Street}, \city{Santa Cruz},\postcode{95064}, \state{CA}, \country{USA}}}
 
-    \begin{keyword}
-        Multivariate extremes \sep Peak over threshold models \sep
-        Bayesian non-parametric models \sep Dirichlet process mixtures
-    \end{keyword}
-\end{frontmatter}
+\abstract{\input{./abstract}}
+\keywords{Multivariate extremes, Peak over threshold models,
+                Bayesian non-parametric models, Dirichlet process mixtures}
+
+% \Author{Peter Trubey\Affil{1} and Bruno Sans\'o\Affil{1}}
+% \AuthorRunning{Peter Trubey and Bruno Sans\'o}
+
+% \Affiliations{
+% \item Department of Statistics,
+%       University Of California Santa Cruz,
+%       USA
+% }
+
+% \CorrAddress{Peter Trubey,
+%             Department of Statistics,
+%             University of California Santa Cruz,
+%             1156 High Street, 
+%             Santa Cruz, CA 95064,
+%             USA}
+% \CorrEmail{ptrubey@ucsc.edu}
+% \CorrPhone{831-459-1448}
+% \CorrFax{831-459-4829}
+
+% \Title{Bayesian non-parametric inference for multivariate peaks-over-threshold models}
+% \TitleRunning{Bayesian non-parametric inference for multivariate PoT models}
+
+% \Abstract{
+%     \input{./abstract}
+% }
+
+% \Keywords{
+%  Multivariate extremes; Peak over threshold models; 
+%  Bayesian non-parametric models; Dirichlet process mixtures
+% }
+
+\maketitle
+https://www.overleaf.com/project/612e69c665059f5fd7998d90
+% \begin{frontmatter}
+%     %\title{Inference for multivariate peaks over threshold models}
+%     \title{Bayesian non-parametric inference for multivariate peaks-over-threshold models}
+%     \author{Peter Trubey\corref{cor1}}
+%     \ead{ptrubey@ucsc.edu}
+
+%     \author{Bruno Sans\'o}
+%     \ead{bruno@soe.ucsc.edu}
+%     \address{Department of Statistics, University of California, Santa Cruz,  USA}
+%     \cortext[cor1]{Corresponding author}
+
+%     \begin{abstract}
+%         \input{abstract}
+%     \end{abstract}
+
+%     \begin{keyword}
+%         Multivariate extremes \sep Peak over threshold models \sep
+%         Bayesian non-parametric models \sep Dirichlet process mixtures
+%     \end{keyword}
+% \end{frontmatter}
 
 %--------content
-\input{introduction}
+\input{./introduction}
 
-\input{methodology}
+\input{./methodology}
 
-\input{evaluation}
+\input{./evaluation}
 
-\input{results}
+\input{./results}
 
-\input{conclusion}
+\input{./conclusion}
 
@@
 
-\bibliographystyle{elsarticle-harv}
+%\bibliographystyle{elsarticle-harv}
 \bibliography{./refs}

### 2022-06-15 methodology_projgamma.tex
first pass at springer-nature class

@@
   Unfortunately, in the limit when $p\rightarrow \infty$, the 
-  projection transformation is not differentiable. Thus, a simple
-  formula like Equation \ref{eqn:projgamma} for the
-  projected gamma density on ${\mathbb S}_\infty^{d-1}$ is not 
-  available.
-Our strategy consists of describing the angular distribution $\Phi$ using a sample based approach with the following steps: (i) Apply the transformation in Equation (\ref{eqn:standardization}) to the original data; (ii) Obtain the subsample of the standardized observations that satisfy $R>1$; (iii) Take a finite $p$ and project the observations onto ${\mathbb S}_p^{d-1}$; (iv) Fit the model in Equation (\ref{eqn:PGmix}) to the resulting data and obtain samples from the fitted model; (v) Project the resulting samples onto ${\mathbb S}_\infty^{d-1}$.
-  For step (iv) we use a Bayesian approach that is implemented using a purposely developed Markov chain Monte Carlo that is described in the next section.
+  projection transformation is not differentiable. Thus, a closed form expression 
+  like Equation (\ref{eqn:projgamma}) for the projected gamma density on ${\mathbb S}_\infty^{d-1}$ is not 
+  available. Instead, we observe that for a sufficiently large $p$, $\mathbb{S}_p^{d-1}$ will approach
+  $\mathbb{S}_{\infty}^{d-1}$.  With that in mind, our strategy consists of describing the angular
+  distribution $\Phi$ using a sample based approach with the following steps: (i) Apply the 
+  transformation in Equation (\ref{eqn:standardization}) to the original data; (ii) Obtain the 
+  subsample of the standardized observations that satisfy $R>1$; (iii) Take a finite $p$ and 
+  project the observations onto ${\mathbb S}_p^{d-1}$; 
+  (iv) Fit the model in Equation (\ref{eqn:PGmix}) to the resulting 
+  data and obtain samples from the fitted model; 
+  (v) Project the resulting samples onto ${\mathbb S}_\infty^{d-1}$.
+  For step (iv) we use a Bayesian approach that is implemented using a purposely developed Markov chain Monte Carlo sampler described in the next section.
 
@@
     \caption{The positive orthant of the $p$-norm sphere.}
-  \begin{subfigure}[b]{0.47\textwidth}
+  \begin{subfigure}[b]{0.4\textwidth}
     \includegraphics[width=\textwidth]{./images/p_sphere}
@@
   %
-  \begin{subfigure}[b]{0.47\textwidth}
+  \begin{subfigure}[b]{0.4\textwidth}
     \includegraphics[width=\textwidth]{./images/p_project}
@@
     into stochastically assigned clusters, where members of a cluster share distributional parameters. Building out 
-    the methods of inference for Equation~\ref{eqn:dppg}, 
+    the methods of inference for Equation~(\ref{eqn:dppg}), 
     let $n_j^{-i}$ be the number of observations in cluster $j$ not including observation $i$.  Let $J^{-i}$ be 
@@
 If $G_0$ is not a conjugate prior for the kernel density, the integral in the above formula may not be
-available in closed form. We solve this by using Algorithm 8 from \cite{neal2000}: by Monte Carlo integration, 
+available in closed form. We sidestep this using Algorithm 8 from \cite{neal2000}: by Monte Carlo integration, 
 we draw $m$ candidate clusters, $\bm{\alpha}_j,\bm{\beta}_j$ for $j = J^{-i} + 1,\ldots, J^{-i} + m$ from $G_0$. 
-Then, we sample the cluster indicator $\delta_i$ from extant or candidate clusters, where
+Then, we sample the cluster indicator $\gamma_i$ from extant or candidate clusters, where
 the probability of cluster membership is proportional to
@@
 %{\bf you first have to say that you sample from this distribution. What is the ``candidate cluster''?}
-then $\delta_i = J = J^{- i} + 1$, and the associated cluster parameters are saved.
+then $\gamma_i = J = J^{- i} + 1$, and the associated cluster parameters are saved.
 
@@
 We also explore a restricted form of this model, where $\beta_{\ell} = 1$ for all $\ell$.
-Under this model, we use the full conditional in Equation~\ref{eqn:alpha1update} for all $\ell$, and omit inference on $\bm{\zeta},\bm{\sigma}$.
+Under this model, we use the full conditional in Equation~(\ref{eqn:alpha1update}) for all $\ell$, and omit inference on $\bm{\zeta},\bm{\sigma}$.
 We refer to this model as the \emph{projected restricted Gamma--Gamma} (PRG--G) model.

### 2022-06-18 introduction.tex
statistics and computing submission

@@
 distribution. It follows that
-\[
-    \lim_{n\rightarrow\infty}\prob{\bm{a}^{-1}_n (\bm{W} - \bm{b}_n)
-    \leq \bm{w} | \bm{W} \not\leq \bm{b}_n)} = \frac{\log G(\bm{w} 
-    \wedge \bm{0}) - \log G(\bm{w}) }{\log G(\bm{0})} = H(\bm{w}).
-\]
+\begin{equation*}
+    \begin{aligned}
+    \lim\limits_{n\rightarrow\infty} &\prob{\bm{a}^{-1}_n (\bm{W} - \bm{b}_n) 
+   \leq \bm{w} \mid \bm{W} \not\leq \bm{b}_n)} \\ 
+     &\hspace{0.5cm}= \frac{\log G(\bm{w} \wedge \bm{0}) 
+    - \log G(\bm{w}) }{\log G(\bm{0})} = H(\bm{w}).
+    \end{aligned}
+\end{equation*}
 $H$ is the multivariate Pareto distribution. \cite{RoSeWa2018a}
@@
 in the tail, for two random variables $Z_1$ and $Z_2$, with marginal
-distributions $F_1$ and $F_2$ is given by
-\[	\chi_{12} = \lim_{u\uparrow 1} \prob{F_1(Z_1)>u|F_2(Z_2)>u}  \]
-\citep{coles2001}.
+distributions $F_1$ and $F_2$ is given by \citep{coles2001}.
+\[
+    \chi_{12} = \lim\limits_{u\uparrow 1} \prob{F_1(Z_1)>u\mid F_2(Z_2)>u}  
+\]
 $\chi_{12}$ provides information about the distribution of extremes for the variable $Z_1$
@@
 \begin{equation*}
-\prob{R>\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}}
-    = \expect{1\wedge\left(\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}\right)^{-1}}
-    = \expect{\frac{V_\jmath}{z_\jmath}\wedge\frac{V_\ell}{z_\ell}}
+    \begin{aligned}
+    &\hspace{-0.5cm}\prob{R>\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}}
+    = \expect{1\wedge\left(\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}\right)^{-1}}\\
+    &= \expect{\frac{V_\jmath}{z_\jmath}\wedge\frac{V_\ell}{z_\ell}}
     = (1 - u)\expect{\frac{V_\jmath}{\expect{V_\jmath}}\wedge\frac{V_\ell}{\expect{V_\ell}}},
+    \end{aligned}
 \end{equation*}
@@
 bounded and $z_i\rightarrow\infty$. The proof is completed by noting
-that $\prob{F_i(Z_i)>u} = 1 - u.\hfill\Box$
+that $\prob{F_i(Z_i)>u} = 1 - u. \hfill\Box$
 
@@
 \begin{equation} \label{eqn:condsurv}
-    \mathrm{Pr}\left[\bigcap_{\ell \in \alpha} Z_\ell > z_\ell \right|\left. \bigcap_{\ell\not\in\alpha} Z_\ell > z_\ell\right] =
-      \frac{\expect{\bigwedge_{k = 1}^d 1\wedge \frac{V_k}{z_k}}}{\expect{\bigwedge_{k \not\in\alpha}1\wedge\frac{V_k}{z_k}}}.
+    \begin{aligned}
+    &\hspace{-0.5cm}\mathrm{Pr}\left[\bigcap_{\ell \in \alpha} Z_\ell > z_\ell \mid \bigcap_{\ell\not\in\alpha} Z_\ell > z_\ell\right] \\
+    &= \frac{\expect{\bigwedge_{k = 1}^d 1\wedge \frac{V_k}{z_k}}}{\expect{\bigwedge_{k \not\in\alpha}1\wedge\frac{V_k}{z_k}}}.
+    \end{aligned}
   \end{equation}

### 2022-06-18 main.tex
minor update
statistics and computing submission

--- 2022-05-27 main.tex
@@
-\documentclass[12pt,authoryear]{elsarticle}
-%\documentclass[12pt]{article}
-% \usepackage[margin=1in]{geometry}
-% \usepackage[margin=0.5in]{caption}
+\documentclass[iicol,sn-basic]{sn-jnl}
+
 \usepackage{amsmath}
@@
 %\usepackage{graphicx}
-\usepackage{xcolor}
+%\usepackage{xcolor}
 \usepackage{booktabs}
 \usepackage{soul}
-\usepackage[ruled]{algorithm2e}
+%\usepackage[ruled]{algorithm2e}
 \usepackage{bm}
@@
 \newcommand{\bigmid}{\,\middle\vert\,}
-% \newcommand{\bruno}[1]{\textcolor{blue}{#1}} % Bruno Sanso edits - blue
+% \newcommand{\bruno}[1]{\textcolor{blue}{#1}} % Bruno edits - blue
 % \newcommand{\makenote}[1]{\textcolor{red}{#1}} % Peter edits - red
@@
 \newcommand{\Chi}{\mbox{\Large$\chi$}}
+
 \newcommand{\norm}[1]{\left\lVert #1 \right\rVert}
@@
 
-\SetAlCapNameFnt{\footnotesize}
-\SetAlCapFnt{\footnotesize}
-\SetAlCapSty{}
+\jyear{2022}
+\theoremstyle{thmstyleone}
+
+\raggedbottom
 
@@
 
-\begin{frontmatter}
-    %\title{Inference for multivariate peaks over threshold models}
-    \title{Bayesian non-parametric inference for multivariate peaks-over-threshold models}
-    \author{Peter Trubey\corref{cor1}}
-    \ead{ptrubey@ucsc.edu}
+\title[~\hspace{6.75in}BNP inference for multivariate PoT]{Bayesian non-parametric inference for multivariate peaks-over-threshold models}
 
-    \author{Bruno Sans\'o}
-    \ead{bruno@soe.ucsc.edu}
-    \address{Department of Statistics, University of California, Santa Cruz,  USA}
-    \cortext[cor1]{Corresponding author}
+\author*{\fnm{Peter} \sur{Trubey}}\email{ptrubey@ucsc.edu}
+\author{\fnm{Bruno} \sur{Sans\'{o}}}\email{bruno@soe.ucsc.edu}
 
-    \begin{abstract}
-        \input{abstract}
-    \end{abstract}
+\affil{\orgdiv{Department of Statistics}, \orgname{University of California Santa Cruz}, 
+            \orgaddress{\street{1156 High Street}, \city{Santa Cruz},\postcode{95064}, \state{CA}, \country{USA}}}
 
-    \begin{keyword}
-        Multivariate extremes \sep Peak over threshold models \sep
-        Bayesian non-parametric models \sep Dirichlet process mixtures
-    \end{keyword}
-\end{frontmatter}
+\abstract{\input{./abstract}}
+\keywords{Multivariate extremes, Peak over threshold models,
+                Bayesian non-parametric models, Dirichlet process mixtures}
+
+\maketitle
 
 %--------content
-\input{introduction}
+\input{./introduction}
 
-\input{methodology}
+\input{./methodology}
 
-\input{evaluation}
+\input{./evaluation}
 
-\input{results}
+\input{./results}
 
-\input{conclusion}
+\input{./conclusion}
 
 \section*{Acknowledgements}
-This material is based upon work supported by the National Science Foundation under Grant No. SES-2050012 and DMS-2153277.
+This material is based upon work supported by the National Science Foundation
+    under Grant No. SES-2050012 and DMS-2153277.
 
-\bibliographystyle{elsarticle-harv}
 \bibliography{./refs}

### 2022-06-18 methodology_projgamma.tex
statistics and computing submission

@@
   \label{joint}
-    f(r,\bm{ y}) = \prod_{\ell = 1}^{d}
+    \begin{aligned}
+    f(r,\bm{ y}) = &\prod_{\ell = 1}^{d}
       \left[\frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell})}(ry_{\ell})^{\alpha_{\ell} - 1}
-          \exp\lbrace-\beta_{\ell}ry_{\ell}\rbrace\right]
-      \times r^{d-1}\left[y_d +
+          \exp\lbrace-\beta_{\ell}ry_{\ell}\rbrace\right] \\
+      &\hspace{0.5cm}\times r^{d-1}\left[y_d +
             {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right].
+    \end{aligned}
   \end{equation}
@@
     \label{eqn:projgamma}
-    \text{PG}(\bm{ y}\mid\bm{ \alpha},\bm{ \beta}) =
+    \begin{aligned}
+    \text{PG}(\bm{ y}\mid\bm{ \alpha},\bm{ \beta}) &=
           \prod_{\ell = 1}^d\left[\frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell})}
-                y_{\ell}^{\alpha_{\ell} - 1}\right]
-      \times \left[y_d +
-          {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right]
-      \times \frac{\Gamma({\textstyle\sum}_{\ell = 1}^d\alpha_{\ell})}{\left({\textstyle\sum}_{\ell = 1}^d
-                    \beta_{\ell}y_{\ell}\right)^{{\scriptstyle\sum_{\ell = 1}^d \alpha_{\ell}}}} ,
+                y_{\ell}^{\alpha_{\ell} - 1}\right]\\
+      &\hspace{0.5cm}\times \left[y_d +
+          {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right]\\
+      &\hspace{0.5cm}\times \frac{\Gamma({\textstyle\sum}_{\ell = 1}^d\alpha_{\ell})}{\left({\textstyle\sum}_{\ell = 1}^d
+                    \beta_{\ell}y_{\ell}\right)^{{\scriptstyle\sum_{\ell = 1}^d \alpha_{\ell}}}},
+    \end{aligned}
   \end{equation}
@@
   \begin{equation} \label{eqn:PGmix}     
-     f(\bm{y}) = \int_\Theta \mathcal{PG}(\bm{y}|\bm{\theta}) dG(\bm{\theta}), 
+     f(\bm{y}) = \int_\Theta \mathcal{PG}(\bm{y}\mid \bm{\theta}) dG(\bm{\theta}), 
   \end{equation} 
@@
 In particular, assuming a Dirichlet process prior for $G$, we have a hierarchical formulation of the mixture model that, for a vector of observations $\bm{y}_i$, is given by
-  \begin{equation}\label{eqn:dppg}    \bm{y}_i \sim \text{PG}(\bm{y}_i|\bm{\theta}_i) , \;\;\; \bm{\theta}_i \sim G, \;\;\; G \sim \mathcal{DP}(\eta, G_0),
+  \begin{equation}\label{eqn:dppg}
+    \begin{aligned}
+    \bm{y}_i &\sim \text{PG}(\bm{y}_i\mid \bm{\theta}_i)\\
+    \bm{\theta}_i &\sim G
+    \end{aligned}
+    ~\hspace{1cm}
+    \begin{aligned}
+    G &\sim \mathcal{DP}(\eta, G_0),
+    \end{aligned}
   \end{equation}
@@
     \text{Pr}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
-        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
-        \eta\int\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)dG_0(\bm{\alpha}_j,\bm{\beta}_j)\hspace{0.5cm} &\text{~for~}j = J^{-i} + 1.
+        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\\ % \hspace{0.5cm} \\ % &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
+        \eta\int\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)dG_0(\bm{\alpha}_j,\bm{\beta}_j),%\hspace{0.5cm} % &\text{~for~}j = J^{-i} + 1.
         \end{cases}
 \end{equation*}
-Note the top case is iterating over extant clusters, and the bottom case is for a \emph{new} cluster. 
-If $G_0$ is not a conjugate prior for the kernel density, the integral in the above formula may not be
-available in closed form. We sidestep this using Algorithm 8 from \cite{neal2000}: by Monte Carlo integration, 
-we draw $m$ candidate clusters, $\bm{\alpha}_j,\bm{\beta}_j$ for $j = J^{-i} + 1,\ldots, J^{-i} + m$ from $G_0$. 
-Then, we sample the cluster indicator $\gamma_i$ from extant or candidate clusters, where
-the probability of cluster membership is proportional to
+where the top case is iterating over extant clusters $j = 1,\ldots, J^{-i}$, and the bottom case is
+    for a \emph{new} cluster. If $G_0$ is not a conjugate prior for the kernel density, the integral in the 
+    above formula may not be available in closed form. We sidestep this using Algorithm 8 from \cite{neal2000}: 
+    by Monte Carlo integration, we draw $m$ candidate clusters, $\bm{\alpha}_j,\bm{\beta}_j$ for
+    $j = J^{-i} + 1,\ldots, J^{-i} + m$ from $G_0$. Then, we sample the cluster indicator 
+    $\gamma_i$ from extant or candidate clusters, where the probability of cluster membership is proportional to
 \begin{equation}
     \text{Pr}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
-        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
-        \frac{\eta}{m}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace J^{-i} + 1,\ldots, J^{-i} + m\rbrace.
+        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\\ % \hspace{0.5cm}  % &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
+        \frac{\eta}{m}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right). % \hspace{0.5cm} &\text{~for~}j \in \lbrace J^{-i} + 1,\ldots, J^{-i} + m\rbrace.
         \end{cases}
 \end{equation}
-If a candidate cluster is selected,
+Again, the top case is iterating over extant clusters, and now the bottom case is iterating 
+    over new \emph{candidate} clusters.  If a candidate cluster is selected,
 %{\bf you first have to say that you sample from this distribution. What is the ``candidate cluster''?}
@@
 \begin{equation}
-    f(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{Y},\bm{r},\bm{\delta},\ldots) \propto \prod_{i:\gamma_i = j}\prod_{\ell = 1}^d\mathcal{G}\left(r_iy_{i\ell}\mid\alpha_{j\ell},\beta_{j\ell}\right) \times dG_0(\bm{\alpha}_j,\bm{\beta}_j).
+    \begin{aligned}
+    &\hspace{-1cm}f(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{Y},\bm{r},\bm{\delta},\ldots) \\
+    &\propto \prod_{i:\gamma_i = j}\prod_{\ell = 1}^d\mathcal{G}\left(r_iy_{i\ell}\mid\alpha_{j\ell},\beta_{j\ell}\right) \\
+    &\hspace{0.5cm} \times dG_0(\bm{\alpha}_j,\bm{\beta}_j).
+    \end{aligned}
 \end{equation}
@@
 \begin{equation}
-    G_0(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{\xi},\bm{\tau},\bm{\zeta},\bm{\sigma}) = \prod_{\ell = 1}^d\mathcal{G}(\alpha_{j\ell}\mid \xi_{\ell},\tau_{\ell})\times\prod_{\ell = 2}^d\mathcal{G}(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}).
+    \begin{aligned}
+    &\hspace{-0.3cm}G_0(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{\xi},\bm{\tau},\bm{\zeta},\bm{\sigma}) \\
+    &= \prod_{\ell = 1}^d\mathcal{G}(\alpha_{j\ell}\mid \xi_{\ell},\tau_{\ell})\times\prod_{\ell = 2}^d\mathcal{G}(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}).
+    \end{aligned}
 \end{equation}
@@
 \begin{equation}
-    \pi(\alpha_{j\ell}\mid \bm{r},\bm{Y},\bm{\gamma},\xi_\ell,\tau_\ell,\zeta_\ell,\sigma_\ell) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i\ell}\right)^{\alpha_{j\ell} - 1}\alpha_{j\ell}^{\xi_\ell - 1}e^{-\tau_\ell \alpha_{j\ell}}\times\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\Gamma^{n_j}(\alpha_{j\ell})\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
+    \label{eqn:alphalupdate}
+    \begin{aligned}
+    &\hspace{-0.5cm}\pi(\alpha_{j\ell}\mid \bm{r},\bm{Y},\bm{\gamma},\xi_\ell,\tau_\ell,\zeta_\ell,\sigma_\ell) \\
+    &\propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i\ell}\right)^{\alpha_{j\ell} - 1}\alpha_{j\ell}^{\xi_\ell - 1}e^{-\tau_\ell \alpha_{j\ell}}\times}{\Gamma^{n_j}(\alpha_{j\ell})} \\
+    &\hspace{0.5cm}\times \frac{\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
+    \end{aligned}
 \end{equation}
@@
     \label{eqn:alpha1update}
-    \pi(\alpha_{j1}\mid\bm{r},\bm{Y},\bm{\gamma},\xi_1,\tau_1) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}\alpha_{j1}^{\xi_1 - 1}e^{-\tau_1\alpha_{j1}}}{\Gamma^{n_j}(\alpha_{j1})}.
+    \begin{aligned}
+    &\hspace{-0.5cm}\pi(\alpha_{j1}\mid\bm{r},\bm{Y},\bm{\gamma},\xi_1,\tau_1) \\
+    &\propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}\alpha_{j1}^{\xi_1 - 1}e^{-\tau_1\alpha_{j1}}}{\Gamma^{n_j}(\alpha_{j1})}.
+    \end{aligned}
 \end{equation}
@@
     \label{eqn:betafc}
-    \beta_{j\ell}\mid\bm{r},\bm{Y},\alpha,\zeta_{\ell},\sigma_{\ell} \sim \mathcal{G}\left(\beta_{j\ell}\bigmid n_j\alpha_{j\ell} + \zeta_\ell, \sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right),
+    \begin{aligned}
+    &\hspace{-0.5cm}\beta_{j\ell}\mid\bm{r},\bm{Y},\alpha,\zeta_{\ell},\sigma_{\ell} \\
+    &\sim \mathcal{G}\left(\beta_{j\ell}\bigmid n_j\alpha_{j\ell} + \zeta_\ell, \sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right),
+    \end{aligned}
 \end{equation}
@@
 \begin{equation}
-    G_0\left(\bm{\alpha}_j,\bm{\beta}_j\mid\bm{\mu},\bm{\Sigma},\zeta,\sigma\right) = \mathcal{LN}\left(\bm{\alpha}_j\mid\bm{\mu},\bm{\Sigma}\right)\times\prod_{\ell = 2}^d\mathcal{G}\left(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}\right).
+    \begin{aligned}
+    &\hspace{-0.5cm}G_0\left(\bm{\alpha}_j,\bm{\beta}_j\mid\bm{\mu},\bm{\Sigma},\zeta,\sigma\right) \\
+    &=\mathcal{LN}\left(\bm{\alpha}_j\mid\bm{\mu},\bm{\Sigma}\right)\times\prod_{\ell = 2}^d\mathcal{G}\left(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}\right).
+    \end{aligned}
 \end{equation}
@@
     \begin{aligned}
-    \pi(\bm{\alpha}_j\mid\bm{Y},\bm{r},\bm{\delta},\bm{\mu},\bm{\Sigma},\bm{\zeta},\bm{\sigma})
-    &\propto
-    \frac{1}{\prod_{\ell = 1}^d\alpha_{j\ell}}\exp\left\lbrace-\frac{1}{2}(\log\bm{\alpha}_j - \mu)^T\bm{\Sigma}^{-1}(\log\bm{\alpha}_j - \mu)\right\rbrace \\
-    &\hspace{-2cm}\times \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}}{\prod_{\ell = 1}^d\Gamma^{n_j}(\alpha_{j\ell})}\times \prod_{\ell = 2}^d\frac{\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
+    &\hspace{-0.5cm}\pi(\bm{\alpha}_j\mid\bm{Y},\bm{r},\bm{\delta},\bm{\mu},\bm{\Sigma},\bm{\zeta},\bm{\sigma})\\
+    &\propto \exp\left\lbrace-\frac{1}{2}(\log\bm{\alpha}_j - \mu)^T\bm{\Sigma}^{-1}(\log\bm{\alpha}_j - \mu)\right\rbrace \\
+    &\hspace{0.5cm}\times\frac{1}{\prod_{\ell = 1}^d\alpha_{j\ell}}\times \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}}{\prod_{\ell = 1}^d\Gamma^{n_j}(\alpha_{j\ell})}\\
+    &\hspace{0.5cm}\times \prod_{\ell = 2}^d\frac{\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
     \end{aligned}
@@
 
-%\subsubsection{Choice of Norm\label{subsubsec:norm}}
-%In the extreme value case, we have defined our task on ${\mathbb S}_{\infty}^{d-1}$, the positive orthant
-%  of the hypersphere defined under the $\mathcal{L}_{\infty}$ norm.  We previously described the
-%  difficulty in establishing a density directly on this geometry.  However, we can build the projected
-%  gamma distribution on ${\mathbb S}_p^{d-1}$ for a large $p$.  We can then project samples from the
-%  posterior predictive distribution for a model built on ${\mathbb S}_p^{d-1}$ onto
-%  ${\mathbb S}_{\infty}^{d-1}$. As $p$ increases, the closer the manifold will be to
-%  ${\mathbb S}_{\infty}^{d-1}$, and thus the closer the projected distribution will be to an idealized
-%  projected Gamma built on ${\mathbb S}_{\infty}^{d-1}$.  In testing, we tried developing models on
-%  $S_p^{d-1}$ for varying levels of $p$, and found that higher $p$ produced higher performing models.
-%  Additionally, the difference in model performance became more apparent as dimensionality increased.
-%  We present results here for models developed on ${\mathbb S}_{10}^{d-1}$.
-
 % EOF

### 2022-06-18 poster_sub.tex
statistics and computing submission

\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[margin=1in]{geometry}

\title{Bayesian non-parametric inference for multivariate peaks-over-threshold models}
\author{Peter Trubey \& Bruno Sans\'{o}}
\date{University of California, Santa Cruz \\ Santa Cruz, CA 95064; USA}

\begin{document}

\maketitle

\begin{abstract}
    \input{./abstract}
\end{abstract}

\paragraph{Keywords:}  Multivariate extremes, Peak over threshold models, Bayesian non-parametric models, Dirichlet process mixtures

\end{document}

% EOF

### 2022-06-18 results.tex
first pass at springer-nature class
statistics and computing submission

@@
-% \begin{figure}[tb]
-% \begin{center}
-%     \centering
-%   \caption{Posterior predictive loss (Left) and energy score (right) for various models on simulated data.  Gen indicates baseline.\label{fig:simpples}}
-%   \begin{subfigure}[b]{0.48\textwidth}
-%     \centering
-%     \includegraphics[height=2in, width=\textwidth]{./images/simulation_ppl}
-%   \end{subfigure}
-%   %
-%   \begin{subfigure}[b]{0.48\textwidth}
-%     \centering
-%     \includegraphics[height=2in, width=\textwidth]{./images/simulation_es}
-%   \end{subfigure}
-% \end{center}
-% \end{figure}
-
-\begin{figure}[htb]
+\begin{figure*}[htb]
     \centering
@@
     \includegraphics[height=3in, width = \textwidth]{./images/simulation_score}
-\end{figure}
+\end{figure*}
 
@@
     we consider simulated datasets on $\mathbb{S}_{\infty}^{d-1}$, for 
-$d$ ranging between 3 and 20. We generated each dataset from a mixture of projected Gammas, with the number of mixture components
-  ranging between 3 and 12.  The generation procedure is detailed in Algorithm~\ref{algo:simulated}.
-  \begin{algorithm}[ht]
-    \footnotesize
+    $d$ ranging between 3 and 20. We generated each dataset from a mixture of projected Gammas, with 
+    the number of mixture components ranging between 3 and 12.  The generation procedure is detailed 
+    in Algorithm~\ref{algo:simulated}.
+    \begin{algorithm}[ht]  % algorithm using algorithm package
         \caption{Simulated Angular Dataset Generation Routine\label{algo:simulated}}
-    \For{$n_{\text{mix}}$ in $[3, 6, 9, 12]$}{
-      Generate $n_{\text{mix}} \times 20$ shape Parameters $\alpha$\\
-      Generate $n_{\text{mix}} \times 20$ Rate Parameters $\beta$\\
-      Generate $500$ Mixture Component Identifiers $\delta$\\
-      \For{$i$ in $1,\ldots,500$}{
-        Generate ${\bf X}_i \sim \prod_{\ell = 1}^d\text{Ga}\left(X_{il}\mid\alpha_{\delta_i,l},\beta_{\delta_i, l}\right)$
-        }
-      \For{$n_{\text{col}}$ in $[3,6,12,20]$}{
-        Project columns 1 to $n_{\text{col}}$ of ${\bf X}$ onto 
-        $\mathcal{S}_{\infty}^{n_{\text{col}} - 1}$ and save.
-        }
-    }
+        \begin{algorithmic}
+        \For{$n_{\text{mix}}$ in $[3, 6, 9, 12]$}
+            \State Generate $n_{\text{mix}} \times 20$ shape Parameters $\alpha$
+            \State Generate $n_{\text{mix}} \times 20$ Rate Parameters $\beta$
+            \State Generate $500$ Mixture Component Identifiers $\delta$
+            \For{$i$ in $1,\ldots,500$}
+                \State Generate ${\bf X}_i \sim \prod_{\ell = 1}^d\text{Ga}\left(X_{il}\mid\alpha_{\delta_i,l},\beta_{\delta_i, l}\right)$
+            \EndFor
+            \For{$n_{\text{col}}$ in $[3,6,12,20]$}
+                \State Project columns 1 to $n_{\text{col}}$ of ${\bf X}$ onto $\mathcal{S}_{\infty}^{n_{\text{col}} - 1}$ and save.
+            \EndFor
+        \EndFor
+        \end{algorithmic}
     \end{algorithm}
+
 Each one of the shape parameters $\alpha$ were generated 
@@
 The scale parameters $\beta$ were obtained as $\beta\sim\text{Unif}(\beta\mid 0.25, 2.5)$.
-
-% \begin{figure}[ht]
-%   \caption{Simulation Energy Score versus Column Count, by Mixture Component Count. Lower is better.\label{fig:simes}}
-%   \centering
-%   \includegraphics[width=0.8\textwidth]{./images/simulation_es}
-% \end{figure}
 
@@
   The results from both scoring criteria are comparable.
-
-% We frequently see all of the gamma-based mixture models accomplish a lower energy score
-%     than the generative distribution.  As \cite{nunez2019} noted, the projected Gamma model
-%     is flexible and can generate multimodal distributions from a single mixture component.
-%     When we allow a mixture model, we allow the possibility to break those modes into 
-%     different mixture components.  This will tend to result in a better energy score, as 
-%     posterior predictive replicates for a given observation in a local model will tend to 
-%     be concentrated in that local mode rather than being spread between multiple modes.  
-%     As for the dominance of projected restricted Gamma, By specifying $\beta_{\ell} := 1$ 
-%     for all $l$, if there were multi-modal mixture components, we are in effect breaking 
-%     the mixture components into separate modes. As for whether this represents 
-%     \emph{overfitting}, If we had reason to believe that the generative distribution for
-%     real data followed this form, then such a case could be made.  However, we are using 
-%     projected Gamma as a parametric stand-in for an unknown distribution. In this sense, 
-%     the argument for overfitting is less clear.
-%     {\bf This paragraph is not very clear. Please make an effort to tell the story more clearly.}
 
@@
 \begin{algorithm}[htb]
-    \footnotesize
     \caption{Data preprocessing to isolate and transform data exhibiting extreme behavior.  $r_i$
@@
     portion is relevant for data correlated in time.\label{algo:processing}}
-  \KwResult{$\bm{ r},\bm{v} : r_i \sim \text{Pareto}(1)$, $\bm{ v}_i \in {\mathbb S}_{\infty}^{d-1}$}
-  \For{$\ell = 1,\ldots,d$}{
-    Set $b_{t,\ell} = \hat{F}_{\ell}^{-1}\left(1 - \frac{1}{t}\right)$.\\
-    With $\bm{ x}_{\ell} > b_{t,\ell}$, fit $a_{\ell}$, $\xi_{\ell}$ via MLE according to generalized Pareto likelihood.\\
-    }
-  \For{$i = 1,\ldots,n$}{
-    Define $z_{i,\ell} = \left(1 + \xi_{\ell}\frac{x_{i,\ell} - b_{t,\ell}}{a_{\ell}}\right)_{+}^{1/\xi_{\ell}}$; $\;\;\;$ then $r_i = \pnorm{\bm{ z}_i}{\infty}$, $\;\;\bm{ v}_i = \frac{\bm{ z}_i}{\pnorm{\bm{ z}_i}{\infty}}$\\
-    }
-  Subset $\bm{ r},\bm{ v}$ such that $r_i \geq 1$\\
-  \If{declustering}{
-    \For{$i = 1,\ldots,n$}{
-      If $r_i \geq 1$ and $r_{i-1} \geq 1$, drop the lesser (and associated $v_i$) from data set.\\
-    }
-  }
+    \begin{algorithmic}
+        \For{$\ell = 1,\ldots,d$}
+            \State Set $b_{t,\ell} = \hat{F}_{\ell}^{-1}\left(1 - \frac{1}{t}\right)$.
+            \State With $\bm{ x}_{\ell} > b_{t,\ell}$, fit $a_{\ell}$, $\xi_{\ell}$ via MLE according to generalized Pareto likelihood.
+        \EndFor
+        \For{$i = 1,\ldots,n$}
+            \State Define $z_{i,\ell} = \left(1 + \xi_{\ell}\frac{x_{i,\ell} - b_{t,\ell}}{a_{\ell}}\right)_{+}^{1/\xi_{\ell}}$; $\;\;\;$ then $r_i = \pnorm{\bm{ z}_i}{\infty}$, $\;\;\bm{ v}_i = \frac{\bm{ z}_i}{\pnorm{\bm{ z}_i}{\infty}}$
+        \EndFor
+        \State Subset $\bm{ r},\bm{ v}$ such that $r_i \geq 1$
+        \If{declustering}
+            \For{$i = 1,\ldots,n$}
+                \State If $r_i \geq 1$ and $r_{i-1} \geq 1$, drop the lesser (and associated $v_i$) from data set.
+            \EndFor
+        \EndIf
+    \end{algorithmic}
 \end{algorithm}
 
-\begin{figure}[ht]%{l}{0.45\textwidth}
+\begin{figure*}[ht]%{l}{0.45\textwidth}
     \centering
@@
 All data is between 0 and 1.\label{fig:erai_data}}
-    \includegraphics[width=.8\linewidth]{./images/data_transformed}
-\end{figure}
+    \includegraphics[width=.7\linewidth]{./images/data_transformed}
+\end{figure*}
 
@@
   both datasets.
-%  {\bf At this point it would be nice to include a figure with a plot of the data after all the pre-processing. You show a couple of panels for the ERA-Iterim and a couple of the ERA5.}
-%  \makenote{Do you mean marginal histograms, or, like, a couple selected ternary scatter-plots?}
 
@@
 
-\begin{figure}[ht]
+\begin{figure*}[htb]
     \centering
@@
     \includegraphics[width=0.9\textwidth]{./images/chi_ij_c}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}[htb]
+\begin{figure*}[htb]
     \centering
@@
     \includegraphics[width=\linewidth]{./images/condsurv_1d}
-\end{figure}
+\end{figure*}
 
-\begin{figure}[htb]
+\begin{figure*}[htb]
     \centering
     \caption{Pairwise conditional survival curves for selected locations, using ERA-Interim, and PRG-G model, conditioning on all other dimensions at greater than 90th percentile (fitted).\label{fig:condsurv2d}}
-    \includegraphics[height=5in, width=5in]{./images/condsurv_2d}
-\end{figure}
+    \includegraphics[height=4in, width=4in]{./images/condsurv_2d}
+\end{figure*}
 

### 2022-08-30 poster.tex
blah

\documentclass[25pt, a0paper, portrait]{tikzposter}
\title{\parbox{\linewidth}{\centering Bayesian Non-Parametric Inference
                        in Multivariate Peaks-over-Thresholds Models}}
\author{Peter Trubey}
\date{September 7, 2022}
\institute{UC Santa Cruz; Department of Statistics}

\usepackage{blindtext}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{comment}

\usetheme{Board}

\begin{document}

\maketitle

\block{~}
{
    \blindtext
}

\begin{columns}
    \column{0.5}
    \block{Multivariate Peaks-over-Thresholds}{
        Consider a $d$-dimensional random vector $\bm{W} = (W_1,\ldots,W_d)$ with cumulative distribution function $F$.  
        Assume the existence of sequences of vectors $\bm{a}_n,\bm{b}_n$ such that 
        \[\lim\limits_{n\to\infty}F^n\left(\bm{a}_n\bm{w} + \bm{b}_n\right) = G(\bm{w}),\] 
        then $G$ is a $d$-variate 
        generalized extreme value distribution.  It follows that
        \[
        \lim\limits_{n\to\infty}\text{Pr}\left(\frac{\bm{W} - \bm{b}_n}{\bm{a}_n} \leq \bm{w}\mid \bm{W} 
            \not\leq \bm{b}_n\right) = \frac{\log G(\bm{w}\wedge \bm{0}) - \log G(\bm{w})}{\log G(\bm{0})} = H(\bm{w}),
            \]
        where $H$ is the multivariate Pareto distribution.  That is, excesses \emph{over} the threshold 
        $b = \lim\limits_{n\to\infty} \bm{b}_n$ will be multivariate Pareto.  This comprises the cornerstone of 
        \emph{peaks over threshold} modelling.  In practice, we take a sufficiently large threshold $\bm{b}$, and model
        excesses over that threshold as generalized Pareto.

        For ease of interpretation, we transform $W$ to standardized marginals. Let 
        \[Z_{\ell} = \left(1 + \xi_{\ell}\frac{w_{\ell} - b_{\ell}}{a_{\ell}}\right))+^{1/\xi_{\ell}},\]
        then $\bm{Z}$ follows the standard multivariate Pareto distribution.  Remark 3.1 of \cite{rootzen2018} shows how
        we can factorize \[\bm{Z} = R\bm{V},\;\;R\in\mathbb{R}_+;\;V\in\mathbb{S}_{\infty}^{d-1},\] with $R$ independent
        of $\bm{V}$. That is, we can factorize $Z$ into \emph{independent} angular and radial components. $R$ will follow
        a standard Pareto distribution, and $\bm{V}$ will be

        will contain all information related to the dependence structure of $\bm{W}$ in extreme regions.  
        For that reason, we seek a flexible distribution to model $\bm{V}$.
        }

    \column{0.5}
    \block{Angular Space}{
        We recognize $\mathbb{S}_{\infty}^{d-1}$ as the positive orthant of the unit hypersphere, under the 
        $\mathcal{L}_{\infty}$ norm.  If we consider the $\mathcal{L}_p$ norm, and then the $\mathbb{L}_{\infty}$ norm is expressed as the limit of the $\mathcal{L}_p$ norm,
        \[
            \lVert \bm{s}\rVert_p = \left(\sum_{\ell = 1}^d s_{\ell}^p\right)^{1/p},\;\;\;
            \lVert \bm{s}\rVert_{\infty} = \lim\limits_{p\to\infty}\lVert \bm{s}\rVert_p = \bigvee_{\ell = 1}^d \bm{s}.\]
        If we define $\mathbb{S}_{\infty}^{d-1}$ using this norm, then we consider it the limit of of an expanding 
        series of manifolds $\mathbb{S}_p^{d-1}$, the positive orthant of the unit hypersphere under the $\mathcal{L}_p$ 
        norm.  That is, for $\bm{s}\in R_+^d$,
        \[ \mathbb{S}_{\infty}^{d-1} = \lim\limits_{p\to\infty}\left\lbrace \bm{x}\;\lVert \bm{x}\rVert_p = 1\right\rbrace.\]
        To establish a distribution on $\mathbb{S}_p^{d-1}$, we project a random vector from 
        $\mathbb{R}_+^d$ onto $\mathbb{S}_p^{d-1}$.  That is, for $\mathbb{x}\in\mathbb{R}_+^d$, let 
        $\bm{y} = \frac{\bm{x}}{\lVert\bm{x}\rVert_p} \in \mathbb{S}_p^{d-1}$.  Then, by letting 
        $y_d = \left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}$, the transformation
        \[T(x_1,\ldots,x_d) = \left(\lVert \bm{x}\rVert p, \frac{x_1}{\lVert \bm{x}\rVert_p}, 
                \ldots, \frac{x_{d-1}}{\lVert \bm{x}\rVert_p}\right)\]
        is invertible with
        \[T^{-1}(r,y_1,\ldots,y_{d-1}) = 
            \left(ry_1,\ldots,ry_{d-1},r\left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}\right).\]
        The determinant of the Jacobian of this transformation takes the form
        \[r^{d-1}\left[\left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}} +
            \sum_{\ell = 1}^{d-1}y_{\ell}^p\left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p} - 1}\right].\]
        Integrating out $r$ will yield a distribution on $\mathbb{S}_p^{d-1}$.
        With this transformation, we can project a density defined on $\mathbb{R}_+^{d-1}$ onto $\mathbb{S}_{p}^{d-1}$
        for any finite $p$.  Note that for $p=\infty$, the transformation is no longer differentiable.  Thus, our plan
        becomes to project $\bm{V}$ onto a large but finite $p$.
        }
    \block{Projected Gamma Family}
    {
    Given the form of the Jacobian, a natural distribution to consider in $\mathbb{R}_+^d$ is given by the product of 
    independent gamma distributions.  Let $\bm{X} \sim \prod_{\ell = 1}^d\text{Ga}(X_{\ell}\mid\alpha_{\ell},\beta_{\ell})$.  
    Using the transformation previously described, we have the joint density:
    \[
        f(r,\bm{y}\mid\bm{\alpha},\bm{\beta}) = \prod_{\ell = 1}^d \left[ \frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell}}
            (ry_{\ell})^{\alpha_{\ell} - 1}\exp\lbrace-\beta_{\ell}ry_{\ell}\rbrace\right]\times 
            r^{d-1}\left[\left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}} +
            \sum_{\ell = 1}^{d-1}y_{\ell}^p\left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p} - 1}\right].
    \]
    Integrating out $r$ yields the \emph{Projected Gamma} density,
    \[
    \text{PG}\left(\bm{y}\mid\bm{\alpha},\bm{\beta}\right) = \prod_{\ell = 1}^d
        \left[\frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell})}\right]
        \times \left[\left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}} +
            \sum_{\ell = 1}^{d-1}y_{\ell}^p\left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p} - 1}\right]
        \times 
        \frac{\Gamma\left(\sum_{\ell = 1}^d\alpha_{\ell}\right)}{\left(\sum_{\ell = 1}^d
                \beta_{\ell}y_{\ell}\right)^{\sum_{\ell = 1}^d\alpha_{\ell}}}
    \]
    }

\end{columns}
\block{Angular Distributions}{Here, \blindtext \vspace{4cm}}
    \note[
        targetoffsetx=-9cm, 
        targetoffsety=-6.5cm, 
        width=0.4\linewidth
        ]
        {e-mail \texttt{welcome@overleaf.com}}
\begin{columns}
    \column{0.5}
    \block{A figure}
    {
        \begin{tikzfigure}
            \includegraphics[width=0.4\textwidth]{images/ar}
        \end{tikzfigure}
    }
    \column{0.5}
    \block{Description of the figure}{\blindtext}
\end{columns}

\end{document}

### 2022-08-30 poster.tex
blah

@@
-\documentclass[25pt, a0paper, portrait]{tikzposter}
+\documentclass[20pt, a0paper, portrait]{tikzposter}
 \title{\parbox{\linewidth}{\centering Bayesian Non-Parametric Inference
@@
 
-\block{~}
-{
-    \blindtext
-}
-
 \begin{columns}
@@
         then $G$ is a $d$-variate 
-        generalized extreme value distribution.  It follows that
+        generalized extreme value distribution.  It follows that\\
         \[
@@
             \not\leq \bm{b}_n\right) = \frac{\log G(\bm{w}\wedge \bm{0}) - \log G(\bm{w})}{\log G(\bm{0})} = H(\bm{w}),
-            \]
+            \]\\
         where $H$ is the multivariate Pareto distribution.  That is, excesses \emph{over} the threshold 
@@
         For ease of interpretation, we transform $W$ to standardized marginals. Let 
-        \[Z_{\ell} = \left(1 + \xi_{\ell}\frac{w_{\ell} - b_{\ell}}{a_{\ell}}\right))+^{1/\xi_{\ell}},\]
+        \[Z_{\ell} = \left(1 + \xi_{\ell}\frac{w_{\ell} - b_{\ell}}{a_{\ell}}\right)_+^{1/\xi_{\ell}},\]
         then $\bm{Z}$ follows the standard multivariate Pareto distribution.  Remark 3.1 of \cite{rootzen2018} shows how
@@
         }
-
-    \column{0.5}
     \block{Angular Space}{
@@
         becomes to project $\bm{V}$ onto a large but finite $p$.
+        }
+    \column{0.5}
+    \block{}
+    {
+        \begin{tikzfigure}
+            \includegraphics[width=0.4\textwidth]{images/stdpareto2d}
+        \end{tikzfigure}
     }

### 2022-09-06 poster.tex
blah

--- 2022-08-30 poster.tex
@@
-\documentclass[25pt, a0paper, portrait]{tikzposter}
-\title{\parbox{\linewidth}{\centering Bayesian Non-Parametric Inference
-                        in Multivariate Peaks-over-Thresholds Models}}
-\author{Peter Trubey}
+\documentclass[20pt, a0paper, landscape,colspace=0.8cm,blockverticalspace=0.8cm,innermargin=0.8cm]{tikzposter}
+\title{\centering Bayesian Non-Parametric Inference
+                        in Multivariate Peaks-over-Thresholds Models}
+\author{Peter Trubey \& Bruno Sans{\'o}}
 \date{September 7, 2022}
@@
 \usepackage{comment}
+\usepackage{nccmath}
+\usepackage{booktabs}
+\usepackage{enumitem}
 
-\usetheme{Board}
+\usetheme{Desert}
 
@@
 
-\block{~}
-{
-    \blindtext
-}
-
 \begin{columns}
-    \column{0.5}
+    \column{0.333}
     \block{Multivariate Peaks-over-Thresholds}{
-        Consider a $d$-dimensional random vector $\bm{W} = (W_1,\ldots,W_d)$ with cumulative distribution function $F$.  
-        Assume the existence of sequences of vectors $\bm{a}_n,\bm{b}_n$ such that 
+        For $d$-dimensional random vector $\bm{W} = (W_1,\ldots,W_d) \sim F$;   
+        assume the existence of sequences of vectors $\bm{a}_n,\bm{b}_n$ such that 
         \[\lim\limits_{n\to\infty}F^n\left(\bm{a}_n\bm{w} + \bm{b}_n\right) = G(\bm{w}),\] 
-        then $G$ is a $d$-variate 
-        generalized extreme value distribution.  It follows that
+        then $G$ is a $d$-variate GEV.  It follows that\\
         \[
@@
             \not\leq \bm{b}_n\right) = \frac{\log G(\bm{w}\wedge \bm{0}) - \log G(\bm{w})}{\log G(\bm{0})} = H(\bm{w}),
-            \]
-        where $H$ is the multivariate Pareto distribution.  That is, excesses \emph{over} the threshold 
-        $b = \lim\limits_{n\to\infty} \bm{b}_n$ will be multivariate Pareto.  This comprises the cornerstone of 
-        \emph{peaks over threshold} modelling.  In practice, we take a sufficiently large threshold $\bm{b}$, and model
-        excesses over that threshold as generalized Pareto.
+            \]\\
+        where $H$ is multivariate Pareto.  In practice, for a sufficiently large
+        threshold $\bm{b}$, we model excesses as generalized Pareto.
 
         For ease of interpretation, we transform $W$ to standardized marginals. Let 
-        \[Z_{\ell} = \left(1 + \xi_{\ell}\frac{w_{\ell} - b_{\ell}}{a_{\ell}}\right))+^{1/\xi_{\ell}},\]
-        then $\bm{Z}$ follows the standard multivariate Pareto distribution.  Remark 3.1 of \cite{rootzen2018} shows how
-        we can factorize \[\bm{Z} = R\bm{V},\;\;R\in\mathbb{R}_+;\;V\in\mathbb{S}_{\infty}^{d-1},\] with $R$ independent
-        of $\bm{V}$. That is, we can factorize $Z$ into \emph{independent} angular and radial components. $R$ will follow
-        a standard Pareto distribution, and $\bm{V}$ will be
+        \[Z_{\ell} = \left(1 + \xi_{\ell}\frac{w_{\ell} - b_{\ell}}{a_{\ell}}\right)_+^{1/\xi_{\ell}},\]
+        then $\bm{Z}$ follows the standard multivariate Pareto distribution.
+        \vspace{1cm}
+        \begin{center}
+        \begin{tikzfigure}[2-dimensional generalized Pareto (left) transformed to standard marginals (right).]
+            \includegraphics[width=0.12\textwidth]{images/genpareto2d}
+            ~
+            \includegraphics[width=0.12\textwidth]{images/stdpareto2d}
+        \end{tikzfigure}
+        \end{center}
 
-        will contain all information related to the dependence structure of $\bm{W}$ in extreme regions.  
-        For that reason, we seek a flexible distribution to model $\bm{V}$.
+        \vspace{1cm}
+        Remark 3.1 of Rootzen et al, 2018 shows how % \cite{rootzen2018} shows how
+        we can factorize \[\bm{Z} =
+            R\bm{V},\;\;R\in\mathbb{R}_+;\;V\in\mathbb{S}_{\infty}^{d-1},\]
+        with $R$ independent of $\bm{V}$. 
+        That is, we can factorize $Z$ into
+        \emph{independent} angular and radial components. $R$ will follow
+        a standard Pareto distribution, and $\bm{V}$ will contain all 
+        information related to the dependence structure of $\bm{W}$ in extreme
+        regions.  For that reason, we seek a flexible distribution to model
+        $\bm{V}$.
+        }
+    \block{Scoring Criteria for distributions on $\mathbb{S}_{\infty}^{d-1}$}
+    {
+    To assess and compare the fidelity we focus on \emph{energy score}.  This has a deceptively simple calculation: the score at observation $i$ is calculated as
+    \[
+        S(P,\bm{x}_i) = \text{E}_p\left[g(\bm{X}_i,\bm{x}_i\right] -
+        \frac{1}{2}\text{E}_P\left[g(\bm{X}_i,\bm{X}_i^{\prime})\right],
+    \]
+    where $g$ is a negative definite kernel.  As the process we are trying to capture operates 
+    on $\mathbb{S}_{\infty}^{d-1}$, ideally $g$ would be the geodesic distance in
+    that space. Let \[\mathbb{C}_\ell^{d-1} = \lbrace
+    \bm{x} : \bm{x} \in \mathbb{S}_{\infty}^{d-1}, x_{\ell} = 1\rbrace\] comprise the $\ell$th
+    face of $\mathbb{S}_{\infty}^{d-1}$.  Geodesic distance in
+    $\mathbb{S}_{\infty}^{d-1}$ is difficult, but for
+    $\bm{a}\in\mathbb{C}_{\ell}^{d-1}$, $\bm{b}\in\mathbb{C}_{\jmath}^{d-1}$, 
+    an upper bound can be established as
+    \[
+        g(\bm{a},\bm{b}) = \min\limits_{\bm{c}\in\mathbb{C}_{\jmath}^{d-1}\cap\mathbb{C}_{\ell}^{d-1}}\lbrace\lVert \bm{c} - \bm{a}\rVert_2 + \lVert \bm{b} - \bm{c}\rVert_2\rbrace.
+    \]
+    Under this construction, $g$ is a negative definite kernel, as it is the sum of two 
+    negative definite kernels.  Calculating $g$ is made faster by rotating 
+    $\bm{b}\in\mathbb{C}_{\jmath}^{d-1}$ into the same hyperplane as $\bm{a}\in\mathbb{C}_{\ell}^{d-1}$.
+    % , 
+    % by setting \[\bm{b}^{\prime} = \begin{cases}b_i &\text{ for }i\neq \jmath,\ell\\ 
+    %     1 &\text{ for }i = \ell \\ 2 - b_{\ell} &\text{ for }i = j  \end{cases},\]
+    % then $g(\bm{a},\bm{b}) = \lVert \bm{a} - \bm{b}^{\prime}\rVert_2$.
     }
 
-    \column{0.5}
+    \column{0.333}
     \block{Angular Space}{
-        We recognize $\mathbb{S}_{\infty}^{d-1}$ as the positive orthant of the unit hypersphere, under the 
-        $\mathcal{L}_{\infty}$ norm.  If we consider the $\mathcal{L}_p$ norm, and then the $\mathbb{L}_{\infty}$ norm is expressed as the limit of the $\mathcal{L}_p$ norm,
+        \begin{center}
+        \begin{tikzfigure}[$\mathbb{S}_p^1$ for various $p$ (left); Projection of angular data onto various $\mathbb{S}_p^1$ (right).]
+            \includegraphics[width=0.12\textwidth]{images/p_sphere}
+            ~
+            \includegraphics[width=0.12\textwidth]{images/p_project}
+        \end{tikzfigure}
+        \end{center}
+
+        We recognize $\mathbb{S}_{\infty}^{d-1}$ as the positive orthant of the unit 
+        hypersphere, under the $\mathcal{L}_{\infty}$ norm.  If we consider the 
+        $\mathcal{L}_p$ norm, and then the $\mathbb{L}_{\infty}$ norm is expressed as the 
+        limit of the $\mathcal{L}_p$ norm,
         \[
-            \lVert \bm{s}\rVert_p = \left(\sum_{\ell = 1}^d s_{\ell}^p\right)^{1/p},\;\;\;
-            \lVert \bm{s}\rVert_{\infty} = \lim\limits_{p\to\infty}\lVert \bm{s}\rVert_p = \bigvee_{\ell = 1}^d \bm{s}.\]
-        If we define $\mathbb{S}_{\infty}^{d-1}$ using this norm, then we consider it the limit of of an expanding 
-        series of manifolds $\mathbb{S}_p^{d-1}$, the positive orthant of the unit hypersphere under the $\mathcal{L}_p$ 
-        norm.  That is, for $\bm{s}\in R_+^d$,
-        \[ \mathbb{S}_{\infty}^{d-1} = \lim\limits_{p\to\infty}\left\lbrace \bm{x}\;\lVert \bm{x}\rVert_p = 1\right\rbrace.\]
+            \lVert \bm{s}\rVert_p = \left({\small\sum}_{\ell = 1}^d
+                s_{\ell}^p\right)^{1/p},\;\;\;\;\;
+            \lVert \bm{s}\rVert_{\infty} = 
+            \lim\limits_{p\to\infty}\lVert \bm{s}\rVert_p = \bigvee_{\ell = 1}^d s_{\ell}.
+        \]
+        If we define $\mathbb{S}_{\infty}^{d-1}$ using this norm, then we consider it the 
+        limit of of an expanding series of manifolds $\mathbb{S}_p^{d-1}$, the positive 
+        orthant of the unit hypersphere under the $\mathcal{L}_p$ norm.  That is, for 
+        $\bm{s}\in R_+^d$,
+        \[ 
+            \mathbb{S}_{\infty}^{d-1} = 
+            \lim\limits_{p\to\infty}\left\lbrace \bm{x}\;\lVert \bm{x}\rVert_p =
+            1\right\rbrace.\
+        \]
         To establish a distribution on $\mathbb{S}_p^{d-1}$, we project a random vector from 
-        $\mathbb{R}_+^d$ onto $\mathbb{S}_p^{d-1}$.  That is, for $\mathbb{x}\in\mathbb{R}_+^d$, let 
+        $\mathbb{R}_+^d$ onto $\mathbb{S}_p^{d-1}$.  That is, for 
+        $\mathbb{x}\in\mathbb{R}_+^d$, let 
         $\bm{y} = \frac{\bm{x}}{\lVert\bm{x}\rVert_p} \in \mathbb{S}_p^{d-1}$.  Then, by letting 
-        $y_d = \left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}$, the transformation
+        $y_d = \left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}$, the
+        transformation
         \[T(x_1,\ldots,x_d) = \left(\lVert \bm{x}\rVert p, \frac{x_1}{\lVert \bm{x}\rVert_p}, 
@@
         \[T^{-1}(r,y_1,\ldots,y_{d-1}) = 
-            \left(ry_1,\ldots,ry_{d-1},r\left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}\right).\]
+            \left(ry_1,\ldots,ry_{d-1},r\left(1 - {\small\sum}_{\ell =
+                1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}\right).\]
         The determinant of the Jacobian of this transformation takes the form
-        \[r^{d-1}\left[\left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}} +
-            \sum_{\ell = 1}^{d-1}y_{\ell}^p\left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p} - 1}\right].\]
+        \[r^{d-1}\left[\left(1 - {\small\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}} +
+            {\small\sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(1 - {\small\sum}_{\ell =
+            1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p} - 1}\right].\]
         Integrating out $r$ will yield a distribution on $\mathbb{S}_p^{d-1}$.
-        With this transformation, we can project a density defined on $\mathbb{R}_+^{d-1}$ onto $\mathbb{S}_{p}^{d-1}$
-        for any finite $p$.  Note that for $p=\infty$, the transformation is no longer differentiable.  Thus, our plan
-        becomes to project $\bm{V}$ onto a large but finite $p$.
+        With this transformation, we can project a density defined on $\mathbb{R}_+^{d-1}$ 
+        onto $\mathbb{S}_{p}^{d-1}$ for any finite $p$.  Note that for $p=\infty$, the
+        transformation is no longer differentiable.  Thus, our plan becomes to project 
+        $\bm{V}$ onto a large but finite $p$, and assess model fidelity on $\mathbb{S}_{\infty}^{d-1}$
         }
-    \block{Projected Gamma Family}
+
+    \block{Future Work in extreme analysis}
+    {
+    \begin{itemize}[noitemsep,nolistsep]
+        \item Novelty detection in extremes using the projected gamma representation:
+            \begin{itemize}[noitemsep,nolistsep]
+                \item Inference in $\mathbb{S}_{\infty}^{d-1}$ introduces new possibilities of applications.
+We hope to create an unsupervised--learning method of spotting anomalies---data that is not \emph{normal}.
+            \end{itemize}
+        \item A variational approach to Dirichlet process mixtures of projected gammas:
+        \begin{itemize}[noitemsep,nolistsep]
+            \item The log-normal centering distribution, and MCMC techniques for sampling DP's in general induce a great deal of computational complexity.
+Our goal is combating this problem through a variational approach.
+        \end{itemize}
+    \end{itemize}
+    }
+\column{0.333}
+\block{Projected Gamma Family \& DP mixtures thereof}
     {
@@
     \[
+        \medmath{
         f(r,\bm{y}\mid\bm{\alpha},\bm{\beta}) = \prod_{\ell = 1}^d \left[ \frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell}}
@@
             \sum_{\ell = 1}^{d-1}y_{\ell}^p\left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p} - 1}\right].
+            }
     \]
@@
     \[
+        \medmath{
     \text{PG}\left(\bm{y}\mid\bm{\alpha},\bm{\beta}\right) = \prod_{\ell = 1}^d
@@
         \frac{\Gamma\left(\sum_{\ell = 1}^d\alpha_{\ell}\right)}{\left(\sum_{\ell = 1}^d
-                \beta_{\ell}y_{\ell}\right)^{\sum_{\ell = 1}^d\alpha_{\ell}}}
+                \beta_{\ell}y_{\ell}\right)^{\sum_{\ell = 1}^d\alpha_{\ell}}}.
+                }
     \]
+    For identifiability, we fix $\beta_1 := 1$. This creates a moderately flexible distribution on
+    $\mathbb{S}_p^{d-1}$ for any finite $p$. The projected gamma family is simple to specify and
+    has very tractable computational properties. Thus, we use it as a building block to model the 
+    angular distribution.  To build a flexible family of distributions on $\mathbb{S}_p^{d-1}$, we build a Dirichlet process mixture of the form
+    \[
+    \bm{y}_i\mid (\bm{\alpha}_i,\bm{\beta}_i) \sim \text{PG}(y_i\mid\bm{\alpha}_i,\bm{\beta}_i), \;\;\;\;
+    (\alpha_i,\beta_i) \mid G \sim G, \;\;\;\; G \sim \text{DP}(\eta,G_0).
+    \]
+    We consider 4 options:
+    \begin{itemize}
+        \item \emph{Projected Gamma} (PG) vs \emph{Projected Restricted Gamma} (PRG) kernel distribution
+        \item \emph{Product of Gammas} (--G) vs \emph{Multivariate Lognormal} (--LN) centering distribution
+    \end{itemize}
+    }
+    \block{Results}{
+    We conducted a simulation study to evaluate the performance of PG vs PRG, and ~--G vs ~--LN.
+
+    \begin{tikzfigure}[Energy Score for recovering distribution of simulated data at a variety of columns, number of mixtures in generating distribution.]
+        \centering 
+        \includegraphics[width=0.25\textwidth,height=0.09\textwidth]{images/simulation_score}
+    \end{tikzfigure}
+
+    We interpret energy score formulated in this way as "lower is better".
+    This indicates model fidelity was negatively affected by the
+    additional flexibility offered in the unrestricted model (PG vs PRG).
+    Interesting to note here is that the Log-normal model offers slightly worse
+    performance as compared to the product of gammas.
+
+    We also tried the models against two datasets for the \emph{integrated vapor transport} (IVT).
+This data captures the total amount of water vapor in the air in an atmospheric column, over a cell grid along California's coast.
+ERA--Interim has 9 columns; and ERA5 has 43 columns.
+
+    \begin{tikzfigure}[Energy Score for fitted models against IVT data]
+        \centering
+        \begin{tabular}{ccc}
+        \toprule
+        Model   & ERA--Interim  & ERA--5\\
+        \midrule
+        PRG--G  & {\bf 0.172}   & 0.751       \\
+        PRG--LN & 0.173         & {\bf 0.714} \\
+        PG--G   & 0.426         & 1.169       \\
+        PG--LN  & 0.294         & 0.799       \\
+        \bottomrule
+        \end{tabular}
+    \end{tikzfigure}
+
+    We see a slight reversal here, as compared to the simulation, where the log-normal models are preferred for the high-dimensional data.
     }
@@
 \end{columns}
-\block{Angular Distributions}{Here, \blindtext \vspace{4cm}}
-    \note[
-        targetoffsetx=-9cm, 
-        targetoffsety=-6.5cm, 
-        width=0.4\linewidth
-        ]
-        {e-mail \texttt{welcome@overleaf.com}}
-\begin{columns}
-    \column{0.5}
-    \block{A figure}
-    {
-        \begin{tikzfigure}
-            \includegraphics[width=0.4\textwidth]{images/ar}
-        \end{tikzfigure}
-    }
-    \column{0.5}
-    \block{Description of the figure}{\blindtext}
-\end{columns}
+
+% \block{Angular Distributions}{Here, \blindtext \vspace{4cm}}
+%     \note[
+%         targetoffsetx=-9cm, 
+%         targetoffsety=-6.5cm, 
+%         width=0.4\linewidth
+%         ]
+%         {e-mail \texttt{welcome@overleaf.com}}
+% \begin{columns}
+%     \column{0.5}
+%     \block{A figure}
+%     {
+%         \begin{tikzfigure}
+%             \includegraphics[width=0.4\textwidth]{images/ar}
+%         \end{tikzfigure}
+%     }
+%     \column{0.5}
+%     \block{Description of the figure}{\blindtext}
+% \end{columns}
 

### 2022-09-06 poster2.tex
blah

--- 2022-09-06 poster.tex
@@
-\documentclass[20pt, a0paper, landscape,colspace=0.8cm,blockverticalspace=0.8cm,innermargin=0.8cm]{tikzposter}
-\title{\centering Bayesian Non-Parametric Inference
-                        in Multivariate Peaks-over-Thresholds Models}
+\documentclass[20pt, a0paper, landscape, colspace=9mm, subcolspace=4mm, blockverticalspace=9mm, innermargin=8mm, margin=0mm]{tikzposter}
+\title{\parbox{\linewidth}{\centering Bayesian Non-Parametric Inference
+                        in Multivariate\\ Peaks-over-Thresholds Models}}
 \author{Peter Trubey \& Bruno Sans{\'o}}
@@
 
+\usepackage[english]{babel}
 \usepackage{blindtext}
@@
 \usepackage{enumitem}
+\usepackage{graphicx}
+\usepackage{wrapfig}
+\usepackage{adjustbox}
 
-\usetheme{Desert}
+\usetheme{Simple}
 
@@
         then $\bm{Z}$ follows the standard multivariate Pareto distribution.
-        \vspace{1cm}
-        \begin{center}
-        \begin{tikzfigure}[2-dimensional generalized Pareto (left) transformed to standard marginals (right).]
-            \includegraphics[width=0.12\textwidth]{images/genpareto2d}
-            ~
-            \includegraphics[width=0.12\textwidth]{images/stdpareto2d}
-        \end{tikzfigure}
-        \end{center}
-
-        \vspace{1cm}
         Remark 3.1 of Rootzen et al, 2018 shows how % \cite{rootzen2018} shows how
@@
         }
-    \block{Scoring Criteria for distributions on $\mathbb{S}_{\infty}^{d-1}$}
+    \block{Angular Space}{
+        \begin{center}
+        \begin{tikzfigure}[$\mathbb{S}_p^1$ for various $p$ (left); Projection of angular data onto various $\mathbb{S}_p^1$ (right).]
+            \includegraphics[width=0.10\textwidth]{images/p_sphere}
+            ~
+            \hspace{1cm}
+            ~
+            \includegraphics[width=0.10\textwidth]{images/p_project}
+        \end{tikzfigure}
+        \end{center}
+        We recognize $\mathbb{S}_{\infty}^{d-1}$ as the positive orthant of the unit 
+        hypersphere, under the $\mathcal{L}_{\infty}$ norm.  If we consider the 
+        $\mathcal{L}_p$ norm, and then the $\mathbb{L}_{\infty}$ norm is expressed as the 
+        limit of the $\mathcal{L}_p$ norm,
+        \[
+            \lVert \bm{s}\rVert_p = \left({\small\sum}_{\ell = 1}^d
+                s_{\ell}^p\right)^{1/p},\;\;\;\;\;
+            \lVert \bm{s}\rVert_{\infty} = 
+            \lim\limits_{p\to\infty}\lVert \bm{s}\rVert_p = \bigvee_{\ell = 1}^d s_{\ell}.
+        \]
+        If we define $\mathbb{S}_{\infty}^{d-1}$ using this norm, then we consider it
+        the limit of of an expanding series of manifolds $\mathbb{S}_p^{d-1}$, the
+        positive orthant of the unit hypersphere under the $\mathcal{L}_p$ norm.  
+        To establish a distribution on $\mathbb{S}_p^{d-1}$, we project a random 
+        vector from $\mathbb{R}_+^d$ onto $\mathbb{S}_p^{d-1}$.  For 
+        $\mathbb{x}\in\mathbb{R}_+^d$, let 
+        $\bm{y} = \frac{\bm{x}}{\lVert\bm{x}\rVert_p} \in \mathbb{S}_p^{d-1}$.  Then, by letting 
+        $y_d = \left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}$, the
+        transformation
+        \[T(x_1,\ldots,x_d) = \bigg(\lVert \bm{x}\rVert p, \frac{x_1}{\lVert \bm{x}\rVert_p}, 
+                \ldots, \frac{x_{d-1}}{\lVert \bm{x}\rVert_p}\bigg)\]
+        is invertible with
+        \[T^{-1}(r,y_1,\ldots,y_{d-1}) = 
+            \big(ry_1,\ldots,ry_{d-1},ry_d\big).\]
+        The determinant of the Jacobian of this transformation takes the form
+        \[r^{d-1}\left[y_d +
+            {\tiny\sum}_{\ell = 1}^{d-1}y_{\ell}^py_d^{1 - p}\right].\]
+        Integrating out $r$ will yield a distribution on $\mathbb{S}_p^{d-1}$.
+        Note that for $p=\infty$, the transformation is no longer differentiable.  Thus, we 
+        project $\bm{V}$ onto a large but finite $p$, and assess model fidelity on 
+        $\mathbb{S}_{\infty}^{d-1}$
+        }
+
+\column{0.333}
+
+\block{Projected Gamma Family \& DP Mixtures Thereof}
+    {
+    Given the form of the Jacobian, a natural distribution to consider in $\mathbb{R}_+^d$ is given by the product of 
+    independent gamma distributions.  Let $\bm{X} \sim \prod_{\ell = 1}^d\text{Ga}(X_{\ell}\mid\alpha_{\ell},\beta_{\ell})$.  
+    Using the transformation previously described, and treating $y_d = \left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}$, we have the joint density:
+    \[
+        f(r,\bm{y}\mid\bm{\alpha},\bm{\beta}) = \prod_{\ell = 1}^d \left[ \frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell}}
+            (ry_{\ell})^{\alpha_{\ell} - 1}\exp\lbrace-\beta_{\ell}ry_{\ell}\rbrace\right]\times 
+            r^{d-1}\left[y_d +
+            \sum_{\ell = 1}^{d-1}y_{\ell}^py_d^{1-p}\right].
+    \]
+    Integrating out $r$ yields the \emph{Projected Gamma} density,
+    \[
+    \text{PG}\left(\bm{y}\mid\bm{\alpha},\bm{\beta}\right) = \prod_{\ell = 1}^d
+        \left[\frac{\beta_{\ell}^{\alpha_{\ell}}y_{\ell}^{\alpha_{\ell} - 1}}{\Gamma(\alpha_{\ell})}\right]
+        \times \left[y_d +
+            \sum_{\ell = 1}^{d-1}y_{\ell}^py_d^{1-p}\right]
+        \times 
+        \frac{\Gamma\left(\sum_{\ell = 1}^d\alpha_{\ell}\right)}{\left(\sum_{\ell = 1}^d
+                \beta_{\ell}y_{\ell}\right)^{\sum_{\ell = 1}^d\alpha_{\ell}}}.
+    \]
+    For identifiability, we fix $\beta_1 := 1$. This creates a moderately flexible distribution on
+    $\mathbb{S}_p^{d-1}$ for any finite $p$. The projected gamma family is simple to specify and
+    has very tractable computational properties. Thus, we use it as a building block to model the 
+    angular distribution.  To build a flexible family of distributions on $\mathbb{S}_p^{d-1}$, we build a Dirichlet process mixture of the form
+    \[
+    \bm{y}_i\mid (\bm{\alpha}_i,\bm{\beta}_i) \sim \text{PG}(y_i\mid\bm{\alpha}_i,\bm{\beta}_i), \;\;\;\;
+    (\bm{\alpha}_i,\bm{\beta}_i) \mid G \sim G, \;\;\;\; G \sim \text{DP}(\eta,G_0).
+    \]
+    We consider 4 options:
+    \begin{itemize}
+        \item \emph{Projected Gamma} (PG) vs \emph{Projected Restricted Gamma} (PRG) kernel distribution
+        \item \emph{Product of Gammas} (--G) vs \emph{Multivariate Lognormal} (--LN) centering distribution
+    \end{itemize}
+    }
+
+\block{Scoring Criteria for Distributions on $\mathbb{S}_{\infty}^{d-1}$}
     {
@@
 
-    \column{0.333}
-    \block{Angular Space}{
-        \begin{center}
-        \begin{tikzfigure}[$\mathbb{S}_p^1$ for various $p$ (left); Projection of angular data onto various $\mathbb{S}_p^1$ (right).]
-            \includegraphics[width=0.12\textwidth]{images/p_sphere}
-            ~
-            \includegraphics[width=0.12\textwidth]{images/p_project}
-        \end{tikzfigure}
-        \end{center}
-
-        We recognize $\mathbb{S}_{\infty}^{d-1}$ as the positive orthant of the unit 
-        hypersphere, under the $\mathcal{L}_{\infty}$ norm.  If we consider the 
-        $\mathcal{L}_p$ norm, and then the $\mathbb{L}_{\infty}$ norm is expressed as the 
-        limit of the $\mathcal{L}_p$ norm,
-        \[
-            \lVert \bm{s}\rVert_p = \left({\small\sum}_{\ell = 1}^d
-                s_{\ell}^p\right)^{1/p},\;\;\;\;\;
-            \lVert \bm{s}\rVert_{\infty} = 
-            \lim\limits_{p\to\infty}\lVert \bm{s}\rVert_p = \bigvee_{\ell = 1}^d s_{\ell}.
-        \]
-        If we define $\mathbb{S}_{\infty}^{d-1}$ using this norm, then we consider it the 
-        limit of of an expanding series of manifolds $\mathbb{S}_p^{d-1}$, the positive 
-        orthant of the unit hypersphere under the $\mathcal{L}_p$ norm.  That is, for 
-        $\bm{s}\in R_+^d$,
-        \[ 
-            \mathbb{S}_{\infty}^{d-1} = 
-            \lim\limits_{p\to\infty}\left\lbrace \bm{x}\;\lVert \bm{x}\rVert_p =
-            1\right\rbrace.\
-        \]
-        To establish a distribution on $\mathbb{S}_p^{d-1}$, we project a random vector from 
-        $\mathbb{R}_+^d$ onto $\mathbb{S}_p^{d-1}$.  That is, for 
-        $\mathbb{x}\in\mathbb{R}_+^d$, let 
-        $\bm{y} = \frac{\bm{x}}{\lVert\bm{x}\rVert_p} \in \mathbb{S}_p^{d-1}$.  Then, by letting 
-        $y_d = \left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}$, the
-        transformation
-        \[T(x_1,\ldots,x_d) = \left(\lVert \bm{x}\rVert p, \frac{x_1}{\lVert \bm{x}\rVert_p}, 
-                \ldots, \frac{x_{d-1}}{\lVert \bm{x}\rVert_p}\right)\]
-        is invertible with
-        \[T^{-1}(r,y_1,\ldots,y_{d-1}) = 
-            \left(ry_1,\ldots,ry_{d-1},r\left(1 - {\small\sum}_{\ell =
-                1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}\right).\]
-        The determinant of the Jacobian of this transformation takes the form
-        \[r^{d-1}\left[\left(1 - {\small\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}} +
-            {\small\sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(1 - {\small\sum}_{\ell =
-            1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p} - 1}\right].\]
-        Integrating out $r$ will yield a distribution on $\mathbb{S}_p^{d-1}$.
-        With this transformation, we can project a density defined on $\mathbb{R}_+^{d-1}$ 
-        onto $\mathbb{S}_{p}^{d-1}$ for any finite $p$.  Note that for $p=\infty$, the
-        transformation is no longer differentiable.  Thus, our plan becomes to project 
-        $\bm{V}$ onto a large but finite $p$, and assess model fidelity on $\mathbb{S}_{\infty}^{d-1}$
-        }
-
-    \block{Future Work in extreme analysis}
+\block{Future Work in Extreme Analysis}
     {
-    \begin{itemize}[noitemsep,nolistsep]
+    \begin{itemize}[noitemsep]
         \item Novelty detection in extremes using the projected gamma representation:
-            \begin{itemize}[noitemsep,nolistsep]
+            \begin{itemize}[noitemsep]
                 \item Inference in $\mathbb{S}_{\infty}^{d-1}$ introduces new possibilities of applications.
@@
         \item A variational approach to Dirichlet process mixtures of projected gammas:
-        \begin{itemize}[noitemsep,nolistsep]
+        \begin{itemize}[noitemsep]
             \item The log-normal centering distribution, and MCMC techniques for sampling DP's in general induce a great deal of computational complexity.
@@
 \column{0.333}
-\block{Projected Gamma Family \& DP mixtures thereof}
-    {
-    Given the form of the Jacobian, a natural distribution to consider in $\mathbb{R}_+^d$ is given by the product of 
-    independent gamma distributions.  Let $\bm{X} \sim \prod_{\ell = 1}^d\text{Ga}(X_{\ell}\mid\alpha_{\ell},\beta_{\ell})$.  
-    Using the transformation previously described, we have the joint density:
-    \[
-        \medmath{
-        f(r,\bm{y}\mid\bm{\alpha},\bm{\beta}) = \prod_{\ell = 1}^d \left[ \frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell}}
-            (ry_{\ell})^{\alpha_{\ell} - 1}\exp\lbrace-\beta_{\ell}ry_{\ell}\rbrace\right]\times 
-            r^{d-1}\left[\left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}} +
-            \sum_{\ell = 1}^{d-1}y_{\ell}^p\left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p} - 1}\right].
-            }
-    \]
-    Integrating out $r$ yields the \emph{Projected Gamma} density,
-    \[
-        \medmath{
-    \text{PG}\left(\bm{y}\mid\bm{\alpha},\bm{\beta}\right) = \prod_{\ell = 1}^d
-        \left[\frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell})}\right]
-        \times \left[\left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}} +
-            \sum_{\ell = 1}^{d-1}y_{\ell}^p\left(1 - \sum_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p} - 1}\right]
-        \times 
-        \frac{\Gamma\left(\sum_{\ell = 1}^d\alpha_{\ell}\right)}{\left(\sum_{\ell = 1}^d
-                \beta_{\ell}y_{\ell}\right)^{\sum_{\ell = 1}^d\alpha_{\ell}}}.
-                }
-    \]
-    For identifiability, we fix $\beta_1 := 1$. This creates a moderately flexible distribution on
-    $\mathbb{S}_p^{d-1}$ for any finite $p$. The projected gamma family is simple to specify and
-    has very tractable computational properties. Thus, we use it as a building block to model the 
-    angular distribution.  To build a flexible family of distributions on $\mathbb{S}_p^{d-1}$, we build a Dirichlet process mixture of the form
-    \[
-    \bm{y}_i\mid (\bm{\alpha}_i,\bm{\beta}_i) \sim \text{PG}(y_i\mid\bm{\alpha}_i,\bm{\beta}_i), \;\;\;\;
-    (\alpha_i,\beta_i) \mid G \sim G, \;\;\;\; G \sim \text{DP}(\eta,G_0).
-    \]
-    We consider 4 options:
-    \begin{itemize}
-        \item \emph{Projected Gamma} (PG) vs \emph{Projected Restricted Gamma} (PRG) kernel distribution
-        \item \emph{Product of Gammas} (--G) vs \emph{Multivariate Lognormal} (--LN) centering distribution
-    \end{itemize}
-    }
+
     \block{Results}{
@@
 
-    \begin{tikzfigure}[Energy Score for recovering distribution of simulated data at a variety of columns, number of mixtures in generating distribution.]
+    \begin{tikzfigure}[Energy Score for recovering distribution of simulated data for a variety of dimensions, number of mixture components in generating distribution (lower is better).]
         \centering 
@@
 
-    We interpret energy score formulated in this way as "lower is better".
-    This indicates model fidelity was negatively affected by the
-    additional flexibility offered in the unrestricted model (PG vs PRG).
-    Interesting to note here is that the Log-normal model offers slightly worse
-    performance as compared to the product of gammas.
+    We see model fidelity was negatively affected by the additional 
+    flexibility offered in the unrestricted model (PG vs PRG).
+    Interesting to note here is that the Log-normal model offers slightly 
+    worse performance as compared to the product of gammas.
 
-    We also tried the models against two datasets for the \emph{integrated vapor transport} (IVT).
-This data captures the total amount of water vapor in the air in an atmospheric column, over a cell grid along California's coast.
-ERA--Interim has 9 columns; and ERA5 has 43 columns.
+    \vspace{1cm}
+    ~
+    % Text and figure Inside the Block
+    \begin{minipage}[c]{0.6\linewidth}
 
-    \begin{tikzfigure}[Energy Score for fitted models against IVT data]
+    We applied our models to two datasets for the \emph{integrated vapor transport} (IVT).
+    This data captures the total amount of water vapor in the air in an atmospheric 
+    column, over a cell grid along California's coast.  ERA--Interim has 9 columns; and 
+    ERA5 has 43 columns.
+
+    We see a slight reversal here, as compared to the simulation, where the log-normal 
+    centering distribution is preferred for the high-dimensional data.  Again, the
+    restricted model performs better than the unrestricted model.
+    \end{minipage}%
+    \begin{adjustbox}{valign=c}
+    \begin{minipage}[t]{0.4\linewidth}
+    \begin{tikzfigure}[Energy Score for fitted models against\\ IVT data (lower is better).]
         \centering
@@
     \end{tikzfigure}
+    \end{minipage}
+    \end{adjustbox}
+    }
+\begin{subcolumns}
+\subcolumn{0.5}
+\block{Conditional Survival}{
+    For a set of dimensions $\Omega\subset\lbrace1,\ldots,d\rbrace$, the conditional probability of survival for set $\Omega$ can be calculated as
+    \[
+    \medmath{
+    \mathrm{Pr}\bigg[\cap_{\ell \in \Omega} Z_\ell > z_\ell \mid \cap_{\ell\not\in\Omega} Z_\ell > z_\ell\bigg]
+    = \frac{\text{E}\left[\wedge_{k = 1}^d 1\wedge \frac{V_k}{z_k}\right]}{\text{E}\left[\wedge_{k \not\in\Omega}1\wedge\frac{V_k}{z_k}\right]}.
+    }
+    \]
+    We see two such sets $\Omega$ here, holding all other dimensions at their respective fitted $90$th percentile.
+    \begin{tikzfigure}[Conditional survival curves for 2 $\Omega$--sets given all others $\geq F^{-1}(0.9)$ (fitted), from ERA-Interim data with PRG--G model.
+Left is original scale, right is standardized margins.]
+    % \begin{wrapfigure}{L}{0.15\textwidth}
+    \centering
+    \includegraphics[width=0.12\textwidth]{./images/condsurv_2d}
+    %     \caption{Conditional survival curves for 2 pairs of dimensions given all others $\geq F^{-1}(0.9)$ (fitted), from ERA-Interim data with PRG--G model.}
+    % \end{wrapfigure}
+    \end{tikzfigure}
+}
+\subcolumn{0.5}
+\block{Extremal Dependence}{
+    A measure to characterize the strength of dependence between two variables is the
+    extremal dependence coefficient.  We establish the second equality
+    as, by standardization, the marginal distribution of $Z_{\jmath}$ is the same as that of $Z_{\ell}$.
+    \[
+        \medmath{
+        \begin{aligned}
+        \chi_{\jmath\ell} &= \lim\limits_{u\uparrow 1}\text{Pr}\left[
+        F_{\jmath}(Z_{\jmath}) > u \mid F_{\ell}(Z_{\ell}) > u\right]\\
+        &= \text{E}\left[
+        \frac{V_{\jmath}}{\text{E}[V_{\jmath}]}\bigwedge
+        \frac{V_{\ell}}{\text{E}[V_{\ell}]}
+        \right].
+        \end{aligned}
+        }
+    \]
+    Using a sample from the posterior predictive of a fitted PRG-G model on the IVT data, we compute this expectation numerically.
 
-    We see a slight reversal here, as compared to the simulation, where the log-normal models are preferred for the high-dimensional data.
+    \begin{tikzfigure}[Pairwise extremal dependence coefficients for IVT data using PRG-G model.]
+    \centering
+    \includegraphics[width=0.95\linewidth]{./images/chi_ij_c}
+    \end{tikzfigure}
+    Effectively, using the model in this way, we are capturing the spatial dependence structure for extremes in the IVT.
     }
+\end{subcolumns}
 

### 2022-12-18 poster2.tex
blah

@@
 
-    \begin{tikzfigure}[Energy Score for recovering distribution of simulated data for a variety of dimensions, number of mixture components in generating distribution (lower is better).]
+    \begin{tikzfigure}[Energy Score vs number of dimensions for recovering distribution of simulated data,\\for varying number of mixture components in generating distribution (lower is better).]
         \centering 
-        \includegraphics[width=0.25\textwidth,height=0.09\textwidth]{images/simulation_score}
+        \includegraphics[width=0.25\textwidth,height=0.08\textwidth]{images/simulation_score_es}
     \end{tikzfigure}
@@
 
-    \begin{tikzfigure}[Pairwise extremal dependence coefficients for IVT data using PRG-G model.]
+    \begin{tikzfigure}[Pairwise extremal dependence coefficients\\ for IVT data using PRG-G model.]
     \centering

### 2023-05-08 pointbypoint.tex
before some cutting

%\documentclass[sn-basic]{sn-jnl}
\documentclass[10pt]{article}
\usepackage{natbib}
\usepackage[margin=1in]{geometry}
\bibliographystyle{plainnat}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{soul}
\usepackage{bm}
\usepackage{subcaption}
\usepackage{wrapfig}

\newcommand{\comment}[1]{\textcolor{teal}{#1}}
\newcommand{\bruno}[1]{\textcolor{red}{#1}}

\begin{document}

\section*{Review 1}
\begin{itemize}
    \item \emph{Moment constraint}: As can be seen, for example, in \cite[Equation 4.1]{einmahl2009}
    there is a moment constraint on the angular measure for any $\mathcal{L}_p$ norm with 
    $p \in [1, \infty]$ (nb including $p = \infty$).  Hence, why is there no moment constraint in 
    the submitted 
    manuscript? Some papers on inference for the angular measure have not taken into consideration 
    that constraint (e.g. \cite{einmahl2001}), though most modern approaches do account for it. If 
    the moment constraint is being ignored here, I would be willing to overlook this provided that 
    this is openly acknowledged---as well as what are the consequences of this. Otherwise, this needs 
    to be further clarified.

    \comment{Both reviewers comment on the moment constraint.  I forget if it's this one, or
    the reference provided by the other reviewer, but I don't think they understood the reference
    directly; they were using the $\mathcal{L}_1$ norm in $p$ dimensions, rather than the 
    $\mathcal{L}_p$ norm as described in our paper.  A massive difference in meaning of notation.
    I'll find the reference again and confirm.}

    We thank the referee for calling our attention to
the issue of moment restrictions. The approach considered in the
paper follows that of \cite{rootzen2018}, consisting on modeling the
distribution of the extreme observations, conditional on them being
above a threshold, say $H$.  The approach considered in
\cite{EiSe2009} focuses in the limiting joint exponent measure. More
precisely, for the two dimensional case, they assume that
\[
	sPr(s^{-1}(Z_1,Z_2) \in A) \rightarrow \mu(A),
\]
where $\mu$ is the exponent measure. The moment conditions are
the result of assuming that both marginals of $\mu$ correspond to a
standard Pareto distribution. Such assumption is not compatible
with the conditional distribution approach considered in our paper.
In fact, as noted in \cite{KiRoSeWa2019}, typically the margins of
$H$ are not univariate Pareto, due to the difference between the 
conditioning events in the joint and the marginal cases. Thus, no
moment conditions arise in our model. This has now been clarified in the paper.

We note in passing that the moment conditions for the spectral
distribution, say $\Phi$, in the two-dimensional case, using the 
infinity norm, imply that 
\[
	\int_0^{\pi/4} \Phi(d\theta) = \int_{\pi/4}^{\pi/2}
\Phi(d\theta).
\]
The former looks like a fairly strong symmetry restriction that is
unlike to be realistic in practice. Thus, imposing moment
conditions would probably result in a very restrictive model.

    \item \emph{Main contribution}: While I appreciate the interesting construction based on the 
    DPM of projected gammas (which itself could be of use for other contexts beyond extremes?), 
    one wonders about what would be the main added value of it with respect to other existing 
    Bayesian approaches (\cite{boldi2007},\cite{guillotte2011},\cite{sabourin2014},\cite{hanson2017}).
    Keeping in mind the computational scope of Statistics and Computing, one wonders if the main 
    advantage is, for example, a computational one? This needs to be clarified—especially in light 
    of the previous comment on the moment constraint. Indeed, one could use any of the previous 
    approaches with pseudo-angles based on the infinity norm.

    \comment{Hanson is clearly different.  They're using Bernstein polynomials effectively defined 
    on $\mathbb{S}_1^{d-1}$.  We can demonstrate that projected gammas built on $\mathbb{S}_{p}$, 
    $p \geq 10$ and $\mathbb{S}_1$ create \emph{very} different gamma shape parameters, for data
    created by the \emph{same} distribution.  I need to check again the other references, but 
    based on their follow-on comment, I'm guessing the previous point needs to be addressed before 
    we can properly justify this point.}

    \bruno{Based on our conversation, you need to look carefully into each one of these references.
    it seems that most of them do not develop a PoT approach. But also, we use a kernel whose natural 
    support is the positive cone (without using transformations) to build a mixture WRT to a random
    measure. This provides the flexibility of mixture for a methods that has computational advantages 
    and scales to moderately large problems, as illustrated in the paper.}

    \comment{\cite{hanson2017}: Bernstein polynomials on $\mathbb{S}_1^{d-1}$, modelling multivariate Frechets.  \cite{boldi2007}: Marginal Frechets, also operating on $\mathbb{S}_1^{d-1}$.  \cite{guillotte2011}: A strictly bivariate model.  \cite{sabourin2014}: }

    \item \emph{Numerical accuracy}:  Keeping in mind the computational scope of Statistics and 
    Computing one would hope for more numerical work than what is presented in §5.1. As far as I
    understand what is presented are single run experiments for each $d$? Monte Carlo evidence on the
    performance of the procedure would be appreciated. On top of this, the current numerical exercise
    ignores the uncertainty that stems from step 1 of the procedure (i.e. estimating the margins).

    \comment{Regarding the uncertainty from step 1, I think we agreed that was basically intractible.
    We may need to explicitly specify as such, and that a practicioner must accept the numerical 
    uncertainty that arises from step 1.  That said, I seem to recall the numbers observed being 
    relatively robust to choice of threshold (for large threshold), the difference mainly coming 
    down to computation time.}

    \bruno{I think we need to conduct a more comprehensive suite of experiments. Regarding the uncertainty 
    coming from Step1, we just need to acknowledge that we are ignoring it.}

    To address this concern, we have \comment{revised (note: future past tense)} the simulation study 
    to feature the average rise in energy score over baseline, averaged over 10 iterations.  The simulated
    datasets are generated as mixtures of multivariate lognormals, projected onto the unit sphere.

    \item \emph{Prior and MCMC}: A lot is left to the reader in §5 in terms of prior specification
    and MCMC settings.  At least some further details would be required.

    \comment{The form of the prior is specified.  I'm guessing he's referring to the actual values
    used for computation?  $\bm{\mu}_0 = \bm{0}$, $\nu = d + 50$, $\bm{\Psi} = \nu I_d$, and so on?}

    \bruno{You need to provide more details here.}

    \item \emph{Future research}: §6 calls for a view of the authors on what could be related 
    exercises—or even extensions of this research and future work. Using the proposed Bayesian 
    multivariate PoT construction for devising a regression model for an an extreme value response 
    on an extreme value covariate is one example of a natural follow-up; see \cite{carvalho2022}
    for related ideas. Conditional versions of the angular measure (e.g. \cite{carvalho2016},
    \cite{castro2018}, \cite{escobar2018}, \cite{mhalla2019}) would lead to a conditional 
    multivariate peaks-over-threshold model, where dependence between exceedances could change 
    along with a covariate. Also, perhaps the here proposed DPM of projected gammas—as well as 
    predictor-dependent versions of it—could be of interest in other settings beyond angular 
    measures?

    \comment{I'm guessing this guy is de Carvalho?}

    \bruno{I have no doubts that this referee is Carvalho. I'll spend some time reading the papers
    that he mentions.}

\end{itemize}

\section*{Review 2}
\begin{itemize}
    \item \emph{Lack of clarity}: The presentation of the probabilistic setting for the multivariate 
    PoT model (Section 2) and estimation method (Section 3) contains some inconsistencies and unclear
    points:
    \begin{enumerate}
        \item Please use a different notation for the random vector distributed according to $F$ 
        (denoted by $W$ in the paper) and for the Peaks-over-threshold limit, that is the standard
        Generalized Pareto vector, also denoted by $W$. This is not correct because the latter has 
        not the same distribution as the former. The confusion goes on throughout the paper, e.g. 
        the standardized variables in Section 3 (after a probability integral transform, Equation
        (3), are denoted by the letter $z$. Notice that the vector $W$ defined as a limit in Section 
        2 does not necessarily follow a standard Generalized Pareto distribution. The relation 
        (and the difference) between standard and non-standard is explained e.g. in [3], Proposition 4. 
        In order to have the representation $W = W_{\infty}V$ as in the paper, with 
        $W_{\infty} \sim \text{Pa}(1)$, one needs to assume that $W$ is indeed standard GP. In other 
        words it is correct to write $Z = RV$ where $Z$ is a properly standardized vector, not $W = RV$.

        \comment{Yeah.  W is a multivariate GPD vector; Z is a multivariate standardized Pareto vector.
        This needs to be fixed.}

        We thank the referee for pointing out the inconsistencies in the notation. We have corrected them 
        to make things clearer. We denote all random variables with capital letters and their values
        with small letters. The random variable corresponding to the conditional limiting distribution
        $H$ is now denoted as $Z$.

        \item About the standardization, Eq. (3): please introduce some background. In the paper it is
        introduced as \emph{’the standardization’} and it is unclear what is the relation between this 
        and the definition of the GPD (beginning of Section 2), unless one already knows about marginal
        standardization methods in multivariate EVT. In any case there are several ways to perform 
        marginal standardization in multivariate EVT.

        We have rewritten the material at the beginning of Section 3, explicitly mentioning the marginal
        univariate generalized Pareto, and how it is used to estimate the shape and scale parameters, 
        which are then used to standardize the observations.

        \item Section 3.2.1: Please provide some minimal background (or references) on the Dirichlet 
        Process Mixture, in particular about clusters of observations occurring naturally in this 
        model. This is indeed standard in Bayesian Non-parametrics, so there should be a precise 
        reference where it is explained.

        \bruno{Peter, you need to come up with additional, and more specific references here.}

    \end{enumerate}

    \item \emph{Organization of paper}:
        \begin{enumerate}
            \item End of Section 2, starting from \emph{A measure that is used to characterize 
            the strength of dependence...}, until the end: the discussion about the dependence 
            coefficient (Coles’s $\chi$) is quite disconnected from the rest at this point. It
            is not useful for Section 3. It is only used in the experiments (Section 5). If I 
            understand well, in Section 5, pairwise $\chi$’s are computed by Monte-Carlo simulation 
            in the whole model, which exemplifies the usefulness of the projected gamma model. 
            However these $\chi$’s may be computed in many different ways, the most simple being 
            the empirical version. It is unclear at this point what is the added value of the 
            model compared with an empirical method.

            \item The statistical model (projected Gamma family) is only introduced in the
            \emph{‘estimation’} section, whereas it is the main building block of the paper.

            We thank the referee for the suggestion to change the layout of the paper. We have
            re-written Section 3 to put a stronger emphasis on the presentation of the
            projected gamma kernel.

        \end{enumerate}

    \item \emph{Unclear Goals}: 
    \begin{enumerate}
        \item There is no proper justification in the paper for the focus on the infinity norm 
        and the associated sphere $\mathbb{S}_{\infty}^{d-1}$. In \cite{ferreira2014}, the focus
        is on the space of bounded processes, and this infinite dimensional sample space is 
        naturally associated with the infinity norm. However in the multivariate case here, the 
        setting is very different and there is no obvious reason why the max norm should be used. 
        Also, by changing the conditioning event, i.e. replacing $\lbrace W \not\leq b_{n}\rbrace$
        (notations of the paper, beginning of Section 2) with 
        $\lbrace \lVert W\rVert_p > t_n\rbrace$, one will obtain a product representation in the 
        limit of the form $\tilde{W} = RV_p$ where $R$ is Pareto and $V_p$ is located on the 
        $\mathcal{L}_p$-sphere. See e.g. one of Resnick’s books, where it is clearly stated that 
        the product form of the limit does not depend on the choice of the norm. I believe that 
        the projected Gamma family on the $L_p$ sphere with a DPM prior is anyways a nice 
        contribution, which is somewhat hidden by this additional projection on the infinity sphere.

        \bruno{We need to get hold of a copy of the right book to see what the result is. If the result 
        is so straightforward, then we need to decided what to do about it, but I am worried that
        it will require a moment constraint.}

        \item Another issue is that it is not clear in the present form of the paper, what is the
        added value of fitting this parametric model with the Bayesian framework and the Gibbs 
        sampler, if the output is mainly bivariate densities and extremal dependence coefficients 
        which could be obtained by standard non parametric methods with very low computational cost?

        \bruno{This is an unfair comment. We do use the full multivariate distributiont to come up
        with the results we present. I'll work on an answer to this one.}
    \end{enumerate}

    \item \emph{Comparison with other methods}: The only comparison study displayed in the paper 
    takes place within the suggested model, only with different prior choice. This is not sufficient 
    for performance comparison in a methodological paper. One may e.g. think of comparing with the
    results obtained with the empirical angular measure or smoothed versions of it in terms of 
    extremal dependence and conditional survival probability.

    \bruno{Well, here we have to bite the bullet. Hopefully there some software that allows us
    to fit the empirical angular measure.}

    \item \emph{Experiments}:
        \begin{enumerate}
            \item Computational Evidence missing: The convergence of the Gibbs sampler is not 
            discussed in the paper.

            \bruno{This one was pointed out by the other referee. So, we need to come up with something.}

            \comment{\st{I'm not sure how to demonstrate convergence in the case of the DP.  I can
            I can demonstrate convergence for certain quantities (e.g., the hierarchical mean,
            and log-determinant of hierarchical covariance matrix, a given observation's distribution
            of parameters, etc.), but in the general stick-breaking configuration, there is no concern
            given to label switching.  The variational approach will probably find a label arrangement
            and stick with it.}}

            \comment{I'm working on adding log-density over time plots to demonstrate convergence.}

            \item Figure 2: The scores from the ‘baseline’ should outperform the scores of
            the fitted models, because the baseline consists of samples generated from the 
            true distribution. This is not the case in Figure 2, which deserves an explanation 
            (it might be a too small sample size, or some coding error at worst).

            \bruno{We need to think about this one.}

            \comment{I think it was an issue of understanding.  I've updated the code, remaking the plots
            is nearly done.}

        \end{enumerate}

    \item \emph{About the absence (or presence) of moments constraint}: The angular measure in
    multivariate extremes with respect to any norm has a moment constraint which precise form depends 
    on the chosen norm and on the standardization, see e.g. 
    \cite{beirlant2006} Chapter 8, Section 8.2.3, Equation 8.20. 
    This seems to contradict the author’s claim that there is no moments constraints in the models, 
    due to the choice of the $\mathcal{L}_{\infty}$-norm. Although it may be the case that their choice
    of standardization removes the moments constraint, I do not understand how the choice the 
    $\mathcal{L}_{\infty}$-norm helps here.

    Another way to see this: for $x \geq 1$, using the representation $W = W_{\infty}V$ as in the
    paper, with $W_{\infty}$ a standard Pareto variable, we get 
    \[\mathbb{P}\left[W_{\ell} > x\right] = \mathbb{E}\left[V_{\ell}\right] / x\]
    (see also Equation (2.20) in \cite{ferreira2014}, with $\omega_0 = 1$).  As a consequence if 
    the distribution of the marginal variable $W_{\ell}$ in the limit is imposed, then so is 
    $\mathbb{E}\left[V_{\ell}\right]$, namely 
    \[ \mathbb{E}\left[V_{\ell}\right] = \mathbb{P}\left[W_{\ell} > 1\right] \]
    It may be the case that, with the standardization Equation (3) in their paper, the marginal
    distributions, i.e. the distributions of the $W_{\ell}$’s in the limit are not entirely determined, 
    but this needs to be clarified.

    \comment{I seem to regard something in the intro to extremes book regarding the lack of a
    moment constraint on the $\mathcal{L}_{\infty}$ norm.  I'll look for it.}

    \bruno{This was dealt with already.}

    \item \emph{Energy Score, kernel function, flaw in the proof}: In the proof of proposition 3, 
    one needs to verify that $g$ is a negative definite kernel, that is, that for all $x_1,\ldots,x_n$
    and $a_1,\ldots,a_n$, \[ \sum_i\sum_ja_ia_jg(x_i,x_j)\leq 0,\]
    (with varying points $x_i,x_j$), not that \[\sum_i\sum_ja_ia_jg(x_1,x_2)\leq 0.\]  Thus the vector $e$
    which allows to break the sum into two parts in the current proof should in fact depend on $(i,j)$, and the proof is not valid.

    \bruno{We need to rewrite the proof. I'll do that.}

    \comment{\st{Yeah, I think I messed this one up here.  I'm going to run simulations to verify it is
    a negative definite kernel, then try and figure out a better proof.  Too bad the trivial one
    didn't work.}}

    \comment{\st{I'm somewhat worried that they will regard the new proof as also problematic, as the second 
    term is dependent on the first term.  We could get around this by splitting $\sum_{i,j}a_ia_jg(x_i,x_j)$ 
    into sets belonging to particular face pairs (e.g., $x_i$ on face $a$, $x_j$ on face $b$) and sum over 
    permutations of $a,b$, but then there's no guarantee that a particular $a_i$ set within a particular 
    $a,b$ face set will sum to 0.}}

    \item \emph{Minor point 1}: About the word \emph{‘Projection’}, e.g. \emph{in 
    ‘To define an angular measure, we are interested in the direction of vectors in $\mathbb{R}_+^d$
    thus, we project them onto $[\ldots]$}: The angular variable variable is not really a projection 
    (at least not an orthogonal one) except when the $\mathcal{L}_2$ norm is used.

    \comment{Indeed, it's not an orthogonal projection.  Even the $\mathcal{L}_2$ norm doesn't 
    produce an orthogonal projection, unless using the trigonometric mapping.  I don't think that's
    a problem.}

    \bruno{I think we need to de-emphasize the word ``projection''. We will call it ``normalization''
    the first time we mention it, and then say that we will use the term ``projection''. It is a
    projection in the sense that it is the point in $\mathbb{S}_p^{d-1}$ that is closest to the original
    point, is it not?}

    \item \emph{Minor point 2}: The first sentence of Section 3 is vague: [\emph{‘Consider a 
    collection of observations $w_l, l = 1,\ldots,n$ that exhibit extreme behavior}].  Probably the
    authors mean that the underlying random vector W is regularly varying, or belongs to some domain 
    of attraction. However only specialists of EVT can guess that.

    \comment{we're using $l$ for iterating observations, rather than dimensions?  Yeah, it's probably
    bad wording.}

    \item \emph{Minor point 3}: (typos): Section 3, first paragraph: Some confusions between $\ell \in \lbrace 1,\ldots, n\rbrace$ and $\ell \in \lbrace 1,\ldots,d\rbrace$

    \comment{same as previous.}
\end{itemize}

\bibliography{./refs}

\end{document}

% EOF

### 2023-05-08 evaluation.tex
statistics and computing submission
before some cutting

@@
 
-In order to assess and compare the estimation of a distribution on ${\mathbb S}_\infty^{d-1}$ we consider the theory of proper scoring rules developed in \cite{gneiting2007}.
-As mentioned in Section~\ref{subsec:projgamma}, our approach does not provide a density on ${\mathbb S}_{\infty}^{d-1}$, restricting our ability to construct model selection criteria to sample-based approaches.
-We consider two scoring methods:  \emph{posterior predictive loss} (PPL) introduced in \cite{gelfand1998}, and \emph{energy score} (ES) introduced in \cite{gneiting2007}.
+In order to assess and compare the estimation of a distribution on ${\mathbb S}_\infty^{d-1}$ we consider 
+    the theory of proper scoring rules developed in \cite{gneiting2007}. As mentioned in 
+    Section~\ref{subsec:projgamma}, our approach does not provide a density on 
+    ${\mathbb S}_{\infty}^{d-1}$, restricting our ability to construct model selection criteria 
+    to sample-based approaches.  To this end, we employ the \emph{energy score} criterion introduced
+    therein.
+    % the We consider two scoring methods:  \emph{posterior predictive loss} (PPL) introduced in 
+    % \cite{gelfand1998}, and \emph{energy score} (ES) introduced in \cite{gneiting2007}.
 
-\subsection{Posterior predictive loss}
-The posterior predictive loss criterion is introduced in \cite{gelfand1998}.  When we
-  assume a squared error loss function, then for the $i$th observation, the posterior predictive
-  loss criterion is computed as
-  \begin{equation}
-    \label{eq:ppl}
-      S_{\text{PPL}}^k\left(P,x_i\right) = \text{Var}_P\left[X_i\right] + \frac{k}{k+1}\left(\text{E}_P\left[X_{i}\right] - x_i\right)^2
-  \end{equation}
-  where $X_i$ is a random variable from the posterior predictive distribution for $x_i$.  The
-  scalar $k$ is a weighting factor that scales the importance of goodness of fit
-  relative to precision.  In fact a small $\text{Var}(X_{i})$ indicates high
-  precision, and a small $(\text{E}[X_{i}] - x_{i})^2$  indicates a good fit.  In our analysis,
-  we take the limit as $k\to\infty$, weighting both
-  equally. Overall, the smaller the score, the better.  Note that the PPLC score is defined 
-  for a univariate distribution.  As we are dealing
-  with a multivariate distribution, we will be taking the posterior predictive loss
-  summed over all dimensions.  That is,
-  \begin{equation}
-    \label{eq:ppl2}
-      S_{\text{PPL}}\left(P,\bm{x}_i\right) = \frac{1}{d}\sum_{\ell = 1}^d\left[\text{Var}_P\left[X_{i\ell}\right] + \left(\text{E}_P\left[X_{i\ell}\right] - x_{i\ell}\right)^2\right] .
-  \end{equation}
-  We report the average $S_{\text{PPL}}$ taken from all observed data.
+% \subsection{Posterior predictive loss}
+% The posterior predictive loss criterion is introduced in \cite{gelfand1998}.  When we
+%   assume a squared error loss function, then for the $i$th observation, the posterior predictive
+%   loss criterion is computed as
+%   \begin{equation}
+%     \label{eq:ppl}
+%     \begin{aligned}
+%       &\hspace{-0.5cm}S_{\text{PPL}}^k\left(P,x_i\right) \\
+%       &= \text{Var}_P\left[X_i\right] + \frac{k}{k+1}\left(\text{E}_P\left[X_{i}\right] - x_i\right)^2
+%     \end{aligned}
+%   \end{equation}
+%   where $X_i$ is a random variable from the posterior predictive distribution for $x_i$.  The
+%   scalar $k$ is a weighting factor that scales the importance of goodness of fit
+%   relative to precision.  In fact a small $\text{Var}(X_{i})$ indicates high
+%   precision, and a small $(\text{E}[X_{i}] - x_{i})^2$  indicates a good fit.  In our analysis,
+%   we take the limit as $k\to\infty$, weighting both
+%   equally. Overall, the smaller the score, the better.  Note that the PPLC score is defined 
+%   for a univariate distribution.  As we are dealing
+%   with a multivariate distribution, we will be taking the posterior predictive loss
+%   summed over all dimensions.  That is,
+%   \begin{equation}
+%     \label{eq:ppl2}
+%     \begin{aligned}
+%       &\hspace{-0.5cm}S_{\text{PPL}}\left(P,\bm{x}_i\right) \\
+%       &= \frac{1}{d}\sum_{\ell = 1}^d\left[\text{Var}_P\left[X_{i\ell}\right] + \left(\text{E}_P\left[X_{i\ell}\right] - x_{i\ell}\right)^2\right].
+%     \end{aligned}
+%   \end{equation}
+%   We report the average $S_{\text{PPL}}$ taken from all observed data.
 
-\subsection{Energy score}
-We focus on \emph{energy scores} defined for a general probability distribution $P$, with finite expectation, as 
+% \subsection{Energy score}
+% We focus on \emph{energy scores} defined for a general probability distribution $P$, with finite expectation, as 
+The energy score criterion defined for a general probability distribution $P$, with finite expectation, is developed
+    as
     \begin{equation}
     \label{eq:es}
-    \text{S}_{\text{ES}}\left(P, \bm{x}_i\right) =  \text{E}_p \left[g\left(\bm{X}_i, \bm{x}_i\right)\right] -
+    \begin{aligned}
+    &\hspace{-0.5cm}\text{S}_{\text{ES}}\left(P, \bm{x}_i\right) \\
+    &=\text{E}_p \left[g\left(\bm{X}_i, \bm{x}_i\right)\right] -
                                     \frac{1}{2}\text{E}_p \left[g\left(\bm{X}_i,\bm{X}_i^{\prime}\right)\right],
+    \end{aligned}
     \end{equation}
@@
 
-In a Euclidean space, these conditions are satisfied by the Euclidean distance. However, in
-  ${\mathbb S}_{\infty}^{d-1}$, Euclidean distance will tend to under-estimate the actual 
-  distance required to travel between two points.  Let
+In a Euclidean space, these conditions are satisfied by the Euclidean distance \citep{berg1984}. 
+    However, for observations on different faces of ${\mathbb S}_{\infty}^{d-1}$, Euclidean 
+    distance will under-estimate the geodesic distance, the actual distance required to travel 
+    between the two points.  Let
 \begin{equation*}
@@
   shortest path between two points on a geometric figure corresponds to the length of a
-  straight line drawn between the points on an appropriate unfolding, rotation, or \emph{net} of the figure from a $d$-dimensional to a $d-1$-dimensional space.
-The optimal net will have the shortest straight line between the points, as long as that line is fully contained within such net.
-As ${\mathbb S}_{\infty}^{d-1}$ has $d$ faces---each face pairwise adjacent, there are $d!$ possible nets.
-However, we are only interested in nets that begin and end on the source and destination faces respectively.
-That reduces the number of nets under consideration to $\sum_{k = 0}^{d-2}\binom{d-2}{k}$.
-This is still computationally burdensome for a large number of dimensions.
+  straight line drawn between the points on an appropriate unfolding, rotation, or 
+  \emph{net} of the figure from a $d$-dimensional to a $d-1$-dimensional space.  The optimal 
+  net will have the shortest straight line between the points, as long as that line is fully 
+  contained within such net. As ${\mathbb S}_{\infty}^{d-1}$ has $d$ faces---each face 
+  pairwise adjacent, there are $d!$ possible nets.  However, we are only interested in nets 
+  that begin and end on the source and destination faces respectively, reducing the number 
+  of nets under consideration to $\sum_{k = 0}^{d-2}\binom{d-2}{k}$.  This is still 
+  computationally burdensome for a large number of dimensions.  
+  \makenote{However, we can efficiently establish an upper bound on the geodesic 
+  length.  We use this upper bound on geodesic distance as the kernel function for
+  the energy score.}
 
-To reduce the computations needed to calculate the energy score we define a kernel as 
-in the following proposition.
-\begin{prop}\label{prop:g}
-Let $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-1}$, for $\ell, \jmath \in \{1, \ldots , d\}$. Define $g$ as
+To calculate the energy score we define the kernel 
     \[  
@@
     \]
-Then $g$  is a negative definite kernel.
-\end{prop}
-
-{\em Proof:}
-When $\bm{a}$ and $\bm{b}$ are on the same face, then $\ell = \jmath$, and $g$ is the Euclidean distance between $\bm{a}$ and $\bm{b}$, which is a  negative definite kernel.
+    where $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and 
+    $\bm{b} \in {\mathbb   C}_{\jmath}^{d-1}$, for $\ell,\jmath\in \{1,\ldots, d\}$.
+%\end{prop}
+%{\em Proof:}
+%When $\bm{a}$ and $\bm{b}$ are on the same face, then $\ell = \jmath$, and $g$ is the Euclidean distance between $\bm{a}$ and $\bm{b}$, which is a  negative definite kernel.
 When $\bm{a}$ and $\bm{b}$ are on separate
-  faces, let 
-  \[\bm{e} = 
-    \argmin_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
-        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
-  \]
-  We have that $g(\bm{a}, \bm{b}) = g(\bm{b}, \bm{a})$ due to the uniqueness of $\bm{e}$ and the symmetry of the Euclidean distance.
-Additionally, given $n$ and  a collection $\alpha_1,\ldots,\alpha_n\in{\mathbb R}$ such that  $\sum_{i = 1}^n\alpha_i = 0$, then,
-  \begin{equation*}
-    \begin{aligned}
-      \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j g(\bm{x}_1,\bm{x}_2) &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j \bigg[\pnorm{\bm{e} - \bm{x}_1}{2} + \pnorm{\bm{x}_2 - \bm{e}}{2}\bigg]\\
-      &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{e} - \bm{x}_1}{2} + \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{x}_2 - \bm{e}}{2} \leq0,
-    \end{aligned}
-  \end{equation*}
-  as the Euclidean norm is a negative definite kernel.$\hfill\Box$
-
-Evaluating $g$ as defined in Proposition \ref{prop:g} requires the solution of a $(d-2)$-dimensional 
+  %faces, let 
+  %\[\bm{e} = 
+    %\argmin_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
+        %\pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
+  %\]
+  %We have that $g(\bm{a}, \bm{b}) = g(\bm{b}, \bm{a})$ due to the uniqueness of $\bm{e}$ and the symmetry of the Euclidean distance.
+Additionally, given $n$ and  a collection $\alpha_1,\ldots,\alpha_n\in{\mathbb R}$ such that  $\sum_{i = 1}^n\alpha_i = 0$, then
+  %\begin{equation*}
+    %\begin{aligned}
+      %&\hspace{-0.5cm}\sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j g(\bm{x}_1,\bm{x}_2)\\
+      %&= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j \bigg[\pnorm{\bm{e} - \bm{x}_1}{2} + \pnorm{\bm{x}_2 - \bm{e}}{2}\bigg]\\
+      %&= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{e} - \bm{x}_1}{2}\\
+      %&\hspace{1cm}+ \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{x}_2 - \bm{e}}{2}\\
+      %&\leq 0
+    %\end{aligned}
+  %\end{equation*}
+  %as the Euclidean norm is a negative definite kernel.$\hfill\Box$
+Evaluating $g$ as described requires the solution of a $(d-2)$-dimensional 
   optimization problem.  The following proposition provides a straightforward solution.
 \begin{prop}\label{prop:rot}
-    Let $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-1}$, for $\ell, \jmath \in \{1, \ldots , d\}$.  For $\ell\neq \jmath$, the 
-  transformation $P_{\jmath\ell}(\cdot)$ required to rotate the $\jmath$th face along the $\ell$th axis 
-  produces the vector $\bm{b}^\prime$, where
+    Let $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-1}$, 
+        for $\ell, \jmath \in \{1, \ldots , d\}$.  For $\ell\neq \jmath$, the 
+  transformation $P_{\jmath\ell}(\cdot)$ required to rotate the $\jmath$th face along the 
+  $\ell$th axis produces the vector $\bm{b}^\prime$, where
   \begin{equation}
@@
   $\bm{c} \in {\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}$,
-     $\pnorm{\bm{b} - \bm{c}}{2} =  \pnorm{\bm{b}^{\prime} - \bm{c}}{2}$. Then, recalling the definition of 
-     $\bm{e}$ in the proof of Proposition \ref{prop:g}, we have that
+     $\pnorm{\bm{b} - \bm{c}}{2} =  \pnorm{\bm{b}^{\prime} - \bm{c}}{2}$ and let
+\[\bm{e} = 
+    \argmin_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
+        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
+\]
+We then have that
   \begin{equation*}
@@
     g(\bm{a}, \bm{b}) &= \pnorm{\bm{a} - \bm{e}}{2} + \pnorm{\bm{b} - \bm{e}}{2}\\
-    &= \pnorm{\bm{a} - \bm{e}}{2} + \pnorm{\bm{b}^{\prime} - \bm{e}}{2} = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}.
+    &= \pnorm{\bm{a} - \bm{e}}{2} + \pnorm{\bm{b}^{\prime} 
+        - \bm{e}}{2} = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}.
     \end{aligned}
   \end{equation*}
-  The last equality is due to the fact that $\bm{a}$ and $\bm{b}'$ belong to the same hyperplane.
-  $\hfill\Box$
+  The last equality is due to the fact that $\bm{a}$ and $\bm{b}'$ belong to the 
+  same hyperplane. $\hfill\Box$
 
-Proposition \ref{prop:rot} provides a computational efficient way to evaluate the proper scoring rule $S_{\text{ES}}$ defined on ${\mathbb S}_\infty^{d-1}$, for each observation.
-For the purpose of model assessment and comparison, we report the average $S_{\text{ES}}$ taken across all observed data.
-In this form, the interpretation of $S_{\text{ES}}$ is similar to that of $S_{\text{PPL}}$: The smaller the score, the better.
+Using the rotation in Proposition \ref{prop:rot} we obtain the following result.
+    \begin{prop}\label{prop:g}
+    g is a negative definite kernel.
+    \end{prop}
+    {\em Proof:} For a given $n$ consider an arbitrary set of points
+    $\bm{a}_1, \ldots , \bm{a}_n \in {\mathbb S}_\infty^{d-1}$, and 
+    $\alpha_1, \ldots , \alpha_n \in {\mathbb R}$, such that $\sum_{i=1}^n
+    \alpha _i = 0$. Then
+    \[
+    \sum_{i,\jmath} \alpha_i \alpha_\jmath g(\bm{a}_i, \bm{a}_\jmath) 
+        = \sum_{i,\jmath} \alpha_i \alpha_\jmath 
+    \|\bm{a}_i - \bm{a}_\jmath^\prime\|_2 \leq 0,
+    \]
+    where $\bm{a}_\jmath^\prime$ is defined as in Proposition \ref{prop:rot}. $\hfill\Box$
+
+Proposition \ref{prop:rot} provides a computational efficient way to evaluate the proper
+    scoring rule $S_{\text{ES}}$ defined on ${\mathbb S}_\infty^{d-1}$, for each 
+    observation. For the purpose of model assessment and comparison, we report the 
+    average $S_{\text{ES}}$ taken across all observed data.  In this form, the 
+    interpretation of $S_{\text{ES}}$ is similar to that of $S_{\text{PPL}}$: The smaller 
+    the score, the better.
+
 % EOF

### 2023-05-08 introduction.tex
before some cutting

@@
 \end{equation*}
-$H$ is the multivariate Pareto distribution. \cite{RoSeWa2018a}
-provide a number of stochastic representations for $H$. In particular
-Remark 1 justifies the representation given in \cite{ferreira2014},
-consisting of taking $\bm{W}$, in the limit, as $\bm{W} = \bm{V}R$ 
-where $R$ and $\bm{V}$ are independent. $R = ||\bm{W}||_\infty$ is 
+$H$ a multivariate Pareto distribution, and corresponds to the joint
+distribution conditional on exceeding a multivariate threshold. In addition to
+a copula function, $H$ depends on two $d$-dimensional vectors of 
+parameters that we denote as $\bm{\xi}$ for the shapes and $\bm{\sigma}$ 
+for the scales. \cite{RoSeWa2018a} provide a number of stochastic 
+representations for $H$.  In particular, denote as $Z$ a
+random variable with distribution $H$ where $\bm{\xi}= 1$ and 
+$\bm{\sigma} = 0$.  Then, Remark 1 justifies the representation in 
+\cite{ferreira2014}, giving $\bm{Z} = \bm{V}R$
+where $R$ and $\bm{V}$ are independent. $R = \|\bm{Z}\|_\infty$ is 
 distributed as a standard Pareto random variable, and $\bm{V} = 
-\bm{W}/||\bm{W}||_\infty$ is a random vector in  
+\bm{Z}/\|\bm{Z}\|_\infty$ is a random vector in  
 $\mathbb{S}_{\infty}^{d-1}$, the positive orthant of the 
-unit  sphere in infinite norm, with distribution $\Phi$. 
-%More  specifically, 
-%using the notation in Theorem 6 of \cite{RoSeWa2018a}, define $\bm{S} 
-%=  \log(\bm{V})$.  Then, according to Definition 5, $\bm{S}$ 
-%is a spectral  random vector. Taking  $E \sim \text{Exp}(1)$, 
-%independent of  $\bm{V}$,  we have that $\bm{Y} = \bm{S} + E$  
-%follows a multivariate GP, according  to Theorem 6.  Then, using 
-%Proposition 4,  $\bm{V}R = e^{\bm{Y}}$ will also follow
-%a multivariate GP. Notice that $R = e^E$ is distributed as 
-%a standard Pareto random variable. 
-$R$ and $\bm{V}$ are referred, respectively, as the {\em radial} 
+unit sphere under $\mathcal{L}_{\infty}$ norm, with distribution $\Phi$. 
+This representation is central to the methods proposed in this paper.
+$R$ and $\bm{V}$ are referred to, respectively, as the {\em radial} 
 and {\em angular} components of $H$. The angular measure controls 
-the dependence  structure of $\bm{W}$  in the tails. In view of 
+the dependence  structure of $\bm{Z}$  in the tails. In view of 
 this, to obtain a  PoT model we seek a flexible model for the 
@@
 
+The approach considered in \cite{rootzen2018} focuses on the limiting
+conditional distribution $H$. An alternative approach consists of assuming
+that regular variation \cite[see, for example,][]{resnick2008extreme} holds for 
+the limiting distribution of $\bm{W}$, implying
+\[
+    \lim\limits_{n\rightarrow\infty} n \prob{n^{-1} \bm{W}\in A} = 
+    \mu(A),
+\]
+where $\mu$ is termed the exponent measure. $\mu$ has the homogeneity property
+$\mu(tA) = t^{-1}\mu(A)$. Letting $\rho = \|W\|_p, p>0$ 
+and $\bm{\theta} = \bm{W}/\rho$, define  $\Psi(B) = \mu(\{\bm{w} : \rho>1, 
+\bm{\theta} \in B\})$, which is referred to as the angular measure. After 
+some manipulations, we obtain that
+\begin{equation}
+\label{exponent-mu}
+    \lim\limits_{r\rightarrow\infty} \text{Pr}\left[\bm{\theta}\in A \rvert \rho>r\right] = 
+    \frac{\Psi(A)}{\Psi({\mathbb S}_p^{d-1})} .
+\end{equation}
+Thus, a model for the exponent measure induces a model for the limiting distribution
+conditional on the observations being above a threshold defined with respect
+to their $p$-norm. The constraint that all marginals of $\mu$ correspond to a
+standard Pareto distribution leads to the so called moment constraints on $\Psi$. 
+Inference for the limiting distribution of the exceedances
+needs to account for the normalizing constant in Equation \ref{exponent-mu}, as well
+the moment constraints. In addition, the moment constraints can impose unrealistic
+symmetries in the data distribution.
+
 %\subection{Pairwise Extremal Dependence\label{subsec:ped}}
-A measure that is used to characterize the strength of the dependence,
-in the tail, for two random variables $Z_1$ and $Z_2$, with marginal
-distributions $F_1$ and $F_2$ is given by \citep{coles2001}.
-\[
-    \chi_{12} = \lim\limits_{u\uparrow 1} \prob{F_1(Z_1)>u\mid F_2(Z_2)>u}  
-\]
-$\chi_{12}$ provides information about the distribution of extremes for the variable $Z_1$
-given that $Z_2$ is very large.  When $\chi_{12}>0$, $Z_1$ and $Z_2$ 
-are said to
-be asymptotically dependent, otherwise they are asymptotically
-independent. The following result provides the asymptotic dependence coefficient between two components of $\bm{Z}$
-after PoT limit. 
-\begin{prop}\label{ppchi}
-Suppose that $\bm{Z} = R\bm{V}$ with $R\sim Pa(1)$,
-$\prob{V_\ell > 0} = 1$ and $\expect{V_\ell}$ exists, for $\ell=1, \ldots ,d$, then
-\begin{equation}
-    \label{eqn:chi_ij}
-	\chi_{\jmath\ell} = \expect{\frac{V_\jmath}{\expect{V_\jmath}} \wedge \frac{V_\ell}{\expect{V_\ell}}}
-\end{equation}
-\end{prop}
-{\em Proof:}
-Denote as $F_\ell$ the marginal distribution of $Z_\ell$. To obtain $\chi_{\jmath\ell}$ we need $Pr(Z_\jmath>z_\jmath,Z_\ell>z_\ell)$, where $z_\ell =
-F_\ell^{-1}(u) = \expect{V_\ell}/(1 - u), \;\ell=1,\dots, d$.
-Using the fact that
-$V_\ell>0, \forall \ell$ almost surely, we have that the former is equal to
-\begin{equation*}
-    \begin{aligned}
-    &\hspace{-0.5cm}\prob{R>\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}}
-    = \expect{1\wedge\left(\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}\right)^{-1}}\\
-    &= \expect{\frac{V_\jmath}{z_\jmath}\wedge\frac{V_\ell}{z_\ell}}
-    = (1 - u)\expect{\frac{V_\jmath}{\expect{V_\jmath}}\wedge\frac{V_\ell}{\expect{V_\ell}}},
-    \end{aligned}
-\end{equation*}
-where the second identity is justified by the fact that $V_i$ is
-bounded and $z_i\rightarrow\infty$. The proof is completed by noting
-that $\prob{F_i(Z_i)>u} = 1 - u. \hfill\Box$
-
-%\subsection{Multivariate Conditional Survival Function\label{subsec:condsurv}}
-Equation \ref{eqn:chi_ij} implies that $\chi_{\jmath\ell}>0$, and so, no asymptotic independence is possible under our proposed model.
+%%A measure that is used to characterize the strength of the dependence,
+%%in the tail, for two random variables $Z_1$ and $Z_2$, with marginal
+%%distributions $F_1$ and $F_2$ is given by \citep{coles2001}.
+%%\[
+%%    \chi_{12} = \lim\limits_{u\uparrow 1} \prob{F_1(Z_1)>u\mid F_2(Z_2)>u}  
+%%\]
+%%$\chi_{12}$ provides information about the distribution of extremes for the variable $Z_1$
+%%given that $Z_2$ is very large.  When $\chi_{12}>0$, $Z_1$ and $Z_2$ 
+%%are said to
+%%be asymptotically dependent, otherwise they are asymptotically
+%%independent. The following result provides the asymptotic dependence coefficient between two components of $\bm{Z}$
+%%after PoT limit. 
+%%\begin{prop}\label{ppchi}
+%%Suppose that $\bm{Z} = R\bm{V}$ with $R\sim Pa(1)$,
+%%$\prob{V_\ell > 0} = 1$ and $\expect{V_\ell}$ exists, for $\ell=1, \ldots ,d$, then
+%%\begin{equation}
+%%    \label{eqn:chi_ij}
+%%	\chi_{\jmath\ell} = \expect{\frac{V_\jmath}{\expect{V_\jmath}} \wedge \frac{V_\ell}{\expect{V_\ell}}}
+%%\end{equation}
+%%\end{prop}
+%%{\em Proof:}
+%%Denote as $F_\ell$ the marginal distribution of $Z_\ell$. To obtain $\chi_{\jmath\ell}$ we need $Pr(Z_\jmath>z_\jmath,Z_\ell>z_\ell)$, where $z_\ell =
+%%F_\ell^{-1}(u) = \expect{V_\ell}/(1 - u), \;\ell=1,\dots, d$.
+%%Using the fact that
+%%$V_\ell>0, \forall \ell$ almost surely, we have that the former is equal to
+%%\begin{equation*}
+%%    \begin{aligned}
+%%    &\hspace{-0.5cm}\prob{R>\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}}
+%%    = \expect{1\wedge\left(\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}\right)^{-1}}\\
+%%    &= \expect{\frac{V_\jmath}{z_\jmath}\wedge\frac{V_\ell}{z_\ell}}
+%%    = (1 - u)\expect{\frac{V_\jmath}{\expect{V_\jmath}}\wedge\frac{V_\ell}{\expect{V_\ell}}},
+%%    \end{aligned}
+%%\end{equation*}
+%%where the second identity is justified by the fact that $V_i$ is
+%%bounded and $z_i\rightarrow\infty$. The proof is completed by noting
+%%that $\prob{F_i(Z_i)>u} = 1 - u. \hfill\Box$
+%%
+%%%\subsection{Multivariate Conditional Survival Function\label{subsec:condsurv}}
+%%Equation \ref{eqn:chi_ij} implies that $\chi_{\jmath\ell}>0$, and so, no asymptotic independence is possible under our proposed model.
 For the analysis of extreme values it is of interest to calculate the multivariate conditional survival function.
 The following result provides the relevant expression, as a function of the angular measure.
-\begin{prop}
-Assume the same conditions of Proposition \ref{ppchi}. 
-Let $\alpha \subset \{1, \ldots ,d\}$ be a collections of indexes. 
-Then     
-\begin{equation} \label{eqn:condsurv}
-    \begin{aligned}
-    &\hspace{-0.5cm}\mathrm{Pr}\left[\bigcap_{\ell \in \alpha} Z_\ell > z_\ell \mid \bigcap_{\ell\not\in\alpha} Z_\ell > z_\ell\right] \\
-    &= \frac{\expect{\bigwedge_{k = 1}^d 1\wedge \frac{V_k}{z_k}}}{\expect{\bigwedge_{k \not\in\alpha}1\wedge\frac{V_k}{z_k}}}.
-    \end{aligned}
-  \end{equation}
-\end{prop}  
-The proof uses a similar approach to the proof of Proposition \ref{ppchi}.
-
+%%\begin{prop}
+%%Assume the same conditions of Proposition \ref{ppchi}. 
+%%Let $\alpha \subset \{1, \ldots ,d\}$ be a collections of indexes. 
+%%Then     
+%%\begin{equation} \label{eqn:condsurv}
+%%    \begin{aligned}
+%%    &\hspace{-0.5cm}\mathrm{Pr}\left[\bigcap_{\ell \in \alpha} Z_\ell > z_\ell \mid \bigcap_{\ell\not\in\alpha} Z_\ell > z_\ell\right] \\
+%%    &= \frac{\expect{\bigwedge_{k = 1}^d 1\wedge \frac{V_k}{z_k}}}{\expect{\bigwedge_{k \not\in\alpha}1\wedge\frac{V_k}{z_k}}}.
+%%    \end{aligned}
+%%  \end{equation}
+%%\end{prop}  
+%%The proof uses a similar approach to the proof of Proposition \ref{ppchi}.
+%%
   % EOF

### 2023-05-08 methodology_projgamma.tex
before some cutting

@@
 \subsection{Projected gamma family\label{subsec:projgamma}}
+To obtain a distribution on 
+${\mathbb S}_{p}^{d-1}=\{\bm{s} : \bm{s} \in {\mathbb R}_{+}^{d}, \| \bm{s}\|_{p} = 1\}$ 
+we start with a vector in 
+$\bm{x} \in {\mathbb R}^d_+$, calculate its $\mathcal{L}_p$-norm  defined as
+  \begin{equation*}
+    \lVert \bm{x} \rVert_p = 
+    \left({\textstyle\sum}_{\ell = 1}^d \lvert s_{\ell}\rvert^p\right)^{\frac{1}{p}},
+  \end{equation*}
+  and normalize it to obtain $\bm{y} = \bm{x}/\|\bm{x}\|_p \in {\mathbb S}_p^{d-1}$. 
+  The absolute and Euclidean norms are obtained for $p=1$ and $p=2$ respectively,
+  and the $\mathcal{L}_{\infty}$ norm can be obtained as a limit, 
+  \begin{equation*}
+    \| \bm{s} \|_{\infty}
+      = \lim\limits_{p\to\infty} \| \bm{s} \|
+      = \underset{\ell = 1}{\overset{d}{\bigvee}}s_{\ell}.
+  \end{equation*}
+
   A natural  distribution to consider in ${\mathbb R}^d_+$ is given by a product of independent
@@
     $\bm{ X} \sim \prod_{\ell = 1}^d\text{Ga}\left(X_{\ell}\mid\alpha_{\ell},\beta_{\ell}\right)$, 
-    $\alpha_\ell$ and $\beta_\ell$ are the shape and scale parameters, respectively. Using Equations (\ref{eqn:pnormt}) -- (\ref{eqn:pnormjac}) we have the joint density
+    $\alpha_\ell$ and $\beta_\ell$ are the shape and scale parameters, respectively. 
+    For any finite $p$, letting 
+   $ y_d = (1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p)^{1/p}$,
+    the transformation
+  \begin{equation}
+    \label{eqn:pnormt}
+    \begin{aligned}
+    T(x_1,\ldots,x_d) &= \left(\pnorm{\bm{x}}{p}, \frac{x_1}{\pnorm{\bm{x}}{p}},
+                          \ldots , \frac{x_{d-1}}{\pnorm{\bm{x}}{p}}\right) \\
+                        &= (r,y_1,\ldots,y_{d-1})
+    \end{aligned}
+  \end{equation}
+  is invertible with
+  \begin{equation}
+    \label{eqn:pnormtinv}
+    \begin{aligned}
+    &\hspace{-0.5cm}T^{-1}\left(r,y_1,\ldots,y_{d-1}\right)\\
+      & = \left(ry_1,\ldots,ry_{d-1}, r\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}\right).
+    \end{aligned}
+  \end{equation}
+  The Jacobian of the transformation takes the form
+  \begin{equation}
+    \label{eqn:pnormjac}
+    \begin{aligned}
+    r^{d-1}\left[\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}} \right. & \\
+        &\hspace{-2cm}+ \left. {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(1 - {\textstyle\sum}_{l=1}^{d-1} y_{\ell}^p\right)^{\frac{1}{p} - 1}\right].
+    \end{aligned}
+  \end{equation}
+  The normalization provided by $T$ maps a vector in ${\mathbb R}_+^d$ onto
+  ${\mathbb S}_p^{d-1} $. With a slight abuse of terminology we refer to it as a
+  projection. Using Equations (\ref{eqn:pnormt}) -- (\ref{eqn:pnormjac}) we have the joint density
   \begin{equation}
@@
   \end{equation}
-  defined for $\bm{y}\in {\mathbb S}_p^{d-1}$, and for any $p>0$. To avoid identifiability problems when estimating the shape and scale parameters, we set $\beta_1 = 1$.
+  defined for $\bm{y}\in {\mathbb S}_p^{d-1}$, and for any finite $p>0$. To avoid identifiability 
+  problems when estimating the shape and scale parameters, we set $\beta_1 = 1$.
   \cite{nunez2019} obtain the density in Equation (\ref{eqn:projgamma})
@@
 \end{figure}
+
+A measure that is used to characterize the strength of the dependence,
+  in the tail, for two random variables $Z_1$ and $Z_2$, with marginal
+  distributions $F_1$ and $F_2$ is given by \citep{coles2001}
+  \[
+    \chi_{12} = \lim\limits_{u\uparrow 1} \prob{F_1(Z_1)>u\mid F_2(Z_2)>u}.   
+  \]
+  $\chi_{12}$ provides information about the distribution of extremes for the variable $Z_1$
+  given that $Z_2$ is very large.  When $\chi_{12}>0$, $Z_1$ and $Z_2$ are said to
+  be asymptotically dependent, otherwise they are asymptotically independent. The following result 
+  provides the asymptotic dependence coefficient between two components of $\bm{Z}$ after PoT limit.
+  \begin{prop}\label{ppchi}
+  Suppose that $\bm{Z} = R\bm{V}$ with $R\sim Pa(1)$,
+  $\prob{V_\ell > 0} = 1$ and $\expect{V_\ell}$ exists, for $\ell=1, \ldots ,d$, then
+    \begin{equation}
+    \label{eqn:chi_ij}
+	\chi_{\jmath\ell} = \expect{\frac{V_\jmath}{\expect{V_\jmath}} \wedge \frac{V_\ell}{\expect{V_\ell}}}
+    \end{equation}
+  \end{prop}
+  {\em Proof:}
+  Denote as $F_\ell$ the marginal distribution of $Z_\ell$. To obtain $\chi_{\jmath\ell}$ we need 
+  $Pr(Z_\jmath>z_\jmath,Z_\ell>z_\ell)$, where $z_\ell = F_\ell^{-1}(u) = \expect{V_\ell}/(1 - u), \;\ell=1,\dots, d$.  
+  Using the fact that $V_\ell>0, \forall \ell$ almost surely, we have that the former is equal to
+  \begin{equation*}
+    \begin{aligned}
+    &\hspace{-0.5cm}\prob{R>\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}}
+    = \expect{1\wedge\left(\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}\right)^{-1}}\\
+    &= \expect{\frac{V_\jmath}{z_\jmath}\wedge\frac{V_\ell}{z_\ell}}
+    = (1 - u)\expect{\frac{V_\jmath}{\expect{V_\jmath}}\wedge\frac{V_\ell}{\expect{V_\ell}}},
+    \end{aligned}
+  \end{equation*}
+  where the second identity is justified by the fact that $V_i$ is bounded and $z_i\rightarrow\infty$. The 
+  proof is completed by noting that $\prob{F_i(Z_i)>u} = 1 - u. \hfill\Box$
+
+Equation \ref{eqn:chi_ij} implies that $\chi_{\jmath\ell}>0$, and so, no asymptotic independence is possible 
+  under our proposed model. For the analysis of extreme values it is of interest to calculate the multivariate 
+  conditional survival function. The following result provides the relevant expression, as a function of the 
+  angular measure.
+  \begin{prop}
+  Assume the same conditions of Proposition \ref{ppchi}.   Let $\alpha \subset \{1, \ldots ,d\}$ be a 
+  collections of indexes.  Then     
+  \begin{equation} \label{eqn:condsurv}
+    \begin{aligned}
+    &\hspace{-0.5cm}\mathrm{Pr}\left[\bigcap_{\ell \in \alpha} 
+        Z_\ell > z_\ell \mid \bigcap_{\ell\not\in\alpha} Z_\ell > z_\ell\right] \\
+    &= \frac{\expect{\bigwedge_{k = 1}^d 1\wedge \frac{V_k}{z_k}}}{\expect{\bigwedge_{k \not\in\alpha}1\wedge\frac{V_k}{z_k}}}.
+    \end{aligned}
+  \end{equation}
+  \end{prop}  
+  The proof uses a similar approach to the proof of Proposition \ref{ppchi}.
+
+Equations \ref{eqn:chi_ij} and \ref{eqn:condsurv} provide relevant tools for inference on the tail 
+  behavior of the joint distribution of the observations. The expressions can be readily calculated 
+  within a sample-based inferential approach like the one considered in the following section.
 

### 2023-05-17 pointbypoint.tex
minor update

@@
 
-    \comment{Both reviewers comment on the moment constraint.  I forget if it's this one, or
-    the reference provided by the other reviewer, but I don't think they understood the reference
-    directly; they were using the $\mathcal{L}_1$ norm in $p$ dimensions, rather than the 
-    $\mathcal{L}_p$ norm as described in our paper.  A massive difference in meaning of notation.
-    I'll find the reference again and confirm.}
+    % \comment{Both reviewers comment on the moment constraint.  I forget if it's this one, or
+    % the reference provided by the other reviewer, but I don't think they understood the reference
+    % directly; they were using the $\mathcal{L}_1$ norm in $p$ dimensions, rather than the 
+    % $\mathcal{L}_p$ norm as described in our paper.  A massive difference in meaning of notation.
+    % I'll find the reference again and confirm.}
 
@@
 
-We note in passing that the moment conditions for the spectral
-distribution, say $\Phi$, in the two-dimensional case, using the 
-infinity norm, imply that 
+We note in passing that the moment conditions for the spectral distribution, say 
+    $\Phi$, in the two-dimensional case, using the infinity norm, imply that 
     \[
@@
     \]
-The former looks like a fairly strong symmetry restriction that is
-unlike to be realistic in practice. Thus, imposing moment
-conditions would probably result in a very restrictive model.
+    The former looks like a fairly strong symmetry restriction that is unlike to 
+    be realistic in practice. Thus, imposing moment conditions would probably 
+    result in a very restrictive model.
 
-    \item \emph{Main contribution}: While I appreciate the interesting construction based on the 
-    DPM of projected gammas (which itself could be of use for other contexts beyond extremes?), 
-    one wonders about what would be the main added value of it with respect to other existing 
-    Bayesian approaches (\cite{boldi2007},\cite{guillotte2011},\cite{sabourin2014},\cite{hanson2017}).
-    Keeping in mind the computational scope of Statistics and Computing, one wonders if the main 
-    advantage is, for example, a computational one? This needs to be clarified—especially in light 
-    of the previous comment on the moment constraint. Indeed, one could use any of the previous 
-    approaches with pseudo-angles based on the infinity norm.
+\item \emph{Main contribution}: While I appreciate the interesting construction 
+    based on the DPM of projected gammas (which itself could be of use for other 
+    contexts beyond extremes?), one wonders about what would be the main added 
+    value of it with respect to other existing Bayesian approaches 
+    (\cite{boldi2007},\cite{guillotte2011},\cite{SaNa2014},\cite{hanson2017}).
+    Keeping in mind the computational scope of Statistics and Computing, one 
+    wonders if the main advantage is, for example, a computational one? This 
+    needs to be clarified—especially in light of the previous comment on the 
+    moment constraint. Indeed, one could use any of the previous approaches with 
+    pseudo-angles based on the infinity norm.
 
-    \comment{Hanson is clearly different.  They're using Bernstein polynomials effectively defined 
-    on $\mathbb{S}_1^{d-1}$.  We can demonstrate that projected gammas built on $\mathbb{S}_{p}$, 
-    $p \geq 10$ and $\mathbb{S}_1$ create \emph{very} different gamma shape parameters, for data
-    created by the \emph{same} distribution.  I need to check again the other references, but 
-    based on their follow-on comment, I'm guessing the previous point needs to be addressed before 
-    we can properly justify this point.}
+We have clarified the distinction between our approach and the prior contributions
+    mentioned.
 
-    \bruno{Based on our conversation, you need to look carefully into each one of these references.
-    it seems that most of them do not develop a PoT approach. But also, we use a kernel whose natural 
-    support is the positive cone (without using transformations) to build a mixture WRT to a random
-    measure. This provides the flexibility of mixture for a methods that has computational advantages 
-    and scales to moderately large problems, as illustrated in the paper.}
+    % \comment{Hanson is clearly different.  They're using Bernstein polynomials 
+    % effectively defined on $\mathbb{S}_1^{d-1}$.  We can demonstrate that 
+    % projected gammas built on $\mathbb{S}_{p}$, $p \geq 10$ and $\mathbb{S}_1$ 
+    % create \emph{very} different gamma shape parameters, for data created by 
+    % the \emph{same} distribution.  I need to check again the other references, 
+    % but based on their follow-on comment, I'm guessing the previous point needs 
+    % to be addressed before we can properly justify this point.}
 
-    \comment{\cite{hanson2017}: Bernstein polynomials on $\mathbb{S}_1^{d-1}$, modelling multivariate Frechets.
-\cite{boldi2007}: Marginal Frechets, also operating on $\mathbb{S}_1^{d-1}$.
-\cite{guillotte2011}: A strictly bivariate model.
-\cite{sabourin2014}: }
+    % \bruno{Based on our conversation, you need to look carefully into each one 
+    % of these references. it seems that most of them do not develop a PoT 
+    % approach. But also, we use a kernel whose natural support is the positive 
+    % cone (without using transformations) to build a mixture WRT to a random
+    % measure. This provides the flexibility of mixture for a methods that has 
+    % computational advantages and scales to moderately large problems, as 
+    % illustrated in the paper.}
 
-    \item \emph{Numerical accuracy}:  Keeping in mind the computational scope of Statistics and 
-    Computing one would hope for more numerical work than what is presented in §5.1. As far as I
-    understand what is presented are single run experiments for each $d$? Monte Carlo evidence on the
-    performance of the procedure would be appreciated. On top of this, the current numerical exercise
-    ignores the uncertainty that stems from step 1 of the procedure (i.e. estimating the margins).
+    \comment{\cite{Hanson2017}: Bernstein polynomials on $\mathbb{S}_1^{d-1}$, 
+    modelling multivariate Frechets.  \cite{boldi2007}: Marginal Frechets, also 
+    operating on $\mathbb{S}_1^{d-1}$.  \cite{guillotte2011}: A strictly bivariate model.  
+    \cite{SaNa2014}: }
 
-    \comment{Regarding the uncertainty from step 1, I think we agreed that was basically intractible.
-    We may need to explicitly specify as such, and that a practicioner must accept the numerical 
-    uncertainty that arises from step 1.  That said, I seem to recall the numbers observed being 
-    relatively robust to choice of threshold (for large threshold), the difference mainly coming 
-    down to computation time.}
+\item \emph{Numerical accuracy}:  Keeping in mind the computational scope of 
+    Statistics and Computing one would hope for more numerical work than what is 
+    presented in §5.1. As far as I understand what is presented are single run 
+    experiments for each $d$? Monte Carlo evidence on the performance of the 
+    procedure would be appreciated. On top of this, the current numerical exercise
+    ignores the uncertainty that stems from step 1 of the procedure 
+    (i.e. estimating the margins).
 
-    \bruno{I think we need to conduct a more comprehensive suite of experiments. Regarding the uncertainty 
-    coming from Step1, we just need to acknowledge that we are ignoring it.}
+To address this concern, we have revised the simulation study to feature the 
+    average rise in energy score over baseline, averaged over 10 replications.  
+    The simulated datasets are generated as mixtures of multivariate lognormals, 
+    projected onto the unit sphere.
 
-    To address this concern, we have \comment{revised (note: future past tense)} the simulation study 
-    to feature the average rise in energy score over baseline, averaged over 10 iterations.  The simulated
-    datasets are generated as mixtures of multivariate lognormals, projected onto the unit sphere.
+    % \comment{Regarding the uncertainty from step 1, I think we agreed that was 
+    % basically intractible. We may need to explicitly specify as such, and that a 
+    % practicioner must accept the numerical uncertainty that arises from step 1.  
+    % That said, I seem to recall the numbers observed being relatively robust to 
+    % choice of threshold (for large threshold), the difference mainly coming down 
+    % to computation time.}
 
-    \item \emph{Prior and MCMC}: A lot is left to the reader in §5 in terms of prior specification
-    and MCMC settings.  At least some further details would be required.
+    % \bruno{I think we need to conduct a more comprehensive suite of experiments. 
+    % Regarding the uncertainty coming from Step1, we just need to acknowledge that 
+    % we are ignoring it.}
 
-    \comment{The form of the prior is specified.  I'm guessing he's referring to the actual values
-    used for computation?  $\bm{\mu}_0 = \bm{0}$, $\nu = d + 50$, $\bm{\Psi} = \nu I_d$, and so on?}
+\item \emph{Prior and MCMC}: A lot is left to the reader in §5 in terms of
+    prior specification and MCMC settings.  At least some further details would 
+    be required.
+
+    \comment{The form of the prior is specified.  I'm guessing he's referring to 
+    the actual values used for computation?  $\bm{\mu}_0 = \bm{0}$, 
+    $\nu = d + 50$, $\bm{\Psi} = \nu I_d$, and so on?}
 
@@
 
-    \item \emph{Future research}: §6 calls for a view of the authors on what could be related 
-    exercises—or even extensions of this research and future work. Using the proposed Bayesian 
-    multivariate PoT construction for devising a regression model for an an extreme value response 
-    on an extreme value covariate is one example of a natural follow-up; see \cite{carvalho2022}
-    for related ideas. Conditional versions of the angular measure (e.g. \cite{carvalho2016},
-    \cite{castro2018}, \cite{escobar2018}, \cite{mhalla2019}) would lead to a conditional 
-    multivariate peaks-over-threshold model, where dependence between exceedances could change 
-    along with a covariate. Also, perhaps the here proposed DPM of projected gammas—as well as 
-    predictor-dependent versions of it—could be of interest in other settings beyond angular 
-    measures?
+\item \emph{Future research}: §6 calls for a view of the authors on what 
+    could be related exercises—or even extensions of this research and future 
+    work. Using the proposed Bayesian multivariate PoT construction for 
+    devising a regression model for an an extreme value response on an extreme 
+    value covariate is one example of a natural follow-up; see 
+    \cite{carvalho2022} for related ideas. Conditional versions of the angular 
+    measure (e.g. \cite{carvalho2016}, \cite{castro2018}, \cite{escobar2018}, 
+    \cite{mhalla2019}) would lead to a conditional multivariate 
+    peaks-over-threshold model, where dependence between exceedances could 
+    change along with a covariate. Also, perhaps the here proposed DPM of 
+    projected gammas—as well as predictor-dependent versions of it—could be of 
+    interest in other settings beyond angular measures?
 
@@
 
-    \bruno{I have no doubts that this referee is Carvalho. I'll spend some time reading the papers
-    that he mentions.}
+    \bruno{I have no doubts that this referee is Carvalho. I'll spend some time 
+    reading the papers that he mentions.}
 
@@
 
+    We have included for comparison some other models established on $\mathcal{S}_{1}^{d-1}$;
+        notably the pairwise betas model developed in \cite{sabourin2013}, as well
+        as the dirichlet distribution.  We have investigated several packages to
+
     \bruno{Well, here we have to bite the bullet. Hopefully there some software that allows us
@@
     (with varying points $x_i,x_j$), not that \[\sum_i\sum_ja_ia_jg(x_1,x_2)\leq 0.\]  Thus the vector $e$
-    which allows to break the sum into two parts in the current proof should in fact depend on $(i,j)$, and the proof is not valid.
+    which allows to break the sum into two parts in the current proof should in fact depend on $(i,j)$, 
+    and the proof is not valid.
 
@@
 
-    \item \emph{Minor point 3}: (typos): Section 3, first paragraph: Some confusions between $\ell \in \lbrace 1,\ldots, n\rbrace$ and $\ell \in \lbrace 1,\ldots,d\rbrace$
+    \item \emph{Minor point 3}: (typos): Section 3, first paragraph: Some confusions between 
+        $\ell \in \lbrace 1,\ldots, n\rbrace$ and $\ell \in \lbrace 1,\ldots,d\rbrace$
 

### 2023-05-17 introduction.tex
OL 2023_10_15
minor update

@@
 Inferential methods for the extreme values of univariate observations
-are well established and software is widely available \cite[see, for example,][]{coles2001}. For variables 
-in one dimension the application of EVT methods considers the asymptotic distribution of either the maxima 
-calculated for regular blocks of data, or the values that exceed a certain threshold. The former leads to
-a Generalized Extreme Value (GEV) distribution, that depends on three parameters. The latter leads to a 
-Generalized Pareto (GP) distribution, that depends on a shape and a scale parameter. Likelihood-based 
-approaches to inference can be readily implemented in both cases.
-In the multivariate case the GEV theory is well developed \citep[see, for example][]{dehaan2006}, but the
-inferential problem is more complicated, as there is no parametric representation of the GEV. This problem
-is inherited by the peaks over threshold (PoT) approach and compounded by the fact that there is no unique definition of an 
-exceedance of a multivariate threshold, as there is an obvious dependence on the norm that is used to
-measure the size of a vector.
+    are well established and software is widely available 
+    \cite[see, for example,][]{coles2001}. For variables in one dimension the 
+    application of EVT methods considers the asymptotic distribution of either the 
+    maxima calculated for regular blocks of data, or the values that exceed a 
+    certain threshold. The former leads to a Generalized Extreme Value (GEV) 
+    distribution, that depends on three parameters. The latter leads to a Generalized 
+    Pareto (GP) distribution, that depends on a shape and a scale parameter. 
+    Likelihood-based approaches to inference can be readily implemented in both cases.
+    In the multivariate case the GEV theory is well developed 
+    \citep[see, for example][]{dehaan2006}, but the inferential problem is more 
+    complicated, as there is no parametric representation of the GEV. This problem
+    is inherited by the peaks over threshold (PoT) approach and compounded by the 
+    fact that there is no unique definition of an exceedance of a multivariate 
+    threshold, as there is an obvious dependence on the norm that is used to measure 
+    the size of a vector.
 
-During the last decade or so, much work has been done in the exploration of the definition 
-  and properties of an appropriate generalization of the univariate GP distribution 
-  to a multivariate setting.  To mention some of the papers on the topic, the work of
-  \citep{rootzen2006} defines the generalized Pareto distribution, with further analysis on these classes
-  of distributions presented in \cite{falk2008} and \cite{michel2008}.  A recent review of the state
-  of the art in multivariate peaks over threshold modelling using generalized Pareto is provided in
-  \cite{rootzen2018} while \cite{RoSeWa2018a} provides insight on the theoretical properties of possible parametrizations.
-These are use in \cite{KiRoSeWa2019} for likelihood-based models for PoT estimation.
-  A frequently used method for describing the dependence
-  in multivariate distributions is to use a copula.
-\cite{renard2007}, and \cite{falk2019} provide successful examples of this approach in an EVT framework.
-\cite{ferreira2014} presents a constructive definition of the Pareto process, that generalizes the GP to an infinite dimensional setting.
-It consists of decomposing the process into independent radial and angular components.
-Such an approach can be used in the finite dimensional case, where the angular component
-  contains the information pertaining to the dependence structure of the random vector.
-  Based on this definition, we present a novel approach for modelling  the angular component
-  with families of distributions that provide flexibility and can be applied in a moderately large dimensional setting.
-  Our focus on the  angular measure is similar to that in \cite{SaNa2014} and \cite{HaCaCh2017}, that consider Bayesian non-parametric approaches.
-Yet, our approach avoids the problem of dealing with the so called moment constraint by considering a constructive definition of the multivariate GP based on the infinity norm.
+During the last decade or so, much work has been done in the exploration of the 
+    definition and properties of an appropriate generalization of the univariate 
+    GP distribution to a multivariate setting.  To mention some of the papers on 
+    the topic, the work of \citep{rootzen2006} defines the generalized Pareto 
+    distribution, with further analysis on these classes of distributions 
+    presented in \cite{falk2008} and \cite{michel2008}.  A recent review of the 
+    state of the art in multivariate peaks over threshold modelling using 
+    generalized Pareto is provided in \cite{rootzen2018} while \cite{RoSeWa2018a} 
+    provides insight on the theoretical properties of possible parametrizations. 
+    These are use in \cite{KiRoSeWa2019} for likelihood-based models for PoT 
+    estimation. A frequently used method for describing the dependence in 
+    multivariate distributions is to use a copula. \cite{renard2007}, and 
+    \cite{falk2019} provide successful  examples of this approach in an EVT 
+    framework. \cite{ferreira2014} presents a constructive definition of the 
+    Pareto process, that generalizes the GP to an infinite dimensional setting. It 
+    consists of decomposing the process into independent radial and angular 
+    components. Such an approach can be used in the finite dimensional case, where 
+    the angular component contains the information pertaining to the dependence 
+    structure of the random vector. Based on this definition, we present a novel 
+    approach for modelling  the angular component with families of distributions 
+    that provide flexibility and can be applied in a moderately large dimensional 
+    setting.  Our focus on the angular measure is similar to that in \cite{boldi2007},
+    \cite{SaNa2014} and \cite{HaCaCh2017}, that consider Bayesian non-parametric 
+    approaches. Yet, our approach differs in that it is established in the 
+    peaks-over-threshold regime, and avoids the problem of dealing with the so 
+    called moment constraint by considering a constructive definition of the 
+    multivariate GP, based on the infinity norm.  Further, the
 
 The reminder of this paper is outlined as follows. 
-Section~\ref{sec:multivariatepot} comprises a brief review of multivariate PoT, detailing the separation of the radial measure from the angular measure.
-  Section~\ref{sec:methodology} details our approach for estimating the angular measure, based on projecting an arbitrary distribution supported in ${\mathbb R}_+^d$ onto unit hyper-spheres defined for $p$-norms.
-Section~\ref{sec:evaluation} develops criteria for model selection in the support of the angular measure.
-Section~\ref{sec:results} explores the efficacy of the proposed approach on a set of simulated data, and, acknowledging the relevance of extreme value theory to climatological events~\citep{jentsch2007,vousdoukas2018,li2019}, estimates the extremal dependence structure for a measure of water vapor flow in the atmosphere, used for identifying atmospheric rivers.
+    Section~\ref{sec:multivariatepot} comprises a brief review of multivariate PoT, 
+      detailing the separation of the radial measure from the angular measure.
+    Section~\ref{sec:methodology} details our approach for estimating the angular 
+      measure, based on projecting an arbitrary distribution supported in ${\mathbb R}_+^d$ 
+      onto unit hyper-spheres defined for $p$-norms. 
+    Section~\ref{sec:evaluation} develops criteria for model selection in the support 
+      of the angular measure.  
+    Section~\ref{sec:results} explores the efficacy of the proposed approach on a set 
+      of simulated data, and, acknowledging the relevance of extreme value theory to 
+      climatological events~\citep{jentsch2007,vousdoukas2018,li2019}, estimates the 
+      extremal dependence structure for a measure of water vapor flow in the atmosphere, 
+      used for identifying atmospheric rivers.  
     Finally, Section~\ref{sec:conclusion} presents our conclusions and discussion.
@@
 Throughout the paper, we adopt the operators $\wedge$ to denote minima, and 
-    the $\vee$ to denote maxima.  Thus $\wedge_i s_i = \min_i s_i$, and $\vee_i s_i = \max_i s_i$.  
-    These operators can also be applied component-wise between vectors, such as
-    $\bm{a}\wedge\bm{b} = (a_1\wedge b_1, a_2\wedge b_2,\ldots)$.  We use 
-    uppercase to indicate random variables, lowercase to indicate points, and
+    the $\vee$ to denote maxima.  Thus $\wedge_i s_i = \min_i s_i$, and 
+    $\vee_i s_i = \max_i s_i$.  These operators can also be applied component-wise 
+    between vectors, such as $\bm{a}\wedge\bm{b} = (a_1\wedge b_1, a_2\wedge b_2,\ldots)$.  
+    We use uppercase to indicate random variables, lowercase to indicate points, and
     bold-face to indicate vectors or matrices thereof.
@@
 \section{A multivariate PoT model\label{sec:multivariatepot}}
-%To develop a multivariate PoT model for extreme values consider a 
-%$d$-dimensional random vector $\bm{W} = (W_1, \ldots ,W_d)$  with 
-%marginal distributions $F_1(w_1), \dots , F_d(w_d)$, and define its
-%standardization as $\bm{Z} = (Z_1, \ldots , Z_d)$, where $Z_i = 1/ (1 - F_i(W_i)), \; i=1, \ldots , d$.
-%Following the discussion in \cite{goix2017} we assume a form of 
-%regular variation, consisting on assuming the existence of a 
-%limiting measure $\mu$ on $(0,\infty]^d$, such that.
-%\begin{equation}\label{eq:reg-var}
-%        \lim\limits_{n\to\infty}n\prob{\frac{1}{n}\bm{Z}\not\leq \bm{z}} = \mu\left([\bm{0},\bm{z}]^c\right),
-%\end{equation}
-%where the inequality and the interval are taken componentwise. 
-%In this setting, $\mu$ is the asymptotic distribution of $\bm{Z}$ 
-%in extreme regions, and it is  referred to as the \emph{exponent
-%measure}. In addition, we assume that $\lim_{n\rightarrow\infty} 
-%n\prob{\max\{Z_1, \ldots , Z_d\}>n} = 1$. Equation \ref{eq:reg-var}
-%implies that $\mu$ has the homogeneity property $\mu(t\cdot) = 
-%\frac{1}{t}\mu(\cdot)$i, and it can be shown that  $\mu$ 
-%can be decomposed into a radial and an angular component. In fact, 
-%for $\bm{z} = (z_1, \ldots , z_n) \in (0,\infty]^d$, let $r(\bm{z}) = 
-%||\bm{z}||_\infty$ and $\bm{v}(\bm{z}) = \bm{z}/r(\bm{z})$. Define 
-%$\Phi(B) = \mu(\{\bm{z} : r(\bm{z})>1, \bm{v}(\bm{z}) \in B\})$, for any
-%$B\subset \mathbb{S}_{\infty}^{d-1}$, the positive orthant of the unit 
-%sphere in infinite norm. $\Phi$ is referred to as the {\em angular 
-%measure}. Then 
-%\begin{equation}\label{eq:exp-measure}
-%    \mu\left(\left\lbrace  \bm{z} \in [0,\infty)^d : \lVert \bm{z} \rVert_{\infty} > r,\;\bm{z}/\lVert\bm{z}\rVert_{\infty} \in B\right\rbrace \right) = r^{-1}\Phi(B).
-%\end{equation}
-%Thus, the radial and angular components of the exponent measure are
-%independent. To obtain a multivariate peak over threshold model we 
-%consider the radial and angular component of the random variable $\bm{Z}$
-%denoted, respectively, as $R = \max_{i} Z_i$ and $\bm{V} = \bm{Z}/R$. 
-%Then, using Equations \ref{eq:reg-var} and \ref{eq:exp-measure}, we have
-%that
-
-%To obtain a statistical approach to estimating $\mu$, we follow the approach proposed in .
-%Define  , and assume that it is distributed as a standard Pareto, and thus $R>1$ with probability one.
-Also, define , and assume that $\bm{V}$ has probability distribution $\Phi$ concentrated in $\mathbb{S}_{\infty}^{d-1}$, the positive orthant of the unit sphere in infinite norm.
-Also, assume that $E(V_i)>0$, and that $ \bm{Z}$ and $R$ are independent.
-%We refer to $R$ and $\bm{V}$ as, respectively, the  of $\bm{Z}$. Under those assumptions,  the convergence stated in Equation \ref{eq:reg-var} holds with limit measure $\mu$ determined by
-
-%\citep[e.g.][applied to the finite-dimensional case]{ferreira2014}
-%\[
-%    \mu\left(\left\lbrace  \bm{z} \in [0,\infty)^d : \lVert \bm{z} \rVert_{\infty} > r,\;\bm{z}/\lVert\bm{z}\rVert_{\infty} \in A\right\rbrace \right) = r^{-1}\Phi(A)
-%\]
-%for $r > 0$ and for Borel sets $A\subseteq \mathbb{S}_{\infty}^{d-1}$.
-
-%    $\mu(tA) = \frac{1}{t}\mu(A)$.  Define  $R = \max_{i} Z_i$, and assume that $R>1$ with probability one. 
-%    Also, define $V_i = Z_i/\max_i Z_i$, and assume that $E(V_i)>0$. These two assumptions together with the 
-%    homogeneity of $\mu$ correspond to the second condition in Theorem 2.1 of
-%    \cite{ferreira2014}, applied to the finite-dimensional case.
-%    Thus, we have the factorization $\bm{Z} =  R\bm{V}$, where $R$ and $\bm{V}$  are independent. Moreover,
-%    for some measure $\Phi$, $\mu(\cdot)$ is factorized as
-%    \temp{\begin{equation}\label{mufact}
-%        %\mu\left( \left\lbrace\bm{Z} : V \in A , R > r\right\rbrace \right) = r^{-1}\Phi(A).
-%        \mu\left(\left\lbrace \bm{z}\in \mathbb{R}_+^d : R(\bm{z}) > r, V(\bm{z}) \in A\right\rbrace\right) = r^{-1}\Phi(A)
-%    \end{equation}
-%    for $R(\bm{Z}) > 0$, and Borel sets $A \subseteq \mathbb{S}_{\infty}^{d-1}$.  Hereafter $R$ is referred
-%    to as the radial component, and $\bm{V}$ the angular component of $\bm{Z}$
-%    \st{$R$ describes the radial component, and $\bm{V} \in {\mathbb S}_{\infty}^{d-1}$ the 
-%    angular component of $\bm{Z}$.}}  Here ${\mathbb S}_\infty^{d-1}$ denotes the positive
-%    orthant of the unit sphere in infinite norm.  $\Phi(\cdot)$ is referred to as the \emph{angular measure}.   
-%    From Equation (\ref{mufact}) we see that $R$ is distributed as a standard Pareto random variable. Conditioning on $R$, we have that
-%    \begin{equation} \label{MGP}
-%        \prob{\bm{V} \in B \mid R > r} = \frac{r\prob{\bm{V} \in B, R > r}}{r\prob{R > r}}\hspace{0.2cm}
-%            \xrightarrow[r\to\infty]{~} \hspace{0.2cm} \Phi(B),
-%    \end{equation}
-%as $\lim_{r\rightarrow\infty} r\prob{R>r} = 1$.
-%    Thus, the angular measure corresponds to the distribution of $\bm{V}$, conditional on $R$ being large.
-%    As such, it controls the dependence structure of $\bm{Z}$ in the tails. 
-%    In fact, assuming that $E(V_i)>0$ 
-%    and following \cite{ferreira2014}, in the limit the distribution function of $\bm{Z}$ is given as
-%    \begin{equation*}
-%        \prob{Z_1\leq z_1, \ldots ,Z_d\leq z_d} = \expect{\bigvee_{i=1}^d
-%    \frac{V_i}{z_i\wedge 1}} - \expect{\bigvee_{i=1}^d
-%    \frac{V_i}{z_i}}
-%    \end{equation*}
-%    where the expectation is taken with respect to $\Phi$
-%In view of this, to obtain a PoT model we consider a limiting 
-%distribution for $\bm{Z}$ that corresponds to the distribution
-%of the product of two independent random variables $\tilde{R}$ 
-%and $\tilde{\bm{V}}$, with distributions, respectively, standard 
-%Pareto and $\Phi$. \cite{ferreira2014} refer to such model as a 
-%simple  Pareto process indexed on a finite dimensional set.  We 
-%seek a flexible model for the distribution
-%of $\tilde{\bm{V}} \in {\mathbb S}_{\infty}^{d-1}$.
-
-%An alternative approach to justify the above mentioned limiting
-%distribution, that provides a link to the multivariate GEV and the
-%theory of max-stable distributions, is discussed 
-%in Remark 1 of \cite{RoSeWa2018a}. More specifically, 
-%using the notation in Theorem 6 of \cite{RoSeWa2018a}, define $\bm{S} 
-%=  \log(\tilde{\bm{V}})$.  Then, according to Definition 5, $\bm{S}$ 
-%is a spectral  random vector. Taking  $E \sim \text{Exp}(1)$, 
-%independent of  $\tilde{\bm{V}}$,  we have that $\bm{Y} = \bm{S} + E$  
-%follows a multivariate GP, according  to Theorem 6.  Then, using 
-%Proposition 4,  $\tilde{\bm{V}}\tilde{R} = e^{\bm{Y}}$ will also follow
-%a multivariate GP. Notice that $\tilde{R} = e^E$ is distributed as 
-%a standard Pareto random variable. This is one of several stochastic
-%representations of a multivariate GP presented in \cite{RoSeWa2018a}.
-
 To develop a multivariate PoT model for extreme values consider a 
@@
     conditional distribution $H$. An alternative approach consists of assuming
-that regular variation \cite[see, for example,][]{resnick2008extreme} holds for 
-the limiting distribution of $\bm{W}$, implying
+    that regular variation \cite[see, for example,][]{resnick2008extreme} holds 
+    for the limiting distribution of $\bm{W}$, implying
     \[
@@
     \label{exponent-mu}
-    \lim\limits_{r\rightarrow\infty} \text{Pr}\left[\bm{\theta}\in A \rvert \rho>r\right] = 
+        \lim\limits_{r\rightarrow\infty} 
+        \text{Pr}\left[\bm{\theta}\in A \rvert \rho>r\right] = 
             \frac{\Psi(A)}{\Psi({\mathbb S}_p^{d-1})} .
@@
 
-%\subection{Pairwise Extremal Dependence\label{subsec:ped}}
-%%A measure that is used to characterize the strength of the dependence,
-%%in the tail, for two random variables $Z_1$ and $Z_2$, with marginal
-%%distributions $F_1$ and $F_2$ is given by \citep{coles2001}.
-%%\[
-%%    \chi_{12} = \lim\limits_{u\uparrow 1} \prob{F_1(Z_1)>u\mid F_2(Z_2)>u}  
-%%\]
-%%$\chi_{12}$ provides information about the distribution of extremes for the variable $Z_1$
-%%given that $Z_2$ is very large.  When $\chi_{12}>0$, $Z_1$ and $Z_2$ 
-%%are said to
-%%be asymptotically dependent, otherwise they are asymptotically
-%%independent. The following result provides the asymptotic dependence coefficient between two components of $\bm{Z}$
-%%after PoT limit. 
-%%\begin{prop}\label{ppchi}
-%%Suppose that $\bm{Z} = R\bm{V}$ with $R\sim Pa(1)$,
-%%$\prob{V_\ell > 0} = 1$ and $\expect{V_\ell}$ exists, for $\ell=1, \ldots ,d$, then
-%%\begin{equation}
-%%    \label{eqn:chi_ij}
-%%	\chi_{\jmath\ell} = \expect{\frac{V_\jmath}{\expect{V_\jmath}} \wedge \frac{V_\ell}{\expect{V_\ell}}}
-%%\end{equation}
-%%\end{prop}
-%%{\em Proof:}
-%%Denote as $F_\ell$ the marginal distribution of $Z_\ell$. To obtain $\chi_{\jmath\ell}$ we need $Pr(Z_\jmath>z_\jmath,Z_\ell>z_\ell)$, where $z_\ell =
-%%F_\ell^{-1}(u) = \expect{V_\ell}/(1 - u), \;\ell=1,\dots, d$.
-%%Using the fact that
-%%$V_\ell>0, \forall \ell$ almost surely, we have that the former is equal to
-%%\begin{equation*}
-%%    \begin{aligned}
-%%    &\hspace{-0.5cm}\prob{R>\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}}
-%%    = \expect{1\wedge\left(\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}\right)^{-1}}\\
-%%    &= \expect{\frac{V_\jmath}{z_\jmath}\wedge\frac{V_\ell}{z_\ell}}
-%%    = (1 - u)\expect{\frac{V_\jmath}{\expect{V_\jmath}}\wedge\frac{V_\ell}{\expect{V_\ell}}},
-%%    \end{aligned}
-%%\end{equation*}
-%%where the second identity is justified by the fact that $V_i$ is
-%%bounded and $z_i\rightarrow\infty$. The proof is completed by noting
-%%that $\prob{F_i(Z_i)>u} = 1 - u. \hfill\Box$
-%%
-%%%\subsection{Multivariate Conditional Survival Function\label{subsec:condsurv}}
-%%Equation \ref{eqn:chi_ij} implies that $\chi_{\jmath\ell}>0$, and so, no asymptotic independence is possible under our proposed model.
-For the analysis of extreme values it is of interest to calculate the multivariate conditional survival function.
-The following result provides the relevant expression, as a function of the angular measure.
-%%\begin{prop}
-%%Assume the same conditions of Proposition \ref{ppchi}. 
-%%Let $\alpha \subset \{1, \ldots ,d\}$ be a collections of indexes. 
-%%Then     
-%%\begin{equation} \label{eqn:condsurv}
-%%    \begin{aligned}
-%%    &\hspace{-0.5cm}\mathrm{Pr}\left[\bigcap_{\ell \in \alpha} Z_\ell > z_\ell \mid \bigcap_{\ell\not\in\alpha} Z_\ell > z_\ell\right] \\
-%%    &= \frac{\expect{\bigwedge_{k = 1}^d 1\wedge \frac{V_k}{z_k}}}{\expect{\bigwedge_{k \not\in\alpha}1\wedge\frac{V_k}{z_k}}}.
-%%    \end{aligned}
-%%  \end{equation}
-%%\end{prop}  
-%%The proof uses a similar approach to the proof of Proposition \ref{ppchi}.
-%%
 % EOF

### 2023-05-17 pointbypoint.tex
minor update

@@
 We have clarified the distinction between our approach and the prior contributions
-    mentioned.
+    mentioned.  Setting aside \cite{guillotte2011} which is a strictly bivariate
+    model, the named approaches consider batch extrema and by construction must
+    respect the moment constraint.  As noted, our approach is able to set that 
+    aside, allowing consideration of a wider class of distributions of pseudo-angles.
 
@@
 
-    \comment{\cite{Hanson2017}: Bernstein polynomials on $\mathbb{S}_1^{d-1}$, 
-    modelling multivariate Frechets.  \cite{boldi2007}: Marginal Frechets, also 
-    operating on $\mathbb{S}_1^{d-1}$.  \cite{guillotte2011}: A strictly bivariate model.  
-    \cite{SaNa2014}: }
+    % \comment{\cite{Hanson2017}: Bernstein polynomials on $\mathbb{S}_1^{d-1}$, 
+    % modelling multivariate Frechets.  \cite{boldi2007}: Marginal Frechets, also 
+    % operating on $\mathbb{S}_1^{d-1}$.  \cite{guillotte2011}: A strictly bivariate model.  
+    % \cite{SaNa2014}: }
 
@@
     The simulated datasets are generated as mixtures of multivariate lognormals, 
-    projected onto the unit sphere.
+    projected onto the unit sphere.  As you say, this paper focuses on estimating
+    the spectral distribution, and ignores uncertainty from estimating marginal
+    parameters.  We have clarified as such.
 

### 2023-05-23 results.tex
Updates from Overleaf

@@
     \centering
-    \caption{Energy score (top row) and Posterior Predictive Loss (bottom row) for various models on simulated data with ascending count of mixture components (indicated by plot heading) and number of dimensions (indicated by horizontal axis).
+    % \caption{Energy score (top row) and Posterior Predictive Loss (bottom row) for various models on simulated data with ascending count of mixture components (indicated by plot heading) and number of dimensions (indicated by horizontal axis).
 Gen indicates baseline.\label{fig:simpples}}
-    \includegraphics[height=3in, width = \textwidth]{./images/simulation_score}
+    \caption{Average energy score rise over baseline (on $\mathbb{S}_{\infty}^{d-1}$) for various models on simulated data, with ascending count of mixture components (indicated by plot heading) and number of dimensions (indicated by horizontal axis).
+Note that pairwise betas is a moment-restricted model.\label{fig:simpples}}
+    \includegraphics[height=3in, width = \textwidth]{./images/sim_es_rise}
 \end{figure*}
@@
 \subsection{Simulations\label{subsec:simulated}}
+To evaluate our proposed approach for angular measure estimation 
+    we consider simulated datasets on $\mathbb{S}_{\infty}^{d-1}$,
+    for values of $d$ ranging between 2 and 32.  We generated each
+    dataset as a mixture of multivariate log-normal distributions,
+    projected onto $\mathbb{S}_p^{d-1}$. The generation procedure is detailed 
+    in Algorithm~\ref{algo:simulated}.
 
-To evaluate our proposed approach for angular measure estimation 
-    we consider simulated datasets on $\mathbb{S}_{\infty}^{d-1}$, for 
-    $d$ ranging between 3 and 20. We generated each dataset from a mixture of projected Gammas, with 
-    the number of mixture components ranging between 3 and 12.  The generation procedure is detailed 
-    in Algorithm~\ref{algo:simulated}.
 \begin{algorithm}[ht]  % algorithm using algorithm package
@@
         \begin{algorithmic}
-        \For{$n_{\text{mix}}$ in $[3, 6, 9, 12]$}
-            \State Generate $n_{\text{mix}} \times 20$ shape Parameters $\alpha$
-            \State Generate $n_{\text{mix}} \times 20$ Rate Parameters $\beta$
-            \State Generate $500$ Mixture Component Identifiers $\delta$
-            \For{$i$ in $1,\ldots,500$}
-                \State Generate ${\bf X}_i \sim \prod_{\ell = 1}^d\text{Ga}\left(X_{il}\mid\alpha_{\delta_i,l},\beta_{\delta_i, l}\right)$
+        \For{$n_{\text{iter}}$ in $[1,\ldots,10]$}
+        \For{$n_{\text{mix}}$ in $[1, 2, 4, 8]$}
+            \For{$j$ in $1,\ldots,n_{\text{mix}}$}
+                \State Generate $\mu_{j} \sim \mathcal{N}\left(\bm{0},\bm{I}\right)$
+                \State Generate $\Sigma_{j}\sim\mathcal{IW}\left(70,70 \bm{I}\right)$
             \EndFor
-            \For{$n_{\text{col}}$ in $[3,6,12,20]$}
+            \State Generate $\pi\sim\text{Dirichlet}(\bm{10})$
+            \For{$i$ in $1,\ldots,1000$}
+                \State Generate $\delta_i \sim \text{Categorical}(\pi)$
+                \State Generate $\bm{X}_i \sim \mathcal{LN}\left(\mu_{[\delta_i]},\Sigma_{[\delta_i]}\right)$
+            \EndFor
+            \For{$n_{\text{col}}$ in $[2,4,8,16,24,32]$}
                 \State Project columns 1 to $n_{\text{col}}$ of ${\bf X}$ onto $\mathcal{S}_{\infty}^{n_{\text{col}} - 1}$ and save.
+            \EndFor
         \EndFor
@@
 
-Each one of the shape parameters $\alpha$ were generated 
-$\alpha = \alpha_0 + \alpha_1$ where
-  $\alpha_0 \sim \text{Unif}(\alpha_0\mid 0,4)$, $\alpha_1\sim \text{Gamma}(\alpha_1\mid 1,1)$.
-The scale parameters $\beta$ were obtained as $\beta\sim\text{Unif}(\beta\mid 0.25, 2.5)$.
+where $\mu_j$, $\Sigma_j$ the parameters of the log-normal mixture component 
+    distribution; $\pi$ the probability vector assigning weight to each mixture 
+    component, and $\delta_i$ the mixture component identifier for each simulated 
+    observation.
 
-Figure~\ref{fig:simpples}, shows two sets of panels that illustrate the comparison
+Figure~\ref{fig:simpples}, shows the average rise over baseline in energy score as
+    calculated on $\mathbb{S}_{\infty}^{d-1}$ using the kernel metric described
+    in Proposition~\ref{prop:rot}, for models trained on simulated data.  After training
+    a model, a posterior predictive dataset is generated, and the energy score is
+    calculated as a Monte Carlo approximation of Equation~\ref{eq:es}.
+    The \emph{baseline} value is the energy score of a new dataset from the same
+    generating distribution as the training dataset, evaluated against the training 
+    dataset.  We considered a total of four variations of the Dirichlet-process mixture
+    of projected gammas model: whether to restrict $\beta_{j\ell} := 1$ as the projected
+    \emph{restricted} gamma (DP-PRG) akin to a projected Dirichlet, or allow them to be 
+    learned with a gamma prior (DP-PG); and for the shape parameters whether to use a 
+    gamma centering distribution (-G) or a log-normal centering distribution (-LN).
+    For comparison we also included two other distributions on $\mathbb{S}_{1}^{d-1}$:
+    The Dirichlet distribution (with gamma priors), and the pairwise betas distribution
+    developed in~\cite{COOLEY2010}.  Note that the pairwise betas model respects
+    the moment restriction generally assumed.\makenote{This phrasing needs to be changed.
+    The point I am trying to make is that because the moment-restriction assumption is 
+    discarded, a moment-restricted model is inadequate, as is borne out in the
+    simulation study.}  We observe that the projected gamma models dominate, and within
+    that class of model, the choice of centering distribution is irrelevant.  For the
+    projected \emph{restricted} gamma models, the choice of log-normal centering
+    distribution induces
+
+shows two sets of panels that illustrate the comparison
 between four different models used to fit the simulated data. We produced 16 

### 2023-05-27 pointbypoint.tex
updates

@@
         \bruno{Peter, you need to come up with additional, and more specific references here.}
+        \comment{We have \cite{Ferguson74}, \cite{Antoniak1974}, (foundational) \cite{neal2000}, (methodological relative to auxiliary Gibbs) and \cite{escobar1995} (methodological relative to concentration parameter sampling).
+I'm not sure what additional we need?
+I added an additional reference \citep{ascolani2022}, which concerns the robustness of clustering, but I'm not sure of the relevance.}
 
@@
 
+        \comment{\cite{resnick2008extreme} is available on springerlink.  I'm skimming through it now.}
+
         \item Another issue is that it is not clear in the present form of the paper, what is the
@@
         notably the pairwise betas model developed in \cite{sabourin2013}, as well
-        as the dirichlet distribution.  We have investigated several packages to
+        as the dirichlet distribution.  We have investigated several R and python packages to
+        compute empirical angular measure, or other means of estimating the spectral distribution.
+        Most packages implement a bivariate angular density.\comment{include list}.  I found 
+        \cite{BMAmevt} as an example of a true multivariate method, and tried to use it, but
+        encountered a bug with its output (for distributions with more than 3 dimensions, the
+        dimension of the output parameter space became larger than it should be).  Upon 
+        conferring with the package maintainer and author, they identified the source of the
+        bug and fixed it.  For this reason, we are able to include it as a point of comparison.
 
@@
 
-    \comment{\st{Yeah, I think I messed this one up here.  I'm going to run simulations to verify it is
-    a negative definite kernel, then try and figure out a better proof.  Too bad the trivial one
-    didn't work.}}
-
-    \comment{\st{I'm somewhat worried that they will regard the new proof as also problematic, as the second 
-    term is dependent on the first term.  We could get around this by splitting $\sum_{i,j}a_ia_jg(x_i,x_j)$ 
-    into sets belonging to particular face pairs (e.g., $x_i$ on face $a$, $x_j$ on face $b$) and sum over 
-    permutations of $a,b$, but then there's no guarantee that a particular $a_i$ set within a particular 
-    $a,b$ face set will sum to 0.}}
-
     \item \emph{Minor point 1}: About the word \emph{‘Projection’}, e.g. \emph{in 

### 2023-05-27 methodology_projgamma.tex
blah
updates

@@
     &\hspace{-0.5cm}T^{-1}\left(r,y_1,\ldots,y_{d-1}\right)\\
-      & = \left(ry_1,\ldots,ry_{d-1}, r\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}\right).
+      & = \left(ry_1,\ldots,ry_{d-1}, 
+        r\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}\right).
     \end{aligned}
@@
     r^{d-1}\left[\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}} \right. & \\
-        &\hspace{-2cm}+ \left. {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(1 - {\textstyle\sum}_{l=1}^{d-1} y_{\ell}^p\right)^{\frac{1}{p} - 1}\right].
+        &\hspace{-2cm}+ \left. {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p
+          \left(1 - {\textstyle\sum}_{l=1}^{d-1} y_{\ell}^p\right)^{\frac{1}{p} - 1}\right].
     \end{aligned}
@@
 
-  The projected gamma family is simple to specify and has very tractable computational properties.
-Thus, we use it as a building block for the angular measure $\Phi$ models.
-To build a flexible family of distributions in ${\mathbb S}_p^{d-1}$ we consider mixtures of projected gamma densities defined as
+The projected gamma family is simple to specify and has very tractable 
+  computational properties. Thus, we use it as a building block for the 
+  angular measure $\Phi$ models. To build a flexible family of distributions 
+  in ${\mathbb S}_p^{d-1}$ we consider mixtures of projected gamma densities 
+  defined as
   \begin{equation} \label{eqn:PGmix}     
@@
   \end{equation} 
-  where $\bm{\theta} = (\bm{\alpha}, \bm{\beta})$.
-Following a Bayesian non-parametric approach \citep{Ferguson74,Antoniak1974} we assume that $G$ is drawn from a random measure.
-In particular, assuming a Dirichlet process prior for $G$, we have a hierarchical formulation of the mixture model that, for a vector of observations $\bm{y}_i$, is given by
+where $\bm{\theta} = (\bm{\alpha}, \bm{\beta})$. Following a Bayesian 
+  non-parametric approach \citep{Ferguson74,Antoniak1974} we assume that $G$ 
+  is drawn from a random measure. In particular, assuming a Dirichlet process 
+  prior for $G$, we have a hierarchical formulation of the mixture model that, 
+  for a vector of observations $\bm{y}_i$, is given by
   \begin{equation}\label{eqn:dppg}
@@
   \end{equation}
-  where $\mathcal{DP}$ denotes a Dirichlet process, $\eta$ is the precision parameter, and $G_0$ is the centering distribution.
+  where $\mathcal{DP}$ denotes a Dirichlet process, $\eta$ is the precision 
+  parameter, and $G_0$ is the centering distribution.
 
-  Unfortunately, in the limit when $p\rightarrow \infty$, the 
-  projection transformation is not differentiable. Thus, a closed form expression 
+Unfortunately, in the limit when $p\rightarrow \infty$, the projection 
+  transformation is not differentiable. Thus, a closed form expression 
   like Equation (\ref{eqn:projgamma}) for the projected gamma density on ${\mathbb S}_\infty^{d-1}$ is not 
@@
   (v) Project the resulting samples onto ${\mathbb S}_\infty^{d-1}$.
-  For step (iv) we use a Bayesian approach that is implemented using a purposely developed Markov chain Monte Carlo sampler described in the next section.
+  For step (iv) we use a Bayesian approach that is implemented using a purposely 
+  developed Markov chain Monte Carlo sampler described in the next section.
 
@@
 \subsubsection{Inference for the projected Gamma mixture model}
-In a sample-based inference approach, for a given iteration the Dirichlet process mixture model groups observations
-    into stochastically assigned clusters, where members of a cluster share distributional parameters. Building out 
-    the methods of inference for Equation~(\ref{eqn:dppg}), 
+In a sample-based inference approach, for a given iteration the Dirichlet process mixture model 
+    groups observations into stochastically assigned clusters, where members of a cluster share 
+    distributional parameters \citep{ascolani2022}. Building out the methods of inference for Equation~(\ref{eqn:dppg}), 
     let $n_j^{-i}$ be the number of observations in cluster $j$ not including observation $i$.  Let $J^{-i}$ be 
@@
 \end{equation}
-This model is completed with a normal prior on $\bm{\mu}$, an inverse Wishart prior on $\bm{\Sigma}$, and Gamma priors on $\zeta_{\ell}$ and $\sigma_{\ell}$, and $\eta$.
+This model is completed with a normal prior on $\bm{\mu}$, an inverse Wishart prior on
+$\bm{\Sigma}$, and Gamma priors on $\zeta_{\ell}$ and $\sigma_{\ell}$, and $\eta$.  
 This model is denoted as the \emph{projected Gamma--log-normal} (PG--LN) model.  
-We also explore a restricted Gamma form of this model as above, where $\beta_{\ell} = 1$ for all $\ell$.
-This is denoted as the \emph{projected restricted Gamma--log-normal} (PRG--LN) model.
-Updates for $\bm{\alpha}$ can be accomplished using a joint Metropolis step, where $\beta_{j\ell}$ for $\ell = 2,\ldots,d$ have been integrated out of the log-density.
-That is,
+We also explore a restricted Gamma form of this model as above, where 
+$\beta_{\ell} = 1$ for all $\ell$.  This is denoted as the 
+\emph{projected restricted gamma--log-normal} (PRG--LN) model.  Updates for 
+$\bm{\alpha}$ can be accomplished using a joint Metropolis step, where 
+$\beta_{j\ell}$ for $\ell = 2,\ldots,d$ have been integrated out of the 
+log-density.  That is,
 \begin{equation*}
@@
 
+To speed convergence and effectively explore the parameter space, we employ parallel 
+tempering \citep{earl2005} for the log-normal models. This technique runs parallel
+chains at ascending temperatures--that is, for chain $s$, the posterior density is 
+exponentiated to the reciprocal of temperature $t_s$.  For the \emph{cold} chain, 
+$t_1 := 1$.  Let $E_s$ be the log-posterior density under the current parameter 
+state for chain $s$.  Then states between chains $r,s$ are exchanged via a Metropolis 
+step with probability 
+\[
+\alpha_{rs} = \min
+\left[1, \exp\left\lbrace(t_{r}^{-1} - t_{s}^{-1})(E_r - E_s)\right\rbrace\right].
+\]
+Higher temperatures serve to \emph{flatten} the posterior distribution, increasing the 
+speed at which the chain explores parameter space.
+
+Convergence of the MCMC chain is assessed by stabilization of the log-posterior 
+density. For all of our datasets, none of the models required more than 40 thousand 
+iterations to achieve stability, so we use that as a burn-in, and thereafter sample 
+an additional 10 thousand iterations. We then thin the chain by 10 to collect 1000 
+total samples with which to generate posterior predictive densities.
+
 % EOF

### 2023-05-27 notation.tex
updates

\documentclass{article}
\usepackage[margin=2cm]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}

\begin{document}

\begin{description}
\item[$\bm{W},\bm{w}$] - Random, observed variable following multivariate generalized Pareto distribution.
\item[$\bm{Z},\bm{z}$] - Random, observed variable following multivariate standard Pareto distribution.
\item[$\bm{X},\bm{x}$] - Generic random, observed variable in $\mathcal{R}_+^d$.
\item[$\bm{Y},\bm{y}$] - Random, observed variable on $\mathbb{S}_p^{d-1}$.
\item[$\bm{V},\bm{v}$] - Random, observed variable on $\mathbb{S}_{\infty}^{d-1}$; angular component.
\item[$R,r$] - Random, observed variable on $\mathbb{R}_+$; radial component.
\item[$i$] - Iterator over observations.
\item[$j$] - Iterator over mixture components.
\item[$\ell$] - Iterator over dimensions.
\item[$d$] - Number of dimensions.

\end{description}

\end{document}
% EOF

### 2023-05-27 results.tex
updates

@@
 \section{Data illustrations\label{sec:results}} 
+\makenote{this title seems weird?}
 
-\subsection{Simulations\label{subsec:simulated}}
+\subsection{Simulation Study\label{subsec:simulated}}
 To evaluate our proposed approach for angular measure estimation 
@@
     dataset as a mixture of multivariate log-normal distributions,
-    projected onto $\mathbb{S}_p^{d-1}$. The generation procedure is detailed 
-    in Algorithm~\ref{algo:simulated}.
+    projected onto $\mathbb{S}_{\infty}^{d-1}$. The generation 
+    procedure is detailed in Algorithm~\ref{algo:simulated}.  For each
+    model, the data is then projected onto the appropriate $\mathbb{S}_p^{d-1}$
+    for model fitting--dirichlet and pairwise betas being fit on 
+    $\mathbb{S}_1^{d-1}$, and the projected gamma models being fit on 
+    $\mathbb{S}_{10}^{d-1}$.
 
-\begin{algorithm}[ht]  % algorithm using algorithm package
-        \caption{Simulated Angular Dataset Generation Routine\label{algo:simulated}}
+\begin{algorithm}[h]  % algorithm using algorithm package
+        \caption{Simulated Angular Dataset Generation Routine\label{algo:simulated}.
+        $\mu_j$, $\Sigma_j$ are the parameters of the mixture component distribution; 
+        $\pi$ is the probability vector assigning weight mixture components; $\delta_i$ 
+        is the mixture component identifier for each simulated observation.}
         \begin{algorithmic}
@@
             \For{$j$ in $1,\ldots,n_{\text{mix}}$}
-                \State Generate $\mu_{j} \sim \mathcal{N}\left(\bm{0},\bm{I}\right)$
-                \State Generate $\Sigma_{j}\sim\mathcal{IW}\left(70,70 \bm{I}\right)$
+                \State Generate $\mu_{j} \sim \mathcal{N}_{32}\left(\bm{0},\bm{I}\right)$
+                \State Generate $\Sigma_{j}\sim\mathcal{IW}_{32}\left(70,70 \bm{I}\right)$
             \EndFor
@@
 
-where $\mu_j$, $\Sigma_j$ the parameters of the log-normal mixture component 
-    distribution; $\pi$ the probability vector assigning weight to each mixture 
-    component, and $\delta_i$ the mixture component identifier for each simulated 
-    observation.
-
-Figure~\ref{fig:simpples}, shows the average rise over baseline in energy score as
+Figure~\ref{fig:simpples} shows the average rise over baseline in energy score as
     calculated on $\mathbb{S}_{\infty}^{d-1}$ using the kernel metric described
@@
     that class of model, the choice of centering distribution is irrelevant.  For the
-    projected \emph{restricted} gamma models, the choice of log-normal centering
-    distribution induces
-
-shows two sets of panels that illustrate the comparison
-between four different models used to fit the simulated data. We produced 16 
-different data sets, one for each of four different numbers of mixture components 
-and four different dimensions. We considered three different models: DP mixtures 
-of projected gamma; Projected restricted gamma; and projected restricted gamma 
-with a multivariate log-normal prior. We calculated $S^\infty_{\text{PPL}}$ and 
-$S_{\text{ES}}$ using predictive samples from each of fitted models. To provide 
-a comparative baseline, we calculated the scores using samples generated 
-directly using the same model and parameters that produced the simulated 
-observations. The results are denoted as \emph{Gen} in the figure. 
-We observe that the projected
-  restricted Gamma model dominates in most situations, but the difference between projected restricted
-  Gamma and the other gamma-based models tends to shrink as the number of dimensions increases.
-  Alternatively, that difference appears to grow as the number of mixture components increases. 
-  The results from both scoring criteria are comparable.
+    projected \emph{restricted} gamma models, the choice of a multivariate log-normal 
+    centering distribution negatively affects model fidelity.
 
@@
     that tracks the flow of the total water volume in a column of air 
-over a given area \citep{ralph2017}.  IVT is increasingly used in the study of atmospheric rivers because of its direct 
-relationship with orographically induced precipitation
-\citep{neiman2009water}. Atmospheric rivers (AR) are elongated areas of high 
+    over a given area \citep{ralph2017}.  IVT is increasingly used in 
+    the study of atmospheric rivers because of its direct relationship 
+    with orographically induced precipitation \citep{neiman2009water}. 
+    Atmospheric rivers (AR) are elongated areas of high 
     local concentration of water vapor in the atmosphere that transport water
@@
     along the coast of California.  The ERA-Interim data used covers the time period 
-1979 through 2014 (37 years) omitting leap days, and eight grid cells that correspond to the coast of California.  The ERA5 data
-covers the time period 1979 through 2019 (42 years) with the same restriction, and 
-47 grid cells for the coast of California.  This gives us the opportunity to illustrate the performance of our method 
-in multivariate settings of very different dimensions. Figure~\ref{fig:gridlocs} provides a visual representation of the area these grid cells represent.
+    1979 through 2014 (37 years) omitting leap days, and eight grid cells that 
+    correspond to the coast of California.  The ERA5 data covers the time period 
+    1979 through 2019 (42 years) with the same restriction, and  47 grid cells for 
+    the coast of California.  This gives us the opportunity to illustrate the 
+    performance of our method in multivariate settings of very different dimensions. 
+    Figure~\ref{fig:gridlocs} provides a visual representation of the area these grid 
+    cells represent.
 
@@
   \centering
-  \caption{Model comparison metrics: Posterior~Predictive~Loss~($S_{\text{PPL}}$) and 
-    Energy~Score~($S_{\text{ES}}$) criteria from fitted models against the IVT data.  
-    Lower is better. \label{tab:dev}}
-  \include{./tables/table_dev}
+  \caption{Energy score criterion from fitted models against the IVT data. 
+    Lower is better.\label{tab:dev}}
+  \include{./tables/table_dev_updated}
 \end{table}
 
-Table~\ref{tab:dev} shows the values of the estimated posterior predictive loss and energy scores for the different models considered.
-We observe that, as in the case of the simulation study shown in Figures~\ref{fig:simpples}, the preferred model is the projected restricted
-  gamma models.  Penalization for the unrestricted gamma model is 
-  stronger for the real data than for the simulated data. We also observe that
-  the restricted gamma model with the log-normal prior performs significantly better 
-  than the restricted gamma model with the gamma prior on the higher dimensional data. This is the opposite of the result obtained in the simulation example.
+Table~\ref{tab:dev} shows the values of the estimated posterior predictive loss and energy 
+    scores for the different models considered. We observe that, as in the case of the simulation 
+    study shown in Figures~\ref{fig:simpples}, the preferred model is the projected restricted 
+    gamma models, though for the lower-dimensional ERA-Interim data, all models perform comparably.
+    Interestingly, this preference for the restricted gamma form is the opposite of the result 
+    obtained in the simulation example.
 
@@
     \centering
-    \caption{Pairwise conditional survival curves for selected locations, using ERA-Interim, and PRG-G model, conditioning on all other dimensions at greater than 90th percentile (fitted).\label{fig:condsurv2d}}
+    \caption{Pairwise conditional survival curves for selected locations, using 
+        ERA-Interim, and PRG-G model, conditioning on all other dimensions at greater 
+        than 90th percentile (fitted).\label{fig:condsurv2d}}
     \includegraphics[height=4in, width=4in]{./images/condsurv_2d}

### 2023-06-14 pointbypoint.tex
Updates from Overleaf

@@
 \newcommand{\bruno}[1]{\textcolor{red}{#1}}
+
+% TODO
+%
+% in pointbypoint, add description of why parallel tempering for lognormal and why not in gamma.
+% additional description of prior elicitation
+% future work in conclusion.
+% Responses for already answered comments referencing previous comments.
+%
 
@@
 
-We have clarified the distinction between our approach and the prior contributions
-    mentioned.  Setting aside \cite{guillotte2011} which is a strictly bivariate
-    model, the named approaches consider batch extrema and by construction must
-    respect the moment constraint.  As noted, our approach is able to set that 
-    aside, allowing consideration of a wider class of distributions of pseudo-angles.
+\comment{Please review this response.  I've done a little googling, but I can't figure
+    out how one would sample from the Bernstein polynomial model.  Maybe after a lot of
+    integration you could sort out some inverse CDF transform?}
+
+We thank the reviewer for inviting us to expound on differences between our approach
+    and prior contributions. We have clarified the distinctions between our approach 
+    and the prior contributions
+    mentioned.  \cite{guillotte2011} is a strictly bivariate model which assumes
+    unit Fr\'{e}chet margins.  \cite{boldi2007} similarly assumes unit Fr\'{e}chet
+    margins, but builds the spectral density representation on the unit simplex with
+    a finite mixture of Dirichlets.  \cite{SaNa2014} assumes the same, but implements
+    a reversible jump MCMC to allow varying the number of mixture components.
+    \cite{hanson2017} similarly assumes unit Fr\'{e}chet margins, and to represent
+    the spectral density implements a mixture of Bernstein polynomial densities.
+    The distinction as compared to these methods is three-fold: the projected gamma does
+    offer computational advantages as inference is straightforward, and perhaps as 
+    importantly it is simple to sample from.  As a Dirichlet process model, inference
+    on component weights is also straightforward.  It is highly flexible, with no
+    imposed restriction on the number of dimensions.  Finally, as stated in the previous
+    comment, the moment restriction used in those papers is incompatible with our model.
+    This allows us to consider a wider class of distributions on pseudo-angles.
+
+    % Setting aside \cite{guillotte2011} which is a strictly bivariate
+    % model, the named approaches consider batch extrema and by construction must
+    % respect the moment constraint.  As noted, our approach is able to set that 
+    % aside, allowing consideration of a wider class of distributions of pseudo-angles.
+    % \bruno{\bf Where is this in the paper?}
 
@@
 
-    % \comment{\cite{Hanson2017}: Bernstein polynomials on $\mathbb{S}_1^{d-1}$, 
-    % modelling multivariate Frechets.  \cite{boldi2007}: Marginal Frechets, also 
-    % operating on $\mathbb{S}_1^{d-1}$.  \cite{guillotte2011}: A strictly bivariate model.  
-    % \cite{SaNa2014}: }
-
 \item \emph{Numerical accuracy}:  Keeping in mind the computational scope of 
@@
     parameters.  We have clarified as such.
-
-    % \comment{Regarding the uncertainty from step 1, I think we agreed that was 
-    % basically intractible. We may need to explicitly specify as such, and that a 
-    % practicioner must accept the numerical uncertainty that arises from step 1.  
-    % That said, I seem to recall the numbers observed being relatively robust to 
-    % choice of threshold (for large threshold), the difference mainly coming down 
-    % to computation time.}
-
-    % \bruno{I think we need to conduct a more comprehensive suite of experiments. 
-    % Regarding the uncertainty coming from Step1, we just need to acknowledge that 
-    % we are ignoring it.}
+    \comment{I'm not sure how to address this.  I added a comment in the paper
+    that explicitly stated that uncertainty in marginal parameters is outside the
+    scope of the paper.  As a practical concern, every paper I have seen assumed
+    pre-standardized marginals, which implies estimating marginal parameters and
+    ignoring the uncertainty stemming from that estimation.}
 
@@
 
-    \comment{The form of the prior is specified.  I'm guessing he's referring to 
-    the actual values used for computation?  $\bm{\mu}_0 = \bm{0}$, 
-    $\nu = d + 50$, $\bm{\Psi} = \nu I_d$, and so on?}
+    \comment{We thank you for drawing our attention to this omission.  We have included the
+    details of the prior parameters at the start of section 5.}
 
-    \bruno{You need to provide more details here.}
+    % \comment{The form of the prior is specified.  I'm guessing he's referring to 
+    % the actual values used for computation?  $\bm{\mu}_0 = \bm{0}$, 
+    % $\nu = d + 50$, $\bm{\Psi} = \nu I_d$, and so on?}
+
+    % \bruno{You need to provide more details here.}
 
@@
 
-        \bruno{Peter, you need to come up with additional, and more specific references here.}
-        \comment{We have \cite{Ferguson74}, \cite{Antoniak1974}, (foundational) \cite{neal2000}, (methodological relative to auxiliary Gibbs) and \cite{escobar1995} (methodological relative to concentration parameter sampling).
-I'm not sure what additional we need?
-I added an additional reference \citep{ascolani2022}, which concerns the robustness of clustering, but I'm not sure of the relevance.}
+        To ameliorate this shortfall, in addition to the foundational papers for the DP
+        already included, we add \cite{muller2015} to provide a modern approachable reference on
+        and associated clustering, as well as \cite{ascolani2022} which provides a discussion of
+        robustness of clusters generated under the DP.
+
+        % \bruno{Peter, you need to come up with additional, and more specific references here.}
+        % \comment{We have \cite{Ferguson74}, \cite{Antoniak1974}, (foundational) \cite{neal2000},
+        % (methodological relative to auxiliary Gibbs) and \cite{escobar1995} (methodological 
+        % relative to concentration parameter sampling).  I'm not sure what additional we need? 
+        % I added an additional reference \citep{ascolani2022}, which concerns the robustness of
+        % clustering, but I'm not sure of the relevance.}
 
@@
         notably the pairwise betas model developed in \cite{sabourin2013}, as well
-        as the dirichlet distribution.  We have investigated several R and python packages to
+        as the Dirichlet distribution.  We have investigated several R and python packages to
         compute empirical angular measure, or other means of estimating the spectral distribution.
-        Most packages implement a bivariate angular density.\comment{include list}.  I found 
-        \cite{BMAmevt} as an example of a true multivariate method, and tried to use it, but
-        encountered a bug with its output (for distributions with more than 3 dimensions, the
-        dimension of the output parameter space became larger than it should be).  Upon 
+        To be clear, our analysis requires the ability to generate a sample from a posterior
+        predictive distribution, or at least sample from a fitted density, projected onto 
+        $\mathbb{S}_{\infty}^{d-1}$. Most packages implement a bivariate angular density, 
+        including \verb|ExtremalDep|~\citep{ExtremalDep}, \verb|BAMBI|~\citep{BAMBI}, 
+        \comment{expand list}.  Some, such as
+        \verb|evd|~\citep{evd} implement a multivariate generalized Pareto density with a 
+        parametric dependence structure, but I didn't see a way to learn the dependence
+        structure in isolation, as well as sample from it.  I found 
+        \cite{BMAmevt} as an example of a true multivariate density on $\mathbb{S}_1^{d-1}$, 
+        and on first attempt to use
+        it I encountered a bug with its output (for distributions with more than 3 dimensions, 
+        the dimension of the output parameter space became larger than it should be).  Upon 
         conferring with the package maintainer and author, they identified the source of the
@@
 
-            \bruno{This one was pointed out by the other referee. So, we need to come up with something.}
+            We thank the reviewers for bringing this omission to our attention.  We have now
+            included a discussion of MCMC convergence among the models in section 5.
 
-            \comment{\st{I'm not sure how to demonstrate convergence in the case of the DP.  I can
-            I can demonstrate convergence for certain quantities (e.g., the hierarchical mean,
-            and log-determinant of hierarchical covariance matrix, a given observation's distribution
-            of parameters, etc.), but in the general stick-breaking configuration, there is no concern
-            given to label switching.  The variational approach will probably find a label arrangement
-            and stick with it.}}
+            % \bruno{This one was pointed out by the other referee. So, we need to come up with something.}
 
-            \comment{I'm working on adding log-density over time plots to demonstrate convergence.}
+            % \comment{\st{I'm not sure how to demonstrate convergence in the case of the DP.  I can
+            % I can demonstrate convergence for certain quantities (e.g., the hierarchical mean,
+            % and log-determinant of hierarchical covariance matrix, a given observation's distribution
+            % of parameters, etc.), but in the general stick-breaking configuration, there is no concern
+            % given to label switching.  The variational approach will probably find a label arrangement
+            % and stick with it.}}
+
+            % \comment{I'm working on adding log-density over time plots to demonstrate convergence.}
 
@@
 
-            \bruno{We need to think about this one.}
-
-            \comment{I think it was an issue of understanding.  I've updated the code, remaking the plots
-            is nearly done.}
+            We have converted the graphs to use average rise in score over baseline (as calculated
+            by average of model on 10 distributions), rather than the score directly on a single
+            sample. We have verified that instances of displayed values dropping below zero are 
+            within MCMC error of zero.\comment{I need to say this better.}
 
@@
     but this needs to be clarified.
-
-    \comment{I seem to regard something in the intro to extremes book regarding the lack of a
-    moment constraint on the $\mathcal{L}_{\infty}$ norm.  I'll look for it.}
 

### 2023-06-14 conclusion.tex
pgpareto github
statistics and computing submission

@@
 \section{Conclusion\label{sec:conclusion}}
-In this paper, we have built upon the definition of the multivariate Pareto described in \cite{ferreira2014}
-  to establish a useful representation of its dependence structure, through
-  the distribution of its angular component, which is supported on the positive orthant of the unit
-  hypersphere under the $\mathcal{L}_{\infty}$ norm.  Due to the 
-  inherent difficulty of obtaining a distribution with support on ${\mathbb S}^{d-1}_\infty$
-  our method projects distributions, supported on the positive quadrant and
+In this paper, we have built upon the definition of the multivariate Pareto 
+    described in \cite{ferreira2014} to establish a useful representation of 
+    its dependence structure, through the distribution of its angular component,
+    which is supported on the positive orthant of the unit hypersphere under 
+    the $\mathcal{L}_{\infty}$ norm.  Due to the inherent difficulty of 
+    obtaining a distribution with support on ${\mathbb S}^{d-1}_\infty$ our 
+    method projects distributions, supported on the positive quadrant and 
     based on products of independent gammas,  onto the manifold 
     ${\mathbb S}_{p}^{d-1}$. Samples of the resulting probability distribution 
-  are then projected onto ${\mathbb S}_{\infty}^{d-1}$.  
-  As ${\mathbb S}_{p}^{d-1}$ converges to ${\mathbb S}_{\infty}^{d-1}$ as $
-  p\to\infty$, we expect the resampling to be efficient for large enough
+    are then projected onto ${\mathbb S}_{\infty}^{d-1}$.  As 
+    ${\mathbb S}_{p}^{d-1}$ converges to ${\mathbb S}_{\infty}^{d-1}$ as 
+    $p\to\infty$, we expect the resampling to be efficient for large enough
     $p$. In fact, our exploration of the simulated and real data indicate
@@
 We explored the behavior of the proposed model using simulated data from
-    mixtures of projected Gammas with varying degrees of complexity---both in terms of dimensionality and 
-    in number of mixture components.  From this, we learned that the additional flexibility 
-    offered by varying the $\beta$ term---the rate parameter of the Gamma distribution---does not result 
-    in increased model fidelity in a mixture setting.  Within the tested range, between 3 and 20 
-    dimensions, we observed that the additional information provided by a log--normal 
-    centering distribution did not translate to additional model fidelity.  However, for 
-    real data, with 47 dimensions, we did observe an effect.
+    mixtures of projected Gammas with varying degrees of complexity---both 
+    in terms of dimensionality and in number of mixture components.  From this,
+    we observed that the additional flexibility offered by varying the $\beta$
+    term---the rate parameter of the gamma distribution---can offer additional
+    model fidelity.  This result is not borne out in our real data example,
+    for which the simpler model is vastly preferred.  Additionally, we observe
+    that the choice of the log-normal centering distribution and additional
+    flexibility it offers did not translate to additional model fidelity.  This
+    is borne out in the real data, where on the ERA--Interim dataset it performed
+    only comparably, but on ERA--5 it performed significantly worse.
 
-The computations in this paper were performed on a desktop computer with an AMD Ryzen 5950X processor.
-  The program is largely single-threaded, so computation time is not dependent on available core count.  In each
-  case, we run the MCMC chain for 50\,000 iterations, with a burn-in of 20\,000 samples.  Fitting the PG--G model
-  on the ERA5 dataset took approximately 30 minutes.  Work is in progress to optimize the code, and explore
-  parallelization where possible.  We are also exploring alternative computational approaches that will make it
-  feasible to tackle very high dimensional problems. In fact, to elaborate on the study of IVT, there is a need 
-  to consider several hundreds, if not thousands, of grid cells over the Pacific Ocean in order to obtain a 
-  good description of atmospheric events responsible for large storm activity over California.
+    % From this,
+    % we learned that the additional flexibility offered by varying the $\beta$ 
+    % term---the rate parameter of the Gamma distribution---does not result in 
+    % increased model fidelity in a mixture setting.  Within the tested range, 
+    % between 3 and 20 dimensions, we observed that the additional information 
+    % provided by a log--normal centering distribution did not translate to 
+    % additional model fidelity.  However, for real data, with 47 dimensions, 
+    % we did observe an effect.
 
-% {\bf ( describe
-% the computer)} and in each case we run the MCMC for {\bf XXX} iterations
-% with a burn in of {\bf XXX} samples. For the large simulated data and the
-% ERA5 dataset the process of fitting the model took {\bd YYY} minutes. Work is in progress to optimize the
-% code to perform the MCMC {\bf (insert some comment here if you think it is appropriate)}. We are also 
-% exploring alternative computational approaches that will make it feasible to tackle very high dimensional
-% problems. In fact, to elaborate on the study of IVT, there is a need to consider several hundreds, if not
-% thousands, of grid cells over the Pacific Ocean in order to obtain a good decsription of atmospheric events responsible for large storm activity over California.
+The computations in this paper were performed on a desktop computer with an 
+    AMD Ryzen 5950X processor. The program is largely single-threaded, so 
+    computation time is not dependent on available core count.  In each case, 
+    we run the MCMC chain for 50\,000 iterations, with a burn-in of 40\,000 
+    samples.  Fitting the PG--G model on the ERA5 dataset took approximately 
+    15 minutes.  Work is in progress to optimize the code, and explore 
+    parallelization where possible.  We are also exploring alternative 
+    computational approaches that will make it feasible to tackle very high 
+    dimensional problems. In fact, to elaborate on the study of IVT, there is 
+    a need to consider several hundreds, if not thousands, of grid cells over
+    the Pacific Ocean in order to obtain a good description of atmospheric 
+    events responsible for large storm activity over California.  Moving beyond
+    this model
 
-%We used integrated vapor transport data---a measure of water in the atmosphere---from the ERA-Interim and ERA5
-%    models as a motivating example for our model.  This data subsets California into a series of grid cells,
-%    and provides daily IVT values for each cell.  For the ERA-Interim data, having determined from
-%    Table~\ref{tab:dev} that the optimal model presented was PRG--G, we used this fitted model to develop
-%    $\Chi$, a measure of pairwise extremal dependence between grid cells.  This provides a useful summary
-%    statistic for describing the relationship between two sites.  Granted this relationship is pairwise, it
-%    can not represent the full dependence structure; so recognizing these limits, we also presented 
-%    conditional survival 
-%    functions---$\text{P}\left[Z_{\ell} > z_{\ell}\mid Z_{\neg\ell} > z_{\neg\ell}\right]$---which
-%    provide a more complete understanding of the dependence structure of $\bm{Z}$.
+% §6 calls for a view of the authors on what 
+%     could be related exercises—or even extensions of this research and future 
+%     work. Using the proposed Bayesian multivariate PoT construction for 
+%     devising a regression model for an an extreme value response on an extreme 
+%     value covariate is one example of a natural follow-up; see 
+%     \cite{carvalho2022} for related ideas. Conditional versions of the angular 
+%     measure (e.g. \cite{carvalho2016}, \cite{castro2018}, \cite{escobar2018}, 
+%     \cite{mhalla2019}) would lead to a conditional multivariate 
+%     peaks-over-threshold model, where dependence between exceedances could 
+%     change along with a covariate. Also, perhaps the here proposed DPM of 
+%     projected gammas—as well as predictor-dependent versions of it—could be of 
+%     interest in other settings beyond angular measures?
+
+% EOF

### 2023-06-14 evaluation.tex
OL 2023_10_15

@@
     therein.
-    % the We consider two scoring methods:  \emph{posterior predictive loss} (PPL) introduced in 
-    % \cite{gelfand1998}, and \emph{energy score} (ES) introduced in \cite{gneiting2007}.
 
-% \subsection{Posterior predictive loss}
-% The posterior predictive loss criterion is introduced in \cite{gelfand1998}.  When we
-%   assume a squared error loss function, then for the $i$th observation, the posterior predictive
-%   loss criterion is computed as
-%   \begin{equation}
-%     \label{eq:ppl}
-%     \begin{aligned}
-%       &\hspace{-0.5cm}S_{\text{PPL}}^k\left(P,x_i\right) \\
-%       &= \text{Var}_P\left[X_i\right] + \frac{k}{k+1}\left(\text{E}_P\left[X_{i}\right] - x_i\right)^2
-%     \end{aligned}
-%   \end{equation}
-%   where $X_i$ is a random variable from the posterior predictive distribution for $x_i$.  The
-%   scalar $k$ is a weighting factor that scales the importance of goodness of fit
-%   relative to precision.  In fact a small $\text{Var}(X_{i})$ indicates high
-%   precision, and a small $(\text{E}[X_{i}] - x_{i})^2$  indicates a good fit.  In our analysis,
-%   we take the limit as $k\to\infty$, weighting both
-%   equally. Overall, the smaller the score, the better.  Note that the PPLC score is defined 
-%   for a univariate distribution.  As we are dealing
-%   with a multivariate distribution, we will be taking the posterior predictive loss
-%   summed over all dimensions.  That is,
-%   \begin{equation}
-%     \label{eq:ppl2}
-%     \begin{aligned}
-%       &\hspace{-0.5cm}S_{\text{PPL}}\left(P,\bm{x}_i\right) \\
-%       &= \frac{1}{d}\sum_{\ell = 1}^d\left[\text{Var}_P\left[X_{i\ell}\right] + \left(\text{E}_P\left[X_{i\ell}\right] - x_{i\ell}\right)^2\right].
-%     \end{aligned}
-%   \end{equation}
-%   We report the average $S_{\text{PPL}}$ taken from all observed data.
-
-% \subsection{Energy score}
-% We focus on \emph{energy scores} defined for a general probability distribution $P$, with finite expectation, as 
-The energy score criterion defined for a general probability distribution $P$, with finite expectation, is developed
-    as
+The energy score criterion defined for a general probability distribution $P$, with finite expectation, 
+    is developed as
     \begin{equation}
@@
 In a Euclidean space, these conditions are satisfied by the Euclidean distance \citep{berg1984}. 
-    However, for observations on different faces of ${\mathbb S}_{\infty}^{d-1}$, Euclidean 
+    However, for observations on different faces of ${\mathbb S}_{\infty}^{d-1}$, the Euclidean 
     distance will under-estimate the geodesic distance, the actual distance required to travel 
@@
 \end{equation*}
-comprise the $\ell$th \emph{face}.  For points on the same face, Euclidean distance 
+    comprise the $\ell$th \emph{face}.  For points on the same face, the Euclidean distance 
     corresponds to the length of the shortest possible path in ${\mathbb S}_{\infty}^{d-1}$.  
@@
     $\bm{b} \in {\mathbb   C}_{\jmath}^{d-1}$, for $\ell,\jmath\in \{1,\ldots, d\}$.
-%\end{prop}
-%{\em Proof:}
-%When $\bm{a}$ and $\bm{b}$ are on the same face, then $\ell = \jmath$, and $g$ is the Euclidean distance between $\bm{a}$ and $\bm{b}$, which is a  negative definite kernel.
-When $\bm{a}$ and $\bm{b}$ are on separate
-  %faces, let 
-  %\[\bm{e} = 
-    %\argmin_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
-        %\pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
-  %\]
-  %We have that $g(\bm{a}, \bm{b}) = g(\bm{b}, \bm{a})$ due to the uniqueness of $\bm{e}$ and the symmetry of the Euclidean distance.
-Additionally, given $n$ and  a collection $\alpha_1,\ldots,\alpha_n\in{\mathbb R}$ such that  $\sum_{i = 1}^n\alpha_i = 0$, then
-  %\begin{equation*}
-    %\begin{aligned}
-      %&\hspace{-0.5cm}\sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j g(\bm{x}_1,\bm{x}_2)\\
-      %&= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j \bigg[\pnorm{\bm{e} - \bm{x}_1}{2} + \pnorm{\bm{x}_2 - \bm{e}}{2}\bigg]\\
-      %&= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{e} - \bm{x}_1}{2}\\
-      %&\hspace{1cm}+ \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{x}_2 - \bm{e}}{2}\\
-      %&\leq 0
-    %\end{aligned}
-  %\end{equation*}
-  %as the Euclidean norm is a negative definite kernel.$\hfill\Box$
 Evaluating $g$ as described requires the solution of a $(d-2)$-dimensional 
-  optimization problem.  The following proposition provides a straightforward solution.
+  optimization problem.  The following proposition provides a straightforward approach.
 \begin{prop}\label{prop:rot}
@@
     observation. For the purpose of model assessment and comparison, we report the 
-    average $S_{\text{ES}}$ taken across all observed data.  In this form, the 
-    interpretation of $S_{\text{ES}}$ is similar to that of $S_{\text{PPL}}$: The smaller 
+    average $S_{\text{ES}}$ taken across all observed data, and notice that the smaller 
     the score, the better.

### 2023-06-14 methodology_projgamma.tex

@@
 where $\bm{\theta} = (\bm{\alpha}, \bm{\beta})$. Following a Bayesian 
-  non-parametric approach \citep{Ferguson74,Antoniak1974} we assume that $G$ 
+  non-parametric approach \citep{Ferguson74,Antoniak1974,muller2015},
+  % \st{\bruno{\bf Peter:
+  % Please provide a more recent and  approacheable reference, possibly a book. If you can't
+  % find one  ask Thanos or Juhee, our BNP experts.} 
+  we assume that $G$ 
   is drawn from a random measure. In particular, assuming a Dirichlet process 
@@
     &\hspace{-0.5cm}\pi(\alpha_{j\ell}\mid \bm{r},\bm{Y},\bm{\gamma},\xi_\ell,\tau_\ell,\zeta_\ell,\sigma_\ell) \\
-    &\propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i\ell}\right)^{\alpha_{j\ell} - 1}\alpha_{j\ell}^{\xi_\ell - 1}e^{-\tau_\ell \alpha_{j\ell}}\times}{\Gamma^{n_j}(\alpha_{j\ell})} \\
+    &\propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i\ell}\right)^{\alpha_{j\ell} - 1}\alpha_{j\ell}^{\xi_\ell - 1}e^{-\tau_\ell \alpha_{j\ell}}}{\Gamma^{n_j}(\alpha_{j\ell})} \\
     &\hspace{0.5cm}\times \frac{\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
@@
 
-To speed convergence and effectively explore the parameter space, we employ parallel 
-tempering \citep{earl2005} for the log-normal models. This technique runs parallel
+To effectively employ the sample space for a joint Metropolis step, as well as
+    speeding convergence, we employ a parallel tempering algorithm\citep{earl2005} 
+    for the log-normal models. This technique runs parallel MCMC
     chains at ascending temperatures--that is, for chain $s$, the posterior density is 
@@
     \]
-Higher temperatures serve to \emph{flatten} the posterior distribution, increasing the 
-speed at which the chain explores parameter space.
-
-Convergence of the MCMC chain is assessed by stabilization of the log-posterior 
-density. For all of our datasets, none of the models required more than 40 thousand 
-iterations to achieve stability, so we use that as a burn-in, and thereafter sample 
-an additional 10 thousand iterations. We then thin the chain by 10 to collect 1000 
-total samples with which to generate posterior predictive densities.
+    Higher temperatures serve to \emph{flatten} the posterior distribution, meaning hotter
+    chains have a higher probability of making a given transition, or will make larger 
+    transitions.  As such, they will more quickly explore the parameter space, and share
+    information gained through state exchange.
 

### 2023-06-14 results.tex

@@
 \section{Data illustrations\label{sec:results}} 
-\makenote{this title seems weird?}
+We apply the aforementioned models to generated angular data, as well as a data analysis
+    of atmospheric data.
+    Assessing convergence of an MCMC chain for a large-dimensional model is difficult.
+    For our part, we assess convergence by stabilization of the log-posterior 
+    density. For all of our datasets, none of the models required more than 40 thousand 
+    iterations to achieve stability, so we use that as a burn-in, and thereafter sample 
+    an additional 10 thousand iterations. We then thin the chain by 10 to collect 1000 
+    total samples with which to generate posterior predictive densities.
+    Implementation details play a role here: posterior samples from the gamma-centered 
+    models are moderately autocorrelated as compared to the log-normal centered models.
+    This is a benefit of the parallel tempering algorithm employed in fitting the 
+    log-normal models, but parallel tempering necessitates an additional computational cost,
+    growing linearly with the number of temperatures.  We used a temperature ladder
+    of $t_s = 1.3^{s}$, for $s \in \lbrace0,1,\ldots,5$.  This was verified to produce
+    acceptable swap probabilities both for the simulated data, and real data.
+
+Our hyperprior parameters are set as follows: for the gamma-centered models (PG--G, PRG--G), 
+    the shape parameter for the centering 
+    distribution~$\xi_{\ell}\sim \mathcal{G}\left(1,1\right)$, and
+    rate parameter~$\tau_{\ell}\sim\mathcal{G}\left(2,2\right)$.
+    For the log-normal centered models (PG--LN, PRG--LN), the centering distribution's
+    log-mean $\bm{\mu}\sim\mathcal{N}_d\left(0,\bm{I}_d\right)$, and covariance 
+    matrix~$\Sigma\sim\mathcal{IW}\left(d + 10, (d+10)\bm{I}_{d}\right)$.  These values are 
+    intended such that draws from the prior for $\Sigma$ will weakly tend towards the 
+    identity matrix.
+    For models learning rate parameters $\beta_{j\ell}$ (PG--G, PG--LN), the centering
+    distribution's shape parameter $\zeta\sim\mathcal{G}\left(1,1\right)$ and rate parameter
+    $\sigma\sim\mathcal{G}\left(2,2\right)$.  The choice of $\mathcal{G}(2,2)$ for rate
+    parameters was meant to draw estimates for the value away from 0 for numerical stability.
 
@@
     The Dirichlet distribution (with gamma priors), and the pairwise betas distribution
-    developed in~\cite{COOLEY2010}.  Note that the pairwise betas model respects
-    the moment restriction generally assumed.\makenote{This phrasing needs to be changed.
-    The point I am trying to make is that because the moment-restriction assumption is 
-    discarded, a moment-restricted model is inadequate, as is borne out in the
-    simulation study.}  We observe that the projected gamma models dominate, and within
-    that class of model, the choice of centering distribution is irrelevant.  For the
-    projected \emph{restricted} gamma models, the choice of a multivariate log-normal 
-    centering distribution negatively affects model fidelity.
+    developed in~\cite{COOLEY2010}.  
+    % Note that the pairwise betas model respects
+    % the moment restriction generally assumed.\makenote{This phrasing needs to be changed.
+    % The point I am trying to make is that because the moment-restriction assumption is 
+    % discarded, a moment-restricted model is inadequate, as is borne out in the
+    % simulation study.} \bruno{\bf I would avoid getting into the details of what the
+    % pairwise beta model does. More important is to say why we chose it, and how available 
+    % it is} 
+    We observe that for simulated data using Algorithm~\ref{algo:simulated}, the 
+    projected gamma models dominate, and within that class of model, the choice of 
+    centering distribution is irrelevant.  For the projected \emph{restricted} 
+    gamma models, the choice of a multivariate log-normal centering distribution 
+    appears to negatively affect model fidelity.
 
@@
 
+% Table~\ref{tab:dev} shows the values of the estimated posterior predictive loss and energy 
+%     scores for the different models considered. We observe that, as in the case of the simulation 
+%     study shown in Figures~\ref{fig:simpples}, the preferred model is the projected restricted 
+%     gamma models, though for the lower-dimensional ERA-Interim data, all models perform comparably.
+%     Interestingly, this preference for the restricted gamma form is the opposite of the result 
+%     obtained in the simulation example.
 Table~\ref{tab:dev} shows the values of the estimated posterior predictive loss and energy 
-    scores for the different models considered. We observe that, as in the case of the simulation 
-    study shown in Figures~\ref{fig:simpples}, the preferred model is the projected restricted 
+    scores for the different models considered. We observe that, contrary to the results in the
+    simulation study in Figure~\ref{fig:simpples}, the preferred model is the projected restricted 
     gamma models, though for the lower-dimensional ERA-Interim data, all models perform comparably.
@@
     Figure~\ref{fig:condsurv2d} presents the bi-variate conditional survival function,
-    conditioning on all other dimensions.  These results illustrate quantitatively how extremal dependence
-    affects the shape of the conditional survival curves.  The two top panels represent the joint survival function 
-    between grid locations 4 and 5, which are shown in Figure~\ref{fig:chi_ij} to exhibit strong
-    extremal dependence.  We observe that the joint survival surface 
-    is strongly convex.  The bottom panels represent the joint survival surface between grid locations 1 and 5, 
-    which exhibited low extremal dependence.  In this case the shape of the contours tend to be concave, quite different from the shapes observed in the top panels.
+    conditioning on all other dimensions.  These results illustrate quantitatively how 
+    extremal dependence affects the shape of the conditional survival curves.  The two 
+    top panels represent the joint survival function 
+    between grid locations 4 and 5, which are shown in Figure~\ref{fig:chi_ij} to exhibit 
+    strong  extremal dependence.  We observe that the joint survival surface 
+    is strongly convex.  The bottom panels represent the joint survival surface between 
+    grid locations 1 and 5, which exhibited low extremal dependence.  In this case the 
+    shape of the contours tend to be concave, quite different from the shapes observed 
+    in the top panels.
 
-Using our proposed scoring criteria, we explored the effect of the choice of $p$ on the final results. Using the simulated data, generated
-    from a mixture of projected Gammas, we were unable to observe sizeable differences in the scores for $p$ 
-    ranging between 1 and 15.   However, 
-    for the IVT data, we observed a drop in the energy score associated 
-    with higher $p$, with diminishing effect as $p$ increased.  
-    We observed no significant differences in the performance
-    of the model that uses $p=10$, which corresponds to the analysis 
-    presented, relative to the one that uses $p=15$.
+Using our proposed scoring criteria, we explored the effect of the choice of $p$ on the 
+    final results. Using the simulated data, generated from a mixture of projected Gammas, 
+    we were unable to observe sizeable differences in the scores for $p$ ranging between 
+    1 and 15.   However, for the IVT data, we observed a drop in the energy score associated 
+    with higher $p$, with diminishing effect as $p$ increased.  We observed no significant 
+    differences in the performance of the model that uses $p=10$, which corresponds to the 
+    analysis presented, relative to the one that uses $p=15$.
 % EOF

### 2023-06-15 pointbypoint.tex
update

@@
 
+            As we considered parallel tempering to be necessary in fitting the multivariate-lognormal
+            model, as we're using a $d$-dimensional joint proposal step, it is readily apparent that 
+            convergence will appear \emph{cleaner} on the log-density plots for the lognormal models 
+            relative to the gamma models, as parallel tempering also ameliorates autocorrelation
+            issues.  Below are the log-density by iteration plots for the ERA-5 data, on the two 
+            unrestricted projected gamma models; left uses the product of gammas centering distribution, 
+            right uses the log-normal centering distribution.  The red line indicates the end of the 
+            burn-in period, and we only keep every tenth sample after.
+
+            \includegraphics[width=0.45\textwidth]{images/ERA-5-PG-G.pdf}
+            \includegraphics[width=0.45\textwidth]{images/ERA-5-PG-LN.pdf}
+
+            We could say that parallel tempering would improve autocorrelation for the gamma model,
+            but sampling longer and thinning more would serve the same effect.  And since it also
+            reaches convergence faster, we could also simply use less burn-in.
+
             % \bruno{This one was pointed out by the other referee. So, we need to come up with something.}

### 2023-06-15 abstract.tex
statistics and computing submission
update

@@
-% We consider a constructive definition of the multivariate Pareto that factorizes the 
-% random vector into a radial component and an independent angular component; The former
-% following a univariate Pareto distribution, and the latter defined on the surface of the 
-% positive orthant of the unit hypercube.  In this paper, we  propose a method for
-% inferring the distribution of the angular component.  We identify its support 
-% as the limit of the positive orthants of the unit $p$--norm spheres,
-% and introduce a projected gamma family of
-% distributions defined as the projection of a vector of independent gamma random variables
-% onto the $p$--norm sphere.  This family serves as a building block for a flexible family
-% of distributions obtained as a Dirichlet process mixture of projected gammas.  For 
-% model assessment and comparison, we discuss model scoring methods appropriate 
-% to distributions on the unit hypercube.  In particular, working with the energy score
-% criterion, we develop a kernel metric appropriate to the hypercube that produces a 
-% proper scoring rule.  We then present a simulation study to compare 
-% different modeling choices using the proposed scoring rules.  Finally we apply our 
-% approach to describe the dependence structure of the extreme values of the magnitude 
-% of the integrated vapor  transport (IVT), a variable that describes the rate of flow 
-% of moisture in the atmosphere  along the coast of California for the years of 1979 
-% through 2020.  We find a clear but heterogeneous geographical dependence.
-
 We consider a constructive definition of the multivariate Pareto that factorizes the 
@@
 following a univariate Pareto distribution, and the latter defined on the surface of the 
-positive orthant of the unit hypercube.  In this paper, we  propose a method for
-inferring the distribution of the angular component.  We identify its support 
-as the limit of the positive orthant of the unit $p$--norm spheres,
-and introduce a projected gamma family of
-distributions defined as the projection of a vector of independent gamma random variables
-onto the $p$--norm sphere.  This family serves as a building block for a flexible family
-of distributions obtained as a Dirichlet process mixture of projected gammas.  For 
-model assessment, we discuss scoring methods appropriate 
+positive orthant of the infinity norm unit hypercube.  In this paper, we  propose a method 
+for inferring the distribution of the angular component.  We identify its support 
+as the limit of the positive orthant of the unit $p$--norm spheres, and introduce a 
+projected gamma family of distributions defined through the normalization a vector of 
+independent gamma random variables to the $p$--norm sphere.  This family serves as a 
+building block for a flexible family of distributions obtained as a Dirichlet process 
+mixture of projected gammas.  For model assessment, we discuss scoring methods appropriate 
 to distributions on the unit hypercube.  In particular, working with the energy score
-criterion, we develop a kernel metric that produces a 
-proper scoring rule, and present a simulation study to compare 
-different modeling choices using the proposed scoring rules.  Finally we apply our 
-approach to describe the dependence structure of the extreme values of the magnitude 
-of the integrated vapor transport (IVT), data describing the flow 
+criterion, we develop a kernel metric that produces a  proper scoring rule, and present 
+a simulation study to compare  different modeling choices using the proposed scoring rules.  
+Finally we apply our  approach to describe the dependence structure of the extreme values of 
+the magnitude  of the integrated vapor transport (IVT), data describing the flow 
 of atmospheric moisture along the coast of California for the years of 1979 

### 2023-06-15 evaluation.tex
update

@@
   computationally burdensome for a large number of dimensions.  
-  \makenote{However, we can efficiently establish an upper bound on the geodesic 
+  However, we can efficiently establish an upper bound on the geodesic 
   length.  We use this upper bound on geodesic distance as the kernel function for
-  the energy score.}
+  the energy score.
 
@@
   $\bm{c} \in {\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}$,
-     $\pnorm{\bm{b} - \bm{c}}{2} =  \pnorm{\bm{b}^{\prime} - \bm{c}}{2}$ and let
-\[\bm{e} = 
-    \argmin_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
-        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
-\]
+     $\pnorm{\bm{b} - \bm{c}}{2} =  \pnorm{\bm{b}^{\prime} - \bm{c}}{2}$. %and let
+%\[\bm{e} = 
+%    \argmin_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
+%        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
+%\]
 We then have that
@@
     \begin{aligned}
-    g(\bm{a}, \bm{b}) &= \pnorm{\bm{a} - \bm{e}}{2} + \pnorm{\bm{b} - \bm{e}}{2}\\
-    &= \pnorm{\bm{a} - \bm{e}}{2} + \pnorm{\bm{b}^{\prime} 
-        - \bm{e}}{2} = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}.
+    g(\bm{a},\bm{b}) = &\min_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
+        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\} \\
+     &\min_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
+        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}^{\prime}-\bm{c}}{2} \right\}\\
+    &= \pnorm{\bm{a} - \bm{b}^{\prime}}{2}.
     \end{aligned}
@@
     \]
-    where $\bm{a}_\jmath^\prime$ is defined as in Proposition \ref{prop:rot}. $\hfill\Box$
+    where $\bm{a}_\jmath^\prime$ is defined as in Proposition \ref{prop:rot}. 
+    The last equality holds as $\|\bm{x} - \bm{x}^\prime\|_2, \bm{x},
+    \bm{x}^{\prime} \in {\mathbb R}^d$ is negative definite
+    \citep{gneiting2007}$\hfill\Box$
 

### 2023-06-15 introduction.tex
update

@@
     In the multivariate case the GEV theory is well developed 
-    \citep[see, for example][]{dehaan2006}, but the inferential problem is more 
-    complicated, as there is no parametric representation of the GEV. This problem
-    is inherited by the peaks over threshold (PoT) approach and compounded by the 
-    fact that there is no unique definition of an exceedance of a multivariate 
-    threshold, as there is an obvious dependence on the norm that is used to measure 
-    the size of a vector.
+    \citep[see, for example][]{dehaan2006}, but the inferential problem is  
+    complicated by the fact that there is no parametric representation of the GEV. 
+    This problem is inherited by the peaks over threshold (PoT) approach and 
+    compounded by the  fact that there is no unique definition of an exceedance of 
+    a multivariate  threshold, as there is an obvious dependence on the norm that 
+    is used to measure  the size of a vector.
 
@@
     called moment constraint by considering a constructive definition of the 
-    multivariate GP, based on the infinity norm.  Further, the
+    multivariate GP, based on the infinity norm.
 
@@
     \end{equation*}
-    $H$ a multivariate Pareto distribution, and corresponds to the joint
+    $H$ is a multivariate Pareto distribution, and corresponds to a joint
     distribution conditional on exceeding a multivariate threshold. In addition to

### 2023-06-15 methodology_projgamma.tex
update

@@
 
-Unfortunately, in the limit when $p\rightarrow \infty$, the projection 
+Unfortunately, in the limit when $p\rightarrow \infty$, the normalizing 
   transformation is not differentiable. Thus, a closed form expression 
@@
 
-To effectively employ the sample space for a joint Metropolis step, as well as
-    speeding convergence, we employ a parallel tempering algorithm\citep{earl2005} 
+To effectively explore the sample space with a joint Metropolis step, as well as
+    to speed convergence, we implement a parallel tempering algorithm\citep{earl2005} 
     for the log-normal models. This technique runs parallel MCMC
-    chains at ascending temperatures--that is, for chain $s$, the posterior density is 
+    chains at ascending temperatures. That is, for chain $s$, the posterior density is 
     exponentiated to the reciprocal of temperature $t_s$.  For the \emph{cold} chain, 

### 2023-06-15 results.tex
update

@@
 \section{Data illustrations\label{sec:results}} 
-We apply the aforementioned models to generated angular data, as well as a data analysis
-    of atmospheric data.
-    Assessing convergence of an MCMC chain for a large-dimensional model is difficult.
-    For our part, we assess convergence by stabilization of the log-posterior 
-    density. For all of our datasets, none of the models required more than 40 thousand 
-    iterations to achieve stability, so we use that as a burn-in, and thereafter sample 
-    an additional 10 thousand iterations. We then thin the chain by 10 to collect 1000 
-    total samples with which to generate posterior predictive densities.
-    Implementation details play a role here: posterior samples from the gamma-centered 
-    models are moderately autocorrelated as compared to the log-normal centered models.
-    This is a benefit of the parallel tempering algorithm employed in fitting the 
-    log-normal models, but parallel tempering necessitates an additional computational cost,
-    growing linearly with the number of temperatures.  We used a temperature ladder
-    of $t_s = 1.3^{s}$, for $s \in \lbrace0,1,\ldots,5$.  This was verified to produce
-    acceptable swap probabilities both for the simulated data, and real data.
+We apply the aforementioned models to simulated angular data. We then consider the analysis
+    of atmospheric data. To tackle the difficult problem of assessing the convergence 
+    an MCMC chain for a large-dimensional model, we monitor the log-posterior 
+    density. In all the examples considered MCMC samples produced stable tarces of the
+    log-posterior in less than 40 thousand  iterations. Thus, so we use that as a burn-in, 
+    and thereafter sample 10 thousand additional iterations. We then thin the chain by retaining one
+    every ten samples, to obtain 1000 total samples. These are used to generate samples from the 
+    posterior predictive densities. We used two different strategies implemment the MCMC samplers.
+    For the models whose DP prior is centered around a log-normal distribution we used parallel 
+    tempering. For the models gamma-centered models, we used a single chain. The temperature ladder was set as $t_s = 1.3^{s}$, 
+    for $s \in \lbrace0,1,\ldots,5$. This was set empirically in order to produce
+    acceptable swap probabilities both for the simulated data, and real data. Our attempts to use
+    a single chain produced sample with very slow mixing. On the other hand, from our examples 
+    we observe that posterior samples from the gamma-centered models, that do not use parallel 
+    tempering, show slightly stronger autocorrelation than those of the log-normal centered 
+    models. Parallel tempering produces chains with good mixing properties, but has a computational 
+    cost that  grows linearly with the number of temperatures. In view of this, we use it for the
 
@@
     distribution's shape parameter $\zeta\sim\mathcal{G}\left(1,1\right)$ and rate parameter
-    $\sigma\sim\mathcal{G}\left(2,2\right)$.  The choice of $\mathcal{G}(2,2)$ for rate
-    parameters was meant to draw estimates for the value away from 0 for numerical stability.
+    $\sigma\sim\mathcal{G}\left(2,2\right)$.  The choice of the $\mathcal{G}(2,2)$ for rate
+    parameters places no mass near 0, in order to draw estimates for the value away from 
+    0 for numerical stability.
 

### 2023-06-20 pointbypoint.tex
blah

--- 2023-06-14 pointbypoint.tex
@@
 \newcommand{\bruno}[1]{\textcolor{red}{#1}}
-
-% TODO
-%
-% in pointbypoint, add description of why parallel tempering for lognormal and why not in gamma.
-% additional description of prior elicitation
-% future work in conclusion.
-% Responses for already answered comments referencing previous comments.
-%
+\newcommand{\response}[1]{\textcolor{blue}{#1}}
 
@@
 
-We thank the referee for calling our attention to
+\response{We thank the referee for calling our attention to
     the issue of moment restrictions. The approach considered in the
@@
     moment conditions arise in our model. This has now been clarified in the paper.
+    }
 
-We note in passing that the moment conditions for the spectral distribution, say 
+\response{We note in passing that the moment conditions for the spectral distribution, say 
     $\Phi$, in the two-dimensional case, using the infinity norm, imply that 
@@
     result in a very restrictive model.
+    }
 
@@
 
-\comment{Please review this response.  I've done a little googling, but I can't figure
-    out how one would sample from the Bernstein polynomial model.  Maybe after a lot of
-    integration you could sort out some inverse CDF transform?}
-
-We thank the reviewer for inviting us to expound on differences between our approach
-    and prior contributions. We have clarified the distinctions between our approach 
-    and the prior contributions
-    mentioned.  \cite{guillotte2011} is a strictly bivariate model which assumes
+\response{We thank the reviewer for inviting us to be explicit about the advantages of 
+    our approach. \cite{guillotte2011} is a strictly bivariate model which assumes
     unit Fr\'{e}chet margins.  \cite{boldi2007} similarly assumes unit Fr\'{e}chet
@@
     This allows us to consider a wider class of distributions on pseudo-angles.
+    We have added a sentence in the introduction of the paper to highlight our 
+    contribution.
+    }
 
@@
 
-To address this concern, we have revised the simulation study to feature the 
+\response{To address this concern, we have revised the simulation study to feature the 
     average rise in energy score over baseline, averaged over 10 replications.  
     The simulated datasets are generated as mixtures of multivariate lognormals, 
-    projected onto the unit sphere.  As you say, this paper focuses on estimating
-    the spectral distribution, and ignores uncertainty from estimating marginal
-    parameters.  We have clarified as such.
-    \comment{I'm not sure how to address this.  I added a comment in the paper
-    that explicitly stated that uncertainty in marginal parameters is outside the
-    scope of the paper.  As a practical concern, every paper I have seen assumed
-    pre-standardized marginals, which implies estimating marginal parameters and
-    ignoring the uncertainty stemming from that estimation.}
+    projected onto the unit sphere.  The results are now presented in a new set
+    of plots in Figure 2. As for the lack of propagation of the uncertainty
+    related to the estimation of the Pareto marginals, we acknowledge that this is
+    indeed the case. A comprehensive approach is in theory possible, by using the
+    full Bayesian formalism. Yet the positive value function in the marginals 
+    introduce data dependent parameter space constraints that are essentially
+    impossible to deal with when assuming a flexible model for the angular
+    measure. Even for a simple parametric model, the exploration of the posterior
+    distribution becomes very challenging in high dimensions. We have a added
+    a statement pointing this issue in the section with the concluding remarks.
+    }
 
@@
 
-    \comment{We thank you for drawing our attention to this omission.  We have included the
-    details of the prior parameters at the start of section 5.}
+    \response{We thank the reviewers for drawing our attention to this omission.  We have 
+    included the details of the prior parameters at the start of section 5.}
 
@@
 
-    \comment{I'm guessing this guy is de Carvalho?}
-
-    \bruno{I have no doubts that this referee is Carvalho. I'll spend some time 
-    reading the papers that he mentions.}
+\response{We thank the referee for the very interesting suggestions for extensions 
+    of our proposed approach. We have a included a discussion of these ideas
+    in the final section of the paper.
+    }
 
@@
 
-        \comment{Yeah.  W is a multivariate GPD vector; Z is a multivariate standardized Pareto vector.
-        This needs to be fixed.}
-
-        We thank the referee for pointing out the inconsistencies in the notation. We have corrected them 
+\response{We thank the referee for pointing out the inconsistencies in the notation. We have corrected them 
         to make things clearer. We denote all random variables with capital letters and their values
         with small letters. The random variable corresponding to the conditional limiting distribution
-        $H$ is now denoted as $Z$.
+        $H$ is now denoted as $Z$. We have re-written Section 2 where the backgournd material on multivariate
+        PoT is presented.
+        }
 
@@
 
-        We have rewritten the material at the beginning of Section 3, explicitly mentioning the marginal
+\response{We have rewritten the material at the beginning of Section 3, explicitly mentioning the marginal
         univariate generalized Pareto, and how it is used to estimate the shape and scale parameters, 
         which are then used to standardize the observations.
+        }
 
@@
 
-        To ameliorate this shortfall, in addition to the foundational papers for the DP
+\response{To ameliorate this shortfall, in addition to the foundational papers for the DP
         already included, we add \cite{muller2015} to provide a modern approachable reference on
@@
         robustness of clusters generated under the DP.
+        }
 
@@
 
-            We thank the referee for the suggestion to change the layout of the paper. We have
+\response{We thank the referee for the suggestion to change the layout of the paper. We have
             re-written Section 3 to put a stronger emphasis on the presentation of the
             projected gamma kernel.
+            }
 
@@
 
-        \bruno{We need to get hold of a copy of the right book to see what the result is. If the result 
-        is so straightforward, then we need to decided what to do about it, but I am worried that
-        it will require a moment constraint.}
-
-        \comment{\cite{resnick2008extreme} is available on springerlink.  I'm skimming through it now.}
+\response{We apologize for the lack of clarity of our presentation of the background that supports 
+        our proposed model. We have rewritten Section 2 making clear that there are two possible
+        approaches to obtaining a multivariate PoT model. One is to assume a max-stability condition 
+        that leads to a conditional multivariate Pareto distribution. The other is to assume a 
+        condition of regular variation. We follow the former, and use the fact that there are a
+        number of representations of the resulting multivariate Pareto. One of them is based on 
+        using the infinity norm to represent the random variables that follow such distribution 
+        as the product of an angular and a radial component. As for the idea that a it is possible 
+        to use the $\tilde{W} = RV_p$, where $R$ is a standard Pareto and $V_p$ is located on the 
+        $\mathcal{L}_p$-sphere, as a representation of the multivariate Pareto distribution that 
+        we term $H$ in the paper, we have  not been able to find any results to justify it for 
+        $p\neq \infty$. It is possible to assume regular variation and obtain an equation like
+        Equation (1) in the paper for a $p$-norm. That, though, would entail either dealing with,
+        or ignoring, the moment constraints and the normalizing constant. Having said that, we
+        acknowledge that our model proposes a family of distributions on the $\mathcal{L}_p$-sphere
+        that can be useful in its own right. One possible application is that of inference for
+        directional data in many dimensions. That is mentioned in the final section of the manuscript.
+        }
 
@@
 
-        \bruno{This is an unfair comment. We do use the full multivariate distributiont to come up
-        with the results we present. I'll work on an answer to this one.}
+\response{The focus of our paper is to develop a model that is flexible and computationally tractable
+        for multivariate PoT extreme value analysis in many dimensions. We discuss and compute the
+        properties of the pair-wise extremal coefficients, as it is an important summary of multivariate
+        extreme value analysis. Yet, our inferential results are based on the full joint multivariate 
+        distribution, and use the whole vector of observations. It is not a pair-wise analysis. In 
+        particular, the conditional survival functions considered in Proposition 2 and illustrated in
+        Figure 7 make use of the whole joint distribution.
+        }
     \end{enumerate}
@@
 
-    We have included for comparison some other models established on $\mathcal{S}_{1}^{d-1}$;
-        notably the pairwise betas model developed in \cite{sabourin2013}, as well
-        as the Dirichlet distribution.  We have investigated several R and python packages to
-        compute empirical angular measure, or other means of estimating the spectral distribution.
-        To be clear, our analysis requires the ability to generate a sample from a posterior
-        predictive distribution, or at least sample from a fitted density, projected onto 
-        $\mathbb{S}_{\infty}^{d-1}$. Most packages implement a bivariate angular density, 
-        including \verb|ExtremalDep|~\citep{ExtremalDep}, \verb|BAMBI|~\citep{BAMBI}, 
-        \comment{expand list}.  Some, such as
-        \verb|evd|~\citep{evd} implement a multivariate generalized Pareto density with a 
-        parametric dependence structure, but I didn't see a way to learn the dependence
-        structure in isolation, as well as sample from it.  I found 
-        \cite{BMAmevt} as an example of a true multivariate density on $\mathbb{S}_1^{d-1}$, 
-        and on first attempt to use
-        it I encountered a bug with its output (for distributions with more than 3 dimensions, 
-        the dimension of the output parameter space became larger than it should be).  Upon 
+\response{We thank the referee for prompting us to seek a more thorough comparison of our proposed
+        model with existing models. To this end investigated several R and python packages designed 
+        to compute empirical angular measure in highly multivariate settings, with the ability to 
+        provide predictive uncertainty. We found that most packages implement methods to estimate
+        a bivariate angular density,}  \response{including} \verb|ExtremalDep| \response{\citep{ExtremalDep},}
+        \verb|BAMBI| \response{\citep{BAMBI}. We found that} \verb|evd| \response{\citep{evd} implements a multivariate 
+        generalized Pareto density with a  parametric dependence structure, but it is unclear to us 
+        how to learn the dependence structure and sample from it. 
+        notably} \verb|BMAmevt| \response{\citep{BMAmevt} that implements the 
+        pairwise betas model developed in \cite{sabourin2013}, as well as the Dirichlet distribution,
+        is an example of a true multivariate density on $\mathbb{S}_1^{d-1}$.  Our first attempt to use
+        it found a bug with its output for distributions with more than 3 dimensions.  Upon 
         conferring with the package maintainer and author, they identified the source of the
         bug and fixed it.  For this reason, we are able to include it as a point of comparison.
-
-    \bruno{Well, here we have to bite the bullet. Hopefully there some software that allows us
-    to fit the empirical angular measure.}
+}    
+    \bruno{You need to include in the paper some justification of why the package you used was the only
+    one considered.}
 
@@
 
-            We thank the reviewers for bringing this omission to our attention.  We have now
-            included a discussion of MCMC convergence among the models in section 5.
+\response{We thank the reviewers for bringing this omission to our attention.  We have now
+            included a discussion of MCMC convergence among the models in section 5.  In the current
+            version of the paper we clarify that we used two different strategies to run the MCMC.
+            For the model that have a prior distribution centered around a log-normal we observed 
+            that a single chain Metropolis-Hastings algorithm had very slow mixing. Thus we resorted 
+            to implementing a parallel tempering approach. This improved the mixing, at the cost 
+            of increasing computation times per iterations. In addition, parallel tempering requires
+            tuning of the temperature scale, something that can be time-consuming. The models that
+            use a prior centered around a product of gammas have substantially better mixing than the
+            log-normal centered models, albeit the mixing is not ideal. Yet each iteration is much
+            faster, so consider a strategy of running the chain longer and doing more thinning, than 
+            in the former case. This is now discussed in Section 5. In order to monitor convergence 
+            we used plots of the log-posterior density.
+            Below are the log-density by iteration plots for the ERA-5 data, on the two 
+            unrestricted projected gamma models; left uses the product of gammas centering distribution, 
+            right uses the log-normal centering distribution.  The red line indicates the end of the 
+            burn-in period, and we only keep every tenth sample after.
+            }
+
+            \includegraphics[width=0.45\textwidth]{images/ERA-5-PG-G.pdf}
+            \includegraphics[width=0.45\textwidth]{images/ERA-5-PG-LN.pdf}
+
+            \bruno{You should include, also, a graph of the autocorrelation functons of these two values.
+            Now, there is something that is bit contradicting, you say you run the PG-G longer and do more
+            thinning and the PG-LN shorter and do less thinning. But in both cases you use 40K and thin
+            every 10. How is that?}
 
@@
 
-            We have converted the graphs to use average rise in score over baseline (as calculated
-            by average of model on 10 distributions), rather than the score directly on a single
-            sample. We have verified that instances of displayed values dropping below zero are 
-            within MCMC error of zero.\comment{I need to say this better.}
+\response{The plots in Figure 2 are totally new, as they are based on a new set of experiments 
+            and comparisons.  We now show average rise in score over baseline (as calculated
+            by average of model on 10 simulations), rather than the score directly on a single
+            sample. The averages in the figures are all positive.
+            }
 
@@
 
-    \bruno{This was dealt with already.}
+\response{Please refer to the answer to the first question of Referee 1}
 
@@
 
-    \bruno{We need to rewrite the proof. I'll do that.}
+\response{We thank the referee for pointing out the inconsistency and apologize for the sloppy presentation 
+    of the result. We have rewritten propositions 3 and 4 and their proofs.
+    }
 
@@
 
-    \comment{Indeed, it's not an orthogonal projection.  Even the $\mathcal{L}_2$ norm doesn't 
-    produce an orthogonal projection, unless using the trigonometric mapping.  I don't think that's
-    a problem.}
-
-    \bruno{I think we need to de-emphasize the word ``projection''. We will call it ``normalization''
-    the first time we mention it, and then say that we will use the term ``projection''. It is a
-    projection in the sense that it is the point in $\mathbb{S}_p^{d-1}$ that is closest to the original
-    point, is it not?}
+\response{The transformation that maps a point onto a hypercube is indeed not an ``orthogonal'' projection. 
+    In Section 3.1 we acknowledge that this is an abuse of terminology. 
+    }
 
@@
 
-    \comment{we're using $l$ for iterating observations, rather than dimensions?  Yeah, it's probably
-    bad wording.}
+\response{The introduction of Section 3 has been rewritten.}
 
@@
 
-    \comment{same as previous.}
+\response{See above.}
 \end{itemize}

### 2023-06-20 conclusion.tex
blah

@@
     the $\mathcal{L}_{\infty}$ norm.  Due to the inherent difficulty of 
-    obtaining a distribution with support on ${\mathbb S}^{d-1}_\infty$ our 
-    method projects distributions, supported on the positive quadrant and 
-    based on products of independent gammas,  onto the manifold 
-    ${\mathbb S}_{p}^{d-1}$. Samples of the resulting probability distribution 
-    are then projected onto ${\mathbb S}_{\infty}^{d-1}$.  As 
-    ${\mathbb S}_{p}^{d-1}$ converges to ${\mathbb S}_{\infty}^{d-1}$ as 
-    $p\to\infty$, we expect the resampling to be efficient for large enough
-    $p$. In fact, our exploration of the simulated and real data indicate
-    that the procedure is robust to the choice of moderately large values
-    of $p$.
+    obtaining the likelihood of distributions with support on ${\mathbb S}^{d-1}_\infty$ 
+    our method transforms the data to ${\mathbb S}_{p}^{d-1}$, fits then using mixtures of 
+    products of independent gammas, then transforms the predictions back to 
+    ${\mathbb S}^{d-1}_\infty$. As ${\mathbb S}_{p}^{d-1}$ converges to 
+    ${\mathbb S}_{\infty}^{d-1}$ as  $p\to\infty$, we expect the proposed 
+    resampling to be efficient for large enough $p$. In fact, our exploration 
+    of the simulated and real data indicates that the procedure is robust to 
+    the choice of moderately large values of $p$. 
+    Our method includes two inferential steps. The first consists of the
+    estimation of the marginal Pareto distributions. The second consists 
+    of the estimation of the angular density. Parameter uncertainty incurred in
+    the former is not propagated to the latter. Conceptually, and integrated approach 
+    that accounts for all the estimation uncertainty is conceivable. Unfortunately, 
+    this leads to posterior distributions with complex data dependent restrictions 
+    that  are very difficult to explore, especially in large dimensional setting. In 
+    fact, our attempts to fit a simple parametric model for the marginals and the
+    angular measure jointly in several dimensions were not successful.
+
+In this paper we have focused on a particular representation of the multivariate
+    Pareto distribution for PoT inference on extreme values. To this end, our model 
+    provides a computationally efficient and flexible approach. An interesting extension 
+    of the proposed model is to consider regressions of extreme value responses, due
+    to extreme value inputs following the ideas in  \cite{carvalho2022}. This will 
+    produce PoT based Bayesian non-parametric extreme value regression models. More 
+    generally,  models that allow for covariate-dependent extremal dependence 
+    \citep{mhalla2019} could be considered. In addition, we notice that our approach 
+    is based on flexibly  modeling angular distributions for any $p$-norm. As such, 
+    it can be applied to other problems focused on high dimensional directional 
+    statistics  constrained  to a cone of directions.
 
@@
     term---the rate parameter of the gamma distribution---can offer additional
-    model fidelity.  This result is not borne out in our real data example,
-    for which the simpler model is vastly preferred.  Additionally, we observe
+    model fidelity.  \makenote{However,} this result is not borne out in our 
+    real data example, for which in higher dimensions the simpler model is 
+    strongly preferred.  Additionally, we observe
     that the choice of the log-normal centering distribution and additional
-    flexibility it offers did not translate to additional model fidelity.  This
-    is borne out in the real data, where on the ERA--Interim dataset it performed
-    only comparably, but on ERA--5 it performed significantly worse.
-
-    % From this,
-    % we learned that the additional flexibility offered by varying the $\beta$ 
-    % term---the rate parameter of the Gamma distribution---does not result in 
-    % increased model fidelity in a mixture setting.  Within the tested range, 
-    % between 3 and 20 dimensions, we observed that the additional information 
-    % provided by a log--normal centering distribution did not translate to 
-    % additional model fidelity.  However, for real data, with 47 dimensions, 
-    % we did observe an effect.
+    flexibility it offers did not translate to additional model fidelity.  
+    \makenote{This result is borne out in the real data, where it performed at best 
+    comparably to the gamma model.}  
+    % This
+    % is borne out in the real data, where on the ERA--Interim dataset it performed
+    % only comparably, but on ERA--5 it performed significantly worse.
+    % {\bf I am not sure that this paragraph is appropriate anymore. I think that
+    % the different mixture models considered are pretty much equivalent, so we should
+    % favor the one that is easiest and most robust to fit.}
 
@@
     computational approaches that will make it feasible to tackle very high 
-    dimensional problems. In fact, to elaborate on the study of IVT, there is 
-    a need to consider several hundreds, if not thousands, of grid cells over
-    the Pacific Ocean in order to obtain a good description of atmospheric 
-    events responsible for large storm activity over California.  Moving beyond
-    this model
+    dimensional problems, such as variational Bayes. In fact, to elaborate on 
+    the study of IVT, there is a need to consider several hundreds, if not thousands, 
+    of grid cells over the Pacific Ocean in order to obtain a good description of 
+    atmospheric  events responsible for large storm activity over California.
 

### 2023-06-20 introduction.tex
blah

@@
     called moment constraint by considering a constructive definition of the 
-    multivariate GP, based on the infinity norm.
+    multivariate GP, based on the infinity norm. The approach proposed in this paper
+    adds to the growing literature on Bayesian models for multivariate extreme
+    value analysis (see, for example, \cite{boldi2007},\cite{guillotte2011},
+    \cite{SaNa2014},\cite{hanson2017}), providing a model that has strong computational
+    advantages due its structural simplicitly, achieves flexibility using 
+    a mixture model, imposes no moment constraints, and scales well to moderately 
+    large dimensions.
 

### 2023-06-20 results.tex
blah

@@
     \centering
-    % \caption{Energy score (top row) and Posterior Predictive Loss (bottom row) for various models on simulated data with ascending count of mixture components (indicated by plot heading) and number of dimensions (indicated by horizontal axis).
-Gen indicates baseline.\label{fig:simpples}}
-    \caption{Average energy score rise over baseline (on $\mathbb{S}_{\infty}^{d-1}$) for various models on simulated data, with ascending count of mixture components (indicated by plot heading) and number of dimensions (indicated by horizontal axis).
-Note that pairwise betas is a moment-restricted model.\label{fig:simpples}}
+    \caption{Average energy score rise over baseline (on $\mathbb{S}_{\infty}^{d-1}$) for various 
+    models fitted to simulated data, with ascending count of mixture components (indicated by plot
+    heading) and number of dimensions (indicated by horizontal axis).  Note that pairwise betas is
+    a moment-restricted model.\label{fig:simpples}}
     \includegraphics[height=3in, width = \textwidth]{./images/sim_es_rise}
@@
     an MCMC chain for a large-dimensional model, we monitor the log-posterior 
-    density. In all the examples considered MCMC samples produced stable tarces of the
-    log-posterior in less than 40 thousand  iterations. Thus, so we use that as a burn-in, 
+    density. In all the examples considered, MCMC samples produced stable traces of the
+    log-posterior in less than 40 thousand  iterations. We use that as a burn-in, 
     and thereafter sample 10 thousand additional iterations. We then thin the chain by retaining one
     every ten samples, to obtain 1000 total samples. These are used to generate samples from the 
-    posterior predictive densities. We used two different strategies implemment the MCMC samplers.
+    posterior predictive densities. We used two different strategies to implement the MCMC samplers.
     For the models whose DP prior is centered around a log-normal distribution we used parallel 
-    tempering. For the models gamma-centered models, we used a single chain. The temperature ladder was set as $t_s = 1.3^{s}$, 
-    for $s \in \lbrace0,1,\ldots,5$. This was set empirically in order to produce
-    acceptable swap probabilities both for the simulated data, and real data. Our attempts to use
-    a single chain produced sample with very slow mixing. On the other hand, from our examples 
-    we observe that posterior samples from the gamma-centered models, that do not use parallel 
-    tempering, show slightly stronger autocorrelation than those of the log-normal centered 
-    models. Parallel tempering produces chains with good mixing properties, but has a computational 
-    cost that  grows linearly with the number of temperatures. In view of this, we use it for the
+    tempering. This serves to overcome the very slow mixing that we observed for these cases.
+    The temperature ladder was set as $t_s = 1.3^{s}$,  for $s \in \lbrace0,1,\ldots,5\rbrace$. This was 
+    set empirically in order to produce acceptable swap probabilities both for the simulated data, 
+    and real data. Parallel tempering produces chains with good mixing properties, but has a 
+    computational cost that  grows linearly with the number of temperatures. Thus, for the 
+    gamma-centered models, we used a single chain. We leverage the fast speed of each iteration, 
+    to obtain a large number of samples, that are appropriately thinned to deal with a mild
+    autocorrelation. In summary, the strategy for log-normal centered models is based on a costly
+    sampler with good mixing properties. The strategy for the gamma-centered models is based on
+    a cheap sampler that can be run for a large number of iterations.
 
@@
     projected onto $\mathbb{S}_{\infty}^{d-1}$. The generation 
-    procedure is detailed in Algorithm~\ref{algo:simulated}.  For each
-    model, the data is then projected onto the appropriate $\mathbb{S}_p^{d-1}$
-    for model fitting--dirichlet and pairwise betas being fit on 
-    $\mathbb{S}_1^{d-1}$, and the projected gamma models being fit on 
-    $\mathbb{S}_{10}^{d-1}$.
+    procedure is detailed in Algorithm~\ref{algo:simulated}.  We consider two gamma-centered
+    and two log-normal centered DP mixture models, with and without restrictions in each
+    case. We also consider a Dirichlet distribution and 
+    a pairwise betas model \citep{COOLEY2010}, fitted using the \verb|BMAmevt| 
+    package in R~\citep{BMAmevt}. For the DP mixture
+    models, the data are projected onto $\mathbb{S}_{10}^{d-1}$. For the other
+    two models they are projected on $\mathbb{S}_1^{d-1}$.
 
-\begin{algorithm}[h]  % algorithm using algorithm package
+\begin{algorithm}[ht]  % algorithm using algorithm package
         \caption{Simulated Angular Dataset Generation Routine\label{algo:simulated}.
@@
             \EndFor
-            \State Generate $\pi\sim\text{Dirichlet}(\bm{10})$
+            \State Generate $\pi\sim\text{Dirichlet}(\bm{10}_{n_{\text{mix}}})$
             \For{$i$ in $1,\ldots,1000$}
@@
     generating distribution as the training dataset, evaluated against the training 
-    dataset.  We considered a total of four variations of the Dirichlet-process mixture
-    of projected gammas model: whether to restrict $\beta_{j\ell} := 1$ as the projected
-    \emph{restricted} gamma (DP-PRG) akin to a projected Dirichlet, or allow them to be 
-    learned with a gamma prior (DP-PG); and for the shape parameters whether to use a 
-    gamma centering distribution (-G) or a log-normal centering distribution (-LN).
-    For comparison we also included two other distributions on $\mathbb{S}_{1}^{d-1}$:
-    The Dirichlet distribution (with gamma priors), and the pairwise betas distribution
-    developed in~\cite{COOLEY2010}.  
+    dataset.  
+%    We considered a total of four variations of the Dirichlet-process mixture
+%    of projected gammas model: whether to restrict $\beta_{j\ell} := 1$ as the projected
+%    \emph{restricted} gamma (DP-PRG) akin to a projected Dirichlet, or allow them to be 
+%    learned with a gamma prior (DP-PG); and for the shape parameters whether to use a 
+%    gamma centering distribution (-G) or a log-normal centering distribution (-LN).
+%    For comparison we also included two other distributions on $\mathbb{S}_{1}^{d-1}$:
+%    The Dirichlet distribution (with gamma priors), and the pairwise betas distribution
+%    developed in~\cite{COOLEY2010}.  
     % Note that the pairwise betas model respects
@@
     % it is}
-    We observe that for simulated data using Algorithm~\ref{algo:simulated}, the 
-    projected gamma models dominate, and within that class of model, the choice of 
-    centering distribution is irrelevant.  For the projected \emph{restricted} 
-    gamma models, the choice of a multivariate log-normal centering distribution 
-    appears to negatively affect model fidelity.
+    \makenote{It was anticipated that the log-normal centering distribution would place
+    more mass of the  DP into regions of the support with active clusters, which would 
+    generate new clusters more readily.  For the simulated data,} 
+    we observe that the projected gamma models dominate, and within that class
+    of model, the choice of centering distribution is immaterial.  For the 
+    projected \emph{restricted} gamma models, the choice of a multivariate log-normal 
+    centering distribution appears to \makenote{somewhat} negatively affect model 
+    fidelity.  \makenote{Additionally, the log-normal models with the additional computational cost
+    of multivariate normal evaluation and parallel tempering, take approximately six times
+    longer to sample relative to the gamma models.}
+    % \bruno{The comparison needs some information about the computing times.}
 
@@
 
-Our data corresponds to daily average values for the IVT magnitude
+Our data correspond to daily average values for the IVT magnitude
     along the coast of California.  The ERA-Interim data used covers the time period 
     1979 through 2014 (37 years) omitting leap days, and eight grid cells that 
-    correspond to the coast of California.  The ERA5 data covers the time period 
+    correspond to the coast of California.  The ERA5 data cover the time period 
     1979 through 2019 (42 years) with the same restriction, and  47 grid cells for 
@@
     Figure~\ref{fig:gridlocs} provides a visual representation of the area these grid 
-    cells represent.
+    cells cover.
 
@@
 %     obtained in the simulation example.
-Table~\ref{tab:dev} shows the values of the estimated posterior predictive loss and energy 
+Table~\ref{tab:dev} shows the values of the estimated energy 
     scores for the different models considered. We observe that, contrary to the results in the
@@
     gamma models, though for the lower-dimensional ERA-Interim data, all models perform comparably.
-    Interestingly, this preference for the restricted gamma form is the opposite of the result 
-    obtained in the simulation example.
 

### 2023-07-07 pointbypoint.tex
prep for submission

@@
 \begin{document}
+
+Dear Professor Jasra,
+
+\vskip 1.5cm
+
+Please consider a revised version of the manuscript `Bayesian non-parametric inference for 
+multivariate peaks-over-threshold models''. We have carefully considered the comments made
+by the referees, and produced a new version of the paper that we believe shows substantial
+improvement with respect to the original version. We notice that addressing the reviewers' 
+comments required significant
+    effort in redeveloping the simulation study.  To that end, we converted the generating distribution
+    for the simulations from mixtures of projected gammas, to mixtures of projected multivariate 
+    log-normals so that we are not merely recovering the generating distribution.  Further, the analysis 
+    from the simulation study has been made significantly more robust using multiple 
+    multiple replicates of the generating distribution.  The results are now reflected in a new figure
+    that should make them more easily interpretable.  After a thorough search for available software,
+    suitable to perform a comparison with our proposed method, we added two models for 
+    comparison: one based on Dirichlet distributions, and one based on pairwise beta distributions.  In
+    addition, we have followed the referees' advice of sharpening the presentation, including
+    justification, notation and proofs of the stated results.
+
+We believe all concerns raised by the reviewers have been addressed, and we have elaborated our responses
+    to each raised issue below.  Thank you for your time and effort in consideration of our manuscript.
+
+% The proof of negative definiteness of our kernel for the energy score has been fixed.  The rotation that
+%     previously served as a convenient method of calculation serves in the proof itself to verify negative
+%     definiteness.  The ordering of the section has been revised to reflect this.
 
@@
     indeed the case. A comprehensive approach is in theory possible, by using the
-    full Bayesian formalism. Yet the positive value function in the marginals 
-    introduce data dependent parameter space constraints that are essentially
+    full Bayesian formalism. Yet, the positive value function in the marginals 
+    introduce data dependent constraints for the parameter space. These are essentially
     impossible to deal with when assuming a flexible model for the angular
@@
 
-\response{To ameliorate this shortfall, in addition to the foundational papers for the DP
-        already included, we add \cite{muller2015} to provide a modern approachable reference on
-        and associated clustering, as well as \cite{ascolani2022} which provides a discussion of
-        robustness of clusters generated under the DP.
+\response{To include non-parametric modeling background background to the list of references, in 
+        addition to the foundational papers for the DP already included, we added \cite{muller2015} 
+        to provide a modern approachable and comprehensive reference,  as well as the more specific
+        reference \cite{ascolani2022},  which provides a discussion of robustness of clusters 
+        generated under the DP.
         }
@@
         using the infinity norm to represent the random variables that follow such distribution 
-        as the product of an angular and a radial component. As for the idea that a it is possible 
+        as the product of an angular and a radial component. As for the idea that it is possible 
         to use the $\tilde{W} = RV_p$, where $R$ is a standard Pareto and $V_p$ is located on the 
@@
 \response{We thank the referee for prompting us to seek a more thorough comparison of our proposed
-        model with existing models. To this end investigated several R and python packages designed 
+        model with existing models. To this end we investigated several R and python packages designed 
         to compute empirical angular measure in highly multivariate settings, with the ability to 
@@
         a bivariate angular density,}  \response{including} \verb|ExtremalDep| \response{\citep{ExtremalDep},}
-        \verb|BAMBI| \response{\citep{BAMBI}. We found that} \verb|evd| \response{\citep{evd} implements a multivariate 
+        \response{and} \verb|BAMBI| \response{\citep{BAMBI}. We found that} \verb|evd| \response{\citep{evd} implements a multivariate 
         generalized Pareto density with a  parametric dependence structure, but it is unclear to us 
         how to learn the dependence structure and sample from it. 
-        notably} \verb|BMAmevt| \response{\citep{BMAmevt} that implements the 
+        Notably} \verb|BMAmevt| \response{\citep{BMAmevt} implements the 
         pairwise betas model developed in \cite{sabourin2013}, as well as the Dirichlet distribution,
-        is an example of a true multivariate density on $\mathbb{S}_1^{d-1}$.  Our first attempt to use
+        an example of a multivariate density on $\mathbb{S}_1^{d-1}$.  Our first attempt to use
         it found a bug with its output for distributions with more than 3 dimensions.  Upon 
         conferring with the package maintainer and author, they identified the source of the
-        bug and fixed it.  For this reason, we are able to include it as a point of comparison.
+        bug and fixed it.  For this reason, we are able to include it as a point of comparison. The
+        comparison consists of the energy scores. For the IVT data we also include the computing
+        times. Due to the very large nature of the exercise, computations for the simulated examples 
+        were performed in a number of different environments, using different number of processors. Thus
+        computation times are difficult to compare. For the IVT examples all computations were performed 
+        in the same computer, serially. Thus, the comparison of computing times is fair. 
 }
-    \bruno{You need to include in the paper some justification of why the package you used was the only
-    one considered.}
 
@@
             to implementing a parallel tempering approach. This improved the mixing, at the cost 
-            of increasing computation times per iterations. In addition, parallel tempering requires
+            of increasing computation times per iteration. In addition, parallel tempering requires
             tuning of the temperature scale, something that can be time-consuming. The models that
@@
             log-normal centered models, albeit the mixing is not ideal. Yet each iteration is much
-            faster, so consider a strategy of running the chain longer and doing more thinning, than 
+            faster, so we adopt a strategy of running the chain longer and doing more thinning, than 
             in the former case. This is now discussed in Section 5. In order to monitor convergence 
@@
             \includegraphics[width=0.45\textwidth]{images/ERA-5-PG-G.pdf}
-            \includegraphics[width=0.45\textwidth]{images/ERA-5-PG-LN.pdf}
+            \includegraphics[width=0.45\textwidth]{images/ERA-5-PG-LN.pdf} \\
 
-            \bruno{You should include, also, a graph of the autocorrelation functons of these two values.
-            Now, there is something that is bit contradicting, you say you run the PG-G longer and do more
-            thinning and the PG-LN shorter and do less thinning. But in both cases you use 40K and thin
-            every 10. How is that?}
+            \response{The choice of 40,000 iterations burn-in, and thinning the resulting chain, 
+            provides a consistent strategy that produces satisfactory posterior samples for both 
+            models.}
+
+            % \bruno{You should include, also, a graph of the autocorrelation functons of these two values.
+            % Now, there is something that is bit contradicting, you say you run the PG-G longer and do more
+            % thinning and the PG-LN shorter and do less thinning. But in both cases you use 40K and thin
+            % every 10. How is that?}
 

### 2023-07-07 conclusion.tex
prep for submission

@@
     of the estimation of the angular density. Parameter uncertainty incurred in
-    the former is not propagated to the latter. Conceptually, and integrated approach 
+    the former is not propagated to the latter. Conceptually, an integrated approach 
     that accounts for all the estimation uncertainty is conceivable. Unfortunately, 
     this leads to posterior distributions with complex data dependent restrictions 
-    that  are very difficult to explore, especially in large dimensional setting. In 
+    that  are very difficult to explore, especially in large dimensional settings. In 
     fact, our attempts to fit a simple parametric model for the marginals and the
@@
 
-We explored the behavior of the proposed model using simulated data from
-    mixtures of projected Gammas with varying degrees of complexity---both 
-    in terms of dimensionality and in number of mixture components.  From this,
-    we observed that the additional flexibility offered by varying the $\beta$
-    term---the rate parameter of the gamma distribution---can offer additional
-    model fidelity.  \makenote{However,} this result is not borne out in our 
-    real data example, for which in higher dimensions the simpler model is 
-    strongly preferred.  Additionally, we observe
-    that the choice of the log-normal centering distribution and additional
-    flexibility it offers did not translate to additional model fidelity.  
-    \makenote{This result is borne out in the real data, where it performed at best 
-    comparably to the gamma model.}  
+%We explored the behavior of the proposed model using simulated data from
+    %mixtures of projected Gammas with varying degrees of complexity---both 
+    %in terms of dimensionality and in number of mixture components.  From this,
+    %we observed that the additional flexibility offered by varying the $\beta$
+    %term---the rate parameter of the gamma distribution---can offer additional
+    %model fidelity.  \makenote{However,} this result is not borne out in our 
+    %real data example, for which in higher dimensions the simpler model is 
+    %strongly preferred.  Additionally, we observe
+    %that the choice of the log-normal centering distribution and additional
+    %flexibility it offers did not translate to additional model fidelity.  
+    %\makenote{This result is borne out in the real data, where it performed at best 
+    %comparably to the gamma model.}  
     % This

### 2023-07-07 introduction.tex
prep for submission

@@
 
-The reminder of this paper is outlined as follows. 
+The remainder of this paper is outlined as follows. 
     Section~\ref{sec:multivariatepot} comprises a brief review of multivariate PoT, 
@@
       measure, based on projecting an arbitrary distribution supported in ${\mathbb R}_+^d$ 
-      onto unit hyper-spheres defined for $p$-norms. 
+      onto unit hyper-spheres defined by $p$-norms. 
     Section~\ref{sec:evaluation} develops criteria for model selection in the support 
@@
     Inference for the limiting distribution of the exceedances
-    needs to account for the normalizing constant in Equation \ref{exponent-mu}, as well
+    needs to account for the normalizing constant in Equation (\ref{exponent-mu}), as well
     the moment constraints. In addition, the moment constraints can impose unrealistic

### 2023-07-07 methodology_projgamma.tex
blah
prep for submission

@@
 \subsection{Projected gamma family\label{subsec:projgamma}}
-To obtain a distribution on 
-${\mathbb S}_{p}^{d-1}=\{\bm{s} : \bm{s} \in {\mathbb R}_{+}^{d}, \| \bm{s}\|_{p} = 1\}$ 
-we start with a vector in 
-$\bm{x} \in {\mathbb R}^d_+$, calculate its $\mathcal{L}_p$-norm  defined as
+The $\mathcal{L}_p$-norm, for $p>0$ is  defined as
   \begin{equation*}
@@
   \end{equation*}
-  and normalize it to obtain $\bm{y} = \bm{x}/\|\bm{x}\|_p \in {\mathbb S}_p^{d-1}$. 
 The absolute and Euclidean norms are obtained for $p=1$ and $p=2$ respectively,
@@
 
+To obtain a distribution on 
+${\mathbb S}_{p}^{d-1}=\{\bm{s} : \bm{s} \in {\mathbb R}_{+}^{d}, \| \bm{s}\|_{p} = 1\}$ 
+we start with a vector in 
+$\bm{x} \in {\mathbb R}^d_+$, and normalize it to obtain 
+$\bm{y} = \bm{x}/\|\bm{x}\|_p \in {\mathbb S}_p^{d-1}$. 
 A natural  distribution to consider in ${\mathbb R}^d_+$ is given by a product of independent
@@
     $\alpha_\ell$ and $\beta_\ell$ are the shape and scale parameters, respectively. 
-    For any finite $p$, letting 
+    For any finite $p>0$, letting 
    $ y_d = (1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p)^{1/p}$,
@@
       &\hspace{0.5cm}\times r^{d-1}\left[y_d +
-            {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right].
+            {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p y_d^{1 - p}\right].
     \end{aligned}
@@
       &\hspace{0.5cm}\times \left[y_d +
-          {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right]\\
+          {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p y_d^{1 - p}\right]\\
       &\hspace{0.5cm}\times \frac{\Gamma({\textstyle\sum}_{\ell = 1}^d\alpha_{\ell})}{\left({\textstyle\sum}_{\ell = 1}^d
@@
   for $p=2$ as a multivariate distribution for directional data, using spherical coordinates.  
-  For $\bm{ y}\in {\mathbb S}_1^{d-1}$, and $\beta_{\ell} = \beta$ for all $\ell$, the density in Equation (\ref{eqn:projgamma}) corresponds to that of a Dirichlet distribution.
+  For $\bm{ y}\in {\mathbb S}_1^{d-1}$, and $\beta_{\ell} = \beta$ for all $\ell$, the density in
+  Equation (\ref{eqn:projgamma}) corresponds to that of a Dirichlet distribution.
 
@@
     groups observations into stochastically assigned clusters, where members of a cluster share 
-    distributional parameters \citep{ascolani2022}. Building out the methods of inference for Equation~(\ref{eqn:dppg}), 
+    distributional parameters \citep{muller2015,ascolani2022}. Building out the methods of inference for Equation~(\ref{eqn:dppg}), 
     let $n_j^{-i}$ be the number of observations in cluster $j$ not including observation $i$.  Let $J^{-i}$ be 
-    the number of extant clusters, not including any singleton that observation $i$ is in. Under this model, 
+    the number of extant clusters, not including any singleton containing observation $i$. Under this model, 
     the probability of cluster membership for a given observation is proportional to
@@
 We also assume a Gamma prior on $\eta$, that is updated via the procedure outlined in \cite{escobar1995}.
-We refer to this model as the \emph{projected Gamma--Gamma} (PG--G) model.
+We refer to this model as the \emph{projected gamma--gamma} (PG--G) model.
 An advantage of the PG--G model is that, thanks to conjugacy, the rate parameters can easily be integrated out for inference on $\bm{\alpha}_j$.
@@
 \end{equation}
-for $\ell = 2,\ldots, d$.
-Updating $\beta_{j\ell}$ is done via a Gibbs step.
-The hyper-parameters $\xi_{\ell},\tau_{\ell},\zeta_{\ell},\sigma_{\ell}$ follow similar Gamma-Gamma update relationships.
-We also explore a restricted form of this model, where $\beta_{\ell} = 1$ for all $\ell$.
-Under this model, we use the full conditional in Equation~(\ref{eqn:alpha1update}) for all $\ell$, and omit inference on $\bm{\zeta},\bm{\sigma}$.
-We refer to this model as the \emph{projected restricted Gamma--Gamma} (PRG--G) model.
+for $\ell = 2,\ldots, d$.  Updating $\beta_{j\ell}$ is done via a Gibbs step.  The hyper-parameters
+$\xi_{\ell},\tau_{\ell},\zeta_{\ell},\sigma_{\ell}$ follow similar gamma-gamma update relationships.
+We also explore a restricted form of this model, where $\beta_{\ell} = 1$ for all $\ell$.  Under this 
+model, we use the full conditional in Equation~(\ref{eqn:alpha1update}) for all $\ell$, and omit 
+inference on $\bm{\zeta},\bm{\sigma}$.  We refer to this model as the 
+\emph{projected restricted gamma--gamma} (PRG--G) model.
 
-The second form of centering distribution we explore is a multivariate log-normal distribution on the shape parameters $\bm{\alpha}_j$, with independent Gamma $\beta_{j\ell}$ rate parameters.  
+The second form of centering distribution we explore is a multivariate log-normal distribution on the shape parameters $\bm{\alpha}_j$, with independent gamma $\beta_{j\ell}$ rate parameters.  
     \begin{equation}
@@
     $\bm{\Sigma}$, and Gamma priors on $\zeta_{\ell}$ and $\sigma_{\ell}$, and $\eta$.  
-    This model is denoted as the \emph{projected Gamma--log-normal} (PG--LN) model.  
+    This model is denoted as the \emph{projected gamma--log-normal} (PG--LN) model.  
     We also explore a restricted Gamma form of this model as above, where 

### 2023-07-07 results.tex
prep for submission

@@
     $\sigma\sim\mathcal{G}\left(2,2\right)$.  The choice of the $\mathcal{G}(2,2)$ for rate
-    parameters places no mass near 0, in order to draw estimates for the value away from 
+    parameters places little mass near 0, in order to draw estimates for the value away from 
     0 for numerical stability.
@@
     projected onto $\mathbb{S}_{\infty}^{d-1}$. The generation 
-    procedure is detailed in Algorithm~\ref{algo:simulated}.  We consider two gamma-centered
+    procedure is detailed in Algorithm~\ref{algo:simulated}. We produced 
+    ten replicates of each configuration. We consider two gamma-centered
     and two log-normal centered DP mixture models, with and without restrictions in each
-    case. We also consider a Dirichlet distribution and 
-    a pairwise betas model \citep{COOLEY2010}, fitted using the \verb|BMAmevt| 
-    package in R~\citep{BMAmevt}. For the DP mixture
+    case. To perform a comparative analysis we fitted 
+   % \makenote{Comparative analysis using posterior predictive samples necessitates
+    %alternative models having defined methods of sampling from their respective posterior
+    %predictive distributions, for models defined on some normalization of the positive cone.  
+    %To that end, we also consider the Dirichlet distribution, as well as} 
+    %We also consider a Dirichlet distribution and 
+    the pairwise betas model proposed in \cite{COOLEY2010}. We chose this model for comparison
+    as it is implemented in the readily available package \verb|BMAmevt| 
+    in R~\citep{BMAmevt}. \verb|BMAmevt| provides samples from the predictive
+    posterior distribution. These are needed for the calculation of the energy
+    scores that are at the basis of our comparison. In addition, \verb|BMAmevt| can 
+    be fitted to moderately large multivariate observations. For the DP mixture
     models, the data are projected onto $\mathbb{S}_{10}^{d-1}$. For the other
-    two models they are projected on $\mathbb{S}_1^{d-1}$.
+    two models they are projected on $\mathbb{S}_1^{d-1}$.   We sampled each model for 50,000 
+    iterations, dropping the first 40,000 as burn-in, and thinning to keep every 10th iteration
+    after.  These settings were intended to provide a consistent sampling strategy that would
+    work with every model, even if inefficient for some.
 
@@
     a model, a posterior predictive dataset is generated, and the energy score is
-    calculated as a Monte Carlo approximation of Equation~\ref{eq:es}.
+    calculated as a Monte Carlo approximation of Equation~\ref{eq:es}.  In our analysis,
+    after burn-in and thinning, we had 1,000 replicates from the posterior distribution, and
+    generated 10 posterior predictive replicates per iteration.
     The \emph{baseline} value is the energy score of a new dataset from the same
@@
     % it is}
-    \makenote{It was anticipated that the log-normal centering distribution would place
-    more mass of the  DP into regions of the support with active clusters, which would 
-    generate new clusters more readily.  For the simulated data,} 
-    we observe that the projected gamma models dominate, and within that class
-    of model, the choice of centering distribution is immaterial.  For the 
-    projected \emph{restricted} gamma models, the choice of a multivariate log-normal 
-    centering distribution appears to \makenote{somewhat} negatively affect model 
-    fidelity.  \makenote{Additionally, the log-normal models with the additional computational cost
-    of multivariate normal evaluation and parallel tempering, take approximately six times
-    longer to sample relative to the gamma models.}
+   % \makenote{It was anticipated that the log-normal centering distribution would place
+    %more mass of the  DP into regions of the support with active clusters, which would 
+    %generate new clusters more readily.}  
+    For the simulated data, we observe that the projected gamma models dominate the other two
+    options considered, regardless of the choice of centering distribution.  The  projected 
+    restricted gamma models with a multivariate log-normal centering distribution appears to 
+    be dominated by the models based on the alternative centering distributions. Moreover, the
+    performance deteriorates with the increase in dimensionality. Additionally, models
+    centered around the log-normal distribution incur in the computational cost of 
+    multivariate normal evaluation and parallel tempering, taking approximately six times
+    longer to sample relative to the gamma models.  We also note that the 
+    computational cost  of the pairwise betas model grows combinatorially, with a 
+    sample space of dimension $\binom{d}{2} + 1$. In our testing, \verb|BMAmevt| is 
+    substantially faster than any of our proposed DP mixture models for low dimensions, 
+    however for examples with high number of dimensions, the computational time for 
+    \verb|BMAmevt| was comparable or greater than that for DP-PG-G.  We compare computing
+    times in our data analysis in Table~\ref{tab:sampletime}.
     % \bruno{The comparison needs some information about the computing times.}
@@
 \begin{figure}[tb]
-    \begin{center}
     \caption{Grid cell locations for ERA-Interim (left) and ERA5 (right).\label{fig:gridlocs}}
-    \begin{subfigure}[b]{0.48\textwidth}
     \centering
-        \includegraphics[height=2.5in]{./images/erai_grid}
-    \end{subfigure}
-    %
-    \begin{subfigure}[b]{0.48\textwidth}
+    \begin{minipage}{0.25\textwidth}
     \centering
-        \includegraphics[height=2.5in]{./images/era5_grid}
-    \end{subfigure}
-    \end{center}
+    \includegraphics[width=0.99\linewidth]{./images/erai_grid}
+    \end{minipage}%
+    \begin{minipage}{0.25\textwidth}
+    \centering
+    \includegraphics[width=0.99\linewidth]{./images/era5_grid}
+    \end{minipage}  
+    % \caption{Grid cell locations for ERA-Interim (left) and ERA5 (right).\label{fig:gridlocs}}
+    % \begin{center}
+    % \begin{subfigure}{0.48\textwidth}
+    %     \centering 
+    %     \includegraphics[height=2.5in]{./images/erai_grid}
+    % \end{subfigure}
+    % %
+    % \begin{subfigure}{0.48\textwidth}
+    %     \centering
+    %     \includegraphics[height=2.5in]{./images/era5_grid}
+    % \end{subfigure}
+    % \end{center}
 \end{figure}
@@
 
-% Table~\ref{tab:dev} shows the values of the estimated posterior predictive loss and energy 
-%     scores for the different models considered. We observe that, as in the case of the simulation 
-%     study shown in Figures~\ref{fig:simpples}, the preferred model is the projected restricted 
-%     gamma models, though for the lower-dimensional ERA-Interim data, all models perform comparably.
-%     Interestingly, this preference for the restricted gamma form is the opposite of the result 
-%     obtained in the simulation example.
+\begin{table}[htb]
+  \centering
+  \caption{Time to sample (in minutes) 50,000 iterations for various models\label{tab:sampletime}}
+  \include{./tables/table_sampletime}
+\end{table}
+
 Table~\ref{tab:dev} shows the values of the estimated energy 
@@
     gamma models, though for the lower-dimensional ERA-Interim data, all models perform comparably.
+    Table~\ref{tab:sampletime} shows the computing times needed to fit the different models to the
+    two datasets.  We see the effect of dimensionality on the various models; for gamma centered
+    models it grows linearly; for the log--normal centered model, it will grow superlinearly as matrix
+    inversion becomes the most costly operation.  For \verb|BMAmevt|, its parameter space grows
+    combinatorically with the number of dimensions, and thus so does computational complexity and
+    sampling time.
+    % \bruno{Insert here the table that you sent me in the email and comment the results.}
 
@@
   between $0.181$ and $0.840$ for the ERA5 data.  The greater range in dependence scores observed 
-  with the ERA5 data versus ERA-Interim speaks to the greater granularity of the ERA5 data--distance between locations is a strong contributor to the strength of the 
+  with the ERA5 data versus ERA-Interim speaks to the greater granularity of the ERA5 data,
+  indicating that distance between locations is a strong contributor to the strength of the 
   pairwise asymptotic dependence. The highest coefficients are $0.759$ for 

### 2023-11-26 main_comparison.tex
any updates

\documentclass[iicol,sn-basic]{sn-jnl}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{soul}
\usepackage{bm}
\usepackage{subcaption}
\usepackage{wrapfig}
\usepackage{anysize}
\usepackage{cancel}
\marginsize{1in}{1in}{1in}{1in}

\newtheorem{prop}{Proposition}
\DeclareMathOperator{\tr}{Tr}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\bigmid}{\,\middle\vert\,}

\newcommand{\findcite}{\textcolor{red}{[Find Citation]}}
\newcommand{\needcite}{\findcite}
\newcommand{\Chi}{\mbox{\Large$\chi$}}
\newcommand{\mathcolorbox}[2]{\colorbox{#1}{$\displaystyle #2$}}

\newcommand{\norm}[1]{\left\lVert #1 \right\rVert}
\newcommand{\inorm}[1]{\norm{#1}_{\infty}}
\newcommand{\pnorm}[2]{\norm{#1}_{#2}}
\newcommand{\prob}[1]{\text{Pr}\left[#1\right]}
\newcommand{\expect}[1]{\text{E}\left[#1\right]}

\jyear{2023}
\theoremstyle{thmstyleone}

\raggedbottom

\begin{document}

\title[~\hspace{6.75in}BNP inference for multivariate PoT]{Bayesian non-parametric inference for multivariate peaks-over-threshold models}

\author*{\fnm{Peter} \sur{Trubey}}\email{ptrubey@ucsc.edu}
\author{\fnm{Bruno} \sur{Sans\'{o}}}\email{bruno@soe.ucsc.edu}

\affil{\orgdiv{Department of Statistics}, \orgname{University of California Santa Cruz},
\orgaddress{\street{1156 High Street}, \city{Santa Cruz},\postcode{95064}, \state{CA}, \country{USA}}}

\abstract{\bf{\colorbox{red}{Note: Assume tables, algorithms and figures have changed.  Highlighting changes}}

\bf{\colorbox{red}{in these environments is not straightforward.}}

We consider a constructive definition of the multivariate Pareto that 
factorizes the random vector into a radial component and an independent angular 
component; The former following a univariate Pareto distribution, and the latter 
defined on the surface of the positive orthant of the infinity norm unit hypercube.  In this 
paper, we  propose a method for inferring the distribution of the angular 
component.  We identify its support as the limit of the positive orthant of the 
unit $p$--norm spheres, and introduce a projected gamma family of distributions 
defined \st{as}\hl{through} the \st{projec}\hl{normaliza}tion of a vector of independent gamma random variables \st{on}to 
the $p$--norm sphere.  This family serves as a building block for a flexible 
family of distributions obtained as a Dirichlet process mixture of projected 
gammas.  For model assessment, we discuss scoring methods appropriate to 
distributions on the unit hypercube.  In particular, working with the energy 
score criterion, we develop a kernel metric that produces a proper scoring rule, 
and present a simulation study to compare different modeling choices using the 
proposed scoring rules.  Finally we apply our approach to describe the 
dependence structure of the extreme values of the magnitude of the integrated 
vapor transport (IVT), data describing the flow of atmospheric moisture along 
the coast of California for the years of 1979 through 2020.  We find clear but 
heterogeneous geographical dependence.}
\keywords{Multivariate extremes, Peak over threshold models, Bayesian non-parametric models, Dirichlet process mixtures}

\maketitle

%--------content

\section{Introduction}

The statistical analysis of extreme values focuses on inference for rare events that correspond to the tails of probability distributions. As such, it is a key ingredient in the risk assessment of phenomena that can have strong societal impacts like floods, heat waves, high concentration of pollutants, crashes in the financial markets, among others. The fundamental challenge of extreme value theory (EVT) is to use information, collected over limited periods of time, to extrapolate to long time horizons. This sets EVT apart from most of statistical inference, where the focus is on the bulk of the distribution. Extrapolation to the tails of the distributions is possible thanks to theoretical results that give asymptotic descriptions of the probability distributions of extreme values.

Inferential methods for the extreme values of univariate observations are well established and software is widely available \cite[see, for example,][]{coles2001}. For variables in one dimension the application of EVT methods considers the asymptotic distribution of either the maxima calculated for regular blocks of data, or the values that exceed a certain threshold. The former leads to a Generalized Extreme Value (GEV) distribution, that depends on three parameters. The latter leads to a Generalized Pareto (GP) distribution, that depends on a shape and a scale parameter.  Likelihood-based approaches to inference can be readily implemented in both cases. In the multivariate case the GEV theory is well developed \citep[see, for example][]{dehaan2006}, but the inferential problem is \st{more} complicated\st{, as} \hl{by the fact that} there is no parametric representation of the GEV. This problem is inherited by the peaks over threshold (PoT) approach and compounded by the fact that there is no unique definition of an exceedance of a multivariate threshold, as there is an obvious dependence on the norm that is used to measure the size of a vector.

During the last decade or so, much work has been done in the exploration of the definition and properties
of an appropriate generalization of the univariate GP distribution to a multivariate setting.  To mention
some of the papers on the topic, the work of \citep{rootzen2006} defines the generalized Pareto distribution,
with further analysis on these classes of distributions presented in \cite{falk2008} and \cite{michel2008}.  
A recent review of the state of the art in multivariate peaks over threshold modelling using generalized 
Pareto is provided in \cite{rootzen2018} while \cite{RoSeWa2018a} provides insight on the theoretical 
properties of possible parametrizations. These are use in \cite{KiRoSeWa2019} for likelihood-based models 
for PoT estimation. A frequently used method for describing the dependence in multivariate distributions is
to use a copula. \cite{renard2007}, and \cite{falk2019} provide successful examples of this approach in an 
EVT framework. \cite{ferreira2014} presents a constructive definition of the Pareto process, that generalizes
the GP to an infinite dimensional setting. It consists of decomposing the process into independent radial and 
angular components. Such an approach can be used in the finite dimensional case, where the angular component 
contains the information pertaining to the dependence structure of the random vector. Based on this definition, 
we present a novel approach for modelling  the angular component with families of distributions that provide 
flexibility and can be applied in a moderately large dimensional setting.  Our focus on the angular measure is 
similar to that in \hl{\mbox{\cite{boldi2007}}}, \cite{SaNa2014} and \cite{HaCaCh2017}, that consider Bayesian 
non-parametric approaches. Yet, our approach \hl{differs in that it is established in the peaks-over-threshold 
regime, and }avoids the problem of dealing with the so called moment constraint by considering a constructive 
definition of the multivariate GP, based on the infinity norm. \hl{The approach proposed in this paper adds to 
the growing literature on Bayesian models for multivariate extreme value analysis (see, for example, 
\mbox{\cite{boldi2007}}, \mbox{\cite{guillotte2011}}, \mbox{\cite{SaNa2014}}, \mbox{\cite{hanson2017}}), providing a model that has strong computational advantages due its structural simplicitly, achieves flexibility using a mixture model, imposes no moment constraints, and scales well to moderately large dimensions.}

The remainder of this paper is outlined as follows. Section~\ref{sec:multivariatepot} comprises a brief review of multivariate PoT, detailing the separation of the radial measure from the angular measure. Section~\ref{sec:methodology} details our approach for estimating the angular measure, based on projecting an arbitrary distribution supported in ${\mathbb R}_+^d$ onto unit hyper-spheres defined \st{for}\hl{by} $p$-norms. Section~\ref{sec:evaluation} develops criteria for model selection in the support of the angular measure.  Section~\ref{sec:results} explores the efficacy of the proposed approach on a set of simulated data, and, acknowledging the relevance of extreme value theory to climatological events~\citep{jentsch2007,vousdoukas2018,li2019}, estimates the extremal dependence structure for a measure of water vapor flow in the atmosphere, used for identifying atmospheric rivers.  Finally, Section~\ref{sec:conclusion} presents our conclusions and discussion.

Throughout the paper, we adopt the operators $\wedge$ to denote minima, and the $\vee$ to denote maxima.  Thus $\wedge_i s_i = \min_i s_i$, and $\vee_i s_i = \max_i s_i$.  These operators can also be applied component-wise between vectors, such as $\bm{a}\wedge\bm{b} = (a_1\wedge b_1, a_2\wedge b_2,\ldots)$.  We use uppercase to indicate random variables, lowercase to indicate points, and bold-face to indicate vectors or matrices thereof.

\section{A multivariate PoT model\label{sec:multivariatepot}}
To develop a multivariate PoT model for extreme values consider a $d$-dimensional random vector $\bm{W} = (W_1, \ldots ,W_d)$ with cumulative distribution $F$. Following \cite{RoSeWa2018a}, assume that there exists sequences of vectors $\bm{a}_n$ and $\bm{b}_n$, and a $d$-variate distribution $G$ such that $\lim_{n\rightarrow\infty} F^n(\bm{a}_n \bm{w} + \bm{b}_n) = G(\bm{w})$. $G$ is a $d$-variate generalized extreme value distribution. It follows that
\begin{equation*}
\begin{aligned}
\lim\limits_{n\rightarrow\infty} &\prob{\bm{a}^{-1}_n (\bm{W} - \bm{b}_n)
\leq \bm{w} \mid \bm{W} \not\leq \bm{b}_n)} \\
&\hspace{0.5cm}= \frac{\log G(\bm{w} \wedge \bm{0})
- \log G(\bm{w}) }{\log G(\bm{0})} = H(\bm{w}).
\end{aligned}
\end{equation*}
$H$ is \st{the}\hl{a} multivariate Pareto distribution\hl{, and corresponds to a joint distribution conditional on exceeding a multivariate threshold. In addition to a copula function, $H$ depends on two $d$-dimensional vectors of parameters that we denote as $\bm{\xi}$ for the shapes and $\bm{\sigma}$ for the scales}. \cite{RoSeWa2018a} provide a number of stochastic representations for $H$. In particular\hl{, denote as $Z$ a random variable with distribution $H$ where $\bm{\xi}= 1$ and $\bm{\sigma} = 0$.  Then,} Remark 1 justifies the representation \st{given } in \cite{ferreira2014}, \st{consisting of taking $\bm{W}$, in the limit, as $\bm{W} = \bm{V}R$} \hl{giving $\bm{Z} = \bm{V}R$} where $R$ and $\bm{V}$ are independent. \st{$R = ||\bm{W}||_\infty$}\hl{$R = \|\bm{Z}\|_\infty$} is distributed as a standard Pareto random variable, and \st{$\bm{V} = \bm{W}/||\bm{W}||_\infty$}\hl{$\bm{V} = \bm{Z}/\|\bm{Z}\|_\infty$} is a random vector in $\mathbb{S}_{\infty}^{d-1}$, the positive orthant of the unit sphere \st{in infinite}\hl{under $\mathcal{L}_{\infty}$} norm, with distribution $\Phi$.  \hl{This representation is central to the methods proposed in this paper.  }$R$ and $\bm{V}$ are referred\hl{ to}, respectively, as the {\em radial} and {\em angular} components of $H$. The angular measure controls the dependence  structure of \st{ $\bm{W}$} \hl{$\bm{Z}$}  in the tails. In view of this, to obtain a  PoT model we seek a flexible model for the distribution of $\bm{V} \in {\mathbb S}_{\infty}^{d-1}$.

\hl{The approach considered in \mbox{\cite{rootzen2018}} focuses on the limiting conditional distribution $H$. An alternative approach consists of assuming that regular variation \mbox{\cite[see, for example,][]{resnick2008extreme}} holds for the limiting distribution of $\bm{W}$, implying}
\[
\mathcolorbox{yellow}{\lim\limits_{n\rightarrow\infty} n \prob{n^{-1} \bm{W}\in A} = \mu(A),}
\]
\hl{where $\mu$ is termed the exponent measure. $\mu$ has the homogeneity property $\mu(tA) = t^{-1}\mu(A)$. Letting $\rho = \|W\|_p, p>0$ and $\bm{\theta} = \bm{W}/\rho$, define  $\Psi(B) = \mu(\{\bm{w} : \rho>1, \bm{\theta} \in B\})$, which is referred to as the angular measure. After some manipulations, we obtain that}
\begin{equation}
\label{exponent-mu}
\mathcolorbox{yellow}{\lim\limits_{r\rightarrow\infty}\text{Pr}\left[\bm{\theta}\in A \rvert \rho>r\right] =\frac{\Psi(A)}{\Psi({\mathbb S}_p^{d-1})}.}
\end{equation}
\hl{Thus, a model for the exponent measure induces a model for the limiting distribution conditional on the observations being above a threshold defined with respect to their $p$-norm. The constraint that all marginals of $\mu$ correspond to a standard Pareto distribution leads to the so called moment constraints on $\Psi$. Inference for the limiting distribution of the exceedances needs to account for the normalizing constant in Equation~(\mbox{\ref{exponent-mu}}), as well the moment constraints. In addition, the moment constraints can impose unrealistic symmetries in the data distribution}

\section{Estimation of the angular measure\label{sec:methodology}}

Our approach to estimating the PoT distribution considers two steps: First we estimate the shape and scale parameters for the multivariate Pareto distribution, using the univariate marginals; Then we focus on the dependence structure in extreme regions by proposing a flexible model for the distribution of $\bm{V}$. Consider a collection of observations  $\bm{w}_i \in {\mathbb R}^d, i = 1, \ldots, n$ that exhibit extreme behavior.  We start by setting a large threshold $b_{t,\ell}$ for the  $\ell$-th marginal, $\ell = 1, \ldots,d$. Then, the distribution of the observations for the $\ell$-th component, conditional on exceeding the threshold, can be approximated as $1 - (1 + \xi_\ell   (w_{i\ell} - b_{t,\ell})/\sigma_\ell)_+^{1/\xi_\ell}$, where $(\cdot)_+$ indicates the positive part function.  We proceed by setting a threshold equal to the empirical $(1-1/t)$-quantile. Thus $b_{t,l}  = \hat{F}^{-1}_{\ell}(1 - 1/t)$.  We then use the approximate exceedance distribution to estimate $\xi_\ell$ and $\sigma_\ell$, for each $\ell$, using likelihood based methods. Then, in order to estimate the angular distribution, we use those estimates to standardize each of the marginals. The standardization yields
\begin{equation}
\label{eqn:standardization}
z_{i\ell} = \left(1 + \xi_{\ell}\frac{w_{i\ell} - b_{t,\ell}}{a_{\ell}}\right)_{+}^{1/\xi_{\ell}}\; .
\end{equation}
Note that $z_{i\ell}> 1$ implies that $w_{i\ell} > b_{t,\ell}$, meaning that the observation $\bm{w}_i$ is extreme in the $\ell$-th dimension. Thus, $r_i = \|\bm{z}_i\|_\infty > 1$ implies that at least one dimension has an extreme observation, and corresponds to a very extreme observation when $t$ is large. We focus on  the observations that are such that $r_i > 1$. These provide a sub-sample of the stardardized original sample. We define $\bm{v}_i = \bm{z}_i /r_i \in  \mathbb{S}_{\infty}^{d-1}$. These vectors are used for  the estimation of $\Phi$.

\subsection{Projected gamma family\label{subsec:projgamma}}
The $\mathcal{L}_p$-norm, for $p>0$ is  defined as
\begin{equation*}
\lVert \bm{x} \rVert_p =
\left({\textstyle\sum}_{\ell = 1}^d \lvert s_{\ell}\rvert^p\right)^{\frac{1}{p}},
\end{equation*}
The absolute and Euclidean norms are obtained for $p=1$ and $p=2$ respectively, and the $\mathcal{L}_{\infty}$ norm can be obtained as a limit,
\begin{equation*}
\| \bm{s} \|_{\infty}
= \lim\limits_{p\to\infty} \| \bm{s} \|
= \underset{\ell = 1}{\overset{d}{\bigvee}}s_{\ell}.
\end{equation*}

\st{To define an angular measure,  we are interested in the direction of vectors in ${\mathbb R}_{+}^d$, thus, we project them onto}
\begin{equation*}
{\mathbb S}_{p}^{d-1} = \left\lbrace \bm{s} : \bm{s} \in {\mathbb R}_{+}^{d}, \lVert \bm{s}\rVert_{p} = 1\right\rbrace,
\end{equation*}
\st{the $\mathcal{L}_p$-norm unit hypersphere.  Given $\bm{x}\in {\mathbb R}^d_+$ the projection onto ${\mathbb S}_{p}^{d-1}$ is given as $\bm{y} = \bm{x} / \lVert \bm{x}\rVert_p \in {\mathbb S}_{p}^{d-1}$. Observations on one hypersphere can be projected without loss of information onto another by dividing by the defining norm of the target hypersphere. In fact, for finite $p$, letting $y_d = \left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}$, the transformation}

\hl{To obtain a distribution on ${\mathbb S}_{p}^{d-1}=\{\bm{s} : \bm{s} \in {\mathbb R}_{+}^{d}, \| \bm{s}\|_{p} = 1\}$ we start with a vector in $\bm{x} \in {\mathbb R}^d_+$, and normalize it to obtain $\bm{y} = \bm{x}/\|\bm{x}\|_p \in {\mathbb S}_p^{d-1}$. A natural  distribution to consider in ${\mathbb R}^d_+$ is given by a product of independent univariate Gamma distributions. Let $\bm{ X} \sim \prod_{\ell = 1}^d\text{Ga}\left(X_{\ell}\mid\alpha_{\ell},\beta_{\ell}\right)$, $\alpha_\ell$ and $\beta_\ell$ are the shape and scale parameters, respectively. For any finite $p>0$, letting $ y_d = (1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p)^{1/p}$, the transformation}
\begin{equation}
\label{eqn:pnormt}
\begin{aligned}
T(x_1,\ldots,x_d) &= \left(\pnorm{\bm{x}}{p}, \frac{x_1}{\pnorm{\bm{x}}{p}}, \ldots , \frac{x_{d-1}}{\pnorm{\bm{x}}{p}}\right) \\
&= (r,y_1,\ldots,y_{d-1})
\end{aligned}
\end{equation}
is invertible with
\begin{equation}
\label{eqn:pnormtinv}
\begin{aligned}
&\hspace{-0.5cm}T^{-1}\left(r,y_1,\ldots,y_{d-1}\right)\\
& = \left(ry_1,\ldots,ry_{d-1}, r\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}\right).
\end{aligned}
\end{equation}
The Jacobian of the transformation takes the form
\begin{equation}
\label{eqn:pnormjac}
\begin{aligned}
r^{d-1}\left[\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}} \right. & \\
&\hspace{-2cm}+ \left. {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(1 - {\textstyle\sum}_{l=1}^{d-1} y_{\ell}^p\right)^{\frac{1}{p} - 1}\right].
\end{aligned}
\end{equation}
\st{Equations (\mbox{\ref{eqn:pnormt}}) -- (\mbox{\ref{eqn:pnormjac}}) can be used to define a density on ${\mathbb R}^d_+$, and then obtain a density on ${\mathbb S}_{p}^{d-1}$, for any finite $p>0$.}

\st{A natural  distribution to consider in ${\mathbb R}^d_+$ is given by a product of independent univariate Gamma distributions. Let $\bm{ X} \sim \prod_{\ell = 1}^d\text{Ga}\left(X_{\ell}\mid\alpha_{\ell},\beta_{\ell}\right)$, $\alpha_\ell$ and $\beta_\ell$ are the shape and scale parameters, respectively.}
\hl{The normalization provided by $T$ maps a vector in ${\mathbb R}_+^d$ onto
${\mathbb S}_p^{d-1} $. With a slight abuse of terminology we refer to it as a
projection.} Using Equations (\ref{eqn:pnormt}) -- (\ref{eqn:pnormjac}) we have the joint density
\begin{equation}
\label{joint}
\begin{aligned}
f(r,\bm{ y}) = &\prod_{\ell = 1}^{d}
\left[\frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell})}(ry_{\ell})^{\alpha_{\ell} - 1}
\exp\lbrace-\beta_{\ell}ry_{\ell}\rbrace\right] \\
&\hspace{0.5cm}\times r^{d-1}\left[y_d +
{\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p y_d^{1 - p}\right].
\end{aligned}
\end{equation}
Integrating out $r$ yields the resulting \emph{Projected Gamma} density
\begin{equation}
\label{eqn:projgamma}
\begin{aligned}
\text{PG}(\bm{ y}\mid\bm{ \alpha},\bm{ \beta}) &=
\prod_{\ell = 1}^d\left[\frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell})}
y_{\ell}^{\alpha_{\ell} - 1}\right]\\
&\hspace{0.5cm}\times \left[y_d +
{\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p y_d^{1 - p}\right]\\
&\hspace{0.5cm}\times \frac{\Gamma({\textstyle\sum}_{\ell = 1}^d\alpha_{\ell})}{\left({\textstyle\sum}_{\ell = 1}^d
\beta_{\ell}y_{\ell}\right)^{{\scriptstyle\sum_{\ell = 1}^d \alpha_{\ell}}}},
\end{aligned}
\end{equation}
defined for $\bm{y}\in {\mathbb S}_p^{d-1}$, and for any \hl{finite }$p>0$. To avoid identifiability problems when estimating the shape and scale parameters, we set $\beta_1 = 1$.  \cite{nunez2019} obtain the density in Equation (\ref{eqn:projgamma}) for $p=2$ as a multivariate distribution for directional data, using spherical coordinates.  For $\bm{ y}\in {\mathbb S}_1^{d-1}$, and $\beta_{\ell} = \beta$ for all $\ell$, the density in Equation~(\ref{eqn:projgamma}) corresponds to that of a Dirichlet distribution.

The projected gamma family is simple to specify and has very tractable computational properties. Thus, we use it as a building block for the angular measure $\Phi$ models. To build a flexible family of distributions in ${\mathbb S}_p^{d-1}$ we consider mixtures of projected gamma densities defined as
\begin{equation} \label{eqn:PGmix}
f(\bm{y}) = \int_\Theta \mathcal{PG}(\bm{y}\mid \bm{\theta}) dG(\bm{\theta}),
\end{equation}
where $\bm{\theta} = (\bm{\alpha}, \bm{\beta})$. Following a Bayesian non-parametric approach \citep{Ferguson74,Antoniak1974}\hl{\mbox{\citep{muller2015}}}, we assume that $G$ is drawn from a random measure. In particular, assuming a Dirichlet process prior for $G$, we have a hierarchical formulation of the mixture model that, for a vector of observations $\bm{y}_i$, is given by
\begin{equation}\label{eqn:dppg}
\begin{aligned}
\bm{y}_i &\sim \text{PG}(\bm{y}_i\mid \bm{\theta}_i)\\
\bm{\theta}_i &\sim G
\end{aligned}
~\hspace{1cm}
\begin{aligned}
G &\sim \mathcal{DP}(\eta, G_0),
\end{aligned}
\end{equation}
where $\mathcal{DP}$ denotes a Dirichlet process, $\eta$ is the precision parameter, and $G_0$ is the centering distribution.

Unfortunately, in the limit when $p\rightarrow \infty$, the \st{projection}\hl{normalizing} transformation is not differentiable. Thus, a closed form expression like Equation (\ref{eqn:projgamma}) for the projected gamma density on ${\mathbb S}_\infty^{d-1}$ is not available. Instead, we observe that for a sufficiently large $p$, $\mathbb{S}_p^{d-1}$ will approach $\mathbb{S}_{\infty}^{d-1}$.  With that in mind, our strategy consists of describing the angular distribution $\Phi$ using a sample based approach with the following steps: (i) Apply the transformation in Equation (\ref{eqn:standardization}) to the original data; (ii) Obtain the subsample of the standardized observations that satisfy $R>1$; (iii) Take a finite $p$ and project the observations onto ${\mathbb S}_p^{d-1}$; (iv) Fit the model in Equation (\ref{eqn:PGmix}) to the resulting data and obtain samples from the fitted model; (v) Project the resulting samples onto ${\mathbb S}_\infty^{d-1}$. For step (iv) we use a Bayesian approach that is implemented using a purposely developed Markov chain Monte Carlo sampler described in the next section.
\begin{figure}[ht]
\centering
\caption{The positive orthant of the $p$-norm sphere.}
\begin{subfigure}[b]{0.4\textwidth}
\includegraphics[width=\textwidth]{./images/p_sphere}
\caption{${\mathbb S}_p^1$ for $p = 1,\ldots, 10$\label{fig:psphere}}
\end{subfigure}
%
\begin{subfigure}[b]{0.4\textwidth}
\includegraphics[width=\textwidth]{./images/p_project}
\caption{Projection of data on ${\mathbb S}_{\infty}^1$ onto ${\mathbb S}_p^1$\label{fig:pproject}}
\end{subfigure}
\end{figure}

A measure that is used to characterize the strength of the dependence, in the tail, for two random variables $Z_1$ and $Z_2$, with marginal distributions $F_1$ and $F_2$ is given by \citep{coles2001}
\[
\chi_{12} = \lim\limits_{u\uparrow 1} \prob{F_1(Z_1)>u\mid F_2(Z_2)>u}.
\]
$\chi_{12}$ provides information about the distribution of extremes for the variable $Z_1$ given that $Z_2$ is very large.  When $\chi_{12}>0$, $Z_1$ and $Z_2$ are said to be asymptotically dependent, otherwise they are asymptotically independent. The following result provides the asymptotic dependence coefficient between two components of $\bm{Z}$ after PoT limit.
\begin{prop}\label{ppchi}
Suppose that $\bm{Z} = R\bm{V}$ with $R\sim Pa(1)$,
$\prob{V_\ell > 0} = 1$ and $\expect{V_\ell}$ exists, for $\ell=1, \ldots ,d$, then
\begin{equation}
\label{eqn:chi_ij}
\chi_{\jmath\ell} = \expect{\frac{V_\jmath}{\expect{V_\jmath}} \wedge \frac{V_\ell}{\expect{V_\ell}}}
\end{equation}
\end{prop}
{\em Proof:}
Denote as $F_\ell$ the marginal distribution of $Z_\ell$. To obtain $\chi_{\jmath\ell}$ we need $Pr(Z_\jmath>z_\jmath,Z_\ell>z_\ell)$, where $z_\ell = F_\ell^{-1}(u) = \expect{V_\ell}/(1 - u), \;\ell=1,\dots, d$.  Using the fact that $V_\ell>0, \forall \ell$ almost surely, we have that the former is equal to
\begin{equation*}
\begin{aligned}
&\hspace{-0.5cm}\prob{R>\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}}
= \expect{1\wedge\left(\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}\right)^{-1}}\\
&= \expect{\frac{V_\jmath}{z_\jmath}\wedge\frac{V_\ell}{z_\ell}}
= (1 - u)\expect{\frac{V_\jmath}{\expect{V_\jmath}}\wedge\frac{V_\ell}{\expect{V_\ell}}},
\end{aligned}
\end{equation*}
where the second identity is justified by the fact that $V_i$ is bounded and $z_i\rightarrow\infty$. The proof is completed by noting that $\prob{F_i(Z_i)>u} = 1 - u. \hfill\Box$

Equation \ref{eqn:chi_ij} implies that $\chi_{\jmath\ell}>0$, and so, no asymptotic independence is possible under our proposed model. For the analysis of extreme values it is of interest to calculate the multivariate conditional survival function. The following result provides the relevant expression, as a function of the angular measure.
\begin{prop}
Assume the same conditions of Proposition \ref{ppchi}. Let $\alpha \subset \{1, \ldots ,d\}$ be a collections of indexes. Then
\begin{equation} \label{eqn:condsurv}
\begin{aligned}
&\hspace{-0.5cm}\mathrm{Pr}\left[\bigcap_{\ell \in \alpha} Z_\ell > z_\ell \mid \bigcap_{\ell\not\in\alpha} Z_\ell > z_\ell\right] \\
&= \frac{\expect{\bigwedge_{k = 1}^d 1\wedge \frac{V_k}{z_k}}}{\expect{\bigwedge_{k \not\in\alpha}1\wedge\frac{V_k}{z_k}}}.
\end{aligned}
\end{equation}
\end{prop}
The proof uses a similar approach to the proof of Proposition \ref{ppchi}.

\hl{Equations \mbox{\ref{eqn:chi_ij}} and \mbox{\ref{eqn:condsurv}} provide relevant tools for inference on the tail behavior of the joint distribution of the observations. The expressions can be readily calculated within a sample-based inferential approach like the one considered in the following section.}

\subsubsection{Inference for the projected Gamma mixture model}
In a sample-based inference approach, for a given iteration the Dirichlet process mixture model groups observations into stochastically assigned clusters, where members of a cluster share distributional parameters\hl{ \mbox{\citep{muller2015,ascolani2022}}}. Building out the methods of inference for Equation~(\ref{eqn:dppg}), let $n_j^{-i}$ be the number of observations in cluster $j$ not including observation $i$.  Let $J^{-i}$ be the number of extant clusters, not including any singleton \st{that} \hl{containing} observation $i$ \st{is in}. Under this model, the probability of cluster membership for a given observation is proportional to
\begin{equation*}
\text{Pr}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\\ % \hspace{0.5cm} \\ % &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
\eta\int\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)dG_0(\bm{\alpha}_j,\bm{\beta}_j),%\hspace{0.5cm} % &\text{~for~}j = J^{-i} + 1.
\end{cases}
\end{equation*}
where the top case is iterating over extant clusters $j = 1,\ldots, J^{-i}$, and the bottom case is for a \emph{new} cluster. If $G_0$ is not a conjugate prior for the kernel density, the integral in the above formula may not be available in closed form. We sidestep this using Algorithm 8 from \cite{neal2000}: by Monte Carlo integration, we draw $m$ candidate clusters, $\bm{\alpha}_j,\bm{\beta}_j$ for $j = J^{-i} + 1,\ldots, J^{-i} + m$ from $G_0$. Then, we sample the cluster indicator $\gamma_i$ from extant or candidate clusters, where the probability of cluster membership is proportional to
\begin{equation}
\text{Pr}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\\ % \hspace{0.5cm}  % &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
\frac{\eta}{m}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right). % \hspace{0.5cm} &\text{~for~}j \in \lbrace J^{-i} + 1,\ldots, J^{-i} + m\rbrace.
\end{cases}
\end{equation}
Again, the top case is iterating over extant clusters, and now the bottom case is iterating over new \emph{candidate} clusters.  If a candidate cluster is selected, then $\gamma_i = J = J^{- i} + 1$, and the associated cluster parameters are saved.

A key feature of the the projected Gamma distribution is its computational properties. We augment $\mathcal{PG}(\bm{y}_i\mid\bm{\alpha}_i,\bm{\beta}_i) $ by introducing a latent radial component $r_i$, for each observation. Using Equation \eqref{joint} we observe that the full conditional of $r_i$ is easy to sample from, as it is given as
\begin{equation}
r_i\mid\bm{\alpha}_i,\bm{\beta}_i \sim \mathcal{G}\left(r_i \bigmid \sum_{\ell = 1}^d\alpha_{i\ell}, \sum_{\ell = 1}^d\beta_{\ell} y_{i\ell}\right).
\end{equation}
Moreover,  the full conditional for $\bm{\alpha}_j,\bm{\beta}_j$ is then proportional to
\begin{equation}
\begin{aligned}
&\hspace{-1cm}f(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{Y},\bm{r},\bm{\delta},\ldots) \\
&\propto \prod_{i:\gamma_i = j}\prod_{\ell = 1}^d\mathcal{G}\left(r_iy_{i\ell}\mid\alpha_{j\ell},\beta_{j\ell}\right) \\
&\hspace{0.5cm} \times dG_0(\bm{\alpha}_j,\bm{\beta}_j).
\end{aligned}
\end{equation}
Note that the ordering of the products can be reversed in the likelihood, indicating that given $\bm{r}$, the likelihood becomes separable by dimension.  We first consider a centering distribution given by a product of independent Gammas:
\begin{equation}
\begin{aligned}
&\hspace{-0.3cm}G_0(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{\xi},\bm{\tau},\bm{\zeta},\bm{\sigma}) \\
&= \prod_{\ell = 1}^d\mathcal{G}(\alpha_{j\ell}\mid \xi_{\ell},\tau_{\ell})\times\prod_{\ell = 2}^d\mathcal{G}(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}).
\end{aligned}
\end{equation}
This model is completed with independent Gamma priors on $\xi_{\ell}$, $\tau_{\ell}$, $\zeta_{\ell}$, $\sigma_{\ell}$.  We also assume a Gamma prior on $\eta$, that is updated via the procedure outlined in \cite{escobar1995}.  We refer to this model as the \emph{projected gamma--gamma} (PG--G) model.  An advantage of the PG--G model is that, thanks to conjugacy, the rate parameters can easily be integrated out for inference on $\bm{\alpha}_j$.  Then, the full conditional for $\alpha_{j\ell}$ takes the form
\begin{equation}
\label{eqn:alphalupdate}
\begin{aligned}
&\hspace{-0.5cm}\pi(\alpha_{j\ell}\mid \bm{r},\bm{Y},\bm{\gamma},\xi_\ell,\tau_\ell,\zeta_\ell,\sigma_\ell) \\
&\propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i\ell}\right)^{\alpha_{j\ell} - 1}\alpha_{j\ell}^{\xi_\ell - 1}e^{-\tau_\ell \alpha_{j\ell}}}{\Gamma^{n_j}(\alpha_{j\ell})} \\
&\hspace{0.5cm}\times \frac{\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
\end{aligned}
\end{equation}
for $\ell = 2,\ldots,d$.  For $\ell = 1$, as $\beta_{1} := 1$, the full conditional takes the simpler form
\begin{equation}
\label{eqn:alpha1update}
\begin{aligned}
&\hspace{-0.5cm}\pi(\alpha_{j1}\mid\bm{r},\bm{Y},\bm{\gamma},\xi_1,\tau_1) \\
&\propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}\alpha_{j1}^{\xi_1 - 1}e^{-\tau_1\alpha_{j1}}}{\Gamma^{n_j}(\alpha_{j1})}.
\end{aligned}
\end{equation}
Samples of $\alpha_{j\ell}$ can thus be obtained using a Metropolis step. In our analysis, we first transform $\alpha_{j\ell}$ to the log scale, and use a normal proposal density.  The full conditional for $\beta$ is
\begin{equation}
\label{eqn:betafc}
\begin{aligned}
&\hspace{-0.5cm}\beta_{j\ell}\mid\bm{r},\bm{Y},\alpha,\zeta_{\ell},\sigma_{\ell} \\
&\sim \mathcal{G}\left(\beta_{j\ell}\bigmid n_j\alpha_{j\ell} + \zeta_\ell, \sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right),
\end{aligned}
\end{equation}
for $\ell = 2,\ldots, d$.  Updating $\beta_{j\ell}$ is done via a Gibbs step.  The hyper-parameters $\xi_{\ell},\tau_{\ell},\zeta_{\ell},\sigma_{\ell}$ follow similar gamma-gamma update relationships.  We also explore a restricted form of this model, where $\beta_{\ell} = 1$ for all $\ell$.  Under this model, we use the full conditional in Equation~(\ref{eqn:alpha1update}) for all $\ell$, and omit inference on $\bm{\zeta},\bm{\sigma}$.  We refer to this model as the \emph{projected restricted gamma--gamma} (PRG--G) model.

The second form of centering distribution we explore is a multivariate log-normal distribution on the shape parameters $\bm{\alpha}_j$, with independent gamma $\beta_{j\ell}$ rate parameters.
\begin{equation}
\begin{aligned}
&\hspace{-0.5cm}G_0\left(\bm{\alpha}_j,\bm{\beta}_j\mid\bm{\mu},\bm{\Sigma},\zeta,\sigma\right) \\
&=\mathcal{LN}\left(\bm{\alpha}_j\mid\bm{\mu},\bm{\Sigma}\right)\times\prod_{\ell = 2}^d\mathcal{G}\left(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}\right).
\end{aligned}
\end{equation}
This model is completed with a normal prior on $\bm{\mu}$, an inverse Wishart prior on $\bm{\Sigma}$, and Gamma priors on $\zeta_{\ell}$ and $\sigma_{\ell}$, and $\eta$.  This model is denoted as the \emph{projected gamma--log-normal} (PG--LN) model.  We also explore a restricted Gamma form of this model as above, where
$\beta_{\ell} = 1$ for all $\ell$.  This is denoted as the \emph{projected restricted gamma--log-normal} (PRG--LN) model.  Updates for $\bm{\alpha}$ can be accomplished using a joint Metropolis step, where $\beta_{j\ell}$ for $\ell = 2,\ldots,d$ have been integrated out of the log-density.  That is,
\begin{equation*}
\begin{aligned}
&\hspace{-0.5cm}\pi(\bm{\alpha}_j\mid\bm{Y},\bm{r},\bm{\delta},\bm{\mu},\bm{\Sigma},\bm{\zeta},\bm{\sigma})\\
&\propto \exp\left\lbrace-\frac{1}{2}(\log\bm{\alpha}_j - \mu)^T\bm{\Sigma}^{-1}(\log\bm{\alpha}_j - \mu)\right\rbrace \\
&\hspace{0.5cm}\times\frac{1}{\prod_{\ell = 1}^d\alpha_{j\ell}}\times \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}}{\prod_{\ell = 1}^d\Gamma^{n_j}(\alpha_{j\ell})}\\
&\hspace{0.5cm}\times \prod_{\ell = 2}^d\frac{\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
\end{aligned}
\end{equation*}
The inferential forms for $\beta_{j\ell}$ and its priors are the same as for PG--G.  The normal prior for $\mu$ is conjugate for the log-normal $\bm{\alpha}_j$, and can be sampled via a Gibbs step.  Finally, the inverse Wishart prior for $\Sigma$ is again conjugate to the log-normal $\bm{\alpha}_j$, implying that it can also be sampled via a Gibbs step.

\hl{To effectively explore the sample space with a joint Metropolis step, as well as to speed convergence, we implement a parallel tempering algorithm\mbox{\citep{earl2005}} for the log-normal models. This technique runs parallel MCMC chains at ascending temperatures. That is, for chain $s$, the posterior density is exponentiated to the reciprocal of temperature $t_s$.  For the \emph{cold} chain, $t_1 := 1$.  Let $E_s$ be the log-posterior density under the current parameter state for chain $s$.  Then states between chains $r,s$ are exchanged via a Metropolis step with probability}
\[
\mathcolorbox{yellow}{\alpha_{rs} = \min\left[1, \exp\left\lbrace(t_{r}^{-1} - t_{s}^{-1})(E_r - E_s)\right\rbrace\right].}
\]
\hl{Higher temperatures serve to \emph{flatten} the posterior distribution, meaning hotter chains have a higher probability of making a given transition, or will make larger transitions.  As such, they will more quickly explore the parameter space, and share information gained through state exchange.}

\section{Scoring criteria for distributions in ${\mathbb S}_\infty^{d-1}$\label{sec:evaluation}}
In order to assess and compare the estimation of a distribution on ${\mathbb S}_\infty^{d-1}$ we consider the theory of proper scoring rules developed in \cite{gneiting2007}. As mentioned in Section~\ref{subsec:projgamma}, our approach does not provide a density on ${\mathbb S}_{\infty}^{d-1}$, restricting our ability to construct model selection criteria to sample-based approaches.  \st{We consider two scoring methods:  \emph{posterior predictive loss} (PPL) introduced in \mbox{\cite{gelfand1998}}, and \emph{energy score} (ES) introduced in \mbox{\cite{gneiting2007}}.}\hl{To this end, we employ the \emph{energy score} criterion introduced therein.}

\st{The posterior predictive loss criterion is introduced in \mbox{\cite{gelfand1998}}.  When we assume a squared error loss function, then for the $i$th observation, the posterior predictive loss criterion is computed as}\begin{equation}
\label{eq:ppl}
\begin{aligned}
&\hspace{-0.5cm}\cancel{S_{\text{PPL}}^k\left(P,x_i\right)} \\
&\cancel{=\text{Var}_P\left[X_i\right] + \frac{k}{k+1}\left(\text{E}_P\left[X_{i}\right] - x_i\right)^2}
\end{aligned}
\end{equation}
\st{where $X_i$ is a random variable from the posterior predictive distribution for $x_i$.  The scalar $k$ is a weighting factor that scales the importance of goodness of fit relative to precision.  In fact a small $\text{Var}(X_{i})$ indicates high precision, and a small $(\text{E}[X_{i}] - x_{i})^2$  indicates a good fit.  In our analysis, we take the limit as $k\to\infty$, weighting both equally. Overall, the smaller the score, the better.  Note that the PPLC score is defined for a univariate distribution.  As we are dealing with a multivariate distribution, we will be taking the posterior predictive loss summed over all dimensions.  That is,}
\begin{equation}
\label{eq:ppl2}
\begin{aligned}
&\hspace{-0.5cm}\cancel{S_{\text{PPL}}\left(P,\bm{x}_i\right)} \\
&\cancel{=\frac{1}{d}\sum_{\ell = 1}^d\left[\text{Var}_P\left[X_{i\ell}\right] + \left(\text{E}_P\left[X_{i\ell}\right] - x_{i\ell}\right)^2\right].}
\end{aligned}
\end{equation}
\st{We report the average $S_{\text{PPL}}$ taken from all observed data.}

\st{We focus on \emph{energy scores}} \hl{The energy score criterion} defined for a general probability distribution $P$, with finite expectation, \hl{is developed} as
\begin{equation}
\label{eq:es}
\begin{aligned}
&\hspace{-0.5cm}\text{S}_{\text{ES}}\left(P, \bm{x}_i\right) \\
&=\text{E}_p \left[g\left(\bm{X}_i, \bm{x}_i\right)\right] -
\frac{1}{2}\text{E}_p \left[g\left(\bm{X}_i,\bm{X}_i^{\prime}\right)\right],
\end{aligned}
\end{equation}
where $g$ is a kernel function. The score defined in Equation \eqref{eq:es} can be evaluated using samples from $P$, with the help of the law of large numbers.   Moreover, Theorem~4 in \cite{gneiting2007}, states that if $g(\cdot,\cdot)$ is a negative   definite kernel, then $\text{S}(P,\bm{x})$ is a \emph{proper} scoring rule.  Recall that a real valued function $g$ is a negative definite kernel if it is symmetric in its arguments, and $\sum_{i=1}^n\sum_{j=1}^na_ia_jg(x_i,x_j)\leq 0$ for all positive integers $n$, and any collection $a_1,\ldots,a_n\in{\mathbb R}$ such that  $\sum_{i = 1}^na_i = 0$.

In a Euclidean space, these conditions are satisfied by the Euclidean distance \hl{\mbox{\citep{berg1984}}}. However, \hl{for observations on different faces of }\st{in } ${\mathbb S}_{\infty}^{d-1}$, \hl{the }Euclidean distance will \st{tend to} under-estimate the \hl{geodesic distance, the }actual distance required to travel between \hl{the }two points.  Let
\begin{equation*}
{\mathbb C}_{\ell}^{d-1} = \lbrace \bm{x} : \bm{x} \in {\mathbb S}_{\infty}^{d-1}, x_{\ell} = 1\rbrace
\end{equation*}
comprise the $\ell$th \emph{face}.  For points on the same face, \hl{the }Euclidean distance corresponds to the length of the shortest possible path in ${\mathbb S}_{\infty}^{d-1}$.  For points on different faces, the Euclidean distance is a lower bound for that length.

For a finite $p$, the shortest connecting path between two points in ${\mathbb S}_p^{d-1}$ is the minimum geodesic; its length satisfying the definition of a distance.  Thus its length can be used as a negative definite kernel for the purpose of defining an energy score. Unfortunately as $p\to\infty$ the resulting surface ${\mathbb S}_{\infty}^{d-1}$ is not differentiable, implying that routines to calculate geodesics are not readily available.  However, as ${\mathbb S}_{\infty}^{d-1}$ is a portion of a $d$-cube, we can borrow a result from geometry \citep{pappas1989} stating that the length of the shortest path between two points on a geometric figure corresponds to the length of a straight line drawn between the points on an appropriate unfolding, rotation, or \emph{net} of the figure from a $d$-dimensional to a $d-1$-dimensional space.  The optimal net will have the shortest straight line between the points, as long as that line is fully contained within such net. As ${\mathbb S}_{\infty}^{d-1}$ has $d$ faces---each face pairwise adjacent, there are $d!$ possible nets.  However, we are only interested in nets that begin and end on the source and destination faces respectively\st{.  That reduces}\hl{, reducing} the number of nets under consideration to $\sum_{k = 0}^{d-2}\binom{d-2}{k}$.  This is still computationally burdensome for a large number of dimensions.\hl{  However, we can efficiently establish an upper bound on the geodesic length.  We use this upper bound on geodesic distance as the kernel function for the energy score.}

\st{To reduce the computations needed to calculate the energy score we define a kernel as in the following proposition.  Let $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-1}$, for $\ell, \jmath \in \{1, \ldots , d\}$. Define $g$ as}
\hl{To calculate the energy score we define the kernel}
\[
g(\bm{a},\bm{b}) = \min_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{
\pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
\]
\st{Then $g$  is a negative definite kernel.}

\st{{\em Proof:}}
\st{When $\bm{a}$ and $\bm{b}$ are on the same face, then $\ell = \jmath$, and $g$ is the Euclidean distance between $\bm{a}$ and $\bm{b}$, which is a  negative definite kernel.  When $\bm{a}$ and $\bm{b}$ are on separate faces, let}
\[\cancel{\bm{e} =
\argmin_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{
\pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.}
\]
\st{We have that $g(\bm{a}, \bm{b}) = g(\bm{b}, \bm{a})$ due to the uniqueness of $\bm{e}$ and the symmetry of the Euclidean distance. Additionally, given $n$ and  a collection $\alpha_1,\ldots,\alpha_n\in{\mathbb R}$ such that  $\sum_{i = 1}^n\alpha_i = 0$, then}
\begin{equation*}
    \begin{aligned}
    &\cancel{\hspace{-0.5cm}\sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j g(\bm{x}_1,\bm{x}_2)}\\
    &\cancel{= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j \bigg[\pnorm{\bm{e} - \bm{x}_1}{2} + \pnorm{\bm{x}_2 - \bm{e}}{2}\bigg]}\\
    &\cancel{= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{e} - \bm{x}_1}{2}}\\
    &\cancel{\hspace{1cm}+ \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{x}_2 - \bm{e}}{2}}\\
    &\cancel{\leq 0}
    \end{aligned}
    \end{equation*}
\st{as the Euclidean norm is a negative definite kernel.$\hfill\Box$}

\hl{where $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and
$\bm{b} \in {\mathbb   C}_{\jmath}^{d-1}$, for $\ell,\jmath\in \{1,\ldots, d\}$.
Evaluating $g$ as described  requires the solution of a $(d-2)$-dimensional optimization problem.  The following proposition provides a straightforward approach.}
\begin{prop}\label{prop:rot}
Let $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-1}$,
for $\ell, \jmath \in \{1, \ldots , d\}$.  For $\ell\neq \jmath$, the
transformation $P_{\jmath\ell}(\cdot)$ required to rotate the $\jmath$th face along the
$\ell$th axis produces the vector $\bm{b}^\prime$, where
\begin{equation}
\label{eqn:rotation}
\bm{b}^{\prime}_i = P_{\jmath\ell}(\bm{b}) =
\begin{cases}
b_{i} &\text{for }i\neq \jmath,\ell\\
1 &\text{for }i = \ell\\
2 - b_{\ell} &\text{for }i = \jmath
\end{cases}.
\end{equation}
Then $g(\bm{a},\bm{b}) = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}$.
\end{prop}
{\em Proof:}
Notice that for
$\bm{c} \in {\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}$,
$\pnorm{\bm{b} - \bm{c}}{2} =  \pnorm{\bm{b}^{\prime} - \bm{c}}{2}$.\st{ Then, recalling the definition of $\bm{e}$ in the proof of Proposition \mbox{\ref{prop:g}}, we}\hl{We then} have that
\begin{equation*}
\begin{aligned}
g(\bm{a},\bm{b}) = &\min_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{
\pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\} \\
= &\min_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{
\pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}^{\prime}-\bm{c}}{2} \right\}\\
= & \pnorm{\bm{a} - \bm{b}^{\prime}}{2}.
\end{aligned}
\end{equation*}
The last equality is due to the fact that $\bm{a}$ and $\bm{b}'$ belong to the same hyperplane. $\hfill\Box$

Using the rotation in Proposition \ref{prop:rot} we obtain the following result.
\begin{prop}\label{prop:g}
g is a negative definite kernel.
\end{prop}
{\em Proof:} For a given $n$ consider an arbitrary set of points
$\bm{a}_1, \ldots , \bm{a}_n \in {\mathbb S}_\infty^{d-1}$, and
$\alpha_1, \ldots , \alpha_n \in {\mathbb R}$, such that $\sum_{i=1}^n
\alpha _i = 0$. Then
\[
\sum_{i,\jmath} \alpha_i \alpha_\jmath g(\bm{a}_i, \bm{a}_\jmath)
= \sum_{i,\jmath} \alpha_i \alpha_\jmath
\|\bm{a}_i - \bm{a}_\jmath^\prime\|_2 \leq 0,
\]
where $\bm{a}_\jmath^\prime$ is defined as in Proposition \ref{prop:rot}.
The last equality holds as $\|\bm{x} - \bm{x}^\prime\|_2, \bm{x},
\bm{x}^{\prime} \in {\mathbb R}^d$ is negative definite
\citep{gneiting2007}$\hfill\Box$

Proposition \ref{prop:rot} provides a computational efficient way to evaluate the proper scoring rule $S_{\text{ES}}$ defined on ${\mathbb S}_\infty^{d-1}$, for each observation. For the purpose of model assessment and comparison, we report the average $S_{\text{ES}}$ taken across all observed data, and notice that the smaller the score, the better.

\begin{figure*}[htb]
\centering
\caption{Average energy score rise over baseline (on $\mathbb{S}_{\infty}^{d-1}$) for various
models fitted to simulated data, with ascending count of mixture components (indicated by plot
heading) and number of dimensions (indicated by horizontal axis).  Note that pairwise betas is
a moment-restricted model.\label{fig:simpples}}
\includegraphics[height=3in, width = \textwidth]{./images/sim_es_rise}
\end{figure*}

\section{Data illustrations\label{sec:results}}
\hl{We apply the aforementioned models to simulated angular data. We then consider the analysis of atmospheric data. To tackle the difficult problem of assessing the convergence an MCMC chain for a large-dimensional model, we monitor the log-posterior density. In all the examples considered, MCMC samples produced stable traces of the log-posterior in less than 40 thousand  iterations. We use that as a burn-in, and thereafter sample 10 thousand additional iterations. We then thin the chain by retaining one every ten samples, to obtain 1000 total samples. These are used to generate samples from the posterior predictive densities. We used two different strategies to implement the MCMC samplers. For the models whose DP prior is centered around a log-normal distribution we used parallel tempering. This serves to overcome the very slow mixing that we observed for these cases. The temperature ladder was set as $t_s = 1.3^{s}$,  for $s \in \lbrace0,1,\ldots,5\rbrace$. This was set empirically in order to produce acceptable swap probabilities both for the simulated data, and real data. Parallel tempering produces chains with good mixing properties, but has a computational cost that  grows linearly with the number of temperatures. Thus, for the gamma-centered models, we used a single chain. We leverage the fast speed of each iteration, to obtain a large number of samples, that are appropriately thinned to deal with a mild autocorrelation. In summary, the strategy for log-normal centered models is based on a costly sampler with good mixing properties. The strategy for the gamma-centered models is based on a cheap sampler that can be run for a large number of iterations.}

\hl{Our hyperprior parameters are set as follows: for the gamma-centered models (PG--G, PRG--G), the shape parameter for the centering distribution~$\xi_{\ell}\sim \mathcal{G}\left(1,1\right)$, and rate parameter~$\tau_{\ell}\sim\mathcal{G}\left(2,2\right)$. For the log-normal centered models (PG--LN, PRG--LN), the centering distribution's log-mean $\bm{\mu}\sim\mathcal{N}_d\left(0,\bm{I}_d\right)$, and covariance matrix~$\Sigma\sim\mathcal{IW}\left(d + 10, (d+10)\bm{I}_{d}\right)$.  These values are intended such that draws from the prior for $\Sigma$ will weakly tend towards the identity matrix. For models learning rate parameters $\beta_{j\ell}$ (PG--G, PG--LN), the centering distribution's shape parameter $\zeta\sim\mathcal{G}\left(1,1\right)$ and rate parameter $\sigma\sim\mathcal{G}\left(2,2\right)$.  The choice of the $\mathcal{G}(2,2)$ for rate parameters places little mass near 0, in order to draw estimates for the value away from 0 for numerical stability.}

\subsection{Simulation Study\label{subsec:simulated}}
To evaluate our proposed approach for angular measure estimation we consider simulated datasets on $\mathbb{S}_{\infty}^{d-1}$, for values of $d$ ranging between \st{3 and 20}\hl{2 and 32}.  We generated each dataset as a mixture of \st{projected gammas}\hl{multivariate log-normal distributions, projected onto $\mathbb{S}_{\infty}^{d-1}$}\st{with the number of mixture components ranging between 3 and 12}. The generation procedure is detailed in Algorithm~\mbox{\ref{algo:simulated}}. \hl{We produced ten replicates of each configuration. We consider two gamma-centered and two log-normal centered DP mixture models, with and without restrictions in each case. To perform a comparative analysis we fitted the pairwise betas model proposed in\mbox{\cite{COOLEY2010}}}. 
\hl{We chose this model for comparison as it is implemented in the readily available package \mbox{\texttt{BMAmevt}} in R~\mbox{\citep{BMAmevt}}. \mbox{\texttt{BMAmevt}} provides samples from the predictive posterior distribution.}  
\hl{These are needed for the calculation of the energy scores that are at the basis of our comparison. In addition, \mbox{\texttt{BMAmevt}} can be fitted to moderately large multivariate observations.}
\hl{For the DP mixture models, the data are projected onto $\mathbb{S}_{10}^{d-1}$. For the other two models they are projected on $\mathbb{S}_1^{d-1}$.   We sampled each model for 50,000 iterations, dropping the first 40,000 as burn-in, and thinning to keep every 10th iteration after.  These settings were intended to provide a consistent sampling strategy that would work with every model, even if inefficient for some.}

\begin{algorithm}[ht]  % algorithm using algorithm package
\caption{Simulated Angular Dataset Generation Routine\label{algo:simulated}.
$\mu_j$, $\Sigma_j$ are the parameters of the mixture component distribution;
$\pi$ is the probability vector assigning weight mixture components; $\delta_i$
is the mixture component identifier for each simulated observation.}
\begin{algorithmic}
\For{$n_{\text{iter}}$ in $[1,\ldots,10]$}
\For{$n_{\text{mix}}$ in $[1, 2, 4, 8]$}
\For{$j$ in $1,\ldots,n_{\text{mix}}$}
\State Generate $\mu_{j} \sim \mathcal{N}_{32}\left(\bm{0},\bm{I}\right)$
\State Generate $\Sigma_{j}\sim\mathcal{IW}_{32}\left(70,70 \bm{I}\right)$
\EndFor
\State Generate $\pi\sim\text{Dirichlet}(\bm{10}_{n_{\text{mix}}})$
\For{$i$ in $1,\ldots,1000$}
\State Generate $\delta_i \sim \text{Categorical}(\pi)$
\State Generate $\bm{X}_i \sim \mathcal{LN}\left(\mu_{[\delta_i]},\Sigma_{[\delta_i]}\right)$
\EndFor
\For{$n_{\text{col}}$ in $[2,4,8,16,24,32]$}
\State Project columns 1 to $n_{\text{col}}$ of ${\bf X}$ onto $\mathcal{S}_{\infty}^{n_{\text{col}} - 1}$ and save.
\EndFor
\EndFor
\EndFor
\end{algorithmic}
\end{algorithm}

\st{Each one of the shape parameters $\alpha$ were generated $\alpha = \alpha_0 + \alpha_1$ where $\alpha_0 \sim \text{Unif}(\alpha_0\mid 0,4)$, $\alpha_1\sim \text{Gamma}(\alpha_1\mid 1,1)$.  The scale parameters $\beta$ were obtained as $\beta\sim\text{Unif}(\beta\mid 0.25, 2.5)$.}

\st{Figure~\mbox{\ref{fig:simpples}}, shows two sets of panels that illustrate the comparison between four different models used to fit the simulated data. We produced 16 different data sets, one for each of four different numbers of mixture components and four different dimensions. We considered three different models: DP mixtures of projected gamma; Projected restricted gamma; and projected restricted gamma with a multivariate log-normal prior. We calculated $S^\infty_{\text{PPL}}$ and $S_{\text{ES}}$ using predictive samples from each of fitted models. To provide a comparative baseline, we calculated the scores using samples generated directly using the same model and parameters that produced the simulated observations. The results are denoted as \emph{Gen} in the figure. We observe that the projected restricted Gamma model dominates in most situations, but the difference between projected restricted Gamma and the other gamma-based models tends to shrink as the number of dimensions increases. Alternatively, that difference appears to grow as the number of mixture components increases. The results from both scoring criteria are comparable.}

\hl{Figure~\mbox{\ref{fig:simpples}} shows the average rise over baseline in energy score as calculated on $\mathbb{S}_{\infty}^{d-1}$ using the kernel metric described in Proposition~\mbox{\ref{prop:rot}}, for models trained on simulated data.  After training a model, a posterior predictive dataset is generated, and the energy score is calculated as a Monte Carlo approximation of Equation~\mbox{\ref{eq:es}}.  In our analysis, after burn-in and thinning, we had 1,000 replicates from the posterior distribution, and generated 10 posterior predictive replicates per iteration. The \emph{baseline} value is the energy score of a new dataset from the same generating distribution as the training dataset, evaluated against the training dataset.  For the simulated data, we observe that the projected gamma models dominate the other two options considered, regardless of the choice of centering distribution.  The  projected restricted gamma models with a multivariate log-normal centering distribution appears to be dominated by the models based on the alternative centering distributions. Moreover, the performance deteriorates with the increase in dimensionality. Additionally, models centered around the log-normal distribution incur in the computational cost of multivariate normal evaluation and parallel tempering, taking approximately six times longer to sample relative to the gamma models.  We also note that the computational cost  of the pairwise betas model grows combinatorially, with a sample space of dimension $\binom{d}{2} + 1$. In our testing, \texttt{BMAmevt} is substantially faster than any of our proposed DP mixture models for low dimensions, however for examples with high number of dimensions, the computational time for \texttt{BMAmevt} was comparable or greater than that for DP-PG-G.  We compare computing times in our data analysis in Table~\mbox{\ref{tab:sampletime}}.}

\subsection{Integrated Vapor Transport\label{subsec:ivt}}
The \emph{integrated vapor transport} (IVT) is a two component vector that tracks the flow of the total water volume in a column of air over a given area \citep{ralph2017}.  IVT is increasingly used in the study of atmospheric rivers because of its direct relationship with orographically induced precipitation \citep{neiman2009water}. Atmospheric rivers (AR) are elongated areas of high local concentration of water vapor in the atmosphere that transport water from the tropics around the world. AR can cause extreme precipitation,  something that is usually associated with very large values of the IVT magnitude over a whole geographical area. In spite of this, AR are fundamental for the water supply of areas like California. Thus the importance of understanding the extreme behavior of IVT, including extreme tail dependence. We consider datasets that correspond to IVT estimated at two different spatial resolutions. The coarse resolution dataset is obtained from the European Centre for Medium-Range Weather Forecasts (ECMWF) Interim reanalysis (ERA-Interim) \citep{berrisford2011atmospheric,dee2011era}. The high resolution dataset corresponds to the latest ECMWF observational product, ERA5 \citep{hersbach2020era5}.

\begin{figure}[tb]
\caption{Grid cell locations for ERA-Interim (left) and ERA5 (right).\label{fig:gridlocs}}
\centering
\begin{minipage}{0.25\textwidth}
\centering
\includegraphics[width=0.99\linewidth]{./images/erai_grid}
\end{minipage}%
\begin{minipage}{0.25\textwidth}
\centering
\includegraphics[width=0.99\linewidth]{./images/era5_grid}
\end{minipage}
\end{figure}

Our data correspond to daily average values for the IVT magnitude along the coast of California.  The ERA-Interim data used covers the time period 1979 through 2014 (37 years) omitting leap days, and eight grid cells that correspond to the coast of California.  The ERA5 data cover the time period 1979 through 2019 (42 years) with the same restriction, and  47 grid cells for the coast of California.  This gives us the opportunity to illustrate the performance of our method in multivariate settings of very different dimensions. Figure~\ref{fig:gridlocs} provides a visual representation of the area these grid cells \st{represent}\hl{cover}.

\begin{algorithm}[htb]
\caption{Data preprocessing to isolate and transform data exhibiting extreme behavior.  $r_i$ represents the radial component, and $\bm{v}_i$ the angular component.  The declustering portion is relevant for data correlated in time.\label{algo:processing}}
\begin{algorithmic}
\For{$\ell = 1,\ldots,d$}
\State Set $b_{t,\ell} = \hat{F}_{\ell}^{-1}\left(1 - \frac{1}{t}\right)$.
\State With $\bm{ x}_{\ell} > b_{t,\ell}$, fit $a_{\ell}$, $\xi_{\ell}$ via MLE according to generalized Pareto likelihood.
\EndFor
\For{$i = 1,\ldots,n$}
\State Define $z_{i,\ell} = \left(1 + \xi_{\ell}\frac{x_{i,\ell} - b_{t,\ell}}{a_{\ell}}\right)_{+}^{1/\xi_{\ell}}$; $\;\;\;$ then $r_i = \pnorm{\bm{ z}_i}{\infty}$, $\;\;\bm{ v}_i = \frac{\bm{ z}_i}{\pnorm{\bm{ z}_i}{\infty}}$
\EndFor
\State Subset $\bm{ r},\bm{ v}$ such that $r_i \geq 1$
\If{declustering}
\For{$i = 1,\ldots,n$}
\State If $r_i \geq 1$ and $r_{i-1} \geq 1$, drop the lesser (and associated $v_i$) from data set.
\EndFor
\EndIf
\end{algorithmic}
\end{algorithm}

\begin{figure*}[ht]%{l}{0.45\textwidth}
\centering
\caption{Pairwise plots from ERA-Interim data after transformation and projection to ${\mathbb S}_{\infty}^{7}$.  Down the diagonal are marginal kernel densities, with two-dimensional histograms on the off-diagonal.  In those plots, red indicates a higher density.  All data is between 0 and 1.\label{fig:erai_data}}
\includegraphics[width=.7\linewidth]{./images/data_transformed}
\end{figure*}

Fitting our models to the IVT data requires some pre-processing. First, we subset the data to the rainy season, which in California runs roughly from November to March.  Following the approach described in Section \ref{sec:methodology} we estimate the shape and scale parameters of a univariate GP, in each dimension, using maximum likelihood. We set the threshold in each dimensions $\ell$ as $b_{t,\ell} = \hat{F}_{\ell}^{-1}(1 - t^{-1})$, where $\hat{F}$ is the empirical CDF and $t=20$, that corresponds to the $95$ percentile. We then use the transformation in Equation \eqref{eqn:standardization} to standardize the observations.  Dividing each standardized observation by its $\mathcal{L}_{\infty}$ norm, we obtain a projection onto $\mathcal{S}_{\infty}^{d-1}$. As the data correspond to a daily time series, the observations are temporally correlated.  For each group of consecutive standardized vectors $z_i$ such  that  $\inorm{z_i} > 1$, we retain only the vector with the largest $\mathcal{L}_{\infty}$ norm.  The complete procedure is outlined in Algorithm~\ref{algo:processing}.

After subsetting the ERA-Interim data to the rainy season we have $5587$ observations. After the processing and declustering described in Algorithm~\ref{algo:processing}, this number reduces to $511$ observations. A pairwise plot of the transformed data after processing and declustering is presented in Figure~\ref{fig:erai_data}.  From this, we note that the marginal densities display strong similarities, with a large spike near 0 and a small spike near 1. A value of 1 in a particular axis indicates that the standardized threshold exceedance was largest in that dimension.  The off-diagonal plots correspond to pairwise density plots.  We observe that some site pairs, such as $(1,2)$, $(7,8)$, and especially $(4,5)$ have the bulk of their data concentrated in a small arc along the $45^{\circ}$, while other site combinations such as $(3,6)$, $(2,7)$, or $(1,8)$ the data is split, favoring one side or the other of the $45^{\circ}$ line. For the ERA5 data, after subsetting we have $6342$ observations, which reduces to $532$ observations after processing and declustering. We fit the PG--G, PRG--G, PG--LN, and PRG--LN models to both datasets.

\begin{table}[htb]
\centering
\caption{Energy score criterion from fitted models against the IVT data. Lower is better.\label{tab:dev}}
\newlength{\templen}
\setlength\templen{\tabcolsep}
\setlength\tabcolsep{3.5pt}
\begin{tabular}{c|ccccc}
\toprule
Source & \begin{tabular}{@{}c@{}}Pairwise \\ Betas\end{tabular} & PG-G & PG-LN & PRG-G & PRG-LN\\
\midrule
\begin{tabular}{@{}c@{}}ERA- \\ Interim\end{tabular} & $0.8620$ & $0.8003$ & $0.7986$ & $\bm{0.7966}$ & $0.7970$\\
ERA5 & $2.0311$ & $1.6404$ & $1.5576$ & $\bm{1.4349}$ & $1.5051$\\
\bottomrule
\end{tabular}
\setlength\tabcolsep{\templen}
\end{table}

\begin{table}[htb]
\centering
\caption{Time to sample (in minutes) 50,000 iterations for various models\label{tab:sampletime}}
\setlength\templen{\tabcolsep}
\setlength\tabcolsep{3.5pt}
\begin{tabular}{c|ccccc}
\toprule
Source & \begin{tabular}{@{}c@{}}Pairwise \\ Betas\end{tabular} & PG-G & PG-LN & PRG-G & PRG-LN\\
\midrule
\begin{tabular}{@{}c@{}}ERA- \\ Interim\end{tabular} & $1.5$ & $16.3$ & $66.5$ & $14.8$ & $52.9$\\
ERA5 & $53.1$ & $19.4$ & $153.4$ & $24.6$ & $121.4$\\
\bottomrule
\end{tabular}
\setlength\tabcolsep{\templen}
\end{table}

Table~\ref{tab:dev} shows the values of the estimated \st{posterior predictive loss and }energy scores for the different models considered. We observe that, \st{as in}\hl{contrary to} the \st{case of}\hl{results in} the simulation study \st{shown} in Figure\st{s}~\ref{fig:simpples}, the preferred model is the projected restricted gamma models{\st .  Penalization for the unrestricted gamma model is stronger for the real data than for the simulated data. We also observe that the restricted gamma model with the log-normal prior performs significantly better than the restricted gamma model with the gamma prior on the higher dimensional data. This is the opposite of the result obtained in the simulation example.}{\hl, though for the lower-dimensional ERA-Interim data, all models perform comparably.  Table~\ref{tab:sampletime} shows the computing times needed to fit the different models to the two datasets.  We see the effect of dimensionality on the various models; for gamma centered models it grows linearly; for the log--normal centered model, it will grow superlinearly as matrix inversion becomes the most costly operation.  For \texttt{BMAmevt}, its parameter space grows combinatorically with the number of dimensions, and thus so does computational complexity and sampling time.}

\begin{figure*}[htb]
\centering
\caption{Pairwise extremal dependence coefficients for IVT data using the PRG--G model.\label{fig:chi_ij}}
\includegraphics[width=0.9\textwidth]{./images/chi_ij_c}
\end{figure*}

We consider an exploration of the pairwise extremal dependence using Monte Carlo estimates of the coefficients in  Equation~\ref{eqn:chi_ij}. For this we use samples obtained from the PRG--G model. Figure~\ref{fig:chi_ij} provides a graphical analysis of the results. The coefficients achieve values between $0.286$ and $0.759$ for the ERA-Interim data and between $0.181$ and $0.840$ for the ERA5 data.  The greater range in dependence scores observed with the ERA5 data versus ERA-Interim speaks to the greater granularity of the ERA5 data\st{--}\hl{, indicating that }distance between locations is a strong contributor to the strength of the pairwise asymptotic dependence. The highest coefficients are $0.759$ for locations 4 and 5 in the ERA-Interim data and $0.840$ for locations 1 and 2 in the ERA5 data.  Clearly, pairwise asymptotic dependence coefficients tell a limited story, as a particular dependence may include more than two locations.   We can, however, glean some information from the patterns that emerge in two dimensions.  For the ERA-Interim data, we observe a possible cluster between cells 5-8, indicating a strong dependence among these cells.  Analogously, for the ERA5 data, we observe three possible groups of locations.

\begin{figure*}[htb]
\centering
\caption{Conditional survival curves for selected locations, using ERA-Interim, and PRG-G model,  conditioning on all other dimensions at greater than 90th percentile (fitted)\label{fig:condsurv1d}. The left panel uses original units. Right panel uses standardized units.}
\includegraphics[width=\linewidth]{./images/condsurv_1d}
\end{figure*}

\begin{figure*}[htb]
\centering
\caption{Pairwise conditional survival curves for selected locations, using
ERA-Interim, and PRG-G model, conditioning on all other dimensions at greater
than 90th percentile (fitted).\label{fig:condsurv2d}}
\includegraphics[height=4in, width=4in]{./images/condsurv_2d}
\end{figure*}

Figure~\ref{fig:condsurv1d} shows, for the ERA-Interim data under the PRG--G model, the conditional survival curve defined in Equation~\ref{eqn:condsurv}, for one dimension, conditioned on all other dimensions being greater than their (fitted) $90$th percentile.  Figure~\ref{fig:condsurv2d} presents the bi-variate conditional survival function, conditioning on all other dimensions.  These results illustrate quantitatively how extremal dependence affects the shape of the conditional survival curves.  The two top panels represent the joint survival function between grid locations 4 and 5, which are shown in Figure~\ref{fig:chi_ij} to exhibit strong  extremal dependence.  We observe that the joint survival surface is strongly convex.  The bottom panels represent the joint survival surface between grid locations 1 and 5, which exhibited low extremal dependence.  In this case the shape of the contours tend to be concave, quite different from the shapes observed in the top panels.

Using our proposed scoring criteria, we explored the effect of the choice of $p$ on the final results. Using the simulated data, generated from a mixture of projected Gammas, we were unable to observe sizeable differences in the scores for $p$ ranging between 1 and 15.   However, for the IVT data, we observed a drop in the energy score associated with higher $p$, with diminishing effect as $p$ increased.  We observed no significant differences in the performance of the model that uses $p=10$, which corresponds to the analysis presented, relative to the one that uses $p=15$.

\section{Conclusion\label{sec:conclusion}}
In this paper, we have built upon the definition of the multivariate Pareto described in \cite{ferreira2014} to establish a useful representation of its dependence structure, through the distribution of its angular component, which is supported on the positive orthant of the unit hypersphere under the $\mathcal{L}_{\infty}$ norm.  Due to the inherent difficulty of obtaining \st{a}\hl{the likelihood of} distribution\hl{s} with support on ${\mathbb S}^{d-1}_\infty$ our method \st{projects distributions, supported on the positive quadrant and based on products of independent gammas,  onto the manifold ${\mathbb S}_{p}^{d-1}$. Samples of the resulting probability distribution are then projected onto ${\mathbb S}_{\infty}^{d-1}$. }\hl{transforms the data to ${\mathbb S}_{p}^{d-1}$, fits then using mixtures of products of independent gammas, then transforms the predictions back to ${\mathbb S}^{d-1}_\infty$.} As ${\mathbb S}_{p}^{d-1}$ converges to ${\mathbb S}_{\infty}^{d-1}$ as  $p\to\infty$, we expect the \hl{proposed }resampling to be efficient for large enough $p$. In fact, our exploration of the simulated and real data indicates that the procedure is robust to the choice of moderately large values of $p$. \hl{Our method includes two inferential steps. The first consists of the estimation of the marginal Pareto distributions. The second consists of the estimation of the angular density. Parameter uncertainty incurred in the former is not propagated to the latter. Conceptually, an integrated approach that accounts for all the estimation uncertainty is conceivable. Unfortunately, this leads to posterior distributions with complex data dependent restrictions that are very difficult to explore, especially in large dimensional settings. In fact, our attempts to fit a simple parametric model for the marginals and the angular measure jointly in several dimensions were not successful.}

\st{We explored the behavior of the proposed model using simulated data from mixtures of projected Gammas with varying degrees of complexity---both in terms of dimensionality and in number of mixture components.  From this, we learned that the additional flexibility offered by varying the $\beta$ term---the rate parameter of the Gamma distribution---does not result in increased model fidelity in a mixture setting.  Within the tested range, between 3 and 20 dimensions, we observed that the additional information provided by a log--normal centering distribution did not translate to additional model fidelity.  However, for real data, with 47 dimensions, we did observe an effect.}

\hl{In this paper we have focused on a particular representation of the multivariate Pareto distribution for PoT inference on extreme values. To this end, our model  provides a computationally efficient and flexible approach. An interesting extension of the proposed model is to consider regressions of extreme value responses, due to extreme value inputs following the ideas in  \mbox{\cite{carvalho2022}}. This will produce PoT based Bayesian non-parametric extreme value regression models. More generally,  models that allow for covariate-dependent extremal dependence \mbox{\citep{mhalla2019}} could be considered. In addition, we notice that our approach is based on flexibly  modeling angular distributions for any $p$-norm. As such, it can be applied to other problems focused on high dimensional directional statistics constrained to a cone of directions.}

The computations in this paper were performed on a desktop computer with an AMD Ryzen 5950X processor. The program is largely single-threaded, so computation time is not dependent on available core count.  In each case, we run the MCMC chain for 50\,000 iterations, with a burn-in of 40\,000 samples.  Fitting the PG--G model on the ERA5 dataset took approximately 15 minutes.  Work is in progress to optimize the code, and explore parallelization where possible.  We are also exploring alternative computational approaches that will make it feasible to tackle very high dimensional problems, such as variational Bayes. In fact, to elaborate on the study of IVT, there is a need to consider several hundreds, if not thousands, of grid cells over the Pacific Ocean in order to obtain a good description of atmospheric  events responsible for large storm activity over California.

\section*{Acknowledgements}
This material is based upon work supported by the National Science Foundation under Grant No. SES-2050012 and DMS-2153277.

\bibliography{./refs}

\end{document}

% EOF

### 2024-02-01 abstract_sbss.tex
changes

We consider a constructive definition of the multivariate Pareto that factorizes the 
random vector into a radial component and an independent angular component; The former
following a univariate Pareto distribution, and the latter defined on the surface of the 
positive orthant of the infinity norm unit hypercube.   We propose a method 
for inferring the distribution of the angular component, by identifying its support 
as the limit of the positive orthant of the unit $p$--norm spheres, and introduce a 
projected gamma family of distributions defined through the normalization of a vector of 
independent random gammas to the space.   This serves in constructing a flexible family of
distributions obtained as a Dirichlet process mixture of projected gammas.  For model 
assessment, we discuss scoring methods appropriate to distributions on the unit hypercube.  
In particular, working with the energy score criterion, we develop a kernel metric that 
produces a proper scoring rule, and present 
a simulation study to compare different modeling choices using the proposed metric.    
Using our approach, we describe the dependence structure of extreme values in the
integrated vapor transport (IVT), data describing the flow of atmospheric moisture 
along the coast of California.  We find clear but heterogeneous geographical dependence.

### 2024-02-01 conclusion.tex
changes

@@
     described in \cite{ferreira2014} to establish a useful representation of 
-    its dependence structure, through the distribution of its angular component,
+    its dependence structure through the distribution of its angular component,
     which is supported on the positive orthant of the unit hypersphere under 
-    the $\mathcal{L}_{\infty}$ norm.  Due to the inherent difficulty of 
+    the $\mathcal{L}_{\infty}$ norm, $\mathbb{S}_{\infty}^{d-1}$.
+    Due to the inherent difficulty of 
     obtaining the likelihood of distributions with support on ${\mathbb S}^{d-1}_\infty$ 
@@
     Our method includes two inferential steps. The first consists of the
-    estimation of the marginal Pareto distributions. The second consists 
+    estimation of the marginal Pareto distributions; the second consists 
     of the estimation of the angular density. Parameter uncertainty incurred in
@@
     statistics  constrained  to a cone of directions.
+
+Establishing the angular measure
+    specifically in $\mathbb{S}_{\infty}^{d-1}$ provides two benefits over 
+    $\mathbb{S}_p^{d-1}$.  First, the transformation to 
+    $\mathbb{S}_{\infty}^{d-1}$ is unique.  Observe in 
+    Equation~\eqref{eqn:pnormtinv}, establishing $y_d$ as a function
+    of $y_1,\ldots,y_{d-1}$ can be replaced in form with any $y_{\ell}$, 
+    indicating there are $d$ equivalent transformations, each yielding a different
+    Jacobian and, for $p>1$, potentially resulting in a different calculated density.  
+    Second, evaluation of geodesic distance on $\mathbb{S}_p^{d-1}$ is not 
+    straightforward.  However, we have demonstrated a computationally efficient 
+    upper bound on geodesic distance on $\mathbb{S}_{\infty}^{d-1}$.
 

### 2024-02-01 introduction.tex
changes

@@
     \cite{falk2019} provide successful examples of this approach in an EVT 
-    framework. \cite{ferreira2014} presents a constructive definition of the 
+    framework.
+
+    \cite{ferreira2014} presents a constructive definition of the 
     Pareto process, that generalizes the GP to an infinite dimensional setting. It 
@@
     Inference for the limiting distribution of the exceedances
-    needs to account for the normalizing constant in Equation (\ref{exponent-mu}), as well
-    the moment constraints. In addition, the moment constraints can impose unrealistic
-    symmetries in the data distribution.
+    needs to account for the normalizing constant in Equation~\eqref{exponent-mu}, as well
+    the moment constraints.
+
+    % In addition, the moment constraints can impose unrealistic
+    % symmetries in the data distribution.
 

### 2024-02-01 main_sbss.tex
sbss initial
changes

--- 2022-06-18 main.tex
@@
-\documentclass[iicol,sn-basic]{sn-jnl}
+\documentclass[11pt]{article}
 
-\usepackage{amsmath}
-\usepackage{amssymb}
-%\usepackage{natbib}
-\usepackage{amsthm}
+\usepackage{amsmath,amssymb,amsfonts,amsthm}
+\usepackage{natbib}
 \usepackage{hyperref}
-%\usepackage{graphicx}
-%\usepackage{xcolor}
+\usepackage{graphicx}
+\usepackage{xcolor}
 \usepackage{booktabs}
 \usepackage{soul}
-%\usepackage[ruled]{algorithm2e}
 \usepackage{bm}
@@
 \usepackage{wrapfig}
-% \usepackage{setspace}
-% \linespread{1.8}
+\usepackage{setspace}
+\linespread{1.8}
+
+\usepackage{multirow}%
+\usepackage{mathrsfs}%
+\usepackage{textcomp}%
+\usepackage{manyfoot}%
+\usepackage{algorithm}%
+\usepackage{algorithmicx}%
+\usepackage{algpseudocode}%
 \usepackage{anysize}
-\marginsize{1in}{1in}{1in}{1in}
+\usepackage[margin=1.75cm]{geometry}
 
@@
 \newcommand{\bigmid}{\,\middle\vert\,}
-% \newcommand{\bruno}[1]{\textcolor{blue}{#1}} % Bruno edits - blue
-% \newcommand{\makenote}[1]{\textcolor{red}{#1}} % Peter edits - red
+\newcommand{\bruno}[1]{\textcolor{red}{#1}} % Bruno edits - red
+\newcommand{\makenote}[1]{\textcolor{teal}{#1}} % Peter edits - teal
 
@@
 
-\jyear{2022}
-\theoremstyle{thmstyleone}
+\raggedbottom
 
-\raggedbottom
+\title{Bayesian Non-parametric Inference for Multivariate\\ Peaks-over-threshold Models}
+
+% \author{Peter Trubey \& Bruno Sans\'{o}\\ Department of Statistics, UC Santa Cruz}
+\author{}
+\date{}
 
@@
 
-\title[~\hspace{6.75in}BNP inference for multivariate PoT]{Bayesian non-parametric inference for multivariate peaks-over-threshold models}
+\maketitle
 
-\author*{\fnm{Peter} \sur{Trubey}}\email{ptrubey@ucsc.edu}
-\author{\fnm{Bruno} \sur{Sans\'{o}}}\email{bruno@soe.ucsc.edu}
-
-\affil{\orgdiv{Department of Statistics}, \orgname{University of California Santa Cruz}, 
-            \orgaddress{\street{1156 High Street}, \city{Santa Cruz},\postcode{95064}, \state{CA}, \country{USA}}}
-
-\abstract{\input{./abstract}}
-\keywords{Multivariate extremes, Peak over threshold models,
-                Bayesian non-parametric models, Dirichlet process mixtures}
+\abstract{\input{./abstract_sbss}}
+% \keywords{Multivariate extremes, Peak over threshold models,
+%                 Bayesian non-parametric models, Dirichlet process mixtures}
 
@@
 
+\bibliographystyle{plainnat}
 \bibliography{./refs}

### 2024-02-01 methodology_projgamma.tex
changes

@@
   ${\mathbb S}_p^{d-1} $. With a slight abuse of terminology we refer to it as a
-  projection. Using Equations (\ref{eqn:pnormt}) -- (\ref{eqn:pnormjac}) we have the joint density
+  projection. Using Equations \eqref{eqn:pnormt}--\eqref{eqn:pnormjac} we have the joint density
   \begin{equation}
@@
   problems when estimating the shape and scale parameters, we set $\beta_1 = 1$.
-  \cite{nunez2019} obtain the density in Equation (\ref{eqn:projgamma})
+  \cite{nunez2019} obtain the density in Equation~\eqref{eqn:projgamma}
   for $p=2$ as a multivariate distribution for directional data, using spherical coordinates.  
   For $\bm{ y}\in {\mathbb S}_1^{d-1}$, and $\beta_{\ell} = \beta$ for all $\ell$, the density in
-  Equation (\ref{eqn:projgamma}) corresponds to that of a Dirichlet distribution.
+  Equation~\eqref{eqn:projgamma} corresponds to that of a Dirichlet distribution.
 
@@
   transformation is not differentiable. Thus, a closed form expression 
-  like Equation (\ref{eqn:projgamma}) for the projected gamma density on ${\mathbb S}_\infty^{d-1}$ is not 
+  like Equation~\eqref{eqn:projgamma} for the projected gamma density on ${\mathbb S}_\infty^{d-1}$ is not 
   available. Instead, we observe that for a sufficiently large $p$, $\mathbb{S}_p^{d-1}$ will approach
@@
   distribution $\Phi$ using a sample based approach with the following steps: (i) Apply the 
-  transformation in Equation (\ref{eqn:standardization}) to the original data; (ii) Obtain the 
+  transformation in Equation~\eqref{eqn:standardization} to the original data; (ii) Obtain the 
   subsample of the standardized observations that satisfy $R>1$; (iii) Take a finite $p$ and 
   project the observations onto ${\mathbb S}_p^{d-1}$; 
-  (iv) Fit the model in Equation (\ref{eqn:PGmix}) to the resulting 
+  (iv) Fit the model in Equation~\eqref{eqn:PGmix} to the resulting 
   data and obtain samples from the fitted model; 
@@
 
-Equation \ref{eqn:chi_ij} implies that $\chi_{\jmath\ell}>0$, and so, no asymptotic independence is possible 
+Equation~\eqref{eqn:chi_ij} implies that $\chi_{\jmath\ell}>0$, and so, no asymptotic independence is possible 
   under our proposed model. For the analysis of extreme values it is of interest to calculate the multivariate 
@@
 
-Equations \ref{eqn:chi_ij} and \ref{eqn:condsurv} provide relevant tools for inference on the tail 
+Equations~\eqref{eqn:chi_ij} and \eqref{eqn:condsurv} provide relevant tools for inference on the tail 
   behavior of the joint distribution of the observations. The expressions can be readily calculated 
@@
     groups observations into stochastically assigned clusters, where members of a cluster share 
-    distributional parameters \citep{muller2015,ascolani2022}. Building out the methods of inference for Equation~(\ref{eqn:dppg}), 
+    distributional parameters \citep{muller2015,ascolani2022}. Building out the methods of inference for Equation~\eqref{eqn:dppg}, 
     let $n_j^{-i}$ be the number of observations in cluster $j$ not including observation $i$.  Let $J^{-i}$ be 
@@
 $\mathcal{PG}(\bm{y}_i\mid\bm{\alpha}_i,\bm{\beta}_i) $ by introducing a latent radial component $r_i$, 
-for each observation. Using Equation \eqref{joint} we observe that the 
+for each observation. Using Equation~\eqref{joint} we observe that the 
 full conditional of $r_i$ is easy to sample from, as it is given as
@@
 \begin{equation}
+    \label{eqn:alphabetafc}
     \begin{aligned}
@@
 \end{equation}
-Note that the ordering of the products can be reversed in the likelihood, indicating that given $\bm{r}$, 
-the likelihood becomes separable by dimension.
+Note that the ordering of the products can be reversed in Equation~\eqref{eqn:alphabetafc}, 
+indicating that with appropriate choice of centering distribution, the full conditionals 
+for $\bm{\alpha}_j,\bm{\beta}_j$ become separable by dimension.
 We first consider a centering distribution given by a product of independent Gammas:
@@
 We also assume a Gamma prior on $\eta$, that is updated via the procedure outlined in \cite{escobar1995}.  
-We refer to this model as the \emph{projected gamma--gamma} (PG--G) model.
-An advantage of the PG--G model is that, thanks to conjugacy, the rate parameters can easily be integrated out for inference on $\bm{\alpha}_j$.
+We refer to this model as the \emph{projected gamma--gamma} (PG--G) model.  An advantage of the PG--G model 
+is that, thanks to conjugacy, the rate parameters can easily be integrated out for inference on $\bm{\alpha}_j$.  
 Then, the full conditional for $\alpha_{j\ell}$ takes the form
@@
 We also explore a restricted form of this model, where $\beta_{\ell} = 1$ for all $\ell$.  Under this 
-model, we use the full conditional in Equation~(\ref{eqn:alpha1update}) for all $\ell$, and omit 
+model, we use the full conditional in Equation~\eqref{eqn:alpha1update} for all $\ell$, and omit 
 inference on $\bm{\zeta},\bm{\sigma}$.  We refer to this model as the 

### 2024-02-01 results.tex
changes

@@
 
-\begin{algorithm}[ht]  % algorithm using algorithm package
+\begin{algorithm}  % algorithm using algorithm package
         \caption{Simulated Angular Dataset Generation Routine\label{algo:simulated}.
@@
     a model, a posterior predictive dataset is generated, and the energy score is
-    calculated as a Monte Carlo approximation of Equation~\ref{eq:es}.  In our analysis,
+    calculated as a Monte Carlo approximation of Equation~\eqref{eq:es}.  In our analysis,
     after burn-in and thinning, we had 1,000 replicates from the posterior distribution, and
@@
   $b_{t,\ell} = \hat{F}_{\ell}^{-1}(1 - t^{-1})$, where $\hat{F}$ is the empirical CDF and $t=20$, 
-  that corresponds to the $95$ percentile. We then use the transformation in Equation
-  \eqref{eqn:standardization} to standardize the observations.  Dividing each standardized observation
+  that corresponds to the $95$ percentile. We then use the transformation in Equation~\eqref{eqn:standardization} 
+  to standardize the observations.  Dividing each standardized observation
   by its $\mathcal{L}_{\infty}$ norm, we obtain a projection onto $\mathcal{S}_{\infty}^{d-1}$. As 
@@
 We consider an exploration of the pairwise extremal dependence using Monte Carlo estimates of the 
-  coefficients in  Equation~\ref{eqn:chi_ij}. For this we use samples obtained from the PRG--G model.
+  coefficients in  Equation~\eqref{eqn:chi_ij}. For this we use samples obtained from the PRG--G model.
   Figure~\ref{fig:chi_ij} provides a graphical analysis of the results. 
@@
 Figure~\ref{fig:condsurv1d} shows, for the ERA-Interim data under the PRG--G model,  
-    the conditional survival curve defined in Equation~\ref{eqn:condsurv}, for one dimension, 
+    the conditional survival curve defined in Equation~\eqref{eqn:condsurv}, for one dimension, 
     conditioned on all other dimensions being greater than their (fitted) $90$th percentile. 

### 2024-02-08 conclusion.tex
init switch to entropy

@@
     straightforward.  However, we have demonstrated a computationally efficient 
-    upper bound on geodesic distance on $\mathbb{S}_{\infty}^{d-1}$.
+    upper bound on geodesic distance on $\mathbb{S}_{\infty}^{d-1}$.  Accepting
+    these foibles, it would be interesting to explore the distribution on 
+    $\mathbb{S}_{p}^{d-1}$,
 
@@
 The computations in this paper were performed on a desktop computer with an 
-    AMD Ryzen 5950X processor. The program is largely single-threaded, so 
+    AMD Ryzen 5000 series processor. The program is largely single-threaded, so 
     computation time is not dependent on available core count.  In each case, 

### 2024-02-20 cover.tex
entropy submission RC1

@@
 
-\begin{letter}{Editorial Board \\ Computational Statistics \& Data Analysis}
+\begin{letter}{Editorial Board \\ \emph{Entropy}}
 
@@
 
-We submit to you our manuscript for review.  It details a method of inference for the dependence structure of
-extremes in a multivariate peaks over thresholds model based mixtures of projections of vectors of gamma variables, 
-along with an appropriate model selection criteria and applications of the fitted model.  A simulation study 
-is performed to explore properties of the distribution, and the model then applied to atmospheric data.
+We submit to you our manuscript for inclusion in the special issue of \emph{Entropy}.  It details a method of 
+inference for the dependence structure of extremes in a multivariate peaks over thresholds model, based on a 
+mixture of projections of vectors of gamma variables.  Included are an appropriate model selection 
+criterion, and applications of the fitted model.  A simulation study is performed to explore properties of the 
+distribution, and the model then applied to atmospheric data.
 
-We confirm the content of this work is original, and is not under consideration at any other journal.
+We confirm that neither the manuscript nor any parts of its content are currently under consideration or published in another journal.
 
-We further confirm that there are no competing interests to declare.
+All authors have approved the manuscript and agree with its submission to \emph{Entropy}.
 

### 2024-02-20 evaluation.tex
changes
entropy submission RC1

@@
-\section{Scoring criteria for distributions in ${\mathbb S}_\infty^{d-1}$\label{sec:evaluation}}
+\section{Scoring criteria for distributions on the infinity--norm sphere\label{sec:evaluation}}
 
@@
     \label{eq:es}
-    \begin{aligned}
-    &\hspace{-0.5cm}\text{S}_{\text{ES}}\left(P, \bm{x}_i\right) \\
-    &=\text{E}_p \left[g\left(\bm{X}_i, \bm{x}_i\right)\right] -
+    \text{S}_{\text{ES}}\left(P, \bm{x}_i\right)
+    =\text{E}_p \left[g\left(\bm{X}_i, \bm{x}_i\right)\right] -
                                     \frac{1}{2}\text{E}_p \left[g\left(\bm{X}_i,\bm{X}_i^{\prime}\right)\right],
-    \end{aligned}
     \end{equation}
-where $g$ is a kernel function. The score defined in Equation \eqref{eq:es} can be evaluated
+where $g$ is a kernel function. The score defined in Equation~\eqref{eq:es} can be evaluated
   using samples from $P$, with the help of the law of large numbers.
@@
     between the two points.  Let
-\begin{equation*}
+    \[
     {\mathbb C}_{\ell}^{d-1} = \lbrace \bm{x} : \bm{x} \in {\mathbb S}_{\infty}^{d-1}, x_{\ell} = 1\rbrace
-\end{equation*}
+    \]
     comprise the $\ell$th \emph{face}.  For points on the same face, the Euclidean distance 
@@
 We then have that
-  \begin{equation*}
+    \[
     \begin{aligned}
-    g(\bm{a},\bm{b}) = &\min_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
+    g(\bm{a},\bm{b}) &= \min_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
         \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\} \\
-     &\min_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
+     &= \min_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
         \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}^{\prime}-\bm{c}}{2} \right\}\\
-    &= \pnorm{\bm{a} - \bm{b}^{\prime}}{2}.
+    &=\pnorm{\bm{a} - \bm{b}^{\prime}}{2}.
     \end{aligned}
-  \end{equation*}
+    \]
   The last equality is due to the fact that $\bm{a}$ and $\bm{b}'$ belong to the 

### 2024-02-20 main.tex
before some cutting
asd
prep for submission
entropy submission RC1

@@
-\documentclass[iicol,sn-basic]{sn-jnl}
-
-\usepackage{amsmath}
-\usepackage{amssymb}
-%\usepackage{natbib}
-\usepackage{amsthm}
-\usepackage{hyperref}
-%\usepackage{graphicx}
-%\usepackage{xcolor}
-\usepackage{booktabs}
-\usepackage{soul}
-%\usepackage[ruled]{algorithm2e}
+%  LaTeX support: latex@mdpi.com 
+%  For support, please attach all files needed for compiling as well as the log file, and specify your operating system, LaTeX version, and LaTeX editor.
+%=================================================================
+\documentclass[entropy,article,submit,pdftex,moreauthors]{Definitions/mdpi} 
+%=================================================================
+% MDPI internal commands - do not modify
+\firstpage{1} 
+\makeatletter 
+\setcounter{page}{\@firstpage} 
+\makeatother
+\pubvolume{1}
+\issuenum{1}
+\articlenumber{0}
+\pubyear{2024}
+\copyrightyear{2024}
+%\externaleditor{Academic Editor: Firstname Lastname}
+\datereceived{ } 
+\daterevised{ } % Comment out if no revised date
+\dateaccepted{ } 
+\datepublished{ } 
+%\datecorrected{} % For corrected papers: "Corrected: XXX" date in the original paper.
+%\dateretracted{} % For corrected papers: "Retracted: XXX" date in the original paper.
+\hreflink{https://doi.org/} % If needed use \linebreak
+%\doinum{}
+%\pdfoutput=1 % Uncommented for upload to arXiv.org
+%\CorrStatement{yes}  % For updates
+%=================================================================
+% Add packages and commands here.
+The following packages are loaded in our class file: fontenc, inputenc, calc, indentfirst, fancyhdr, graphicx, epstopdf, lastpage, ifthen, float, amsmath, amssymb, lineno, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, tabto, xcolor, colortbl, soul, multirow, microtype, tikz, totcount, changepage, attrib, upgreek, array, tabularx, pbox, ragged2e, tocloft, marginnote, marginfix, enotez, amsthm, natbib, hyperref, cleveref, scrextend, url, geometry, newfloat, caption, draftwatermark, seqsplit
+% cleveref: load \crefname definitions after \begin{document}
 \usepackage{bm}
+\usepackage{algorithm}
+\usepackage{algorithmicx}
+\usepackage{algpseudocode}
 \usepackage{subcaption}
-\usepackage{wrapfig}
-% \usepackage{setspace}
-% \linespread{1.8}
-\usepackage{anysize}
-\marginsize{1in}{1in}{1in}{1in}
-
+\captionsetup[sub]{position=bottom,
+        labelfont={bf, small, stretch=1.17},
+        labelsep=space,
+        textfont={small, stretch=1.5},
+        aboveskip=6pt, 
+        belowskip=6pt,
+        singlelinecheck=off,
+        justification=justified}   
+%=================================================================
+% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition, Notation, Assumption
+%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).
 \newtheorem{prop}{Proposition}
@@
 \newcommand{\bigmid}{\,\middle\vert\,}
-% \newcommand{\bruno}[1]{\textcolor{blue}{#1}} % Bruno edits - blue
-% \newcommand{\makenote}[1]{\textcolor{red}{#1}} % Peter edits - red
-
+\newcommand{\bruno}[1]{\textcolor{red}{#1}} % Bruno edits - red
+\newcommand{\makenote}[1]{\textcolor{teal}{#1}} % Peter edits - teal
 \newcommand{\findcite}{\textcolor{red}{[Find Citation]}}
@@
 \newcommand{\Chi}{\mbox{\Large$\chi$}}
-
 \newcommand{\norm}[1]{\left\lVert #1 \right\rVert}
@@
 \newcommand{\expect}[1]{\text{E}\left[#1\right]}
+%=================================================================
+% Full title of the paper (Capitalized)
+\Title{Bayesian non-parametric inference for multivariate peaks-over-threshold models}
 
-\jyear{2022}
-\theoremstyle{thmstyleone}
+% MDPI internal command: Title for citation in the left column
+\TitleCitation{BNP inference for multivariate PoT}
 
-\raggedbottom
+% Author Orchid ID: enter ID or remove command
+\newcommand{\orcidauthorA}{0000-0002-7549-2033} % Add \orcidA{} behind the author's name
+\newcommand{\orcidauthorB}{0000-0001-5436-8028} % Add \orcidB{} behind the author's name
 
+% Authors, for the paper (add full first names)
+\Author{Peter Trubey $^{1}$\orcidA{} and Bruno Sans\'{o} $^{2}$\orcidB{}}
+
+%\longauthorlist{yes}
+
+% MDPI internal command: Authors, for metadata in PDF
+\AuthorNames{Peter Trubey and Bruno Sans\'{o}}
+
+% MDPI internal command: Authors, for citation in the left column
+\AuthorCitation{Trubey, P.; Sans\'{o}, B.}
+
+% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
+\address{%
+$^{1}$ \quad Department of Statistics, University of California, Santa Cruz, ptrubey@ucsc.edu\\
+$^{2}$ \quad Department of Statistics, University of California, Santa Cruz, bsanso@ucsc.edu
+}
+% $^{2}$ \quad Affiliation 2; e-mail@e-mail.com}
+
+% Contact information of the corresponding author
+\corres{Correspondence: ptrubey@ucsc.edu}
+
+% Current address and/or shared authorship
+% The commands \thirdnote{} till \eighthnote{} are available for further notes
+
+%\simplesumm{} % Simple summary
+
+%\conference{} % An extended version of a conference paper
+
+% Abstract (Do not insert blank lines, i.e. \\) 
+% \abstract{A single paragraph of about 200 words maximum.
+For research articles, abstracts should give a pertinent overview of the work.
+We strongly encourage authors to use the following style of structured abstracts, but without headings: (1) Background: place the question addressed in a broad context and highlight the purpose of the study; (2) Methods: describe briefly the main methods or treatments applied; (3) Results: summarize the article's main findings; (4) Conclusions: indicate the main conclusions or interpretations.
+The abstract should be an objective representation of the article, it must not contain results which are not presented and substantiated in the main text and should not exaggerate the main conclusions.}
+\abstract{We consider a constructive definition of the multivariate Pareto that factorizes the 
+random vector into a radial component and an independent angular component; The former
+following a univariate Pareto distribution, and the latter defined on the surface of the 
+positive orthant of the infinity norm unit hypercube.   We propose a method 
+for inferring the distribution of the angular component, by identifying its support 
+as the limit of the positive orthant of the unit $p$--norm spheres, and introduce a 
+projected gamma family of distributions defined through the normalization of a vector of 
+independent random gammas to the space.   This serves in constructing a flexible family of
+distributions obtained as a Dirichlet process mixture of projected gammas.  For model 
+assessment, we discuss scoring methods appropriate to distributions on the unit hypercube.  
+In particular, working with the energy score criterion, we develop a kernel metric that 
+produces a proper scoring rule, and present 
+a simulation study to compare different modeling choices using the proposed metric.    
+Using our approach, we describe the dependence structure of extreme values in the
+integrated vapor transport (IVT), data describing the flow of atmospheric moisture 
+along the coast of California.  We find clear but heterogeneous geographical dependence.}
+
+% Keywords
+\keyword{Multivariate extremes, Peak over threshold models, Bayesian non-parametric models, Dirichlet process mixtures}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{document}
 
-\title[~\hspace{6.75in}BNP inference for multivariate PoT]{Bayesian non-parametric inference for multivariate peaks-over-threshold models}
-
-\author*{\fnm{Peter} \sur{Trubey}}\email{ptrubey@ucsc.edu}
-\author{\fnm{Bruno} \sur{Sans\'{o}}}\email{bruno@soe.ucsc.edu}
-
-\affil{\orgdiv{Department of Statistics}, \orgname{University of California Santa Cruz}, 
-            \orgaddress{\street{1156 High Street}, \city{Santa Cruz},\postcode{95064}, \state{CA}, \country{USA}}}
-
-\abstract{\input{./abstract}}
-\keywords{Multivariate extremes, Peak over threshold models,
-                Bayesian non-parametric models, Dirichlet process mixtures}
-
-\maketitle
-
-%--------content
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \input{./introduction}
@@
 
-\section*{Acknowledgements}
-This material is based upon work supported by the National Science Foundation
-    under Grant No. SES-2050012 and DMS-2153277.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\vspace{6pt} 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\authorcontributions{Conceptualization, methodology, visualization, writing, P.T.; Supervision, editing, B.S.  All authors have read and agreed to the published version of the manuscript.}
+
+\funding{This material is based upon work supported by the National Science Foundation under Grant No. SES-2050012 and DMS-2153277.}
+
+% \dataavailability{We encourage all authors of articles published in MDPI journals to share their research data.
+In this section, please provide details regarding where data supporting reported results can be found, including links to publicly archived datasets analyzed or generated during the study.
+Where no new data were created, or where data is unavailable due to privacy or ethical restrictions, a statement is still required.
+Suggested Data Availability Statements are available in section ``MDPI Research Data Policies'' at \url{https://www.mdpi.com/ethics}.}
+
+\conflictsofinterest{The authors declare no conflicts of interest.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{adjustwidth}{-\extralength}{0cm}
 
@@
 
+\PublishersNote{}
+\end{adjustwidth}
 \end{document}
-
-% EOF

### 2024-02-20 methodology_projgamma.tex
entropy submission RC1

@@
 The $\mathcal{L}_p$-norm, for $p>0$ is  defined as
-  \begin{equation*}
+  \[
     \lVert \bm{x} \rVert_p = 
     \left({\textstyle\sum}_{\ell = 1}^d \lvert s_{\ell}\rvert^p\right)^{\frac{1}{p}},
-  \end{equation*}
+  \]
 The absolute and Euclidean norms are obtained for $p=1$ and $p=2$ respectively,
   and the $\mathcal{L}_{\infty}$ norm can be obtained as a limit, 
-  \begin{equation*}
+  \[
     \| \bm{s} \|_{\infty}
@@
       = \underset{\ell = 1}{\overset{d}{\bigvee}}s_{\ell}.
-  \end{equation*}
+  \]
 
@@
     \label{eqn:pnormt}
-    \begin{aligned}
-    T(x_1,\ldots,x_d) &= \left(\pnorm{\bm{x}}{p}, \frac{x_1}{\pnorm{\bm{x}}{p}},
-                          \ldots , \frac{x_{d-1}}{\pnorm{\bm{x}}{p}}\right) \\
-                        &= (r,y_1,\ldots,y_{d-1})
-    \end{aligned}
+    T(x_1,\ldots,x_d) = \left(\pnorm{\bm{x}}{p}, \frac{x_1}{\pnorm{\bm{x}}{p}},
+                          \ldots , \frac{x_{d-1}}{\pnorm{\bm{x}}{p}}\right) = (r,y_1,\ldots,y_{d-1})
   \end{equation}
@@
     \label{eqn:pnormtinv}
-    \begin{aligned}
-    &\hspace{-0.5cm}T^{-1}\left(r,y_1,\ldots,y_{d-1}\right)\\
-      & = \left(ry_1,\ldots,ry_{d-1}, 
+    T^{-1}\left(r,y_1,\ldots,y_{d-1}\right)
+       = \left(ry_1,\ldots,ry_{d-1}, 
         r\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}\right).
-    \end{aligned}
   \end{equation}
@@
     \label{eqn:pnormjac}
-    \begin{aligned}
-    r^{d-1}\left[\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}} \right. & \\
-        &\hspace{-2cm}+ \left. {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p
+    r^{d-1}\left[\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}} \right.
+        + \left. {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p
           \left(1 - {\textstyle\sum}_{l=1}^{d-1} y_{\ell}^p\right)^{\frac{1}{p} - 1}\right].
-    \end{aligned}
   \end{equation}
@@
   \label{joint}
-    \begin{aligned}
-    f(r,\bm{ y}) = &\prod_{\ell = 1}^{d}
+    f(r,\bm{ y}) = \prod_{\ell = 1}^{d}
       \left[\frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell})}(ry_{\ell})^{\alpha_{\ell} - 1}
-          \exp\lbrace-\beta_{\ell}ry_{\ell}\rbrace\right] \\
-      &\hspace{0.5cm}\times r^{d-1}\left[y_d +
+          \exp\lbrace-\beta_{\ell}ry_{\ell}\rbrace\right] \times r^{d-1}\left[y_d +
             {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p y_d^{1 - p}\right].
-    \end{aligned}
   \end{equation}
@@
     \label{eqn:projgamma}
-    \begin{aligned}
-    \text{PG}(\bm{ y}\mid\bm{ \alpha},\bm{ \beta}) &=
+    \text{PG}(\bm{ y}\mid\bm{ \alpha},\bm{ \beta}) =
           \prod_{\ell = 1}^d\left[\frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell})}
-                y_{\ell}^{\alpha_{\ell} - 1}\right]\\
-      &\hspace{0.5cm}\times \left[y_d +
-          {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p y_d^{1 - p}\right]\\
-      &\hspace{0.5cm}\times \frac{\Gamma({\textstyle\sum}_{\ell = 1}^d\alpha_{\ell})}{\left({\textstyle\sum}_{\ell = 1}^d
+                y_{\ell}^{\alpha_{\ell} - 1}\right] \times \left[y_d +
+          {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p y_d^{1 - p}\right]
+      \times \frac{\Gamma({\textstyle\sum}_{\ell = 1}^d\alpha_{\ell})}{\left({\textstyle\sum}_{\ell = 1}^d
                     \beta_{\ell}y_{\ell}\right)^{{\scriptstyle\sum_{\ell = 1}^d \alpha_{\ell}}}},
-    \end{aligned}
   \end{equation}
@@
   \begin{equation}\label{eqn:dppg}
-    \begin{aligned}
-    \bm{y}_i &\sim \text{PG}(\bm{y}_i\mid \bm{\theta}_i)\\
-    \bm{\theta}_i &\sim G
-    \end{aligned}
-    ~\hspace{1cm}
-    \begin{aligned}
-    G &\sim \mathcal{DP}(\eta, G_0),
-    \end{aligned}
+    \bm{y}_i \sim \text{PG}(\bm{y}_i\mid \bm{\theta}_i)\hspace{1cm}
+    \bm{\theta}_i \sim G\hspace{1cm}
+    G \sim \mathcal{DP}(\eta, G_0)
   \end{equation}
@@
 
-\begin{figure}[ht]
+\begin{figure}[t]
     \centering
-    \caption{The positive orthant of the $p$-norm sphere.}
   \begin{subfigure}[b]{0.4\textwidth}
@@
   \end{subfigure}
+    \caption{The positive orthant of the $p$--norm sphere for $d=2$.}
 \end{figure}
@@
   Using the fact that $V_\ell>0, \forall \ell$ almost surely, we have that the former is equal to
-  \begin{equation*}
-    \begin{aligned}
-    &\hspace{-0.5cm}\prob{R>\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}}
+  \[
+    %\begin{aligned}
+    %&\hspace{-0.5cm}
+    \prob{R>\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}}
     = \expect{1\wedge\left(\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}\right)^{-1}}\\
-    &= \expect{\frac{V_\jmath}{z_\jmath}\wedge\frac{V_\ell}{z_\ell}}
-    = (1 - u)\expect{\frac{V_\jmath}{\expect{V_\jmath}}\wedge\frac{V_\ell}{\expect{V_\ell}}},
-    \end{aligned}
-  \end{equation*}
+    %&
+    = \expect{\frac{V_\jmath}{z_\jmath}\wedge\frac{V_\ell}{z_\ell}}
+    = (1 - u)\expect{\frac{V_\jmath}{\expect{V_\jmath}}\wedge\frac{V_\ell}{\expect{V_\ell}}}
+    %\end{aligned}
+  \]
   where the second identity is justified by the fact that $V_i$ is bounded and $z_i\rightarrow\infty$. The 
@@
   \begin{equation} \label{eqn:condsurv}
-    \begin{aligned}
-    &\hspace{-0.5cm}\mathrm{Pr}\left[\bigcap_{\ell \in \alpha} 
-        Z_\ell > z_\ell \mid \bigcap_{\ell\not\in\alpha} Z_\ell > z_\ell\right] \\
-    &= \frac{\expect{\bigwedge_{k = 1}^d 1\wedge \frac{V_k}{z_k}}}{\expect{\bigwedge_{k \not\in\alpha}1\wedge\frac{V_k}{z_k}}}.
-    \end{aligned}
+    \mathrm{Pr}\left[\bigcap_{\ell \in \alpha} 
+        Z_\ell > z_\ell \mid \bigcap_{\ell\not\in\alpha} Z_\ell > z_\ell\right]
+    = \frac{\expect{\bigwedge_{k = 1}^d 1\wedge \frac{V_k}{z_k}}}{\expect{\bigwedge_{k \not\in\alpha}1\wedge\frac{V_k}{z_k}}}.
   \end{equation}
@@
     the probability of cluster membership for a given observation is proportional to
-\begin{equation*}
+    \[
     \text{Pr}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
@@
         \end{cases}
-\end{equation*}
+    \]
     where the top case is iterating over extant clusters $j = 1,\ldots, J^{-i}$, and the bottom case is
@@
     \label{eqn:alphabetafc}
-    \begin{aligned}
-    &\hspace{-1cm}f(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{Y},\bm{r},\bm{\delta},\ldots) \\
-    &\propto \prod_{i:\gamma_i = j}\prod_{\ell = 1}^d\mathcal{G}\left(r_iy_{i\ell}\mid\alpha_{j\ell},\beta_{j\ell}\right) \\
-    &\hspace{0.5cm} \times dG_0(\bm{\alpha}_j,\bm{\beta}_j).
-    \end{aligned}
+    f(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{Y},\bm{r},\bm{\delta},\ldots)
+    \propto \prod_{i:\gamma_i = j}\prod_{\ell = 1}^d\mathcal{G}\left(r_iy_{i\ell}\mid\alpha_{j\ell},\beta_{j\ell}\right)
+    \times dG_0(\bm{\alpha}_j,\bm{\beta}_j).
 \end{equation}
@@
 \begin{equation}
-    \begin{aligned}
-    &\hspace{-0.3cm}G_0(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{\xi},\bm{\tau},\bm{\zeta},\bm{\sigma}) \\
-    &= \prod_{\ell = 1}^d\mathcal{G}(\alpha_{j\ell}\mid \xi_{\ell},\tau_{\ell})\times\prod_{\ell = 2}^d\mathcal{G}(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}).
-    \end{aligned}
+    G_0(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{\xi},\bm{\tau},\bm{\zeta},\bm{\sigma})
+    = \prod_{\ell = 1}^d\mathcal{G}(\alpha_{j\ell}\mid \xi_{\ell},\tau_{\ell})\times\prod_{\ell = 2}^d\mathcal{G}(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}).
 \end{equation}
@@
     \label{eqn:alphalupdate}
-    \begin{aligned}
-    &\hspace{-0.5cm}\pi(\alpha_{j\ell}\mid \bm{r},\bm{Y},\bm{\gamma},\xi_\ell,\tau_\ell,\zeta_\ell,\sigma_\ell) \\
-    &\propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i\ell}\right)^{\alpha_{j\ell} - 1}\alpha_{j\ell}^{\xi_\ell - 1}e^{-\tau_\ell \alpha_{j\ell}}}{\Gamma^{n_j}(\alpha_{j\ell})} \\
-    &\hspace{0.5cm}\times \frac{\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
-    \end{aligned}
+    \pi(\alpha_{j\ell}\mid \bm{r},\bm{Y},\bm{\gamma},\xi_\ell,\tau_\ell,\zeta_\ell,\sigma_\ell)
+    \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i\ell}\right)^{\alpha_{j\ell} - 1}\alpha_{j\ell}^{\xi_\ell - 1}e^{-\tau_\ell \alpha_{j\ell}}}{\Gamma^{n_j}(\alpha_{j\ell})}\times \frac{\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
 \end{equation}
@@
     \label{eqn:alpha1update}
-    \begin{aligned}
-    &\hspace{-0.5cm}\pi(\alpha_{j1}\mid\bm{r},\bm{Y},\bm{\gamma},\xi_1,\tau_1) \\
-    &\propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}\alpha_{j1}^{\xi_1 - 1}e^{-\tau_1\alpha_{j1}}}{\Gamma^{n_j}(\alpha_{j1})}.
-    \end{aligned}
+    \pi(\alpha_{j1}\mid\bm{r},\bm{Y},\bm{\gamma},\xi_1,\tau_1)
+    \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}\alpha_{j1}^{\xi_1 - 1}e^{-\tau_1\alpha_{j1}}}{\Gamma^{n_j}(\alpha_{j1})}.
 \end{equation}
@@
     \label{eqn:betafc}
-    \begin{aligned}
-    &\hspace{-0.5cm}\beta_{j\ell}\mid\bm{r},\bm{Y},\alpha,\zeta_{\ell},\sigma_{\ell} \\
-    &\sim \mathcal{G}\left(\beta_{j\ell}\bigmid n_j\alpha_{j\ell} + \zeta_\ell, \sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right),
-    \end{aligned}
+    \beta_{j\ell}\mid\bm{r},\bm{Y},\alpha,\zeta_{\ell},\sigma_{\ell}
+    \sim \mathcal{G}\left(\beta_{j\ell}\bigmid n_j\alpha_{j\ell} + \zeta_\ell, \sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right),
 \end{equation}
@@
     \begin{equation}
-        \begin{aligned}
-        &\hspace{-0.5cm}G_0\left(\bm{\alpha}_j,\bm{\beta}_j\mid\bm{\mu},\bm{\Sigma},\zeta,\sigma\right) \\
-        &=\mathcal{LN}\left(\bm{\alpha}_j\mid\bm{\mu},\bm{\Sigma}\right)\times\prod_{\ell = 2}^d\mathcal{G}\left(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}\right).
-        \end{aligned}
+        G_0\left(\bm{\alpha}_j,\bm{\beta}_j\mid\bm{\mu},\bm{\Sigma},\zeta,\sigma\right)
+        =\mathcal{LN}\left(\bm{\alpha}_j\mid\bm{\mu},\bm{\Sigma}\right)\times\prod_{\ell = 2}^d\mathcal{G}\left(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}\right).
     \end{equation}
@@
     log-density.  That is,
-    \begin{equation*}
+    \[
         \begin{aligned}
-        &\hspace{-0.5cm}\pi(\bm{\alpha}_j\mid\bm{Y},\bm{r},\bm{\delta},\bm{\mu},\bm{\Sigma},\bm{\zeta},\bm{\sigma})\\
-        &\propto \exp\left\lbrace-\frac{1}{2}(\log\bm{\alpha}_j - \mu)^T\bm{\Sigma}^{-1}(\log\bm{\alpha}_j - \mu)\right\rbrace \\
-        &\hspace{0.5cm}\times\frac{1}{\prod_{\ell = 1}^d\alpha_{j\ell}}\times \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}}{\prod_{\ell = 1}^d\Gamma^{n_j}(\alpha_{j\ell})}\\
-        &\hspace{0.5cm}\times \prod_{\ell = 2}^d\frac{\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
+        \pi(\bm{\alpha}_j\mid\bm{Y},\bm{r},\bm{\delta},\bm{\mu},\bm{\Sigma},\bm{\zeta},\bm{\sigma})
+        &\propto \exp\left\lbrace-\frac{1}{2}(\log\bm{\alpha}_j - \mu)^T\bm{\Sigma}^{-1}(\log\bm{\alpha}_j - \mu)\right\rbrace
+        \times\frac{1}{\prod_{\ell = 1}^d\alpha_{j\ell}}\\
+        &\hspace{1cm}\times \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}}{\prod_{\ell = 1}^d\Gamma^{n_j}(\alpha_{j\ell})}
+        \times \prod_{\ell = 2}^d\frac{\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
         \end{aligned}
-    \end{equation*}
+    \]
     The inferential forms for $\beta_{j\ell}$ and its priors are the same as for PG--G.  The 

### 2024-04-10 abstract.tex
entropy submission RC1
push for mdpi resubmission

@@
 for inferring the distribution of the angular component.  We identify its support 
-as the limit of the positive orthant of the unit $p$--norm spheres, and introduce a 
+as the limit of the positive orthant of the unit $p$-norm spheres, and introduce a 
 projected gamma family of distributions defined through the normalization a vector of 
-independent gamma random variables to the $p$--norm sphere.  This family serves as a 
+independent gamma random variables to the $p$-norm sphere.  This family serves as a 
 building block for a flexible family of distributions obtained as a Dirichlet process 
@@
 through 2020.  We find clear but heterogeneous geographical dependence.
-
-% EOF

### 2024-04-10 conclusion.tex
push for mdpi resubmission

@@
 
-Establishing the angular measure
-    specifically in $\mathbb{S}_{\infty}^{d-1}$ provides two benefits over 
-    $\mathbb{S}_p^{d-1}$.  First, the transformation to 
-    $\mathbb{S}_{\infty}^{d-1}$ is unique.  Observe in 
-    Equation~\eqref{eqn:pnormtinv}, establishing $y_d$ as a function
-    of $y_1,\ldots,y_{d-1}$ can be replaced in form with any $y_{\ell}$, 
-    indicating there are $d$ equivalent transformations, each yielding a different
-    Jacobian and, for $p>1$, potentially resulting in a different calculated density.  
-    Second, evaluation of geodesic distance on $\mathbb{S}_p^{d-1}$ is not 
+Developing an angular measure specifically in $\mathbb{S}_{\infty}^{d-1}$ provides 
+    two benefits over $\mathbb{S}_p^{d-1}$.  First, the transformation to 
+    $\mathbb{S}_{\infty}^{d-1}$ is unique.  Recall that 
+    Equation~\eqref{eqn:pnormtinv} gives $y_d$ as a function
+    of $y_1,\ldots,y_{d-1}$. An analogous expression can be obtained for
+    any $y_{\ell}$. This indicates that there are $d$ equivalent transformations, 
+    each yielding a different
+    Jacobian and, for $p>1$, potentially resulting in a different density.  
+    Second, evaluation of geodesic distances on $\mathbb{S}_p^{d-1}$ is not 
     straightforward.  However, we have demonstrated a computationally efficient 
@@
 
-%We explored the behavior of the proposed model using simulated data from
-    %mixtures of projected Gammas with varying degrees of complexity---both 
-    %in terms of dimensionality and in number of mixture components.  From this,
-    %we observed that the additional flexibility offered by varying the $\beta$
-    %term---the rate parameter of the gamma distribution---can offer additional
-    %model fidelity.  \makenote{However,} this result is not borne out in our 
-    %real data example, for which in higher dimensions the simpler model is 
-    %strongly preferred.  Additionally, we observe
-    %that the choice of the log-normal centering distribution and additional
-    %flexibility it offers did not translate to additional model fidelity.  
-    %\makenote{This result is borne out in the real data, where it performed at best 
-    %comparably to the gamma model.}  
-    % This
-    % is borne out in the real data, where on the ERA--Interim dataset it performed
-    % only comparably, but on ERA--5 it performed significantly worse.
-    % {\bf I am not sure that this paragraph is appropriate anymore. I think that
-    % the different mixture models considered are pretty much equivalent, so we should
-    % favor the one that is easiest and most robust to fit.}
-
 The computations in this paper were performed on a desktop computer with an 
@@
     we run the MCMC chain for 50\,000 iterations, with a burn-in of 40\,000 
-    samples.  Fitting the PG--G model on the ERA5 dataset took approximately 
+    samples.  Fitting the PG-G model on the ERA5 dataset took approximately 
     15 minutes.  Work is in progress to optimize the code, and explore 

### 2024-04-10 introduction.tex
entropy submission RC1
push for mdpi resubmission

--- 2023-07-07 introduction.tex
@@
     provides insight on the theoretical properties of possible parametrizations. 
-    These are use in \cite{KiRoSeWa2019} for likelihood-based models for PoT 
+    These are used in \cite{KiRoSeWa2019} for likelihood-based models for PoT 
     estimation. A frequently used method for describing the dependence in 
@@
     approaches. Yet, our approach differs in that it is established in the 
-    peaks-over-threshold regime, and avoids the problem of dealing with the so 
-    called moment constraint by considering a constructive definition of the 
+    peaks-over-threshold regime, and uses a constructive definition of the 
     multivariate GP, based on the infinity norm. The approach proposed in this paper
@@
     advantages due its structural simplicitly, achieves flexibility using 
-    a mixture model, imposes no moment constraints, and scales well to moderately 
-    large dimensions.
+    a mixture model, and scales well to moderately large dimensions.
 
@@
     Section~\ref{sec:methodology} details our approach for estimating the angular 
-      measure, based on projecting an arbitrary distribution supported in ${\mathbb R}_+^d$ 
+      measure, based on transforming an arbitrary 
+      distribution supported in ${\mathbb R}_+^d$ 
       onto unit hyper-spheres defined by $p$-norms. 
@@
     between vectors, such as $\bm{a}\wedge\bm{b} = (a_1\wedge b_1, a_2\wedge b_2,\ldots)$.  
+    Similarly, we apply inequality and arithmetic operators operators to vectors, for
+    example $\bm{a}\leq \bm{b}$, and interpret them component-wise.
     We use uppercase to indicate random variables, lowercase to indicate points, and
@@
     $d$-dimensional random vector $\bm{W} = (W_1, \ldots ,W_d)$ with
-    cumulative distribution $F$. Following \cite{RoSeWa2018a}, assume 
+    cumulative distribution $F$. A common assumption on 
+    $\bm{W}$ is that it is in the so called domain of attraction of 
+    a multivariate max-stable distribution, $G$. Thus, following \cite{RoSeWa2018a}, assume 
     that there exists sequences of vectors $\bm{a}_n$ and $\bm{b}_n$,
-    and a $d$-variate distribution $G$ such that 
-    $\lim_{n\rightarrow\infty} F^n(\bm{a}_n \bm{w} + \bm{b}_n) = 
+    such that $\lim_{n\rightarrow\infty} F^n(\bm{a}_n \bm{w} + \bm{b}_n) = 
     G(\bm{w})$. $G$ is a $d$-variate generalized extreme value 
-    distribution. It follows that
-    \begin{equation*}
-        \begin{aligned}
-        \lim\limits_{n\rightarrow\infty} &\prob{\bm{a}^{-1}_n (\bm{W} - \bm{b}_n) 
-      \leq \bm{w} \mid \bm{W} \not\leq \bm{b}_n)} \\ 
-        &\hspace{0.5cm}= \frac{\log G(\bm{w} \wedge \bm{0}) 
-        - \log G(\bm{w}) }{\log G(\bm{0})} = H(\bm{w}).
-        \end{aligned}
-    \end{equation*}
-    $H$ is a multivariate Pareto distribution, and corresponds to a joint
-    distribution conditional on exceeding a multivariate threshold. In addition to
-    a copula function, $H$ depends on two $d$-dimensional vectors of 
-    parameters that we denote as $\bm{\xi}$ for the shapes and $\bm{\sigma}$ 
-    for the scales. \cite{RoSeWa2018a} provide a number of stochastic 
-    representations for $H$.  In particular, denote as $Z$ a
-    random variable with distribution $H$ where $\bm{\xi}= 1$ and 
-    $\bm{\sigma} = 0$.  Then, Remark 1 justifies the representation in 
-    \cite{ferreira2014}, giving $\bm{Z} = \bm{V}R$
-    where $R$ and $\bm{V}$ are independent. $R = \|\bm{Z}\|_\infty$ is 
-    distributed as a standard Pareto random variable, and $\bm{V} = 
-    \bm{Z}/\|\bm{Z}\|_\infty$ is a random vector in  
+    distribution. Notice that, even though the univariate marginals are obtained from 
+    a three parameter family, there is no parametric form to represent $G$.
+    Taking logarithms and expanding, we have that 
+    \[  
+        \lim\limits_{n\rightarrow\infty} n(1 - F(\bm{a}_n \bm{w} + \bm{b}_n)) = 
+        -\log G(\bm{w}), 
+    \]
+    $\forall \bm{w\in{\mathbb R}^d}$ such that $G(\bm{w})>0$. It follows that
+    \begin{multline*}
+        \lim\limits_{n\rightarrow\infty} \prob{\bm{a}^{-1}_n (\bm{W} - \bm{b}_n) 
+      \leq \bm{w} \mid \bm{W} \not\leq \bm{b}_n)} = 
+         \frac{\log G(\bm{w} \wedge \bm{0}) 
+        - \log G(\bm{w}) }{\log G(\bm{0})} = H(\bm{w}),
+    \end{multline*}
+    where $\bm{a}_n^{-1}$ indicates element-wise inversion, and 
+    $\{\bm{W} \not\leq \bm{b}_n\}$ denotes the set where at least one coordinate is above 
+    the corresponding  component of $\bm{b}_n$.  
+    $H$ is a multivariate Pareto distribution.  It corresponds to a joint
+    distribution conditional on exceeding a multivariate threshold. $H$
+    is defined by a non-parametric function governing the multivariate dependence, and 
+    two $d$-dimensional vectors of 
+    parameters that control the shapes and scales of the
+    marginals. We denote these as $\bm{\xi}$ for the shapes and $\bm{\sigma}$ 
+    for the scales.  \cite{RoSeWa2018a} provides a number of stochastic 
+    representations for $H$.  In this paper we focus in a
+    particular one that is proposed in \cite{ferreira2014}. To this end, 
+    we denote as $Z$ a random variable with distribution $H$ where $\bm{\xi}= 1$ 
+    and $\bm{\sigma} = 0$.  Then, $\bm{Z} = R\bm{V}$ where $R$ and $\bm{V}$ 
+    are independent. $R = \|\bm{Z}\|_\infty= \vee_{i=1}^d Z_i$ is
+    distributed as a standard Pareto random variable, and 
+    $\bm{V} = \bm{Z}/\|\bm{Z}\|_\infty$ is a random vector in  
     $\mathbb{S}_{\infty}^{d-1}$, the positive orthant of the 
@@
     this, to obtain a  PoT model we seek a flexible model for the 
-    distribution of $\bm{V} \in {\mathbb S}_{\infty}^{d-1}$.
+    distribution of $\bm{V} \in {\mathbb S}_{\infty}^{d-1}$, based
+    on a Bayesian non-parametric model.
 
 The approach considered in \cite{rootzen2018} focuses on the limiting
-    conditional distribution $H$. An alternative approach consists of assuming
+    conditional distribution $H$. An alternative approach to obtaining a limiting 
+    PoT distribution consists of assuming
     that regular variation \cite[see, for example,][]{resnick2008extreme} holds 
-    for the limiting distribution of $\bm{W}$, implying
+    for the limiting distribution of $\bm{W}$, implying that
     \[
@@
     \]
-    where $\mu$ is termed the exponent measure. $\mu$ has the homogeneity property
-    $\mu(tA) = t^{-1}\mu(A)$. Letting $\rho = \|W\|_p, p>0$ 
+    for some measure $\mu$ that is referred to as the exponent measure. 
+    $\mu$ has the homogeneity property $\mu(tA) = t^{-1}\mu(A)$. 
+    Letting $\rho = \|W\|_p, p>0$ 
     and $\bm{\theta} = \bm{W}/\rho$, define  $\Psi(B) = \mu(\{\bm{w} : \rho>1, 
@@
     to their $p$-norm. The constraint that all marginals of $\mu$ correspond to a
-    standard Pareto distribution leads to the so called moment constraints on $\Psi$. 
+    standard Pareto distribution leads to the so called moment constraints on $\Psi$, consisting
+    of
+    \[
+    \int_{{\mathbb S}_p^{d-1}} w_i \,d\Psi(\bm{w}) = \frac{1}{d}, \;\; i=1, \ldots , d.
+    \]
     Inference for the limiting distribution of the exceedances
-    needs to account for the normalizing constant in Equation (\ref{exponent-mu}), as well
-    the moment constraints. In addition, the moment constraints can impose unrealistic
-    symmetries in the data distribution.
+    needs to account for the normalizing constant in Equation~\eqref{exponent-mu}, as well
+    the moment constraints. Because of these issues, in this paper
+    we prefer to follow the limiting conditional distribution approach. An example of 
+    the application of the regular variation approach, using $p=1$ is developed in \cite{SaNa2014}.
 

### 2024-04-10 main.tex
push for mdpi resubmission

@@
 \newcommand{\bruno}[1]{\textcolor{red}{#1}} % Bruno edits - red
-\newcommand{\makenote}[1]{\textcolor{teal}{#1}} % Peter edits - teal
+\newcommand{\makenote}[1]{\textcol:or{teal}{#1}} % Peter edits - teal
 \newcommand{\findcite}{\textcolor{red}{[Find Citation]}}
@@
 for inferring the distribution of the angular component, by identifying its support 
-as the limit of the positive orthant of the unit $p$--norm spheres, and introduce a 
+as the limit of the positive orthant of the unit $p$-norm spheres, and introduce a 
 projected gamma family of distributions defined through the normalization of a vector of 

### 2024-04-10 methodology_projgamma.tex
push for mdpi resubmission

@@
 \subsection{Projected gamma family\label{subsec:projgamma}}
-The $\mathcal{L}_p$-norm, for $p>0$ is  defined as
+
+At the core of our PoT method is the development of a distribution 
+    on ${\mathbb S}_{p}^{d-1}=\{\bm{s} : \bm{s} \in 
+    {\mathbb R}_{+}^{d}, \| \bm{s}\|_{p} = 1\}$, where, for $p>0$, $\|\cdot\|_p$ is 
+    the $\mathcal{L}_p$-norm, of a vector $\bm{x}\in {\mathbb R}^d$, defined as
     \[
         \lVert \bm{x} \rVert_p = 
-    \left({\textstyle\sum}_{\ell = 1}^d \lvert s_{\ell}\rvert^p\right)^{\frac{1}{p}},
+        \left({\textstyle\sum}_{\ell = 1}^d \lvert x_{\ell}\rvert^p\right)^{\frac{1}{p}}.
     \]
@@
     \[
-    \| \bm{s} \|_{\infty}
-      = \lim\limits_{p\to\infty} \| \bm{s} \|
-      = \underset{\ell = 1}{\overset{d}{\bigvee}}s_{\ell}.
+        \| \bm{x} \|_{\infty}
+            = \lim\limits_{p\to\infty} \| \bm{x} \|
+            = \underset{\ell = 1}{\overset{d}{\bigvee}}x_{\ell}.
     \]
@@
 To obtain a distribution on 
-${\mathbb S}_{p}^{d-1}=\{\bm{s} : \bm{s} \in {\mathbb R}_{+}^{d}, \| \bm{s}\|_{p} = 1\}$ 
+${\mathbb S}_{p}^{d-1}$ 
 we start with a vector in 
@@
     T(x_1,\ldots,x_d) = \left(\pnorm{\bm{x}}{p}, \frac{x_1}{\pnorm{\bm{x}}{p}},
-                          \ldots , \frac{x_{d-1}}{\pnorm{\bm{x}}{p}}\right) = (r,y_1,\ldots,y_{d-1})
+                          \ldots , \frac{x_{d-1}}{\pnorm{\bm{x}}{p}}\right) 
+                          = (r,y_1,\ldots,y_{d-1})
   \end{equation}
@@
     \label{eqn:pnormtinv}
-    T^{-1}\left(r,y_1,\ldots,y_{d-1}\right)
-       = \left(ry_1,\ldots,ry_{d-1}, 
+    T^{-1}\left(r,y_1,\ldots,y_{d-1}\right)=
+        \left(ry_1,\ldots,ry_{d-1}, 
         r\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}\right).
@@
   \begin{equation}\label{eqn:dppg}
-    \bm{y}_i \sim \text{PG}(\bm{y}_i\mid \bm{\theta}_i)\hspace{1cm}
-    \bm{\theta}_i \sim G\hspace{1cm}
+    \bm{y}_i \sim \text{PG}(\bm{y}_i\mid \bm{\theta}_i)\hspace{.5cm}
+    \bm{\theta}_i \sim G\hspace{.5cm}
     G \sim \mathcal{DP}(\eta, G_0)
@@
   \end{subfigure}
-    \caption{The positive orthant of the $p$--norm sphere for $d=2$.}
+    \caption{The positive orthant of the $p$-norm sphere for $d=2$.}
 \end{figure}
+
+\subsection{Tail probabilities for the PoT model}
 
@@
   be asymptotically dependent, otherwise they are asymptotically independent. The following result 
-  provides the asymptotic dependence coefficient between two components of $\bm{Z}$ after PoT limit.
+  provides the asymptotic dependence coefficient between two components of $\bm{Z}$
+  for our proposed PoT model.
   \begin{prop}\label{ppchi}
@@
    \[ 
-    %\begin{aligned}
     %&\hspace{-0.5cm}
     \prob{R>\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}}
-    = \expect{1\wedge\left(\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}\right)^{-1}}\\
+    = \expect{1\wedge\left(\frac{z_\jmath}{V_\jmath}\vee\frac{z_\ell}{V_\ell}\right)^{-1}}
     %&
@@
     = (1 - u)\expect{\frac{V_\jmath}{\expect{V_\jmath}}\wedge\frac{V_\ell}{\expect{V_\ell}}}
-    %\end{aligned}
     \]
@@
 \subsubsection{Inference for the projected Gamma mixture model}
-In a sample-based inference approach, for a given iteration the Dirichlet process mixture model 
+
+To perform inference for our proposed PoT model we develop a iterative 
+    sample-based approach.  We implement a Markov chain Monte Carlo method that, for a given iteration,  
     groups observations into stochastically assigned clusters, where members of a cluster share 
-    distributional parameters \citep{muller2015,ascolani2022}. Building out the methods of inference for Equation~\eqref{eqn:dppg}, 
-    let $n_j^{-i}$ be the number of observations in cluster $j$ not including observation $i$.  Let $J^{-i}$ be 
+    distributional parameters \citep{muller2015,ascolani2022}. Building out the methods of inference for 
+    Equation~\eqref{eqn:dppg}, 
+    let $n_j^{(-i)}$ be the number of observations in cluster $j$ not including observation $i$.  Let $J^{(-i)}$ be 
     the number of extant clusters, not including any singleton containing observation $i$. Under this model, 
@@
     \text{Pr}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
-        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\\ % \hspace{0.5cm} \\ % &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
+        n_j^{(-i)}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\\ % \hspace{0.5cm} \\ % &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
         \eta\int\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)dG_0(\bm{\alpha}_j,\bm{\beta}_j),%\hspace{0.5cm} % &\text{~for~}j = J^{-i} + 1.
@@
     \]
-    where the top case is iterating over extant clusters $j = 1,\ldots, J^{-i}$, and the bottom case is
+    where the top case is iterating over extant clusters $j = 1,\ldots, J^{(-i)}$, and the bottom case is
     for a \emph{new} cluster. If $G_0$ is not a conjugate prior for the kernel density, the integral in the 
@@
     by Monte Carlo integration, we draw $m$ candidate clusters, $\bm{\alpha}_j,\bm{\beta}_j$ for
-    $j = J^{-i} + 1,\ldots, J^{-i} + m$ from $G_0$. Then, we sample the cluster indicator 
+    $j = J^{(-i)} + 1,\ldots, J^{(-i)} + m$ from $G_0$. Then, we sample the cluster indicator 
     $\gamma_i$ from extant or candidate clusters, where the probability of cluster membership is proportional to
@@
     \text{Pr}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
-        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\\ % \hspace{0.5cm}  % &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
+        n_j^{(-i)}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\\ % \hspace{0.5cm}  % &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
         \frac{\eta}{m}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right). % \hspace{0.5cm} &\text{~for~}j \in \lbrace J^{-i} + 1,\ldots, J^{-i} + m\rbrace.
@@
 %{\bf you first have to say that you sample from this distribution. What is the ``candidate cluster''?}
-    then $\gamma_i = J = J^{- i} + 1$, and the associated cluster parameters are saved.
+    then $\gamma_i = J = J^{(- i)} + 1$, and the associated cluster parameters are saved.
 
@@
 indicating that with appropriate choice of centering distribution, the full conditionals 
-for $\bm{\alpha}_j,\bm{\beta}_j$ become separable by dimension.
+for $\bm{\alpha}_j,\bm{\beta}_j$ can become separable by dimension, and thus inference on
+$\alpha_{j\ell},\beta_{j\ell}$ can be done in parallel for all $j$, $\ell$.
 We first consider a centering distribution given by a product of independent Gammas:
@@
 We also assume a Gamma prior on $\eta$, that is updated via the procedure outlined in \cite{escobar1995}.  
-We refer to this model as the \emph{projected gamma--gamma} (PG--G) model.  An advantage of the PG--G model 
-is that, thanks to conjugacy, the rate parameters can easily be integrated out for inference on $\bm{\alpha}_j$.  
-Then, the full conditional for $\alpha_{j\ell}$ takes the form
+We refer to this model as the \emph{projected gamma--gamma} (PG-G) model.  An advantage of the PG-G model 
+is that, thanks to conjugacy, the rate parameters $\beta_{j\ell}$ can easily be integrated out for inference 
+on $\bm{\alpha}_j$.  Then, the full conditional for $\alpha_{j\ell}$ takes the form
 \begin{equation}
@@
 inference on $\bm{\zeta},\bm{\sigma}$.  We refer to this model as the 
-\emph{projected restricted gamma--gamma} (PRG--G) model.
+\emph{projected restricted gamma--gamma} (PRG-G) model.
 
@@
     $\bm{\Sigma}$, and Gamma priors on $\zeta_{\ell}$ and $\sigma_{\ell}$, and $\eta$.  
-    This model is denoted as the \emph{projected gamma--log-normal} (PG--LN) model.  
+    This model is denoted as the \emph{projected gamma--log-normal} (PG-LN) model.  
     We also explore a restricted Gamma form of this model as above, where 
     $\beta_{\ell} = 1$ for all $\ell$.  This is denoted as the 
-    \emph{projected restricted gamma--log-normal} (PRG--LN) model.  Updates for 
+    \emph{projected restricted gamma--log-normal} (PRG-LN) model.  Updates for 
     $\bm{\alpha}$ can be accomplished using a joint Metropolis step, where 
@@
     \]
-    The inferential forms for $\beta_{j\ell}$ and its priors are the same as for PG--G.  The 
+    The inferential forms for $\beta_{j\ell}$ and its priors are the same as for PG-G.  The 
   normal prior for $\mu$ is conjugate for the log-normal $\bm{\alpha}_j$, and can be sampled

### 2024-04-10 results.tex
entropy submission RC1
push for mdpi resubmission

@@
-\begin{figure*}[htb]
-    \centering
-    \caption{Average energy score rise over baseline (on $\mathbb{S}_{\infty}^{d-1}$) for various 
-    models fitted to simulated data, with ascending count of mixture components (indicated by plot
-    heading) and number of dimensions (indicated by horizontal axis).  Note that pairwise betas is
-    a moment-restricted model.\label{fig:simpples}}
-    \includegraphics[height=3in, width = \textwidth]{./images/sim_es_rise}
-\end{figure*}
-
 \section{Data illustrations\label{sec:results}} 
@@
 
-Our hyperprior parameters are set as follows: for the gamma-centered models (PG--G, PRG--G), 
+Our hyperprior parameters are set as follows: for the gamma-centered models (PG-G, PRG-G), 
     the shape parameter for the centering 
@@
     rate parameter~$\tau_{\ell}\sim\mathcal{G}\left(2,2\right)$.
-    For the log-normal centered models (PG--LN, PRG--LN), the centering distribution's
+    For the log-normal centered models (PG-LN, PRG-LN), the centering distribution's
     log-mean $\bm{\mu}\sim\mathcal{N}_d\left(0,\bm{I}_d\right)$, and covariance 
@@
     identity matrix.
-    For models learning rate parameters $\beta_{j\ell}$ (PG--G, PG--LN), the centering
+    For models learning rate parameters $\beta_{j\ell}$ (PG-G, PG-LN), the centering
     distribution's shape parameter $\zeta\sim\mathcal{G}\left(1,1\right)$ and rate parameter
@@
     0 for numerical stability.
-
-\subsection{Simulation Study\label{subsec:simulated}}
-To evaluate our proposed approach for angular measure estimation 
-    we consider simulated datasets on $\mathbb{S}_{\infty}^{d-1}$,
-    for values of $d$ ranging between 2 and 32.  We generated each
-    dataset as a mixture of multivariate log-normal distributions,
-    projected onto $\mathbb{S}_{\infty}^{d-1}$. The generation 
-    procedure is detailed in Algorithm~\ref{algo:simulated}. We produced 
-    ten replicates of each configuration. We consider two gamma-centered
-    and two log-normal centered DP mixture models, with and without restrictions in each
-    case. To perform a comparative analysis we fitted 
-   % \makenote{Comparative analysis using posterior predictive samples necessitates
-    %alternative models having defined methods of sampling from their respective posterior
-    %predictive distributions, for models defined on some normalization of the positive cone.  
-    %To that end, we also consider the Dirichlet distribution, as well as} 
-    %We also consider a Dirichlet distribution and 
-    the pairwise betas model proposed in \cite{COOLEY2010}. We chose this model for comparison
-    as it is implemented in the readily available package \verb|BMAmevt| 
-    in R~\citep{BMAmevt}. \verb|BMAmevt| provides samples from the predictive
-    posterior distribution. These are needed for the calculation of the energy
-    scores that are at the basis of our comparison. In addition, \verb|BMAmevt| can 
-    be fitted to moderately large multivariate observations. For the DP mixture
-    models, the data are projected onto $\mathbb{S}_{10}^{d-1}$. For the other
-    two models they are projected on $\mathbb{S}_1^{d-1}$.   We sampled each model for 50,000 
-    iterations, dropping the first 40,000 as burn-in, and thinning to keep every 10th iteration
-    after.  These settings were intended to provide a consistent sampling strategy that would
-    work with every model, even if inefficient for some.
 
@@
     \end{algorithm}
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[height=2.5in, width = 0.9\textwidth]{./images/sim_es_rise}
+    \caption{Average energy score rise over baseline (on $\mathbb{S}_{\infty}^{d-1}$) for various 
+    models fitted to simulated data, with ascending count of mixture components (indicated by plot
+    heading) and number of dimensions (indicated by horizontal axis).  Note that pairwise betas is
+    a moment-restricted model.\label{fig:simpples}}
+\end{figure}
+
+\subsection{Simulation Study\label{subsec:simulated}}
+The challenging problem in multivariate EVT is to capture the dependence 
+    structure of the limiting distribution.  To this end, we focus our 
+    simulation study specifically on the angular component.  To evaluate 
+    our proposed approach for angular measure estimation 
+    we consider simulated datasets on $\mathbb{S}_{\infty}^{d-1}$,
+    for values of $d$ ranging between 2 and 32.  We generated each
+    dataset as a mixture of multivariate log-normal distributions,
+    projected onto $\mathbb{S}_{\infty}^{d-1}$. The generation 
+    procedure is detailed in Algorithm~\ref{algo:simulated}. We produced 
+    ten replicates of each configuration. We consider two gamma-centered
+    and two log-normal centered DP mixture models, with and without restrictions in each
+    case. To perform a comparative analysis we fitted 
+   % \makenote{Comparative analysis using posterior predictive samples necessitates
+    %alternative models having defined methods of sampling from their respective posterior
+    %predictive distributions, for models defined on some normalization of the positive cone.  
+    %To that end, we also consider the Dirichlet distribution, as well as} 
+    %We also consider a Dirichlet distribution and 
+    the pairwise betas model proposed in \cite{COOLEY2010}. We chose this model for comparison
+    as it is similarly works to capture a complex dependence structure on
+    an $\mathbb{S}_p^{d-1}$ sphere, albeit with $p = 1$, and is implemented in the readily 
+    available package \verb|BMAmevt| in R~\citep{BMAmevt}, which can provide samples from the 
+    posterior predictive distribution. These samples are needed for the calculation of the energy
+    scores that are at the basis of our comparison. In addition, \verb|BMAmevt| can 
+    be fitted to moderately large multivariate observations. For the DP mixture
+    models, the data are projected onto $\mathbb{S}_{10}^{d-1}$. For the other
+    two models they are projected on $\mathbb{S}_1^{d-1}$.   We sampled each model for 50,000 
+    iterations, dropping the first 40,000 as burn-in, and thinning to keep every 10th iteration
+    after.  These settings were intended to provide a consistent sampling strategy that would
+    work with every model, even if inefficient for some.
 
@@
     computational cost of the pairwise betas model grows combinatorially, with a 
-    sample space of dimension $\binom{d}{2} + 1$. In our testing, \verb|BMAmevt| is 
-    substantially faster than any of our proposed DP mixture models for low dimensions, 
-    however for examples with high number of dimensions, the computational time for 
-    \verb|BMAmevt| was comparable or greater than that for DP-PG-G.  We compare computing
+    sample space of dimension $\binom{d}{2} + 1$.  By comparison, the sample space for 
+    PG-G and PRG-G are $2(J+1)d$ and $(J+1)d$ respectively, where $J$ is the number of extant
+    clusters, with much of that inference able to be done in parallel.  
+    In our testing, for low-dimensional problems, \verb|BMAmevt| was 
+    substantially faster than any of our proposed DP mixture models.  However, for examples with 
+    high numbers of dimensions, the computational time for 
+    \verb|BMAmevt| was greater than that for PG-G.  We compare computing
     times in our data analysis in Table~\ref{tab:sampletime}.
@@
 \begin{figure}[tb]
-    \caption{Grid cell locations for ERA-Interim (left) and ERA5 (right).\label{fig:gridlocs}}
     \centering
@@
     \end{minipage}
-    % \caption{Grid cell locations for ERA-Interim (left) and ERA5 (right).\label{fig:gridlocs}}
-    % \begin{center}
-    % \begin{subfigure}{0.48\textwidth}
-    %     \centering 
-    %     \includegraphics[height=2.5in]{./images/erai_grid}
-    % \end{subfigure}
-    % %
-    % \begin{subfigure}{0.48\textwidth}
-    %     \centering
-    %     \includegraphics[height=2.5in]{./images/era5_grid}
-    % \end{subfigure}
-    % \end{center}
+    \caption{Grid cell locations for ERA-Interim (left) and ERA5 (right).\label{fig:gridlocs}}
 \end{figure}
@@
 
-\begin{figure*}[ht]%{l}{0.45\textwidth}
+\begin{figure}[ht]%{l}{0.45\textwidth}
     \centering
+    \includegraphics[width=.7\linewidth]{./images/data_transformed}
     \caption{Pairwise plots from ERA-Interim data after transformation and projection to ${\mathbb S}_{\infty}^{7}$.
@@
 All data is between 0 and 1.\label{fig:erai_data}}
-    \includegraphics[width=.7\linewidth]{./images/data_transformed}
-\end{figure*}
+\end{figure}
 
@@
   ERA5 data, after subsetting we have $6342$ observations, which reduces to $532$ observations 
-  after processing and declustering. We fit the PG--G, PRG--G, PG--LN, and PRG--LN models to 
+  after processing and declustering. We fit the PG-G, PRG-G, PG-LN, and PRG-LN models to 
   both datasets.
 
-\begin{table}[htb]
+% \begin{table}[htb]
+%   \centering
+%   \caption{Energy score criterion from fitted models against the IVT data. 
+%     Lower is better.\label{tab:dev}}
+%   \include{./tables/table_dev_updated}
+% \end{table}
+
+% \begin{table}[htb]
+%   \centering
+%   \caption{Time to sample (in minutes) 50,000 iterations for various models\label{tab:sampletime}}
+%   \include{./tables/table_sampletime}
+% \end{table}
+
+\begin{table}[t]
     \centering
-  \caption{Energy score criterion from fitted models against the IVT data. 
-    Lower is better.\label{tab:dev}}
+    \caption{Model fit assessment and computation time on ERA-Interim and ERA5 data.}
+    \begin{subtable}{0.7\textwidth}
+    \centering
     \include{./tables/table_dev_updated}
-\end{table}
-
-\begin{table}[htb]
+    \caption{Energy score criterion from fitted models against the IVT data. Lower is better.\label{tab:dev}}
+    \end{subtable}
+    \\
+    \bigskip
+    \begin{subtable}{0.7\textwidth}
     \centering
+    \include{./tables/table_sampletime}
     \caption{Time to sample (in minutes) 50,000 iterations for various models\label{tab:sampletime}}
-  \include{./tables/table_sampletime}
+    \end{subtable}%
 \end{table}
@@
     two datasets.  We see the effect of dimensionality on the various models; for gamma centered
-    models it grows linearly; for the log--normal centered model, it will grow superlinearly as matrix
+    models it grows linearly; for the log-normal centered model, it will grow superlinearly as matrix
     inversion becomes the most costly operation.  For \verb|BMAmevt|, its parameter space grows
@@
 
-\begin{figure*}[htb]
+\begin{figure}[b]
     \centering
-    \caption{Pairwise extremal dependence coefficients for IVT data using the PRG--G model.\label{fig:chi_ij}}
     \includegraphics[width=0.9\textwidth]{./images/chi_ij_c}
-\end{figure*}
+    \caption{Pairwise extremal dependence coefficients for IVT data using the PRG-G model.\label{fig:chi_ij}}
+\end{figure}
 
 We consider an exploration of the pairwise extremal dependence using Monte Carlo estimates of the 
-  coefficients in  Equation~\eqref{eqn:chi_ij}. For this we use samples obtained from the PRG--G model.
+  coefficients in  Equation~\eqref{eqn:chi_ij}. For this we use samples obtained from the PRG-G model.
   Figure~\ref{fig:chi_ij} provides a graphical analysis of the results. 
@@
 
-\begin{figure*}[htb]
+\begin{figure}[t]
     \centering
+    \includegraphics[width=\linewidth]{./images/condsurv_1d}
     \caption{Conditional survival curves for selected locations, using ERA-Interim, and PRG-G model,  conditioning on all other dimensions at greater than 90th percentile (fitted)\label{fig:condsurv1d}.
@@
 Right panel uses standardized units.}
-    \includegraphics[width=\linewidth]{./images/condsurv_1d}
-\end{figure*}
+\end{figure}
 
-\begin{figure*}[htb]
+\begin{figure}[t]
     \centering
+    \includegraphics[height=4in, width=4in]{./images/condsurv_2d}
     \caption{Pairwise conditional survival curves for selected locations, using 
@@
         than 90th percentile (fitted).\label{fig:condsurv2d}}
-    \includegraphics[height=4in, width=4in]{./images/condsurv_2d}
-\end{figure*}
+\end{figure}
 
-Figure~\ref{fig:condsurv1d} shows, for the ERA-Interim data under the PRG--G model,  
+Figure~\ref{fig:condsurv1d} shows, for the ERA-Interim data under the PRG-G model,  
     the conditional survival curve defined in Equation~\eqref{eqn:condsurv}, for one dimension, 
